### 2021-08-15 paper/paper.Rmd
start outline

---
title: Individual Injury and Representational Harms (?)
authors:
  - name: Cory McCartan
    thanks: Footnote stuff
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Enter the text of your abstract here.
keywords:
  - partisan fairness
  - representation
  - redistricting simulation
  - standing
bibliography: references.bib
biblio-style: unsrt
output: 
  rticles::arxiv_article:
    keep_tex: true
---

# Introduction

- introduce partisan symmetry
  - Katz, King, Rosenblatt 2019; Deford et al 2021; Katz, King, Rosenblatt 2021

- Election law side is very indivudalized
  - Standing discussions 
    - Lujan v Defenders of Wildlife (1992)
    - Shaw v. Reno (1993) -- racial gerrymandering
    - Issacharoff and Karlan 1998

- Recent work
  - EGap
  - Partisan Dislocation

- Power of Ensemble Analysis
  - Recent advances allow us to make claims about harm based in counterfactuals that are realistic and statistically useful

# Towards Individualized Representational Harm

# Connections to the Partisan Symmetry Standard

# Measures of Individual Injury

# Differences Between Classes of Fairness Metrics

# Discussion

### 2021-08-16 R/toy/00_toy_setup.R
up

@@ -1,0 +1,12 @@
+# libs ----
+library(sf)
+library(tidyverse)
+library(redist)
+
+# make grid ----
+bb <-st_sfc(st_polygon(list(rbind(c(0,0), c(1,0), c(1,1), c(0,0)))))
+grid <- st_make_grid(bb, n = 4)
+tb <- st_as_sf(grid)
+
+# add votes ----
+

### 2021-08-21 R/00_setup_utils.R
progress

@@ -3,5 +3,34 @@ library(geomander)
 library(redist)
 library(scales)
 library(wacolors)
+library(here)
 
 theme_repr = function() theme_bw(base_family="Times New Roman", base_size=11)
+
+# downloads data for state `abbr` to `folder/{abbr}_2020_*.csv` and returns path to file
+download_redistricting_file = function(abbr, folder) {
+    abbr = tolower(abbr)
+    url_vtd = paste0("https://raw.githubusercontent.com/alarm-redist/census-2020/",
+                     "main/census-vest-2020/", abbr, "_2020_vtd.csv")
+    url_block = paste0("https://raw.githubusercontent.com/alarm-redist/census-2020/",
+                       "main/census-vest-2020/", abbr, "_2020_block.csv")
+
+    path = paste0(folder, "/", basename(url_vtd))
+    resp = download.file(url_vtd, path)
+    if (resp != 0) {
+        path = paste0(folder, "/", basename(url_block))
+        resp = download.file(url_block, path)
+        if (resp != 0)  {
+            stop("No files available for ", abbr)
+        }
+    }
+    path
+}
+
+# adds precinct shapefile geometry to downloaded data
+join_shapefile = function(data) {
+    geom_d = PL94171::pl_get_vtd(data$state[1]) %>%
+        select(GEOID20, area_land=ALAND20, area_water=AWATER20, geometry)
+    left_join(data, geom_d, by="GEOID20") %>%
+        sf::st_as_sf()
+}

### 2021-08-21 R/02_sim_iowa.R

@@ -1,0 +1,47 @@
+iowa = read_rds(here("data/IA_cd_final_vtd_20.rds")) %>%
+    redist_map(pop_tol=0.001, ndists=4, adj=.$adj)
+
+N_sim = 5e3
+plans = redist_smc(iowa, N_sim, verbose=FALSE)
+#pl_opt_dem = redist_shortburst(iowa, scorer_group_pct(iowa, ndv, ndv+nrv, 2))
+opt_dem = c(4, 4, 2, 4, 4, 4, 2, 2, 2, 3, 2, 2, 2, 2, 4, 3, 2, 4, 2, 4, 2, 2, 3, 2, 1, 4, 4, 2, 3, 2, 2, 2, 2, 2, 2, 4, 2, 2, 1, 2, 2, 2, 4, 3, 2, 2, 4, 4, 3, 1, 4, 3, 3, 4, 2, 4, 3, 3, 4, 2, 4, 4, 4, 1, 4, 2, 4, 4, 4, 3, 4, 2, 4, 2, 4, 2, 1, 4, 1, 4, 2, 3, 4, 4, 1, 4, 4, 4, 4, 4, 4, 4, 4, 2, 2, 2, 4, 2, 2)
+#pl_opt_rep = redist_shortburst(iowa, scorer_group_pct(iowa, nrv, ndv+nrv, 4))
+opt_rep = c(4, 1, 3, 1, 4, 3, 2, 4, 2, 3, 2, 2, 4, 4, 4, 3, 2, 4, 3, 1, 2, 3, 3, 4, 4, 1, 1, 3, 2, 2, 3, 2, 3, 3, 2, 1, 4, 2, 4, 4, 4, 2, 4, 2, 3, 4, 4, 3, 3, 4, 1, 2, 3, 2, 2, 2, 3, 2, 1, 2, 1, 1, 1, 4, 1, 2, 4, 1, 1, 2, 2, 2, 1, 4, 4, 4, 1, 4, 2, 1, 4, 3, 4, 2, 4, 2, 1, 4, 1, 1, 1, 2, 1, 4, 2, 3, 4, 2, 4)
+
+plans = plans %>%
+    subset_sampled() %>%
+    add_reference(opt_dem, "dem_gerry") %>%
+    add_reference(opt_rep, "rep_gerry") %>%
+    mutate(dev = plan_parity(iowa),
+           comp = distr_compactness(iowa),
+           dem = group_frac(iowa, ndv, ndv + nrv),
[…]

### 2021-08-24 paper/paper.Rmd
up
progress
Update paper.Rmd

@@
 ---
-title: Individual Injury and Representational Harms (?)
+title: Individual Representation: Harm and Utility(?)
 authors:
@@
 
+Understanding fairness in electoral systems has long suffered from a focus on parties over the individuals who vote for them.
+There is no doubt that parties are interesting units in electoral discussions.
+They offer stability over candidates, information for the under-informed, and shortcuts for understanding preferred vote choices.
+Yet, court histories and academic traditions value the harm of the individual over the harm to a transient political body, such as the party.
+
+In this article, we offer an approach to understanding partisan fairness through a lens of individual harm and utility.
+This does not supplant existing understandings of partisan fairness, but instead offers an alternative rooted in understanding how the individual potential voters are impacted by districting plans.
+
+Several existing methods of redistricting simulation rely on randomness, such as Random Seed and Grow (Chen and Rodden 2013) or Recombination (Deford et al. 2021).
+Moving beyond these hueristic methods, we can use sampling methods that have defined target distributions to ensure that we characterize the space of applicable plans properly.
+Methods which operationalize a statistics-based approach include Fifield et al. 2020, McCartan and Imai (2020), and Autrey et al (2020).
+Using these methods advance us beyond the critiques of Katz, King, and Roseblatt (2020) by sampling plans using weights within a Gibbs distribution, rather than assuming that uniform sampling weights alone are sufficient.
+
 -   introduce partisan symmetry
+
     -   Katz, King, Rosenblatt 2019; Deford et al 2021; Katz, King, Rosenblatt 2021
@@
 -   Election law side is very indivudalized
+
     -   Standing discussions
+
         -   Lujan v Defenders of Wildlife (1992)
@@
 
-- Recent work
+-   Recent, related work
+
     -   EGap
@@
 
+-   Important to consider benefits and harms to individual voters
+
 -   Power of Ensemble Analysis
+
     -   Recent advances allow us to make claims about harm based in counterfactuals that are realistic and statistically useful
+
+# Voter Utility in Election Outcomes
+
+1.
 

### 2021-08-26 paper/paper.Rmd

@@
 Several existing methods of redistricting simulation rely on randomness, such as Random Seed and Grow (Chen and Rodden 2013) or Recombination (Deford et al. 2021).
-Moving beyond these hueristic methods, we can use sampling methods that have defined target distributions to ensure that we characterize the space of applicable plans properly.
+Moving beyond these heuristic methods, we can use sampling methods that have defined target distributions to ensure that we characterize the space of applicable plans properly.
 Methods which operationalize a statistics-based approach include Fifield et al. 2020, McCartan and Imai (2020), and Autrey et al (2020).
 Using these methods advance us beyond the critiques of Katz, King, and Roseblatt (2020) by sampling plans using weights within a Gibbs distribution, rather than assuming that uniform sampling weights alone are sufficient.
+
+Metrics often associated with partisan symmetry, such as TODO, are indicative of violations of partisan symmetry.
+Notably, these are only defined for full legislatures (see Katz, King, and Rosenblatt (2021)).
+As such, theory-based interpretations of these measures do not indicate signs of unfairness to parities in the congressional district setting, unless they are consistent across many, many states.
+We offer measures of partisan harm which are rooted directly in the representational harms inflicted upon individuals by those who choose to draw a amap.
+
+These measures are deeply connected to the law and constitutional understandings of partisan fairness, due to their related roots in the individual.
 

### 2021-08-29 paper/paper.Rmd
section 2 drafted

@@
 ---
-title: Individual Representation: Harm and Utility(?)
+title: Individualized Measures of Partisan Fairness and Utility in Redistricting (??)
 authors:
@@
     keep_tex: true
+    includes:
+      in_header: "header.tex"
+    latex_engine: pdflatex
 ---
+
+**Title suggestions**
+
+-   Individual Representation: Harm and Utility
+
+-   Redistricting and Individual Representational Harm
+
+-   Fairness and Utility in Redistricting: an Individual Perspective
+
+-   Individualized Measures of Fairness and Utility in Redistricting
 
@@
 
-1.
+The utility voters gain from a particular election outcome naturally varies across the electorate, depending on factors such as political engagement, partisanship, race and ethnicity, individual policy preferences, individual candidate attributes, and more.
+Here, we restrict our attention to what we believe are the two dominant contributors to partisan utility: which party represents a voter's district, and how many overall seats a voter's preferred party wins.
+
+## Notation and Setup
+
+Let $V$ be the set of voters in the region and time period under consideration, who can vote in a set of legislative elections $E$.
+Each election involves the selection of one of a set of political parties $P$ to represent each of a set of legislative districts, indexed by $D$.
+We take these four sets to be fixed (which does not require that voters do not move, or that parties contest every election).
+
+For each election $e\in E$, every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is formally represented by a function $\xi_e:V\surjto D$.
+The set of all possible districting plans is $\Xi_e$.
+
+Voters either vote for a candidate running in their district, or abstain from voting.
+Notationally, we have a set of functions $\vote_e:V\to P\cup\{\textsc{abstain}\}$.
+Votes may be tallied at the district level, either in total or by party: \begin{align*}
+    \votes_e&: D\to \N, \qquad 
+    \votes_e(d) = \sum_{v\in\xi_e^{-1}(d)} \ind\{\vote_e(v)\in P\} \\
+    \votes_e^{(p)}&: D\to \N, \qquad 
+    \votes_e^{(p)}(d) = \sum_{v\in\xi_e^{-1}(d)} \ind\{\vote_e(v) = p\}.
+\end{align*} The political party of the candidate with the most votes wins each district, as captured by the function $$
+    \win_e: D\to P, \qquad \win_e(d) = \argmax_{p\in P} \votes_e^{(p)}(d).
+$$Thus an individual voter $v\in V$ is represented after election $e$ by political party $\win_e\circ\xi_e(v)$, and each political party can be said to win a number of seats, formalized as$$
+\seats_e:P\to\N, \qquad \seats_e(p) = \sum_{d\in D} \ind\{\win_e(d) = p\}.
+$$
+
+## Global Outcome Utility
+
+We posit that a utility of the global election outcome to a voter is an increasing function of the share of total seats that the voter's preferred party wins.
+Their global outcome utility function for election $e\in E$ can thus be written$$
+u_e^{(G)}:V\to\R, \qquad u_e^{(G)}(v) = h\qty(\frac{\seats_e\circ\vote_e(v)}{|D|}),
+$$
+
+for $h:[0,1]\bijto\R$ an increasing function.
+
+What kind of electoral systems does this utility function encourage, if the goal is maximizing total voter utility?
+It depends on the form of $h$, as the following result describes.
+
+\begin{prop}
+
+Let $\{s_p:p\in P\}$ be a set of party popular vote shares, and $f:[0,1]\bijto[0,1]$ be a rule for assigning each party a share of seats according to their share of the overall vote.
+Define the total utility for this rule as $$
+U_e^{(G)}(f) = \sum_{v\in V} h\qty(f\qty(s_{\vote_e(v)})).
+$$Then if $h$ is convex, $U_e^{(G)}$ is maximized for $f(x)=\ind\{x\ge \half\}$, and if $h$ is concave, $U_e^{(G)}$ is maximized by an increasing $f$.
+In particular, if $h(x)=\log(x)$ and every party receives at least one vote, then $f(x)=x$ maximizes $U_e^{(G)}$.
+
+\end{prop}
+
+Thus, if we are willing to adopt the logarithm as the functional specification of $u_e^{(G)}$, then this global outcome utility measure will encourage districting systems that create proportionality between party vote shares and seat shares.
+
+## Local Outcome Utility
+
+The specification of the utility of the local election outcome is more straightforward.
+We define$$
+u_e^{(L)}:V\to\R, \qquad u_e^{(L)}(v) = \ind\{\win_e\circ\xi_e(v) = \vote_e(v)\},
+$$that is, a voter gains utility only if they winner of their legislative district belongs to their preferred political party.
+
+Attempting to maximize the total local utility across the electorate would lead to a districting plan where every voter is assigned to a district with other voters of the same party preference; in this way, every voter's utility would be $1$.
+Whether such an districting plan is possible depends on the geographical layout of voters and the legal constraints on the districting process---must districts be contiguous, balanced in population, constructed from whole precincts, geographically compact, and so on.
+But the general tendency of this utility measure (in the context of a single election, at least), like the global utility measure, will be to push plans towards proportionality.
+The local measure will additionally reward what have been called *bipartisan gerrymanders*, where there are few or no competitive districts.
 
@@
 # Discussion
+
+\appendix
+
+# Proofs of Propositions

### 2021-08-29 paper/paper.Rmd
add appendix proof

@@
 
-\begin{prop}
-
+```{=tex}
+\begin{restatable}{prop}{propglobal} \label{prop:global}
 Let $\{s_p:p\in P\}$ be a set of party popular vote shares, and $f:[0,1]\bijto[0,1]$ be a rule for assigning each party a share of seats according to their share of the overall vote.
-Define the total utility for this rule as $$
+Define the total utility for this rule as \[
     U_e^{(G)}(f) = \sum_{v\in V} h\qty(f\qty(s_{\vote_e(v)})).
-$$Then if $h$ is convex, $U_e^{(G)}$ is maximized for $f(x)=\ind\{x\ge \half\}$, and if $h$ is concave, $U_e^{(G)}$ is maximized by an increasing $f$.
+\] 
+Then if $h$ is convex, $U_e^{(G)}$ is maximized for $f(x)=\ind\{x\ge \half\}$, and if $h$ is concave, $U_e^{(G)}$ is maximized by an increasing $f$.
 In particular, if $h(x)=\log(x)$ and every party receives at least one vote, then $f(x)=x$ maximizes $U_e^{(G)}$.
 
-\end{prop}
-
+\end{restatable}
+```
 Thus, if we are willing to adopt the logarithm as the functional specification of $u_e^{(G)}$, then this global outcome utility measure will encourage districting systems that create proportionality between party vote shares and seat shares.
@@
 # Proofs of Propositions
+
+```{=tex}
+\propglobal*
+
+\begin{proof}
+\end{proof}
+```

### 2021-08-29 paper/paper.Rmd
fix math error

@@
 \begin{restatable}{prop}{propglobal} \label{prop:global}
-Let $\{s_p:p\in P\}$ be a set of party popular vote shares, and $f:[0,1]\bijto[0,1]$ be a rule for assigning each party a share of seats according to their share of the overall vote.
+Let $S=\{s_p:p\in P\}$ be a set of party popular vote shares, and $f:[0,1]\bijto[0,1]$ be a rule for assigning each party a share of seats according to their share of the popular vote.
 Define the total utility for this rule as \[
@@
 \] 
-Then if $h$ is convex, $U_e^{(G)}$ is maximized for $f(x)=\ind\{x\ge \half\}$, and if $h$ is concave, $U_e^{(G)}$ is maximized by an increasing $f$.
+Then if $h$ is convex, $U_e^{(G)}$ is maximized for $f(x)=\ind\{x\ge \max S\}$, and if $h$ is concave, the maximizer of $U_e^{(G)}$ will be an increasing function.
 In particular, if $h(x)=\log(x)$ and every party receives at least one vote, then $f(x)=x$ maximizes $U_e^{(G)}$.

### 2021-08-30 paper/paper.Rmd
mini schematic

@@
 ---
+
+```{r SETUP, include=FALSE}
+library(knitr)
+library(here)
+
+opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE,
+               fig.path=here("paper/figures/"), fig.align="center",
+               fig.width=(8.5-2*1)/2, out.width="100%", fig.asp=0.8)
+```
 
@@
 
-# Voter Utility in Election Outcomes
+## Running Examples
 
-The utility voters gain from a particular election outcome naturally varies across the electorate, depending on factors such as political engagement, partisanship, race and ethnicity, individual policy preferences, individual candidate attributes, and more.
+To illustrate the notation and measures we introduce, this article will make use of three running examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichussets.
+Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
+Schematics for these examples are shown in Figure \ref{fig:mini-schematic}.
+
+```{r mini-schematic, fig.cap="Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts. The total voting population for each example is 180."}
+include_graphics(here("paper/figures/minis_schematic.pdf"))
+```
+
+## Outline of the Paper
+
+Section \ref{sec:voter-util} provides the necessary notation and introduces our measures of voter utility.
+Section \ref{sec:indiv-harm} builds a measure of individual harm from this notion of utility.
+Section \ref{sec:empirical} applies this measure to a real-world example.
+Section \ref{sec:discussion} concludes.
+
+# Voter Utility in Election Outcomes {#sec:voter-util}
+
+The utility voters gain from a particular election outcome naturally varies across the electorate, depending on factors such as political engagement, partisanship, race and ethnicity, individual policy preferences, and individual candidate attributes.
 Here, we restrict our attention to what we believe are the two dominant contributors to partisan utility: which party represents a voter's district, and how many overall seats a voter's preferred party wins.
@@
 \] 
-Then if $h$ is convex, $U_e^{(G)}$ is maximized for $f(x)=\ind\{x\ge \max S\}$, and if $h$ is concave, the maximizer of $U_e^{(G)}$ will be an increasing function.
+Then if $h$ is convex, $U_e^{(G)}$ is maximized for $f(x)=\ind\{x\ge \max S\}$, and if $h$ is strictly concave, the maximizer of $U_e^{(G)}$ will be a strictly increasing function.
 In particular, if $h(x)=\log(x)$ and every party receives at least one vote, then $f(x)=x$ maximizes $U_e^{(G)}$.
@@
 
-# Towards Individualized Representational Harm
+# Towards Individualized Representational Harm {#sec:indiv-harm}
 
@@
 
-# Measures of Individual Injury
-
 # Differences Between Classes of Fairness Metrics
 
-# Discussion
+# Real-world Example {#sec:empirical}
+
+# Discussion {#sec:discussion}
 

### 2021-08-30 R/00_setup_utils.R

@@ -4,10 +4,12 @@ library(geomander)
 library(redist)
 library(sf)
 library(scales)
+library(patchwork)
 library(wacolors)
 library(here)
 
-theme_repr = function() theme_bw(base_family="Times New Roman", base_size=11)
+theme_repr = function() theme_bw(base_family="Times", base_size=10)
+theme_repr_map = function() theme_void(base_family="Times", base_size=10)
 
 GOP_DEM = c("#A0442C", "#B25D4C", "#C27568", "#D18E84", "#DFA8A0",
             "#EBC2BC",  "#F6DCD9", "#F9F9F9", "#DAE2F4", "#BDCCEA",

### 2021-08-30 R/toy/00_toy_setup.R

@@ -40,7 +40,7 @@ mininois = grid %>%
     relocate(geometry, .after=rep) %>%
     redist_map(ndists=3, pop_tol=0.1)
 
-# functions and helpers
+# people for plotting -------
 
 make_people = function(pop, dem, geometry, row, col, ...) {
     url = str_glue("http://hydra.nat.uni-magdeburg.de/packing/csq/txt/csq{pop}.txt")
@@ -64,8 +64,28 @@ minichusetts_people = minichusetts %>%
 minissouri_people = minissouri %>%
     pmap_dfr(make_people)
 
+# functions and helpers -------
+plot_state = function(map, people, qty=NULL, ppl_size=2, ...) {
+    redist.plot.map(map, fill=qty, ...) +
+        scale_fill_party_c() +
+        geom_sf(data=map, size=0.8, fill=NA, color="white") +
+        geom_sf(aes(color=dem), data=people, size=ppl_size) +
+        scale_color_party_d(guide="none") +
+        theme_repr_map() +
+        theme(plot.title = element_text(hjust = 0.5))
+}
+
 sim_toy = function(map, N=500) {
     redist_smc(map, N, verbose=F) %>%
         mutate(comp = distr_compactness(map),
                dem = group_frac(map, dem, pop))
+}
+
+# initial plots -------
+if (!file.exists(fig_path <- here("paper/figures/minis_schematic.pdf"))) {
+    list(plot_state(mininois, mininois_people) + labs(title="(a) Mininois"),
+         plot_state(minissouri, minissouri_people) + labs(title="(b) Minissouri"),
+         plot_state(minichusetts, minichusetts_people) + labs(title="(c) Minichusetts")) %>%
+        wrap_plots(nrow=1)
+    ggsave(fig_path, width=6.5, height=2.5)
 }

### 2021-08-30 paper/paper.Rmd
harm section

@@
   - redistricting simulation
-  - standing
+  - individual standing
 bibliography: references.bib
@@
 As such, theory-based interpretations of these measures do not indicate signs of unfairness to parities in the congressional district setting, unless they are consistent across many, many states.
-We offer measures of partisan harm which are rooted directly in the representational harms inflicted upon individuals by those who choose to draw a amap.
+We offer measures of partisan harm which are rooted directly in the representational harms inflicted upon individuals by those who choose to draw a map.
 
@@
 Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
-Schematics for these examples are shown in Figure \ref{fig:mini-schematic}.
+Each example state has the same population and should be divided into 3 districts with population between 54 and 66.
+Voters' preferences are assumed to be stable across elections, and turnout is 100%.
+Schematics for these examples are shown in Figure \ref{fig:mini-schematic}(a)--(c).
 
@@
 Section \ref{sec:voter-util} provides the necessary notation and introduces our measures of voter utility.
-Section \ref{sec:indiv-harm} builds a measure of individual harm from this notion of utility.
-Section \ref{sec:empirical} applies this measure to a real-world example.
+Section \ref{sec:indiv-harm} builds a measure of individual harm and a derived measure of partisan fairness from this notion of utility.
+Section \ref{sec:estimating} discusses estimating these measures from observed election results.
+Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
+Section \ref{sec:empirical} applies the proposed measures to a real-world example to illustrate their value and these comparsions.
 Section \ref{sec:discussion} concludes.
@@
 
-Let $V$ be the set of voters in the region and time period under consideration, who can vote in a set of legislative elections $E$.
-Each election involves the selection of one of a set of political parties $P$ to represent each of a set of legislative districts, indexed by $D$.
-We take these four sets to be fixed (which does not require that voters do not move, or that parties contest every election).
+Let $\Voters$ be the set of voters in the region and time period under consideration, who vote to select a candidate from one of a set of political parties $\Parties$ to represent each of a set of legislative districts, indexed by $\Distrs$.
+We take these three sets to be fixed.
 
-For each election $e\in E$, every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is formally represented by a function $\xi_e:V\surjto D$.
-The set of all possible districting plans is $\Xi_e$.
+Voters vote in elections, each of which produces a specific outcome $e$ from the set $\Elecs$ of all possible election outcomes.
+A specific election contains information on how each voter voted.
 
-Voters either vote for a candidate running in their district, or abstain from voting.
-Notationally, we have a set of functions $\vote_e:V\to P\cup\{\textsc{abstain}\}$.
+Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is formally represented by a function $q:\Voters\surjto \Distrs$.
+The set of all possible districting plans is $\Plans$.[^1]
+
+[^1]: Note that these districting plans do not depend on the election; as a result, we implicitly assume that voters do not move residences during the period under consideration.
+    This is not a necessary restriction and can be dealt with formally by first mapping voters to precincts (varying by election) and then mapping precincts to districts (constant across elections).
+    Once we aggregate or average across all voters in a party or district, the differences between these two approaches disappear.
+
+In an election, voters either vote for a candidate running in their district, or abstain from voting.
+Notationally, we have a set of functions $\vote:\Voters\times\Elecs\to \Parties\cup\{\textsc{abstain}\}$.
 Votes may be tallied at the district level, either in total or by party: \begin{align*}
-    \votes_e&: D\to \N, \qquad 
-    \votes_e(d) = \sum_{v\in\xi_e^{-1}(d)} \ind\{\vote_e(v)\in P\} \\
-    \votes_e^{(p)}&: D\to \N, \qquad 
-    \votes_e^{(p)}(d) = \sum_{v\in\xi_e^{-1}(d)} \ind\{\vote_e(v) = p\}.
+    \votes&: \Distrs\times\Elecs \to \N, \qquad 
+    \votes(d, e) = \sum_{v\in q^{-1}(d)} \ind_P(\vote(v, e)) \\
+    \votes^{(p)}&: \Distrs\times\Elecs\times\Plans \to \N, \qquad 
+    \votes^{(p)}(d, e, q) = \sum_{v\in q^{-1}(d)} \ind_{\{p\}}(\vote(v, e)).
 \end{align*} The political party of the candidate with the most votes wins each district, as captured by the function $$
-    \win_e: D\to P, \qquad \win_e(d) = \argmax_{p\in P} \votes_e^{(p)}(d).
-$$Thus an individual voter $v\in V$ is represented after election $e$ by political party $\win_e\circ\xi_e(v)$, and each political party can be said to win a number of seats, formalized as$$
-\seats_e:P\to\N, \qquad \seats_e(p) = \sum_{d\in D} \ind\{\win_e(d) = p\}.
+    \win: \Distrs\times\Elecs\times\Plans\to P, \qquad 
+    \win(d, e, q) = \argmax_{p\in\Parties} \votes^{(p)}(d, e, q).
+$$Thus an individual voter $v\in V$ is represented after election $e$ by political party $\win(q(v), e, q)$, and each political party can be said to win a number of seats, formalized as$$
+    \seats:\Parties\times\Elecs\times\Plans\to\N, 
+    \qquad \seats(p,e,q) = \sum_{d\in\Distrs} \ind_{\{p\}}(\win(d,e,q)).
 $$
+
+**[CM: better letter/concept pairings possible?]**
 
@@
 We posit that a utility of the global election outcome to a voter is an increasing function of the share of total seats that the voter's preferred party wins.
-Their global outcome utility function for election $e\in E$ can thus be written$$
-u_e^{(G)}:V\to\R, \qquad u_e^{(G)}(v) = h\qty(\frac{\seats_e\circ\vote_e(v)}{|D|}),
-$$
+Their global outcome utility function for an election outcome $e\in\Elecs$ and districting plan $q\in\Plans$ can thus be written$$
+    u^{(G)}:\Voters\times\Elecs\times\Plans\to\R, 
+    \qquad u^{(G)}(v,e,q) = g\qty(\frac{\seats(\vote(v, e), e, q)}{|\Distrs|}),
+$$for $g:[0,1]\bijto\R$ an increasing function.
 
-for $h:[0,1]\bijto\R$ an increasing function.
-
-What kind of electoral systems does this utility function encourage, if the goal is maximizing total voter utility?
+What kind of electoral systems does this utility function encourage, if the goal is maximizing total voter utility (or, equivalently, average voter utility)?
 It depends on the form of $h$, as the following result describes.
@@
 \begin{restatable}{prop}{propglobal} \label{prop:global}
-Let $S=\{s_p:p\in P\}$ be a set of party popular vote shares, and $f:[0,1]\bijto[0,1]$ be a rule for assigning each party a share of seats according to their share of the popular vote.
+Let $S=\{s_p:p\in\Parties\}$ be a set of party popular vote shares, and $f:[0,1]\bijto[0,1]$ be a rule for assigning each party a share of seats according to their share of the popular vote.
 Define the total utility for this rule as \[
-    U_e^{(G)}(f) = \sum_{v\in V} h\qty(f\qty(s_{\vote_e(v)})).
+    u_{\mathrm{tot}}^{(G)}(f, e) \dfeq \sum_{v\in\Voters} g\qty(f\qty(s_{\vote(v, e)})).
 \] 
-Then if $h$ is convex, $U_e^{(G)}$ is maximized for $f(x)=\ind\{x\ge \max S\}$, and if $h$ is strictly concave, the maximizer of $U_e^{(G)}$ will be a strictly increasing function.
-In particular, if $h(x)=\log(x)$ and every party receives at least one vote, then $f(x)=x$ maximizes $U_e^{(G)}$.
+Then if $g$ is convex, $u_{\mathrm{tot}}^{(G)}$ is maximized for $f(x)=\ind_{[\max S, 1]}(x)$, and if $g$ is strictly concave, the maximizer of $u_{\mathrm{tot}}^{(G)}$ will be a strictly increasing function.
+In particular, if $g(x)=\log(x)$ and every party receives at least one vote, then $f(x)=x$ maximizes $u_{\mathrm{tot}}^{(G)}$.
 
@@
 ```
-Thus, if we are willing to adopt the logarithm as the functional specification of $u_e^{(G)}$, then this global outcome utility measure will encourage districting systems that create proportionality between party vote shares and seat shares.
+Thus, if we are willing to adopt the logarithm as the functional specification of $u^{(G)}$, then this global outcome utility measure will encourage districting systems that create proportionality between party vote shares and seat shares.
+
+*(running examples).*
 
@@
 We define$$
-u_e^{(L)}:V\to\R, \qquad u_e^{(L)}(v) = \ind\{\win_e\circ\xi_e(v) = \vote_e(v)\},
+    u^{(L)}:\Voters\times\Elecs\times\Plans\to\R, 
+    \qquad u^{(L)}(v,e,q) = \ind\{\vote(v, e)=\win(q(v), e, q)\},
 $$that is, a voter gains utility only if they winner of their legislative district belongs to their preferred political party.
 
-Attempting to maximize the total local utility across the electorate would lead to a districting plan where every voter is assigned to a district with other voters of the same party preference; in this way, every voter's utility would be $1$.
+Attempting to maximize the average local utility across the electorate would lead to a districting plan where every voter is assigned to a district with other voters of the same party preference; in this way, every voter's utility would be $1$.
 Whether such an districting plan is possible depends on the geographical layout of voters and the legal constraints on the districting process---must districts be contiguous, balanced in population, constructed from whole precincts, geographically compact, and so on.
@@
 
+*(running examples).*
+
+## Multiple Elections or Districting Plans
+
+The utility measures above are defined for a single election outcome $e\in \Elecs$ and districting plan $q\in\Plans$.
+But often we will want to consider utility averaged over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans.
+
+To allow this, we replace the fixed election districting plan with random variables $(E,Q)\sim \pi$, where $\pi$ is a joint probability measure on election outcomes and districting plans.
+The randomness in election outcomes comes from variations in factors such as voter preferences, the political environment, candidate quality, and campaign events.
+The randomness in districting plans reflects the constraints and preferences of the plan-drawing process.
+As we will discuss in Section \ref{sec:estimating}, generally we will assume that $E\indep Q$, though this is not required for our theoretical analysis.
+We will also consider a voter $V$ taken uniformly at random from $\Voters$, with $V\indep\pi$.
+
+Then we may define the utilities for a (randomly selected) voter as $$
+    U^{(G)} \dfeq u^{(G)}(V, E, Q) \qand
+    U^{(L)} \dfeq u^{(L)}(V, E, Q).
+$$Thus, the utility for a single voter $v$ becomes $U^{(L)}\mid V=v$; the electorate's average utility for election outcome $e$ becomes $E[U^{(L)}\mid E=e]$, and so on.
+
 # Towards Individualized Representational Harm {#sec:indiv-harm}
 
-# Connections to the Partisan Symmetry Standard
+Utility may be defined solely in terms of realized election outcomes and districting plans.
+Harm, we argue, cannot.
+Harm is not the same as disutility, for to be harmed by a decision a voter must have a better outcome in the counterfactual world where another decision was made.
+Put another way, in the context of districting plans, if no alternative districting plan could provide more utility to a voter, then that voter was not harmed by the districting plan chosen.
+
+## Individual Harm
+
+To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
+Then we may define the harm to a voter as$$
+    h:\Voters\times\Elecs\times\Plans^2\to\R, 
+    \qquad h(v,e,q,\tilde q) = 
+        \ind\{\vote(v, e)\neq\win(q(v), e, q)\}
+        \ind\{\vote(v, e)=\win(\tilde q(v), e, \tilde q)\},
+$$that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.
+
+As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.
+Make the counterfactual plan a random variable $\tilde Q$, with $(E,Q,\tilde Q)\sim\pi$ such that $Q\indep\tilde Q$.
+Then the harm to a random voter is $H \dfeq h(V, E, Q, \tilde Q)$.
+
+The expected harm to a specific voter then becomes $$
+   \E[H\mid V=v] = \Pr\qty(\vote(v, E)\neq\win(Q(v), E, Q)
+    \text{ and }\vote(v, E)=\win(\tilde Q(v), E, \tilde Q)),
+$$
+
+that is, the probability that the voter is misrepresented under the adopted plan but would not have been under a counterfactual.
+
+*(Need some explanation as to why we are dropping the global measure at this point. B/c it boils down to a difference in seats?)*
+
+*(running examples).*
+
+## Fairness
+
+In practice, few voters will experience zero harm.
+Most voters living in a district with a political preference not matching their own could draw a districting plan, however contrived, which puts them in a district dominated by voters of their party.
+So merely targeting zero harm for all voters is not an adequate representation of fairness.
+Rather, intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
+**[CM: need to rephrase here. awkward]**
+
+From here onward we will restrict our attention to the case of two parties, so $\Parties=\{A,B\}$, since that will simplify the notation and exposition.
+But the same concepts and definitions naturally carry over to the multiparty setting.
+
+Continuing with the above notation, the average fairness of $Q$ versus counterfactual $\tilde Q$ may be defined as $$
+    F = \E[H\mid \vote(V, E) = A] - \E[H\mid \vote(V, E) = B];
+$$obviously the ordering of $A$ and $B$ is immaterial.
+
+Generally, we will want to evaluate the fairness of a specific districting plan $q$ versus a distribution of counterfactuals.
+These counterfactuals could arise from a posited probability distribution capturing the factors relevant to a non-partisan redistricting process, or they could be a set of alternative redistricting plans that were considered as part of the legislative process leading to a final enacted plan.
+We adopt the notation $F(q)$ for this scenario: $$
+    F(q) \dfeq \E[H\mid \vote(V, E) = A, Q=q] - \E[H\mid \vote(V, E) = B, Q=q],
+$$ so that we recover $F=\E[F(Q)]$.
+
+[**CM: better to define just** $F(Q)$**, since we will rarely compare two sets of plans?**]
+
+*(running examples).*
+
+# Estimating the Harm and Fairness Measures {#sec:estimating}
+
+# Connections to Partisan Symmetry and Fairness Standards {#sec:connections}
 

### 2021-08-30 R/toy/00_toy_setup.R

@@ -65,14 +65,22 @@ minissouri_people = minissouri %>%
     pmap_dfr(make_people)
 
 # functions and helpers -------
-plot_state = function(map, people, qty=NULL, ppl_size=2, ...) {
-    redist.plot.map(map, fill=qty, ...) +
+plot_state = function(map, people, plan=NULL, qty=NULL, ppl_size=2, ...) {
+    p = redist.plot.map(map, fill=qty, ...) +
         scale_fill_party_c() +
         geom_sf(data=map, size=0.8, fill=NA, color="white") +
         geom_sf(aes(color=dem), data=people, size=ppl_size) +
         scale_color_party_d(guide="none") +
         theme_repr_map() +
         theme(plot.title = element_text(hjust = 0.5))
+    if (!is.null(plan)) {
+        distrs = map %>%
+            mutate(.plan = plan) %>%
+            group_by(.plan) %>%
+            summarize()
+        p = p + geom_sf(data=distrs, fill=NA, color="black")
+    }
+    p
 }
 
 sim_toy = function(map, N=500) {
@@ -89,3 +97,5 @@ if (!file.exists(fig_path <- here("paper/figures/minis_schematic.pdf"))) {
         wrap_plots(nrow=1)
     ggsave(fig_path, width=6.5, height=2.5)
 }
+
+plan = c(2, 2, 2, 3, 3, 3, 2, 2, 2, 3, 3, 3, 2, 2, 2, 3, 3, 3, 2, 2, 2, 3, 3, 3, 1, 1, 1, 3, 3, 3, 1, 1, 1, 3, 3, 3)

### 2021-08-30 R/toy/01_sim_mininois.R

@@ -1,4 +1,14 @@
-pl_mininois = sim_toy(mininois)
+pl_mininois = sim_toy(mininois, N=2000)
+
+pl_mininois %>%
+    mutate(u_g_dem = rep(utility_global(., dem), each=3),
+           u_g_rep = rep(utility_global(., dem, invert=T), each=3),
+           u_g = total(u_g_dem, u_g_rep, mininois, dem, rep)) %>%
+    arrange(desc(u_g)) %>%
+    as.matrix() %>%
+    `[`(, 1) %>%
+    datapasta::dpasta()
+x = c(2, 2, 2, 3, 3, 3, 2, 2, 2, 3, 3, 3, 2, 2, 2, 3, 3, 3, 2, 2, 2, 3, 3, 3, 1, 1, 3, 3, 3, 3, 1, 1, 1, 3, 3, 3)
 
 redist.plot.plans(pl_mininois, 1, mininois, dem) +
     scale_fill_party_c() +

### 2021-08-30 paper/paper.Rmd
add informal line
better connect utility and harm

@@
 To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
-Then we may define the harm to a voter as$$
-    h:\Voters\times\Elecs\times\Plans^2\to\R, 
-    \qquad h(v,e,q,\tilde q) = 
-        \ind\{\vote(v, e)\neq\win(q(v), e, q)\}
+Then we may define the harm to a voter as $h:\Voters\times\Elecs\times\Plans^2\to\R$, \begin{align*}
+    h(v,e,q,\tilde q) &= (1 - u^{(L)}(v, e, q))\cdot u^{(L)}(v, e, \tilde q) \\
+    &= \ind\{\vote(v, e)\neq\win(q(v), e, q)\}
         \ind\{\vote(v, e)=\win(\tilde q(v), e, \tilde q)\},
-$$that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.
+\end{align*}that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.
 
@@
 $$obviously the ordering of $A$ and $B$ is immaterial.
+Informally, this quantity is the difference in the probability that a voter of each party is represented by the other party but counterfactually would have been represented by their own party.
 

### 2021-08-30 paper/paper.Rmd
reorder
typo
add examples and change notation

@@
 
-Let $\Voters$ be the set of voters in the region and time period under consideration, who vote to select a candidate from one of a set of political parties $\Parties$ to represent each of a set of legislative districts, indexed by $\Distrs$.
+Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate from one of a set of political parties $\Parties$ to represent each of a set of legislative districts, indexed by $\Distrs$.
 We take these three sets to be fixed.
 
-Voters vote in elections, each of which produces a specific outcome $e$ from the set $\Elecs$ of all possible election outcomes.
-A specific election contains information on how each voter voted.
+Voters vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
+In an election, voters either vote for a candidate running in their district, or abstain from voting.
+So we write $v:\Voters\to\Parties\cup\{\textsc{abstain}\}$, with $v(i)$ the party that voter $i\in\Voters$ voted for in election $v$.
 
@@
 
-In an election, voters either vote for a candidate running in their district, or abstain from voting.
-Notationally, we have a set of functions $\vote:\Voters\times\Elecs\to \Parties\cup\{\textsc{abstain}\}$.
 Votes may be tallied at the district level, either in total or by party: \begin{align*}
     \votes&: \Distrs\times\Elecs \to \N, \qquad 
-    \votes(d, e) = \sum_{v\in q^{-1}(d)} \ind_P(\vote(v, e)) \\
+    \votes(d, v) = \sum_{i\in q^{-1}(d)} \ind_P(v(i)) \\
     \votes^{(p)}&: \Distrs\times\Elecs\times\Plans \to \N, \qquad 
-    \votes^{(p)}(d, e, q) = \sum_{v\in q^{-1}(d)} \ind_{\{p\}}(\vote(v, e)).
+    \votes^{(p)}(d, v, q) = \sum_{i\in q^{-1}(d)} \ind_{\{p\}}(v(i)).
 \end{align*} The political party of the candidate with the most votes wins each district, as captured by the function $$
     \win: \Distrs\times\Elecs\times\Plans\to P, \qquad 
-    \win(d, e, q) = \argmax_{p\in\Parties} \votes^{(p)}(d, e, q).
-$$Thus an individual voter $v\in V$ is represented after election $e$ by political party $\win(q(v), e, q)$, and each political party can be said to win a number of seats, formalized as$$
+    \win(d, v, q) = \argmax_{p\in\Parties} \votes^{(p)}(d, v, q).
+$$Thus an individual voter $i\in\Voters$ is represented after election $v$ by political party $\win(q(i), v, q)$, and each political party can be said to win a number of seats, formalized as$$
     \seats:\Parties\times\Elecs\times\Plans\to\N, 
-    \qquad \seats(p,e,q) = \sum_{d\in\Distrs} \ind_{\{p\}}(\win(d,e,q)).
+    \qquad \seats(p,v,q) = \sum_{d\in\Distrs} \ind_{\{p\}}(\win(d,v,q)).
 $$
-
-**[CM: better letter/concept pairings possible?]**
 
@@
     u^{(G)}:\Voters\times\Elecs\times\Plans\to\R, 
-    \qquad u^{(G)}(v,e,q) = g\qty(\frac{\seats(\vote(v, e), e, q)}{|\Distrs|}),
+    \qquad u^{(G)}(i,v,q) = g\qty(\frac{\seats(v(i), v, q)}{|\Distrs|}),
 $$for $g:[0,1]\bijto\R$ an increasing function.
@@
 
-*(running examples).*
-
 ## Local Outcome Utility
@@
     u^{(L)}:\Voters\times\Elecs\times\Plans\to\R, 
-    \qquad u^{(L)}(v,e,q) = \ind\{\vote(v, e)=\win(q(v), e, q)\},
+    \qquad u^{(L)}(i,v,q) = \ind\{v(i)=\win(q(i), e, q)\},
 $$that is, a voter gains utility only if they winner of their legislative district belongs to their preferred political party.
@@
 
-*(running examples).*
+## Examples
+
+To illustrate these two utility measures, Figure \ref{fig:mini-util-opt} shows three plans for our running examples which maximize both global (with $g=\log$) and local utility.[^2]
+In Mininois and Minissouri, the overall division of seats is as close to proportional as possible.
+In Minichusetts, however, it is impossible to give any seats to Republicans (as long as districts are constructed from precincts); all plans have the same utility.
+In none of the three states is it possible to create a perfect bipartisan gerrymander, but the plans for Mininois and Minissouri get as close as possible by packing Democratic voters into cities (and Republican voters into rural areas, where feasible).
+
+[^2]: These examples present a slight complication---what to do when a party wins no seats?
+    The easiest solution, perhaps, is to define the utility to that party's voters to be $-1$; this will make the utility well-defined while still encouraging proportionality.
+
+```{r mini-util-opt, fig.cap="Utility-maximizing redistricting plans for the three running examples, with each district shaded by its partisan lean."}
+include_graphics(here("paper/figures/minis_util_opt.pdf"))
+```
 
@@
 
-The utility measures above are defined for a single election outcome $e\in \Elecs$ and districting plan $q\in\Plans$.
+The utility measures above are defined for a single election outcome $v\in \Elecs$ and districting plan $q\in\Plans$.
 But often we will want to consider utility averaged over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans.
 
-To allow this, we replace the fixed election districting plan with random variables $(E,Q)\sim \pi$, where $\pi$ is a joint probability measure on election outcomes and districting plans.
+To allow this, we replace the fixed election districting plan with random variables $(V,Q)\sim \pi$, where $\pi$ is a joint probability measure on election outcomes and districting plans.
 The randomness in election outcomes comes from variations in factors such as voter preferences, the political environment, candidate quality, and campaign events.
 The randomness in districting plans reflects the constraints and preferences of the plan-drawing process.
-As we will discuss in Section \ref{sec:estimating}, generally we will assume that $E\indep Q$, though this is not required for our theoretical analysis.
-We will also consider a voter $V$ taken uniformly at random from $\Voters$, with $V\indep\pi$.
+As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V\indep Q$, though this is not required for our theoretical analysis.
+We will also consider a voter $I$ taken uniformly at random from $\Voters$, with $I\indep\pi$.
 
 Then we may define the utilities for a (randomly selected) voter as $$
-    U^{(G)} \dfeq u^{(G)}(V, E, Q) \qand
-    U^{(L)} \dfeq u^{(L)}(V, E, Q).
-$$Thus, the utility for a single voter $v$ becomes $U^{(L)}\mid V=v$; the electorate's average utility for election outcome $e$ becomes $E[U^{(L)}\mid E=e]$, and so on.
+    U^{(G)} \dfeq u^{(G)}(I, V, Q) \qand
+    U^{(L)} \dfeq u^{(L)}(I, V, Q).
+$$Thus, the utility for a single voter $v$ becomes $U^{(L)}\mid I=i$; the electorate's average utility for election outcome $e$ becomes $\E[U^{(L)}\mid V=v]$, and so on.
 
@@
 Then we may define the harm to a voter as $h:\Voters\times\Elecs\times\Plans^2\to\R$, \begin{align*}
-    h(v,e,q,\tilde q) &= (1 - u^{(L)}(v, e, q))\cdot u^{(L)}(v, e, \tilde q) \\
-    &= \ind\{\vote(v, e)\neq\win(q(v), e, q)\}
-        \ind\{\vote(v, e)=\win(\tilde q(v), e, \tilde q)\},
+    h(i,v,q,\tilde q) &= (1 - u^{(L)}(i, v, q))\cdot u^{(L)}(i, v, \tilde q) \\
+    &= \ind\{v(i)\neq\win(q(i), e, q)\}
+        \ind\{v(i)=\win(\tilde q(i), e, \tilde q)\},
 \end{align*}that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.
 
+*(Need some explanation as to why we are dropping the global measure at this point. B/c it boils down to a difference in seats?)*
+
 As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.
-Make the counterfactual plan a random variable $\tilde Q$, with $(E,Q,\tilde Q)\sim\pi$ such that $Q\indep\tilde Q$.
-Then the harm to a random voter is $H \dfeq h(V, E, Q, \tilde Q)$.
+Make the counterfactual plan a random variable $\tilde Q$, with $(V,Q,\tilde Q)\sim\pi$ such that $Q\indep\tilde Q$.
+Then the harm to a random voter is $H \dfeq h(I, V, Q, \tilde Q)$.
 
 The expected harm to a specific voter then becomes $$
-   \E[H\mid V=v] = \Pr\qty(\vote(v, E)\neq\win(Q(v), E, Q)
-    \text{ and }\vote(v, E)=\win(\tilde Q(v), E, \tilde Q)),
+   \E[H\mid I=i] = \Pr\qty(V(i)\neq\win(Q(i), V, Q)
+    \text{ and }V(i)=\win(\tilde Q(i), V, \tilde Q)),
 $$
@@
 
-*(Need some explanation as to why we are dropping the global measure at this point. B/c it boils down to a difference in seats?)*
+```{r mini-harm, fig.cap="Three potential districting plans for Mininois and Minissouris, showing district partisan lean (top row) and total voter harm versus the counterfactuals in Figure 2(a)-(b) (bottom row; darker shading indicates more harm)."}
+include_graphics(here("paper/figures/mini_harm.pdf"))
+```
 
-*(running examples).*
+Figure \ref{fig:mini-harm} illustrates this measure for a set of hypothetical gerrymanders, comparing to the set of optimal plans from Figure \ref{fig:mini-util-opt}.
+As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.
 
@@
 Continuing with the above notation, the average fairness of $Q$ versus counterfactual $\tilde Q$ may be defined as $$
-    F = \E[H\mid \vote(V, E) = A] - \E[H\mid \vote(V, E) = B];
+    F = \E[H\mid V(I) = A] - \E[H\mid V(I) = B];
 $$obviously the ordering of $A$ and $B$ is immaterial.
@@
 We adopt the notation $F(q)$ for this scenario: $$
-    F(q) \dfeq \E[H\mid \vote(V, E) = A, Q=q] - \E[H\mid \vote(V, E) = B, Q=q],
+    F(q) \dfeq \E[H\mid V(I) = A, Q=q] - \E[H\mid V(I) = B, Q=q],
 $$ so that we recover $F=\E[F(Q)]$.
@@
 
-*(running examples).*
+In the districting plan examples of Figure \ref{fig:mini-harm}, the fairness of the Democratic gerrymander of Mininois is $-0.494$, the fairness of the Democratic gerrymander of Minissouri is $-0.415$, and the fairness of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
+In other words, the average Republican voter in Mininois is 49.4% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).
 

### 2021-08-30 R/01_metrics.R

@@ -130,3 +130,17 @@ total = function(util_1, util_2, map, group_1, group_2) {
     voters_2 = sum(eval_tidy(enquo(group_2), map))
     (util_1 * voters_1 + util_2 * voters_2) / (voters_1 + voters_2)
 }
+
+scorer_util_g = function(map, group_pop, total_pop) {
+    group_pop = eval_tidy(enquo(group_pop), map)
+    total_pop = eval_tidy(enquo(total_pop), map)
+    n1 = sum(group_pop)
+    n2 = sum(total_pop) - n1
+    ndists = attr(map, "ndists")
+    fn = function(plans) {
+        seats = colSums(redist:::group_pct(plans, group_pop, total_pop, ndists) > 0.5)
+        (flog(seats)*n1 + flog(ndists - seats)*n2) / (n1 + n2)
+    }
+    class(fn) = c("redist_scorer", "function")
+    fn
+}

### 2021-08-30 R/toy/01_sim_mininois.R

@@ -1,17 +1,0 @@
-pl_mininois = sim_toy(mininois, N=2000)
-
-pl_mininois %>%
-    mutate(u_g_dem = rep(utility_global(., dem), each=3),
-           u_g_rep = rep(utility_global(., dem, invert=T), each=3),
-           u_g = total(u_g_dem, u_g_rep, mininois, dem, rep)) %>%
-    arrange(desc(u_g)) %>%
-    as.matrix() %>%
-    `[`(, 1) %>%
-    datapasta::dpasta()
-x = c(2, 2, 2, 3, 3, 3, 2, 2, 2, 3, 3, 3, 2, 2, 2, 3, 3, 3, 2, 2, 2, 3, 3, 3, 1, 1, 3, 3, 3, 3, 1, 1, 1, 3, 3, 3)
-
-redist.plot.plans(pl_mininois, 1, mininois, dem) +
-    scale_fill_party_c() +
-    geom_sf(data=mininois, size=0.8, fill=NA, color="white") +
-    geom_sf(aes(color=dem), data=mininois_people, size=5) +
-    scale_color_party_d(guide="none")

### 2021-08-30 R/toy/01_util.R

@@ -1,0 +1,31 @@
+pl_opt = list(
+    mininois = c(2, 2, 3, 3, 3, 3,
+                 2, 2, 3, 3, 3, 3,
+                 2, 2, 2, 3, 3, 3,
+                 2, 2, 2, 3, 3, 3,
+                 1, 1, 1, 3, 3, 3,
+                 1, 1, 1, 3, 3, 3),
+    minissouri = c(3, 3, 3, 3, 3, 3,
+                   2, 2, 3, 3, 3, 3,
+                   2, 2, 3, 3, 3, 3,
+                   2, 2, 2, 2, 1, 1,
+                   2, 2, 2, 1, 1, 1,
+                   2, 2, 2, 1, 1, 1),
+    minichusetts = rep(c(1, 1, 2, 2, 3, 3), 6)
+)
+
+dem_opt = imap(pl_opt, function(pl, nm) {
+    with(get(nm), tapply(dem, pl, sum)/tapply(pop, pl, sum))[pl]
+})
+
+plot_opt = imap(names(pl_opt), function(nm, i) {
+    plot_state(get(nm), get(paste0(nm, "_people")),
+               pl_opt[[nm]], dem_opt[[nm]]) +
+        labs(title=paste(c("(a)", "(b)", "(c)")[i], str_to_title(nm)))
+})
+
+wrap_plots(plot_opt) +
+    plot_layout(guides="collect")
+ggsave(here("paper/figures/minis_util_opt.pdf"), width=7, height=2.5)
+
+

### 2021-08-30 R/toy/02_harm_fairness.R

@@ -1,0 +1,70 @@
+# Mininois
+pl_gerry = c(2, 2, 2, 3, 3, 1,
+             2, 2, 2, 3, 3, 1,
+             2, 2, 2, 3, 3, 1,
+             2, 2, 3, 3, 3, 1,
+             2, 3, 3, 3, 1, 1,
+             1, 1, 1, 1, 1, 1)
+dem_gerry = with(mininois, tapply(dem, pl_gerry, sum)/tapply(pop, pl_gerry, sum))[pl_gerry]
+dem_harm = weighted.mean((dem_gerry<0.5)*(dem_opt$mininois>0.5), mininois$dem)
+rep_harm = weighted.mean((dem_gerry>0.5)*(dem_opt$mininois<0.5), mininois$rep)
+fairness_gerry = dem_harm - rep_harm
+cat("Mininois D gerrymander fairness: ", round(fairness_gerry, 3), "\n")
+
+p1 = plot_state(mininois, mininois_people,
+           pl_gerry, dem_gerry) +
+    labs(title="Democratic gerrymander\nof Mininois")
+p2 = plot_state(mininois, mininois_people,
+           pl_gerry, mininois$rep*round(1-dem_opt$mininois)) +
+    scale_fill_gradient(low="#e5e6e5", high="#666666", guide="none") +
+    labs(title="Total voter harm vs. 2(a)")
+
+# Minissouri
+pl_gerry = c(1, 2, 2, 2, 2, 2,
+             1, 2, 3, 3, 3, 2,
+             1, 3, 3, 3, 3, 2,
+             1, 3, 3, 3, 3, 2,
+             1, 3, 3, 3, 1, 2,
+             1, 1, 1, 1, 1, 2)
+dem_gerry = with(minissouri, tapply(dem, pl_gerry, sum)/tapply(pop, pl_gerry, sum))[pl_gerry]
+harm_gerry = minissouri$dem*(dem_gerry<0.5)*(dem_opt$minissouri>0.5) +
+    minissouri$rep*(dem_gerry>0.5)*(dem_opt$minissouri<0.5)
+dem_harm = weighted.mean((dem_gerry<0.5)*(dem_opt$minissouri>0.5), minissouri$dem)
+rep_harm = weighted.mean((dem_gerry>0.5)*(dem_opt$minissouri<0.5), minissouri$rep)
+fairness_gerry = dem_harm - rep_harm
+cat("Minissouri D gerrymander fairness: ", round(fairness_gerry, 3), "\n")
+
[…]

### 2021-08-30 R/toy/03_sims.R

@@ -1,0 +1,3 @@
+pl_mininois = sim_toy(mininois, N=2000)
+
+opt = redist_shortburst(mininois, scorer_group_pct(mininois, dem, pop, 3))

### 2021-08-31 paper/paper.Rmd
cleanup + fig labels

@@
 
-Votes may be tallied at the district level, either in total or by party: \begin{align*}
-    \votes&: \Distrs\times\Elecs \to \N, \qquad 
-    \votes(d, v) = \sum_{i\in q^{-1}(d)} \ind_P(v(i)) \\
-    \votes^{(p)}&: \Distrs\times\Elecs\times\Plans \to \N, \qquad 
+Votes may be tallied at the district level, by party, by defining for each party $p\in\Parties$ $$
+    \votes^{(p)}: \Distrs\times\Elecs\times\Plans \to \N, \qquad 
     \votes^{(p)}(d, v, q) = \sum_{i\in q^{-1}(d)} \ind_{\{p\}}(v(i)).
-\end{align*} The political party of the candidate with the most votes wins each district, as captured by the function $$
+$$ The political party of the candidate with the most votes wins each district, as captured by the function $$
     \win: \Distrs\times\Elecs\times\Plans\to P, \qquad 
@@
 What kind of electoral systems does this utility function encourage, if the goal is maximizing total voter utility (or, equivalently, average voter utility)?
-It depends on the form of $h$, as the following result describes.
+It depends on the form of $g$, as the following result describes.
 
@@
 ```
-Thus, if we are willing to adopt the logarithm as the functional specification of $u^{(G)}$, then this global outcome utility measure will encourage districting systems that create proportionality between party vote shares and seat shares.
+Thus, if we are willing to adopt the logarithm for the functional specification of $u^{(G)}$, then this global outcome utility measure will encourage districting systems that create proportionality between party vote shares and seat shares.
 
@@
 
-The expected harm to a specific voter then becomes $$
+The expected harm to an individual $i\in\Voters$ voter then becomes $$
    \E[H\mid I=i] = \Pr\qty(V(i)\neq\win(Q(i), V, Q)
@@
 
-In practice, few voters will experience zero harm.
-Most voters living in a district with a political preference not matching their own could draw a districting plan, however contrived, which puts them in a district dominated by voters of their party.
-So merely targeting zero harm for all voters is not an adequate representation of fairness.
-Rather, intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
-**[CM: need to rephrase here. awkward]**
-
 From here onward we will restrict our attention to the case of two parties, so $\Parties=\{A,B\}$, since that will simplify the notation and exposition.
@@
 
-Continuing with the above notation, the average fairness of $Q$ versus counterfactual $\tilde Q$ may be defined as $$
+Intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
+This is easily expressed using our notation above; the average fairness of $Q$ versus counterfactual $\tilde Q$ may be defined as $$
     F = \E[H\mid V(I) = A] - \E[H\mid V(I) = B];

### 2021-08-31 R/toy/01_util.R

@@ -18,10 +18,24 @@ dem_opt = imap(pl_opt, function(pl, nm) {
     with(get(nm), tapply(dem, pl, sum)/tapply(pop, pl, sum))[pl]
 })
 
+coords = list(
+    mininois = list(x=c(1/4, 1/6, 2/3),
+                    y=c(1.04, -0.04, -0.04)),
+    minissouri = list(x=c(3/4, 1/4, 1/2),
+                      y=c(1.04, 1.04, -0.04)),
+    minichusetts = list(x=c(1/6, 1/2, 5/6),
+                        y=c(-0.04, 1.04, -0.04))
+)
+
 plot_opt = imap(names(pl_opt), function(nm, i) {
-    plot_state(get(nm), get(paste0(nm, "_people")),
-               pl_opt[[nm]], dem_opt[[nm]]) +
+    p = suppressMessages(plot_state(get(nm), get(paste0(nm, "_people")),
+                                    pl_opt[[nm]], dem_opt[[nm]])) +
         labs(title=paste(c("(a)", "(b)", "(c)")[i], str_to_title(nm)))
+    for (i in 1:3) {
+        p = add_lab(p, fmt_lean(dem_opt[[nm]][as.character(i)]),
+                    coords[[nm]]$x[i], coords[[nm]]$y[i])
+    }
+    p
 })
 
 wrap_plots(plot_opt) +

### 2021-08-31 R/toy/02_harm_fairness.R

@@ -11,13 +11,21 @@ rep_harm = weighted.mean((dem_gerry>0.5)*(dem_opt$mininois<0.5), mininois$rep)
 fairness_gerry = dem_harm - rep_harm
 cat("Mininois D gerrymander fairness: ", round(fairness_gerry, 3), "\n")
 
+coords = list(x=c(1/2, 2/3, 1/4),
+              y=c(1.04, -0.04, -0.04))
+suppressMessages({
 p1 = plot_state(mininois, mininois_people,
            pl_gerry, dem_gerry) +
     labs(title="Democratic gerrymander\nof Mininois")
+for (i in 1:3) {
+    p1 = add_lab(p1, fmt_lean(dem_gerry[as.character(i)]),
+                coords$x[i], coords$y[i])
+}
 p2 = plot_state(mininois, mininois_people,
            pl_gerry, mininois$rep*round(1-dem_opt$mininois)) +
     scale_fill_gradient(low="#e5e6e5", high="#666666", guide="none") +
     labs(title="Total voter harm vs. 2(a)")
+})
 
 # Minissouri
 pl_gerry = c(1, 2, 2, 2, 2, 2,
@@ -34,13 +42,21 @@ rep_harm = weighted.mean((dem_gerry>0.5)*(dem_opt$minissouri<0.5), minissouri$re
 fairness_gerry = dem_harm - rep_harm
 cat("Minissouri D gerrymander fairness: ", round(fairness_gerry, 3), "\n")
 
+coords = list(x=c(1/2, 1/2, 1/2),
+              y=c(1.04, -0.04, 1/2))
+suppressMessages({
 p3 = plot_state(minissouri, minissouri_people,
            pl_gerry, dem_gerry) +
     labs(title="Democratic gerrymander\nof Minissouri")
+for (i in 1:3) {
+    p3 = add_lab(p3, fmt_lean(dem_gerry[as.character(i)]),
+                coords$x[i], coords$y[i])
+}
 p4 = plot_state(minissouri, minissouri_people,
            pl_gerry, harm_gerry) +
     scale_fill_gradient(low="#e5e6e5", high="#666666", guide="none") +
     labs(title="Total voter harm vs. 2(b)")
+})
 
 pl_gerry = c(3, 2, 2, 2, 2, 2,
              3, 1, 1, 1, 2, 2,
[…]

### 2021-08-31 paper/paper.Rmd
start estimation section

@@
 
-From here onward we will restrict our attention to the case of two parties, so $\Parties=\{A,B\}$, since that will simplify the notation and exposition.
+From here onward we will restrict our attention to the case of two parties---concretely, $\Parties=\{\dem,\rep\}$---since that will simplify the notation and exposition.
 But the same concepts and definitions naturally carry over to the multiparty setting.
@@
 This is easily expressed using our notation above; the average fairness of $Q$ versus counterfactual $\tilde Q$ may be defined as $$
-    F = \E[H\mid V(I) = A] - \E[H\mid V(I) = B];
-$$obviously the ordering of $A$ and $B$ is immaterial.
+    F = \E[H\mid V(I) = \dem] - \E[H\mid V(I) = \rep];
+$$obviously the ordering of $\dem$ and $\rep$ is immaterial.
 Informally, this quantity is the difference in the probability that a voter of each party is represented by the other party but counterfactually would have been represented by their own party.
@@
 We adopt the notation $F(q)$ for this scenario: $$
-    F(q) \dfeq \E[H\mid V(I) = A, Q=q] - \E[H\mid V(I) = B, Q=q],
+    F(q) \dfeq \E[H\mid V(I) = \dem, Q=q] - \E[H\mid V(I) = \rep, Q=q],
 $$ so that we recover $F=\E[F(Q)]$.
@@
 
-# Estimating the Harm and Fairness Measures {#sec:estimating}
+# Calculating the Harm and Fairness Measures {#sec:estimating}
+
+The measures $H$ and $F(q)$ defined above require specification of a probability measure on elections and districting plans, $\pi$.
+Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
+Either way, a collection of past election data and an enacted districting plan alone are not enough to calculate $H$ and $F(q)$.
+
+While not strictly necessary, modelling of $V$ and $\tilde Q$ independently will simplify the situation considerably.
+Practically, this translates to the assumption that individual voters would vote for the same political party no matter the districting plan.
+To the extent that voters reward or punish an incumbent party for the districting plan they adopt, or that an adopted districting plan affects which candidates run in a district, this assumption is violated.
+But as our suggested model for election outcomes will not account for candidate factors or policy preferences, the marginal effect on our conclusions will be minimal.
+
+## Modelling Districting Plans
+
+## Modelling Election Outcomes
 

### 2021-08-31 paper/paper.Rmd
explore NJ

@@
 ---
-title: Individualized Measures of Partisan Fairness and Utility in Redistricting (??)
+title: Individualized Measures of Partisan Fairness and Utility in Redistricting
 authors:

### 2021-08-31 R/00_setup_utils.R

@@ -14,9 +14,10 @@ theme_repr_map = function() theme_void(base_family="Times", base_size=10)
 GOP_DEM = c("#A0442C", "#B25D4C", "#C27568", "#D18E84", "#DFA8A0",
             "#EBC2BC",  "#F6DCD9", "#F9F9F9", "#DAE2F4", "#BDCCEA",
             "#9FB6DE", "#82A0D2", "#638BC6", "#3D77BB", "#0063B1")
-scale_fill_party_c = function(name="Democratic share", ...) {
+scale_fill_party_c = function(name="Democratic share", midpoint=0.5, limits=0:1,
+                              labels=percent, oob=squish, ...) {
     scale_fill_gradient2(name=name, ..., low = GOP_DEM[1], high = GOP_DEM[15],
-                         midpoint=0.5, limits=0:1, labels=percent)
+                         midpoint=midpoint, limits=limits, labels=labels, oob=oob)
 }
 scale_color_party_d = function(...) {
     scale_color_manual(..., values=c(GOP_DEM[2], GOP_DEM[14]),

### 2021-08-31 R/03_sim_nj.R

@@ -1,0 +1,41 @@
+nj = read_rds(here("data/NJ_cd_final_vtd_20.rds")) %>%
+    redist_map(pop_tol=0.01, ndists=12, adj=.$adj)
+
+N_sim = 4e2
+plans = redist_smc(nj, N_sim, counties=county, pop_temper=0.01, verbose=FALSE)
+
+plans = plans %>%
+    mutate(dev = plan_parity(nj),
+           comp = distr_compactness(nj),
+           dem = group_frac(nj, ndv, ndv + nrv),
+           competitive = competitiveness(nj, nrv, ndv, ),
+           mean_med = partisan_metrics(nj, "MeanMedian", nrv, ndv),
+           pbias = partisan_metrics(nj, "Bias", nrv, ndv))
+
+m_dem = district_group(plans, dem)
+
+ker = k_t()
+pl_sum = plans %>%
+    group_by(draw) %>%
+    summarize(n_dem = sum(dem > 0.5),
+              n_dem_smooth = sum(ker(dem)),
+              across(c(dev, comp, competitive:pbias), ~ .[1])) %>%
+    mutate(u_loc_dem = utility_local(nj, ndv, m_dem, ker),
+           u_loc_rep = utility_local(nj, ndv, m_dem, ker, invert=TRUE),
+           u_glb_dem = utility_global(plans, dem, ker),
+           u_glb_rep = utility_global(plans, dem, ker, invert=TRUE),
+           h_dem = harm_v1(nj, ndv, m_dem, idx_2=-1:-2, kernel=ker),
+           h_rep = harm_v1(nj, ndv, m_dem, idx_2=-1:-2, kernel=ker, invert=TRUE),
+           u_loc_log = total(log(u_loc_dem), log(u_loc_rep), nj, ndv, nrv),
+           u_loc = total(u_loc_dem, u_loc_rep, nj, ndv, nrv),
+           u_glb = total(u_glb_dem, u_glb_rep, nj, ndv, nrv),
+           h = total(h_dem, h_rep, nj, ndv, nrv))
+
+plot(nj, rowMeans(m_Dem > 0.5)) + scale_fill_party_c()
+
+redist.plot.scatter(pl_sum, u_glb, h_dem - h_rep)
+redist.plot.scatter(pl_sum, u_loc, h_dem - h_rep)
+
+plot(plans, dem, size=0.2)
+hist(plans_diversity(plans)*log(12))
+redist.plot.plans(plans, 1:6, nj)

### 2021-09-01 paper/paper.Rmd
title

@@
 ---
-title: Individualized Measures of Partisan Fairness and Utility in Redistricting
+title: Measuring Individual Partisan Harm in Redistricting
 authors:
@@
 ```
-
-**Title suggestions**
-
--   Individual Representation: Harm and Utility
-
--   Redistricting and Individual Representational Harm
-
--   Fairness and Utility in Redistricting: an Individual Perspective
-
--   Individualized Measures of Fairness and Utility in Redistricting
 

### 2021-09-02 paper/paper.Rmd
try paragraph on why no global

@@
 
-*(Need some explanation as to why we are dropping the global measure at this point. B/c it boils down to a difference in seats?)*
+Harm is defined using the local, not global, utility measure, because inherently, global utility (and thus harm) will be constant for all voters of a particular party.
+In fact, a harm measure defined analogously using global utility would essentially be a difference in seats won by a voter's party between plans $q$ and $\tilde q$ (the exact form would be affected by the function $g$).
+This measure is not particularly interesting, and nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation **[CM: cite Rucho & LWV of PA]**.
+**[CM: revisit ¶ based on NJ findings].**
 

### 2021-09-03 paper/paper.Rmd
reforming intro

@@
 
+What is the question?
+
+Districting is a complicated process.
+Residents of a state or locality must be partitioned into districts.
+These districts must adhere to many constraints, often requiring components such as contiguity and equal population.
+In drawing the districts, it's easy to lose sight of the individuals impacted by district drawing.
+Often, we consider how parties succeed or fail under the electoral environments generated by enacted plans.
+Yet this misses that parties are formed by voters who ally with parties.
+Rather than considering parties, we ask how individuals are harmed or benefited by the districting process?
+
+Why does it matter?
+
+Why is it hard?
+Decades of
+
+What did we do?
+
+What did we learn?
+
+What is the paper's contribution?
+
+```{=html}
+<!--
 Understanding fairness in electoral systems has long suffered from a focus on parties over the individuals who vote for them.
@@
     -   Recent advances allow us to make claims about harm based in counterfactuals that are realistic and statistically useful
-
+-->
+```
 ## Running Examples

### 2021-09-06 paper/paper.Rmd
fixed metrics; WI sim

@@
 The expected harm to an individual $i\in\Voters$ voter then becomes $$
-   \E[H\mid I=i] = \Pr\qty(V(i)\neq\win(Q(i), V, Q)
-    \text{ and }V(i)=\win(\tilde Q(i), V, \tilde Q)),
+   \E[H\mid I=i] = \Pr[V(i)\neq\win(Q(i), V, Q)
+    \text{ and }V(i)=\win(\tilde Q(i), V, \tilde Q)],
 $$
@@
 
-In the districting plan examples of Figure \ref{fig:mini-harm}, the fairness of the Democratic gerrymander of Mininois is $-0.494$, the fairness of the Democratic gerrymander of Minissouri is $-0.415$, and the fairness of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
-In other words, the average Republican voter in Mininois is 49.4% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).
+In the districting plan examples of Figure \ref{fig:mini-harm}, the fairness of the Democratic gerrymander of Mininois is $-0.519$, the fairness of the Democratic gerrymander of Minissouri is $-0.415$, and the fairness of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
+In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).
 

### 2021-09-06 R/00_setup_utils.R

@@ -30,21 +30,46 @@ scale_color_party_d = function(...) {
 
 # 'fixed' log that handles 0 and negative numbers
 flog = function(x) if_else(x <= 0, -1, suppressWarnings(log(x)))
+plapl = function(x) 0.5 * (1 + sign(x) * (1 - exp(-2*abs(x))))
+qlapl = function(p) -0.5 * sign(p - 0.5) * log(1 - 2*abs(p - 0.5))
 k_step = function(x) x > 0.5
 # historical congressional national shifts
 d_hist = local({
     dem_hist = c(0.575, 0.559, 0.537, 0.505, 0.552, 0.521, 0.543, 0.533, 0.521, 0.501, 0.447, 0.482, 0.473, 0.471, 0.452, 0.468, 0.523, 0.532, 0.449, 0.488, 0.455, 0.48, 0.534)
     gop_hist = c(0.407, 0.423, 0.448, 0.479, 0.434, 0.47, 0.444, 0.456, 0.443, 0.451, 0.515, 0.482, 0.484, 0.476, 0.5, 0.494, 0.443, 0.426, 0.517, 0.476, 0.512, 0.491, 0.448)
-    dem_tpp = dem_hist / (dem_hist + gop_hist)
+    dem_tpp = qlapl(dem_hist / (dem_hist + gop_hist))
     list(sd=sd(dem_tpp), df=length(dem_hist)-1)
 })
+sd_app = d_hist$sd * sd(rt(1e5, df=d_hist$df))
 k_t = function(sd=d_hist$sd) {
-    function(x) pt((x-0.5)/sd, df=d_hist$df)
+    function(x) pt(qlapl(x)/sd, df=d_hist$df)
+}
+e_approx = function(a, b) {
+    y = sd_app * (dnorm(a/sd_app) - dnorm(b/sd_app)) / (pnorm(b/sd_app) - pnorm(a/sd_app))
+    if_else(is.finite(y), y,
+            if_else(is.finite(a),
+                    if_else(is.finite(b), 0.5*(a+b), a), b))
 }
 
-plot_cds = function(map, pl, county, abbr) {
+
+plot_cds = function(map, pl, county, abbr, qty="ndv") {
     plan = as.factor(redist:::color_graph(get_adj(map), as.integer(pl)))
     places = suppressMessages(tigris::places(abbr, cb=TRUE))
+
+    if (qty == "dem") {
+        dvote = map$dem
+        rvote = map$rep
[…]

### 2021-09-06 R/01_metrics.R

@@ -25,32 +25,62 @@ district_group = function(plans, group) {
 #' @param group_pop column of `map` containing the group population of each precinct.
 #' @param distr_grp a matrix from `district_group()`
 #' @param kernel a vectorized function mapping [0, 1] to [0, 1]
-#' @param invert if `TRUE`, calculate utility for the out-group
+#' @param ker_mean a vectorized function returning the mean x value under the kernel distribution on the interval [a, b]
 #'
 #' @returns a numeric vector
-utility_local = function(map, group_pop, distr_grp, kernel=k_step, invert=FALSE) {
+utility_local = function(map, group_pop, other_pop, distr_grp, kernel=k_step, ker_mean=e_approx) {
     m = kernel(distr_grp)
-    if (invert) m = 1 - m
-    voters = eval_tidy(enquo(group_pop), map)
-    as.numeric(voters %*% m) / sum(voters)
+    vote_a = eval_tidy(enquo(group_pop), map)
+    vote_b = eval_tidy(enquo(other_pop), map)
+    voters = vote_a + vote_b
+    share = vote_a / (voters + 1e-6)
+
+    if (identical(kernel, k_step)) {
+        as.numeric(vote_a %*% m + vote_b %*% (1 - m)) / sum(voters)
+    } else {
+        u_a = as.numeric(vote_a %*% m)
+        u_b = as.numeric(vote_b %*% (1 - m))
+        # compute adjustment for vote shifts
+        rdgrp = round(distr_grp, 3)
+        vals = unique(as.numeric(rdgrp))
+        lookup = match(rdgrp, vals)
+
+        adj_a = ker_mean(-qlapl(vals), rep(Inf, length(vals))) * kernel(vals)
+        adj_b = ker_mean(qlapl(vals), rep(Inf, length(vals))) * kernel(1-vals)
+        slope = 1 - 2*abs(share - 0.5)
+        adj_a = as.numeric((voters * slope) %*% matrix(adj_a[lookup], nrow=nrow(distr_grp)))
+        adj_b = as.numeric((voters * slope) %*% matrix(adj_b[lookup], nrow=nrow(distr_grp)))
+
+        (u_a + u_b + adj_a + adj_b) / sum(voters)
+    }
 }
 
 #' Voter utility for global representation
 #'
 #' @param plans a `redist_plans` object.
 #' @param group column of `plans` containing the group share of each district.
[…]

### 2021-09-06 R/03_sim_nj.R

@@ -14,33 +14,44 @@ prop_seats = round(attr(nj, "ndists") * statewide)
 
 pl = calc_plans_stats(plans, nj, ndv, nrv)
 
-plot(nj, rowMeans(m_dem > 0.5)) + scale_fill_party_c()
+pdf("out/nj_vars.pdf", 9, 9)
+expl_vars(pl, e_dem, u_glb, u_loc, f, h_dem, h, egap, pbias, mean_med, competitive)
+dev.off()
 
-redist.plot.scatter(pl_sum, u_glb, h_dem - h_rep)
-redist.plot.scatter(pl_sum, u_loc, h_dem - h_rep)
-redist.plot.scatter(pl_sum, u_loc, u_glb)
-redist.plot.scatter(pl_sum, h_rep, h2_rep)
-redist.plot.scatter(pl_sum, h_dem - h_rep, h2_dem - h2_rep)
-redist.plot.scatter(pl_sum, h_dem - h_rep, h)
-redist.plot.scatter(pl_sum, h_dem - h_rep, n_dem_smooth)
-redist.plot.scatter(pl_sum, h, n_dem_smooth)
-redist.plot.scatter(pl_sum, h, u_loc)
-redist.plot.scatter(pl_sum, h, u_glb)
-redist.plot.scatter(pl_sum, u_glb*0.5 + u_loc*0.5, n_dem_smooth)
-redist.plot.scatter(pl_sum, u_loc, egap)
-redist.plot.scatter(pl_sum, u_loc, mean_med)
-redist.plot.scatter(pl_sum, h_dem - h_rep, egap)
+select(pl$plan, e_dem, u_glb, u_loc, f, egap, pbias, mean_med, competitive) %>%
+    #pairs(cex=0.2)
+    pairsD3::pairsD3(cex=1.5, opacity=0.5, group=pl$plan$n_dem)
 
-ggplot(pl_sum, aes(h_dem - h_rep, u_loc, color=as.factor(round(egap, 1)))) +
+
+plot(nj, rowMeans(pl$mat > 0.5)) + scale_fill_party_c()
+
+
+idx_loc = as.integer(as.character(filter(pl$plan, u_loc == max(u_loc))$draw[1]))
+idx_glb = as.integer(as.character(filter(pl$plan, u_glb == max(u_glb))$draw[1]))
+idx_both = mutate(pl$plan, score = scale(u_loc)[,1] +scale(u_glb)[,1]) %>%
[…]

### 2021-09-06 R/04_sim_wi.R

@@ -1,0 +1,49 @@
+wi = read_rds(here("data/WI_cd_final_vtd_20.rds")) %>%
+    redist_map(pop_tol=0.01, ndists=8, adj=.$adj) %>%
+    mutate(across(contains("dv"), ~ coalesce(., 0)),
+           across(contains("rv"), ~ coalesce(., 0)))
+
+if (!file.exists(sim_path <- here("data/wi_sims.rds"))) {
+    N_sim = 1000
+    plans = redist_smc(wi, N_sim, counties=county, pop_temper=0.01, verbose=TRUE)
+    write_rds(plans, sim_path, compress="xz")
+} else {
+    plans = read_rds(sim_path)
+}
+
+statewide = with(wi, sum(ndv)/sum(ndv+nrv))
+prop_seats = round(attr(wi, "ndists") * statewide)
+
+pl = calc_plans_stats(plans, wi, ndv, nrv)
+
+pdf("out/wi_vars.pdf", 9, 9)
+expl_vars(pl, e_dem, u_glb, u_loc, f, h_dem, h, egap, pbias, mean_med, competitive)
+dev.off()
+
+plot(wi, rowMeans(pl$mat > 0.5)) + scale_fill_party_c()
+plot(wi, rowMeans(pl$mat)) + scale_fill_party_c(limits=c(0.3, 0.7))
+
+idx_loc = as.integer(as.character(filter(pl$plan, u_loc == max(u_loc))$draw[1]))
+idx_glb = as.integer(as.character(filter(pl$plan, u_glb == max(u_glb))$draw[1]))
+idx_both = mutate(pl$plan, score = scale(u_loc)[,1] +scale(u_glb)[,1]) %>%
+    filter(score == max(score)) %>%
+    pull(draw) %>%
+    `[`(1) %>%
+    as.character() %>%
+    as.integer()
+
+p1 = plot_cds(wi, as.matrix(pl$distr)[,idx_loc], county, "WI") + labs(title="Best Local")
+p2 = plot_cds(wi, as.matrix(pl$distr)[,idx_glb], county, "WI") + labs(title="Best Global")
+p3 = plot_cds(wi, as.matrix(pl$distr)[,idx_both], county, "WI") + labs(title="Best Overall")
+p3 + {if (idx_both != idx_loc) p1 } + {if (idx_both != idx_glb) p2 } +
+    plot_layout(guides="collect") & theme(legend.position="bottom")
+
+
+pl$distr %>%
[…]

### 2021-09-06 R/toy/01_util.R

@@ -1,9 +1,9 @@
 pl_opt = list(
     mininois = c(2, 2, 3, 3, 3, 3,
                  2, 2, 3, 3, 3, 3,
+                 2, 2, 3, 3, 3, 3,
                  2, 2, 2, 3, 3, 3,
-                 2, 2, 2, 3, 3, 3,
-                 1, 1, 1, 3, 3, 3,
+                 2, 1, 1, 3, 3, 3,
                  1, 1, 1, 3, 3, 3),
     minissouri = c(3, 3, 3, 3, 3, 3,
                    2, 2, 3, 3, 3, 3,

### 2021-09-06 R/toy/03_sims.R

@@ -1,17 +1,15 @@
-pl_mininois = redist_smc(mininois, 2000, verbose=F)
-
-opt = redist_shortburst(mininois, scorer_group_pct(mininois, dem, pop, 3))
+pl_mininois = redist_smc(mininois, 2000, verbose=F) %>%
+    add_reference(pl_opt$mininois, "opt") %>%
+    calc_plans_stats(mininois, dem, rep, ker=k_step)
 
 pl_minissouri = redist_smc(minissouri, 2000, verbose=F) %>%
     add_reference(pl_opt$minissouri, "opt") %>%
-    calc_plans_stats(minissouri, dem, rep)
+    calc_plans_stats(minissouri, dem, rep, ker=k_step)
 
+redist.plot.scatter(pl_mininois$plan, u_glb, u_loc)
+redist.plot.scatter(pl_mininois$plan, n_dem, u_glb)
 
-redist.plot.scatter(pl_sum_mini, u_glb, h_dem - h_rep)
-redist.plot.scatter(pl_sum_mini, u_loc, h_dem - h_rep)
-redist.plot.scatter(pl_sum_mini, u_loc, u_glb)
-redist.plot.scatter(pl_sum_mini, h_dem - h_rep, h)
-redist.plot.scatter(pl_sum_mini, h_dem - h_rep, n_dem_smooth)
-redist.plot.scatter(pl_sum_mini, h, n_dem_smooth)
-redist.plot.scatter(pl_sum_mini, u_glb*0.15 + u_loc*0.85, n_dem_smooth)
-redist.plot.scatter(pl_sum_mini, h_dem - h_rep, egap)
+idxs = c(1, with(pl_mininois$plan, which(u_loc==max(u_loc))))
+dm = plan_distances(pl_mininois$distr)[idxs, idxs]
+
+as.matrix(pl_mininois$plan)[,110]

### 2021-09-07 paper/paper.Rmd
outlining

@@
 bibliography: references.bib
-biblio-style: unsrt
+biblio-style: apalike
 output: 
@@
 Define the total utility for this rule as \[
-    u_{\mathrm{tot}}^{(G)}(f, e) \dfeq \sum_{v\in\Voters} g\qty(f\qty(s_{\vote(v, e)})).
+    u_{\mathrm{tot}}^{(G)}(f, v) \dfeq \sum_{i\in\Voters} g\qty(f\qty(s_{v(i)})).
 \] 
@@
     u^{(L)}:\Voters\times\Elecs\times\Plans\to\R, 
-    \qquad u^{(L)}(i,v,q) = \ind\{v(i)=\win(q(i), e, q)\},
+    \qquad u^{(L)}(i,v,q) = \ind\{v(i)=\win(q(i), v, q)\},
 $$that is, a voter gains utility only if they winner of their legislative district belongs to their preferred political party.
@@
 But the general tendency of this utility measure (in the context of a single election, at least), like the global utility measure, will be to push plans towards proportionality.
-The local measure will additionally reward what have been called *bipartisan gerrymanders*, where there are few or no competitive districts.
+Unlike with the global measure, a specific kind of proportionality is preferred: that which creates as many lopsided districts as possible (what is often termed a *bipartisan gerrymander*).
+This point is discussed further in Section \ref{sec:competitive} below.
 
@@
 We will also consider a voter $I$ taken uniformly at random from $\Voters$, with $I\indep\pi$.
+This has the effect of treating voters exchangeably and valuing their personal utilities equally.
 
@@
     U^{(L)} \dfeq u^{(L)}(I, V, Q).
-$$Thus, the utility for a single voter $v$ becomes $U^{(L)}\mid I=i$; the electorate's average utility for election outcome $e$ becomes $\E[U^{(L)}\mid V=v]$, and so on.
+$$Thus, the utility for a single voter $v$ becomes $U^{(L)}\mid I=i$; the electorate's average utility for election outcome $v$ becomes $\E[U^{(L)}\mid V=v]$, and so on.
 
@@
     h(i,v,q,\tilde q) &= (1 - u^{(L)}(i, v, q))\cdot u^{(L)}(i, v, \tilde q) \\
-    &= \ind\{v(i)\neq\win(q(i), e, q)\}
-        \ind\{v(i)=\win(\tilde q(i), e, \tilde q)\},
+    &= \ind\{v(i)\neq\win(q(i), v, q)\}
+        \ind\{v(i)=\win(\tilde q(i), v, \tilde q)\},
 \end{align*}that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.
@@
 In fact, a harm measure defined analogously using global utility would essentially be a difference in seats won by a voter's party between plans $q$ and $\tilde q$ (the exact form would be affected by the function $g$).
-This measure is not particularly interesting, and nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation **[CM: cite Rucho & LWV of PA]**.
-**[CM: revisit ¶ based on NJ findings].**
+This measure is not particularly interesting, and nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation [@2018league; @2019rucho; @2019lewis].
+**[CM: revisit ¶ based on NJ findings]****.**
 
@@
 
+Despite being derived from a local harm measure, $F(Q)$ is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
+Yet unlike a simple measure of seats, $F(Q)$ is shifted to an interpretable scale, where a value of zero corresponds to partisan fairness.
+In other words, $F(Q)$ provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
+This connection further supports the use of total seats as a summary statistic in redistricting simulation methods, as advocated by @deford2020symmetry.
+
+```{=tex}
+\begin{restatable}{prop}{propfairness} \label{prop:fairness}
+Basically, if turnout is equal across districts, for a districting plan $q\in\Plans$ and political party $p\in\Parties$, we have
+\begin{align*}
+    F(q) = \E[\seats(p, V, Q)\mid Q=q] - \E[\seats(p, V, \tilde Q)].
+\end{align*}
+
+\end{restatable}
+```
 # Calculating the Harm and Fairness Measures {#sec:estimating}
@@
 
+1.  discrete and no universal metric structure
+
+2.  algorithms developed to sample from maximum-entropy distributions [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020]
+
+3.  impose constraints required by law and tradition; then samples are exchangeable
+
 ## Modelling Election Outcomes
 
-# Connections to Partisan Symmetry and Fairness Standards {#sec:connections}
+1.  complex, with many unknown future factors
 
-# Differences Between Classes of Fairness Metrics
+2.  need simple model that will capture primary factor (nationwide political environment).
+    Ideally would be easy to implement
+
+3.  we propose **Proportionate Uniform Shift**, where a constant fraction of voters of one party are assumed to switch to the other.
+
+    1.  Not mathematically equivalent Butler's Uniform Partisan Swing [@butler1951swing], but near-identical for most precincts and districts.
+
+    2.  Highly accurate in out-of-sample prediction, considering simplicity [@katz2020].
+
+    3.  Can be viewed as mapping vote shares through a Laplace CDF centered at 0.5 with rate 2
+
+    4.  Model shift amounts as Normal.
+        Fit to historical variability in nationwide vote shares to get $\mathrm{t}_{22}(0, 0.03723)$.
+
+4.  More complicated models certainly possible and straightforward (if not particularly efficient) to compute.
+    Could e.g. model district deviations as a systematic national shift plus a local idiosyncratic effect, like the Stochastic Uniform Partisan Swing of @gelman1994unified.
+
+5.  Formulas to actually compute these things are in the Appendix:
+
+    1.  Global utility is discrete, so we can compute the range of shifts that produce each number of seats, and the average vote share for those ranges.
+        Combine these
+
+    2.  To be computationally efficient, local utility requires linearizing the vote share curve.
+
+    3.  For voter fairness, PUS means that relative share of party's voters in each precinct remains the same.
+        Need only to apply a kernel function (composition of Laplace ICDF + model shift CDF)
+
+# Connections to Partisan Symmetry and Fairness, Standards {#sec:connections}
+
+Over the past decade [**CM: correct?!**], researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
+Specifically, these measures aim to quantify deviations from partisan symmetry, fairness, or competitiveness.
+Some of the most commonly used are perhaps the *efficiency gap* [@stephanopoulos2015], *partisan bias* [@tufte1973pb; @king1987democratic], and the *mean-median* measure [@mcdonald2015unfair; @wang2016three].
+All of these measures may be estimated using district vote shares.
+
+## The Efficiency Gap
+
+1.  has been criticized [@tamcho2017egap; @bernstein2017egap; @deford2020symmetry]
+
+## Partisan Bias
+
+1.  Integrating yields a related measure of @nagle2015measures, which has also been criticized by @deford2020symmetry.
+2.  Focuses on only one part of S-V curve, which may not be relevant
+
+## Mean-Median
+
+1.  Like PB, focuses on only one part of S-V curve, which may not be relevant
+
+## Districting Plan Competitiveness {#sec:competitive}
+
+The notion of district *competitiveness*, while related to the slope of the seats-vote curve around $\half$, is not the same as partisan symmetry.
+However, one commonly used measure of competitiveness, from @tamcho2016, is related to our local utility measure $U^{(L)}$.
+The measure of @tamcho2016, which we will denote by $\mathrm{comp}_\alpha(v,q)$ contains two terms: one, $T_p(v,q)$, measuring the closeness of district vote shares to 50%, and the other, $T_e(v,q)$, measuring the closeness of the overall share of seats to 50%; the relative importance of these two terms is controlled by $\alpha$.
+It is this first term which his closely related to $U^{(L)}$, as stated in Proposition \ref{prop:comp} below.
+
+```{=tex}
+\begin{restatable}{prop}{propcomp} \label{prop:comp}
+Basically, if turnout is equal across districts, for an election $v\in\Elecs$ and districting plan $q\in\Plans$ and two political parties, we have
+\begin{align*}
+    \E[U^{(L)}\mid V=v, Q=q] &= \frac{1}{2|D|}\sum_{d\in D}
+        \qty|\frac{\votes^{(p_a)}}{\votes^{(p_b)} + \votes^{(p_b)}}-\half| \\
+    &= \half T_p(v,q) = \frac{\mathrm{comp}_\alpha(v,q)}{1+\alpha T_e(v_q)}.
+\end{align*}
+
+\end{restatable}
+```
+Importantly, $\mathrm{comp}_\alpha(v,q)$ is designed to be *minimized*---smaller values correspond to more competitive districts.
+But, as briefly discussed above, this actually runs counter to the local utility measure, which encourages lopsided districts with as many similarly-minded voters as possible.
+@brunell2010 discusses this point in much more detail, arguing that competitive districts should be avoided where possible.
+Buttressing our choice of local utility function, he analyzes ANES survey data to show that voters approve more highly of their representative and of Congress overall when their representative votes in alignment with their preferences.
 
 # Real-world Example {#sec:empirical}
+
+1.  New Jersey
+
+    1.  Not a 50-50 state
+
+    2.  moderate number of districts
+
+```{r nj-vars, fig.cap="Various measures for an ensemble of New Jersey districting plans.", fig.width=6.5, fig.height=6.5}
+include_graphics(here("out/nj_vars.pdf"))
+```
 
@@
 \end{proof}
+
+\propfairness*
+
+\begin{proof}
+\end{proof}
+
+\propcomp*
+
+\begin{proof}
+\end{proof}
 ```

### 2021-09-07 R/01_metrics.R

@@ -173,19 +173,3 @@ calc_plans_stats = function(plans, map, dem, gop, ker=k_t()) {
 
     list(distr=select(plans, draw:total_pop, dem), plan=pl_sum, mat=m_dem)
 }
-
-
-# global utility scorer
-scorer_util_g = function(map, group_pop, total_pop) {
-    group_pop = eval_tidy(enquo(group_pop), map)
-    total_pop = eval_tidy(enquo(total_pop), map)
-    n1 = sum(group_pop)
-    n2 = sum(total_pop) - n1
-    ndists = attr(map, "ndists")
-    fn = function(plans) {
-        seats = colSums(redist:::group_pct(plans, group_pop, total_pop, ndists) > 0.5)
-        (flog(seats)*n1 + flog(ndists - seats)*n2) / (n1 + n2)
-    }
-    class(fn) = c("redist_scorer", "function")
-    fn
-}

### 2021-09-07 paper/paper.Rmd
comment out temp figure

@@
 This measure is not particularly interesting, and nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation [@2018league; @2019rucho; @2019lewis].
-**[CM: revisit ¶ based on NJ findings]****.**
+**[CM: revisit ¶ based on NJ findings].**
 
@@
 ```{r nj-vars, fig.cap="Various measures for an ensemble of New Jersey districting plans.", fig.width=6.5, fig.height=6.5}
-include_graphics(here("out/nj_vars.pdf"))
+#include_graphics(here("out/nj_vars.pdf"))
 ```

### 2021-09-07 paper/paper.Rmd
legal cites better

@@
   - individual standing
+citation_package: biblatex
 bibliography: references.bib
-biblio-style: apalike
+biblio-style: apa
 output: 
@@
 However, one commonly used measure of competitiveness, from @tamcho2016, is related to our local utility measure $U^{(L)}$.
-The measure of @tamcho2016, which we will denote by $\mathrm{comp}_\alpha(v,q)$ contains two terms: one, $T_p(v,q)$, measuring the closeness of district vote shares to 50%, and the other, $T_e(v,q)$, measuring the closeness of the overall share of seats to 50%; the relative importance of these two terms is controlled by $\alpha$.
+The measure of @tamcho2016, which we will denote by $\mathrm{comp}_\alpha(v,q)$, contains two terms: one, $T_p(v,q)$, measuring the closeness of district vote shares to 50%, and the other, $T_e(v,q)$, measuring the closeness of the overall share of seats to 50%; the relative importance of these two terms is controlled by $\alpha$.
 It is this first term which his closely related to $U^{(L)}$, as stated in Proposition \ref{prop:comp} below.
@@
 
+# References {.unnumbered}
+
+::: {#refs}
+:::
+
+\clearpage
+
 \appendix
@@
 ```
+# Computing Utility, Harm, and Fairness from Precinct-level Election Data
+
+formulas here.

### 2021-09-08 paper/paper.Rmd
typo
calc/estimation section drafted

@@
 It depends on the form of $g$, as the following result describes.
+Proofs of this and all other propositions are deferred to Appendix \ref{app:proofs}.
 
@@
 This measure is not particularly interesting, and nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation [@2018league; @2019rucho; @2019lewis].
-**[CM: revisit ¶ based on NJ findings].**
+**[CM: revisit ¶ based on NJ findings]****.**
 
@@
 
-The expected harm to an individual $i\in\Voters$ voter then becomes $$
+The expected harm to an individual $i\in\Voters$ then becomes $$
    \E[H\mid I=i] = \Pr[V(i)\neq\win(Q(i), V, Q)
@@
 
-The measures $H$ and $F(q)$ defined above require specification of a probability measure on elections and districting plans, $\pi$.
+The measures $U^{(G)}, U^{(L)}, H$ and $F$ defined above require specification of a probability measure on elections and districting plans, $\pi$.
 Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
-Either way, a collection of past election data and an enacted districting plan alone are not enough to calculate $H$ and $F(q)$.
+Either way, a collection of past election data and an enacted districting plan alone are not enough to calculate these measures.
 
-While not strictly necessary, modelling of $V$ and $\tilde Q$ independently will simplify the situation considerably.
+While not strictly necessary, modeling of $V$ and $\tilde Q$ independently will simplify the situation considerably.
 Practically, this translates to the assumption that individual voters would vote for the same political party no matter the districting plan.
@@
 
-## Modelling Districting Plans
+## Modeling Districting Plans
 
-1.  discrete and no universal metric structure
+It is difficult to construct a tractable probability distribution for districting plans, since the space of plans is discrete and lacks a natural metric structure.
+Thankfully, in recent years a number of algorithms have been developed which sample districting plans from specific distributions [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
+These distributions can generally be characterized as maximum-entropy distributions on the space of all valid districting plans, in combination with several moment constraints.
 
-2.  algorithms developed to sample from maximum-entropy distributions [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020]
+For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans which are geographically contiguous, have district populations within a deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
+These maximum-entropy distributions put as few restraints as possible on the districting sampling process, subject to a set of constraints imposed by the researcher.
+If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by partisan considerations, to use in estimating $H$ and $F$.
 
-3.  impose constraints required by law and tradition; then samples are exchangeable
+## Modeling Election Outcomes
 
-## Modelling Election Outcomes
+Election results present a different set of modeling challenges, given the large number of factors that influence the outcome, and the general unpredictability of future events.
+As far as estimating measures like $U^{(G)}, U^{(L)}, H$, and $F$, we would ideally use a relatively simple model that captures the primary drivers of election outcomes.
+In the interest of fairness, the model should not err on the side of simplicity and symmetry, and avoid making any biasing assumptions about the future (e.g., that demographic shifts will favor one party, or that past trends in election returns will continue in the same direction).
 
-1.  complex, with many unknown future factors
+We propose a *proportionate uniform shift* model, which focuses on swings in the national political environment as the main determinant of election outcomes.
+Each election outcome over the period of interest (generally ten years or less) is modeled as a random deviation from a baseline outcome, and these deviations are assumed to consist of a constant fraction of voters of one party in each precinct switching to the other.
+The baseline itself can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
 
-2.  need simple model that will capture primary factor (nationwide political environment).
-    Ideally would be easy to implement
+The proportionate uniform shift model is not mathematically equivalent to the uniform partisan swing model of @butler1951swing, but is near-identical for most precincts and districts.
+Despite their simplicity, uniform-swing--type models are highly accurate for out-of-sample prediction [@katz2020], and are intuitive and easy to implement.
 
-3.  we propose **Proportionate Uniform Shift**, where a constant fraction of voters of one party are assumed to switch to the other.
+The proportionate shift, where we assume that a constant fraction of *voters of one party* in each precinct switch to the other party, has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
+First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
+Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
+For measures like $H$ and $F$ which consider voters in each party separately, this helpfully simplifies computations.
 
-    1.  Not mathematically equivalent Butler's Uniform Partisan Swing [@butler1951swing], but near-identical for most precincts and districts.
+The proportionate shift can be mathematically framed as linear shift on a transformed scale, analogous to the generalized linear model setting.
+Specifically, we define $\gamma:[0,1]\to\R$, $$
+\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
+$$ which is the inverse CDF of a Laplace with rate 2.
+Then a party's vote share in precinct $k$, notated here as $s_k$, is modeled as $$
+\gamma(s_k) = \gamma(\overline s_k) + \eps, \qquad \eps \sim \Norm(0, \sigma^2_{\mathrm{hist}}),
+$$ where $\overline{s}_k$ is the party baseline, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
+For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.
 
-    2.  Highly accurate in out-of-sample prediction, considering simplicity [@katz2020].
-
-    3.  Can be viewed as mapping vote shares through a Laplace CDF centered at 0.5 with rate 2
-
-    4.  Model shift amounts as Normal.
-        Fit to historical variability in nationwide vote shares to get $\mathrm{t}_{22}(0, 0.03723)$.
-
-4.  More complicated models certainly possible and straightforward (if not particularly efficient) to compute.
-    Could e.g. model district deviations as a systematic national shift plus a local idiosyncratic effect, like the Stochastic Uniform Partisan Swing of @gelman1994unified.
-
-5.  Formulas to actually compute these things are in the Appendix:
-
-    1.  Global utility is discrete, so we can compute the range of shifts that produce each number of seats, and the average vote share for those ranges.
-        Combine these
-
-    2.  To be computationally efficient, local utility requires linearizing the vote share curve.
-
-    3.  For voter fairness, PUS means that relative share of party's voters in each precinct remains the same.
-        Need only to apply a kernel function (composition of Laplace ICDF + model shift CDF)
+More complicated election outcome models are certainly possible and straightforward to use with the measures proposed here.
+For example, one could model district deviations as a systematic national shift plus a local idiosyncratic effect, as is done in the Stochastic Uniform Partisan Swing model of @gelman1994unified.
+For ease and efficiency of computation, we stick to the simple proportionate uniform shift model here.
+Appendix \ref{app:formulas} contains the specific formulas used to compute $U^{(G)}, U^{(L)}, H$, and $F$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
 
@@
 
-# Proofs of Propositions
+# Proofs of Propositions {#app:proofs}
 
@@
 ```
-# Computing Utility, Harm, and Fairness from Precinct-level Election Data
+# Computing Utility, Harm, and Fairness from Precinct-level Election Data {#app:formulas}
 
-formulas here.
+1.  Global utility is discrete, so we can compute the range of shifts that produce each number of seats, and the average vote share for those ranges.
+    Combine these
+
+2.  To be computationally efficient, local utility requires linearizing the vote share curve.
+
+3.  For voter fairness, PUS means that relative share of party's voters in each precinct remains the same.
+    Need only to apply a kernel function (composition of Laplace ICDF + model shift CDF)
+
+Formulas here.

### 2021-09-08 R/03_sim_nj.R

@@ -23,6 +23,30 @@ select(pl$plan, e_dem, u_glb, u_loc, f, egap, pbias, mean_med, competitive) %>%
     pairsD3::pairsD3(cex=1.5, opacity=0.5, group=pl$plan$n_dem)
 
 
+d_plot = pl$plan %>%
+    mutate(`terc_Local Utility` = ntile(u_loc, 3),
+           `terc_Global Utility` = ntile(u_glb, 3),
+           terc_Fairness = ntile(f, 3)) %>%
+    select(draw, n_dem, e_dem, u_loc, u_glb, f, starts_with("terc_")) %>%
+    as_tibble()
+d_plot = number_by(pl$distr, dem) %>%
+    left_join(d_plot, by="draw") %>%
+    pivot_longer(starts_with("terc_"), names_to="qty",
+                 values_to="tercile", names_prefix="terc_")
+
+ggplot(d_plot, aes(as.factor(district), dem, fill=as.factor(tercile))) +
+    facet_grid(qty ~ .) +
+    geom_hline(yintercept=0.5, lty="dashed") +
+    geom_boxplot(size=0.3, outlier.size=0.1) +
+    scale_y_continuous("Democratic two-party share", labels=percent) +
+    labs(x="Districts, ordered by Democratic share",
+         fill="Tercile\nof measure") +
+    scale_fill_wa_d("sound_sunset", which=c(1, 8, 13)) +
+    theme_repr()
+ggsave("paper/figures/nj_terciles.pdf", width=6.5, height=6)
+
+
+
 plot(nj, rowMeans(pl$mat > 0.5)) + scale_fill_party_c()
 
 

### 2021-09-11 paper/paper.Rmd
add intro; join refs

@@
 
-What is the question?
-
 Districting is a complicated process.
-Residents of a state or locality must be partitioned into districts.
+Residents of a state or locality must be partitioned into districts based on the geographies that they live in.
 These districts must adhere to many constraints, often requiring components such as contiguity and equal population.
@@
 
-Why does it matter?
+Decades of literature on redistricting have focused on the fairness to parties.
+This literature defers to symmetric outcomes for parties, commonly referred to as partisan symmetry \cite{a:bunch:of:king}.
+Partisan symmetry is a useful idea which greatly improves academic understanding of the redistricting process.
+We Why is it hard?
+Understanding the redistricting process requires a feasible and reliable understanding of the counterfactual set of redistricting plans.
+Early methodologies for simulating plans relied on randomness to generate many different plans and enhanced understanding of redistricting greatly \\citep{ciri:etal:00, chen:rodd:13, defO:duch:solo:21.
+These methods greatly enhanced the understanding of redistricting within contexts of a particular state with set populations.
+Yet, without a known target distribution, these heuristic approaches to simulation fall do not offer necessary guarantees for confidence in the simulated plans to represent the total space of possible plans \cite{fife:etal:20b}.
+Recent advances in Markov Chain methods as applied to redistricting allow us move beyond simulation towards sampling plans from this definable space \\cite{fife:etal:20, mcca:imai:20, autr:etal:21}.
 
-Why is it hard?
-Decades of
+Using sampling methods, we can then develop a reliable counterfactual that can be tuned to consider the samping constraints imposed by the laws and practices of a state.
+This counterfactual allows us to characterize the likely plans from a state.
+Often, we can would then call this an ensemble analysis, which is a flexible approach to sampling redistricting plans by simulation, relying on large numbers of plans for stability \cite{matt:etal:21} or sufficient diversity of nearly-independent plans \cite{kenn:etal:forth}.
+using this counterfactual, we thus create measures of harm, utility, and fairness which are based on a single plan compared to a set of other reasonable and feasible plans which encorporate the necessary legal constraints to make a plan viable.
 
-What did we do?
+We offer a reminder that geographic context is key to understanding common problems in political science..
+Elections themselves are geographically determined.
+Recent research offers important resolution of the degree of partisan sorting in the US \cite{brow:enos:something}.
+By combining those understandings with the necessities of redistricting, we are able to expand the localized aspect of harm.
+Further, we suggest that while the efficiency gap may not incorporate geographic information, it often helps capture individual harm better than other common measures.
+This is a great benefit as an indicator to detect extreme gerrymanders in cases where sufficiently local geographic data to simulate plans is not available.\
+We further clarify that common existing measures of partisan fairness -- while they may capture the difficulties offered to a party -- do not necessarily capture the representational harms incurred by individuals who live in districts which do not at all represent their preferences.
 
-What did we learn?
-
-What is the paper's contribution?
+Our metrics offer a framework for understanding approximately individual harms, benefits, and net fairness.
+Using counterfactual plans to understand this space allows us to properly calibrate understandings of the impact of redistricting plans on individual understanding.
+Thus, we create measures which capture the possibility that no plans exist which favor one community \cite{duch:massachusetts}.
+Using this class of metrics which describe the individual affect via representation allow us to better align legalistic interpretations of harm with political science understandings.
 

### 2021-09-11 paper/paper.Rmd
NJ work

--- 2021-09-08 paper/paper.Rmd
@@
 
-# Real-world Example {#sec:empirical}
+# Application: New Jersey {#sec:empirical}
 
-1.  New Jersey
+To illustrate the connections in the preceding section and to demonstrate the value of the proposed measures in a real-world setting, we provide a brief analysis of congressional districting in the state of New Jersey.
+As discussed above, many gerrymandering metrics function well in highly competitive states, but may produce more counterintuitive results in other states.
+New Jersey, with an average Democratic two-party vote share of 56.6% across our two baseline elections (2016 presidential and 2018 senatorial), will demonstrate that this is not a problem unique to lopsided states.
+Additionally, New Jersey has 12 congressional districts, which is enough to produce some variability in seats outcomes across counterfactual plans, but not so many as to render trivial the challenges of discreteness.
 
-    1.  Not a 50-50 state
-
-    2.  moderate number of districts
+```{r nj-maps, fig.cap="Baseline vote patterns, and two hypothetical gerrymanders.", fig.width=6.5, fig.height=3.5}
+include_graphics(here("paper/figures/nj_maps.pdf"))
+```
 

### 2021-09-11 R/00_setup_utils.R

@@ -96,14 +96,14 @@ plot_cds = function(map, pl, county, abbr, qty="ndv") {
         theme_void()
 }
 
-expl_vars = function(pl, ...) {
+expl_vars = function(pl, labels, ...) {
     panel.hist <- function(x, ...) {
         usr <- par("usr"); on.exit(par(usr))
         par(usr = c(usr[1:2], 0, 1.5) )
         h <- hist(x, plot = FALSE)
         breaks <- h$breaks; nB <- length(breaks)
         y <- h$counts; y <- y/max(y)
-        rect(breaks[-nB], 0, breaks[-1], y, col="gray")
+        rect(breaks[-nB], 0, breaks[-1], y, family="Times", col="gray")
     }
     panel.cor <- function(x, y, digits = 2, prefix = "", ...) {
         usr <- par("usr"); on.exit(par(usr))
@@ -111,17 +111,17 @@ expl_vars = function(pl, ...) {
         r <- abs(cor(x, y, use="complete.obs"))
         txt <- format(c(r, 0.123456789), digits = digits)[1]
         txt <- paste0(prefix, txt)
-        text(0.5, 0.5, txt, cex=2.5*sqrt(r), offset=0, adj=c(0.5, 0.5))#cex.cor)
+        text(0.5, 0.5, txt, cex=2.5*sqrt(r), family="Times", offset=0, adj=c(0.5, 0.5))#cex.cor)
     }
     panel.points <- function(x, y, ...) {
-        points(x, y, ...)
+        points(x, y, family="Times", ...)
         abline(h=0, v=0, lty="dashed")
     }
-    seats_vec = as.integer(as.factor(pl$plan$n_dem))
-    select(pl$plan, ...) %>%
-        pairs(cex=0.2, gap=0.5,
-              lower.panel=panel.cor, diag.panel=panel.hist, upper.panel=panel.points,
-              col=wa_pal("sea_star", n=max(seats_vec))[seats_vec])
+    seats_vec = as.integer(as.factor(pl$n_dem))
+    pairs(select(pl, -n_dem), labels=labels,
+          cex=0.2, gap=0.5, family="Times", cex.labels=0.9, ...,
+          lower.panel=panel.cor, diag.panel=panel.hist, upper.panel=panel.points,
+          col=wa_pal("sea_star", n=max(seats_vec))[seats_vec])
 }
 
 # downloads data for state `abbr` to `folder/{abbr}_2020_*.csv` and returns path to file

### 2021-09-11 R/03_sim_nj.R

@@ -4,7 +4,17 @@ nj = read_rds(here("data/NJ_cd_final_vtd_20.rds")) %>%
 if (!file.exists(sim_path <- here("data/nj_sims.rds"))) {
     N_sim = 1000
     plans = redist_smc(nj, N_sim, counties=county, pop_temper=0.01, verbose=TRUE)#FALSE)
-    write_rds(plans, sim_path, compress="xz")
+
+    # gerrymanders
+    set.seed(5118)
+    opt_dem = redist_shortburst(nj, scorer_group_pct(nj, ndv, ndv+nrv, 11))
+    set.seed(5118)
+    opt_rep = redist_shortburst(nj, scorer_group_pct(nj, nrv, ndv+nrv, 6))
+
+    plans %>%
+        add_reference(last_plan(opt_dem), "dem_gerry") %>%
+        add_reference(last_plan(opt_rep), "rep_gerry") %>%
+        write_rds(sim_path, compress="xz")
 } else {
     plans = read_rds(sim_path)
 }
@@ -14,8 +24,38 @@ prop_seats = round(attr(nj, "ndists") * statewide)
 
 pl = calc_plans_stats(plans, nj, ndv, nrv)
 
+# Initial plots
+p1 = ggplot(nj, aes(fill=ndv/(ndv+nrv))) +
+    geom_sf(size=0, color="#00000000") +
+    scale_fill_party_c(limits=c(0.15, 0.85)) +
+    labs(title="(a) Partisan patterns") +
+    theme_repr_map() +
+    theme(plot.title = element_text(hjust = 0.5))
+p2 = plot_cds(nj, as.matrix(plans)[,"dem_gerry"], county, "NJ") +
+    labs(title="(b) Democratic gerrymander") +
+    scale_fill_party_c(limits=c(0.15, 0.85)) +
+    theme_repr_map() +
+    theme(plot.title = element_text(hjust = 0.5))
+p3 = plot_cds(nj, as.matrix(plans)[,"rep_gerry"], county, "NJ") +
+    labs(title="(c) Republican gerrymander") +
+    scale_fill_party_c(limits=c(0.15, 0.85)) +
+    theme_repr_map() +
+    theme(plot.title = element_text(hjust = 0.5))
+p = p1 + p2 + p3 + plot_layout(guides="collect")
+ggsave(here("paper/figures/nj_maps.pdf"), plot=p, width=8, height=4.5)
+
+
+
[…]

### 2021-09-11 paper/paper.Rmd
merge

@@
   - individual standing
-citation_package: biblatex
 bibliography: references.bib
-biblio-style: apa
+biblio-style: apalike
 output: 
@@
 Decades of literature on redistricting have focused on the fairness to parties.
-This literature defers to symmetric outcomes for parties, commonly referred to as partisan symmetry \cite{a:bunch:of:king}.
+This literature defers to symmetric outcomes for parties, commonly referred to as partisan symmetry [@katz2020].
 Partisan symmetry is a useful idea which greatly improves academic understanding of the redistricting process.
-We Why is it hard?
+Why is it hard?
 Understanding the redistricting process requires a feasible and reliable understanding of the counterfactual set of redistricting plans.
-Early methodologies for simulating plans relied on randomness to generate many different plans and enhanced understanding of redistricting greatly \\citep{ciri:etal:00, chen:rodd:13, defO:duch:solo:21.
+Early methodologies for simulating plans relied on randomness to generate many different plans and enhanced understanding of redistricting greatly [@ciri:etal:00; @chen:rodd:13; @defO:duch:solo:21].
 These methods greatly enhanced the understanding of redistricting within contexts of a particular state with set populations.
-Yet, without a known target distribution, these heuristic approaches to simulation fall do not offer necessary guarantees for confidence in the simulated plans to represent the total space of possible plans \cite{fife:etal:20b}.
-Recent advances in Markov Chain methods as applied to redistricting allow us move beyond simulation towards sampling plans from this definable space \\cite{fife:etal:20, mcca:imai:20, autr:etal:21}.
+Yet, without a known target distribution, these heuristic approaches to simulation fall do not offer necessary guarantees for confidence in the simulated plans to represent the total space of possible plans [@fifield2020essential].
+Recent advances in Markov Chain methods as applied to redistricting allow us move beyond simulation towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @deford2019; @mccartan2020].
 
-Using sampling methods, we can then develop a reliable counterfactual that can be tuned to consider the samping constraints imposed by the laws and practices of a state.
+Using sampling methods, we can then develop a reliable counterfactual that can be tuned to consider the sampling constraints imposed by the laws and practices of a state.
 This counterfactual allows us to characterize the likely plans from a state.
 Often, we can would then call this an ensemble analysis, which is a flexible approach to sampling redistricting plans by simulation, relying on large numbers of plans for stability \cite{matt:etal:21} or sufficient diversity of nearly-independent plans \cite{kenn:etal:forth}.
-using this counterfactual, we thus create measures of harm, utility, and fairness which are based on a single plan compared to a set of other reasonable and feasible plans which encorporate the necessary legal constraints to make a plan viable.
+using this counterfactual, we thus create measures of harm, utility, and fairness which are based on a single plan compared to a set of other reasonable and feasible plans which incorporate the necessary legal constraints to make a plan viable.
 
@@
 Further, we suggest that while the efficiency gap may not incorporate geographic information, it often helps capture individual harm better than other common measures.
-This is a great benefit as an indicator to detect extreme gerrymanders in cases where sufficiently local geographic data to simulate plans is not available.\
+This is a great benefit as an indicator to detect extreme gerrymanders in cases where sufficiently local geographic data to simulate plans is not available.
 We further clarify that common existing measures of partisan fairness -- while they may capture the difficulties offered to a party -- do not necessarily capture the representational harms incurred by individuals who live in districts which do not at all represent their preferences.

### 2021-09-11 paper/paper.Rmd
fix refs

@@
 Understanding the redistricting process requires a feasible and reliable understanding of the counterfactual set of redistricting plans.
-Early methodologies for simulating plans relied on randomness to generate many different plans and enhanced understanding of redistricting greatly [@ciri:etal:00; @chen:rodd:13; @defO:duch:solo:21].
+Early methodologies for simulating plans relied on randomness to generate many different plans and enhanced understanding of redistricting greatly [@ciri:darl:orou:00; @chen:rodd:13; @deford2019].
 These methods greatly enhanced the understanding of redistricting within contexts of a particular state with set populations.
@@
 This counterfactual allows us to characterize the likely plans from a state.
-Often, we can would then call this an ensemble analysis, which is a flexible approach to sampling redistricting plans by simulation, relying on large numbers of plans for stability \cite{matt:etal:21} or sufficient diversity of nearly-independent plans \cite{kenn:etal:forth}.
+Often, we can would then call this an ensemble analysis, which is a flexible approach to sampling redistricting plans by simulation, relying on large numbers of plans for stability [@matt:hers:21].
 using this counterfactual, we thus create measures of harm, utility, and fairness which are based on a single plan compared to a set of other reasonable and feasible plans which incorporate the necessary legal constraints to make a plan viable.
@@
 Elections themselves are geographically determined.
-Recent research offers important resolution of the degree of partisan sorting in the US \cite{brow:enos:something}.
+Recent research offers important resolution of the degree of partisan sorting in the US [@enos:brown:21].
 By combining those understandings with the necessities of redistricting, we are able to expand the localized aspect of harm.
@@
 Using counterfactual plans to understand this space allows us to properly calibrate understandings of the impact of redistricting plans on individual understanding.
-Thus, we create measures which capture the possibility that no plans exist which favor one community \cite{duch:massachusetts}.
+Thus, we create measures which capture the possibility that no plans exist which favor one community [@duch:18].
 Using this class of metrics which describe the individual affect via representation allow us to better align legalistic interpretations of harm with political science understandings.
+
+TODO: Add findings!
 

### 2021-09-11 paper/paper.Rmd
remove question
sims + updates

--- 2021-09-11 paper/paper.Rmd
@@
 
-```{r nj-vars, fig.cap="Various measures for an ensemble of New Jersey districting plans.", fig.width=6.5, fig.height=6.5}
-#include_graphics(here("out/nj_vars.pdf"))
-```
+Figure \ref{nj-maps} shows the baseline Democratic vote share by precinct, along with two hypothetical gerrymanders of the state, which were generated using the short-burst optimization approach of @cannon2020.
 

### 2021-09-11 paper/paper.Rmd

--- 2021-09-11 paper/paper.Rmd
@@
 Partisan symmetry is a useful idea which greatly improves academic understanding of the redistricting process.
-Why is it hard?
+
 Understanding the redistricting process requires a feasible and reliable understanding of the counterfactual set of redistricting plans.
@@
 Yet, without a known target distribution, these heuristic approaches to simulation fall do not offer necessary guarantees for confidence in the simulated plans to represent the total space of possible plans [@fifield2020essential].
-Recent advances in Markov Chain methods as applied to redistricting allow us move beyond simulation towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @deford2019; @mccartan2020].
+Recent advances in Markov Chain methods as applied to redistricting allow us move beyond simulation towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @mccartan2020].
 
@@
 
-```{r nj-vars, fig.cap="Various measures for an ensemble of New Jersey districting plans.", fig.width=6.5, fig.height=6.5}
-#include_graphics(here("out/nj_vars.pdf"))
-```
+Figure \ref{fig:nj-maps} shows the baseline Democratic vote share by precinct, along with two hypothetical gerrymanders of the state, which were generated using the short-burst optimization approach of @cannon2020.
 

### 2021-09-11 paper/paper.Rmd
pairs

@@
 Figure \ref{fig:nj-maps} shows the baseline Democratic vote share by precinct, along with two hypothetical gerrymanders of the state, which were generated using the short-burst optimization approach of @cannon2020.
+The Democratic gerrymander has 11 Democratic seats and 1 Republican seat at baseline, while the Republican gerrymander has 6 Democratic and 6 Republican seats.
 
@@
 
+```{=tex}
 \clearpage
-
 \appendix
-
+\renewcommand\thefigure{\thesection.\arabic{figure}}
+\setcounter{figure}{0}
+```
 # Proofs of Propositions {#app:proofs}
@@
 Formulas here.
+
+# Relationship Between Districting Metrics in New Jersey {#app:nj}
+
+```{r nj-pairs, fig.cap="Scatterplots for all pairs of metrics.
+Color indicates the number of Democratic seats won (ranging from 7 to 10).
+Below the diagonal, numbers indicate the absolute value of the correlation between variables.", fig.width=7, fig.height=7}
+include_graphics(here("paper/figures/nj_pairs.pdf"))
+```

### 2021-09-11 R/02_run_toy.R

@@ -1,0 +1,1 @@
+lapply(sort(Sys.glob(here::here("R/**/*.R"))), source)

### 2021-09-11 R/02_sim_iowa.R

@@ -1,61 +1,0 @@
-iowa = read_rds(here("data/IA_cd_final_vtd_20.rds")) %>%
-    redist_map(pop_tol=0.001, ndists=4, adj=.$adj)
-
-N_sim = 5e3
-plans = redist_smc(iowa, N_sim, verbose=FALSE)
-#pl_opt_dem = redist_shortburst(iowa, scorer_group_pct(iowa, ndv, ndv+nrv, 2))
-opt_dem = c(4, 4, 2, 4, 4, 4, 2, 2, 2, 3, 2, 2, 2, 2, 4, 3, 2, 4, 2, 4, 2, 2, 3, 2, 1, 4, 4, 2, 3, 2, 2, 2, 2, 2, 2, 4, 2, 2, 1, 2, 2, 2, 4, 3, 2, 2, 4, 4, 3, 1, 4, 3, 3, 4, 2, 4, 3, 3, 4, 2, 4, 4, 4, 1, 4, 2, 4, 4, 4, 3, 4, 2, 4, 2, 4, 2, 1, 4, 1, 4, 2, 3, 4, 4, 1, 4, 4, 4, 4, 4, 4, 4, 4, 2, 2, 2, 4, 2, 2)
-#pl_opt_rep = redist_shortburst(iowa, scorer_group_pct(iowa, nrv, ndv+nrv, 4))
-opt_rep = c(4, 1, 3, 1, 4, 3, 2, 4, 2, 3, 2, 2, 4, 4, 4, 3, 2, 4, 3, 1, 2, 3, 3, 4, 4, 1, 1, 3, 2, 2, 3, 2, 3, 3, 2, 1, 4, 2, 4, 4, 4, 2, 4, 2, 3, 4, 4, 3, 3, 4, 1, 2, 3, 2, 2, 2, 3, 2, 1, 2, 1, 1, 1, 4, 1, 2, 4, 1, 1, 2, 2, 2, 1, 4, 4, 4, 1, 4, 2, 1, 4, 3, 4, 2, 4, 2, 1, 4, 1, 1, 1, 2, 1, 4, 2, 3, 4, 2, 4)
-
-plans = plans %>%
-    subset_sampled() %>%
-    add_reference(opt_dem, "dem_gerry") %>%
-    add_reference(opt_rep, "rep_gerry") %>%
-    mutate(dev = plan_parity(iowa),
-           comp = distr_compactness(iowa),
-           dem = group_frac(iowa, ndv, ndv + nrv),
[…]

### 2021-09-11 R/03_sim_nj.R

@@ -24,7 +24,7 @@ prop_seats = round(attr(nj, "ndists") * statewide)
 
 pl = calc_plans_stats(plans, nj, ndv, nrv)
 
-# Initial plots
+# Initial plots ----
 p1 = ggplot(nj, aes(fill=ndv/(ndv+nrv))) +
     geom_sf(size=0, color="#00000000") +
     scale_fill_party_c(limits=c(0.15, 0.85)) +
@@ -45,23 +45,14 @@ p = p1 + p2 + p3 + plot_layout(guides="collect")
 ggsave(here("paper/figures/nj_maps.pdf"), plot=p, width=8, height=4.5)
 
 
+# variables plot -----
+meas_labels = labels=c("Expected\nDem. seats", expression(U^G), expression(U^L),
+                       "F(q)", "H", "Efficiency\ngap", "Partisan\nbias", "Mean-median")
+select(pl$plan, n_dem, e_dem, u_glb, u_loc, f, h, egap, pbias, mean_med)
 
 
-pdf("out/nj_vars.pdf", 9, 9)
-expl_vars(pl, e_dem, u_glb, u_loc, f, f_glb, h, egap, pbias, mean_med, competitive)
-dev.off()
 
-pdf("out/nj_vars2.pdf", 9, 9)
-subset_sampled(pl$plan) %>%
-    select(n_dem, e_dem, u_glb, u_loc, f, h, egap, pbias, mean_med) %>%
-    expl_vars(labels=c("Expected\nDem. seats", expression(U^G), expression(U^L),
-                       "F(q)", "H", "Efficiency\ngap", "Partisan\nbias", "Mean-median"))
-dev.off()
-
-select(pl$plan, e_dem, u_glb, u_loc, f, egap, pbias, mean_med, competitive) %>%
-    #pairs(cex=0.2)
-    pairsD3::pairsD3(cex=1.5, opacity=0.5, group=pl$plan$n_dem)
-
+# terciles plot -----
 
 d_plot = pl$plan %>%
     mutate(`terc_Local Utility` = ntile(u_loc, 3),
@@ -86,67 +77,9 @@ ggplot(d_plot, aes(as.factor(district), dem, fill=as.factor(tercile))) +
 ggsave("paper/figures/nj_terciles.pdf", width=6.5, height=6)
 
 
-
-plot(nj, rowMeans(pl$mat > 0.5)) + scale_fill_party_c()
-
-
[…]

### 2021-09-11 paper/paper.Rmd
nj vars

@@
 
-```{r nj-maps, fig.cap="Baseline vote patterns, and two hypothetical gerrymanders.", fig.width=6.5, fig.height=3.5}
+```{r nj-maps, fig.cap="Baseline vote patterns, and two hypothetical gerrymanders. Census-designated places are shaded in gray.", fig.width=6.5, fig.height=3.5}
 include_graphics(here("paper/figures/nj_maps.pdf"))
@@
 The Democratic gerrymander has 11 Democratic seats and 1 Republican seat at baseline, while the Republican gerrymander has 6 Democratic and 6 Republican seats.
+
+We sample 1,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
+For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.
+The distribution of these measures is shown in Figure \ref{fig:nj-vars}.
+
+```{r nj-vars, fig.cap="Distributions of districting measures for simulated New Jersey distriting plans, the Republican gerrymander (red), and the Democratic gerrymander (blue).", fig.width=6.5, fig.height=3.25}
+include_graphics(here("paper/figures/nj_vars.pdf"))
+```
 

### 2021-09-11 R/03_sim_nj.R

@@ -46,9 +46,37 @@ ggsave(here("paper/figures/nj_maps.pdf"), plot=p, width=8, height=4.5)
 
 
 # variables plot -----
-meas_labels = labels=c("Expected\nDem. seats", expression(U^G), expression(U^L),
-                       "F(q)", "H", "Efficiency\ngap", "Partisan\nbias", "Mean-median")
-select(pl$plan, n_dem, e_dem, u_glb, u_loc, f, h, egap, pbias, mean_med)
+meas_labels = c(e_dem="Expected\nDem. seats", u_glb=expression(U^G), u_loc=expression(U^L),
+                f="F(q)", h="H", egap="Efficiency\ngap", pbias="Partisan\nbias", mean_med="Mean-median")
+plot_var = function(nm, i) { suppressMessages({
+    lab = meas_labels[nm]
+    if (str_detect(as.character(lab), "\n"))
+        lab = str_replace(lab, "\n", " ")
+    p = hist(pl$plan, !!ensym(nm), fill="#aaaaaa", bins=32) +
+        scale_color_party_d(guide="none") +
+        labs(x=NULL, title=lab) +
+        scale_y_continuous(name = if (i %% 4 == 1) "Fraction of plans" else NULL,
+                           labels=function(x) percent(x, 1),
+                           limits=c(0, 0.405), expand=expansion(mult=c(0, 0.05))) +
+        theme_repr() +
+        theme(plot.title=element_text(size=10),
+              panel.grid.minor=element_blank(),
+              panel.grid.major.x=element_blank())
+    if (min(pl$plan[[nm]]) < 0 & max(pl$plan[[nm]]) > 0)
+        p = p + geom_vline(xintercept=0, lty="dashed", size=0.25)
+    if (nm == "e_dem")
+        p = p + geom_vline(xintercept=statewide*12, lty="dashed", size=0.25) +
+            annotate("label", x=statewide*12, y=0.35, label="Proportional share",
+                     family="Times", size=2.5, hjust=-0.05, label.size=0,
+                     label.r=unit(0, "pt"), label.padding=unit(1, "pt"))
+    if (i %% 4 != 1)
+        p = p + theme(axis.text.y=element_blank(),
+                      axis.ticks.y=element_blank())
+    p
+})}
+p = imap(names(meas_labels), plot_var) %>%
+    wrap_plots(nrow=2, ncol=4)
+ggsave(here("paper/figures/nj_vars.pdf"), plot=p, width=8, height=4)
 
 
 

### 2021-09-12 paper/paper.Rmd
work on nj

@@
 To illustrate the connections in the preceding section and to demonstrate the value of the proposed measures in a real-world setting, we provide a brief analysis of congressional districting in the state of New Jersey.
-As discussed above, many gerrymandering metrics function well in highly competitive states, but may produce more counterintuitive results in other states.
+As discussed above, many gerrymandering measures function well in highly competitive states, but may produce more counterintuitive results in other states.
 New Jersey, with an average Democratic two-party vote share of 56.6% across our two baseline elections (2016 presidential and 2018 senatorial), will demonstrate that this is not a problem unique to lopsided states.
@@
 ```
+
+The efficiency gap and $F(q)$ both correctly classify the gerrymandered plans as such, and with the correct sign.
+Both gerrymanders also have a higher average harm than any of the simulated plans.
+In contrast, under the partisan bias measure the Republican gerrymander is completely unbiased.
+And according to the mean-median measure both gerrymanders strongly favor Democrats.
+For the utility measures, the Democratic plan is a low-utility outlier.
+But the Republican plan is near the median for both global and local utility.
+
+How to square this with the finding that the Republican plan is an unfair gerrymander which harms more voters than expected under a neutral process?
+As the distribution of expected Democratic seats shows, Democrats are structurally favored in New Jersey, able to win a larger fraction of seats than their statewide vote share under a neutral process.
+Plans which actively counteract this to try to draw proportional districts, like the Republican gerrymander, will necessarily harm Democratic voters versus the neutral counterfactual.
+But in doing so they will create a more proportional plan that will tend to pack voters of each party into more homogeneous districts, increasing both global and local utility.
 

### 2021-09-12 paper/paper.Rmd
edits

@@
 
-```{r nj-maps, fig.cap="Baseline vote patterns, and two hypothetical gerrymanders. Census-designated places are shaded in gray.", fig.width=6.5, fig.height=3.5}
+```{r nj-maps, fig.cap="Baseline vote patterns, and two hypothetical gerrymanders. Census-designated places are shaded in gray and county lines in white.", fig.width=6.5, fig.height=3.5}
 include_graphics(here("paper/figures/nj_maps.pdf"))
@@
 Both gerrymanders also have a higher average harm than any of the simulated plans.
-In contrast, under the partisan bias measure the Republican gerrymander is completely unbiased.
-And according to the mean-median measure both gerrymanders strongly favor Democrats.
+In contrast, under the partisan bias measure the Republican gerrymander is completely unbiased, and according to the mean-median measure, both gerrymanders strongly favor Democrats.
+
 For the utility measures, the Democratic plan is a low-utility outlier.
 But the Republican plan is near the median for both global and local utility.
-
 How to square this with the finding that the Republican plan is an unfair gerrymander which harms more voters than expected under a neutral process?

### 2021-09-12 paper/paper.Rmd
NJ

@@
 
+```{r nj-district-shares, fig.cap="Distribution of district vote shares, with districts ordrered by Democratic two-part vote share, and divided by quantile for each of three districting measures.", fig.width=6.5, fig.height=4.875}
+include_graphics(here("paper/figures/nj_district_shares.pdf"))
+```
+
+This is manifested in the district-level vote shares shown in Figure \ref{fig:nj-district-shares}.
+Within each districting plan, we number districts in order of their Democratic vote share, with district 1 the least Democratic.
+The figure shows the distribution of these vote shares for each ordered district, but broken out according to each plan's values of $F(q)$, $U^{(G)}$, and $U^{(L)}$.
+
 # Discussion {#sec:discussion}

### 2021-09-12 R/03_sim_nj.R

@@ -2,19 +2,19 @@ nj = read_rds(here("data/NJ_cd_final_vtd_20.rds")) %>%
     redist_map(pop_tol=0.01, ndists=12, adj=.$adj)
 
 if (!file.exists(sim_path <- here("data/nj_sims.rds"))) {
-    N_sim = 1000
-    plans = redist_smc(nj, N_sim, counties=county, pop_temper=0.01, verbose=TRUE)#FALSE)
+    N_sim = 10000
+    plans = redist_smc(nj, N_sim, counties=county, pop_temper=0.005, verbose=TRUE)#FALSE)
 
     # gerrymanders
     set.seed(5118)
-    opt_dem = redist_shortburst(nj, scorer_group_pct(nj, ndv, ndv+nrv, 11))
+    opt_dem = redist_shortburst(nj, scorer_group_pct(nj, ndv, ndv+nrv, 11), max_bursts=800)
     set.seed(5118)
-    opt_rep = redist_shortburst(nj, scorer_group_pct(nj, nrv, ndv+nrv, 6))
+    opt_rep = redist_shortburst(nj, scorer_group_pct(nj, nrv, ndv+nrv, 6), max_bursts=800)
 
-    plans %>%
+    plans = plans %>%
         add_reference(last_plan(opt_dem), "dem_gerry") %>%
-        add_reference(last_plan(opt_rep), "rep_gerry") %>%
-        write_rds(sim_path, compress="xz")
+        add_reference(last_plan(opt_rep), "rep_gerry")
+    write_rds(plans, sim_path, compress="xz")
 } else {
     plans = read_rds(sim_path)
 }
@@ -42,7 +42,8 @@ p3 = plot_cds(nj, as.matrix(plans)[,"rep_gerry"], county, "NJ") +
     theme_repr_map() +
     theme(plot.title = element_text(hjust = 0.5))
 p = p1 + p2 + p3 + plot_layout(guides="collect")
-ggsave(here("paper/figures/nj_maps.pdf"), plot=p, width=8, height=4.5)
+if (!file.exists(path <- here("paper/figures/nj_maps.pdf")))
+    ggsave(path, plot=p, width=8, height=4.5)
 
 
 # variables plot -----
@@ -52,7 +53,7 @@ plot_var = function(nm, i) { suppressMessages({
     lab = meas_labels[nm]
     if (str_detect(as.character(lab), "\n"))
[…]

### 2021-09-12 paper/paper.Rmd
nj example plots

--- 2021-09-12 paper/paper.Rmd
@@
 
+```{r nj-district-shares, fig.cap="Distribution of district vote shares, with districts ordrered by baseline Democratic two-part vote share.
+Displayed on top are the district vote shares for six example plans, each extremal for some combination of districting measures.", fig.width=6.5, fig.height=3.9}
+include_graphics(here("paper/figures/nj_district_shares.pdf"))
+```
+
+This is manifested in the district-level vote shares shown in Figure \ref{fig:nj-district-shares}.
+Within each districting plan, we number districts in order of their Democratic vote share, with district 1 the least Democratic.
+The figure shows these vote shares for six selected districting plans, and compares them to the distribution of shares for the entire sample.
+The selected plans themselves are shown in Appendix \ref{app:nj}.
+
 # Discussion {#sec:discussion}
@@
 
-# Relationship Between Districting Metrics in New Jersey {#app:nj}
+# Districting Measures in New Jersey {#app:nj}
 
-```{r nj-pairs, fig.cap="Scatterplots for all pairs of metrics.
+```{r nj-pairs, fig.cap="Scatterplots for all pairs of measures.
 Color indicates the number of Democratic seats won (ranging from 7 to 10).
@@
 ```
+
+```{r nj-ex-maps, fig.cap="Example districting plans selected according to their scores on $F(q)$, $U^{(G)}$, and $U^{(L)}$.
+District shares for these example plans are shown in Figure \\ref{fig:nj-district-shares}.", fig.width=7, fig.height=7}
+include_graphics(here("paper/figures/nj_ex_maps.pdf"))
+```

### 2021-09-12 R/03_sim_nj.R

@@ -82,42 +82,6 @@ if (!file.exists(path <- here("paper/figures/nj_vars.pdf")))
 
 
 
-# terciles plot -----
-
-d_plot = pl$plan %>%
-    subset_sampled() %>%
-    mutate(terc_u_loc = ntile(u_loc, 5),
-           terc_u_glb = ntile(u_glb, 5),
-           terc_f = ntile(-abs(f), 5)) %>%
-    mutate(across(starts_with("terc_"), ~ if_else(. %in% 2:4, 3L, .))) %>%
-    select(draw, n_dem, e_dem, u_loc, u_glb, f, starts_with("terc_")) %>%
-    as_tibble()
-d_plot = number_by(pl$distr, dem) %>%
-    inner_join(d_plot, by="draw") %>%
-    pivot_longer(starts_with("terc_"), names_to="qty",
-                 values_to="tercile", names_prefix="terc_")
-
-terc_labs = c("Bottom 20%", "", "Middle 60%", "", "Top 20%")
-meas_labels = c(u_glb="U^(G)", u_loc="U^(L)", f="-abs(F(q))")
-p = d_plot %>%
-    mutate(qty = meas_labels[qty],
-           tercile = ordered(terc_labs[tercile], levels=terc_labs[c(1, 3, 5)])) %>%
-ggplot(aes(as.factor(district), dem, fill=tercile)) +
-    facet_grid(. ~ qty, labeller=label_parsed) +
-    geom_hline(yintercept=0.5, lty="dashed") +
-    geom_boxplot(size=0.2, outlier.size=0.05, width=0.8) +
-    #geom_violin(bw=0.001, size=0.2) +
-    scale_y_continuous("Democratic two-party share", labels=\(x) percent(x, 1)) +
-    labs(x="Districts, ordered by Democratic share",
-         fill="Quantile\nof measure") +
-    scale_fill_wa_d("sea_star", which=c(3, 9, 15)) +
-    theme_repr() +
-    theme(legend.position=c(0.9, 0.15),
-          legend.background=element_blank())
-if (!file.exists(path <- here("paper/figures/nj_district_shares.pdf")))
-    ggsave(path, plot=p, width=8, height=4.5)
-
-
 # best / worst
 find_best = function(qty1, qty2=NULL) {
     p = subset_sampled(pl$plan)
@@ -131,24 +95,21 @@ find_best = function(qty1, qty2=NULL) {
     }
 }
 
-comp_plans = list(
[…]

### 2021-09-13 paper/paper.Rmd
long awaited section 5

@@
 ## Outline of the Paper
+
+<!--  Section 1 here should really be what we're interested in I think -->
 
@@
 
-Over the past decade [**CM: correct?!**], researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
+In the years since Vieth [@2004vieth] researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
 Specifically, these measures aim to quantify deviations from partisan symmetry, fairness, or competitiveness.
@@
 All of these measures may be estimated using district vote shares.
+We now discuss the relationship between our measure and these three measures.
+<!-- chris thinks we shouldn't cite wang here because he copied mcdonald -->
 
@@
 
-1.  has been criticized [@tamcho2017egap; @bernstein2017egap; @deford2020symmetry]
+The efficiency gap interrogates how a party's votes were wasted in an election under a districting system.
+By comparing the wasted votes for Democrats to the wasted votes for Republicans, rescaled by the total votes cast, it offers a signed percent.
+Interpretation of the efficiency gap is intended to be straightforward and usable by courts.
+The sign indicates that a plan favors one party over another, while the magnitude indicates the degree of preference.
+The general recommendation for use in court is that a bias amounting to the size of two Congressional seats or 8% for state legislative plans would be deemed unacceptable.
+When we unpack the idea behind the efficiency gap, we see that it shares significant intuition with our measure $F(q)$.
+Both measures try to understand how individuals are rendered unable to effectively participate in democratic processes.
 
+Various criticisms of the efficiency gap have been made due to its potential for volatility in noncompetitive elections [@tamcho2017egap; @bernstein2017egap; @deford2020symmetry].
+By directly including simulated plans and electoral data in the formation of $F(q)$, despite shared intuition with the efficiency gap, our measure is not prey to this susceptibility.
+
+```{=html}
+<!--
+Further literature indicates that while the efficiency gap was originally introduced as a measure of partisan symmetry, "it does not imply a coherent seats--votes curve" and thus cannot be a measure of partisan symmetry [@katz2020].
+-->
+```
 ## Partisan Bias
 
+Partisan bias can be considered the classic measure of partisan symmetry.
+The bias is the number of seats that would need to be reallocated to make the system symmetric at a chosen statewide proportion of votes, often 0.50.
+It should not be surprising then that these measures differ often and greatly.
+
+Further, by relying on alternative districting plans, we do not necessitate the use of partisan swings.
+This is of great benefit in scenarios where swings must be so significant that they stray far from the data.
+Instead, the measures of harm and utility rely on $votes^{p}$ which can be model-free estimates of electoral performance of a party (namely historical election returns) or can be some sort of model, such as precinct-level election forecasts.
+Of course, the votes could even be generated by notions of partisan swings, but as a choice rather than a requirement.
+
+Notably, as the quantity of interest is not partisan symmetry, our measures do not require the existence of a seats-votes curve.
+This is consequential for the study of congressional redistricting.
+As a seats-votes curve is defined for a legislature, rather than a delegation, to understand the impact of single state's congressional map, you would be forced to consider how every other state drew their maps [@katz:king:rose:21a].
+Thus, if this taken as the definition of a seats-votes curve, then measures of partisan symmetry lack diagnostic power to identify gerrymanders for the delegations that academics, legal scholars, and the general public are often focused on.
+The individualized focus of our measures are, however, portable to congressional, state legislative, or any other form of districting and elections.
+
+```{=html}
+<!--
 1.  Integrating yields a related measure of @nagle2015measures, which has also been criticized by @deford2020symmetry.
 2.  Focuses on only one part of S-V curve, which may not be relevant
+-->
+```
+## Mean-Median Difference
 
-## Mean-Median
+Our measures greatly differ from the Mean-Median Difference.
+With the focus on individuals, the considerations of the mean district are only related to individuals in-so-far as they describe the general propensity to be harmed or benefited by a particular map.
+The consideration of a median district is similarly distanced.
+As with partisan bias, as our measures do not attempt to measure partisan symmetry, our measures can be applied to congressional delegations.
 
+```{=html}
+<!--
 1.  Like PB, focuses on only one part of S-V curve, which may not be relevant
-
+-->
+```
 ## Districting Plan Competitiveness {#sec:competitive}
@@
 
+We should say something about the importance in developing further measures which utilize simulation approaches
+
 # References {.unnumbered}

### 2021-09-13 paper/paper.Rmd
equity ideas

@@
 
-<!--  Section 1 here should really be what we're interested in I think -->
-
 Section \ref{sec:voter-util} provides the necessary notation and introduces our measures of voter utility.
@@
 
+```{=html}
+<!--  
+With Harm defined, we can thus define Partisan Equity as: $E[H|V(I) = DEM] = E[H|V(I) = REP]$
+
+Thus Fairness can be described as the signed movement away from equity.
+This naming works because it's condition on the votes and geographic distribution that you have, rather than some neutral midpoint that would be hard to justify.
+
+This clarifies our QOI and by later focusing on fairness, that gives us the analogous connection between bias (symmetry) and fairness (equity).
+
+Later, that also gives room in the discussion of the EGAP to talk about why you could describe a utility function based on $Pr(Vote not wasted)$, which 
+-->
+```
 As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.

### 2021-09-13 paper/paper.Rmd
finish drafting NJ

@@
   - name: Cory McCartan
-    thanks: Footnote stuff
+    thanks: |
+      To whom correspondence should be addressed. 
+      The authors thank ____ for helpful comments and discussion.
     department: Department of Statistics
@@
 
-```{=html}
-<!--
-1.  Like PB, focuses on only one part of S-V curve, which may not be relevant
--->
-```
 ## Districting Plan Competitiveness {#sec:competitive}
@@
 
+The fairest, best overall utility, and high global, low local utility plans have 8 Democratic and 4 Republican seats at baseline, but the best-utility plan has more extreme vote shares, i.e., less competitive districts, while the high global, low local utility plan has vote shares closer to 50%, with the fairest plan falling somewhere in the middle.
+
+In contrast, the low global, high local utility plan has 9 Democratic seats, 1 Republican seat, and 2 highly competitive seats at baseline.
+While generally local utility encourages proportionality and noncompetitive districts, conditional on low global utility (poor proportionality) the local utility measure encourages a Democratic gerrymander which satisfies the state's majority of Democratic voters, at the expense of the minority of Republican voters.
+
+The fair but low local utility plan has a similar vote share pattern to the fair plan, but overall with more competitive districts, which are generally penalized by the local utility measure.
+Conversely, the unfair but high local utility plan has 9 Democratic seats, 2 Republican seats, and 1 highly competitive seat at baseline---an unfair Democratic gerrymander---while keeping district vote shares relatively noncompetitive.
+
 # Discussion {#sec:discussion}

### 2021-09-22 paper/paper.Rmd
edit section 3-5.1

@@
 To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
+These counterfactuals could arise from a posited probability distribution capturing the factors relevant to a non-partisan redistricting process, or they could be a set of alternative redistricting plans that were considered as part of the legislative process leading to a final enacted plan.
 Then we may define the harm to a voter as $h:\Voters\times\Elecs\times\Plans^2\to\R$, \begin{align*}
@@
 
-## Fairness
+## Differential Harm and Fairness
 
-From here onward we will restrict our attention to the case of two parties---concretely, $\Parties=\{\dem,\rep\}$---since that will simplify the notation and exposition.
+Intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
+This is easily expressed using our notation above; $Q$ is fair in comparison to counterfactual $\tilde Q$ if $\E[H\mid V(I) = p]$ is constant across parties $p\in\Parties$.
+
+To quantify deviations from fairness, we introduce of a measure of \textit{differential harm}, defined as $$
+  \Delta^{p_1}_{p_2}H(q) \dfeq = \E[H\mid V(I)=p_1, Q=q] - \E[H\mid V(I)=p_2, Q=q],
+$$ for two parties $p_1$ and $p_2$.
+So $\Delta^{p_1}_{p_2}H(q)=0$ for all $p_1,p_2\in\Parties$ if and only if $q$ is fair.
+From here onward we will restrict our attention to the case of two parties---concretely, $\Parties=\{\dem,\rep\}$---and simply write $\Delta H(q)=\Delta H^\dem_\rep(q)$.
 But the same concepts and definitions naturally carry over to the multiparty setting.
 
-Intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
-This is easily expressed using our notation above; the average fairness of $Q$ versus counterfactual $\tilde Q$ may be defined as $$
-    F = \E[H\mid V(I) = \dem] - \E[H\mid V(I) = \rep];
-$$obviously the ordering of $\dem$ and $\rep$ is immaterial.
-Informally, this quantity is the difference in the probability that a voter of each party is represented by the other party but counterfactually would have been represented by their own party.
-
-Generally, we will want to evaluate the fairness of a specific districting plan $q$ versus a distribution of counterfactuals.
-These counterfactuals could arise from a posited probability distribution capturing the factors relevant to a non-partisan redistricting process, or they could be a set of alternative redistricting plans that were considered as part of the legislative process leading to a final enacted plan.
-We adopt the notation $F(q)$ for this scenario: $$
-    F(q) \dfeq \E[H\mid V(I) = \dem, Q=q] - \E[H\mid V(I) = \rep, Q=q],
-$$ so that we recover $F=\E[F(Q)]$.
-
-[**CM: better to define just** $F(Q)$**, since we will rarely compare two sets of plans?**]
-
-In the districting plan examples of Figure \ref{fig:mini-harm}, the fairness of the Democratic gerrymander of Mininois is $-0.519$, the fairness of the Democratic gerrymander of Minissouri is $-0.415$, and the fairness of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
+In the districting plan examples of Figure \ref{fig:mini-harm}, the differential harm of the Democratic gerrymander of Mininois is $-0.519$, the differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
 In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).
 
-Despite being derived from a local harm measure, $F(Q)$ is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
-Yet unlike a simple measure of seats, $F(Q)$ is shifted to an interpretable scale, where a value of zero corresponds to partisan fairness.
-In other words, $F(Q)$ provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
+Despite being derived from a local harm measure, $\Delta H(Q)$ is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
+Yet unlike a simple measure of seats, $\Delta H(Q)$ is shifted to an interpretable scale, where a value of zero corresponds to partisan fairness.
+In other words, $\Delta H(Q)$ provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
 This connection further supports the use of total seats as a summary statistic in redistricting simulation methods, as advocated by @deford2020symmetry.
@@
 \begin{restatable}{prop}{propfairness} \label{prop:fairness}
-Basically, if turnout is equal across districts, for a districting plan $q\in\Plans$ and political party $p\in\Parties$, we have
+If turnout is equal across districts, i.e., if $\sum_{p\in\Parties} \votes^{(p)}(d,v,q)$ is constant across $d$, $v$, and $q$, then for a specific districting plan $q\in\Plans$ and political party $p\in\Parties$, we have
 \begin{align*}
-    F(q) = \E[\seats(p, V, Q)\mid Q=q] - \E[\seats(p, V, \tilde Q)].
+    \Delta H(q) = \E[\seats(p, V, Q)\mid Q=q] - \E[\seats(p, V, \tilde Q)].
 \end{align*}
@@
 ```
-# Calculating the Harm and Fairness Measures {#sec:estimating}
+# Estimating the Harm Measures {#sec:estimating}
 
-The measures $U^{(G)}, U^{(L)}, H$ and $F$ defined above require specification of a probability measure on elections and districting plans, $\pi$.
+The measures $U^{(G)}, U^{(L)}, H$ and $\Delta H^(q)$ defined above require specification of a probability distribution on elections and districting plans, $\pi$.
 Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
-Either way, a collection of past election data and an enacted districting plan alone are not enough to calculate these measures.
+Either way, past election data and an enacted districting plan alone are not enough to calculate these measures.
 
@@
 For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans which are geographically contiguous, have district populations within a deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
-These maximum-entropy distributions put as few restraints as possible on the districting sampling process, subject to a set of constraints imposed by the researcher.
-If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by partisan considerations, to use in estimating $H$ and $F$.
+These maximum-entropy distributions put as few conditions as possible on the districting sampling process, as long as a set of constraints imposed by the researcher are met.
+If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by partisan considerations, to use in estimating $H$ and $\Delta H(q)$.
 
@@
 Election results present a different set of modeling challenges, given the large number of factors that influence the outcome, and the general unpredictability of future events.
-As far as estimating measures like $U^{(G)}, U^{(L)}, H$, and $F$, we would ideally use a relatively simple model that captures the primary drivers of election outcomes.
-In the interest of fairness, the model should not err on the side of simplicity and symmetry, and avoid making any biasing assumptions about the future (e.g., that demographic shifts will favor one party, or that past trends in election returns will continue in the same direction).
+As far as estimating measures like $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$, we would ideally use a relatively simple model that captures the primary drivers of election outcomes.
+In the interest of fairness, the model should not make any biasing assumptions about the future (e.g., that demographic shifts will favor one party, or that past trends in election returns will continue in the same direction).
 
@@
 
-The proportionate shift, where we assume that a constant fraction of *voters of one party* in each precinct switch to the other party, has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
+Proportionate shift, where we assume that a constant fraction of *voters of one party* in each precinct switch to the other party, has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
 First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
@@
 
-The proportionate shift can be mathematically framed as linear shift on a transformed scale, analogous to the generalized linear model setting.
+The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlikea generalized linear model.
 Specifically, we define $\gamma:[0,1]\to\R$, $$
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace with rate 2.
+$$ which is the inverse CDF of a Laplace with rate 2.[^3]
 Then a party's vote share in precinct $k$, notated here as $s_k$, is modeled as $$
@@
 
+[^3]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+
 More complicated election outcome models are certainly possible and straightforward to use with the measures proposed here.
@@
 For ease and efficiency of computation, we stick to the simple proportionate uniform shift model here.
-Appendix \ref{app:formulas} contains the specific formulas used to compute $U^{(G)}, U^{(L)}, H$, and $F$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
+Appendix \ref{app:formulas} contains the specific formulas used to compute $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
 
@@
 
-In the years since Vieth [@2004vieth] researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
+In the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
 Specifically, these measures aim to quantify deviations from partisan symmetry, fairness, or competitiveness.
-Some of the most commonly used are perhaps the *efficiency gap* [@stephanopoulos2015], *partisan bias* [@tufte1973pb; @king1987democratic], and the *mean-median* measure [@mcdonald2015unfair; @wang2016three].
-All of these measures may be estimated using district vote shares.
-We now discuss the relationship between our measure and these three measures.
-<!-- chris thinks we shouldn't cite wang here because he copied mcdonald -->
+Some of the most commonly used are perhaps the *efficiency gap* [@stephanopoulos2015], *partisan bias* [@tufte1973pb; @king1987democratic], and the *mean-median* measure [@mcdonald2015unfair].
+All of these measures may be estimated using district vote shares alone, and as we discuss next, are to varying degrees related to our proposed measures.
 
@@
 
-The efficiency gap interrogates how a party's votes were wasted in an election under a districting system.
-By comparing the wasted votes for Democrats to the wasted votes for Republicans, rescaled by the total votes cast, it offers a signed percent.
-Interpretation of the efficiency gap is intended to be straightforward and usable by courts.
-The sign indicates that a plan favors one party over another, while the magnitude indicates the degree of preference.
-The general recommendation for use in court is that a bias amounting to the size of two Congressional seats or 8% for state legislative plans would be deemed unacceptable.
-When we unpack the idea behind the efficiency gap, we see that it shares significant intuition with our measure $F(q)$.
-Both measures try to understand how individuals are rendered unable to effectively participate in democratic processes.
+The efficiency gap ($EG$) interrogates how a party's votes are wasted under a districting plan.
+By subtracting the wasted votes for Democrats from those for Republicans, and normalizing by the total votes cast, it offers a signed measure that is simple to interpret: $EG=0.05$, for example, indicates that 5% more Democratic votes are wasted than Republican ones.
+The general recommendation is that an efficiency gap larger than two Congressional seats (i.e., 2 divided by the number of districts), or 0.08 for state legislative plans, indicates a partisan gerrymander.
 
-Various criticisms of the efficiency gap have been made due to its potential for volatility in noncompetitive elections [@tamcho2017egap; @bernstein2017egap; @deford2020symmetry].
-By directly including simulated plans and electoral data in the formation of $F(q)$, despite shared intuition with the efficiency gap, our measure is not prey to this susceptibility.
+A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
 
-```{=html}
-<!--
-Further literature indicates that while the efficiency gap was originally introduced as a measure of partisan symmetry, "it does not imply a coherent seats--votes curve" and thus cannot be a measure of partisan symmetry [@katz2020].
--->
-```
+First, $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$ are all averaged over possible election outcomes, and so do not suffer from discreteness.
+And as discussed above, $U^{(G)}$ and $U^{(L)}$ both reward proportionality.
+Of course, proportionality is not always possible, given a state's political geography.
+The harm measures $H$ and $\Delta H(q)$ both reflect this by comparing to a set of counterfactual districting plans.
+This addresses one inherent shortcoming of the efficiency gap, illustrated best by the state of Massachusetts.
+According to the efficiency gap, Massachussetts' congressional districting plan is a Democratic gerrymander ($EG=-0.14$).
+But it is impossible to draw a single Republican congressional district in the state.
+Consequently $H$ and $\Delta H(q)$ are everywhere 0.
+
+Finally, the fact that the efficiency gap is minimized for districts with 75%--25% vote shares is a direct consequence of comparing the raw number of wasted votes, rather than the share within each party.
+In contrast, $\Delta H(q)$ is calculated by comparing harm averaged first within each party, avoiding this deficiency.
+
+As we will see in the application below, often there is substantial correlation between the efficiency gap and $\Delta H(q)$.
+Yet in more unusual or tricky cases, the efficiency gap may point to a partisan gerrymander while $\Delta H(q)$ indicates a fair outcome or an unfair advantage for the opposite party.
+
 ## Partisan Bias
@@
 
-The efficiency gap and $F(q)$ both correctly classify the gerrymandered plans as such, and with the correct sign.
+The efficiency gap and $\Delta H(q)$ both correctly classify the gerrymandered plans as such, and with the correct sign.
 Both gerrymanders also have a higher average harm than any of the simulated plans.
@@
 
-```{r nj-ex-maps, fig.cap="Example districting plans selected according to their scores on $F(q)$, $U^{(G)}$, and $U^{(L)}$.
+```{r nj-ex-maps, fig.cap="Example districting plans selected according to their scores on $\\Delta H(q)$, $U^{(G)}$, and $U^{(L)}$.
 District shares for these example plans are shown in Figure \\ref{fig:nj-district-shares}.", fig.width=7, fig.height=7}

### 2021-10-12 paper/paper.Rmd
small updates

@@
 This literature defers to symmetric outcomes for parties, commonly referred to as partisan symmetry [@katz2020].
-Partisan symmetry is a useful idea which greatly improves academic understanding of the redistricting process.
+Partisan symmetry offers the idea that the number of seats awarded to a party by a set of votes would be the same if the labels on the parties were swapped.
+Partisan symmetry is a useful quantity of interest which has been widely used to improve academic and legal understanding of the redistricting process.
+We do not challenge the value of partisan symmetry, but instead suggest that a different question should be asked to describe the fairness of district based democracies.
 
-Understanding the redistricting process requires a feasible and reliable understanding of the counterfactual set of redistricting plans.
+We argue that understanding the redistricting process requires a feasible and reliable understanding of the counterfactual set of redistricting plans.
+Our metrics offer a framework for understanding approximately individual harms, benefits, and net fairness.
+Using counterfactual plans to understand this space allows us to properly calibrate understandings of the impact of redistricting plans on individual understanding.
+Thus, we create measures which capture the possibility that no plans exist which favor one community [@duch:18].
+Using this class of metrics which describe the individual affect via representation consequently allow us to better align legalistic interpretations of harm for standing with political science understandings.
+
+To describe a reasonable counterfactual to measure our quantity of interest, we rely on advances in redistricting simulation methodologies.
 Early methodologies for simulating plans relied on randomness to generate many different plans and enhanced understanding of redistricting greatly [@ciri:darl:orou:00; @chen:rodd:13; @deford2019].
@@
 Yet, without a known target distribution, these heuristic approaches to simulation fall do not offer necessary guarantees for confidence in the simulated plans to represent the total space of possible plans [@fifield2020essential].
-Recent advances in Markov Chain methods as applied to redistricting allow us move beyond simulation towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @mccartan2020].
+Recent advances in Monte Carlo methods allow us move beyond simulation towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @mccartan2020].
 
-Using sampling methods, we can then develop a reliable counterfactual that can be tuned to consider the sampling constraints imposed by the laws and practices of a state.
+Using these sampling methods, we can then develop a reliable counterfactual that can be tuned to consider the sampling constraints imposed by the laws and practices of a state.
 This counterfactual allows us to characterize the likely plans from a state.
@@
 
-We offer a reminder that geographic context is key to understanding common problems in political science..
+We offer a reminder that geographic context is key to understanding common problems in political science.
 Elections themselves are geographically determined.
@@
 We further clarify that common existing measures of partisan fairness -- while they may capture the difficulties offered to a party -- do not necessarily capture the representational harms incurred by individuals who live in districts which do not at all represent their preferences.
-
-Our metrics offer a framework for understanding approximately individual harms, benefits, and net fairness.
-Using counterfactual plans to understand this space allows us to properly calibrate understandings of the impact of redistricting plans on individual understanding.
-Thus, we create measures which capture the possibility that no plans exist which favor one community [@duch:18].
-Using this class of metrics which describe the individual affect via representation allow us to better align legalistic interpretations of harm with political science understandings.
 
@@
 
-```{r mini-schematic, fig.cap="Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts. The total voting population for each example is 180."}
+```{r mini-schematic, fig.cap="Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts.
+The total voting population for each example is 180.
+White lines delineate precincts."}
 include_graphics(here("paper/figures/minis_schematic.pdf"))
@@
 Attempting to maximize the average local utility across the electorate would lead to a districting plan where every voter is assigned to a district with other voters of the same party preference; in this way, every voter's utility would be $1$.
-Whether such an districting plan is possible depends on the geographical layout of voters and the legal constraints on the districting process---must districts be contiguous, balanced in population, constructed from whole precincts, geographically compact, and so on.
+Whether such an districting plan is possible depends on the geographical layout of voters and the legal constraints on the districting process---districts must be contiguous, balanced in population, constructed from whole precincts, geographically compact, and so on.
 But the general tendency of this utility measure (in the context of a single election, at least), like the global utility measure, will be to push plans towards proportionality.
@@
 
-```{r mini-util-opt, fig.cap="Utility-maximizing redistricting plans for the three running examples, with each district shaded by its partisan lean."}
+```{r mini-util-opt, fig.cap="Utility-maximizing redistricting plans for the three running examples, with each district shaded by its partisan lean.
+Black lines indicate districts.
+Red and blue dots indicate Republicans and Democrats, respectively."}
 include_graphics(here("paper/figures/minis_util_opt.pdf"))
@@
 In fact, a harm measure defined analogously using global utility would essentially be a difference in seats won by a voter's party between plans $q$ and $\tilde q$ (the exact form would be affected by the function $g$).
-This measure is not particularly interesting, and nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation [@2018league; @2019rucho; @2019lewis].
-**[CM: revisit ¶ based on NJ findings]****.**
+A global harm measure is not particularly interesting, nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation [@2018league; @2019rucho; @2019lewis].
+**[CM: revisit ¶ based on NJ findings].**
 
@@
 
-```{r mini-harm, fig.cap="Three potential districting plans for Mininois and Minissouris, showing district partisan lean (top row) and total voter harm versus the counterfactuals in Figure 2(a)-(b) (bottom row; darker shading indicates more harm)."}
+```{r mini-harm, fig.cap="Three potential districting plans for Mininois and Minissouris, showing district partisan lean (top row) and total voter harm versus the counterfactuals in Figure 2(a)-(b) (bottom row; darker shading indicates more harm).**CK: Should there be a second scale here?**"}
 include_graphics(here("paper/figures/mini_harm.pdf"))
@@
 These maximum-entropy distributions put as few conditions as possible on the districting sampling process, as long as a set of constraints imposed by the researcher are met.
-If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by partisan considerations, to use in estimating $H$ and $\Delta H(q)$.
+If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan considerations, to use in estimating $H$ and $\Delta H(q)$.
 
@@
 Further, by relying on alternative districting plans, we do not necessitate the use of partisan swings.
-This is of great benefit in scenarios where swings must be so significant that they stray far from the data.
+This is of great benefit in scenarios where swings must be so significant in magnitude that they stray far from the data.
 Instead, the measures of harm and utility rely on $votes^{p}$ which can be model-free estimates of electoral performance of a party (namely historical election returns) or can be some sort of model, such as precinct-level election forecasts.
@@
 We sample 1,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
-For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.
+For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^4]
 The distribution of these measures is shown in Figure \ref{fig:nj-vars}.
+
+[^4]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature[@katz21].
 

### 2021-10-12 paper/paper.Rmd
round 1 edits

--- 2021-09-22 paper/paper.Rmd
@@
 
-We posit that a utility of the global election outcome to a voter is an increasing function of the share of total seats that the voter's preferred party wins.
+We posit that the utility of the global election outcome to a voter is a strictly increasing function of the share of total seats that the voter's preferred party wins.
 Their global outcome utility function for an election outcome $e\in\Elecs$ and districting plan $q\in\Plans$ can thus be written$$
@@
     \qquad u^{(G)}(i,v,q) = g\qty(\frac{\seats(v(i), v, q)}{|\Distrs|}),
-$$for $g:[0,1]\bijto\R$ an increasing function.
+$$for $g:[0,1]\bijto\R$ a strictly increasing function.
 
@@
     U^{(L)} \dfeq u^{(L)}(I, V, Q).
-$$Thus, the utility for a single voter $v$ becomes $U^{(L)}\mid I=i$; the electorate's average utility for election outcome $v$ becomes $\E[U^{(L)}\mid V=v]$, and so on.
+$$Thus, the local utility for a single voter $i$ becomes $U^{(L)}\mid I=i$; the electorate's average utility for election outcome $v$ becomes $\E[U^{(L)}\mid V=v]$, and so on.
 
@@
 Harm is defined using the local, not global, utility measure, because inherently, global utility (and thus harm) will be constant for all voters of a particular party.
-In fact, a harm measure defined analogously using global utility would essentially be a difference in seats won by a voter's party between plans $q$ and $\tilde q$ (the exact form would be affected by the function $g$).
-This measure is not particularly interesting, and nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation [@2018league; @2019rucho; @2019lewis].
-**[CM: revisit ¶ based on NJ findings]****.**
+As we discuss below in Section \ref{subsec:diff-harm}, locally-based harm is connected to global electoral outcomes, anyway.
 
-```{=html}
-<!--  
-With Harm defined, we can thus define Partisan Equity as: $E[H|V(I) = DEM] = E[H|V(I) = REP]$
-
-Thus Fairness can be described as the signed movement away from equity.
-This naming works because it's condition on the votes and geographic distribution that you have, rather than some neutral midpoint that would be hard to justify.
-
-This clarifies our QOI and by later focusing on fairness, that gives us the analogous connection between bias (symmetry) and fairness (equity).
-
-Later, that also gives room in the discussion of the EGAP to talk about why you could describe a utility function based on $Pr(Vote not wasted)$, which 
--->
-```
 As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.
@@
 
-## Differential Harm and Fairness
+## Differential Harm and Fairness {#subsec:diff-harm}
 
 Intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
-This is easily expressed using our notation above; $Q$ is fair in comparison to counterfactual $\tilde Q$ if $\E[H\mid V(I) = p]$ is constant across parties $p\in\Parties$.
+This is easily expressed using our notation above; $Q$ is *fair* in comparison to counterfactual $\tilde Q$ if $\E[H\mid V(I) = p]$ is constant across parties $p\in\Parties$.
 
 To quantify deviations from fairness, we introduce of a measure of \textit{differential harm}, defined as $$
-  \Delta^{p_1}_{p_2}H(q) \dfeq = \E[H\mid V(I)=p_1, Q=q] - \E[H\mid V(I)=p_2, Q=q],
+  \Delta^{p_1}_{p_2}H(q) \dfeq \E[H\mid V(I)=p_1, Q=q] - \E[H\mid V(I)=p_2, Q=q],
 $$ for two parties $p_1$ and $p_2$.
 So $\Delta^{p_1}_{p_2}H(q)=0$ for all $p_1,p_2\in\Parties$ if and only if $q$ is fair.
-From here onward we will restrict our attention to the case of two parties---concretely, $\Parties=\{\dem,\rep\}$---and simply write $\Delta H(q)=\Delta H^\dem_\rep(q)$.
+From here onward we will restrict our attention to the case of two parties---concretely, $\Parties=\{\dem,\rep\}$---and simply write $\Delta H(q)=\Delta^\dem_\rep H(q)$.
 But the same concepts and definitions naturally carry over to the multiparty setting.
@@
 
-The measures $U^{(G)}, U^{(L)}, H$ and $\Delta H^(q)$ defined above require specification of a probability distribution on elections and districting plans, $\pi$.
+The measures $U^{(G)}, U^{(L)}, H$ and $\Delta H(q)$ defined above require specification of a probability distribution on elections and districting plans, $\pi$.
 Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace with rate 2.[^3]
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^3]
 Then a party's vote share in precinct $k$, notated here as $s_k$, is modeled as $$
@@
 A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
+These concerns to not apply to our proposed measures.
 

### 2021-10-12 paper/paper.Rmd

--- 2021-10-12 paper/paper.Rmd
@@
 Harm is defined using the local, not global, utility measure, because inherently, global utility (and thus harm) will be constant for all voters of a particular party.
-In fact, a harm measure defined analogously using global utility would essentially be a difference in seats won by a voter's party between plans $q$ and $\tilde q$ (the exact form would be affected by the function $g$).
-A global harm measure is not particularly interesting, nor is it novel, being a standard measure of partisan harm in simulation-based redistricting litigation, up to a monotonic transformation [@2018league; @2019rucho; @2019lewis].
-**[CM: revisit ¶ based on NJ findings].**
+As we discuss below in Section \ref{subsec:diff-harm}, locally-based harm is connected to global electoral outcomes, anyway.
 
-```{=html}
-<!--  
-With Harm defined, we can thus define Partisan Equity as: $E[H|V(I) = DEM] = E[H|V(I) = REP]$
-
-Thus Fairness can be described as the signed movement away from equity.
-This naming works because it's condition on the votes and geographic distribution that you have, rather than some neutral midpoint that would be hard to justify.
-
-This clarifies our QOI and by later focusing on fairness, that gives us the analogous connection between bias (symmetry) and fairness (equity).
-
-Later, that also gives room in the discussion of the EGAP to talk about why you could describe a utility function based on $Pr(Vote not wasted)$, which 
--->
-```
 As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.

### 2021-10-14 paper/paper.Rmd
merge
edits

@@
 ---
-title: Measuring Individual Partisan Harm in Redistricting
+title: Individual Partisan Harm and Utility in Redistricting
 authors:
@@
 
-```{r mini-harm, fig.cap="Three potential districting plans for Mininois and Minissouris, showing district partisan lean (top row) and total voter harm versus the counterfactuals in Figure 2(a)-(b) (bottom row; darker shading indicates more harm).**CK: Should there be a second scale here?**"}
+```{r mini-harm, fig.cap="Three potential districting plans for Mininois and Minissouris, showing district partisan lean (top row) and total voter harm versus the counterfactuals in Figure 2(a)-(b) (bottom row; darker shading indicates more harm)."}
 include_graphics(here("paper/figures/mini_harm.pdf"))
@@
 
-Partisan bias can be considered the classic measure of partisan symmetry.
-The bias is the number of seats that would need to be reallocated to make the system symmetric at a chosen statewide proportion of votes, often 0.50.
-It should not be surprising then that these measures differ often and greatly.
+Partisan bias is probably the most widely-used measure of partisan symmetry.
+The bias, denoted by $\beta(s)$ is the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
+Like our proposed measures, this quantity is defined independent of electoral data and can be estimated using a uniform swing model.
 
-Further, by relying on alternative districting plans, we do not necessitate the use of partisan swings.
-This is of great benefit in scenarios where swings must be so significant in magnitude that they stray far from the data.
-Instead, the measures of harm and utility rely on $votes^{p}$ which can be model-free estimates of electoral performance of a party (namely historical election returns) or can be some sort of model, such as precinct-level election forecasts.
-Of course, the votes could even be generated by notions of partisan swings, but as a choice rather than a requirement.
+In competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
+But in more lopsided states, $\beta(\half)$ can give unintuitive results, which, if not interpreted carefully, can lead practictioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.
+
+Fundamentally, however, $\beta(s)$ measures partisan symmetry, and not partisan harm or utility.
+It should not be surprising, then, that it can and does "disagree" about the partisan effects of a particular plan.
+These "disagreements" are really just differences in the target quantity of interest.
+
+```{=html}
+<!--
+1.  Integrating yields a related measure of @nagle2015measures, which has also been criticized by @deford2020symmetry.
+2.  Focuses on only one part of S-V curve, which may not be relevant
 
@@
 The individualized focus of our measures are, however, portable to congressional, state legislative, or any other form of districting and elections.
-
-```{=html}
-<!--
-1.  Integrating yields a related measure of @nagle2015measures, which has also been criticized by @deford2020symmetry.
-2.  Focuses on only one part of S-V curve, which may not be relevant
 -->

### 2021-10-14 paper/paper.Rmd
discuss

--- 2021-10-12 paper/paper.Rmd
@@
 Using this class of metrics which describe the individual affect via representation consequently allow us to better align legalistic interpretations of harm for standing with political science understandings.
+Standing itself is an individual endeavor. Individuals must show that they themselves have suffered injury directly, not just via the larger system they partake in [@1992lujan].
 
@@
 
-We should say something about the importance in developing further measures which utilize simulation approaches
+This paper offers an attempt to bridge the quantitative gap between votes and representation.
+We break from existing redistricting literature and focus on the relationship between votes and representation by copartisans, rather than the relationship between votes and seats.
+This develops a new quantity of interest, partisan fairness, different from the standard of partisan symmetry for consideration.
+Chiefly, this centers measurement strategies around how individual voters in a given geography would have likely faired under counterfactual plans, not how parties would have faired under symmetric elections.
+
+This offers new purchase into the foundational questions of how to quantity the harms and benefits incurred to voters in district-based democracies--where the rules of defining districts are fixed, but the choice of how to how to draw the districts remains a policy question.
+This approach is flexible and we offer a baseline selection of clear and flexible, yet interesting, measures for improving the understanding of redistricting.
+Further work needs to be done to characterize how the potential rules of redistricting and other aspects, like the number or size of districts, impact these possibilities.
 
@@
 ```
+<!--
 # Computing Utility, Harm, and Fairness from Precinct-level Election Data {#app:formulas}
@@
 Formulas here.
-
+-->
 # Districting Measures in New Jersey {#app:nj}

### 2021-10-14 paper/paper.Rmd
Update paper.Rmd
abstract

@@
       To whom correspondence should be addressed. 
-      The authors thank ____ for helpful comments and discussion.
+      The authors thank {you} for helpful comments and discussion.
     department: Department of Statistics
@@
 abstract: |
-  Enter the text of your abstract here.
+  Standards for assessing gerrymandering have often focused on the symmetric treatment of parties by electoral systems.
+By focusing on counterfactual redistricting plans, we develop measures of utility in district-based representation.
+We use this to introduce an alternative quantity of interest, partisan fairness, and measures of deviations from this, differential harm.
+We offer strategies for estimating these measures by leveraging advances in redistricting sampling methodologies which can allow for generating large numbers of legal and feasible plans from a specified target distribution which encapsulates the stated legal constraints for a districting plan.
+We apply our measures to an example in New Jersey and demonstrate how the basis in counterfactual districting plans allows for straightforward interpretation.
 keywords:
@@
     latex_engine: pdflatex
+editor_options: 
+  markdown: 
+    wrap: sentence
 ---
@@
 Using this class of metrics which describe the individual affect via representation consequently allow us to better align legalistic interpretations of harm for standing with political science understandings.
-Standing itself is an individual endeavor. Individuals must show that they themselves have suffered injury directly, not just via the larger system they partake in [@1992lujan].
+Standing itself is an individual endeavor.
+Individuals must show that they themselves have suffered injury directly, not just via the larger system they partake in [@1992lujan].
 
@@
 ```
+```{=html}
 <!--
@@
 -->
+```
 # Districting Measures in New Jersey {#app:nj}

### 2021-10-14 paper/paper.Rmd
intro

--- 2021-10-14 paper/paper.Rmd
@@
 
+How do you measure a partisan gerrymander?
+Researchers, courts, and the public may agree that gerrymandering involves the manipulation of district boundaries for political gain, but beyond this vague definition are liable to disagree on a particular operationalization.
+
+Decades of academic literature on redistricting have focused on the fairness to parties.
+The normative standard for redistricting in this line of research is for electoral outcomes to be symmetric across political parties---what is commonly referred to as *partisan symmetry* [@katz2020].
+A variety of theoretical measures have been developed to capture this notion and to measure quantitatively any deviations from it.
+These measures may generally (and, as some have argued, necessarily) be derived from the *seats-votes curve*, which relates the expected number of seats a party wins to its share of votes received.
+
+The courts, meanwhile, have focused more on the impact an alleged gerrymander has on individual voters.
+To proceed with a legal challenge to a gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the gerrymander [@1964reynolds; @1962baker].
+
+Crucially, for a voter to be harmed by a districting plan, there must exist another districting plan which remedies the harm.
+An example we will return to frequently is that of a Massachusetts Republican: based on residential patterns of party voters, it is not possible to draw a Republican congressional district in Massachusetts, and so a Republican voter cannot be harmed by a districting plan that yields a fully Democratic delegation [@duch:18].
+
+We argue that understanding the redistricting process requires a feasible and reliable understanding of a counterfactual set of redistricting plans.
+To that end, we define two individual utility measures for districting plans: a *local utility* capturing the value to a voter of having a representative of the same party, and a *global utility* reflecting the value in maximizing the number of seats awarded to a voter's chosen party.
+From these two utility measures, we derive a measure of harm, based on counterfactual plans, and a notion of individual partisan fairness, which occurs when harms are evenly distributed across voters of each party.
+Deviations from fairness are measured by *differential harm*, which, as we show, is intimately connected to the number of seats won by each party.
+
+Taken together, our measures offer a framework for understanding the individual impacts of a districting plan, and how these impacts are borne by members of each party.
+The use of counterfactual plans allows us to better relate legalistic interpretations of harm to those of political science.
+
+To estimate these measures requires generating counterfactual districting plans, for which we rely on advances in redistricting simulation methodology.
+Early methodologies for simulating plans relied on randomness to generate many different districting plans [@ciri:darl:orou:00; @chen:rodd:13; @deford2019].
+These methods greatly enhanced the understanding of redistricting within contexts of a particular state with set populations.
+Yet, lacking a specific target probability distribution, these heuristic approaches to simulation do not allow for probabilistic interpretations, and risk misrepresenting the total space of possible districting plans [@fifield2020essential].
+Recent advances in Monte Carlo methods allow us move beyond random simulation and towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @mccartan2020].
+We can thus develop a reliable set of counterfactual plans that can be tuned to reflect the constraints imposed by the laws and practices of a particular state [@matt:hers:21].
+
+We offer a reminder that geographic context is key to understanding common problems in political science; elections themselves are geographically determined.
+Recent research offers important resolution of the degree of partisan sorting in the US [@enos:brown:21].
+By combining those understandings with the necessities of redistricting, we are able to expand the localized aspect of harm.
+Further, we suggest that while the efficiency gap may not incorporate geographic information, it often helps capture individual harm better than other common measures.
+This is a great benefit as an indicator to detect extreme gerrymanders in cases where sufficiently local geographic data to simulate plans is not available.
+We further clarify that common existing measures of partisan fairness -- while they may capture the difficulties offered to a party -- do not necessarily capture the representational harms incurred by individuals who live in districts which do not at all represent their preferences.
+
+```{=html}
+<!--
 Districting is a complicated process.
@@
 
-Decades of literature on redistricting have focused on the fairness to parties.
-This literature defers to symmetric outcomes for parties, commonly referred to as partisan symmetry [@katz2020].
-Partisan symmetry offers the idea that the number of seats awarded to a party by a set of votes would be the same if the labels on the parties were swapped.
-Partisan symmetry is a useful quantity of interest which has been widely used to improve academic and legal understanding of the redistricting process.
+Partisan symmetry is a useful quantity of interest, which has been widely used to improve academic and legal understanding of the redistricting process.
 We do not challenge the value of partisan symmetry, but instead suggest that a different question should be asked to describe the fairness of district based democracies.
 
-We argue that understanding the redistricting process requires a feasible and reliable understanding of the counterfactual set of redistricting plans.
-Our metrics offer a framework for understanding approximately individual harms, benefits, and net fairness.
-Using counterfactual plans to understand this space allows us to properly calibrate understandings of the impact of redistricting plans on individual understanding.
-Thus, we create measures which capture the possibility that no plans exist which favor one community [@duch:18].
-Using this class of metrics which describe the individual affect via representation consequently allow us to better align legalistic interpretations of harm for standing with political science understandings.
-
-To describe a reasonable counterfactual to measure our quantity of interest, we rely on advances in redistricting simulation methodologies.
-Early methodologies for simulating plans relied on randomness to generate many different plans and enhanced understanding of redistricting greatly [@ciri:darl:orou:00; @chen:rodd:13; @deford2019].
-These methods greatly enhanced the understanding of redistricting within contexts of a particular state with set populations.
-Yet, without a known target distribution, these heuristic approaches to simulation fall do not offer necessary guarantees for confidence in the simulated plans to represent the total space of possible plans [@fifield2020essential].
-Recent advances in Monte Carlo methods allow us move beyond simulation towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @mccartan2020].
-
-Using these sampling methods, we can then develop a reliable counterfactual that can be tuned to consider the sampling constraints imposed by the laws and practices of a state.
-This counterfactual allows us to characterize the likely plans from a state.
-Often, we can would then call this an ensemble analysis, which is a flexible approach to sampling redistricting plans by simulation, relying on large numbers of plans for stability [@matt:hers:21].
-using this counterfactual, we thus create measures of harm, utility, and fairness which are based on a single plan compared to a set of other reasonable and feasible plans which incorporate the necessary legal constraints to make a plan viable.
-
-We offer a reminder that geographic context is key to understanding common problems in political science.
-Elections themselves are geographically determined.
-Recent research offers important resolution of the degree of partisan sorting in the US [@enos:brown:21].
-By combining those understandings with the necessities of redistricting, we are able to expand the localized aspect of harm.
-Further, we suggest that while the efficiency gap may not incorporate geographic information, it often helps capture individual harm better than other common measures.
-This is a great benefit as an indicator to detect extreme gerrymanders in cases where sufficiently local geographic data to simulate plans is not available.
-We further clarify that common existing measures of partisan fairness -- while they may capture the difficulties offered to a party -- do not necessarily capture the representational harms incurred by individuals who live in districts which do not at all represent their preferences.
-
-TODO: Add findings!
-
-```{=html}
-<!--
 Understanding fairness in electoral systems has long suffered from a focus on parties over the individuals who vote for them.
@@
 Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
-Section \ref{sec:empirical} applies the proposed measures to a real-world example to illustrate their value and these comparsions.
+Section \ref{sec:empirical} applies the proposed measures to congressional redistricting in the state of New Jersey to illustrate their value and these comparisons.
 Section \ref{sec:discussion} concludes.
@@
 Like our proposed measures, this quantity is defined independent of electoral data and can be estimated using a uniform swing model.
+Each value of $s$ yields information about a different point on the seats-votes curve.
 
@@
 
-Our measures greatly differ from the Mean-Median Difference.
-With the focus on individuals, the considerations of the mean district are only related to individuals in-so-far as they describe the general propensity to be harmed or benefited by a particular map.
-The consideration of a median district is similarly distanced.
-As with partisan bias, as our measures do not attempt to measure partisan symmetry, our measures can be applied to congressional delegations.
+The mean-median difference straightforwardly measures the difference between the mean and median district vote shares.
+In practice, the mean-median differences is moderately correlated with other partisan redistricting measures.
+However, considerations of the mean and median district are only related to individuals insofar as they describe the general propensity to be harmed or benefited by a particular map.
+As [@katz2020] discuss, the mean-median difference can be interpreted as an estimator of *votes-denominated bias*, another value derived from the seats-votes curve.
 

### 2021-10-14 paper/paper.Rmd

@@
 abstract: |
-  Enter the text of your abstract here.
+  Most existing standards for assessing partisan gerrymandering focus on the symmetric treatment of parties by electoral systems.
+We develop measures of utility and harm for individual voters by considering counterfactual redistricting plans, and use these notions to construct a normative standard of partisan fairness which is centered on the individual.
+We estimate these measures from real-world data by leveraging recent advances in redistricting sampling methodology, which allow researchers to generate large numbers of districting plans which reflect local legal and practical requirements.
+We apply our measures to congressional redistricting in New Jersey, and compare them to existing measures of partisan symmetry and competitiveness.
 keywords:
@@
 
-```{=html}
-<!--
-Districting is a complicated process.
-Residents of a state or locality must be partitioned into districts based on the geographies that they live in.
-These districts must adhere to many constraints, often requiring components such as contiguity and equal population.
-In drawing the districts, it's easy to lose sight of the individuals impacted by district drawing.
-Often, we consider how parties succeed or fail under the electoral environments generated by enacted plans.
-Yet this misses that parties are formed by voters who ally with parties.
-Rather than considering parties, we ask how individuals are harmed or benefited by the districting process?
-
-Partisan symmetry is a useful quantity of interest, which has been widely used to improve academic and legal understanding of the redistricting process.
-We do not challenge the value of partisan symmetry, but instead suggest that a different question should be asked to describe the fairness of district based democracies.
-
-Understanding fairness in electoral systems has long suffered from a focus on parties over the individuals who vote for them.
-There is no doubt that parties are interesting units in electoral discussions.
-They offer stability over candidates, information for the under-informed, and shortcuts for understanding preferred vote choices.
-Yet, court histories and academic traditions value the harm of the individual over the harm to a transient political body, such as the party.
-
-In this article, we offer an approach to understanding partisan fairness through a lens of individual harm and utility.
-This does not supplant existing understandings of partisan fairness, but instead offers an alternative rooted in understanding how the individual potential voters are impacted by districting plans.
-
-Several existing methods of redistricting simulation rely on randomness, such as Random Seed and Grow (Chen and Rodden 2013) or Recombination (Deford et al. 2021).
-Moving beyond these heuristic methods, we can use sampling methods that have defined target distributions to ensure that we characterize the space of applicable plans properly.
-Methods which operationalize a statistics-based approach include Fifield et al. 2020, McCartan and Imai (2020), and Autrey et al (2020).
-Using these methods advance us beyond the critiques of Katz, King, and Roseblatt (2020) by sampling plans using weights within a Gibbs distribution, rather than assuming that uniform sampling weights alone are sufficient.
-
-Metrics often associated with partisan symmetry, such as TODO, are indicative of violations of partisan symmetry.
-Notably, these are only defined for full legislatures (see Katz, King, and Rosenblatt (2021)).
-As such, theory-based interpretations of these measures do not indicate signs of unfairness to parities in the congressional district setting, unless they are consistent across many, many states.
-We offer measures of partisan harm which are rooted directly in the representational harms inflicted upon individuals by those who choose to draw a map.
-
-These measures are deeply connected to the law and constitutional understandings of partisan fairness, due to their related roots in the individual.
-
--   introduce partisan symmetry
-
-    -   Katz, King, Rosenblatt 2019; Deford et al 2021; Katz, King, Rosenblatt 2021
-
--   Election law side is very indivudalized
-
-    -   Standing discussions
-
-        -   Lujan v Defenders of Wildlife (1992)
-        -   Shaw v. Reno (1993) -- racial gerrymandering
-        -   Issacharoff and Karlan 1998
-
--   Recent, related work
-
-    -   EGap
-    -   Partisan Dislocation
-
--   Important to consider benefits and harms to individual voters
-
--   Power of Ensemble Analysis
-
-    -   Recent advances allow us to make claims about harm based in counterfactuals that are realistic and statistically useful
--->
-```
 ## Running Examples
@@
 
-We should say something about the importance in developing further measures which utilize simulation approaches
+This paper is an attempt to bridge the quantitative gap between votes and representation.
+We break from existing redistricting literature and focus on the relationship between votes and representation by copartisans, rather than the relationship between votes and seats.
+This focus naturally leads to a new normative standard of partisan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters of each party.
+This may occur regardless of whether the districting plan achieves partisan symmetry.
+
+This approach offers new purchase into the foundational questions of how to quantify the harms and benefits that accrue to voters in district-based democracies, where the rules of defining districts are fixed, but the choice of how to how to draw the districts remains a policy question.
+While the overall framework is flexible, we offer a selection of simple yet interesting measures for improving the individual partisan implications of a districting plan.
+Further work is needed to characterize how existing redistricting requirements and practices, such as the number or size of districts, or encouragement of competitive districts, interact with these measures.
 
@@
 ```
+```{=html}
+<!--
 # Computing Utility, Harm, and Fairness from Precinct-level Election Data {#app:formulas}
@@
 Formulas here.
-
+-->
+```
 # Districting Measures in New Jersey {#app:nj}

### 2021-10-14 paper/paper.Rmd
-basically

@@
 \begin{restatable}{prop}{propcomp} \label{prop:comp}
-Basically, if turnout is equal across districts, for an election $v\in\Elecs$ and districting plan $q\in\Plans$ and two political parties, we have
+If turnout is equal across districts, for an election $v\in\Elecs$ and districting plan $q\in\Plans$ and two political parties, we have
 \begin{align*}

### 2021-10-14 paper/paper.Rmd
fix appendices
small edits

@@
 We develop measures of utility and harm for individual voters by considering counterfactual redistricting plans, and use these notions to construct a normative standard of partisan fairness which is centered on the individual.
-We estimate these measures from real-world data by leveraging recent advances in redistricting sampling methodology, which allow researchers to generate large numbers of districting plans which reflect local legal and practical requirements.
+We calculate these measures on real-world data by leveraging recent advances in redistricting sampling methodology, which allow researchers to generate large numbers of districting plans which reflect local legal and practical requirements.
 We apply our measures to congressional redistricting in New Jersey, and compare them to existing measures of partisan symmetry and competitiveness.
@@
 
-We argue that understanding the redistricting process requires a feasible and reliable understanding of a counterfactual set of redistricting plans.
+We argue that understanding the redistricting process requires a counterfactual set of feasible redistricting plans.
 To that end, we define two individual utility measures for districting plans: a *local utility* capturing the value to a voter of having a representative of the same party, and a *global utility* reflecting the value in maximizing the number of seats awarded to a voter's chosen party.
@@
 
-To estimate these measures requires generating counterfactual districting plans, for which we rely on advances in redistricting simulation methodology.
+To calculate these measures in practice requires generating counterfactual districting plans, for which we rely on advances in redistricting simulation methodology.
 Early methodologies for simulating plans relied on randomness to generate many different districting plans [@ciri:darl:orou:00; @chen:rodd:13; @deford2019].
@@
 ```{=tex}
-\clearpage
+\vspace{1in}
 \appendix
@@
 ```
+# Computing Utility, Harm, and Fairness from Precinct-level Election Data {#app:formulas}
+
 ```{=html}
 <!--
-# Computing Utility, Harm, and Fairness from Precinct-level Election Data {#app:formulas}
 

### 2021-10-20 paper/paper.Rmd
1-1 and many-many appendix
@CoryMcCartan does this fit in the main paper or should it be appendix material?

If main paper, we probably want a figure that shows the idea?

@@
 ```
+
+# Misrepresentation Map
+
+The intuition that drives these measures is the misrepresentation map.
+Consider a state that is partitioned into smaller geographies, typically called precincts or voter tabulation districts.
+These units act as the atoms of redistricting.
+We assume that they are not split in the redistricting process.
+Thus, districts are made of whole precincts.
+For a given precinct, we can thus calculate the expected party share of the district that a precinct is in for any given plan.
+The expected party share can take many forms, as mentioned in *Modeling Election Outcomes*.
+By combining the information about the geographic precinct, we can measure the expected party share that an individual living in that precinct would expect to experience.
+
+Visually, we can consider a map colored by party share of the district that each precinct is assigned to.
+Most often, we represent this in a blue to red gradient which offers the typical Democratic or Republican representative dynamic.
+Given a counterfactual, we can then take the difference of the expected party share, precinct-by-precinct.
+This offers a precinct-level coloration based on difference in expected party share at the district level.
+In the typical two-party scenario, using Democratic two-party vote share, a positive difference would indicate that the individuals in those precincts are likely to be represented by Democrats.
+A negative difference would indicate that the individuals in those precincts are likely to be represented by Republicans.
+Thus, positives would be map-matically represented by blues and negatives by reds.
+
+Note that this takes a similar visual form as a dislocation map, though these are conceptually distinct [@defoeubarodd21].
+The dislocation map indicates within a single plan how nearby individuals are likely or unlikely to be represented by the same party.
+
+We focus largely on the case of a misrepresentation map that is the difference between a single enacted districting plan and an ensemble of alternate, legal districting plans.
+This does not need to be the case, however.
+A misrepresentation map can indicate the difference between two singular maps, calculating the difference in expected district-level party representation.
+This would allow direct computation of utility, fairness, or other metrics based on the difference in the two maps.
+Similarly, given two sets of ensembles, likely generated under different target distributions, we can create a similar misrepresentation map that would allow us to estimate interesting measures, such as the differential harm caused by the inclusion of an additional criteria in state district drawing rules.

### 2021-10-26 paper/paper.Rmd
comments aprw

@@
       To whom correspondence should be addressed. 
-      The authors thank {you} for helpful comments and discussion.
+      The authors thank participants of the Harvard American Politics Research Workshop for helpful comments and discussion.
     department: Department of Statistics
@@
 How do you measure a partisan gerrymander?
-Researchers, courts, and the public may agree that gerrymandering involves the manipulation of district boundaries for political gain, but beyond this vague definition are liable to disagree on a particular operationalization.
+Researchers, courts, and the public may agree that gerrymandering involves the manipulation of district boundaries for political gain, but beyond this vague definition, most are liable to disagree on a particular operationalization.
 

### 2021-11-05 docs/aprw-2021.10.26.Rmd
response, uncondensed

---
title: "APRW Comments"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Major Comments

-   Utility function designs have an assumption that past reform does not impact anything

-   Why not measure the greatest good for the greatest number as your focus

-   Why use New Jersey?
    (We should probably use Illinois. No party data for Missouri.)

    -   If we show NJ, we should include the enacted plan

-   Is Harm too strong a word?

    -   For general things, maybe?
    -   If we want to tie it to court-speak, they use harm and injury for this, so it fits.

-   Not 100% clear why the certain maps end up the ways that they do: unpack this

    -   Why might local maximizing not just be super packed (ie compactness and political geography)

        -   if this is a strength, use it!

-   Setting utility as 0 for nonvoters misses something.
    How might we compare voters vs eligible voters in a state?

    -   Use party registration

        -   if FL, then we get race, party, and turnout all in one

-   For framing the contribution, spend more time thinking about the harm and drop the global aspects

-   Can we consider more complex utility functions?

    -   Mix across types of utility
    -   Basic mixing of local and global (rather than pure)

-   Should think about how ideology mattesr

    -   Use a school district?
    -   Lean into a locality for a cleaner example where party is different

-   How does precinct heterogeneity map into this (Chris needs to work more on that party vs race presentation)

### Steve Comments

-   Drop the whole utility thing

-   lean into how it relates to existing voters

    -   If we weight by voters, do we recover some egap approximation or transformation?

-   Add in an aspect of racial mapping?

    -   include optimization over two dimension -- racial and partisan harm

        -   large 1990s-2000s literature on the costs and benefits of MMDs (Washington etc)

-   How does adding constraints to a simulation change the dimensions of harm?

    -   If we want tot htink about constraints more in how the law does, then spend more time on that
    -   Also consider if this is how *we* think about it

-   Ask: What does equal protection imply?

    -   If you start at equal protection, then you need to think about turnout
    -   That's where there's purchase to talk about differing turnout

-   The big open world for this is if we can get better understanding of what makes a district performant for racial party reasons

    -   Can we use this mechanism to create a threshold?
    -   Mix party with census race
    -   Race x candidate data and code from [Shiro via Jeremiah](https://wholeads.us/research/2020-ballot-demographics/)

## Presentation Comments

-   Make clearer the generalization to probabilities
-   Put the utility maximization together on the slides with the harm for the mini cases

## Paper Comments ()

-   Paper seems very motivated by the court emphasis on individual representation

    -   Seems like a strong approach to trying to understand the microfoundations of redistricting

-   Paper offers lots of measures but then seems to indicate that what we care about is fairness

    -   not totally clear why fairness is an *individual* measure (seems very party oriented)
    -   Can this be tied more clearly to individual harm?

-   Nonvoters and can't voters utility of 0

    -   Spend more time justifying this (is it standing?)
    -   How could we extend to asymmetric partisan preferences?

-   Rawls social welfare (??)

-   Expand more onto the idea that global considerations don't really matter for fairness

-   Does local utility best capture partisan cheerleaders

-   Expand to higher moments of harm

    -   Do we maybe sometimes care more about minimizing the change in harm as swings occur?

-   Spend more time discussing the actual interpretation of what the harm measure gives you

    -   Doesn't have the super clear % seats interpretation, so this seems very useful

-   Justify the use of `g() = log()`.

-   I wrote "systematicity" and underlined it from Jacob's discussion (why? I don't know.)

-   Is there room for some basic discussion of tiered preferences?

-   For discussing New Jersey, spend some time talking about the when/where/why/so what of the differences across measures?
    [less meaningful if we focus on partisan harm]

### 2021-11-05 paper/paper.Rmd
response slides with sections + add note to paper

@@
 We offer a reminder that geographic context is key to understanding common problems in political science; elections themselves are geographically determined.
+<!-- people super hate this sentence above, which is fair, though is a problem in this lit -->
 Recent research offers important resolution of the degree of partisan sorting in the US [@enos:brown:21].

### 2022-01-18 paper/paper.Rmd
edit
start revisions

--- 2021-10-26 paper/paper.Rmd
@@
 ---
-title: Individual Partisan Harm and Utility in Redistricting
+title: Individual and Differential Harm in Redistricting
 authors:
@@
 abstract: |
-  Most existing standards for assessing partisan gerrymandering focus on the symmetric treatment of parties by electoral systems.
-We develop measures of utility and harm for individual voters by considering counterfactual redistricting plans, and use these notions to construct a normative standard of partisan fairness which is centered on the individual.
-We calculate these measures on real-world data by leveraging recent advances in redistricting sampling methodology, which allow researchers to generate large numbers of districting plans which reflect local legal and practical requirements.
-We apply our measures to congressional redistricting in New Jersey, and compare them to existing measures of partisan symmetry and competitiveness.
+  Abstract here.
 keywords:
-  - partisan fairness
   - representation
@@
 
-How do you measure a partisan gerrymander?
-Researchers, courts, and the public may agree that gerrymandering involves the manipulation of district boundaries for political gain, but beyond this vague definition, most are liable to disagree on a particular operationalization.
+-   what makes a good redistricting plan?
 
-Decades of academic literature on redistricting have focused on the fairness to parties.
-The normative standard for redistricting in this line of research is for electoral outcomes to be symmetric across political parties---what is commonly referred to as *partisan symmetry* [@katz2020].
-A variety of theoretical measures have been developed to capture this notion and to measure quantitatively any deviations from it.
-These measures may generally (and, as some have argued, necessarily) be derived from the *seats-votes curve*, which relates the expected number of seats a party wins to its share of votes received.
+-   on the partisan side, lots of attention on the fairness to parties
 
-The courts, meanwhile, have focused more on the impact an alleged gerrymander has on individual voters.
-To proceed with a legal challenge to a gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the gerrymander [@1964reynolds; @1962baker].
+    -   partisan symmetry [@katz2020]
 
-Crucially, for a voter to be harmed by a districting plan, there must exist another districting plan which remedies the harm.
-An example we will return to frequently is that of a Massachusetts Republican: based on residential patterns of party voters, it is not possible to draw a Republican congressional district in Massachusetts, and so a Republican voter cannot be harmed by a districting plan that yields a fully Democratic delegation [@duch:18].
+    -   seats-votes curve
 
-We argue that understanding the redistricting process requires a counterfactual set of feasible redistricting plans.
-To that end, we define two individual utility measures for districting plans: a *local utility* capturing the value to a voter of having a representative of the same party, and a *global utility* reflecting the value in maximizing the number of seats awarded to a voter's chosen party.
-From these two utility measures, we derive a measure of harm, based on counterfactual plans, and a notion of individual partisan fairness, which occurs when harms are evenly distributed across voters of each party.
-Deviations from fairness are measured by *differential harm*, which, as we show, is intimately connected to the number of seats won by each party.
+-   on the racial side, ... ?
+    (VRA focus? looking at race of candidates vs choice? what? I don't know)
 
-Taken together, our measures offer a framework for understanding the individual impacts of a districting plan, and how these impacts are borne by members of each party.
-The use of counterfactual plans allows us to better relate legalistic interpretations of harm to those of political science.
+-   courts, on the other hand, have focused on individual standing [@1964reynolds; @1962baker]
 
-To calculate these measures in practice requires generating counterfactual districting plans, for which we rely on advances in redistricting simulation methodology.
-Early methodologies for simulating plans relied on randomness to generate many different districting plans [@ciri:darl:orou:00; @chen:rodd:13; @deford2019].
-These methods greatly enhanced the understanding of redistricting within contexts of a particular state with set populations.
-Yet, lacking a specific target probability distribution, these heuristic approaches to simulation do not allow for probabilistic interpretations, and risk misrepresenting the total space of possible districting plans [@fifield2020essential].
-Recent advances in Monte Carlo methods allow us move beyond random simulation and towards sampling plans from this definable space [@fifield2020mcmc; @carter2019; @mccartan2020].
-We can thus develop a reliable set of counterfactual plans that can be tuned to reflect the constraints imposed by the laws and practices of a particular state [@matt:hers:21].
+-   **we argue that individual standing/harm is an excellent lens to unify the study and measurement of redistricting plans**
 
-We offer a reminder that geographic context is key to understanding common problems in political science; elections themselves are geographically determined.
-Recent research offers important resolution of the degree of partisan sorting in the US [@enos:brown:21].
-By combining those understandings with the necessities of redistricting, we are able to expand the localized aspect of harm.
-Further, we suggest that while the efficiency gap may not incorporate geographic information, it often helps capture individual harm better than other common measures.
-This is a great benefit as an indicator to detect extreme gerrymanders in cases where sufficiently local geographic data to simulate plans is not available.
-We further clarify that common existing measures of partisan fairness -- while they may capture the difficulties offered to a party -- do not necessarily capture the representational harms incurred by individuals who live in districts which do not at all represent their preferences.
+-   core idea: counterfactuals + voter choice.
+    Is a voter more likely to be represented by the candidate of their choice under a different redistricting plan?
+    If so, they've been harmed.
 
-## Running Examples
+    -   modern simulation methods allow us to produce realistic counterfactuals
 
-To illustrate the notation and measures we introduce, this article will make use of three running examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichussets.
+    -   by analyzing how harms fall differentially along party, race, ideological, etc. lines, we can connect individual harm to more global measures based on parties, etc.
+
+## Outline of the Paper
+
+Section \ref{sec:motivation} motivates our framework for measuring harm and illustrates the key concepts in the context of partisan gerrymandering for three simple examples.
+Section \ref{sec:indiv-harm} formalizes the intuition of Section \ref{sec:motivation} and defines a general measure of individual and differential harm.
+Section \ref{sec:estimating} discusses estimating these measures from observed election results.
+Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
+Section \ref{sec:empirical} applies the proposed measures to congressional representation in New Jersey and primary elections for the New York City Council.
+Section \ref{sec:discussion} concludes.
+
+# Motivation {#sec:motivation}
+
+To illustrate the concepts and measures we introduce, we will make use of three running examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichussets.
 Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
@@
 
-## Outline of the Paper
+## Minichusetts: counterfactuals matter
 
-Section \ref{sec:voter-util} provides the necessary notation and introduces our measures of voter utility.
-Section \ref{sec:indiv-harm} builds a measure of individual harm and a derived measure of partisan fairness from this notion of utility.
-Section \ref{sec:estimating} discusses estimating these measures from observed election results.
-Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
-Section \ref{sec:empirical} applies the proposed measures to congressional redistricting in the state of New Jersey to illustrate their value and these comparisons.
-Section \ref{sec:discussion} concludes.
+-   40% GOP but no GOP districts.
+    Is this unfair?
 
-# Voter Utility in Election Outcomes {#sec:voter-util}
+-   E-gap would say so.
+    Partisan symmetry: ?.
 
-The utility voters gain from a particular election outcome naturally varies across the electorate, depending on factors such as political engagement, partisanship, race and ethnicity, individual policy preferences, and individual candidate attributes.
-Here, we restrict our attention to what we believe are the two dominant contributors to partisan utility: which party represents a voter's district, and how many overall seats a voter's preferred party wins.
+-   But impossible to draw a GOP district
 
-## Notation and Setup
+-   Without a remedy (counterfactual), there can be no harm
 
-Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate from one of a set of political parties $\Parties$ to represent each of a set of legislative districts, indexed by $\Distrs$.
-We take these three sets to be fixed.
+## Mininois: individual harm
 
-Voters vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
-In an election, voters either vote for a candidate running in their district, or abstain from voting.
-So we write $v:\Voters\to\Parties\cup\{\textsc{abstain}\}$, with $v(i)$ the party that voter $i\in\Voters$ voted for in election $v$.
+-   show a fair map and a gerrymandered map
 
-Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is formally represented by a function $q:\Voters\surjto \Distrs$.
-The set of all possible districting plans is $\Plans$.[^1]
+-   zoom into 1 or 2 precincts & calculate individual harm of D & R voters in those precincts
 
-[^1]: Note that these districting plans do not depend on the election; as a result, we implicitly assume that voters do not move residences during the period under consideration.
-    This is not a necessary restriction and can be dealt with formally by first mapping voters to precincts (varying by election) and then mapping precincts to districts (constant across elections).
-    Once we aggregate or average across all voters in a party or district, the differences between these two approaches disappear.
+-   show map of differential harm (with
 
-Votes may be tallied at the district level, by party, by defining for each party $p\in\Parties$ $$
-    \votes^{(p)}: \Distrs\times\Elecs\times\Plans \to \N, \qquad
-    \votes^{(p)}(d, v, q) = \sum_{i\in q^{-1}(d)} \ind_{\{p\}}(v(i)).
-$$ The political party of the candidate with the most votes wins each district, as captured by the function $$
-    \win: \Distrs\times\Elecs\times\Plans\to P, \qquad
-    \win(d, v, q) = \argmax_{p\in\Parties} \votes^{(p)}(d, v, q).
-$$Thus an individual voter $i\in\Voters$ is represented after election $v$ by political party $\win(q(i), v, q)$, and each political party can be said to win a number of seats, formalized as$$
-    \seats:\Parties\times\Elecs\times\Plans\to\N,
-    \qquad \seats(p,v,q) = \sum_{d\in\Distrs} \ind_{\{p\}}(\win(d,v,q)).
-$$
+## Minissouri: differential harm
 
-## Global Outcome Utility
+-   show a fair map and a gerrymandered map
 
-We posit that the utility of the global election outcome to a voter is a strictly increasing function of the share of total seats that the voter's preferred party wins.
-Their global outcome utility function for an election outcome $e\in\Elecs$ and districting plan $q\in\Plans$ can thus be written$$
-    u^{(G)}:\Voters\times\Elecs\times\Plans\to\R,
-    \qquad u^{(G)}(i,v,q) = g\qty(\frac{\seats(v(i), v, q)}{|\Distrs|}),
-$$for $g:[0,1]\bijto\R$ a strictly increasing function.
+-   tally harm by party
 
-What kind of electoral systems does this utility function encourage, if the goal is maximizing total voter utility (or, equivalently, average voter utility)?
-It depends on the form of $g$, as the following result describes.
-Proofs of this and all other propositions are deferred to Appendix \ref{app:proofs}.
-
-```{=tex}
-\begin{restatable}{prop}{propglobal} \label{prop:global}
-Let $S=\{s_p:p\in\Parties\}$ be a set of party popular vote shares, and $f:[0,1]\bijto[0,1]$ be a rule for assigning each party a share of seats according to their share of the popular vote.
-Define the total utility for this rule as \[
-    u_{\mathrm{tot}}^{(G)}(f, v) \dfeq \sum_{i\in\Voters} g\qty(f\qty(s_{v(i)})).
-\]
-Then if $g$ is convex, $u_{\mathrm{tot}}^{(G)}$ is maximized for $f(x)=\ind_{[\max S, 1]}(x)$, and if $g$ is strictly concave, the maximizer of $u_{\mathrm{tot}}^{(G)}$ will be a strictly increasing function.
-In particular, if $g(x)=\log(x)$ and every party receives at least one vote, then $f(x)=x$ maximizes $u_{\mathrm{tot}}^{(G)}$.
-
-\end{restatable}
-```
-Thus, if we are willing to adopt the logarithm for the functional specification of $u^{(G)}$, then this global outcome utility measure will encourage districting systems that create proportionality between party vote shares and seat shares.
-
-## Local Outcome Utility
-
-The specification of the utility of the local election outcome is more straightforward.
-We define$$
-    u^{(L)}:\Voters\times\Elecs\times\Plans\to\R,
-    \qquad u^{(L)}(i,v,q) = \ind\{v(i)=\win(q(i), v, q)\},
-$$that is, a voter gains utility only if they winner of their legislative district belongs to their preferred political party.
-
-Attempting to maximize the average local utility across the electorate would lead to a districting plan where every voter is assigned to a district with other voters of the same party preference; in this way, every voter's utility would be $1$.
-Whether such an districting plan is possible depends on the geographical layout of voters and the legal constraints on the districting process---districts must be contiguous, balanced in population, constructed from whole precincts, geographically compact, and so on.
-But the general tendency of this utility measure (in the context of a single election, at least), like the global utility measure, will be to push plans towards proportionality.
-Unlike with the global measure, a specific kind of proportionality is preferred: that which creates as many lopsided districts as possible (what is often termed a *bipartisan gerrymander*).
-This point is discussed further in Section \ref{sec:competitive} below.
-
-## Examples
-
-To illustrate these two utility measures, Figure \ref{fig:mini-util-opt} shows three plans for our running examples which maximize both global (with $g=\log$) and local utility.[^2]
-In Mininois and Minissouri, the overall division of seats is as close to proportional as possible.
-In Minichusetts, however, it is impossible to give any seats to Republicans (as long as districts are constructed from precincts); all plans have the same utility.
-In none of the three states is it possible to create a perfect bipartisan gerrymander, but the plans for Mininois and Minissouri get as close as possible by packing Democratic voters into cities (and Republican voters into rural areas, where feasible).
-
-[^2]: These examples present a slight complication---what to do when a party wins no seats?
-    The easiest solution, perhaps, is to define the utility to that party's voters to be $-1$; this will make the utility well-defined while still encouraging proportionality.
-
-```{r mini-util-opt, fig.cap="Utility-maximizing redistricting plans for the three running examples, with each district shaded by its partisan lean.
-Black lines indicate districts.
-Red and blue dots indicate Republicans and Democrats, respectively."}
-include_graphics(here("paper/figures/minis_util_opt.pdf"))
-```
-
-## Multiple Elections or Districting Plans
-
-The utility measures above are defined for a single election outcome $v\in \Elecs$ and districting plan $q\in\Plans$.
-But often we will want to consider utility averaged over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans.
-
-To allow this, we replace the fixed election districting plan with random variables $(V,Q)\sim \pi$, where $\pi$ is a joint probability measure on election outcomes and districting plans.
-The randomness in election outcomes comes from variations in factors such as voter preferences, the political environment, candidate quality, and campaign events.
-The randomness in districting plans reflects the constraints and preferences of the plan-drawing process.
-As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V\indep Q$, though this is not required for our theoretical analysis.
-We will also consider a voter $I$ taken uniformly at random from $\Voters$, with $I\indep\pi$.
-This has the effect of treating voters exchangeably and valuing their personal utilities equally.
-
-Then we may define the utilities for a (randomly selected) voter as $$
-    U^{(G)} \dfeq u^{(G)}(I, V, Q) \qand
-    U^{(L)} \dfeq u^{(L)}(I, V, Q).
-$$Thus, the local utility for a single voter $i$ becomes $U^{(L)}\mid I=i$; the electorate's average utility for election outcome $v$ becomes $\E[U^{(L)}\mid V=v]$, and so on.
-
-# Towards Individualized Representational Harm {#sec:indiv-harm}
-
-Utility may be defined solely in terms of realized election outcomes and districting plans.
-Harm, we argue, cannot.
-Harm is not the same as disutility, for to be harmed by a decision a voter must have a better outcome in the counterfactual world where another decision was made.
-Put another way, in the context of districting plans, if no alternative districting plan could provide more utility to a voter, then that voter was not harmed by the districting plan chosen.
-
-## Individual Harm
-
-To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
-These counterfactuals could arise from a posited probability distribution capturing the factors relevant to a non-partisan redistricting process, or they could be a set of alternative redistricting plans that were considered as part of the legislative process leading to a final enacted plan.
-Then we may define the harm to a voter as $h:\Voters\times\Elecs\times\Plans^2\to\R$, \begin{align*}
-    h(i,v,q,\tilde q) &= (1 - u^{(L)}(i, v, q))\cdot u^{(L)}(i, v, \tilde q) \\
-    &= \ind\{v(i)\neq\win(q(i), v, q)\}
-        \ind\{v(i)=\win(\tilde q(i), v, \tilde q)\},
-\end{align*}that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.
-
-Harm is defined using the local, not global, utility measure, because inherently, global utility (and thus harm) will be constant for all voters of a particular party.
-As we discuss below in Section \ref{subsec:diff-harm}, locally-based harm is connected to global electoral outcomes, anyway.
-
-As with the utility measures, this harm measure naturally extends to multiple elections and districting plans.
-Make the counterfactual plan a random variable $\tilde Q$, with $(V,Q,\tilde Q)\sim\pi$ such that $Q\indep\tilde Q$.
-Then the harm to a random voter is $H \dfeq h(I, V, Q, \tilde Q)$.
-
-The expected harm to an individual $i\in\Voters$ then becomes $$
-   \E[H\mid I=i] = \Pr[V(i)\neq\win(Q(i), V, Q)
-    \text{ and }V(i)=\win(\tilde Q(i), V, \tilde Q)],
-$$
-
-that is, the probability that the voter is misrepresented under the adopted plan but would not have been under a counterfactual.
+-   end with transition: there are natural avenues for generalization
 
@@
 
-## Differential Harm and Fairness {#subsec:diff-harm}
+In the districting plan examples of Figure \ref{fig:mini-harm}, the differential harm of the Democratic gerrymander of Mininois is $-0.519$, the differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
+In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).
+
+# Individual Representational Harm
+
+-   this section: formalize and generalize the concepts in Section \ref{sec:motivation}.
+
+## Setup
+
+-   Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of legislative districts, indexed by $\Distrs$.
+
+-   Voters vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
+    In an election, voters either vote for a candidate running in their district, or abstain from voting.
+    We write $v_i$ [**CM: should this be** $v(i)$?]
+    to mean the candidate that individual $i$ votes for in election $v$.
+
+-   Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is represented here by a function $q_i$ [**CM:** $q(i)$?]
+    indicating the district that individual $i$ lives in.
+    The set of all possible districting plans is $\Plans$.
+
+-   For an election $v$, we can tally the votes for each candidate; the candidate with the most votes wins.
+    We denote the winner in district $d$ of districting plan $q$ as $\win_d(v,q)$.
+
+-   explain that $v$ and $q$ are necessarily related: a different districting plan may mean different candidates run.
+
+## Harm
+
+To be harmed by a decision a voter must have a better outcome in the counterfactual world where another decision was made.
+To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
+Then we may define the harm to an individual $i$ as $$
+h_i(v,q,\tilde q) = \ind\{v_i \neq \win_{q(i)}(v, q)\}
+        \ind\{v_i = \win_{\tilde q(i)}(v, \tilde q)\},
+$$that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.
+
+## Multiple Elections or Districting Plans
+
+So far, harm is defined for a single election outcome $v\in \Elecs$ and districting plan $q\in\Plans$.
+But realistically we will need to consider harm averaged over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans, or even over a set of individuals with some shared characteristic.
+
+To allow this, we replace the fixed election and districting plans with random variables $V\in\Elecs$ and $Q,\tilde Q\in\Plans$.
+The randomness in election outcomes comes from variations in factors such as voter preferences, the political environment, candidate quality, and campaign events.
+The randomness in districting plans reflects the constraints and preferences of the plan-drawing process.
+As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V$, $Q$, and $\tilde Q$ are mutually independent, though this is not required for our theoretical analysis.
+We will also consider an individual $I$ taken uniformly at random from $\Voters$, with $I\indep V,Q,\tilde Q$.
+
+Then we may define the harm for a (randomly selected) individual as $H \dfeq h_I(V, Q, \tilde Q)$.
+Thus, for example, the harm for a single individual $i$ becomes $H\mid I=i$; and the electorate's expected harm under districting plan $q$ becomes$$
+\E[H\mid Q=q] = \Pr[V(I)\neq\win_{q(I)}(V, q)
+    \text{ and }V(I)=\win_{\tilde Q(I)}(V, \tilde Q)],
+$$that is, the probability that a randomly selected voter is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
+
+**--------------- CM: this is as far as I've gotten. ---------------**
+
+# Differential Harm and Fairness {#subsec:diff-harm}
 
@@
 But the same concepts and definitions naturally carry over to the multiparty setting.
-
-In the districting plan examples of Figure \ref{fig:mini-harm}, the differential harm of the Democratic gerrymander of Mininois is $-0.519$, the differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
-In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).
 
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^3]
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^1]
 Then a party's vote share in precinct $k$, notated here as $s_k$, is modeled as $$
@@
 
-[^3]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+[^1]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
 
@@
 We sample 1,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
-For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^4]
+For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^2]
 The distribution of these measures is shown in Figure \ref{fig:nj-vars}.
 
-[^4]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature[@katz21].
-
-```{r nj-vars, fig.cap="Distributions of districting measures for simulated New Jersey distriting plans, the Republican gerrymander (red), and the Democratic gerrymander (blue).", fig.width=6.5, fig.height=3.25}
-include_graphics(here("paper/figures/nj_vars.pdf"))
-```
+[^2]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature[@katz21].
 
@@
 ```{=tex}
-\propglobal*
-
-\begin{proof}
-\end{proof}
-
 \propfairness*

### 2022-03-28 paper/paper.Rmd
typos
work

@@
 Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
-Section \ref{sec:empirical} applies the proposed measures to congressional representation in New Jersey and primary elections for the New York City Council.
+Sections \ref{sec:nj} and \ref{sec:boston} apply the proposed measures to partisan gerrymandering of New Jersey congressional districts and racial impacts of district boundaries on Boston City Council primary elections.
 Section \ref{sec:discussion} concludes.
@@
 
-# Individual Representational Harm
+# Individual Representational Harm {#sec:indiv-harm}
 
@@
     In an election, voters either vote for a candidate running in their district, or abstain from voting.
-    We write $v_i$ [**CM: should this be** $v(i)$?]
-    to mean the candidate that individual $i$ votes for in election $v$.
+    We write $v(i)$ to mean the candidate that individual $i$ votes for in election $v$.
 
--   Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is represented here by a function $q_i$ [**CM:** $q(i)$?]
-    indicating the district that individual $i$ lives in.
+-   Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
     The set of all possible districting plans is $\Plans$.
@@
 Then we may define the harm to an individual $i$ as $$
-h_i(v,q,\tilde q) = \ind\{v_i \neq \win_{q(i)}(v, q)\}
-        \ind\{v_i = \win_{\tilde q(i)}(v, \tilde q)\},
+h_i(v,q,\tilde q) = \ind\{v(i) \neq \win_{q(i)}(v, q)\}
+        \ind\{v(i) = \win_{\tilde q(i)}(v, \tilde q)\},
 $$that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.
@@
 
-**--------------- CM: this is as far as I've gotten. ---------------**
-
 # Differential Harm and Fairness {#subsec:diff-harm}
 
-Intuitively, a districting plan is fair if it does not harm the voters of one party more than another, on average.
-This is easily expressed using our notation above; $Q$ is *fair* in comparison to counterfactual $\tilde Q$ if $\E[H\mid V(I) = p]$ is constant across parties $p\in\Parties$.
+A natural extension of studying the average harm to voters is to examine how harm is distributed across certain groups of voters, defined by race, party, or other attributes of interest.
+Let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
+
+Then we say that a plan $Q$ is *fair* in comparison to counterfactual $\tilde Q$ if $\E[H\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.
 
 To quantify deviations from fairness, we introduce of a measure of \textit{differential harm}, defined as $$
-  \Delta^{p_1}_{p_2}H(q) \dfeq \E[H\mid V(I)=p_1, Q=q] - \E[H\mid V(I)=p_2, Q=q],
-$$ for two parties $p_1$ and $p_2$.
-So $\Delta^{p_1}_{p_2}H(q)=0$ for all $p_1,p_2\in\Parties$ if and only if $q$ is fair.
-From here onward we will restrict our attention to the case of two parties---concretely, $\Parties=\{\dem,\rep\}$---and simply write $\Delta H(q)=\Delta^\dem_\rep H(q)$.
-But the same concepts and definitions naturally carry over to the multiparty setting.
+  \Delta_{G_k-G_l}H(q) \dfeq \E[H\mid I\in G_k, Q=q] - \E[H\mid I\in G_l, Q=q],
+$$ for two groups $G_k$ and $G_l$.
+So $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.
+<!--
+From here onward we will restrict our attention to the case of two groups---concretely, $\Groups=\{\dem,\rep\}$---and simply write $\Delta H(q)=\Delta^\dem_\rep H(q)$.
+But the same concepts and definitions naturally carry over to the multigroup setting.
+-->
 
-Despite being derived from a local harm measure, $\Delta H(Q)$ is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
-Yet unlike a simple measure of seats, $\Delta H(Q)$ is shifted to an interpretable scale, where a value of zero corresponds to partisan fairness.
+Despite being derived from a local harm measure, $\Delta H(Q)$ is in fact closely related to the overall number of seats won by each group, as Proposition \ref{prop:fairness} describes below.
+Yet unlike a simple measure of seats, $\Delta H(Q)$ is shifted to an interpretable scale, where a value of zero corresponds to fairness.
 In other words, $\Delta H(Q)$ provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
-This connection further supports the use of total seats as a summary statistic in redistricting simulation methods, as advocated by @deford2020symmetry.
+This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry.
 
@@
 \begin{restatable}{prop}{propfairness} \label{prop:fairness}
-If turnout is equal across districts, i.e., if $\sum_{p\in\Parties} \votes^{(p)}(d,v,q)$ is constant across $d$, $v$, and $q$, then for a specific districting plan $q\in\Plans$ and political party $p\in\Parties$, we have
+Suppose $\Groups=\{G_1,G_2\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
 \begin{align*}
-    \Delta H(q) = \E[\seats(p, V, Q)\mid Q=q] - \E[\seats(p, V, \tilde Q)].
+    \Delta_{G_1-G_2} H(q) = \E[\seats(G_1, V, Q)\mid Q=q] - \E[\seats(G_1, V, \tilde Q)].
 \end{align*}
@@
 
-The measures $U^{(G)}, U^{(L)}, H$ and $\Delta H(q)$ defined above require specification of a probability distribution on elections and districting plans, $\pi$.
+The measures $H$ and $\Delta H(q)$ defined above require specification of a probability distribution on elections and districting plan.
 Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
@@
 Election results present a different set of modeling challenges, given the large number of factors that influence the outcome, and the general unpredictability of future events.
-As far as estimating measures like $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$, we would ideally use a relatively simple model that captures the primary drivers of election outcomes.
+As far as estimating measures like $H$ and $\Delta H(q)$, we would ideally use a relatively simple model that captures the primary drivers of election outcomes.
 In the interest of fairness, the model should not make any biasing assumptions about the future (e.g., that demographic shifts will favor one party, or that past trends in election returns will continue in the same direction).
 
-We propose a *proportionate uniform shift* model, which focuses on swings in the national political environment as the main determinant of election outcomes.
+Here, motivated primarily by computational convenience, we use a *proportionate uniform shift* model, which focuses on swings in the national political environment as the main determinant of election outcomes.
 Each election outcome over the period of interest (generally ten years or less) is modeled as a random deviation from a baseline outcome, and these deviations are assumed to consist of a constant fraction of voters of one party in each precinct switching to the other.
 The baseline itself can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
-
-The proportionate uniform shift model is not mathematically equivalent to the uniform partisan swing model of @butler1951swing, but is near-identical for most precincts and districts.
+The proportionate uniform shift model is nearly identical to the uniform partisan swing model of @butler1951swing for most precincts and districts, but is more computationally convenient and avoids issues of vote shares exceeding 100% or falling below 0%.
 Despite their simplicity, uniform-swing--type models are highly accurate for out-of-sample prediction [@katz2020], and are intuitive and easy to implement.
-
-Proportionate shift, where we assume that a constant fraction of *voters of one party* in each precinct switch to the other party, has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
-First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
-Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
-For measures like $H$ and $F$ which consider voters in each party separately, this helpfully simplifies computations.
-
-The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlikea generalized linear model.
-Specifically, we define $\gamma:[0,1]\to\R$, $$
-\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^1]
-Then a party's vote share in precinct $k$, notated here as $s_k$, is modeled as $$
-\gamma(s_k) = \gamma(\overline s_k) + \eps, \qquad \eps \sim \Norm(0, \sigma^2_{\mathrm{hist}}),
-$$ where $\overline{s}_k$ is the party baseline, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
-For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.
-
-[^1]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+More details about our proportionate uniform shift model may be found in Appendix \ref{app:formulas}.
 
@@
 For ease and efficiency of computation, we stick to the simple proportionate uniform shift model here.
-Appendix \ref{app:formulas} contains the specific formulas used to compute $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
+Appendix \ref{app:formulas} contains the specific formulas used to compute $H$, and $\Delta H(q)$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
 
-# Connections to Partisan Symmetry and Fairness, Standards {#sec:connections}
+# Connections to Partisan Symmetry and Fairness Standards {#sec:connections}
 
@@
 
-## Districting Plan Competitiveness {#sec:competitive}
-
-The notion of district *competitiveness*, while related to the slope of the seats-vote curve around $\half$, is not the same as partisan symmetry.
-However, one commonly used measure of competitiveness, from @tamcho2016, is related to our local utility measure $U^{(L)}$.
-The measure of @tamcho2016, which we will denote by $\mathrm{comp}_\alpha(v,q)$, contains two terms: one, $T_p(v,q)$, measuring the closeness of district vote shares to 50%, and the other, $T_e(v,q)$, measuring the closeness of the overall share of seats to 50%; the relative importance of these two terms is controlled by $\alpha$.
-It is this first term which his closely related to $U^{(L)}$, as stated in Proposition \ref{prop:comp} below.
-
-```{=tex}
-\begin{restatable}{prop}{propcomp} \label{prop:comp}
-If turnout is equal across districts, for an election $v\in\Elecs$ and districting plan $q\in\Plans$ and two political parties, we have
-\begin{align*}
-    \E[U^{(L)}\mid V=v, Q=q] &= \frac{1}{2|D|}\sum_{d\in D}
-        \qty|\frac{\votes^{(p_a)}}{\votes^{(p_b)} + \votes^{(p_b)}}-\half| \\
-    &= \half T_p(v,q) = \frac{\mathrm{comp}_\alpha(v,q)}{1+\alpha T_e(v_q)}.
-\end{align*}
-
-\end{restatable}
-```
-Importantly, $\mathrm{comp}_\alpha(v,q)$ is designed to be *minimized*---smaller values correspond to more competitive districts.
-But, as briefly discussed above, this actually runs counter to the local utility measure, which encourages lopsided districts with as many similarly-minded voters as possible.
-@brunell2010 discusses this point in much more detail, arguing that competitive districts should be avoided where possible.
-Buttressing our choice of local utility function, he analyzes ANES survey data to show that voters approve more highly of their representative and of Congress overall when their representative votes in alignment with their preferences.
-
-# Application: New Jersey {#sec:empirical}
+# Application to Partisan Gerrymandering in New Jersey {#sec:nj}
 
@@
 We sample 1,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
-For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^2]
+For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^1]
 The distribution of these measures is shown in Figure \ref{fig:nj-vars}.
 
-[^2]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature[@katz21].
+[^1]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature[@katz21].
 
@@
 
+# Application to Primary Elections in Boston {#sec:boston}
+
 # Discussion {#sec:discussion}
@@
 ```
+
 # Proofs of Propositions {#app:proofs}
 
-```{=tex}
 \propfairness*
@@
 
-\propcomp*
-
-\begin{proof}
-\end{proof}
-```
-# Computing Utility, Harm, and Fairness from Precinct-level Election Data {#app:formulas}
+# Computing Harm and Fairness from Precinct-level Election Data {#app:formulas}
 
@@
 ```
+Proportionate shift, where we assume that a constant fraction of *voters of one party* in each precinct switch to the other party, has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
+First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
+Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
+For measures like $H$ and $F$ which consider voters in each party separately, this helpfully simplifies computations.
+
+The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlikea generalized linear model.
+Specifically, we define $\gamma:[0,1]\to\R$, $$
+\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^2]
+Then a party's vote share in precinct $k$, notated here as $s_k$, is modeled as $$
+\gamma(s_k) = \gamma(\overline s_k) + \eps, \qquad \eps \sim \Norm(0, \sigma^2_{\mathrm{hist}}),
+$$ where $\overline{s}_k$ is the party baseline, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
+For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.
+
+[^2]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+
 # Districting Measures in New Jersey {#app:nj}

### 2022-03-28 R/extra/04_sim_wi.R

@@ -1,74 +1,0 @@
-wi = read_rds(here("data/WI_cd_final_vtd_20.rds")) %>%
-    redist_map(pop_tol=0.01, ndists=8, adj=.$adj) %>%
-    mutate(across(contains("dv"), ~ coalesce(., 0)),
-           across(contains("rv"), ~ coalesce(., 0)))
-
-if (!file.exists(sim_path <- here("data/wi_sims.rds"))) {
-    N_sim = 1000
-    plans = redist_smc(wi, N_sim, counties=county, pop_temper=0.01, verbose=TRUE)
-    write_rds(plans, sim_path, compress="xz")
-} else {
-    plans = read_rds(sim_path)
-}
-
-statewide = with(wi, sum(ndv)/sum(ndv+nrv))
-prop_seats = round(attr(wi, "ndists") * statewide)
-
-pl = calc_plans_stats(plans, wi, ndv, nrv)
-
-pdf("out/wi_vars.pdf", 9, 9)
-expl_vars(pl, e_dem, u_glb, u_loc, f, f_glb, h, egap, pbias, mean_med, competitive)
-dev.off()
-
-
-d_plot = pl$plan %>%
-    mutate(`terc_Local Utility` = ntile(u_loc, 10),
-           `terc_Global Utility` = ntile(u_glb, 10),
-           terc_Fairness = ntile(f, 10)) %>%
-    mutate(across(starts_with("terc_"), ~ if_else(. %in% c(1L, 10L), ., 5L))) %>%
-    select(draw, n_dem, e_dem, u_loc, u_glb, f, starts_with("terc_")) %>%
-    as_tibble()
-d_plot = number_by(pl$distr, dem) %>%
-    left_join(d_plot, by="draw") %>%
-    pivot_longer(starts_with("terc_"), names_to="qty",
-                 values_to="tercile", names_prefix="terc_")
-
-ggplot(d_plot, aes(as.factor(district), dem, fill=as.factor(tercile))) +
-    facet_grid(qty ~ .) +
-    geom_hline(yintercept=0.5, lty="dashed") +
-    geom_boxplot(size=0.3, outlier.size=0.1) +
-    scale_y_continuous("Democratic two-party share", labels=percent) +
-    labs(x="Districts, ordered by Democratic share",
-         fill="Tercile\nof measure") +
-    scale_fill_wa_d("sound_sunset", which=c(1, 8, 13)) +
[…]

### 2022-03-28 data-raw/boston-data-prep.R

@@ -1,0 +1,146 @@
+library(tidyverse)
+library(sf)
+library(here)
+
+# functions -----
+
+# read from Tabula output
+parse_mayor_data = function(data_dir) {
+    map_dfr(list.files(data_dir), function(fname) {
+        ward_no = as.integer(str_extract(fname, "(?<=-)\\d+(?=.csv)")) + 1L
+        d = read_lines(here(data_dir, fname)) %>%
+            str_replace_all(",{2,}", ",") %>%
+            str_c(collapse="\n") %>%
+            read_csv(show_col_types=F) %>%
+            select(-TOTAL)
+
+        d = d[1:(which(d$CANDIDATES == "VOTES CAST") - 1),]
+
+        d %>%
+            pivot_longer(-CANDIDATES, names_to="precinct", values_to="votes") %>%
+            rename(candidate=CANDIDATES) %>%
+            mutate(precinct = str_c(ward_no, "-", precinct)) %>%
+            select(precinct, candidate, votes)
+    })
+}
+
+parse_name = function(x) {
+    case_when(
+        x %in% c("ALL OTHERS", "NO PREFERENCE") ~ "other",
+        x == "ANNISSA ESSAIBI GEORGE" ~ "essaibi_george",
+        x == "JOSEPH R. BIDEN, JR." ~ "biden",
+        TRUE ~ str_to_lower(word(x, -1))
+    )
+}
+
+# precinct rows, candidate columns
+pivot_name = function(d, prefix="") {
+    d %>%
+        mutate(candidate = parse_name(candidate)) %>%
+        group_by(precinct, candidate) %>%
+        summarize(votes = sum(votes), .groups="drop") %>%
+        arrange(desc(votes)) %>%
+        pivot_wider(names_from = candidate,
+                    values_from = votes) %>%
+        slice(str_order(precinct, numeric=TRUE)) %>%
+        rename_with(function(x) str_c(prefix, x), .cols=-precinct)
+}
+
+
+# read in and parse  ------
+
+d_gen = parse_mayor_data("data-raw/boston_returns/tabula-gen_mayor") %>%
+    pivot_name("mayor_gen_")
+d_prelim = parse_mayor_data("data-raw/boston_returns/tabula-prelim_mayor") %>%
+    pivot_name("mayor_prelim_")
+
[…]

### 2022-03-29 paper/paper.Rmd
section 3 and 4 redrafted

@@
       To whom correspondence should be addressed. 
-      The authors thank participants of the Harvard American Politics Research Workshop for helpful comments and discussion.
+      The authors thank participants of the Harvard American Politics 
+      Research Workshop for helpful comments and discussion.
     department: Department of Statistics
@@
 
--   this section: formalize and generalize the concepts in Section \ref{sec:motivation}.
+This section will formalize and generalize the concepts in Section \ref{sec:motivation}, beginning with some notation.
 
@@
 
--   Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of legislative districts, indexed by $\Distrs$.
+Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of legislative districts, indexed by $\Distrs$.
 
--   Voters vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
+Voters vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
 In an election, voters either vote for a candidate running in their district, or abstain from voting.
@@
 
--   Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
+Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
 The set of all possible districting plans is $\Plans$.
 
--   For an election $v$, we can tally the votes for each candidate; the candidate with the most votes wins.
+For an election $v$, we can tally the votes for each candidate; the candidate with the most votes wins.
 We denote the winner in district $d$ of districting plan $q$ as $\win_d(v,q)$.
 
--   explain that $v$ and $q$ are necessarily related: a different districting plan may mean different candidates run.
+This notation obscures the complex relationships and feedback effects between individuals, elections, and districting plans.
+For example, different districting plans may lead to different candidates running, depending on district partisanship and incumbent residency.
+The assumptions we make on these relationships will be made more explicit in Section \ref{sec:estimating}, where we discuss how to estimate our measures in practice.
 
@@
 
-To be harmed by a decision a voter must have a better outcome in the counterfactual world where another decision was made.
+The key intuition from Section \ref{sec:motivation} was that to be harmed by a decision a voter must have a better outcome in the counterfactual world where another decision was made.
 To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
-Then we may define the harm to an individual $i$ as $$
+Then we may define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as $$
 h_i(v,q,\tilde q) = \ind\{v(i) \neq \win_{q(i)}(v, q)\}
         \ind\{v(i) = \win_{\tilde q(i)}(v, \tilde q)\},
-$$that is, a voter is harmed only if they are not represented by the party of their choice *and* they would have been represented by the party of their choice under the counterfactual plan.
+$$that is, a voter is harmed only if they are not represented by the candidate of their choice and they would have been represented by their chosen candidate under the counterfactual plan.
+
+This definition of harm is limited in two key regards.
+First, it considers only individuals who vote; if a person abstains from voting then their calculated harm is zero.
+A natural extension of this definition would be to separate an individual's candidate preference from their vote choice; the latter would be used in calculating election winners, but the former could provide a more broadly-based harm measure.
+Naturally, this approach requires auxiliary data to implement (such as a large-scale survey), since candidate preferences for nonvoters cannot be inferred from election returns.
+
+A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between candidates he or she did not vote for.
+For example, a voter may consider candidates A and B nearly exchangeable, but may strongly dislike candidate C.
+If the voter ultimately voted for candidate A, this harm measure makes no distinction between candidate B or candidate C winning the voter's district.
+This reflects the simplistic nature of the cost function that is implicit in the harm measure.
+The problem could be remedied by replacing the indicator functions with other cost functions that consider ordinal preferences.
+Ballot-level ranked-choice vote data, or candidate ideology scores, could be used to estimate such a cost function.
+Here, to avoid making further assumptions, we limit ourselves to the indicator-based cost function.
+Since many American elections are between just two candidates, this limitation should not prove too restrictive.
+Section \ref{sec:boston} further discusses the implications of this particular specification in the context of primary elections.
 
@@
 
-So far, harm is defined for a single election outcome $v\in \Elecs$ and districting plan $q\in\Plans$.
-But realistically we will need to consider harm averaged over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans, or even over a set of individuals with some shared characteristic.
+As defined above, $h_i(v,q,\tilde q)$ is defined for a single election and single counterfactual districting plan $\tilde q$, but in general we will be interested in outcomes over multiple elections, and possibly a much larger set of counterfactual plans.
+Ideally we would like to be able to average individual harm over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans, or even over a set of individuals with some shared characteristic.
 
-To allow this, we replace the fixed election and districting plans with random variables $V\in\Elecs$ and $Q,\tilde Q\in\Plans$.
-The randomness in election outcomes comes from variations in factors such as voter preferences, the political environment, candidate quality, and campaign events.
-The randomness in districting plans reflects the constraints and preferences of the plan-drawing process.
-As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V$, $Q$, and $\tilde Q$ are mutually independent, though this is not required for our theoretical analysis.
-We will also consider an individual $I$ taken uniformly at random from $\Voters$, with $I\indep V,Q,\tilde Q$.
+To allow for this, we replace the fixed election and counterfactual districting plans with random variables $V\in\Elecs$ and $\tilde Q\in\Plans$.
+The randomness here is primarily a conceptual device to simplify the notation, but one can also think of it as reflecting uncertainty or variation in voter preferences, the political environment, candidate quality, campaign events, and other factors.
+The randomness in districting plans specifically may be conceptualized as the constraints and preferences of the plan-drawing process (i.e., plans which better respect the preferences are given a higher probability).
+As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V$ and $\tilde Q$ are mutually independent, though this is not required for our theoretical analysis.
+Finally, we will also replace $i$ with an individual $I$ taken uniformly at random from $\Voters$, with $I\indep V,\tilde Q$.
 
-Then we may define the harm for a (randomly selected) individual as $H \dfeq h_I(V, Q, \tilde Q)$.
-Thus, for example, the harm for a single individual $i$ becomes $H\mid I=i$; and the electorate's expected harm under districting plan $q$ becomes$$
-\E[H\mid Q=q] = \Pr[V(I)\neq\win_{q(I)}(V, q)
-    \text{ and }V(I)=\win_{\tilde Q(I)}(V, \tilde Q)],
-$$that is, the probability that a randomly selected voter is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
+Then we may define the harm for a (randomly selected) individual under plan $q$ as $H(q) \dfeq h_I(V, q, \tilde Q)$.
+Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in election $v$ becomes$$
+\E[H(q)\mid V=v] = \Pr[v(I)\neq\win_{q(I)}(v, q)
+    \text{ and }v(I)=\win_{\tilde Q(I)}(v, \tilde Q)],
+$$that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
 
-# Differential Harm and Fairness {#subsec:diff-harm}
+## Differential Harm and Fairness {#subsec:diff-harm}
 
-A natural extension of studying the average harm to voters is to examine how harm is distributed across certain groups of voters, defined by race, party, or other attributes of interest.
-Let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
+Single-member districts are often motivated by the value in representation to specific communities with shared interests; indeed, preserving "communities of interest" is one of the most common statutory and constitutional requirements imposed by states on the plan-drawing process [@ncslcriteria].
+It is therefore natural to consider not just the average harm to voters done by a plan, but also how this harm is distributed across certain groups of voters, defined by race, party, or other attributes of interest.
 
-Then we say that a plan $Q$ is *fair* in comparison to counterfactual $\tilde Q$ if $\E[H\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.
+To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
+We will say that a plan $q$ is *fair* in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.
 
-To quantify deviations from fairness, we introduce of a measure of \textit{differential harm}, defined as $$
-  \Delta_{G_k-G_l}H(q) \dfeq \E[H\mid I\in G_k, Q=q] - \E[H\mid I\in G_l, Q=q],
+Fairness is a strict condition; any real-life plan will never be exactly fair.
+To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
+  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
 $$ for two groups $G_k$ and $G_l$.
-So $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.
+So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
+Using our definition of fairness, $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.
 <!--
@@
 
-Despite being derived from a local harm measure, $\Delta H(Q)$ is in fact closely related to the overall number of seats won by each group, as Proposition \ref{prop:fairness} describes below.
-Yet unlike a simple measure of seats, $\Delta H(Q)$ is shifted to an interpretable scale, where a value of zero corresponds to fairness.
-In other words, $\Delta H(Q)$ provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
-This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry.
+It is this measure of differential harm that we propose as a unified framework for evaluating a districting plan across a variety of dimensions.
+It is derived from an individual-level measure of representation and it accounts for the space of counterfactual possibilities.
+Importantly it separates *vote choice* from *group membership*, allowing one to discuss how partisan election outcomes under a specific districting plan create harm for specific groups or their interactions, simply by conditioning $H(q)$ on a set of individuals.
 
-```{=tex}
-\begin{restatable}{prop}{propfairness} \label{prop:fairness}
-Suppose $\Groups=\{G_1,G_2\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
-\begin{align*}
-    \Delta_{G_1-G_2} H(q) = \E[\seats(G_1, V, Q)\mid Q=q] - \E[\seats(G_1, V, \tilde Q)].
-\end{align*}
+Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe next.
 
-\end{restatable}
-```
 # Estimating the Harm Measures {#sec:estimating}
 
-The measures $H$ and $\Delta H(q)$ defined above require specification of a probability distribution on elections and districting plan.
+The measures $H(q)$ and $\Delta H(q)$ defined above require specification of a probability distribution on elections and districting plans.
 Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
@@
 While not strictly necessary, modeling of $V$ and $\tilde Q$ independently will simplify the situation considerably.
-Practically, this translates to the assumption that individual voters would vote for the same political party no matter the districting plan.
+Practically, this translates to the assumption that individual voters would make the same candidate choice no matter the districting plan.
 To the extent that voters reward or punish an incumbent party for the districting plan they adopt, or that an adopted districting plan affects which candidates run in a district, this assumption is violated.
-But as our suggested model for election outcomes will not account for candidate factors or policy preferences, the marginal effect on our conclusions will be minimal.
+But as our suggested model for election outcomes $V$ will be quite coarse, not accounting for candidate factors or policy preferences, the marginal effect of this independence assumptions on our conclusions will be minimal.
 
@@
 These maximum-entropy distributions put as few conditions as possible on the districting sampling process, as long as a set of constraints imposed by the researcher are met.
-If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan considerations, to use in estimating $H$ and $\Delta H(q)$.
+If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
 
@@
 Election results present a different set of modeling challenges, given the large number of factors that influence the outcome, and the general unpredictability of future events.
-As far as estimating measures like $H$ and $\Delta H(q)$, we would ideally use a relatively simple model that captures the primary drivers of election outcomes.
+As far as estimating measures like $H(q)$ and $\Delta H(q)$, we would ideally use a relatively simple model that captures the primary drivers of election outcomes.
 In the interest of fairness, the model should not make any biasing assumptions about the future (e.g., that demographic shifts will favor one party, or that past trends in election returns will continue in the same direction).
@@
 For ease and efficiency of computation, we stick to the simple proportionate uniform shift model here.
-Appendix \ref{app:formulas} contains the specific formulas used to compute $H$, and $\Delta H(q)$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
 
-# Connections to Partisan Symmetry and Fairness Standards {#sec:connections}
+## Calculating Harm and Differential Harm
+
+With a model for election outcomes and districting plans in hand, calculation of $H(q)$ and $\Delta H(q)$ is straightforward.
+For each sample $\tilde q$ of a counterfactual districting plan, we use the election model to come up with a probability that voters in each precinct are represented by the candidate of their choice under $\tilde q$ but not under $q$.
+This probability is a simple difference in CDF functions under our proportionate uniform shift model.
+When the groups used in $\Delta H(q)$ vary within precincts, we must repeat these computations for each group.
+Then by summing across precincts and/or groups and dividing by the appropriate totals, we arrive at values for $H(q)$ and $\Delta H(q)$.
+Appendix \ref{app:formulas} contains the specific formulas used to compute $H(q)$, and $\Delta H(q)$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
+
+# Connections to Existing Partisan and Racial Gerrymandering Metrics {#sec:connections}
 
@@
 
+## Simulation Methods
+
+Despite being derived from individual harm, differential partisan harm is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
+Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale, where a value of zero corresponds to fairness.
+In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
+This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry.
+
+```{=tex}
+\begin{restatable}{prop}{propfairness} \label{prop:fairness}
+Suppose $\Groups=\{\dem,\rep\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
+\begin{align*}
+    \Delta_{\rep-\dem} H(q) = \E[\seats(\dem, V, \tilde Q)] - \E[\seats(\dem, V, q)].
+\end{align*}
+\textbf{CM note: define $\seats$ or replace with sum.}
+
+\end{restatable}
+```
 # Application to Partisan Gerrymandering in New Jersey {#sec:nj}
@@
 For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^1]
-The distribution of these measures is shown in Figure \ref{fig:nj-vars}.
+The distribution of these measures is shown in Figure \ref{fig:nj-pairs}.
 
@@
 
-```{r nj-district-shares, fig.cap="Distribution of district vote shares, with districts ordrered by baseline Democratic two-part vote share.
-Displayed on top are the district vote shares for six example plans, each extremal for some combination of districting measures.", fig.width=6.5, fig.height=3.9}
-include_graphics(here("paper/figures/nj_district_shares.pdf"))
+```{r nj-pairs, fig.cap="Scatterplots for all pairs of measures.
+Color indicates the number of Democratic seats won (ranging from 7 to 10).
+Below the diagonal, numbers indicate the absolute value of the correlation between variables.", fig.width=7, fig.height=7}
+include_graphics(here("paper/figures/nj_pairs.pdf"))
 ```
-
-This is manifested in the district-level vote shares shown in Figure \ref{fig:nj-district-shares}.
-Within each districting plan, we number districts in order of their Democratic vote share, with district 1 the least Democratic.
-The figure shows these vote shares for six selected districting plans, and compares them to the distribution of shares for the entire sample.
-The selected plans themselves are shown in Appendix \ref{app:nj}.
-
-The fairest, best overall utility, and high global, low local utility plans have 8 Democratic and 4 Republican seats at baseline, but the best-utility plan has more extreme vote shares, i.e., less competitive districts, while the high global, low local utility plan has vote shares closer to 50%, with the fairest plan falling somewhere in the middle.
-
-In contrast, the low global, high local utility plan has 9 Democratic seats, 1 Republican seat, and 2 highly competitive seats at baseline.
-While generally local utility encourages proportionality and noncompetitive districts, conditional on low global utility (poor proportionality) the local utility measure encourages a Democratic gerrymander which satisfies the state's majority of Democratic voters, at the expense of the minority of Republican voters.
-
-The fair but low local utility plan has a similar vote share pattern to the fair plan, but overall with more competitive districts, which are generally penalized by the local utility measure.
-Conversely, the unfair but high local utility plan has 9 Democratic seats, 2 Republican seats, and 1 highly competitive seat at baseline---an unfair Democratic gerrymander---while keeping district vote shares relatively noncompetitive.
 
@@
 ```
-
 # Proofs of Propositions {#app:proofs}
 
-\propfairness*
+\propfairness
 
+```{=tex}
 \begin{proof}
 \end{proof}
-
+```
 # Computing Harm and Fairness from Precinct-level Election Data {#app:formulas}
@@
 
-1.  Global utility is discrete, so we can compute the range of shifts that produce each number of seats, and the average vote share for those ranges.
-    Combine these
-
-2.  To be computationally efficient, local utility requires linearizing the vote share curve.
-
-3.  For voter fairness, PUS means that relative share of party's voters in each precinct remains the same.
-    Need only to apply a kernel function (composition of Laplace ICDF + model shift CDF)
+For voter fairness, PUS means that relative share of party's voters in each precinct remains the same. Need only to apply a kernel function (composition of Laplace ICDF + model shift CDF)
 
@@
 ```
-Proportionate shift, where we assume that a constant fraction of *voters of one party* in each precinct switch to the other party, has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
+## Proportionate Uniform Shift Details
+
+The proportionate uniform shift model assumes that a constant fraction of *voters of one party* in each precinct switch to the other party.
+This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
 First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
 Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
-For measures like $H$ and $F$ which consider voters in each party separately, this helpfully simplifies computations.
+For $\Delta H(q)$, which considers voters in each party separately, this helpfully simplifies computations.
 
-The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlikea generalized linear model.
+The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlike a generalized linear model.
 Specifically, we define $\gamma:[0,1]\to\R$, $$
@@
 $$ which is the inverse CDF of a Laplace distribution with rate 2.[^2]
-Then a party's vote share in precinct $k$, notated here as $s_k$, is modeled as $$
-\gamma(s_k) = \gamma(\overline s_k) + \eps, \qquad \eps \sim \Norm(0, \sigma^2_{\mathrm{hist}}),
+Then a party's vote share in precinct $k$ and election $v$, notated here as $s_{kv}$, is modeled as $$
+\gamma(s_{kv}) = \gamma(\overline s_k) + \eps_v, \qquad \eps_v \iid \Norm(0, \sigma^2_{\mathrm{hist}}),
 $$ where $\overline{s}_k$ is the party baseline, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
@@
 # Districting Measures in New Jersey {#app:nj}
-
-```{r nj-pairs, fig.cap="Scatterplots for all pairs of measures.
-Color indicates the number of Democratic seats won (ranging from 7 to 10).
-Below the diagonal, numbers indicate the absolute value of the correlation between variables.", fig.width=7, fig.height=7}
-include_graphics(here("paper/figures/nj_pairs.pdf"))
-```
-
-```{r nj-ex-maps, fig.cap="Example districting plans selected according to their scores on $\\Delta H(q)$, $U^{(G)}$, and $U^{(L)}$.
-District shares for these example plans are shown in Figure \\ref{fig:nj-district-shares}.", fig.width=7, fig.height=7}
-include_graphics(here("paper/figures/nj_ex_maps.pdf"))
-```
 

### 2022-03-29 R/00_setup_utils.R

@@ -33,6 +33,10 @@ flog = function(x) if_else(x <= 0, -1, suppressWarnings(log(x)))
 plapl = function(x) 0.5 * (1 + sign(x) * (1 - exp(-2*abs(x))))
 qlapl = function(p) -0.5 * sign(p - 0.5) * log(1 - 2*abs(p - 0.5))
 k_step = function(x) x > 0.5
+pos_part = function(x) {
+    x[x < 0] = 0
+    x
+}
 # historical congressional national shifts
 d_hist = local({
     dem_hist = c(0.575, 0.559, 0.537, 0.505, 0.552, 0.521, 0.543, 0.533, 0.521, 0.501, 0.447, 0.482, 0.473, 0.471, 0.452, 0.468, 0.523, 0.532, 0.449, 0.488, 0.455, 0.48, 0.534)

### 2022-03-29 R/toy/00_toy_setup.R

@@ -72,8 +72,9 @@ plot_state = function(map, people, plan=NULL, qty=NULL, ppl_size=1.75, ...) {
     p = redist.plot.map(map, fill=qty, ...) +
         scale_fill_party_c(name="Democratic\nshare") +
         geom_sf(data=map, size=0.8, fill=NA, color="white") +
-        geom_sf(aes(color=dem), data=people, size=ppl_size) +
+        geom_sf(aes(color=dem, shape=dem), data=people, size=ppl_size) +
         scale_color_party_d(guide="none") +
+        scale_shape_manual(values=c(15, 19), guide="none") +
         theme_repr_map() +
         theme(plot.title = element_text(hjust = 0.5))
     if (!is.null(plan)) {

### 2022-03-29 paper/paper.Rmd
s2 updates

@@
 
--   40% GOP but no GOP districts.
-    Is this unfair?
+First, we consider the case of Minichusetts.
+Modeled on Massachusetts, Minichusetts contains a remarkable lack of partisan segregation.
+This high level of integration, combined with Democrats receiving the majority of votes, leads to a Republican inability to win any seats under recent historical voting patterns \[@duch:18\].
+For existing measures, this creates a conflict: Republicans win about 40% of the vote, but zero of the nine seats.
+The Efficiency Gap would correctly identify that the enacted map favors Democrats, whereas both Partisan Bias and Mean-Median indicate a pro-Republican bias (ref planscore).
 
--   E-gap would say so.
-    Partisan symmetry: ?.
-
--   But impossible to draw a GOP district
-
--   Without a remedy (counterfactual), there can be no harm
+This motivates the need for a counterfactual.
+How can a plan be biased towards or against a party in a case where the party is unable to do worse or to do better than the plan?
+By relying on counterfactual plans, we can then identify the cases where individuals are not represented by their preferred candidate or party.
+In places like Minichusetts, as no plan can be drawn which improves the partisan representation of Republican voters, then there can be no harm.
 
@@
 
+Not all cases are as simple as those where one party wins all the seats.
+Mininois also has a Democratic majority, but the voters are no longer integrated.
+In a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions \[@chen:rodd:13\] \[@enos:brown:21\].
+This allows for significantly greater variation in the expected partisanship of the districts drawn.
+
 -   show a fair map and a gerrymandered map
 
--   zoom into 1 or 2 precincts & calculate individual harm of D & R voters in those precincts
-
--   show map of differential harm (with
+A Democratic gerrymander in Mininois effectuates Democratic control by splitting its city into with the blades of a pinwheel reaching out into the more Republican, rural remainder of the state..
+If we consider the two plans in Figure TODO, we see that the Democratic gerrymander would likely elect zero Republicans and three Democrats, whereas the fair map would likely elect two Democrats and one Republican.
+Consider the precincts \[4, 1\] (maybe highlighted in a color).
+\[4,1\] is placed in a Republican district in the fair map, while it is in a Democratic district in the gerrymandered map.
+This precinct has 2 Republicans who are not represented by their co-partisan in the gerrymandered plan, but would have been represented by a co-partisan in the fair plan.
+Conversely, the 1 Democrat is represented by their co-partisan in the gerrymandered plan, but not the fair plan.
+Each of these three voters can show a counterfactual where they would be better represented and thus have a claim to harm.
+Yet, if we instead consider any voter who lives in the (other color) \[1,1\] precinct, those voters must always be in a Democratic district given the distribution of voters.
+While the Republicans in that precinct are not represented by their co-partisan, they never would be.
+Those Republicans can't be individually harmed by redistricting.
 
 ## Minissouri: differential harm
+
+We can also consider a less drastic sorting, where medium cities favor Democrats, but no longer dominate the state politics, such as in Minissouri.
+This allows for additional variability, where either party can gerrymander to draw more seats for their party than a neutral process might allow.
 
@@
 
--   end with transition: there are natural avenues for generalization
+While we've begun with a two group case, Republicans and Democrats, this can be further extended.
+Perhaps we are interested in a third party or the interaction between White and Black with Democrats and Republicans.
+This framework allows for us to index by these groups and simply break down the interactions as components of the sum of the harms.
 
@@
 In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).
+
+| State      | Plan                   | Total Harm | Differential Harm | Efficiency Gap | Partisan Bias | Mean-Median | Declination |
+|------------|------------------------|------------|-------------------|----------------|---------------|-------------|-------------|
+| Mininois   | "Fair"                 |            |                   |                |               |             |             |
+| Mininois   | Democratic Gerrymander |            | -0.519            |                |               |             |             |
+| Minissouri | Fair                   |            |                   |                |               |             |             |
+| Minissouri | Democratic Gerrymander |            | -0.415            |                |               |             |             |
+| Minissouri | Republican Gerrymander |            | 0.450             |                |               |             |             |
 

### 2022-03-29 paper/paper.Rmd
work

--- 2022-03-29 paper/paper.Rmd
@@
 This addresses one inherent shortcoming of the efficiency gap, illustrated best by the state of Massachusetts.
-According to the efficiency gap, Massachussetts' congressional districting plan is a Democratic gerrymander ($EG=-0.14$).
+According to the efficiency gap, Massachusetts' congressional districting plan is a Democratic gerrymander ($EG=-0.14$).
 But it is impossible to draw a single Republican congressional district in the state.
@@
 ```
-# Computing Harm and Fairness from Precinct-level Election Data {#app:formulas}
+# Computing Harm Measures from Precinct-level Election Data {#app:formulas}
 

### 2022-03-29 paper/paper.Rmd
get compile working

@@
 Modeled on Massachusetts, Minichusetts contains a remarkable lack of partisan segregation.
-This high level of integration, combined with Democrats receiving the majority of votes, leads to a Republican inability to win any seats under recent historical voting patterns \[@duch:18\].
+This high level of integration, combined with Democrats receiving the majority of votes, leads to a Republican inability to win any seats under recent historical voting patterns [@duch:18].
 For existing measures, this creates a conflict: Republicans win about 40% of the vote, but zero of the nine seats.
@@
 Mininois also has a Democratic majority, but the voters are no longer integrated.
-In a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions \[@chen:rodd:13\] \[@enos:brown:21\].
+In a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions [@chen:rodd:13; @enos:brown:21].
 This allows for significantly greater variation in the expected partisanship of the districts drawn.
@@
 If we consider the two plans in Figure TODO, we see that the Democratic gerrymander would likely elect zero Republicans and three Democrats, whereas the fair map would likely elect two Democrats and one Republican.
-Consider the precincts \[4, 1\] (maybe highlighted in a color).
-\[4,1\] is placed in a Republican district in the fair map, while it is in a Democratic district in the gerrymandered map.
+Consider the precincts $(4, 1)$ (maybe highlighted in a color).
+$(4,1)$ is placed in a Republican district in the fair map, while it is in a Democratic district in the gerrymandered map.
 This precinct has 2 Republicans who are not represented by their co-partisan in the gerrymandered plan, but would have been represented by a co-partisan in the fair plan.
@@
 Each of these three voters can show a counterfactual where they would be better represented and thus have a claim to harm.
-Yet, if we instead consider any voter who lives in the (other color) \[1,1\] precinct, those voters must always be in a Democratic district given the distribution of voters.
+Yet, if we instead consider any voter who lives in the (other color) $(1, 1)$ precinct, those voters must always be in a Democratic district given the distribution of voters.
 While the Republicans in that precinct are not represented by their co-partisan, they never would be.
@@
 
-Figure \ref{fig:mini-harm} illustrates this measure for a set of hypothetical gerrymanders, comparing to the set of optimal plans from Figure \ref{fig:mini-util-opt}.
+Figure \ref{fig:mini-harm} illustrates this measure for a set of hypothetical gerrymanders, comparing to the set of optimal plans from Figure .
 As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.
@@
 In the districting plan examples of Figure \ref{fig:mini-harm}, the differential harm of the Democratic gerrymander of Mininois is $-0.519$, the differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
-In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).
-
-| State      | Plan                   | Total Harm | Differential Harm | Efficiency Gap | Partisan Bias | Mean-Median | Declination |
-|------------|------------------------|------------|-------------------|----------------|---------------|-------------|-------------|
-| Mininois   | "Fair"                 |            |                   |                |               |             |             |
-| Mininois   | Democratic Gerrymander |            | -0.519            |                |               |             |             |
-| Minissouri | Fair                   |            |                   |                |               |             |             |
-| Minissouri | Democratic Gerrymander |            | -0.415            |                |               |             |             |
-| Minissouri | Republican Gerrymander |            | 0.450             |                |               |             |             |
+In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure.
 
@@
 Using our definition of fairness, $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.
-<!--
-From here onward we will restrict our attention to the case of two groups---concretely, $\Groups=\{\dem,\rep\}$---and simply write $\Delta H(q)=\Delta^\dem_\rep H(q)$.
-But the same concepts and definitions naturally carry over to the multigroup setting.
--->
 
@@
 For ease and efficiency of computation, we stick to the simple proportionate uniform shift model here.
+In Section \ref{sec:boston}, however, we consider a more complicated model for primary elections.
 
@@
 ```{=tex}
-\begin{restatable}{prop}{propfairness} \label{prop:fairness}
+\begin{proprep} \label{prop:fairness}
 Suppose $\Groups=\{\dem,\rep\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
@@
 \textbf{CM note: define $\seats$ or replace with sum.}
-
-\end{restatable}
+\end{proprep}
+\begin{proof}
+\end{proof}
 ```
@@
 ```
-# Proofs of Propositions {#app:proofs}
 
-\propfairness
-
-```{=tex}
-\begin{proof}
-\end{proof}
-```
 # Computing Harm Measures from Precinct-level Election Data {#app:formulas}

### 2022-03-29 paper/paper.Rmd
ex figures

@@
 
-To illustrate the concepts and measures we introduce, we will make use of three running examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichussets.
+To illustrate the concepts and measures we introduce, we will make use of three examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichussets.
 Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
@@
 
--   show a fair map and a gerrymandered map
+```{r harm-mininois, fig.cap="Two potential districting plans for Mininois: a Democratic gerrymander (a) and a close-to-proportional plan (b).
+Panel (c) shows the total harm to voters in each precinct done by (a), using (b) as a counterfactual."}
+include_graphics(here("paper/figures/harm_mininois.pdf"))
+```
 
@@
 
-```{r mini-harm, fig.cap="Three potential districting plans for Mininois and Minissouris, showing district partisan lean (top row) and total voter harm versus the counterfactuals in Figure 2(a)-(b) (bottom row; darker shading indicates more harm)."}
-include_graphics(here("paper/figures/mini_harm.pdf"))
+```{r diffharm-minissouri, fig.cap="Three potential districting plans for Minissouris: a Democratic gerrymander (a), a close-to-proportional plan (b), and a Republican gerrymander (c).
+Panels (d) and (f) show the total harm to voters in each precinct done by their corresponding gerrymander, using (b) as a counterfactual.
+Panel (e) summarizes this harm by party and across precincts, showing the average harm done to voters of each party by each gerrymander."}
+include_graphics(here("paper/figures/diffharm_minissouri.pdf"))
 ```

### 2022-03-29 R/toy/01_harm_mininois.R

@@ -12,10 +12,8 @@ pl_gerry = c(2, 2, 2, 3, 3, 1,
              2, 3, 3, 3, 1, 1,
              1, 1, 1, 1, 1, 1)
 
-dem_gerry = with(mininois, tapply(dem, pl_gerry, sum)/tapply(pop, pl_gerry, sum))[pl_gerry]
 dem_fair = with(mininois, tapply(dem, pl_fair, sum)/tapply(pop, pl_fair, sum))[pl_fair]
-dem_harm = weighted.mean((dem_gerry<0.5)*(dem_fair>0.5), mininois$dem)
-rep_harm = weighted.mean((dem_gerry>0.5)*(dem_fair<0.5), mininois$rep)
+dem_gerry = with(mininois, tapply(dem, pl_gerry, sum)/tapply(pop, pl_gerry, sum))[pl_gerry]
 
 coords_fair = list(x=c(1/4, 1/6, 2/3),
                    y=c(1.04, -0.04, -0.04))
@@ -46,7 +44,7 @@ for (i in 1:3) {
 p3 = plot_state(mininois, mininois_people,
            pl_gerry, mininois$rep*round(1 - dem_fair)) +
     scale_fill_gradient("Total\nharm", low="#e5e6e5", high="#666666") +
-    labs(title="(c), Voter harm, (a) vs. (b)") +
+    labs(title="(c) Voter harm, (a) vs. (b)") +
     theme(legend.key.width=unit(0.2, "cm"))
 })
 

### 2022-03-29 paper/paper.Rmd
clarify nonconstant group IDs

@@
 To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
-We will say that a plan $q$ is *fair* in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.
+In general we will not require the group identities to be fixed with respect to elections.[^1]
+This is because some group identities, like partisan preference, may themselves change election-to-election.
+
+[^1]: Formally, we have random indicator variables $\ind\{i\in G_k\}$ for all $i$ and $k$, subject to the constraint that $\sum_k \ind\{i\in G_k\}=1$ for all $i$.
+    These indicator variables may be correlated with $V$ and/or $\tilde Q$, in general.
+
+Once we have defined a set of groups, we say that a plan $q$ is *fair* in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.
 
@@
 In competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
-But in more lopsided states, $\beta(\half)$ can give unintuitive results, which, if not interpreted carefully, can lead practictioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.
+But in more lopsided states, $\beta(\half)$ can give unintuitive results, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.
 
@@
 We sample 1,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
-For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^1]
+For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^2]
 The distribution of these measures is shown in Figure \ref{fig:nj-pairs}.
 
-[^1]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature[@katz21].
+[^2]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature [@katz21].
 
@@
 ```
-
 # Computing Harm Measures from Precinct-level Election Data {#app:formulas}
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^2]
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^3]
 Then a party's vote share in precinct $k$ and election $v$, notated here as $s_{kv}$, is modeled as $$
@@
 
-[^2]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+[^3]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
 

### 2022-04-05 paper/paper.Rmd
sec 6 + data

@@
 Then by summing across precincts and/or groups and dividing by the appropriate totals, we arrive at values for $H(q)$ and $\Delta H(q)$.
-Appendix \ref{app:formulas} contains the specific formulas used to compute $H(q)$, and $\Delta H(q)$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
+Appendix \ref{app:formulas} contains more details, including the specific formulas used to compute $H(q)$, and $\Delta H(q)$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
 
@@
 
-```{r nj-maps, fig.cap="Baseline vote patterns, and two hypothetical gerrymanders. Census-designated places are shaded in gray and county lines in white.", fig.width=6.5, fig.height=3.5}
+```{r nj-maps, fig.cap="Baseline vote patterns, two hypothetical gerrymanders, and one randomly sampled plan.
+Census-designated places are shaded in gray and county lines in white.", fig.width=6.5, fig.height=3}
 include_graphics(here("paper/figures/nj_maps.pdf"))
@@
 We sample 1,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
-For each of these sampled plans and the two gerrymanders, we calculate the utility, harm, and fairness measures, as well as the expected number of Democratic seats, the efficiency gap, partisan bias, and the mean-median score.[^2]
-The distribution of these measures is shown in Figure \ref{fig:nj-pairs}.
+One such plan is shown in Figure \ref{fig:nj-maps}(d).
+For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm,[^2] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
+[^3] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
 
-[^2]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature [@katz21].
+[^2]: As in section \ref{sec:connections}, this is $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
 
-The efficiency gap and $\Delta H(q)$ both correctly classify the gerrymandered plans as such, and with the correct sign.
-Both gerrymanders also have a higher average harm than any of the simulated plans.
-In contrast, under the partisan bias measure the Republican gerrymander is completely unbiased, and according to the mean-median measure, both gerrymanders strongly favor Democrats.
-
-For the utility measures, the Democratic plan is a low-utility outlier.
-But the Republican plan is near the median for both global and local utility.
-How to square this with the finding that the Republican plan is an unfair gerrymander which harms more voters than expected under a neutral process?
-As the distribution of expected Democratic seats shows, Democrats are structurally favored in New Jersey, able to win a larger fraction of seats than their statewide vote share under a neutral process.
-Plans which actively counteract this to try to draw proportional districts, like the Republican gerrymander, will necessarily harm Democratic voters versus the neutral counterfactual.
-But in doing so they will create a more proportional plan that will tend to pack voters of each party into more homogeneous districts, increasing both global and local utility.
+[^3]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature [@katz21].
 
 ```{r nj-pairs, fig.cap="Scatterplots for all pairs of measures.
-Color indicates the number of Democratic seats won (ranging from 7 to 10).
+Point lightness indicates the number of Democratic seats won (ranging from 7 (dark) to 10 (light)).
+The Democratic and Republican gerrymanders are shown with the blue and red lines and crosses.
+The dashed line indicates a value of zero.
 Below the diagonal, numbers indicate the absolute value of the correlation between variables.", fig.width=7, fig.height=7}
@@
 ```
+
+The expected number of Democratic seats and the differential partisan harm measures identify each gerrymanders as an outlier in the correct direction (two-sided $p=0.0002$), and as expected from Proposition \ref{prop:fairness}, these two measures are highly correlated.
+The efficiency gap and declination measures also correctly identify the gerrymanders ($p=0.0002$), though they are much more discrete, and some nonlinearity is evident for declination.
+In contrast, the partisan bias and mean-median measures incorrectly classify the Republican gerrymander as a fair plan and a Democratic-leaning outlier, respectively ($p=0.099$ and $p=0.0022$).
+These measures are weakly correlated with the expected number of Democratic seats, and show substantial variation even for fixed values of total seats won at baseline.
+There is also disagreement between the measures about the natural bias of districting plans in New Jersey: the median sampled plan has a partisan bias of $0.047$, indicating a natural Republican bias, but the median mean-median score is just $0.0069$, and the median differential partisan harm is $-0.13$, indicating a natural Democratic bias.
+Which is correct?
+Democrats win 8.47 seats on average across the simulations, which is more than the 6.79 seats that would be expected under a completely proportional outcome.
+
+Both gerrymanders excessively harm the average voter, with the Democratic plan doing so at a moderately higher rate ($p=0.0002$ for the Democratic plan, $p=0.0006$ for the Republican plan).
+Notably, the plans which minimize total harm are also those which have near-zero differential partisan harm.
+Thus, at least in New Jersey, overall harm reduction is consistent with a fair map on a partisan basis.
+Minimizing bias with other measures, however, could lead one astray: plans which score an efficiency gap of zero can have a large average harm, which may not be distributed fairly between parties.
+Plans with zero partisan bias or mean-median score actually have higher-than-average harm and tend to differentially harm Republicans.
 
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^3]
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^4]
 Then a party's vote share in precinct $k$ and election $v$, notated here as $s_{kv}$, is modeled as $$
@@
 
-[^3]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+[^4]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
 

### 2022-04-05 R/00_setup_utils.R

@@ -100,32 +100,49 @@ plot_cds = function(map, pl, county, abbr, qty="ndv") {
         theme_void()
 }
 
-expl_vars = function(pl, labels, ...) {
+expl_vars = function(pl, labels, refs=character(0), lines_off_diag=FALSE, ...) {
+    n_ref = length(refs)
+    idx = if (n_ref == 0L) seq_len(nrow(pl)) else -seq_len(n_ref)
+
     panel.hist <- function(x, ...) {
         usr <- par("usr"); on.exit(par(usr))
         par(usr = c(usr[1:2], 0, 1.5) )
-        h <- hist(x, plot = FALSE)
+        h <- hist(x[idx], breaks=24, plot=FALSE)
         breaks <- h$breaks; nB <- length(breaks)
         y <- h$counts; y <- y/max(y)
-        rect(breaks[-nB], 0, breaks[-1], y, family="Times", col="gray")
+        rect(breaks[-nB], 0, breaks[-1], y, family="Times",
+             col="#888888", border="#88888800", lwd=0)
+        for (i in seq_len(n_ref)) {
+            abline(v=x[i], col=refs[i], lwd=3.0)
+        }
+    }
+    panel.text <- function(x, y, labels, cex, font, ...) {
+        text(x, y, labels=labels, family="Times", font=2)
     }
     panel.cor <- function(x, y, digits = 2, prefix = "", ...) {
         usr <- par("usr"); on.exit(par(usr))
         par(usr = c(0, 1, 0, 1))
-        r <- abs(cor(x, y, use="complete.obs"))
+        r <- abs(cor(x[idx], y[idx], use="complete.obs"))
         txt <- format(c(r, 0.123456789), digits = digits)[1]
         txt <- paste0(prefix, txt)
         text(0.5, 0.5, txt, cex=2.5*sqrt(r), family="Times", offset=0, adj=c(0.5, 0.5))#cex.cor)
     }
     panel.points <- function(x, y, ...) {
-        points(x, y, family="Times", ...)
+        points(x[idx], y[idx], family="Times", ...)
         abline(h=0, v=0, lty="dashed")
+        for (i in seq_len(n_ref)) {
+            points(x[i], y[i], col=refs[i], pch=3, cex=1.2)
+            if (isTRUE(lines_off_diag))
[…]

### 2022-04-05 R/01_metrics.R

@@ -75,7 +75,8 @@ calc_plans_stats = function(plans, map, dem, gop, ker=k_t()) {
         mutate(dev = plan_parity(map),
                comp = distr_compactness(map),
                dem = group_frac(map, dvote, dvote+rvote),
-               egap = partisan_metrics(map, "EffGap", rvote, dvote))
+               decl = part_decl(., map, dvote, rvote),
+               egap = part_egap(., map, rvote, dvote))
 
     m_dem = district_group(plans, dem)
 
@@ -86,11 +87,11 @@ calc_plans_stats = function(plans, map, dem, gop, ker=k_t()) {
                   pbias = mean(ker(pos_part(pmin(dem - (statewide - 0.5), 1)))) - 0.5,
                   pbias_sw = e_dem/ndists - statewide,
                   mean_med = median(dem) - mean(dem),
-                  across(c(dev, comp, competitive:pbias), ~ .[1])) %>%
+                  across(c(dev, comp, decl, egap), ~ .[1])) %>%
         mutate(h_dem = harm(map, dvote, m_dem, idx_2=idx_2, kernel=ker),
                h_rep = harm(map, rvote, m_dem, idx_2=idx_2, kernel=ker, invert=TRUE),
                dh = h_dem - h_rep,
-               h = total(h_dem, h_rep, map, dvote, rvote))
+               h = h_dem*statewide + h_rep*(1-statewide))
 
     list(distr=select(plans, draw:total_pop, dem), plan=pl_sum, mat=m_dem)
 }

### 2022-04-05 R/03_sim_nj.R

@@ -25,127 +25,63 @@ prop_seats = round(attr(nj, "ndists") * statewide)
 pl = calc_plans_stats(plans, nj, ndv, nrv)
 
 # Initial plots ----
+map_scale = scale_fill_party_c(name="Democratic\nshare", limits=c(0.19, 0.81))
 p1 = ggplot(nj, aes(fill=ndv/(ndv+nrv))) +
     geom_sf(size=0, color="#00000000") +
-    scale_fill_party_c(limits=c(0.15, 0.85)) +
+    map_scale +
     labs(title="(a) Partisan patterns") +
     theme_repr_map() +
     theme(plot.title = element_text(hjust = 0.5))
 p2 = plot_cds(nj, as.matrix(plans)[,"dem_gerry"], county, "NJ") +
-    labs(title="(b) Democratic gerrymander") +
-    scale_fill_party_c(limits=c(0.15, 0.85)) +
+    labs(title="(b) Democratic\ngerrymander") +
+    map_scale +
     theme_repr_map() +
     theme(plot.title = element_text(hjust = 0.5))
 p3 = plot_cds(nj, as.matrix(plans)[,"rep_gerry"], county, "NJ") +
-    labs(title="(c) Republican gerrymander") +
-    scale_fill_party_c(limits=c(0.15, 0.85)) +
+    labs(title="(c) Republican\ngerrymander") +
+    map_scale +
     theme_repr_map() +
     theme(plot.title = element_text(hjust = 0.5))
-p = p1 + p2 + p3 + plot_layout(guides="collect")
+p4 = plot_cds(nj, as.matrix(plans)[,425], county, "NJ") +
+    labs(title="(d) Sample plan") +
+    map_scale +
+    theme_repr_map() +
+    theme(plot.title = element_text(hjust = 0.5))
+p = p1 + p2 + p3 + p4 + plot_layout(guides="collect", nrow=1)
 if (!file.exists(path <- here("paper/figures/nj_maps.pdf")))
-    ggsave(path, plot=p, width=8, height=4.5)
+    ggsave(path, plot=p, width=8, height=3.75)
 
 
-# variables plot -----
-meas_labels = c(e_dem="Expected\nDem. seats", u_glb=expression(U^G), u_loc=expression(U^L),
-                f="ΔH(q)", h="H", egap="Efficiency\ngap", pbias="Partisan\nbias", mean_med="Mean-median")
-plot_var = function(nm, i) { suppressMessages({
[…]

### 2022-04-05 R/05_boston.R

@@ -1,0 +1,182 @@
+library(cmdstanr)
+library(tidybayes)
+library(posterior)
+library(bayesplot)
+
+d = read_rds(here("data/boston_elec.rds"))
+map = redist_map(d, existing_plan=ccd_2010, adj=d$adj)
+
+d_votes = d %>%
+    as_tibble() %>%
+    mutate(vap_other = vap - vap_white - vap_black - vap_hisp) %>%
+    select(precinct, vap, mayor_gen_wu:pres_prelim_castro) %>%
+    pivot_longer(mayor_gen_wu:pres_prelim_castro, names_to="cand", values_to="votes") %>%
+    separate(cand, c("elec", "stage", "candidate"), sep="_", extra="merge") %>%
+    filter(stage == "prelim", candidate != "other") %>%
+    select(-stage)
+d_abstain = d_votes %>%
+    group_by(precinct, elec) %>%
+    summarize(candidate="<abstain/other>",
+              votes = vap[1] - sum(votes),
+              .groups="drop")
+d_votes = bind_rows(d_votes, d_abstain) %>%
+    select(-vap) %>%
+    arrange(precinct, elec, candidate) %>%
+    mutate(precinct = as.factor(precinct),
+           elec = as.factor(elec),
+           candidate = as.factor(candidate)) %>%
+    arrange(elec, precinct)
+# indices of presidential candidates
+pres = which(levels(d_votes$candidate) %in% unique(filter(d_votes, elec=="pres")$candidate)[-1]) - 1L
+
+d_race = d %>%
+    as_tibble() %>%
+    mutate(vap_other = vap - vap_white - vap_black - vap_hisp,
+           across(starts_with("vap_"), ~ . / vap)) %>%
+    select(vap_white:vap_hisp, vap_other)
+    #select(vap_white, vap_other)
+
+d_race = matrix(1, nrow=nrow(d))
+
+m_votes = d_votes %>%
[…]

### 2022-04-05 R/06_boston_fec.R

@@ -1,0 +1,61 @@
+d_raw = read_csv("data-raw/schedule_a-2022-04-05T16_47_45.csv")
+
+candidates = c(
+    C00703975="biden",
+    #C00695510="booker",
+    C00694455="harris",
+    #C00704510="moulton",
+    C00693234="warren",
+    C00659938="yang",
+    #C00694018="gillibrand",
+    #C00696054="williamson",
+    C00696419="klobuchar",
+    C00696948="sanders",
+    C00697441="buttigieg"
+    #C00698050="inslee",
+    #C00698258="hickenlooper",
+    #C00508416="delaney",
+    #C00700906="weld",
+    #C00705186="bennet",
+    #C00706416="bullock",
+    #C00701698="swalwell",
+    #C00701979="ryan",
+    #C00706697="de_blasio",
+    #C00711614="steyer",
+    #C00717033="walsh",
+    #C00728154="bloomberg"
+)
+
+
+d_fec = d_raw %>%
+    transmute(cand=candidates[committee_id],
+              last_name=contributor_last_name,
+              zip=str_sub(contributor_zip, 1, 5)) %>%
+    drop_na(cand) %>%
+    distinct()
+
+
+library(raceproxy)
+
+r_probs = predict_race_sgz_me(last_name, zip, data=d_fec)
+#r_probs = predict_race_sgz(last_name, zip, data=d_fec, iterate=3, regularize=T)
+p_r_est = colSums(r_probs)/sum(r_probs)
+
+fit = model_race(r_probs, cand, zip, data=d_fec, config=list(lr=0.15))
+est = colMeans(fit$p_xr)
+colnames(est) = str_sub(names(r_probs), 4)
+rownames(est) = levels(factor(d_fec$cand))
+tot_row = matrix(colSums(100*est), nrow=1)
+rownames(tot_row) = "TOTAL"
+rbind(round(100*est, 1), tot_row)
+
+(calc_joint_bisgz(r_probs, d_fec$cand) %*% diag(1/p_r_est)) %>%
+    `colnames<-`(colnames(est)) %>%
+    (\(x) round(x*100, 1)) %>%
+    rbind(tot_row)
+
+
+
+
+
+

### 2022-04-05 paper/paper.Rmd
try abstract

@@
 abstract: |
-  Abstract here.
+  We propose a unified framework for evaluating the impact of a districting plan on voters overall and in different groups.
+By aggregating the harm done to individual voters by their inability to elect a candiate of choice, and taking into account the space of counterfactual possibilities, we derive at a set of measures that flexibly adapts to local geographic constraints, different types of elections, and different groups of study.
+The measures have connections to, and distinct advantages over, common redistricting metrics.
+We provide recommendations for estimating these measures from historical election returns, and apply these methods to study partisan gerrymandering of New Jersey congressional districts and the racial impacts of district boundaries for primary elections for the Boston city council.
 keywords:
@@
 ```{r nj-pairs, fig.cap="Scatterplots for all pairs of measures.
-Point lightness indicates the number of Democratic seats won (ranging from 7 (dark) to 10 (light)).
+Point lightness indicates the number of Democratic seats won at baseline (ranging from 7 (dark) to 10 (light)).
 The Democratic and Republican gerrymanders are shown with the blue and red lines and crosses.

### 2022-04-05 paper/paper.Rmd
typos

@@
 
-We sample 1,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
+We sample 10,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
 One such plan is shown in Figure \ref{fig:nj-maps}(d).
@@
 ```{r nj-pairs, fig.cap="Scatterplots for all pairs of measures.
-Point lightness indicates the number of Democratic seats won at baseline (ranging from 7 (dark) to 10 (light)).
+Point lightness indicates the number of Democratic seats won at baseline, ranging from 7 (dark) to 10 (light).
 The Democratic and Republican gerrymanders are shown with the blue and red lines and crosses.

### 2022-04-08 paper/paper.Rmd
fix harm & write calc appendix

@@
 If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
+In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined; there is certainly an element of subjectivity in sampling a set of redistricting plans.
+Nonetheless, sampling methods are widely used in practice.
 
@@
 
-Here, motivated primarily by computational convenience, we use a *proportionate uniform shift* model, which focuses on swings in the national political environment as the main determinant of election outcomes.
+We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting.
+Estimating differential racial harm in the two-party setting can be done by first generatinc ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
+Multiparty or primary elections pose additional modeling challenges.
+In Section \ref{sec:boston} we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.
+
+Motivated primarily by computational convenience, we use a *proportionate uniform shift* model, which focuses on swings in the national political environment as the main determinant of election outcomes.
 Each election outcome over the period of interest (generally ten years or less) is modeled as a random deviation from a baseline outcome, and these deviations are assumed to consist of a constant fraction of voters of one party in each precinct switching to the other.
@@
 For ease and efficiency of computation, we stick to the simple proportionate uniform shift model here.
-In Section \ref{sec:boston}, however, we consider a more complicated model for primary elections.
 
@@
 For each sample $\tilde q$ of a counterfactual districting plan, we use the election model to come up with a probability that voters in each precinct are represented by the candidate of their choice under $\tilde q$ but not under $q$.
-This probability is a simple difference in CDF functions under our proportionate uniform shift model.
 When the groups used in $\Delta H(q)$ vary within precincts, we must repeat these computations for each group.
 Then by summing across precincts and/or groups and dividing by the appropriate totals, we arrive at values for $H(q)$ and $\Delta H(q)$.
-Appendix \ref{app:formulas} contains more details, including the specific formulas used to compute $H(q)$, and $\Delta H(q)$ from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
+Appendix \ref{app:formulas} contains more details, including the specific formulas used to compute $H(q)$ and $\Delta H(q)$ for two-party elections from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
 
@@
 One such plan is shown in Figure \ref{fig:nj-maps}(d).
-For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm,[^2] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
+For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^2] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
 [^3] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
@@
 
-The expected number of Democratic seats and the differential partisan harm measures identify each gerrymanders as an outlier in the correct direction (two-sided $p=0.0002$), and as expected from Proposition \ref{prop:fairness}, these two measures are highly correlated.
+The expected number of Democratic seats and the differential partisan harm measures identify each gerrymanders as an outlier in the correct direction (two-sided $p=0.0002$ for expected seats), and as expected from Proposition \ref{prop:fairness}, these two measures are highly correlated.
 The efficiency gap and declination measures also correctly identify the gerrymanders ($p=0.0002$), though they are much more discrete, and some nonlinearity is evident for declination.
@@
 
-Both gerrymanders excessively harm the average voter, with the Democratic plan doing so at a moderately higher rate ($p=0.0002$ for the Democratic plan, $p=0.0006$ for the Republican plan).
+Both gerrymanders excessively harm the average voter, with the Democratic plan doing so at a moderately higher rate ($p=0.0002$ for the Democratic plan, $p=0.0004$ for the Republican plan).
 Notably, the plans which minimize total harm are also those which have near-zero differential partisan harm.
@@
 
-```{=html}
-<!--
+In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^4]
+Throughout this appendix, let $j$ index the set of electoral precincts or other units.
+For concreteness here and without loss of generality we will use the Democratic party as the reference party, letting $s_{jv}$ denote the Democratic two-party vote share in precinct $j$ in election $v$.
+Write-in or minor third-party candidates can be safely ignored, since the overall winner will be precisely the two-party vote winner.
 
-For voter fairness, PUS means that relative share of party's voters in each precinct remains the same. Need only to apply a kernel function (composition of Laplace ICDF + model shift CDF)
+[^4]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+    Population information comes from Census blocks, which are practically never split.
 
-Formulas here.
--->
-```
 ## Proportionate Uniform Shift Details
 
-The proportionate uniform shift model assumes that a constant fraction of *voters of one party* in each precinct switch to the other party.
+The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^5]
 This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
@@
 Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
-For $\Delta H(q)$, which considers voters in each party separately, this helpfully simplifies computations.
+For $\Delta_{\dem-\rep}H(q)$, which considers voters in each party separately, this helpfully simplifies computations.
+
+[^5]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
 
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^4]
-Then a party's vote share in precinct $k$ and election $v$, notated here as $s_{kv}$, is modeled as $$
-\gamma(s_{kv}) = \gamma(\overline s_k) + \eps_v, \qquad \eps_v \iid \Norm(0, \sigma^2_{\mathrm{hist}}),
-$$ where $\overline{s}_k$ is the party baseline, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^6]
+Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
+\gamma(s_{jv}) = \gamma(\overline s_j) + \eps_v, \qquad \eps_v \iid \Norm(0, \sigma^2_{\mathrm{hist}}),
+$$ where $\overline{s}_j$ is the party baseline in the precinct, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
 For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.
 
-[^4]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+[^6]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
 
-# Districting Measures in New Jersey {#app:nj}
+It will simplify matters following to have notation for the overall Democratic vote share in a district.
+For a precinct $j$, let $q^{-1}(j)$ denote the district it belongs to. Then let \[
+    S_{q^{-1}(j), v}=\frac{\sum_{l\in q^{-1}(j)} s_{lv}\avgvote_l}{\sum_{l\in q^{-1}(j)} \avgvote_l}
+\] denote the vote share in the district, where $\avgvote_l$ counts the total number of votes in precinct $l$ at baseline.
+The baseline vote share in the district is then $\bar S_{q^{-1}(j)}$.
 
-# Misrepresentation Map
+## Estimating $H(q)$ and $\Delta H_{\dem-\rep}(q)$
 
-The intuition that drives these measures is the misrepresentation map.
-Consider a state that is partitioned into smaller geographies, typically called precincts or voter tabulation districts.
-These units act as the atoms of redistricting.
-We assume that they are not split in the redistricting process.
-Thus, districts are made of whole precincts.
-For a given precinct, we can thus calculate the expected party share of the district that a precinct is in for any given plan.
-The expected party share can take many forms, as mentioned in *Modeling Election Outcomes*.
-By combining the information about the geographic precinct, we can measure the expected party share that an individual living in that precinct would expect to experience.
+We will build up estimates of the overall $H(q)$ and $\Delta H_{\dem-\rep}(q)$ from precinct-by-group harm estimates.
+Since we assumed that elections are independent of districting plans, we can first average across elections and then across counterfactual plans.
+For a random voter $I$ in precinct $j$, we have for every plan $q$ that the probability of a Democrat winning their district is $$
+\Pr(\win_{q(I)}(V, q)=\dem\mid I\in j) 
+= \Pr(S_{q^{-1}(j),V} > \half)
+= \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}),
+$$ where $\Phi$ is the standard Normal CDF.
+To estimate this probability, we simply replace $\Phi$ with the CDF of the $t_{22}$ distribution ($F_{t_22}$), and $\sigma^2_\mathrm{hist}$ with $\hat\sigma^2_\mathrm{hist}$.
+Extending one step further, for two plans $q$ and $\tilde q$, we have
 
-Visually, we can consider a map colored by party share of the district that each precinct is assigned to.
-Most often, we represent this in a blue to red gradient which offers the typical Democratic or Republican representative dynamic.
-Given a counterfactual, we can then take the difference of the expected party share, precinct-by-precinct.
-This offers a precinct-level coloration based on difference in expected party share at the district level.
-In the typical two-party scenario, using Democratic two-party vote share, a positive difference would indicate that the individuals in those precincts are likely to be represented by Democrats.
-A negative difference would indicate that the individuals in those precincts are likely to be represented by Republicans.
-Thus, positives would be map-matically represented by blues and negatives by reds.
+\begin{equation}\label{eq:appharm1}
+\Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j) 
+= \Pr\qty(S_{q^{-1}(j),V} < \half, S_{\tilde q^{-1}(j),V} > \half).
+\end{equation}
 
-Note that this takes a similar visual form as a dislocation map, though these are conceptually distinct [@defoeubarodd21].
-The dislocation map indicates within a single plan how nearby individuals are likely or unlikely to be represented by the same party.
+Because of the uniform shift, for any election $v$, $S_{q^{-1}(j),v} - S_{\tilde q^{-1}(j),v}$ is constant.
+Therefore if $S_{q^{-1}(j),v}>S_{\tilde q^{-1}(j),v}$, this probability is zero (if a Democrat wins under $\tilde q$ they will definitely win under $q$, which has a higher Democratic vote share).
+But if $S_{q^{-1}(j),v}<S_{\tilde q^{-1}(j),v}$, we can rewrite \eqref{eq:appharm1} as the probability that $\eps_{V}$ falls in a particular range.
+Specifically, we have in general that
 
-We focus largely on the case of a misrepresentation map that is the difference between a single enacted districting plan and an ensemble of alternate, legal districting plans.
-This does not need to be the case, however.
-A misrepresentation map can indicate the difference between two singular maps, calculating the difference in expected district-level party representation.
-This would allow direct computation of utility, fairness, or other metrics based on the difference in the two maps.
-Similarly, given two sets of ensembles, likely generated under different target distributions, we can create a similar misrepresentation map that would allow us to estimate interesting measures, such as the differential harm caused by the inclusion of an additional criteria in state district drawing rules.
+\begin{equation}\label{eq:appharm2}
+\Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j) 
+= \qty(\Phi\qty(\frac{\gamma(\bar S_{\tilde q^{-1}(j)})}{\sigma^2_\mathrm{hist}})
+- \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
+\end{equation}
+
+where $(\cdot)_+\dfeq \max\{0, \cdot\}$.
+If we condition on $V(I)=\dem$, then \eqref{eq:appharm2} is exactly the expected harm to voter $I$ under plans $q$ and $\tilde q$.
+The calculation is analogous, up to sign changes, for the expected harm conditioning on $V(I)=\rep$.
+This completes the first step: averaging harm within precinct and groups and across elections.
+
+We can then estimate the average across counterfactual plans $\tilde Q$ with the usual Monte Carlo estimator: \[
+    \hat h_{j,\dem}(q) 
+    \dfeq \frac{1}{N_\text{sim}} \sum_{k=1}^{N_\text{sim}} 
+    \qty(\Phi\qty(\frac{\gamma(\bar S_{\tilde q_k^{-1}(j)})}{\sigma^2_\mathrm{hist}})
+        - \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
+\] where $\hat h_{j,\dem}(q)$ is our estimate of the average harm to a Democratic voter in precinct $j$, averaging across $V$ and $\tilde Q$.
+The corresponding $\hat h_{j,\rep}(q)$ is computed similarly.
+
+Our final estimates $\hat H(q)$ and $\Delta_{\dem-\rep}\hat H(q)$ may then be straightforwardly calculated as weighted averages:
+
+\begin{align*}
+\hat H(q) &\dfeq
+\frac{\sum_{j} (\bar s_j\hat h_{j,\dem}(q) + 
+(1-\bar s_j)\hat h_{j,\rep}(q))\avgvote_j}{\sum_{j} \avgvote_j} \qand \\
+\Delta_{\dem-\rep}\hat H(q) &\dfeq
+\frac{\sum_{j} (\hat h_{j,\dem}(q) - \hat h_{j,\rep}(q))\avgvote_j}{\sum_{j} \avgvote_j}.
+\end{align*}
+
+Here, we rely on the fact that even across election shifts, the relative balance of each party's voters across precincts is preserved.
+
+# Principal Components Ecological Inference Model for Boston Primary Elections

### 2022-04-08 R/01_metrics.R

@@ -8,15 +8,9 @@
 #'
 #' @returns a matrix
 district_group = function(plans, group) {
-    m = as.matrix(plans)
-    m_grp = arrange(plans, as.integer(draw), district) %>%
-        pull({{ group }}) %>%
-        matrix(nrow=max(plans$district))
-    m_prec = matrix(nrow=nrow(m), ncol=ncol(m))
-    for (i in seq_len(ncol(m))) {
-        m_prec[, i] = m_grp[, i][m[, i]]
-    }
-    m_prec
+    m_grp = matrix(eval_tidy(enquo(group), plans),
+                   nrow=attr(plans, "ndists"))
+    mat_by_prec(as.matrix(plans), m_grp)
 }
 
 #' Voter harm for local representation
@@ -26,18 +20,18 @@ district_group = function(plans, group) {
 #' @param distr_grp a matrix from `district_group()`
 #' @param idx_1,idx_2 indices for each of the comparison groups
 #' @param kernel a vectorized function mapping [0, 1] to [0, 1]
-#' @param invert if `TRUE`, calculate utility for the out-group
 #'
-#' @returns a numeric vector matching `idx_1`
-harm = function(map, group_pop, distr_grp,
-                   idx_1=seq_len(ncol(distr_grp)), idx_2=seq_len(ncol(distr_grp)),
-                   kernel=k_step, invert=FALSE) {
-    m = kernel(distr_grp[, idx_1, drop=FALSE])
-    counterfactual = rowMeans(kernel(distr_grp[, idx_2, drop=FALSE]))
-    harm_benefit = counterfactual - m
-    invert = 1L - invert*2L
-    voters = eval_tidy(enquo(group_pop), map)
-    as.numeric(voters %*% pos_part(invert * harm_benefit)) / sum(voters)
+#' @returns a numeric matrix matching `idx_1`: first row is harm to in-group,
+#'   second row is out-group.
+harm = function(plans, group_share, dvote, rvote,
+                idx_1=seq_len(ncol(as.matrix(plans))),
+                idx_2=seq_len(ncol(as.matrix(plans))),
+                ker=k_step) {
+    x = matrix(
+        ker(eval_tidy(enquo(group_share), plans)),
+        nrow=attr(plans, "ndists")
+    )
+    harm_helper(as.matrix(plans), x, dvote, rvote, idx_1, idx_2)
 }
 
[…]

### 2022-04-08 R/03_sim_nj.R

@@ -17,6 +17,7 @@ if (!file.exists(sim_path <- here("data/nj_sims.rds"))) {
     write_rds(plans, sim_path, compress="xz")
 } else {
     plans = read_rds(sim_path)
+    attr(plans, "ndists") = 12L
 }
 
 statewide = with(nj, sum(ndv)/sum(ndv+nrv))
@@ -70,7 +71,7 @@ if (!file.exists(path <- here("paper/figures/nj_pairs.pdf"))) {
 cat("p-values:")
 pl$plan %>%
     as_tibble() %>%
-    mutate(across(n_dem:decl, pval)) %>%
+    mutate(across(n_dem:h, pval)) %>%
     print()
 
 if (FALSE) {
@@ -85,3 +86,6 @@ ggplot(pl_best, aes(score, dh, color=e_dem)) +
 
 with(pl_best, cor(score, h, method="pearson"))
 }
+
+rm(nj, plans, pl_best, pl_plot, pl_sum, p, p1, p2, p3, p4, pl,
+   m_dem, statewide, prop_setas, meas_labels)

### 2022-04-08 R/05_boston.R

@@ -13,42 +13,78 @@ d_votes = d %>%
     pivot_longer(mayor_gen_wu:pres_prelim_castro, names_to="cand", values_to="votes") %>%
     separate(cand, c("elec", "stage", "candidate"), sep="_", extra="merge") %>%
     filter(stage == "prelim", candidate != "other") %>%
+    #filter(stage == "prelim") %>%
+    #mutate(elec = paste0(elec, "_", stage)) %>%
     select(-stage) %>%
-    arrange(precinct, elec, candidate) %>%
-    mutate(elec = fct_inorder(elec),
+    #arrange(precinct, elec, candidate) %>%
+    mutate(precinct = fct_inorder(precinct),
+           elec = fct_inorder(elec),
            candidate = fct_inorder(candidate))
 
+vs_mayor = rowwise(d) %>%
+    mutate(x = mayor_gen_wu / sum(c_across(starts_with("mayor_gen")))) %>%
+    pull()
+vs_pres = rowwise(d) %>%
+    mutate(x = (pres_prelim_sanders + pres_prelim_warren) / sum(c_across(starts_with("pres_prelim")))) %>%
+    pull()
+
+#d_abstain = d_votes %>%
+#    group_by(precinct, elec) %>%
+#    summarize(candidate = as.character(str_glue("<abstain {elec[1]}>")),
+#              vap = vap[1],
+#              votes = vap - sum(votes),
+#              .groups="drop")
+#d_votes = bind_rows(d_votes, d_abstain) %>%
+#    arrange(precinct, elec) %>%
+#    mutate(candidate = fct_inorder(candidate))
+
+
 m_votes = d_votes %>%
+    arrange(precinct, candidate) %>%
     group_by(precinct, elec) %>%
     mutate(vs = votes / sum(1e-6 + votes)) %>%
     ungroup() %>%
-    with(., matrix(vs, nrow=n_distinct(precinct), byrow=T))
+    with(., matrix(vs, nrow=n_distinct(precinct), byrow=T)) %>%
+    `colnames<-`(levels(d_votes$candidate))
 d_sum = d_votes %>%
     group_by(elec, candidate) %>%
     summarize(votes = sum(votes)) %>%
     group_by(elec) %>%
     mutate(vs = round(votes / sum(votes), 3)) %>%
     ungroup() %>%
[…]

### 2022-04-08 src/harm.cpp

@@ -1,0 +1,55 @@
+#include <Rcpp.h>
+using namespace Rcpp;
+
+// [[Rcpp::export(rng = false)]]
+NumericMatrix mat_by_prec(IntegerMatrix pl, NumericMatrix x) {
+    int row = pl.nrow();
+    int col = pl.ncol();
+    NumericMatrix m_prec(row, col);
+
+    for (int i = 0; i < col; i++) {
+        for (int j = 0; j < row; j++) {
+            m_prec(j, i) = x(pl(j, i) - 1, i);
+        }
+    }
+
+    return m_prec;
+}
+
+// [[Rcpp::export(rng = false)]]
+NumericMatrix harm_helper(IntegerMatrix pl, NumericMatrix pr,
+                          NumericVector wt_a, NumericVector wt_b,
+                          IntegerVector idx_1, IntegerVector idx_2) {
+    int n1 = idx_1.size();
+    int n2 = idx_2.size();
+    int prec = pl.nrow();
+    NumericMatrix out(3, n1);
+
+    double tot_a = sum(wt_a);
+    double tot_b = sum(wt_b);
+    for (int i = 0; i < n1; i++) {
+        int i1 = idx_1(i) - 1;
+        double pos_accuml = 0.0;
+        double neg_accuml = 0.0;
+        for (int k = 0; k < prec; k++) {
+            double pr_ik = pr(pl(k, i1) - 1, i1);
+            double wt_a_k = wt_a(k);
+            double wt_b_k = wt_b(k);
+            for (int j = 0; j < n2; j++) {
+                int j2 = idx_2(j) - 1;
+                double harm_benefit = pr(pl(k, j2) - 1, j2) - pr_ik;
+                if (harm_benefit >= 0.0) {
+                    pos_accuml += harm_benefit * wt_a_k;
+                } else {
+                    neg_accuml -= harm_benefit * wt_b_k;
+                }
+            }
+        }
+        out(0, i) = pos_accuml / tot_a / n2;
+        out(1, i) = neg_accuml / tot_b / n2;
+        out(2, i) = (pos_accuml + neg_accuml) / (tot_a + tot_b) / n2;
+        Rcpp::checkUserInterrupt();
+    }
+
+    return out;
+}

### 2022-04-08 paper/paper.Rmd
cleanup boston process; rasterize pairs points

@@
 
-The expected number of Democratic seats and the differential partisan harm measures identify each gerrymanders as an outlier in the correct direction (two-sided $p=0.0002$ for expected seats), and as expected from Proposition \ref{prop:fairness}, these two measures are highly correlated.
-The efficiency gap and declination measures also correctly identify the gerrymanders ($p=0.0002$), though they are much more discrete, and some nonlinearity is evident for declination.
-In contrast, the partisan bias and mean-median measures incorrectly classify the Republican gerrymander as a fair plan and a Democratic-leaning outlier, respectively ($p=0.099$ and $p=0.0022$).
+Differential partisan harm identifies each gerrymander as such, with a large negative value for the Democratic gerrymander and a large positive value for the Republican gerrymander.
+These values are relatively directly interpretable; for example, the differential harm score of $-0.22$ means that an average Democratic voter is 22% less likely to be misrepresented than their average Republican counterpartisan.
+The efficiency gap and declination measures also correctly identify the gerrymanders, though they are much more discrete, and declination gives a much larger score in absolute value to the Democratic plan than to the Republican.
+In contrast, the partisan bias and mean-median measures incorrectly classify the Republican gerrymander as a fair plan and a Democratic-leaning outlier.
 These measures are weakly correlated with the expected number of Democratic seats, and show substantial variation even for fixed values of total seats won at baseline.
+
+Considering the distribution of these statistics across the distribution of sampled plans doesn't help much, either.
+The expected number of Democratic seats naturally identifies the gerrymanders (two-sided $p=0.0002$) and the direction of bias, and as expected from Proposition \ref{prop:fairness} this measure is highly correlated with differential partisan harm.
+The gerrymanders are also an outlier under the efficiency gap and declination measures also (both $p=0.0002$).
+While the gerrymanders are both an outlier for the mean-median score, only the Democratic plan is an outlier according to the partisan bias measure ($p=0.0022$, versus $p=0.099$ for the Republican plan).
+
 There is also disagreement between the measures about the natural bias of districting plans in New Jersey: the median sampled plan has a partisan bias of $0.047$, indicating a natural Republican bias, but the median mean-median score is just $0.0069$, and the median differential partisan harm is $-0.13$, indicating a natural Democratic bias.

### 2022-04-08 R/00_setup_utils.R

@@ -100,9 +100,10 @@ plot_cds = function(map, pl, county, abbr, qty="ndv") {
         theme_void()
 }
 
-expl_vars = function(pl, labels, refs=character(0), lines_off_diag=FALSE, ...) {
+expl_vars = function(pl, labels, refs=character(0), rasterize=TRUE,...) {
     n_ref = length(refs)
     idx = if (n_ref == 0L) seq_len(nrow(pl)) else -seq_len(n_ref)
+    tmp <- tempfile()
 
     panel.hist <- function(x, ...) {
         usr <- par("usr"); on.exit(par(usr))
@@ -125,21 +126,38 @@ expl_vars = function(pl, labels, refs=character(0), lines_off_diag=FALSE, ...) {
         r <- abs(cor(x[idx], y[idx], use="complete.obs"))
         txt <- format(c(r, 0.123456789), digits = digits)[1]
         txt <- paste0(prefix, txt)
-        text(0.5, 0.5, txt, cex=2.5*sqrt(r), family="Times", offset=0, adj=c(0.5, 0.5))#cex.cor)
+        text(0.5, 0.5, txt, cex=2.5*sqrt(r), family="Times", offset=0, adj=c(0.5, 0.5))
     }
-    panel.points <- function(x, y, ...) {
-        points(x[idx], y[idx], family="Times", ...)
+    panel.points = function(x, y, ...) {
+        coords <- par("usr")
+        gx <- grconvertX(coords[1:2], "user", "inches")
+        gy <- grconvertY(coords[3:4], "user", "inches")
+        width <- max(gx) - min(gx)
+        height <- max(gy) - min(gy)
+
+        if (rasterize) {
+            ragg::agg_png(tmp, width = width, height = height, units = "in", res = 300, bg = "transparent")
+            par(mar = c(0,0,0,0))
+            plot.new()
+            plot.window(coords[1:2], coords[3:4], mar = c(0,0,0,0), xaxs = "i", yaxs = "i")
+            points(x[idx], y[idx], family="Times", ...)
+            dev.off()
+
+            panel <- png::readPNG(tmp)
+            rasterImage(panel, coords[1], coords[3], coords[2], coords[4])
+        } else {
+            points(x[idx], y[idx], family="Times", ...)
+        }
+
         abline(h=0, v=0, lty="dashed")
         for (i in seq_len(n_ref)) {
[…]

### 2022-04-08 R/03_sim_nj.R

@@ -1,8 +1,8 @@
 nj = read_rds(here("data/NJ_cd_final_vtd_20.rds")) %>%
     redist_map(pop_tol=0.01, ndists=12, adj=.$adj)
 
+N_sim = 10000
 if (!file.exists(sim_path <- here("data/nj_sims.rds"))) {
-    N_sim = 10000
     plans = redist_smc(nj, N_sim, counties=county, pop_temper=0.005, verbose=TRUE)#FALSE)
 
     # gerrymanders
@@ -25,10 +25,18 @@ prop_seats = round(attr(nj, "ndists") * statewide)
 
 pl = calc_plans_stats(plans, nj, ndv, nrv)
 
+# get 100% accurate harm for 2 gerrymanders
+hh = harm(pl$distr, dem, nj$ndv, nj$nrv, idx_1=1:2, idx_2=(2+1:N_sim), ker=k_t())
+pl$plan$h_dem[1:2] = hh[1, ]
+pl$plan$h_rep[1:2] = hh[2, ]
+pl$plan$dh[1:2] = hh[1, ] - hh[2, ]
+pl$plan$h[1:2] = hh[3, ]
+
 # Initial plots ----
 map_scale = scale_fill_party_c(name="Democratic\nshare", limits=c(0.19, 0.81))
+geom_simp = rmapshaper::ms_simplify(nj$geometry, keep_shapes=TRUE)
 p1 = ggplot(nj, aes(fill=ndv/(ndv+nrv))) +
-    geom_sf(size=0, color="#00000000") +
+    geom_sf(aes(geometry=geom_simp), size=0, color="#00000000") +
     map_scale +
     labs(title="(a) Partisan patterns") +
     theme_repr_map() +
@@ -61,10 +69,10 @@ if (!file.exists(path <- here("paper/figures/nj_pairs.pdf"))) {
     pl_plot = pl$plan %>%
         as_tibble() %>%
         select(n_dem, e_dem, dh, h, egap, pbias, mean_med, decl)
-    pdf(path, 8.5, 8.5)
-    bind_rows(head(pl_plot, 2), slice_sample(head(pl_plot, -2), n=1500)) %>%
-        expl_vars(labels=meas_labels, refs=GOP_DEM[c(2, 14)])
-    dev.off()
+    pl_plot = bind_rows(head(pl_plot, 2), slice_sample(head(pl_plot, -2), n=2000))
+
[…]

### 2022-04-08 R/04_fit_boston.R

@@ -1,0 +1,170 @@
+library(cmdstanr)
+library(posterior)
+
+fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
+                    chains=4, warmup=500, iter=250, adapt_delta=0.95, ...) {
+    stan_d = list(
+        N = nrow(tbls$votes),
+        K = length(tbls$cands),
+        L = ncol(tbls$votes),
+        R = ncol(tbls$race),
+
+        vap = tbls$vap,
+        votes = tbls$votes,
+        vote_share = tbls$vote_share,
+        vap_race = tbls$race,
+        elec = as.integer(tbls$elecs),
+
+        Q = as.integer(n_comp),
+        n_id = nrow(id),
+        id_idx = match(id$cand, tbls$cands),
+        id_loc = do.call(rbind, id$loc)[, seq_len(nrow(id)), drop=F],
+        id_scale = do.call(rbind, id$scale)[, seq_len(nrow(id)), drop=F]
+    )
+
+    path_model = here("R/pcei.stan")
+    path_exc = here("R/pcei")
+
+    sm = cmdstan_model(path_model, compile=F)
+    if (file.exists(path_exc)) {
+        if (isTRUE(recompile) || file.info(path_exc)["mtime"] < file.info(path_model)["mtime"]) {
+            file.remove(path_exc)
+        }
+    }
+    sm$compile()
+
+    if (algorithm == "hmc") {
+        fit = sm$sample(data=stan_d, chains=chains,
+                        iter_warmup=warmup, iter_sampling=iter,
+                        init=0, refresh=50, adapt_delta=adapt_delta, ...)
+    } else if (algorithm == "vb") {
+        fit = sm$variational(data=stan_d, algorithm="meanfield", init=0, ...)
+    } else {
+        stop("Algorithm should be `hmc` or `vb`.")
+    }
+
+    draws = as_draws_rvars(fit$draws())
+
+    draws_dim = dim(draws_of(draws$err_mult))
+    races = colnames(tbls$race)
+    names(draws$support) = tbls$cands
+    rownames(draws$loading) = tbls$cands
+    draws$alpha = c(rvar(array(1, dim=draws_dim), with_chains=TRUE), draws$alpha)
+    names(draws$alpha) = tbls$cands
+    names(draws$turnout_overall) = races
+    colnames(draws$turnout) = races
[…]

### 2022-04-08 R/04_viz_nj.R

@@ -1,45 +1,0 @@
-svc::svc(pl$distr, dem) + scale_color_party_d()
-
-
-ggsave('paper/figures/png/svc.png', dpi = 320, height = 6, width = 6)
-
-svc::svc(pl$distr %>% subset_ref(), dem) + scale_color_party_d()
-ggsave('paper/figures/png/svc_ref.png', dpi = 320, height = 6, width = 6)
-
-counterfactual <- function(map, pl, r, d, lim = 1, ref_idx = 1) {
-  r <- rlang::eval_tidy(rlang::enquo(r), data = map)
-  d <- rlang::eval_tidy(rlang::enquo(d), data = map)
-  map %>%
-    mutate(cf = calc_cf(map, pl, r, d, ref_idx = ref_idx)) %>%
-    ggplot() +
-    geom_sf(aes(fill = cf), lwd = 0.005) +
-    theme_void() +
-    scale_fill_party_c('Simulated - Enacted',
-                       midpoint = 0,
-                       limits = c(min(lim, -lim), max(lim, -lim)),
-    )
-}
-
-
-calc_cf <- function(map, pl, r, d, ref_idx) {
-  pl <- pl %>%
-    mutate(pct_dem = group_frac(map, r, r + d))
-
-  m_pl <- as.matrix(pl)
-  m_val <- pl %>%
-    arrange(as.integer(draw), district) %>%
-    pull(pct_dem) %>%
-    matrix(nrow = attr(map, 'ndists'))
-
-  m_prec <- matrix(nrow = nrow(m_pl), ncol = ncol(m_pl))
-  for (i in seq_len(ncol(m_pl))) {
-    m_prec[, i] <- m_val[, i][m_pl[, i]]
-  }
-
-  rowMeans(-m_prec[, -1:-4] + m_prec[, ref_idx])
-}
-
-counterfactual(nj, pl$distr, r = nrv, d = ndv, ref_idx = 1, lim = 0.5)
-ggsave('paper/figures/png/misr_rep.png', dpi = 320, height = 6, width = 4)
-counterfactual(nj, pl$distr, r = nrv, d = ndv, ref_idx = 2, lim = 0.5)
-ggsave('paper/figures/png/misr_dem.png', dpi = 320, height = 6, width = 4)

### 2022-04-08 R/05_boston.R

@@ -1,121 +1,29 @@
-library(cmdstanr)
-library(tidybayes)
-library(posterior)
-library(bayesplot)
+if (!exists("fit_pcei")) source(here("R/04_fit_boston.R"))
 
-d = read_rds(here("data/boston_elec.rds"))
+d = read_rds(here("data/boston_elec.rds")) %>%
+    mutate(ward = as.integer(str_split(precinct, "-", n=2, simplify=T)[,1]))
 map = redist_map(d, existing_plan=ccd_2010, adj=d$adj)
 
-d_votes = d %>%
-    as_tibble() %>%
-    mutate(vap_other = vap - vap_white - vap_black - vap_hisp) %>%
-    select(precinct, vap, mayor_gen_wu:pres_prelim_castro) %>%
-    pivot_longer(mayor_gen_wu:pres_prelim_castro, names_to="cand", values_to="votes") %>%
-    separate(cand, c("elec", "stage", "candidate"), sep="_", extra="merge") %>%
-    filter(stage == "prelim", candidate != "other") %>%
-    #filter(stage == "prelim") %>%
-    #mutate(elec = paste0(elec, "_", stage)) %>%
-    select(-stage) %>%
-    #arrange(precinct, elec, candidate) %>%
-    mutate(precinct = fct_inorder(precinct),
-           elec = fct_inorder(elec),
-           candidate = fct_inorder(candidate))
+d_votes = make_votes_long(d)
+tbls = make_tables(d_votes, cand_thresh=0.02)
 
-vs_mayor = rowwise(d) %>%
-    mutate(x = mayor_gen_wu / sum(c_across(starts_with("mayor_gen")))) %>%
-    pull()
-vs_pres = rowwise(d) %>%
-    mutate(x = (pres_prelim_sanders + pres_prelim_warren) / sum(c_across(starts_with("pres_prelim")))) %>%
-    pull()
+# priors for identification
+id = tribble(~cand, ~loc, ~scale,
+             "wu", c(-1, 0.8, 0.0), c(0.1, 0.2, 1),
+             "essaibi_george", c(1, 1.2, 0.0), c(0.1, 0.25, 1))
 
-#d_abstain = d_votes %>%
-#    group_by(precinct, elec) %>%
-#    summarize(candidate = as.character(str_glue("<abstain {elec[1]}>")),
-#              vap = vap[1],
[…]

### 2022-04-08 R/06_boston_fec.R

@@ -1,61 +1,0 @@
-d_raw = read_csv("data-raw/schedule_a-2022-04-05T16_47_45.csv")
-
-candidates = c(
-    C00703975="biden",
-    #C00695510="booker",
-    C00694455="harris",
-    #C00704510="moulton",
-    C00693234="warren",
-    C00659938="yang",
-    #C00694018="gillibrand",
-    #C00696054="williamson",
-    C00696419="klobuchar",
-    C00696948="sanders",
-    C00697441="buttigieg"
-    #C00698050="inslee",
-    #C00698258="hickenlooper",
-    #C00508416="delaney",
-    #C00700906="weld",
-    #C00705186="bennet",
-    #C00706416="bullock",
-    #C00701698="swalwell",
-    #C00701979="ryan",
-    #C00706697="de_blasio",
-    #C00711614="steyer",
-    #C00717033="walsh",
-    #C00728154="bloomberg"
-)
-
-
-d_fec = d_raw %>%
-    transmute(cand=candidates[committee_id],
-              last_name=contributor_last_name,
-              zip=str_sub(contributor_zip, 1, 5)) %>%
-    drop_na(cand) %>%
-    distinct()
-
-
-library(raceproxy)
-
-r_probs = predict_race_sgz_me(last_name, zip, data=d_fec)
-#r_probs = predict_race_sgz(last_name, zip, data=d_fec, iterate=3, regularize=T)
-p_r_est = colSums(r_probs)/sum(r_probs)
-
-fit = model_race(r_probs, cand, zip, data=d_fec, config=list(lr=0.15))
-est = colMeans(fit$p_xr)
-colnames(est) = str_sub(names(r_probs), 4)
-rownames(est) = levels(factor(d_fec$cand))
-tot_row = matrix(colSums(100*est), nrow=1)
-rownames(tot_row) = "TOTAL"
-rbind(round(100*est, 1), tot_row)
-
-(calc_joint_bisgz(r_probs, d_fec$cand) %*% diag(1/p_r_est)) %>%
-    `colnames<-`(colnames(est)) %>%
-    (\(x) round(x*100, 1)) %>%
-    rbind(tot_row)
-
-
-
-
-
-

### 2022-04-10 paper/paper.Rmd
checkpoint

@@
 Differential partisan harm identifies each gerrymander as such, with a large negative value for the Democratic gerrymander and a large positive value for the Republican gerrymander.
-These values are relatively directly interpretable; for example, the differential harm score of $-0.22$ means that an average Democratic voter is 22% less likely to be misrepresented than their average Republican counterpartisan.
+These values are relatively directly interpretable; for example, the differential harm score of $-0.221$ means that an average Democratic voter is 22.1% less likely to be misrepresented than their average Republican counterpartisan.
 The efficiency gap and declination measures also correctly identify the gerrymanders, though they are much more discrete, and declination gives a much larger score in absolute value to the Democratic plan than to the Republican.

### 2022-04-10 R/00_setup_utils.R

@@ -1,7 +1,6 @@
 library(rlang)
 library(tidyverse)
 library(sf)
-library(geomander)
 library(redist)
 library(scales)
 library(patchwork)

### 2022-04-10 R/04_fit_boston.R

@@ -2,7 +2,7 @@ library(cmdstanr)
 library(posterior)
 
 fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
-                    chains=4, warmup=500, iter=250, adapt_delta=0.95, ...) {
+                    chains=4, warmup=500, iter=500, adapt_delta=0.8, init=0, ...) {
     stan_d = list(
         N = nrow(tbls$votes),
         K = length(tbls$cands),
@@ -36,7 +36,7 @@ fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
     if (algorithm == "hmc") {
         fit = sm$sample(data=stan_d, chains=chains,
                         iter_warmup=warmup, iter_sampling=iter,
-                        init=0, refresh=50, adapt_delta=adapt_delta, ...)
+                        init=0, refresh=100, adapt_delta=adapt_delta,...)
     } else if (algorithm == "vb") {
         fit = sm$variational(data=stan_d, algorithm="meanfield", init=0, ...)
     } else {
@@ -52,8 +52,8 @@ fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
     draws$alpha = c(rvar(array(1, dim=draws_dim), with_chains=TRUE), draws$alpha)
     names(draws$alpha) = tbls$cands
     names(draws$turnout_overall) = races
-    colnames(draws$turnout) = races
-    dimnames(draws$pref_geo) = list(NULL, NULL, races)
+    colnames(draws$turnout_z) = races
+    dimnames(draws$pref_z) = list(NULL, NULL, races)
     rownames(draws$L_p) = races
     rownames(draws$L_t) = races
     names(draws$sigma_p) = races
@@ -103,8 +103,8 @@ print.fit_pcei = function(x, load_len=NULL, corr=FALSE) {
 
     if (isTRUE(corr)) {
         cat("\nRace correlation:\n")
-        m_t = median(draws$L_t %**% t(draws$L_t))
-        m_p = median(draws$L_p %**% t(draws$L_p))
+        m_t = median(x$L_t %**% t(x$L_t))
+        m_p = median(x$L_p %**% t(x$L_p))
         colnames(m_t) = c("turnout", rep("", ncol(m_t)-1))
         colnames(m_p) = c("loading", rep("", ncol(m_p)-1))
         print(round(cbind(m_t, m_p), 2))

### 2022-04-10 R/05_boston.R

@@ -9,42 +9,51 @@ tbls = make_tables(d_votes, cand_thresh=0.02)
 
 # priors for identification
 id = tribble(~cand, ~loc, ~scale,
-             "wu", c(-1, 0.8, 0.0), c(0.1, 0.2, 1),
-             "essaibi_george", c(1, 1.2, 0.0), c(0.1, 0.25, 1))
+             "wu", c(-1, 1.0, 0.0), c(0.1, 0.1, 1),
+             "essaibi_george", c(1, 1.0, 0.0), c(0.1, 0.2, 1))
 
-draws = fit_pcei(tbls, n_comp=2L, id=id, algorithm="hmc")
+fit = fit_pcei(tbls, n_comp=2L, id=id, algorithm="vb", eta=0.2, adapt_engaged=F)
+init_l = lapply(fit, median)
+fit = fit_pcei(tbls, n_comp=2L, id=id, algorithm="hmc")
 
 library(bayesplot)
-mcmc_intervals(as_draws_matrix(draws$support))
-mcmc_intervals(as_draws_matrix(draws$alpha))
-mcmc_intervals(as_draws_matrix(draws$loading[, 1]))
-mcmc_intervals(as_draws_matrix(draws$loading[, 2]))
-median(draws$L_t %**% t(draws$L_t))
-median(draws$L_p %**% t(draws$L_p))
-mcmc_trace(as_draws_matrix(draws$loading[1:2, ]))
-mcmc_trace(as_draws_matrix(draws$support[1:4]))
-mcmc_pairs(as_draws_matrix(draws$loading[1:2, ]))
+hist(unlist(lapply(fit, posterior::rhat)))
+mcmc_intervals(as_draws_matrix(fit$support))
+mcmc_intervals(as_draws_matrix(fit$alpha))
+mcmc_intervals(as_draws_matrix(fit$loading[, 1]))
+mcmc_intervals(as_draws_matrix(fit$loading[, 2]))
+median(fit$L_t %**% t(fit$L_t))
+median(fit$L_p %**% t(fit$L_p))
+mcmc_trace(as_draws_matrix(fit$loading[1:2, ]))
+mcmc_trace(as_draws_matrix(fit$support[1:4]))
+mcmc_trace(as_draws_matrix(fit$turnout_overall))
+mcmc_trace(as_draws_matrix(fit$scale))
+mcmc_pairs(as_draws_matrix(fit$turnout_overall))
[…]

### 2022-04-12 paper/paper.Rmd
rewriting s2

@@
 For existing measures, this creates a conflict: Republicans win about 40% of the vote, but zero of the nine seats.
-The Efficiency Gap would correctly identify that the enacted map favors Democrats, whereas both Partisan Bias and Mean-Median indicate a pro-Republican bias (ref planscore).
+The Efficiency Gap identifies that the enacted map favors Democrats, whereas both Partisan Bias and Mean-Median indicate a pro-Republican bias under reconstituted recent elections.^[This is according to [PlanScore](https://planscore.campaignlegal.org/massachusetts/#!2020-plan-ushouse-eg), run by the Campaign Legal Center.]
 
-This motivates the need for a counterfactual.
-How can a plan be biased towards or against a party in a case where the party is unable to do worse or to do better than the plan?
-By relying on counterfactual plans, we can then identify the cases where individuals are not represented by their preferred candidate or party.
-In places like Minichusetts, as no plan can be drawn which improves the partisan representation of Republican voters, then there can be no harm.
+Yet, these scores lack a counterfactual.
+Functionally, neither Republicans nor Democrats could do better or worse under this plan than any other plan---a counterfactual plan.
+By relying on counterfactual plans, we can then identify the cases where individuals are not represented by their preferred candidate or party, as an alternative to existing partisan bias mesaures.
+In places like Minichusetts, as no plan can be drawn which improves the partisan representation of Republican voters, under a counterfactual-driven approach, then there is no harm or bias.
 
@@
 In a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions [@chen:rodd:13; @enos:brown:21].
-This allows for significantly greater variation in the expected partisanship of the districts drawn.
+This allows for more variation in the expected partisanship of the districts drawn.
 
@@
 
-A Democratic gerrymander in Mininois effectuates Democratic control by splitting its city into with the blades of a pinwheel reaching out into the more Republican, rural remainder of the state..
-If we consider the two plans in Figure TODO, we see that the Democratic gerrymander would likely elect zero Republicans and three Democrats, whereas the fair map would likely elect two Democrats and one Republican.
-Consider the precincts $(4, 1)$ (maybe highlighted in a color).
-$(4,1)$ is placed in a Republican district in the fair map, while it is in a Democratic district in the gerrymandered map.
-This precinct has 2 Republicans who are not represented by their co-partisan in the gerrymandered plan, but would have been represented by a co-partisan in the fair plan.
-Conversely, the 1 Democrat is represented by their co-partisan in the gerrymandered plan, but not the fair plan.
-Each of these three voters can show a counterfactual where they would be better represented and thus have a claim to harm.
-Yet, if we instead consider any voter who lives in the (other color) $(1, 1)$ precinct, those voters must always be in a Democratic district given the distribution of voters.
-While the Republicans in that precinct are not represented by their co-partisan, they never would be.
-Those Republicans can't be individually harmed by redistricting.
+This pattern allows us to build out the intuition behind our proposed measure.
+Each voter belongs to a group, mostly simply a party.
+For this, we offer that a voter is harmed if they are not represented by their party of choice, but have a counterfactual where they would be represented by their party of choice.
+
+Consider the example in Figure \ref{fig:harm-mininois}.
+A Democratic gerrymander in Mininois effectuates Democratic control by splitting its city into with the blades of a pinwheel, reaching out into the more Republican and less population-dense remainder of the state.
+We see that the Democratic gerrymander would likely elect zero Republicans and three Democrats, whereas the proportional map would likely elect two Democrats and one Republican.
+Figure \ref{fig:harm-mininois}, Panel C then illustrates th total harm by district, comparing the gerrymander to the proportional plan. 
+Republican voters who are forced into a Democratic district under the gerrymander, but would not be under a proportional plan are deemed harmed.
+The Republicans who stay in Democratic districts in the proportional plan are not harmed.
 
@@
 This allows for additional variability, where either party can gerrymander to draw more seats for their party than a neutral process might allow.
-
--   show a fair map and a gerrymandered map
-
--   tally harm by party
 

### 2022-04-12 paper/paper.Rmd
Update paper.Rmd

@@
 
-While we've begun with a two group case, Republicans and Democrats, this can be further extended.
-Perhaps we are interested in a third party or the interaction between White and Black with Democrats and Republicans.
+This framework easily extends to additional parties or races. For example, we can index for the interaction between White and Black with Democrats and Republicans.
 This framework allows for us to index by these groups and simply break down the interactions as components of the sum of the harms.

### 2022-04-18 paper/paper.Rmd
revise s2

@@
 
-To illustrate the concepts and measures we introduce, we will make use of three examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichussets.
+To illustrate the concepts and measures we introduce, we will make use of three examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichusetts.
 Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
@@
 
-First, we consider the case of Minichusetts.
-Modeled on Massachusetts, Minichusetts contains a remarkable lack of partisan segregation.
-This high level of integration, combined with Democrats receiving the majority of votes, leads to a Republican inability to win any seats under recent historical voting patterns [@duch:18].
-For existing measures, this creates a conflict: Republicans win about 40% of the vote, but zero of the nine seats.
-The Efficiency Gap identifies that the enacted map favors Democrats, whereas both Partisan Bias and Mean-Median indicate a pro-Republican bias under reconstituted recent elections.^[This is according to [PlanScore](https://planscore.campaignlegal.org/massachusetts/#!2020-plan-ushouse-eg), run by the Campaign Legal Center.]
+Since every precinct in Minichusetts has a Democratic majority, it is impossible to draw a districting plan built from precincts that creates even a single Republican district.
+This mirrors the situation in the actual state of Massachusetts, where a high level of partisan integration and a solid statewide Democratic majority leads to Republican inability to win any congressional seats under recent historical voting patterns [@duch:18].
 
-Yet, these scores lack a counterfactual.
-Functionally, neither Republicans nor Democrats could do better or worse under this plan than any other plan---a counterfactual plan.
-By relying on counterfactual plans, we can then identify the cases where individuals are not represented by their preferred candidate or party, as an alternative to existing partisan bias mesaures.
-In places like Minichusetts, as no plan can be drawn which improves the partisan representation of Republican voters, under a counterfactual-driven approach, then there is no harm or bias.
+This Democratic dominance poses a problem for existing measures of partisan gerrymandering.
+One of the most common measures, the efficiency gap of @stephanopoulos2015, takes a value of $-0.311$ or $-0.317$ for Minichusetts, depending on the configuration of the districts.
+This would make any Minichusetts districting plan an extreme Democratic gerrymander in comparison to the historical range of efficiency gaps observed in states nationwide.[^1]
+
+[^1]: See, e.g., the historical distribution calculated by the Campaign Legal Center's PlanScore tool: [planscore.campaignlegal.org/metrics/efficiencygap/](https://planscore.campaignlegal.org/metrics/efficiencygap/).
+
+But it is impossible to gerrymander if every manipulation of district boundaries produces no change in partisan outcomes.
+What the efficiency gap and other gerrymandering measures lack is an awareness of the space of possible districting plans, plans that *counterfactually* might have been selected.
+Once a set of districting rules has been agreed on (in Minichusetts, that districts must be contiguous and composed of whole precincts), what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
+There can be no gerrymander, and no harm to individual voters, if no improvement in partisan fairness is possible.
 
@@
 
-Not all cases are as simple as those where one party wins all the seats.
-Mininois also has a Democratic majority, but the voters are no longer integrated.
-In a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions [@chen:rodd:13; @enos:brown:21].
-This allows for more variation in the expected partisanship of the districts drawn.
+To explore how to properly integrate counterfactual districting plans into the analysis of a potential gerrymander, we turn to the more interesting state of Mininois, where, in a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions [@chen:rodd:13; @enos:brown:21].
+
+Consider the two plans in Figure \ref{fig:harm-mininois}(a)-(b).
+A Democratic gerrymander (a) effectuates Democratic control by splitting its city into with the blades of a pinwheel, reaching out into the more Republican and less population-dense remainder of the state.
+We see that the Democratic gerrymander would elect zero Republicans and three Democrats, whereas the proportional map (b) would elect two Democrats and one Republican.
 
@@
 
-This pattern allows us to build out the intuition behind our proposed measure.
-Each voter belongs to a group, mostly simply a party.
-For this, we offer that a voter is harmed if they are not represented by their party of choice, but have a counterfactual where they would be represented by their party of choice.
+Suppose now that plan (a) is enacted into law.
+Who does this decision harm, if anyone?
+We offer that a voter is harmed if they are not represented by their party of choice, but have a counterfactual where they would be represented by their party of choice.
+Using plan (b) as a counterfactual, we can see that rural Republicans in the eastern half of Mininois are harmed by plan (a), since under plan (b) they would be represented by a Republican.
+Republicans in the city and in the southwest are unharmed, as are all Democratic voters,, since they are represented by the same party under both plans.
+Figure \ref{fig:harm-mininois}, Panel C illustrates this by plotting the number of voters harmed in each precinct.
 
-Consider the example in Figure \ref{fig:harm-mininois}.
-A Democratic gerrymander in Mininois effectuates Democratic control by splitting its city into with the blades of a pinwheel, reaching out into the more Republican and less population-dense remainder of the state.
-We see that the Democratic gerrymander would likely elect zero Republicans and three Democrats, whereas the proportional map would likely elect two Democrats and one Republican.
-Figure \ref{fig:harm-mininois}, Panel C then illustrates th total harm by district, comparing the gerrymander to the proportional plan. 
-Republican voters who are forced into a Democratic district under the gerrymander, but would not be under a proportional plan are deemed harmed.
-The Republicans who stay in Democratic districts in the proportional plan are not harmed.
+Measuring the harm of the Democratic gerrymander required comparing it to a fair counterfactual plan.
+The choice of counterfactual might seem hopelessly subjective.
+We discuss in Section \ref{sec:estimating} principled ways to define counterfactuals, but for now we hope the value in measuring harm through counterfactuals is clear even if the operationalization is not.
 
@@
 
-We can also consider a less drastic sorting, where medium cities favor Democrats, but no longer dominate the state politics, such as in Minissouri.
-This allows for additional variability, where either party can gerrymander to draw more seats for their party than a neutral process might allow.
+We could quantify the harm a Democratic gerrymander imposed on voters in Mininois by comparing election outcomes to those under a counterfactual proportional plan.
+But our results said nothing about the direction of the harm.
+While it is apparent that the Democratic gerrymander indeed benefits Democrats, tallying up the total harm done to all voters is not enough to prove this.
 
-This framework easily extends to additional parties or races. For example, we can index for the interaction between White and Black with Democrats and Republicans.
-This framework allows for us to index by these groups and simply break down the interactions as components of the sum of the harms.
+However, we can partition voters into groups, and compare how the harm done by a plan falls across those groups.
+If a plan disproportionately harms voters of one group, then we might conclude that that group was the target of a gerrymander.
+Figure \ref{fig:diffharm-minissouri} illustrates this idea for two hypothetical gerrymanders ((a) and (c)) of Minissouri.
+As in Mininois, we can tally the total number of voters harmed in each precinct ((d) and (f)) in comparison to a proportional counterfactual plan (b).
+As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.
 
@@
 
-Figure \ref{fig:mini-harm} illustrates this measure for a set of hypothetical gerrymanders, comparing to the set of optimal plans from Figure .
-As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.
+Panel (e) then splits the voters by party, and computes, for each party, the share of its voters which harmed.
+The Democratic gerrymander (a) harms Republicans at a much higher rate than Democrats, and the opposite can be said of the Republican gerrymander (c).
+We call the gap between these two rates the *differential harm,* and propose it as a general way to account for the impact of a districting plan on particular groups.
 
-In the districting plan examples of Figure \ref{fig:mini-harm}, the differential harm of the Democratic gerrymander of Mininois is $-0.519$, the differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
-In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure.
+The differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander is $0.450$, where positive values indicate Republican bias.
+In other words, under the Democratic gerrymander, the average Republican voter in Mininois is 41.5% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan in Figure \ref{fig:diffharm-minissouri}(b).
 
@@
 This section will formalize and generalize the concepts in Section \ref{sec:motivation}, beginning with some notation.
-
-## Setup
-
 Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of legislative districts, indexed by $\Distrs$.
@@
 To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
-In general we will not require the group identities to be fixed with respect to elections.[^1]
+In general we will not require the group identities to be fixed with respect to elections.[^2]
 This is because some group identities, like partisan preference, may themselves change election-to-election.
 
-[^1]: Formally, we have random indicator variables $\ind\{i\in G_k\}$ for all $i$ and $k$, subject to the constraint that $\sum_k \ind\{i\in G_k\}=1$ for all $i$.
+[^2]: Formally, we have random indicator variables $\ind\{i\in G_k\}$ for all $i$ and $k$, subject to the constraint that $\sum_k \ind\{i\in G_k\}=1$ for all $i$.
     These indicator variables may be correlated with $V$ and/or $\tilde Q$, in general.
@@
 It is derived from an individual-level measure of representation and it accounts for the space of counterfactual possibilities.
-Importantly it separates *vote choice* from *group membership*, allowing one to discuss how partisan election outcomes under a specific districting plan create harm for specific groups or their interactions, simply by conditioning $H(q)$ on a set of individuals.
+Importantly, it separates *vote choice* from *group membership*, allowing one to discuss how partisan election outcomes under a specific districting plan create harm for specific groups or their interactions, simply by conditioning $H(q)$ on a set of individuals.
 
@@
 One such plan is shown in Figure \ref{fig:nj-maps}(d).
-For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^2] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
-[^3] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
+For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^3] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
+[^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
 
-[^2]: As in section \ref{sec:connections}, this is $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
+[^3]: As in section \ref{sec:connections}, this is $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
 
-[^3]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature [@katz21].
+[^4]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature [@katz21].
 
@@
 
-In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^4]
+In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^5]
 Throughout this appendix, let $j$ index the set of electoral precincts or other units.
@@
 
-[^4]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+[^5]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
     Population information comes from Census blocks, which are practically never split.
@@
 
-The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^5]
+The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^6]
 This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
@@
 
-[^5]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
+[^6]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
 
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^6]
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^7]
 Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
@@
 
-[^6]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+[^7]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
 
 It will simplify matters following to have notation for the overall Democratic vote share in a district.
-For a precinct $j$, let $q^{-1}(j)$ denote the district it belongs to. Then let \[
+For a precinct $j$, let $q^{-1}(j)$ denote the district it belongs to.
+Then let $$
     S_{q^{-1}(j), v}=\frac{\sum_{l\in q^{-1}(j)} s_{lv}\avgvote_l}{\sum_{l\in q^{-1}(j)} \avgvote_l}
-\] denote the vote share in the district, where $\avgvote_l$ counts the total number of votes in precinct $l$ at baseline.
+$$ denote the vote share in the district, where $\avgvote_l$ counts the total number of votes in precinct $l$ at baseline.
 The baseline vote share in the district is then $\bar S_{q^{-1}(j)}$.
@@
 
+```{=tex}
 \begin{equation}\label{eq:appharm1}
@@
 \end{equation}
-
+```
 Because of the uniform shift, for any election $v$, $S_{q^{-1}(j),v} - S_{\tilde q^{-1}(j),v}$ is constant.
@@
 
+```{=tex}
 \begin{equation}\label{eq:appharm2}
@@
 \end{equation}
-
+```
 where $(\cdot)_+\dfeq \max\{0, \cdot\}$.
@@
 
-We can then estimate the average across counterfactual plans $\tilde Q$ with the usual Monte Carlo estimator: \[
+We can then estimate the average across counterfactual plans $\tilde Q$ with the usual Monte Carlo estimator: $$
     \hat h_{j,\dem}(q) 
@@
         - \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
-\] where $\hat h_{j,\dem}(q)$ is our estimate of the average harm to a Democratic voter in precinct $j$, averaging across $V$ and $\tilde Q$.
+$$ where $\hat h_{j,\dem}(q)$ is our estimate of the average harm to a Democratic voter in precinct $j$, averaging across $V$ and $\tilde Q$.
 The corresponding $\hat h_{j,\rep}(q)$ is computed similarly.
@@
 
+```{=tex}
 \begin{align*}
@@
 \end{align*}
-
+```
 Here, we rely on the fact that even across election shifts, the relative balance of each party's voters across precincts is preserved.

### 2022-04-19 paper/paper.Rmd
first 2 boston figures

@@
 
+```{r boston-map, fig.cap="Racial composition of the city of Boston (left), and boundaries of the 2010 city coucil districts (right).
+Neighborhood boundaries are overlaid in black on the left map and in white on the right map.", fig.width=6.75, fig.height=3.375}
+include_graphics(here("paper/figures/boston_map.pdf"))
+```
+
+```{r boston-factors, fig.cap="Candidate factor loadings, with approximate 80\\% credible intervals shown as ellipses (left), and geographic variation in factor values by race (right).
+Neighborhood boundaries overlaid in black.
+Factor values in the Black second-dimension map are truncated so that variation in the other maps is visible.", fig.width=7, fig.height=3.5}
+include_graphics(here("paper/figures/boston_factors.pdf"))
+```
+
 # Discussion {#sec:discussion}

### 2022-04-19 R/00_setup_utils.R

@@ -11,9 +11,11 @@ library(Rcpp)
 theme_repr = function() theme_bw(base_family="Times", base_size=10)
 theme_repr_map = function() theme_void(base_family="Times", base_size=10)
 
-PAL_COAST = c("#7BAEA0", "#386276", "#3A4332", "#7A7D6F", "#D9B96E", "#BED4F0")
-PAL_LARCH = c("#D2A554", "#626B5D", "#8C8F9E", "#858753", "#A4BADF", "#D3BEAF")
-PAL = PAL_COAST[c(5, 1, 2, 4, 3, 6)]
+PAL = wa_pal("coast", which=c(5, 1, 2, 4, 3, 6))
+PAL_RACE = wa_pal("skagit", which=c(1, 4, 6, 3))
+names(PAL_RACE) = c("white", "black", "hisp", "other")
+NAMES_RACE = c("White", "Black", "Hispanic", "other")
+
 GOP_DEM = c("#A0442C", "#B25D4C", "#C27568", "#D18E84", "#DFA8A0",
             "#EBC2BC",  "#F6DCD9", "#F9F9F9", "#DAE2F4", "#BDCCEA",
             "#9FB6DE", "#82A0D2", "#638BC6", "#3D77BB", "#0063B1")
@@ -26,6 +28,8 @@ scale_color_party_d = function(...) {
     scale_color_manual(..., values=c(GOP_DEM[2], GOP_DEM[14]),
                        labels=c("Rep.", "Dem."))
 }
+
+scale_fill_ideo = \(...) scale_fill_wa_c(palette="stuart", ...)
 
 # 'fixed' log that handles 0 and negative numbers
 flog = function(x) if_else(x <= 0, -1, suppressWarnings(log(x)))

### 2022-04-19 R/04_fit_boston.R

@@ -1,6 +1,6 @@
 library(cmdstanr)
 library(posterior)
-
+library(tidybayes)
 
 
 fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
@@ -70,9 +70,9 @@ fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
     if (algorithm == "hmc") {
         fit = sm$sample(data=stan_d, chains=chains,
                         iter_warmup=warmup, iter_sampling=iter,
-                        init=0, refresh=100, adapt_delta=adapt_delta,...)
+                        init=init, refresh=100, adapt_delta=adapt_delta,...)
     } else if (algorithm == "vb") {
-        fit = sm$variational(data=stan_d, algorithm="meanfield", init=0, ...)
+        fit = sm$variational(data=stan_d, algorithm="meanfield", init=init, ...)
     } else {
         stop("Algorithm should be `hmc` or `vb`.")
     }
@@ -108,7 +108,7 @@ fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
     dimnames(draws$pref) = list(NULL, NULL, races)
 
     dim(signs) = c(1, n_comp)
-    draws$post_factor = draws$post_factoe * signs
+    draws$post_factor = draws$post_factor * signs
 
     rownames(draws$L_p) = races
     rownames(draws$L_t) = races

### 2022-04-19 R/05_boston.R

@@ -1,5 +1,7 @@
 if (!exists("fit_pcei")) source(here("R/04_fit_boston.R"))
 
+# Load and fit -------------
+
 d = read_rds(here("data/boston_elec.rds")) %>%
     mutate(ward = as.integer(str_split(precinct, "-", n=2, simplify=T)[,1]))
 map = redist_map(d, existing_plan=ccd_2010, adj=d$adj)
@@ -23,6 +25,133 @@ if (!file.exists(fit_path <- here("data-raw/boston_fit_2.rds"))) {
 }
 
 
+# Plots --------
+## map -----
+map$geometry = rmapshaper::ms_simplify(map$geometry, 0.1, keep_shapes=TRUE)
+map$geometry2 = rmapshaper::ms_simplify(d$geometry, 0.04, keep_shapes=TRUE)
+geom_dist = summarize(group_by(map, ccd_2010))
+geom_nbhd = summarize(group_by(map, nbhd)) %>%
+    as_tibble() %>%
+    st_as_sf() %>%
+    mutate(cuml_area = cume_dist(st_area(geometry)))
+nbhd_show = with(geom_nbhd,
+                 (cuml_area >= 0.4 | nbhd %in% c("Downtown", "Back Bay")) &
+                 (!nbhd %in% c("Harbor Islands", "South Boston Waterfront")))
+geom_ward = summarize(group_by(map, ward))
+
+
+map = map %>%
+    mutate(maj_grp = colnames(tbls$race)[apply(tbls$race, 1, which.max)],
+           maj_share = apply(tbls$race, 1, max))
+
+p1 = ggplot(map, aes(fill=maj_grp, alpha=maj_share)) +
+    geom_sf(size=0, color=NA) +
+    geom_sf(data=geom_nbhd, inherit.aes=F, fill=NA, size=0.25, color="black") +
+    geom_sf_text(aes(label=str_wrap(str_to_upper(nbhd), 15)), data=geom_nbhd[nbhd_show,],
+                 inherit.aes=F, size=2.0, color="#ffffffaa", nudge_y=100,
+                 family="DIN Condensed", fontface="bold") +
+    scale_fill_manual(name="Majority racial group",
+                      values=PAL_RACE, labels=NAMES_RACE) +
+    scale_alpha_binned(name="Share of precinct", breaks=c(0.5, 0.7, 0.9),
+                       labels=percent, range=c(0.3, 1), oob=squish) +
+    coord_sf(expand=F) +
+    theme_repr_map() +
[…]

### 2022-04-29 paper/paper.Rmd
boston + AL work

@@
 
-```{r boston-map, fig.cap="Racial composition of the city of Boston (left), and boundaries of the 2010 city coucil districts (right).
+The proposed harm measures are particularly easy to calculate and interpret in a partisan context, since the preferences of each kind of partisan are directly observable as partisan vote shares.
+However, the same framework can be fruitfully applied to more complex settings, providing insights that go beyond those available through traditional methods.
+
+```{r boston-map, fig.cap="Racial composition of the city of Boston (left), and 2010 city coucil districts (right).
 Neighborhood boundaries are overlaid in black on the left map and in white on the right map.", fig.width=6.75, fig.height=3.375}
@@
 ```
+
+The city of Boston is governed by a mayor and a nonpartisan 13-member city council.
+Nine of the councillors are elected from contiguous city council districts, and the remaining four councillors are elected at-large, with all councillors serving two-year terms.
+Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^5]
+This section will study the extent to which the mixed at-large and district council system harms voters in comparison to a fully district-based system (as was the case before 1909 and from 1924-1951), either overall or differentially by racial group.
+Figure \ref{fig:boston-map} shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.
+
+[^5]: City of Boston charter.
+    Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
 
@@
 
-In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^5]
+In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^6]
 Throughout this appendix, let $j$ index the set of electoral precincts or other units.
@@
 
-[^5]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+[^6]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
     Population information comes from Census blocks, which are practically never split.
@@
 
-The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^6]
+The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^7]
 This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
@@
 
-[^6]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
+[^7]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
 
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^7]
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^8]
 Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
@@
 
-[^7]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+[^8]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
 
@@
 # Principal Components Ecological Inference Model for Boston Primary Elections
+
+```{r boston-gen-pred, fig.cap="Model predictions of general election vote shares and turnout, compared to actual general election results.
+Red reference line indicates perfect agreement.", fig.width=6.5}
+include_graphics(here("paper/figures/boston_gen_pred.pdf"))
+```

### 2022-04-29 R/04_fit_boston.R

@@ -174,6 +174,52 @@ print.fit_pcei = function(x, load_len=NULL, corr=FALSE) {
     }
 }
 
+
+predict.fit_pcei = function(object, support, loading, tbls, recompile=FALSE) {
+    n_comp = length(fit$scale)
+    stopifnot(ncol(loading) == n_comp)
+
+
+    lsupport = qlogis(support)
+
+    stan_d = list(
+        Q = n_comp,
+        N = nrow(tbls$votes),
+        K = nrow(loading),
+        R = ncol(tbls$race),
+
+        vap = tbls$vap,
+        vap_race = tbls$race,
+
+        loading = loading,
+        support = lsupport
+    )
+
+    path_model = here("R/pcei_gq.stan")
+    path_exc = here("R/pcei_gq")
+
+    sm = cmdstan_model(path_model, compile=F)
+    if (file.exists(path_exc)) {
+        if (isTRUE(recompile) || file.info(path_exc)["mtime"] < file.info(path_model)["mtime"]) {
+            file.remove(path_exc)
+        }
+    }
+    sm$compile()
+
+    draws_obj = object[c("turnout", "scale", "pref", "err_mult")]
+    colnames(draws_obj$turnout) = NULL
+    dimnames(draws_obj$pref) = list(NULL, NULL, NULL)
+
+    gqs = sm$generate_quantities(posterior::as_draws_array(draws_obj),
+                                 data=stan_d, sig_figs=4)
+
+    votes = as_draws_rvars(gqs$draws())$votes
+    dimnames(votes) = list(NULL, colnames(tbls$race), rownames(loading))
+
+    votes
+}
+
+
 rot_mat = function(deg) {
     deg = deg * pi/180
     matrix(c(cos(deg), sin(deg), -sin(deg), cos(deg)), 2, 2)

### 2022-04-29 R/05_boston.R

@@ -90,7 +90,7 @@ d_load_exp = group_by(d_load, cand, elec, vs) %>%
     mutate(load_2 = if_else(cand=="campbell", load_2-0.1, load_2))
 
 library(ggrepel)
-nice_name = function(x) str_to_title(str_replace(x, "_", "-"))
+nice_name = function(x) str_to_title(str_replace(x, "_", " "))
 nice_elec = function(x) c(mayor="Mayoral", pres="Presidential")[x]
 
 p1 = ggplot(NULL, aes(load_1, load_2)) +
@@ -148,6 +148,103 @@ p2 = ggplot(map_fact, aes(fill=value)) +
 p1 + p2
 
 ggsave(here("paper/figures/boston_factors.pdf"), height=3.5, width=7, device=cairo_pdf)
+
+
+
+# Predictions -------
+
+## Mayoral General Election -------
+wu_gen_total = with(d, sum(mayor_gen_wu)/(sum(mayor_gen_wu) + sum(mayor_gen_essaibi_george)))
+wu_gen = with(d, (mayor_gen_wu)/((mayor_gen_wu) + (mayor_gen_essaibi_george)))
+
+prior_gen = tbls$summary$vs[1:2] / sum(tbls$summary$vs[1:2])
+loading = mean(fit$loading)[1:2,]
+
+est_votes = predict(fit, prior_gen, loading, tbls)
+est_turn_r = 1e-6 + rvar_apply(est_votes, 1:2, rvar_sum)
+est_turn = rvar_apply(est_turn_r, 1, rvar_sum)
+est_vs = est_votes / est_turn_r
+est_votes_total = rvar_apply(est_votes, c(1, 3), rvar_sum)
+est_vs_total = as_tibble(est_votes_total / est_turn)
+
+act_turn = with(d, mayor_gen_wu + mayor_gen_essaibi_george)
+p1 = ggplot(d, aes(x=wu_gen, y=median(est_vs_total$wu))) +
+    geom_point(size=0.6) +
+    geom_abline(slope=1, color="red") +
+    scale_x_continuous("Actual Wu vote share", labels=percent) +
+    scale_y_continuous("Estimated Wu vote share", labels=percent) +
+    theme_repr()
+p2 = ggplot(d, aes(x=act_turn, y=median(est_turn))) +
+    geom_point(size=0.6) +
[…]

### 2022-04-29 R/06_fit_alabama.R

@@ -1,0 +1,108 @@
+library(cmdstanr)
+library(posterior)
+library(tidybayes)
+
+
+fit_ei = function(tbls, recompile=FALSE, algorithm="vb",
+                  chains=4, warmup=1000, iter=500, adapt_delta=0.8, init=0, ...) {
+    stan_d = list(
+        N = nrow(tbls$votes),
+        L = ncol(tbls$votes),
+        R = ncol(tbls$race),
+
+        vap = tbls$vap,
+        votes = tbls$votes,
+        dem_votes = tbls$dem_votes,
+        vap_race = tbls$race
+    )
+
+    path_model = here("R/ei.stan")
+    path_exc = here("R/ei")
+
+    sm = cmdstan_model(path_model, compile=F)
+    if (file.exists(path_exc)) {
+        if (isTRUE(recompile) || file.info(path_exc)["mtime"] < file.info(path_model)["mtime"]) {
+            file.remove(path_exc)
+        }
+    }
+    sm$compile()
+
+    if (algorithm == "hmc") {
+        fit = sm$sample(data=stan_d, chains=chains,
+                        iter_warmup=warmup, iter_sampling=iter,
+                        init=init, refresh=100, adapt_delta=adapt_delta,...)
+    } else if (algorithm == "vb") {
+        fit = sm$variational(data=stan_d, algorithm="meanfield", init=init, ...)
+    } else {
+        stop("Algorithm should be `hmc` or `vb`.")
+    }
+
+    vars = c("lp__", "turnout_overall", "turnout_elec", "turnout", "L_t", "sigma_t",
+             "support_overall", "support_elec", "support", "L_s", "sigma_s")
+    draws = as_draws_rvars(fit$draws(variables=vars))
+
+    draws_dim = dim(draws_of(draws$lp__, with_chains=TRUE))
+    races = colnames(tbls$race)
+
+    names(draws$turnout_overall) = races
+    colnames(draws$turnout) = races
+    names(draws$turnout_elec) = tbls$elecs
+    rownames(draws$L_t) = races
+    names(draws$sigma_t) = races
+
+    names(draws$support_overall) = races
+    colnames(draws$support) = races
+    names(draws$support_elec) = tbls$elecs
+    rownames(draws$L_s) = races
+    names(draws$sigma_s) = races
+
+    attr(draws, "sm") = fit
+    draws
+}
+
+
+
+# Data prep functions ----
+
[…]

### 2022-04-29 R/07_alabama.R

@@ -1,0 +1,62 @@
+if (!exists("fit_ei")) source(here("R/04_fit_boston.R"))
+
+if (!file.exists(path <- here("data/AL_cd_final_vtd_20.rds"))) {
+    al_shp = geomander::get_alarm("AL", geometry=TRUE, epsg=2759)
+    al_shp$geometry = rmapshaper::ms_simplify(al_shp$geometry, keep=0.03, keep_shapes=TRUE)
+    al_cd = read_sf(here("data-raw/al_2020_congress_2021-11-04_2031-06-30/al_2020_congress_2021-11-04_2031-06-30.shp")) %>%
+        st_transform(4269)
+    al_shp$cd_2020 = geomander::geo_match(al_shp, al_cd, method="area", epsg=2759)
+    al_shp = relocate(al_shp, cd_2020, .after=vtd)
+    al_map = redist_map(al_shp, existing_plan=cd_2020, pop_tol=0.005)
+
+    write_rds(al_map, path, compress="xz")
+} else {
+    al_map = read_rds(path)
+}
+
+d_votes = make_votes_long(al_map)
+tbls = make_tables(d_votes)
+
+fit = fit_ei(tbls, algorithm="hmc", init=0, chains=4, adapt_delta=0.99)
+fit = fit_ei(tbls, algorithm="vb", init=0, eta=0.1, adapt_engaged=F, tol_rel_obj=0.005)
+
+median(fit$L_t %**% t(fit$L_t))
+median(fit$L_s %**% t(fit$L_s))
+
+p1 = plot(al_map, median(fit$turnout)[, "white"]) + scale_fill_wa_c(name="Turnout", labels=percent, limits=c(0, 1)) + labs(title="White turnout")
+p2 = plot(al_map, median(fit$turnout)[, "black"]) + scale_fill_wa_c(name="Turnout", labels=percent, limits=c(0, 1)) + labs(title="Black turnout")
+p3 = plot(al_map, median(fit$support)[, "white"]) + scale_fill_party_c() + labs(title="White Dem. support")
+p4 = plot(al_map, median(fit$support)[, "black"]) + scale_fill_party_c() + labs(title="Black Dem. support")
+p1 + p2 + p3 + p4 + plot_layout(nrow=1, guides="collect") & theme_repr_map() & theme(legend.position="bottom")
+
+
+est_ndv = median(fit$support * fit$turnout) * with(al_map, cbind(vap_white, vap_black, vap_other))
[…]

### 2022-05-02 paper/paper.Rmd
work

@@
 
-```{r mini-schematic, fig.cap="Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts.
-The total voting population for each example is 180.
-White lines delineate precincts."}
-include_graphics(here("paper/figures/minis_schematic.pdf"))
+```{r mini-schematic}
+#| fig.cap: |
+#|      Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts.
+#|      The total voting population for each example is 180. 
+#|      White lines delineate precincts.
+include_graphics("figures/minis_schematic.pdf")
 ```
@@
 
-```{r harm-mininois, fig.cap="Two potential districting plans for Mininois: a Democratic gerrymander (a) and a close-to-proportional plan (b).
-Panel (c) shows the total harm to voters in each precinct done by (a), using (b) as a counterfactual."}
-include_graphics(here("paper/figures/harm_mininois.pdf"))
+```{r harm-mininois}
+#| fig.cap: |
+#|      Two potential districting plans for Mininois: a Democratic gerrymander (a) and a close-to-proportional plan (b).
+#|      Panel (c) shows the total harm to voters in each precinct done by (a), using (b) as a counterfactual.
+include_graphics("figures/harm_mininois.pdf")
 ```
@@
 
-```{r diffharm-minissouri, fig.cap="Three potential districting plans for Minissouris: a Democratic gerrymander (a), a close-to-proportional plan (b), and a Republican gerrymander (c).
-Panels (d) and (f) show the total harm to voters in each precinct done by their corresponding gerrymander, using (b) as a counterfactual.
-Panel (e) summarizes this harm by party and across precincts, showing the average harm done to voters of each party by each gerrymander."}
-include_graphics(here("paper/figures/diffharm_minissouri.pdf"))
+```{r diffharm-minissouri}
+#| fig.cap: |
+#|      Three potential districting plans for Minissouris: a Democratic gerrymander (a), 
+#|      a close-to-proportional plan (b), and a Republican gerrymander (c).
+#|      Panels (d) and (f) show the total harm to voters in each precinct done 
+#|      by their corresponding gerrymander, using (b) as a counterfactual.
+#|      Panel (e) summarizes this harm by party and across precincts, showing 
+#|      the average harm done to voters of each party by each gerrymander.
+include_graphics("figures/diffharm_minissouri.pdf")
 ```
@@
 
-```{r nj-maps, fig.cap="Baseline vote patterns, two hypothetical gerrymanders, and one randomly sampled plan.
-Census-designated places are shaded in gray and county lines in white.", fig.width=6.5, fig.height=3}
-include_graphics(here("paper/figures/nj_maps.pdf"))
+```{r nj-maps, fig.width=6.5, fig.height=3}
+#| fig.cap: |
+#|      Baseline vote patterns, two hypothetical gerrymanders, and one randomly sampled plan.
+#|      Census-designated places are shaded in gray and county lines in white.
+include_graphics("figures/nj_maps.pdf")
 ```
@@
 
-We sample 10,000 districting plans using the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
+We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
 One such plan is shown in Figure \ref{fig:nj-maps}(d).
@@
 
-[^4]: We include partisan bias and mean-median in these figures as it is common practice, though it may have theoretical issues when used for a delegation rather than a legislature [@katz21].
-
-```{r nj-pairs, fig.cap="Scatterplots for all pairs of measures.
-Point lightness indicates the number of Democratic seats won at baseline, ranging from 7 (dark) to 10 (light).
-The Democratic and Republican gerrymanders are shown with the blue and red lines and crosses.
-The dashed line indicates a value of zero.
-Below the diagonal, numbers indicate the absolute value of the correlation between variables.", fig.width=7, fig.height=7}
-include_graphics(here("paper/figures/nj_pairs.pdf"))
+```{r nj-pairs, fig.width=7, fig.height=7}
+#| fig.cap: |
+#|      Scatterplots for all pairs of measures.
+#|      The Democratic and Republican gerrymanders are shown with the blue and red lines and crosses.
+#|      The dashed line indicates a value of zero.
+#|      Below the diagonal, numbers indicate the absolute value of the correlation between variables.
+include_graphics("figures/nj_pairs.pdf")
 ```
@@
 
-```{r boston-map, fig.cap="Racial composition of the city of Boston (left), and 2010 city coucil districts (right).
-Neighborhood boundaries are overlaid in black on the left map and in white on the right map.", fig.width=6.75, fig.height=3.375}
-include_graphics(here("paper/figures/boston_map.pdf"))
+```{r boston-map, fig.width=6.75, fig.height=3.375}
+#| fig.cap: |
+#|      Racial composition of the city of Boston (left), and 2010 city coucil districts (right).
+#|      Neighborhood boundaries are overlaid in black on the left map and in white on the right map.
+include_graphics("figures/boston_map.pdf")
 ```
@@
 
-```{r boston-factors, fig.cap="Candidate factor loadings, with approximate 80\\% credible intervals shown as ellipses (left), and geographic variation in factor values by race (right).
-Neighborhood boundaries overlaid in black.
-Factor values in the Black second-dimension map are truncated so that variation in the other maps is visible.", fig.width=7, fig.height=3.5}
-include_graphics(here("paper/figures/boston_factors.pdf"))
+## Approach
+
+To apply our proposed framework to study the council system and and its district boundaries, we need to be able to predict individual voting preferences under counterfactual districting plans and systems.
+This is more difficult than in our study of New Jersey, for two reasons.
+First, the city council is officially nonpartisan and city politics are dominated by the Democratic Party, so voting preferences are more complex than a simple partisan preference.
+To estimate election outcomes, we will need some way of mapping observed support for particular candidates in one election onto counterfactual support for a hypothetical candidate in another election.
+Second, we aim to examine the racial impact of the district system, and thus need to estimate these voting preferences by race.
+Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone tell us how different racial groups voted.
+
+```{r boston-factors, fig.cap="", fig.width=7, fig.height=3.5}
+#| fig.cap: |
+#|      Candidate factor loadings, with approximate 80% credible intervals shown as ellipses (left), 
+#|      and geographic variation in factor values by race (right). 
+#|      Neighborhood boundaries overlaid in black.
+#|      Factor values in the Black second-dimension map are truncated so that variation in the other maps is visible.
+include_graphics("figures/boston_factors.pdf")
 ```
+
+To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^6] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
+Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail and contains additional diagnostic plots.
+
+[^6]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+
+Like all ecological inference models, ours rests on several unverifiable assumptions.
+The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
+As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not strictly met.
+Yet we believe that the model is a useful simplification of reality that captures the major electoral dynamics that are relevant to our question of interest.
+For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates that advanced to the general election, we were able to predict the precinct-level general election turnout and vote shares with considerable accuracy (see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
+We stress that the latent factor ecological inference model is not necessary for this kind of analysis; local survey data that collected ideological preferences and racial information would also suffice.
+
+## Results
 
@@
 
-In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^6]
+In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^7]
 Throughout this appendix, let $j$ index the set of electoral precincts or other units.
@@
 
-[^6]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+[^7]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
     Population information comes from Census blocks, which are practically never split.
@@
 
-The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^7]
+The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^8]
 This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
@@
 
-[^7]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
+[^8]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
 
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^8]
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^9]
 Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
@@
 
-[^8]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+[^9]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
 
@@
 
-# Principal Components Ecological Inference Model for Boston Primary Elections
+# Latent Factor Ecological Inference Model for Boston Primary Elections {#app:lfei}
 
-```{r boston-gen-pred, fig.cap="Model predictions of general election vote shares and turnout, compared to actual general election results.
-Red reference line indicates perfect agreement.", fig.width=6.5}
-include_graphics(here("paper/figures/boston_gen_pred.pdf"))
+```{r boston-gen-pred, fig.width=6.5}
+#| fig.cap: |
+#|      Model predictions of general election vote shares and turnout, compared to actual general election results. 
+#|      Red reference line indicates perfect agreement. 
+#|      Actual and estimated results have Spearman correlations of 0.87 (vote share) and 0.84 (turnout rate).
+include_graphics("figures/boston_gen_pred.pdf")
 ```

### 2022-05-02 R/00_setup_utils.R

@@ -158,12 +158,10 @@ expl_vars = function(pl, labels, refs=character(0), rasterize=TRUE,...) {
         }
     }
 
-    seats_vec = as.integer(as.factor(pl$n_dem))[idx]
-    pairs(select(pl, -n_dem), labels=labels, oma=rep(1.75, 4),
+    pairs(pl, labels=labels, oma=rep(1.75, 4),
           cex=0.10, gap=0.5, family="Times", cex.labels=0.9, ...,
           lower.panel=panel.cor, upper.panel=panel.points,
-          diag.panel=panel.hist,  text.panel=panel.text,
-          col=str_c(wa_pal("foothills", n=max(seats_vec), which=1:12)[seats_vec], "a0"))
+          diag.panel=panel.hist, text.panel=panel.text, col="#22222255")
 }
 
 

### 2022-05-02 R/03_sim_nj.R

@@ -1,9 +1,10 @@
 nj = read_rds(here("data/NJ_cd_final_vtd_20.rds")) %>%
     redist_map(pop_tol=0.01, ndists=12, adj=.$adj)
 
-N_sim = 10000
+N_sim = 2500
 if (!file.exists(sim_path <- here("data/nj_sims.rds"))) {
-    plans = redist_smc(nj, N_sim, counties=county, pop_temper=0.005, verbose=TRUE)#FALSE)
+    plans = redist_smc(nj, N_sim, counties=county, pop_temper=0.005,
+                       runs=4, verbose=TRUE)
 
     # gerrymanders
     set.seed(5118)
@@ -17,13 +18,13 @@ if (!file.exists(sim_path <- here("data/nj_sims.rds"))) {
     write_rds(plans, sim_path, compress="xz")
 } else {
     plans = read_rds(sim_path)
-    attr(plans, "ndists") = 12L
 }
 
 statewide = with(nj, sum(ndv)/sum(ndv+nrv))
 prop_seats = round(attr(nj, "ndists") * statewide)
 
 pl = calc_plans_stats(plans, nj, ndv, nrv)
+pl$plan$chain = by_plan(plans$chain, ndists=12)
 
 # get 100% accurate harm for 2 gerrymanders
 hh = harm(pl$distr, dem, nj$ndv, nj$nrv, idx_1=1:2, idx_2=(2+1:N_sim), ker=k_t())
@@ -34,7 +35,7 @@ pl$plan$h[1:2] = hh[3, ]
 
 # Initial plots ----
 map_scale = scale_fill_party_c(name="Democratic\nshare", limits=c(0.19, 0.81))
-geom_simp = rmapshaper::ms_simplify(nj$geometry, keep_shapes=TRUE)
+geom_simp = rmapshaper::ms_simplify(nj$geometry, keep=0.03, keep_shapes=TRUE)
 p1 = ggplot(nj, aes(fill=ndv/(ndv+nrv))) +
     geom_sf(aes(geometry=geom_simp), size=0, color="#00000000") +
     map_scale +
@@ -68,7 +69,7 @@ meas_labels = c(e_dem="Expected\nDem. seats", dh="Differential harm", h="Average
 if (!file.exists(path <- here("paper/figures/nj_pairs.pdf"))) {
     pl_plot = pl$plan %>%
         as_tibble() %>%
[…]

### 2022-05-02 R/04_fit_boston.R

@@ -3,7 +3,7 @@ library(posterior)
 library(tidybayes)
 
 
-fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
+fit_lfei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
                     chains=4, warmup=1000, iter=500, adapt_delta=0.8, init=0, ...) {
     n_comp = as.integer(n_comp)
     K = length(tbls$cands)
@@ -56,8 +56,8 @@ fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
         loading_cov = loading_cov
     )
 
-    path_model = here("R/pcei.stan")
-    path_exc = here("R/pcei")
+    path_model = here("R/lfei.stan")
+    path_exc = here("R/lfei")
 
     sm = cmdstan_model(path_model, compile=F)
     if (file.exists(path_exc)) {
@@ -119,7 +119,7 @@ fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
     rownames(draws$post_share_race) = tbls$cands
     colnames(draws$post_share_prec) = tbls$cands
 
-    class(draws) = c("fit_pcei", class(draws))
+    class(draws) = c("fit_lfei", class(draws))
     attr(draws, "sm") = fit
     attr(draws, "loading_loc") = loading_loc
     attr(draws, "loading_cov") = loading_cov
@@ -127,9 +127,9 @@ fit_pcei = function(tbls, n_comp=2L, id=NULL, recompile=FALSE, algorithm="vb",
 }
 
 # generics ----
-print.fit_pcei = function(x, load_len=NULL, corr=FALSE) {
+print.fit_lfei = function(x, load_len=NULL, corr=FALSE) {
     n_comp = dim(x$scale)
-    cat("A principal components ecological inference model fit with",
+    cat("A latent factor ecological inference model fit with",
         n_comp, "components.\n\n")
 
     comp_scale = rev(median(x$scale^2))
@@ -175,12 +175,11 @@ print.fit_pcei = function(x, load_len=NULL, corr=FALSE) {
 }
 
 
-predict.fit_pcei = function(object, support, loading, tbls, recompile=FALSE) {
+predict.fit_lfei = function(object, support, loading, tbls, recompile=FALSE) {
     n_comp = length(fit$scale)
     stopifnot(ncol(loading) == n_comp)
 
[…]

### 2022-05-02 R/05_boston.R

@@ -1,10 +1,10 @@
-if (!exists("fit_pcei")) source(here("R/04_fit_boston.R"))
+if (!exists("fit_lfei")) source(here("R/04_fit_boston.R"))
 
 # Load and fit -------------
 
 d = read_rds(here("data/boston_elec.rds")) %>%
     mutate(ward = as.integer(str_split(precinct, "-", n=2, simplify=T)[,1]))
-map = redist_map(d, existing_plan=ccd_2010, adj=d$adj)
+map = redist_map(d, existing_plan=ccd_2010, pop_tol=0.05, adj=d$adj)
 
 d_votes = make_votes_long(d)
 tbls = make_tables(d_votes, cand_thresh=0.02)
@@ -18,7 +18,7 @@ if (!file.exists(fit_path <- here("data-raw/boston_fit_2.rds"))) {
         )
     )
 
-    fit = fit_pcei(tbls, n_comp=2L, id=id, algorithm="hmc", step_size=0.03)
+    fit = fit_lfei(tbls, n_comp=2L, id=id, algorithm="hmc", step_size=0.03)
     save_fit(fit, fit_path)
 } else {
     fit = read_rds(fit_path)
@@ -189,8 +189,13 @@ cat("Wu vote share Spearman correlation:",
     cor(median(est_turn), act_turn, method="spearman"))
 
 
+## Counterfactual -------
 
-support = c(0.5, 0.5)
+n_draws = length(draws_of(fit$lp__))
+draws_dim = dim(draws_of(fit$lp__, with_chains=TRUE))
+
+support = rvar(array(rbeta(n_draws, 20, 20),, dim=draws_dim), with_chains=TRUE)
+support = c(support, 1 - support)
 loading = tribble(~cand, ~f1, ~f2,
                   "Left", -0.0, -1.0,
                   "Right", 0.0, 1.0) %>%
@@ -198,31 +203,40 @@ loading = tribble(~cand, ~f1, ~f2,
     as.matrix()
 
 est_votes = predict(fit, support, loading, tbls)
-win_pr_total = Pr(est_votes_total / est_turn > 0.5)
+est_turn_r = 1e-6 + rvar(rowSums(draws_of(est_votes), dims=3))
+est_turn = 4e-6 + rvar(rowSums(draws_of(est_votes), dims=2))
[…]

### 2022-05-03 paper/paper.Rmd
work on s5

@@
       The authors thank participants of the Harvard American Politics 
-      Research Workshop for helpful comments and discussion.
+      Research Workshop and Imai Research Group  for helpful comments and discussion.
     department: Department of Statistics
@@
 
-In the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
-Specifically, these measures aim to quantify deviations from partisan symmetry, fairness, or competitiveness.
-Some of the most commonly used are perhaps the *efficiency gap* [@stephanopoulos2015], *partisan bias* [@tufte1973pb; @king1987democratic], and the *mean-median* measure [@mcdonald2015unfair].
+The canonical framework for considering partisan fairness is partisan symmetry [@king1987democratic].
+Partisan symmetry implies that the seats, $s$ won by a party with voteshare $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
+Often, the partisan symmetry framework has been applied via partisan bias, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020].
+Partisan bias, denoted by $\beta(s)$ is the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
+To compute partisan bias, researchers have largely relied on assumptions about how the voteshares would swing to reach the alternative voteshares.
+
+More recently, particularly in the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders. 
+The most prominent of these new measures have attempted to offer alternative measures for partisan symmetry.
+The *mean-median* difference looks to relationship between the voteshare of the median seat in a delegation and the mean voteshare of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
+The *declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
+Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015].
+Referred to by Chief Justice Roberts as "sociological gobbledygook" in Gill v.
+Whitford, the efficiency gap looks to measure the votes wasted by the parties and compare the net wasted votes.
+Notably, while the efficiency gap was intended to measure partisan symmetry, it does not do so [@katz2020].
 All of these measures may be estimated using district vote shares alone, and as we discuss next, are to varying degrees related to our proposed measures.
+
+[3] It's just different, but we also think it's better
+Partisan symmetry offers a coherent and theoretically grounded quantity of interest.
+Yet, as a collective, the approaches towards partisan symmetry suffer from a common issue: the lack of a counterfactual.
+
+[4] specific issues with the key approaches and how we improve that.
 

### 2022-05-03 paper/paper.Rmd
missed save
work

@@
 
+-   why counterfactuals?
+
+    -   many factors influence the performance and qualities of an electoral system
+
+    -   understanding how different components of the system contribute requires reaching in and manipulating those components
+
+    -   just like causal inference, you are inherently interested in the counterfactual where some part of the electoral system is different
+
+-   why representational harm?
+
+    -   lit cites on why districts are good
+
+    -   district-based electoral system working well if people are represented by the candidate of their choice
+
+    -   this motivates *representational harm*
+
+    -   we argue that harm is actually a great unifying framework for studying district-based electoral systems
+
 -   what makes a good redistricting plan?
@@
 #| fig.cap: |
-#|      Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts.
+#|      Schematic maps of the three running examples, 
+#|      showing the partisan composition of each of the 36 precincts.
 #|      The total voting population for each example is 180. 
@@
 #| fig.cap: |
-#|      Two potential districting plans for Mininois: a Democratic gerrymander (a) and a close-to-proportional plan (b).
+#|      Two potential districting plans for Mininois: 
+#|      a Democratic gerrymander (a) and a close-to-proportional plan (b).
 #|      Panel (c) shows the total harm to voters in each precinct done by (a), using (b) as a counterfactual.
@@
 #| fig.cap: |
-#|      Three potential districting plans for Minissouris: a Democratic gerrymander (a), 
+#|      Three potential districting plans for Minissouris: 
+#|      a Democratic gerrymander (a), 
 #|      a close-to-proportional plan (b), and a Republican gerrymander (c).
@@
 Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015].
-Referred to by Chief Justice Roberts as "sociological gobbledygook" in Gill v.
-Whitford, the efficiency gap looks to measure the votes wasted by the parties and compare the net wasted votes.
+Referred to by Chief Justice Roberts as "sociological gobbledygook" in Gill v. Whitford, the efficiency gap looks to measure the votes wasted by the parties and compare the net wasted votes.
 Notably, while the efficiency gap was intended to measure partisan symmetry, it does not do so [@katz2020].
@@
 
-[3] It's just different, but we also think it's better
-Partisan symmetry offers a coherent and theoretically grounded quantity of interest.
+[3] It's just different, but we also think it's better Partisan symmetry offers a coherent and theoretically grounded quantity of interest.
 Yet, as a collective, the approaches towards partisan symmetry suffer from a common issue: the lack of a counterfactual.
@@
 [4] specific issues with the key approaches and how we improve that.
+
+[5] bring the mini-examples full circle with a little table of the scores and the harms?
 
@@
 #| fig.cap: |
-#|      Baseline vote patterns, two hypothetical gerrymanders, and one randomly sampled plan.
+#|      Baseline vote patterns, two hypothetical gerrymanders, 
+#|      and one randomly sampled plan.
 #|      Census-designated places are shaded in gray and county lines in white.
@@
 For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^3] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
-[^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
+[\^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
 
@@
 #|      Scatterplots for all pairs of measures.
-#|      The Democratic and Republican gerrymanders are shown with the blue and red lines and crosses.
+#|      The Democratic and Republican gerrymanders are shown 
+#|      with the blue and red lines and crosses.
 #|      The dashed line indicates a value of zero.
@@
 #| fig.cap: |
-#|      Racial composition of the city of Boston (left), and 2010 city coucil districts (right).
+#|      Racial composition of the city of Boston (left), 
+#|      and 2010 city coucil districts (right).
 #|      Neighborhood boundaries are overlaid in black on the left map and in white on the right map.
@@
 Nine of the councillors are elected from contiguous city council districts, and the remaining four councillors are elected at-large, with all councillors serving two-year terms.
-Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^5]
+Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^4]
 This section will study the extent to which the mixed at-large and district council system harms voters in comparison to a fully district-based system (as was the case before 1909 and from 1924-1951), either overall or differentially by racial group.
@@
 
-[^5]: City of Boston charter.
+[^4]: City of Boston charter.
     Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
@@
 
-To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^6] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
+To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^5] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
 Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail and contains additional diagnostic plots.
 
-[^6]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+[^5]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
 
@@
 
-In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^7]
+In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^6]
 Throughout this appendix, let $j$ index the set of electoral precincts or other units.
@@
 
-[^7]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+[^6]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
     Population information comes from Census blocks, which are practically never split.
@@
 
-The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^8]
+The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^7]
 This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
@@
 
-[^8]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
+[^7]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
 
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^9]
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^8]
 Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
@@
 
-[^9]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+[^8]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
 

### 2022-05-03 R/06_fit_alabama.R

@@ -77,7 +77,7 @@ make_votes_long = function(d) {
                elec = fct_inorder(elec))
 }
 
-make_tables = function(d_votes) {
+make_tables = function(d_votes, al_map) {
     m_dem = d_votes %>%
         pivot_wider(names_from=party, values_from=votes) %>%
         arrange(elec, GEOID20) %>%
@@ -92,7 +92,7 @@ make_tables = function(d_votes) {
     colnames(m_votes) = levels(d_votes$elec)
     storage.mode(m_votes) = "integer"
 
-    m_race = d %>%
+    m_race = al_map %>%
         as_tibble() %>%
         mutate(vap_other = vap - vap_white - vap_black,
                across(starts_with("vap_"), ~ . / vap)) %>%
@@ -100,7 +100,7 @@ make_tables = function(d_votes) {
         as.matrix()
     colnames(m_race) = str_sub(colnames(m_race), 5)
 
-    list(vap = as.integer(pmax(d$vap, apply(m_votes, 1, max))),
+    list(vap = as.integer(pmax(al_map$vap, apply(m_votes, 1, max))),
          votes = m_votes,
          dem_votes = m_dem,
          race = m_race,

### 2022-05-04 paper/paper.Rmd
AL and NJ figures

@@
 Who does this decision harm, if anyone?
-We offer that a voter is harmed if they are not represented by their party of choice, but have a counterfactual where they would be represented by their party of choice.
+We offer that a voter is harmed if they are not represented by their chosen party, but would have been so counterfactually.
 Using plan (b) as a counterfactual, we can see that rural Republicans in the eastern half of Mininois are harmed by plan (a), since under plan (b) they would be represented by a Republican.
@@
 
-```{r nj-maps, fig.width=6.5, fig.height=3}
+```{r nj-maps, fig.width=6.5}
 #| fig.cap: |
-#|      Baseline vote patterns, two hypothetical gerrymanders, 
-#|      and one randomly sampled plan.
+#|      Baseline vote patterns (a), two hypothetical gerrymanders (b) and (c), 
+#|      and one randomly sampled plan (d).
 #|      Census-designated places are shaded in gray and county lines in white.
@@
 
+# Application to Voting Rights Litigation in Alabama {#sec:alabama}
+
+```{r al-maps, fig.width=6.5}
+#| fig.cap: |
+#|      Baseline vote patterns (a), 
+#|      fraction of the voting-age population which is Black (b),
+#|      and fraction of voters in each precinct which are harmed by the 2020
+#|      enacted congressional districts versus the set of four remedial plans (c).
+#|      2020 enacted district boundaries are overlaid 
+#|      in black on the left map and in white on the right maps.
+include_graphics("figures/al_maps.pdf")
+```
+
+Differential racial harm: 0.23708
+
+Differential partisan harm: 0.20939
+
+Total harm: 0.089009
+
+Average total harmed: 162390
+
+Average total harmed by group:
+
+| Average total harmed | Democratic | Republican |
+|----------------------|------------|------------|
+| **White**            | 28,476     | 7,867      |
+| **Black**            | 125,635    | 147        |
+| **Other**            | 257        | 7.4        |
+
+: This table is just a placeholder to store the numbers.
+
+[More text after the table.]
+
 # Application to Primary Elections in Boston {#sec:boston}
@@
 #|      and 2010 city coucil districts (right).
-#|      Neighborhood boundaries are overlaid in black on the left map and in white on the right map.
+#|      Neighborhood boundaries are overlaid in black on the 
+#|      left map and in white on the right map.
 include_graphics("figures/boston_map.pdf")

### 2022-05-04 R/03_sim_nj.R

@@ -34,8 +34,8 @@ pl$plan$dh[1:2] = hh[1, ] - hh[2, ]
 pl$plan$h[1:2] = hh[3, ]
 
 # Initial plots ----
-map_scale = scale_fill_party_c(name="Democratic\nshare", limits=c(0.19, 0.81))
-geom_simp = rmapshaper::ms_simplify(nj$geometry, keep=0.03, keep_shapes=TRUE)
+map_scale = scale_fill_party_c(name="Democratic\nshare", limits=c(0.25, 0.75))
+geom_simp = rmapshaper::ms_simplify(nj$geometry, keep=0.025, keep_shapes=TRUE)
 p1 = ggplot(nj, aes(fill=ndv/(ndv+nrv))) +
     geom_sf(aes(geometry=geom_simp), size=0, color="#00000000") +
     map_scale +
@@ -57,9 +57,10 @@ p4 = plot_cds(nj, as.matrix(plans)[,425], county, "NJ") +
     map_scale +
     theme_repr_map() +
     theme(plot.title = element_text(hjust = 0.5))
-p = p1 + p2 + p3 + p4 + plot_layout(guides="collect", nrow=1)
+p = p1 + p2 + p3 + p4 + plot_layout(guides="collect", nrow=1) &
+    theme(plot.margin = unit(rep(0, 4), "cm"))
 if (!file.exists(path <- here("paper/figures/nj_maps.pdf")))
-    ggsave(path, plot=p, width=8, height=3.75)
+    ggsave(path, plot=p, width=7.5, height=3.5)
 
 
 # variables pairs plot -----

### 2022-05-04 R/06_fit_alabama.R

@@ -2,6 +2,31 @@ library(cmdstanr)
 library(posterior)
 library(tidybayes)
 
+load_538_al = function(pl) {
+    url = str_glue("https://projects.fivethirtyeight.com/redistricting-2022-maps/alabama/plaintiffs_plan_{pl}-districts.json")
+    x = suppressWarnings(read_sf(url)$geometry)
+    # MAGIC NUMBERS determined after much painful analysis
+    x = x * matrix(c(39746.27033, 3498.95882, 3477.34807, -39993.28232), nrow=2) + c(-19134951.8329, 9134711.70241)
+    st_crs(x) = 5070
+    st_sf(geometry=st_transform(x, 2759))
+}
+
+make_al_map = function() {
+    al_shp = geomander::get_alarm("AL", geometry=TRUE, epsg=2759)
+    al_shp$geometry = rmapshaper::ms_simplify(al_shp$geometry, keep=0.005, keep_shapes=TRUE)
+    al_cd = read_sf(here("data-raw/al_2020_congress_2021-11-04_2031-06-30/al_2020_congress_2021-11-04_2031-06-30.shp")) %>%
+        st_transform(4269)
+    # add plans
+    al_shp$cd_2020 = geomander::geo_match(al_shp, al_cd, method="area", epsg=2759)
+    al_shp$cd_pet_a = geomander::geo_match(al_shp, load_538_al("a"), method="area")
+    al_shp$cd_pet_b = geomander::geo_match(al_shp, load_538_al("b"), method="area")
+    al_shp$cd_pet_c = geomander::geo_match(al_shp, load_538_al("c"), method="area")
+    al_shp$cd_pet_d = geomander::geo_match(al_shp, load_538_al("d"), method="area")
+
+    al_shp = relocate(al_shp, starts_with("cd_"), .after=vtd)
+    redist_map(al_shp, existing_plan=cd_2020, pop_tol=0.005)
+}
+
 
 fit_ei = function(tbls, recompile=FALSE, algorithm="vb",
                   chains=4, warmup=1000, iter=500, adapt_delta=0.8, init=0, ...) {

### 2022-05-04 R/07_alabama.R

@@ -1,63 +1,117 @@
 if (!exists("fit_ei")) source(here("R/06_fit_alabama.R"))
 
 if (!file.exists(path <- here("data/AL_cd_final_vtd_20.rds"))) {
-    al_shp = geomander::get_alarm("AL", geometry=TRUE, epsg=2759)
-    al_shp$geometry = rmapshaper::ms_simplify(al_shp$geometry, keep=0.03, keep_shapes=TRUE)
-    al_cd = read_sf(here("data-raw/al_2020_congress_2021-11-04_2031-06-30/al_2020_congress_2021-11-04_2031-06-30.shp")) %>%
-        st_transform(4269)
-    al_shp$cd_2020 = geomander::geo_match(al_shp, al_cd, method="area", epsg=2759)
-    al_shp = relocate(al_shp, cd_2020, .after=vtd)
-    al_map = redist_map(al_shp, existing_plan=cd_2020, pop_tol=0.005)
-
+    al_map = make_al_map()
     write_rds(al_map, path, compress="xz")
 } else {
     al_map = read_rds(path)
 }
 
+
 d_votes = make_votes_long(al_map)
 tbls = make_tables(d_votes, al_map)
 
-fit = fit_ei(tbls, algorithm="hmc", init=0, chains=4, adapt_delta=0.99)
-fit = fit_ei(tbls, algorithm="vb", init=0, eta=0.1, adapt_engaged=F, tol_rel_obj=0.005)
+fit = fit_ei(tbls, algorithm="vb", init=0, eta=0.1, adapt_engaged=F, tol_rel_obj=0.002)
 
-median(fit$L_t %**% t(fit$L_t))
-median(fit$L_s %**% t(fit$L_s))
+#  EI outputs--------
+cat("Turnout correlation matrix:\n")
+print(median(fit$L_t %**% t(fit$L_t)))
+cat("Support correlation matrix:\n")
+print(median(fit$L_s %**% t(fit$L_s)))
 
+## Turnout and support maps by race ------
 p1 = plot(al_map, median(fit$turnout)[, "white"]) + scale_fill_wa_c(name="Turnout", labels=percent, limits=c(0, 1)) + labs(title="White turnout")
 p2 = plot(al_map, median(fit$turnout)[, "black"]) + scale_fill_wa_c(name="Turnout", labels=percent, limits=c(0, 1)) + labs(title="Black turnout")
[…]

### 2022-05-04 data-raw/sf-data-prep.R
SF

@@ -1,0 +1,132 @@
+library(tidyverse)
+library(sf)
+library(readxl)
+library(censable)
+library(geomander)
+library(easycensus)
+library(janitor)
+library(redist)
+library(here)
+
+# Shapefiles ------
+d_prec = read_sf(here("data-raw/sf_prec/SF_DOE_PREC_2019.shp")) %>%
+    transmute(precinct = coalesce(PREC_2019, sprintf("U%03d", 1:n())),
+              nbhd = NeighRep,
+              sup_2010 = as.integer(SupDist),
+              geometry = geometry)
+d_bos = read_sf(here("data-raw/sf_bos/SF_BOS_FinalMap.shp")) %>%
+    st_transform(st_crs(d_prec))
+
+d_prec = d_prec %>%
+    mutate(sup_2020 = coalesce(as.integer(d_bos$DISTRICT)[
+        geo_match(d_prec, d_bos, method="area")], 5), # extra bit in the Bay
+        .after=sup_2010)
+
+d_cens = build_dec("block", "CA", county="San Francisco", geometry=TRUE, year=2020) %>%
+    select(-NAME) %>%
+    mutate(blockgroup = str_sub(GEOID, 1, 12),
+           .after = GEOID)
+crosswalk_block = geo_match(d_cens, d_prec, method="area")
+
+# ACS data -----
+d_acs = get_acs_table("block group", "B25008", year=2020,
+                      state="CA", county="San Francisco", drop_total=TRUE) %>%
+    select(GEOID, estimate, tenure) %>%
+    mutate(tenure = str_c(word(tenure, 1, 1), "s")) %>%
+    pivot_wider(names_from=tenure, values_from=estimate)
+d_cens$renters = estimate_down(d_cens$pop, d_acs$renters, match(d_cens$blockgroup, d_acs$GEOID))
+d_cens$owners = estimate_down(d_cens$pop, d_acs$owners, match(d_cens$blockgroup, d_acs$GEOID))
+
+d_cens_prec = d_cens %>%
+    st_drop_geometry() %>%
+    group_by(precinct = d_prec$precinct[crosswalk_block]) %>%
+    summarize(across(c(pop:vap_two, renters, owners), sum)) %>%
+    mutate(rent_frac = renters / (1e-3 + renters + owners),
+           vap_rent = rent_frac * vap,
+           vap_own = (1 - rent_frac) * vap)
+d_prec = left_join(d_prec, d_cens_prec, by="precinct") %>%
[…]

### 2022-05-04 paper/paper.Rmd
SF
tweak

@@
 By aggregating the harm done to individual voters by their inability to elect a candiate of choice, and taking into account the space of counterfactual possibilities, we derive at a set of measures that flexibly adapts to local geographic constraints, different types of elections, and different groups of study.
-The measures have connections to, and distinct advantages over, common redistricting metrics.
-We provide recommendations for estimating these measures from historical election returns, and apply these methods to study partisan gerrymandering of New Jersey congressional districts and the racial impacts of district boundaries for primary elections for the Boston city council.
+We provide recommendations for estimating these measures from historical election returns and discuss connections to commonly used redistricting metrics.
+Finally, we apply our framework to study partisan gerrymandering in New Jersey, voting rights litigation in Alabama, and the racial impacts of city council district boundaries in Boston.
 keywords:
   - representation
-  - redistricting simulation
-  - individual standing
+  - redistricting
+  - voting rights
+  - individual harm
 bibliography: references.bib
@@
 For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^3] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
-[\^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
+The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
 
@@
 #|      fraction of the voting-age population which is Black (b),
-#|      and fraction of voters in each precinct which are harmed by the 2020
-#|      enacted congressional districts versus the set of four remedial plans (c).
+#|      and the expected fraction of voters in each precinct which are harmed by the 
+#|      2020 enacted congressional districts versus the set of four remedial plans (c).
 #|      2020 enacted district boundaries are overlaid 

### 2022-05-04 paper/paper.Rmd
line
typo

@@
 #| fig.cap: |
-#|      Three potential districting plans for Minissouris: 
+#|      Three potential districting plans for Minissouri: 
 #|      a Democratic gerrymander (a), 
@@
 Despite their simplicity, uniform-swing--type models are highly accurate for out-of-sample prediction [@katz2020], and are intuitive and easy to implement.
-More details about our proportionate uniform shift model may be found in Appendix \ref{app:formulas}.
+More details about our proportionate uniform shift model and how it may be used to efficiently estimate the harm measures may be found in Appendix \ref{app:formulas}.
 

### 2022-05-04 paper/paper.Rmd
s5 finally

--- 2022-05-04 paper/paper.Rmd
@@
 
--   why counterfactuals?
-
-    -   many factors influence the performance and qualities of an electoral system
-
-    -   understanding how different components of the system contribute requires reaching in and manipulating those components
-
-    -   just like causal inference, you are inherently interested in the counterfactual where some part of the electoral system is different
-
--   why representational harm?
-
-    -   lit cites on why districts are good
-
-    -   district-based electoral system working well if people are represented by the candidate of their choice
-
-    -   this motivates *representational harm*
-
-    -   we argue that harm is actually a great unifying framework for studying district-based electoral systems
-
 -   what makes a good redistricting plan?
@@
 #| fig.cap: |
-#|      Schematic maps of the three running examples, 
-#|      showing the partisan composition of each of the 36 precincts.
+#|      Schematic maps of the three running examples, showing the partisan composition of each of the 36 precincts.
 #|      The total voting population for each example is 180. 
@@
 #| fig.cap: |
-#|      Two potential districting plans for Mininois: 
-#|      a Democratic gerrymander (a) and a close-to-proportional plan (b).
+#|      Two potential districting plans for Mininois: a Democratic gerrymander (a) and a close-to-proportional plan (b).
 #|      Panel (c) shows the total harm to voters in each precinct done by (a), using (b) as a counterfactual.
@@
 Who does this decision harm, if anyone?
-We offer that a voter is harmed if they are not represented by their chosen party, but would have been so counterfactually.
+We offer that a voter is harmed if they are not represented by their party of choice, but have a counterfactual where they would be represented by their party of choice.
 Using plan (b) as a counterfactual, we can see that rural Republicans in the eastern half of Mininois are harmed by plan (a), since under plan (b) they would be represented by a Republican.
@@
 #| fig.cap: |
-#|      Three potential districting plans for Minissouris: 
-#|      a Democratic gerrymander (a), 
+#|      Three potential districting plans for Minissouris: a Democratic gerrymander (a), 
 #|      a close-to-proportional plan (b), and a Republican gerrymander (c).
@@
 Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015].
-Referred to by Chief Justice Roberts as "sociological gobbledygook" in Gill v. Whitford, the efficiency gap looks to measure the votes wasted by the parties and compare the net wasted votes.
+Referred to by Chief Justice Roberts as "sociological gobbledygook" in Gill v.
+Whitford, the efficiency gap looks to measure the votes wasted by the parties and compare the net wasted votes.
 Notably, while the efficiency gap was intended to measure partisan symmetry, it does not do so [@katz2020].
@@
 
-[3] It's just different, but we also think it's better Partisan symmetry offers a coherent and theoretically grounded quantity of interest.
-Yet, as a collective, the approaches towards partisan symmetry suffer from a common issue: the lack of a counterfactual.
+Partisan symmetry offers a coherent and theoretically grounded quantity of interest.
+The idea of partisan symmetry is rooted in assessing how parties would have performed under counterfactual voteshares.
+Yet, this suffers from a core inflexibility: it does not allow for variation in districting.
+Districting is the fundamental question of district-based democracies. 
+*For whom can you vote? Who do you vote with?*
+Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation. 
+The natural measurement approach to this directly incorporates alternatives under the logic of a counterfactual.
 
-[4] specific issues with the key approaches and how we improve that.
+Consider the major issues that have been raised over existing measures.
+Partisan bias ($\beta(\half)$) can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.
+This has been called by some the Utah paradox [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
+A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
+These stem from the omission of the alternatives.
 
-[5] bring the mini-examples full circle with a little table of the scores and the harms?
+By looking to alternatives, we ground the idea of zero in a balance of the harms, rather than a purely theoretical way.
+This resolves the directionality problem common to other measures.
+The use of a counterfactual in measuring the quantity of interest grounds the bias in the rules of what districts can be drawn.
+This allows us to draw key strength from the impetus of the redistricting simulation literature.
+Yet, the simplicity of the framework does not restrict only to the use of simulation methods.
+Further, this framework does not need to rely on ideas of proportionality, which has no legal role in American democracy.
+Nor does it suffer from discreteness in ensembles, as now both voteshares and districts lines shift.
 
-## The Efficiency Gap
+Despite these differences, there are also key empirical similarities in many cases.
+The efficiency gap and $\Delta H(q)$ are often quite well correlated under competitive states, as wasted votes and partisan harm are rooted in similar logics.
+Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
+Where our framework rises above these though is especially in non-competitive states, where insurmountable advantages are rendered as zeroes and do not contribute to harm.
 
-The efficiency gap ($EG$) interrogates how a party's votes are wasted under a districting plan.
-By subtracting the wasted votes for Democrats from those for Republicans, and normalizing by the total votes cast, it offers a signed measure that is simple to interpret: $EG=0.05$, for example, indicates that 5% more Democratic votes are wasted than Republican ones.
-The general recommendation is that an efficiency gap larger than two Congressional seats (i.e., 2 divided by the number of districts), or 0.08 for state legislative plans, indicates a partisan gerrymander.
+<!-- ## Simulation Methods -->
 
-A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
-These concerns to not apply to our proposed measures.
-
-First, $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$ are all averaged over possible election outcomes, and so do not suffer from discreteness.
-And as discussed above, $U^{(G)}$ and $U^{(L)}$ both reward proportionality.
-Of course, proportionality is not always possible, given a state's political geography.
-The harm measures $H$ and $\Delta H(q)$ both reflect this by comparing to a set of counterfactual districting plans.
-This addresses one inherent shortcoming of the efficiency gap, illustrated best by the state of Massachusetts.
-According to the efficiency gap, Massachusetts' congressional districting plan is a Democratic gerrymander ($EG=-0.14$).
-But it is impossible to draw a single Republican congressional district in the state.
-Consequently $H$ and $\Delta H(q)$ are everywhere 0.
-
-Finally, the fact that the efficiency gap is minimized for districts with 75%--25% vote shares is a direct consequence of comparing the raw number of wasted votes, rather than the share within each party.
-In contrast, $\Delta H(q)$ is calculated by comparing harm averaged first within each party, avoiding this deficiency.
-
-As we will see in the application below, often there is substantial correlation between the efficiency gap and $\Delta H(q)$.
-Yet in more unusual or tricky cases, the efficiency gap may point to a partisan gerrymander while $\Delta H(q)$ indicates a fair outcome or an unfair advantage for the opposite party.
-
-## Partisan Bias
-
-Partisan bias is probably the most widely-used measure of partisan symmetry.
-The bias, denoted by $\beta(s)$ is the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
-Like our proposed measures, this quantity is defined independent of electoral data and can be estimated using a uniform swing model.
-Each value of $s$ yields information about a different point on the seats-votes curve.
-
-In competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
-But in more lopsided states, $\beta(\half)$ can give unintuitive results, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.
-
-Fundamentally, however, $\beta(s)$ measures partisan symmetry, and not partisan harm or utility.
-It should not be surprising, then, that it can and does "disagree" about the partisan effects of a particular plan.
-These "disagreements" are really just differences in the target quantity of interest.
-
-```{=html}
-<!--
-1.  Integrating yields a related measure of @nagle2015measures, which has also been criticized by @deford2020symmetry.
-2.  Focuses on only one part of S-V curve, which may not be relevant
-
-Notably, as the quantity of interest is not partisan symmetry, our measures do not require the existence of a seats-votes curve.
-This is consequential for the study of congressional redistricting.
-As a seats-votes curve is defined for a legislature, rather than a delegation, to understand the impact of single state's congressional map, you would be forced to consider how every other state drew their maps [@katz:king:rose:21a].
-Thus, if this taken as the definition of a seats-votes curve, then measures of partisan symmetry lack diagnostic power to identify gerrymanders for the delegations that academics, legal scholars, and the general public are often focused on.
-The individualized focus of our measures are, however, portable to congressional, state legislative, or any other form of districting and elections.
+<!-- Despite being derived from individual harm, differential partisan harm is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
 -->
-```
-## Mean-Median Difference
-
-The mean-median difference straightforwardly measures the difference between the mean and median district vote shares.
-In practice, the mean-median differences is moderately correlated with other partisan redistricting measures.
-However, considerations of the mean and median district are only related to individuals insofar as they describe the general propensity to be harmed or benefited by a particular map.
-As [@katz2020] discuss, the mean-median difference can be interpreted as an estimator of *votes-denominated bias*, another value derived from the seats-votes curve.
-
-## Simulation Methods
-
-Despite being derived from individual harm, differential partisan harm is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
-Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale, where a value of zero corresponds to fairness.
-In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
-This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry.
+<!-- Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale, where a value of zero corresponds to fairness. -->
+<!-- In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes. -->
+<!-- This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry. -->
 
@@
 
-```{r nj-maps, fig.width=6.5}
+```{r nj-maps, fig.width=6.5, fig.height=3}
 #| fig.cap: |
-#|      Baseline vote patterns (a), two hypothetical gerrymanders (b) and (c), 
-#|      and one randomly sampled plan (d).
+#|      Baseline vote patterns, two hypothetical gerrymanders, and one randomly sampled plan.
 #|      Census-designated places are shaded in gray and county lines in white.
@@
 For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^3] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
-[\^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
+[^4] The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
 
@@
 #|      Scatterplots for all pairs of measures.
-#|      The Democratic and Republican gerrymanders are shown 
-#|      with the blue and red lines and crosses.
+#|      The Democratic and Republican gerrymanders are shown with the blue and red lines and crosses.
 #|      The dashed line indicates a value of zero.
@@
 
-# Application to Voting Rights Litigation in Alabama {#sec:alabama}
-
-```{r al-maps, fig.width=6.5}
-#| fig.cap: |
-#|      Baseline vote patterns (a), 
-#|      fraction of the voting-age population which is Black (b),
-#|      and fraction of voters in each precinct which are harmed by the 2020
-#|      enacted congressional districts versus the set of four remedial plans (c).
-#|      2020 enacted district boundaries are overlaid 
-#|      in black on the left map and in white on the right maps.
-include_graphics("figures/al_maps.pdf")
-```
-
-Differential racial harm: 0.23708
-
-Differential partisan harm: 0.20939
-
-Total harm: 0.089009
-
-Average total harmed: 162390
-
-Average total harmed by group:
-
-| Average total harmed | Democratic | Republican |
-|----------------------|------------|------------|
-| **White**            | 28,476     | 7,867      |
-| **Black**            | 125,635    | 147        |
-| **Other**            | 257        | 7.4        |
-
-: This table is just a placeholder to store the numbers.
-
-[More text after the table.]
-
 # Application to Primary Elections in Boston {#sec:boston}
@@
 #| fig.cap: |
-#|      Racial composition of the city of Boston (left), 
-#|      and 2010 city coucil districts (right).
-#|      Neighborhood boundaries are overlaid in black on the 
-#|      left map and in white on the right map.
+#|      Racial composition of the city of Boston (left), and 2010 city coucil districts (right).
+#|      Neighborhood boundaries are overlaid in black on the left map and in white on the right map.
 include_graphics("figures/boston_map.pdf")
@@
 Nine of the councillors are elected from contiguous city council districts, and the remaining four councillors are elected at-large, with all councillors serving two-year terms.
-Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^4]
+Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^5]
 This section will study the extent to which the mixed at-large and district council system harms voters in comparison to a fully district-based system (as was the case before 1909 and from 1924-1951), either overall or differentially by racial group.
@@
 
-[^4]: City of Boston charter.
+[^5]: City of Boston charter.
     Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
@@
 
-To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^5] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
+To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^6] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
 Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail and contains additional diagnostic plots.
 
-[^5]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+[^6]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
 
@@
 
+For the menagerie, we proffer a new beast. *can we make the whole discussion a zoo metaphor*
 This paper is an attempt to bridge the quantitative gap between votes and representation.

### 2022-05-04 paper/paper.Rmd

--- 2022-05-04 paper/paper.Rmd
@@
 
-[3] It's just different, but we also think it's better Partisan symmetry offers a coherent and theoretically grounded quantity of interest.
-Yet, as a collective, the approaches towards partisan symmetry suffer from a common issue: the lack of a counterfactual.
+Partisan symmetry offers a coherent and theoretically grounded quantity of interest.
+The idea of partisan symmetry is rooted in assessing how parties would have performed under counterfactual voteshares.
+Yet, this suffers from a core inflexibility: it does not allow for variation in districting.
+Districting is the fundamental question of district-based democracies.
+*For whom can you vote? Who do you vote with?*
+Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation.
+The natural measurement approach to this directly incorporates alternatives under the logic of a counterfactual.
 
-[4] specific issues with the key approaches and how we improve that.
+Consider the major issues that have been raised over existing measures.
+Partisan bias ($\beta(\half)$) can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.
+This has been called by some the Utah paradox [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
+A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
+These stem from the omission of the alternatives.
 
-[5] bring the mini-examples full circle with a little table of the scores and the harms?
+By looking to alternatives, we ground the idea of zero in a balance of the harms, rather than a purely theoretical way.
+This resolves the directionality problem common to other measures.
+The use of a counterfactual in measuring the quantity of interest grounds the bias in the rules of what districts can be drawn.
+This allows us to draw key strength from the impetus of the redistricting simulation literature.
+Yet, the simplicity of the framework does not restrict only to the use of simulation methods.
+Further, this framework does not need to rely on ideas of proportionality, which has no legal role in American democracy.
+Nor does it suffer from discreteness in ensembles, as now both voteshares and districts lines shift.
 
-## The Efficiency Gap
+Despite these differences, there are also key empirical similarities in many cases.
+The efficiency gap and $\Delta H(q)$ are often quite well correlated under competitive states, as wasted votes and partisan harm are rooted in similar logics.
+Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
+Where our framework rises above these though is especially in non-competitive states, where insurmountable advantages are rendered as zeroes and do not contribute to harm.
 
-The efficiency gap ($EG$) interrogates how a party's votes are wasted under a districting plan.
-By subtracting the wasted votes for Democrats from those for Republicans, and normalizing by the total votes cast, it offers a signed measure that is simple to interpret: $EG=0.05$, for example, indicates that 5% more Democratic votes are wasted than Republican ones.
-The general recommendation is that an efficiency gap larger than two Congressional seats (i.e., 2 divided by the number of districts), or 0.08 for state legislative plans, indicates a partisan gerrymander.
+<!-- ## Simulation Methods -->
 
-A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
-These concerns to not apply to our proposed measures.
-
-First, $U^{(G)}, U^{(L)}, H$, and $\Delta H(q)$ are all averaged over possible election outcomes, and so do not suffer from discreteness.
-And as discussed above, $U^{(G)}$ and $U^{(L)}$ both reward proportionality.
-Of course, proportionality is not always possible, given a state's political geography.
-The harm measures $H$ and $\Delta H(q)$ both reflect this by comparing to a set of counterfactual districting plans.
-This addresses one inherent shortcoming of the efficiency gap, illustrated best by the state of Massachusetts.
-According to the efficiency gap, Massachusetts' congressional districting plan is a Democratic gerrymander ($EG=-0.14$).
-But it is impossible to draw a single Republican congressional district in the state.
-Consequently $H$ and $\Delta H(q)$ are everywhere 0.
-
-Finally, the fact that the efficiency gap is minimized for districts with 75%--25% vote shares is a direct consequence of comparing the raw number of wasted votes, rather than the share within each party.
-In contrast, $\Delta H(q)$ is calculated by comparing harm averaged first within each party, avoiding this deficiency.
-
-As we will see in the application below, often there is substantial correlation between the efficiency gap and $\Delta H(q)$.
-Yet in more unusual or tricky cases, the efficiency gap may point to a partisan gerrymander while $\Delta H(q)$ indicates a fair outcome or an unfair advantage for the opposite party.
-
-## Partisan Bias
-
-Partisan bias is probably the most widely-used measure of partisan symmetry.
-The bias, denoted by $\beta(s)$ is the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
-Like our proposed measures, this quantity is defined independent of electoral data and can be estimated using a uniform swing model.
-Each value of $s$ yields information about a different point on the seats-votes curve.
-
-In competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
-But in more lopsided states, $\beta(\half)$ can give unintuitive results, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.
-
-Fundamentally, however, $\beta(s)$ measures partisan symmetry, and not partisan harm or utility.
-It should not be surprising, then, that it can and does "disagree" about the partisan effects of a particular plan.
-These "disagreements" are really just differences in the target quantity of interest.
-
-```{=html}
-<!--
-1.  Integrating yields a related measure of @nagle2015measures, which has also been criticized by @deford2020symmetry.
-2.  Focuses on only one part of S-V curve, which may not be relevant
-
-Notably, as the quantity of interest is not partisan symmetry, our measures do not require the existence of a seats-votes curve.
-This is consequential for the study of congressional redistricting.
-As a seats-votes curve is defined for a legislature, rather than a delegation, to understand the impact of single state's congressional map, you would be forced to consider how every other state drew their maps [@katz:king:rose:21a].
-Thus, if this taken as the definition of a seats-votes curve, then measures of partisan symmetry lack diagnostic power to identify gerrymanders for the delegations that academics, legal scholars, and the general public are often focused on.
-The individualized focus of our measures are, however, portable to congressional, state legislative, or any other form of districting and elections.
+<!-- Despite being derived from individual harm, differential partisan harm is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
 -->
-```
-## Mean-Median Difference
-
-The mean-median difference straightforwardly measures the difference between the mean and median district vote shares.
-In practice, the mean-median differences is moderately correlated with other partisan redistricting measures.
-However, considerations of the mean and median district are only related to individuals insofar as they describe the general propensity to be harmed or benefited by a particular map.
-As [@katz2020] discuss, the mean-median difference can be interpreted as an estimator of *votes-denominated bias*, another value derived from the seats-votes curve.
-
-## Simulation Methods
-
-Despite being derived from individual harm, differential partisan harm is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
-Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale, where a value of zero corresponds to fairness.
-In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
-This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry.
+<!-- Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale, where a value of zero corresponds to fairness. -->
+<!-- In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes. -->
+<!-- This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry. -->
 

### 2022-05-05 paper/paper.Rmd
sf work

@@
 Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
-Sections \ref{sec:nj} and \ref{sec:boston} apply the proposed measures to partisan gerrymandering of New Jersey congressional districts and racial impacts of district boundaries on Boston City Council primary elections.
+Sections \ref{sec:nj}, \ref{sec:alabama}, and \ref{sec:boston} apply the proposed measures to partisan gerrymandering of New Jersey congressional districts, voting rights litigation for Alabama congressional districts, and racial impacts of district boundaries on Boston City Council primary elections.
 Section \ref{sec:discussion} concludes.
@@
 Districting is the fundamental question of district-based democracies.
-*For whom can you vote? Who do you vote with?*
-Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation.
+*For whom can you vote? Who do you vote with?* Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation.
 The natural measurement approach to this directly incorporates alternatives under the logic of a counterfactual.
@@
 -->
+
 <!-- Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale, where a value of zero corresponds to fairness. -->
+
 <!-- In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes. -->
+
 <!-- This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry. -->
@@
 
-For the menagerie, we proffer a new beast. *can we make the whole discussion a zoo metaphor*
-This paper is an attempt to bridge the quantitative gap between votes and representation.
+For the menagerie, we proffer a new beast.
+*can we make the whole discussion a zoo metaphor* This paper is an attempt to bridge the quantitative gap between votes and representation.
 We break from existing redistricting literature and focus on the relationship between votes and representation by copartisans, rather than the relationship between votes and seats.

### 2022-05-05 R/00_setup_utils.R

@@ -58,6 +58,11 @@ e_approx = function(a, b) {
                     if_else(is.finite(b), 0.5*(a+b), a), b))
 }
 
+rot_mat = function(deg) {
+    deg = deg * pi/180
+    matrix(c(cos(deg), sin(deg), -sin(deg), cos(deg)), 2, 2)
+}
+
 
 plot_cds = function(map, pl, county, abbr, qty="ndv") {
     plan = as.factor(redist:::color_graph(get_adj(map), as.integer(pl)))
@@ -186,3 +191,4 @@ Rcpp::sourceCpp("src/harm.cpp", cacheDir="src")
 
 if (!exists("harm")) source(here("R/01_metrics.R"))
 if (!exists("mininois")) source(here("R/toy/00_toy_setup.R"))
+source(here("R/04_appl_data.R"))

### 2022-05-05 R/04_appl_data.R

@@ -1,0 +1,132 @@
+load_538_al = function(pl) {
+    url = str_glue("https://projects.fivethirtyeight.com/redistricting-2022-maps/alabama/plaintiffs_plan_{pl}-districts.json")
+    x = suppressWarnings(read_sf(url)$geometry)
+    # MAGIC NUMBERS determined after much painful analysis
+    x = x * matrix(c(39746.27033, 3498.95882, 3477.34807, -39993.28232), nrow=2) + c(-19134951.8329, 9134711.70241)
+    st_crs(x) = 5070
+    st_sf(geometry=st_transform(x, 2759))
+}
+
+make_al_map = function() {
+    al_shp = geomander::get_alarm("AL", geometry=TRUE, epsg=2759)
+    al_shp$geometry = rmapshaper::ms_simplify(al_shp$geometry, keep=0.005, keep_shapes=TRUE)
+    al_cd = read_sf(here("data-raw/al_2020_congress_2021-11-04_2031-06-30/al_2020_congress_2021-11-04_2031-06-30.shp")) %>%
+        st_transform(4269)
+    # add plans
+    al_shp$cd_2020 = geomander::geo_match(al_shp, al_cd, method="area", epsg=2759)
+    al_shp$cd_pet_a = geomander::geo_match(al_shp, load_538_al("a"), method="area")
+    al_shp$cd_pet_b = geomander::geo_match(al_shp, load_538_al("b"), method="area")
+    al_shp$cd_pet_c = geomander::geo_match(al_shp, load_538_al("c"), method="area")
+    al_shp$cd_pet_d = geomander::geo_match(al_shp, load_538_al("d"), method="area")
+
+    al_shp = relocate(al_shp, starts_with("cd_"), .after=vtd)
+    redist_map(al_shp, existing_plan=cd_2020, pop_tol=0.005)
+}
+
+make_tables_alabama = function(d_votes, al_map) {
+    m_dem = d_votes %>%
+        pivot_wider(names_from=party, values_from=votes) %>%
+        arrange(elec, GEOID20) %>%
+        with(., matrix(dem, nrow=n_distinct(GEOID20))) %>%
+        `colnames<-`(levels(d_votes$elec))
+    storage.mode(m_dem) = "integer"
+
+    m_votes = d_votes %>%
+        group_by(GEOID20, elec) %>%
+        summarize(votes = sum(votes), .groups="drop") %>%
+        with(., matrix(votes, nrow=n_distinct(GEOID20), byrow=T))
[…]

### 2022-05-05 R/06_alabama.R

@@ -1,4 +1,4 @@
-if (!exists("fit_ei")) source(here("R/06_fit_alabama.R"))
+if (!exists("fit_ei")) source(here("R/05_fit_ei.R"))
 
 if (!file.exists(path <- here("data/AL_cd_final_vtd_20.rds"))) {
     al_map = make_al_map()
@@ -8,8 +8,17 @@ if (!file.exists(path <- here("data/AL_cd_final_vtd_20.rds"))) {
 }
 
 
-d_votes = make_votes_long(al_map)
-tbls = make_tables(d_votes, al_map)
+d_votes = d %>%
+    as_tibble() %>%
+    mutate(vap_other = vap - vap_white - vap_black) %>%
+    select(GEOID20, pre_16_rep_tru:uss_20_dem_jon) %>%
+    pivot_longer(pre_16_rep_tru:uss_20_dem_jon, names_to="cand", values_to="votes") %>%
+    separate(cand, c("elec", "year", "party", "candidate"), sep="_", extra="merge") %>%
+    mutate(elec = str_c(elec, "_", year)) %>%
+    select(-candidate, -year) %>%
+    mutate(GEOID20 = fct_inorder(GEOID20),
+           elec = fct_inorder(elec))
+tbls = make_tables_alabama(d_votes, al_map)
 
 fit = fit_ei(tbls, algorithm="vb", init=0, eta=0.1, adapt_engaged=F, tol_rel_obj=0.002)
 

### 2022-05-05 R/06_fit_alabama.R

@@ -1,133 +1,0 @@
-library(cmdstanr)
-library(posterior)
-library(tidybayes)
-
-load_538_al = function(pl) {
-    url = str_glue("https://projects.fivethirtyeight.com/redistricting-2022-maps/alabama/plaintiffs_plan_{pl}-districts.json")
-    x = suppressWarnings(read_sf(url)$geometry)
-    # MAGIC NUMBERS determined after much painful analysis
-    x = x * matrix(c(39746.27033, 3498.95882, 3477.34807, -39993.28232), nrow=2) + c(-19134951.8329, 9134711.70241)
-    st_crs(x) = 5070
-    st_sf(geometry=st_transform(x, 2759))
-}
-
-make_al_map = function() {
-    al_shp = geomander::get_alarm("AL", geometry=TRUE, epsg=2759)
-    al_shp$geometry = rmapshaper::ms_simplify(al_shp$geometry, keep=0.005, keep_shapes=TRUE)
-    al_cd = read_sf(here("data-raw/al_2020_congress_2021-11-04_2031-06-30/al_2020_congress_2021-11-04_2031-06-30.shp")) %>%
-        st_transform(4269)
-    # add plans
-    al_shp$cd_2020 = geomander::geo_match(al_shp, al_cd, method="area", epsg=2759)
-    al_shp$cd_pet_a = geomander::geo_match(al_shp, load_538_al("a"), method="area")
-    al_shp$cd_pet_b = geomander::geo_match(al_shp, load_538_al("b"), method="area")
-    al_shp$cd_pet_c = geomander::geo_match(al_shp, load_538_al("c"), method="area")
-    al_shp$cd_pet_d = geomander::geo_match(al_shp, load_538_al("d"), method="area")
-
-    al_shp = relocate(al_shp, starts_with("cd_"), .after=vtd)
-    redist_map(al_shp, existing_plan=cd_2020, pop_tol=0.005)
-}
-
-
-fit_ei = function(tbls, recompile=FALSE, algorithm="vb",
-                  chains=4, warmup=1000, iter=500, adapt_delta=0.8, init=0, ...) {
-    stan_d = list(
-        N = nrow(tbls$votes),
-        L = ncol(tbls$votes),
-        R = ncol(tbls$race),
-
-        vap = tbls$vap,
-        votes = tbls$votes,
-        dem_votes = tbls$dem_votes,
-        vap_race = tbls$race
-    )
-
[…]

### 2022-05-05 R/07_boston.R

@@ -1,4 +1,4 @@
-if (!exists("fit_lfei")) source(here("R/04_fit_boston.R"))
+if (!exists("fit_lfei")) source(here("R/05_fit_ei.R"))
 
 # Load and fit -------------
 
@@ -6,8 +6,18 @@ d = read_rds(here("data/boston_elec.rds")) %>%
     mutate(ward = as.integer(str_split(precinct, "-", n=2, simplify=T)[,1]))
 map = redist_map(d, existing_plan=ccd_2010, pop_tol=0.05, adj=d$adj)
 
-d_votes = make_votes_long(d)
-tbls = make_tables(d_votes, cand_thresh=0.02)
+d_votes =  d %>%
+    as_tibble() %>%
+    mutate(vap_other = vap - vap_white - vap_black - vap_hisp) %>%
+    select(precinct, mayor_gen_wu:pres_prelim_castro) %>%
+    pivot_longer(mayor_gen_wu:pres_prelim_castro, names_to="cand", values_to="votes") %>%
+    separate(cand, c("elec", "stage", "candidate"), sep="_", extra="merge") %>%
+    filter(stage == "prelim", candidate != "other") %>%
+    select(-stage) %>%
+    mutate(precinct = fct_inorder(precinct),
+           elec = fct_inorder(elec),
+           candidate = fct_inorder(candidate))
+tbls = make_tables_boston(d_votes, cand_thresh=0.02)
 
 if (!file.exists(fit_path <- here("data-raw/boston_fit_2.rds"))) {
     id = list(

### 2022-05-05 R/08_sf.R

@@ -1,0 +1,94 @@
+if (!exists("fit_lfei")) source(here("R/05_fit_ei.R"))
+
+# Load and fit -------------
+
+d = read_rds(here("data/sf_elec.rds"))
+map = redist_map(d, existing_plan=sup_2020, pop_tol=0.05, adj=d$adj)
+
+d_votes =  d %>%
+    as_tibble() %>%
+    select(precinct, atty_loftus:p21_no) %>%
+    pivot_longer(atty_loftus:p21_no, names_to="cand", values_to="votes") %>%
+    separate(cand, c("elec", "candidate"), sep="_", extra="merge") %>%
+    mutate(precinct = fct_inorder(precinct),
+           elec = fct_inorder(elec),
+           candidate = if_else(elec %in% c("p19", "p21"),
+                               str_c(elec, "_", candidate), candidate),
+           candidate = fct_inorder(candidate))
+tbls = make_tables_sf(d_votes, cand_thresh=0.02)
+
+if (!file.exists(fit_path <- here("data-raw/sf_fit.rds"))) {
+    id = list(
+        loc = list(p21_yes=c(-1, 0), p21_no=c(1, 0), biden=c(1, 0.5), buttigieg=c(0, 1)),
+        scale = list(p21_yes=c(0.5, 0.25), p21_no=c(0.5, 0.5), biden=c(1, 1), buttigieg=c(1, 0.5))
+    )
+    # fit = fit_lfei(tbls, n_comp=2L, id=id, algorithm="vb", eta=0.5, adapt_engaged=F, tol_rel_obj=0.005)
+
+    fit = fit_lfei(tbls, n_comp=2L, id=id, algorithm="hmc", step_size=0.1,
+                   inv_metric=read_rds(here("data-raw/sf_metric.rds")),
+                   warmup=500, iter=1000, thin=2)
+    save_fit(fit, fit_path)
+} else {
+    fit = read_rds(fit_path)
+}
+
+
+
+# VALIDATION ---------------
+
+library(bayesplot)
+rhats = unlist(lapply(fit, posterior::rhat))
+qplot(rhats, bins=50)
+mcmc_intervals(as_draws_matrix(fit$support))
+mcmc_intervals(as_draws_matrix(fit$alpha))
+mcmc_intervals(as_draws_matrix(fit$loading[, 1]))
[…]

### 2022-05-06 paper/paper.Rmd
move proof

@@
 A natural extension of this definition would be to separate an individual's candidate preference from their vote choice; the latter would be used in calculating election winners, but the former could provide a more broadly-based harm measure.
-Naturally, this approach requires auxiliary data to implement (such as a large-scale survey), since candidate preferences for nonvoters cannot be inferred from election returns.
+Naturally, this approach requires auxiliary data to implement (such as a large-scale survey), since candidate preferences for nonvoters cannot be directly inferred from election returns.
 
@@
 
-Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe next.
+Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe in the next section.
+First, however, we note two interesting consequences of our definition of differential harm.
 
-# Estimating the Harm Measures {#sec:estimating}
+In the partisan case, with $\Groups=\{\dem,\rep\}$, differential harm is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
+Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale where a value of zero corresponds to fairness.
+In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
+This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry.
+
+```{=tex}
+\begin{proprep} \label{prop:fairness}
+Suppose $\Groups=\{\dem,\rep\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
+\begin{align*}
+    \Delta_{\rep-\dem} H(q) = \E[\seats(\dem, V, \tilde Q)] - \E[\seats(\dem, V, q)].
+\end{align*}
+\textbf{CM note: define $\seats$ or replace with sum.}
+\end{proprep}
+\begin{proof}
+\end{proof}
+```
+# Estimating the Harm Measures
 
@@
 
-<!-- ## Simulation Methods -->
-
-<!-- Despite being derived from individual harm, differential partisan harm is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
--->
-
-<!-- Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale, where a value of zero corresponds to fairness. -->
-
-<!-- In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes. -->
-
-<!-- This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry. -->
-
-```{=tex}
-\begin{proprep} \label{prop:fairness}
-Suppose $\Groups=\{\dem,\rep\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
-\begin{align*}
-    \Delta_{\rep-\dem} H(q) = \E[\seats(\dem, V, \tilde Q)] - \E[\seats(\dem, V, q)].
-\end{align*}
-\textbf{CM note: define $\seats$ or replace with sum.}
-\end{proprep}
-\begin{proof}
-\end{proof}
-```
-# Application to Partisan Gerrymandering in New Jersey {#sec:nj}
+# Application to Partisan Gerrymandering in New Jersey
 

### 2022-05-07 paper/paper.Rmd
alabama section draft

@@
 
+We next turn to litigation in Alabama, Merill v.
+Milligan, where civil rights groups sued Alabama arguing that a second Black opportunity-to-elect district should have been drawn.
+For litigation under Section 2 of the Voting Rights Act of 1965 (VRA), plaintiffs submit one or more remedial maps in order to meet the preconditions set under Thornburg v.
+Gingles (478 US 30) in 1986.
+Remedial maps demonstrate that a new plan could be drawn to comply with the VRA.
+We use four remedial plans submitted by plaintiff's expert, Moon Duchin, which each create a moderately compact, second Black opportunity district while complying with other state redistricting criteria.
+We opt to use remedial plans, as they offer a relevant alternative set of plans ($\tilde{Q}$) to which courts already look when establishing liability.
+
+Figure \ref{fig:al-maps} displays the results of this process. 
+Panels (a) and (b) of \ref{fig:al-maps} display the partisanship and Black voting age percentages of Alabama.
+Unsuprisingly, Black population and Democratic voteshare are highly correlated.
+The remedial plans create a Black opportunity district by connecting Mobile to Montgomery, via the Black Belt region.
+Panel (c) of \ref{fig:al-maps} highlights the total harm from the enacted Alabama plan against the remedial plans.
+As we would expect, the harm largely follows the Black belt, where the Black opportunity district was drawn in each of the remedial plans.
+
 ```{r al-maps, fig.width=6.5}
@@
 
-Differential racial harm: 0.23708
+First, this gives us that the average total harmed is 162,390 people.
+This is equivalent to a total harm of 8.9%.
+Geographically, we can see where voters are harmed.
+For example, in the north of Alabama, near Huntville, despite a larger number of Democratic voters in a Republican district, none of them are harmed here.
+That is, they are also in Republican districts in the remedy plans.
+Along the entirety of the Black Belt, there is substantial harm, most of which comes from Black Democrats.
+There is also a pocket of harmed voters in Tuskaloosa, above the left shoulder of district 7.
 
-Differential partisan harm: 0.20939
+We can then break down the harm into components by party.
+We find that an average of 154,368 Democrats are harmed by the enacted plan.
+Yet, on average, only 8,022 Republicans are harmed by the enacted plan.
+After normalizing by group sizes, this gives a substantial  differential harm of 20.9%.
+Thus, compared to the remedy plans, the enacted plan is much more likely to harm Democrats than Republicans.
 
-Total harm: 0.089009
-
-Average total harmed: 162390
-
-Average total harmed by group:
-
-| Average total harmed | Democratic | Republican |
-|----------------------|------------|------------|
-| **White**            | 28,476     | 7,867      |
-| **Black**            | 125,635    | 147        |
-| **Other**            | 257        | 7.4        |
-
-: This table is just a placeholder to store the numbers.
-
-[More text after the table.]
+Taking this further, we can estimate the harm by race as well.
+To do so, we fit an ecological inference model to estimate party-by-race counts for each precinct.
+Using the estimates, we can see that the harm to Democrats is carried largely by Black Democrats over non-Black Democrats.
+Specifically, 125,635 Black Democrats are harmed, compared to 28,476 white Democrats and 257 other race Democrats.
+In comparison, only 147 Black Republicans, 7,867 white Republicans, and 7 (8?) other race Republicans are harmed on average.
+Thus, once normalized by group size, the differential racial harm is 23.7%, indicating that Black voters are harmed substantially more than white voters.
 

### 2022-05-09 paper/paper.Rmd
edit sec 5

@@
   We propose a unified framework for evaluating the impact of a districting plan on voters overall and in different groups.
-By aggregating the harm done to individual voters by their inability to elect a candiate of choice, and taking into account the space of counterfactual possibilities, we derive at a set of measures that flexibly adapts to local geographic constraints, different types of elections, and different groups of study.
+Considering the harm done to individual voters by their inability to elect their chosen candiate, and taking into account the space of counterfactual possibilities, we derive a set of measures that flexibly adapts to local geographic constraints, different types of elections, and different groups of study.
 We provide recommendations for estimating these measures from historical election returns and discuss connections to commonly used redistricting metrics.
@@
 
+```{=html}
+<!-- Yet, this suffers from a core inflexibility: it does not allow for variation in districting.
+Districting is the fundamental question of district-based democracies.
+*For whom can you vote? Who do you vote with?* Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation.
+The natural measurement approach to this directly incorporates alternatives under the logic of a counterfactual. -->
+```
 ## Outline of the Paper
@@
 Section \ref{sec:indiv-harm} formalizes the intuition of Section \ref{sec:motivation} and defines a general measure of individual and differential harm.
-Section \ref{sec:estimating} discusses estimating these measures from observed election results.
-Section \ref{sec:connections} connects our proposed measures to existing symmetry and fairness measures.
+Section \ref{sec:estimating} discusses estimating these measures from observed election results, and provides what we believe is a sensible and efficient way to calculate the measures in the special case of studying partisan gerrymandering.
+Section \ref{sec:connections} connects our proposed measure to existing practices for analyzing racial and partisan gerrymandering.
 Sections \ref{sec:nj}, \ref{sec:alabama}, and \ref{sec:boston} apply the proposed measures to partisan gerrymandering of New Jersey congressional districts, voting rights litigation for Alabama congressional districts, and racial impacts of district boundaries on Boston City Council primary elections.
@@
 #| fig.cap: |
-#|      Schematic maps of the three running examples,
+#|      Schematic maps of the example states,
 #|      showing the partisan composition of each of the 36 precincts.
 #|      The total voting population for each example is 180.
-#|      White lines delineate precincts.
 include_graphics("figures/minis_schematic.pdf")
@@
 
-Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe in the next section.
-First, however, we note two interesting consequences of our definition of differential harm.
+Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe next.
 
-In the partisan case, with $\Groups=\{\dem,\rep\}$, differential harm is in fact closely related to the overall number of seats won by each party, as Proposition \ref{prop:fairness} describes below.
-Yet unlike a simple measure of seats, differential harm is shifted to an interpretable scale where a value of zero corresponds to fairness.
-In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
-This connection further supports the use of total seats won by each party as a summary statistic in partisan redistricting simulation methods, as advocated by @deford2020symmetry.
-
-```{=tex}
-\begin{proprep} \label{prop:fairness}
-Suppose $\Groups=\{\dem,\rep\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
-\begin{align*}
-    \Delta_{\rep-\dem} H(q) = \E[\seats(\dem, V, \tilde Q)] - \E[\seats(\dem, V, q)].
-\end{align*}
-\textbf{CM note: define $\seats$ or replace with sum.}
-\end{proprep}
-\begin{proof}
-\end{proof}
-```
-# Estimating the Harm Measures
+# Estimating the Harm Measures {#sec:estimating}
 
@@
 We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting.
-Estimating differential racial harm in the two-party setting can be done by first generatinc ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
+Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
 Multiparty or primary elections pose additional modeling challenges.
@@
 
-# Connections to Existing Partisan and Racial Gerrymandering Metrics {#sec:connections}
+# Connections to Current Practice for Analyzing Partisan and Racial Gerrymandering {#sec:connections}
+
+While applicable to the study of redistricting beyond racial and partisan gerrymandering, the proposed harm measures provide an improvement over existing analytic approaches in both of these important contexts.
+
+## Partisan Gerrymandering
 
 The canonical framework for considering partisan fairness is partisan symmetry [@king1987democratic].
-Partisan symmetry implies that the seats, $s$ won by a party with voteshare $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
-Often, the partisan symmetry framework has been applied via partisan bias, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020].
-Partisan bias, denoted by $\beta(s)$ is the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
-To compute partisan bias, researchers have largely relied on assumptions about how the voteshares would swing to reach the alternative voteshares.
+Partisan symmetry implies that the seats, $s$ won by a party with vote share $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
+In practice, the partisan symmetry framework is applied via the partisan bias metric $\beta(s)$, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020], i.e., the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
 
 More recently, particularly in the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
-The most prominent of these new measures have attempted to offer alternative measures for partisan symmetry.
-The *mean-median* difference looks to relationship between the voteshare of the median seat in a delegation and the mean voteshare of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
+The *mean-median* difference looks to relationship between the vote share of the median seat in a delegation and the mean vote share of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
 The *declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
-Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015].
-Referred to by Chief Justice Roberts as "sociological gobbledygook" in Gill v. Whitford, the efficiency gap looks to measure the votes wasted by the parties and compare the net wasted votes.
-Notably, while the efficiency gap was intended to measure partisan symmetry, it does not do so [@katz2020].
-All of these measures may be estimated using district vote shares alone, and as we discuss next, are to varying degrees related to our proposed measures.
+Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015], which measures the difference in the number of the votes wasted by each party.[^3]
 
-Partisan symmetry offers a coherent and theoretically grounded quantity of interest.
-The idea of partisan symmetry is rooted in assessing how parties would have performed under counterfactual voteshares.
-Yet, this suffers from a core inflexibility: it does not allow for variation in districting.
-Districting is the fundamental question of district-based democracies.
-*For whom can you vote? Who do you vote with?* Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation.
-The natural measurement approach to this directly incorporates alternatives under the logic of a counterfactual.
+[^3]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.
 
-Consider the major issues that have been raised over existing measures.
-Partisan bias ($\beta(\half)$) can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.
-This has been called by some the Utah paradox [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
-A number of scholars have criticized the efficiency gap for (1) penalizing proportionality, (2) rewarding 3-to-1 districts and switching sign for more lopsided districts, and (3) being discrete and therefore sensitive to small changes in district vote shares [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
-These stem from the omission of the alternatives.
+The latter four metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in a larger theoretical framework or normative standard (in particular, they do not measure partisan symmetry [@katz2020]).
+As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
+This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^4]
+One advantage of the simpler metrics such as the efficiency gap is that they may quickly calculated using district vote shares alone.
 
-By looking to alternatives, we ground the idea of zero in a balance of the harms, rather than a purely theoretical way.
-This resolves the directionality problem common to other measures.
-The use of a counterfactual in measuring the quantity of interest grounds the bias in the rules of what districts can be drawn.
-This allows us to draw key strength from the impetus of the redistricting simulation literature.
-Yet, the simplicity of the framework does not restrict only to the use of simulation methods.
-Further, this framework does not need to rely on ideas of proportionality, which has no legal role in American democracy.
-Nor does it suffer from discreteness in ensembles, as now both voteshares and districts lines shift.
+[^4]: For partisan bias, both researchers and practitioners have largely relied on simple uniform-shift models similar to the one we use in this work.
+
+All of these existing measures are constructed so that a value of zero indicates no bias, and a value of zero is at least implicitly held up as a normative standard for districting plans.
+This poses a challenge for the measures which do not estimate a theoretical quantity, because it is unclear what the implicit standard is.
+For partisan symmetry this is more explicit; @katz2020 write that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric, if there is a fairer alternative, would not normally be regarded as a legitimate outcome."
+
+But as those authors recognize, in evaluating a districting plan, it is critical to consider "if there is a fairer alternative." The inability of partisan bias and other existing measures to account for the space of counterfactual redistricting plans means that none of them may be interpreted in a vacuum.
+No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the district-based electoral system.
+
+It is this confounding which explains many of the concerns that have been raised over the existing measures.
+The standard calculation of partisan bias, $\beta(\half)$, can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.[^5]
+And none of the measures take into account any aspect of political geography, and therefore suffer in states like Massachusetts, where, for example, the efficiency gap takes a value generally indicating an extreme Democratic gerrymander.
+
+[^5]: This has been called by some the "Utah paradox" [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
 
 Despite these differences, there are also key empirical similarities in many cases.
-The efficiency gap and $\Delta H(q)$ are often quite well correlated under competitive states, as wasted votes and partisan harm are rooted in similar logics.
-Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with $\E[\seats(\dem, V, Q)\mid Q=q]$ and thus $\Delta H(q)$ as well.
-Where our framework rises above these though is especially in non-competitive states, where insurmountable advantages are rendered as zeroes and do not contribute to harm.
+The efficiency gap and differential partisan harm are often highly correlated (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
+Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with differential partisan harm as well.
+Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section \ref{sec:nj}.
 
-# Application to Partisan Gerrymandering in New Jersey
+[^6]: The differences stem from the efficiency gap considering a vote to be wasted if it is in excess of the 50% threshold needed to win, while the harm framework does not penalize such votes, since every such voter is happily represented by their candidate of choice.
+    This definition of wasted votes is also at the root of other criticisms of the efficiency gap, which note its tendency to penalize proportionality and reward 3-to-1 districts [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
+
+Finally, it is worth comparing the proposed measures to the increasingly common application redistricting simulation techniques to cases of partisan gerrymandering.
+Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics.
+While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
+In the worst cases, this can lead to directionally wrong conclusions, as we will see in Section \ref{sec:nj}.
+
+These problems have led @deford2020symmetry, among others, to recommend evaluating the presence of a gerrymander simply by comparing the expected number of seats won under the plan versus the simulations We agree with this recommendation, as it turns out that in the partisan case, with $\Groups=\{\dem,\rep\}$, differential harm is in fact closely related to the overall number of seats won by each party.
+Proposition \ref{prop:fairness} records this formally below.
+However, unlike a simple measure of seats, differential harm is shifted a value of zero corresponds to fairness and rescaled to be directly interpretable as a rate of harm.
+In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
+
+```{=tex}
+\begin{proprep} \label{prop:fairness}
+Suppose $\Groups=\{\dem,\rep\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
+\begin{align*}
+    \Delta_{\rep-\dem} H(q) = \E[\seats(\dem, V, \tilde Q)] - \E[\seats(\dem, V, q)].
+\end{align*}
+\textbf{CM note: define $\seats$ or replace with sum.}
+\end{proprep}
+\begin{proof}
+\end{proof}
+```
+## Racial Gerrymandering
+
+**CHRIS: review please.**
+
+The study of racial gerrymandering is considerably more complicated than that of partisan gerrymandering, due to the difficulty of understanding racial voting preferences when individual votes are secret, and as such no simple formulas exist to measure the racial impacts of a districting plan.
+Indeed, much of the academic literature has focused on the statistical challenge of estimating vote preferences and turnout by race [@king1997solution; @greiner2006ecological; @imai2016improving].
+
+The modern analysis of racial gerrymandering is inseparable from the Voting Rights Act of 1965 (VRA).
+The VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
+Under the current framework established by @1986gingles, to do so, plaintiffs must prove that (a) the minority group is large and geographically compact enough to form a majority in some hypothetical district, (b) the minority group is "politically cohesive," and (c) the majority group can and does vote to deny the minority group its choice of candidate [@hood2018legal] (the so-called *Gingles* criteria).
+Criteria (a) is generally proven through a single example districting plan, while criteria (b) and (c) are shown using the statistical estimates of vote preferences and turnout by race.
+
+While the specific legal requirements for a vote dilution claim will obviously remain front-and-center for court challenges of racial gerrymanders, it is notable that the proposed harm measures unify all three *Gingles* criteria.
+Criteria (b) and (c) captures the misrepresentation of minority voters by a candidate they do not support, while (a) requires the existence of a counterfactual where this would not be the case; together, (a) and (c) establish that minority voters are harmed.
+But the proposed measures can also tackle trickier situations where the *Gingles* criteria may not be clearly met.
+
+One such case arises when there are several distinct minority racial groups with similar preferences.
+In these cases, some courts have found that (a) may be weakened to consider "coalition" districts that still allow minority voters overall to elect candidates of their choice, but there is no clear rule [@stephanopoulos2014south].
+Or when urban White voters have similar partisan preferences to minority voters, (b) and (c) may not be strictly satisfied, especially for two-party general elections.
+Some argue, however, that courts should still rely on the VRA in these cases to protect minority voters, especially in light of differing racial preferences in primary elections [@becker2021computational].
+
+In both these cases, the proposed measures will quantify the extent to which a challenged districting plan differentially harms minority voters, taking into account their geographic and political cohesion, and the individual preferences of every voter.
+Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.
+
+# Application to Partisan Gerrymandering in New Jersey {#sec:nj}
 
@@
 One such plan is shown in Figure \ref{fig:nj-maps}(d).
-For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^3] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
+For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^7] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
 The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
 
-[^3]: As in section \ref{sec:connections}, this is $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
+[^7]: As in section \ref{sec:connections}, this is $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
 
@@
 Differential partisan harm identifies each gerrymander as such, with a large negative value for the Democratic gerrymander and a large positive value for the Republican gerrymander.
-These values are relatively directly interpretable; for example, the differential harm score of $-0.221$ means that an average Democratic voter is 22.1% less likely to be misrepresented than their average Republican counterpartisan.
+These values are relatively directly interpretable; for example, the differential harm score of $-0.221$ means that an average Democratic voter is 22.1% less likely to be misrepresented than their average Republican counter-partisan.
 The efficiency gap and declination measures also correctly identify the gerrymanders, though they are much more discrete, and declination gives a much larger score in absolute value to the Democratic plan than to the Republican.
@@
 
-We next turn to litigation in Alabama, Merill v.
-Milligan, where civil rights groups sued Alabama arguing that a second Black opportunity-to-elect district should have been drawn.
-For litigation under Section 2 of the Voting Rights Act of 1965 (VRA), plaintiffs submit one or more remedial maps in order to meet the preconditions set under Thornburg v.
-Gingles (478 US 30) in 1986.
+We next turn to litigation in Alabama, @2022merrill, where civil rights groups sued Alabama arguing that a second Black opportunity-to-elect district should have been drawn.
+For litigation under Section 2 of the Voting Rights Act of 1965 (VRA), plaintiffs submit one or more remedial maps in order to meet the preconditions set under @1986gingles.
 Remedial maps demonstrate that a new plan could be drawn to comply with the VRA.
@@
 Figure \ref{fig:al-maps} displays the results of this process.
-Panels (a) and (b) of \ref{fig:al-maps} display the partisanship and Black voting age percentages of Alabama.
-Unsuprisingly, Black population and Democratic voteshare are highly correlated.
+Figures \ref{fig:al-maps}(a) and (b) display the partisanship and Black voting age percentages of Alabama.
+Unsurprisingly, Black population and Democratic vote share are highly correlated.
 The remedial plans create a Black opportunity district by connecting Mobile to Montgomery, via the Black Belt region.
-Panel (c) of \ref{fig:al-maps} highlights the total harm from the enacted Alabama plan against the remedial plans.
+Figure \ref{fig:al-maps}(c) highlights the total harm from the enacted Alabama plan against the remedial plans.
 As we would expect, the harm largely follows the Black belt, where the Black opportunity district was drawn in each of the remedial plans.
@@
 Geographically, we can see where voters are harmed.
-For example, in the north of Alabama, near Huntville, despite a larger number of Democratic voters in a Republican district, none of them are harmed here.
+For example, in the north of Alabama, near Huntsville, despite a larger number of Democratic voters in a Republican district, none of them are harmed here.
 That is, they are also in Republican districts in the remedy plans.
 Along the entirety of the Black Belt, there is substantial harm, most of which comes from Black Democrats.
-There is also a pocket of harmed voters in Tuskaloosa, above the left shoulder of district 7.
+There is also a pocket of harmed voters in Tuscaloosa, above the left shoulder of district 7.
 
@@
 Yet, on average, only 8,022 Republicans are harmed by the enacted plan.
-After normalizing by group sizes, this gives a substantial  differential harm of 20.9%.
+After normalizing by group sizes, this gives a substantial differential harm of 0.209.
 Thus, compared to the remedy plans, the enacted plan is much more likely to harm Democrats than Republicans.
@@
 In comparison, only 147 Black Republicans, 7,867 white Republicans, and 7 (8?) other race Republicans are harmed on average.
-Thus, once normalized by group size, the differential racial harm is 23.7%, indicating that Black voters are harmed substantially more than white voters.
+Thus, once normalized by group size, the differential racial harm is 0.237, indicating that Black voters are harmed at a substantially higher rate than white voters.
 
@@
 Nine of the councillors are elected from contiguous city council districts, and the remaining four councillors are elected at-large, with all councillors serving two-year terms.
-Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^4]
+Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^8]
 This section will study the extent to which the mixed at-large and district council system harms voters in comparison to a fully district-based system (as was the case before 1909 and from 1924-1951), either overall or differentially by racial group.
@@
 
-[^4]: City of Boston charter.
+[^8]: City of Boston charter.
     Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
@@
 #| fig.cap: |
-#|      Candidate factor loadings, with approximate 80% credible intervals shown as ellipses (left),
-#|      and geographic variation in factor values by race (right).
+#|      Candidate factor loadings, with approximate 80\% credible intervals shown as
+#|      ellipses (left), and geographic variation in factor values by race (right).
 #|      Neighborhood boundaries overlaid in black.
@@
 
-To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^5] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
+To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^9] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
 Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail and contains additional diagnostic plots.
 
-[^5]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+[^9]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
 
@@
 
-In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^6]
+In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^10]
 Throughout this appendix, let $j$ index the set of electoral precincts or other units.
@@
 
-[^6]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+[^10]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
     Population information comes from Census blocks, which are practically never split.
@@
 
-The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^7]
+The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^11]
 This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
@@
 
-[^7]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
+[^11]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
 
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^8]
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^12]
 Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
@@
 
-[^8]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+[^12]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
 

### 2022-05-09 paper/paper.Rmd
redistricting -> districting

@@
 
--   **we argue that individual standing/harm is an excellent lens to unify the study and measurement of redistricting plans**
+-   **we argue that individual standing/harm is an excellent lens to unify the study and measurement of districting plans**
 
 -   core idea: counterfactuals + voter choice.
-    Is a voter more likely to be represented by the candidate of their choice under a different redistricting plan?
+    Is a voter more likely to be represented by the candidate of their choice under a different districting plan?
     If so, they've been harmed.
@@
 If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
-In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined; there is certainly an element of subjectivity in sampling a set of redistricting plans.
+In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined; there is certainly an element of subjectivity in sampling a set of districting plans.
 Nonetheless, sampling methods are widely used in practice.
@@
 
-But as those authors recognize, in evaluating a districting plan, it is critical to consider "if there is a fairer alternative." The inability of partisan bias and other existing measures to account for the space of counterfactual redistricting plans means that none of them may be interpreted in a vacuum.
+But as those authors recognize, in evaluating a districting plan, it is critical to consider "if there is a fairer alternative." The inability of partisan bias and other existing measures to account for the space of counterfactual districting plans means that none of them may be interpreted in a vacuum.
 No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the district-based electoral system.

### 2022-05-09 paper/paper.Rmd
cleanup

@@
 
-These problems have led @deford2020symmetry, among others, to recommend evaluating the presence of a gerrymander simply by comparing the expected number of seats won under the plan versus the simulations We agree with this recommendation, as it turns out that in the partisan case, with $\Groups=\{\dem,\rep\}$, differential harm is in fact closely related to the overall number of seats won by each party.
+These problems have led @deford2020symmetry, among others, to recommend evaluating the presence of a gerrymander simply by comparing the expected number of seats won under the plan versus the simulations We agree with this recommendation, as it turns out that in the partisan case, differential harm is in fact closely related to the overall number of seats won by each party.
 Proposition \ref{prop:fairness} records this formally below.

### 2022-05-09 paper/paper.Rmd
edit

@@
 Proposition \ref{prop:fairness} records this formally below.
-However, unlike a simple measure of seats, differential harm is shifted a value of zero corresponds to fairness and rescaled to be directly interpretable as a rate of harm.
-In other words, differential harm provides a connection between fairness and seats, establishing a normative guide to overall election outcomes.
+However, unlike a simple measure of seats, differential harm is shifted so that a value of zero corresponds to fairness, and rescaled to be directly interpretable as a fraction of voters harmed.
 

### 2022-05-09 paper/paper.Rmd
AL revision

@@
 bibliography: references.bib
-biblio-style: apalike
+biblio-style: chicago
 output:
@@
 
-```{=html}
-<!-- Yet, this suffers from a core inflexibility: it does not allow for variation in districting.
+<!--# Yet, this suffers from a core inflexibility: it does not allow for variation in districting.
 Districting is the fundamental question of district-based democracies.
-*For whom can you vote? Who do you vote with?* Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation.
-The natural measurement approach to this directly incorporates alternatives under the logic of a counterfactual. -->
-```
+For whom can you vote?
+Who do you vote with?
+Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation.
+The natural measurement approach to this directly incorporates alternatives under the logic of a counterfactual.
+-->
+
 ## Outline of the Paper
@@
 
-[^1]: See, e.g., the historical distribution calculated by the Campaign Legal Center's PlanScore tool: [planscore.campaignlegal.org/metrics/efficiencygap/](https://planscore.campaignlegal.org/metrics/efficiencygap/).
+[^1]: See, e.g., the historical distribution calculated by the Campaign Legal Center's PlanScore tool: <https://planscore.campaignlegal.org/metrics/efficiencygap/>.
 
@@
 Motivated primarily by computational convenience, we use a *proportionate uniform shift* model, which focuses on swings in the national political environment as the main determinant of election outcomes.
-Each election outcome over the period of interest (generally ten years or less) is modeled as a random deviation from a baseline outcome, and these deviations are assumed to consist of a constant fraction of voters of one party in each precinct switching to the other.
+Election outcomes are modeled as a random deviation from a baseline outcome, and these deviations are assumed to consist of a constant fraction of voters of one party in each precinct switching to the other.
 The baseline itself can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
@@
 Partisan symmetry implies that the seats, $s$ won by a party with vote share $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
-In practice, the partisan symmetry framework is applied via the partisan bias metric $\beta(s)$, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020], i.e., the number of seats that would need to be reallocated to reach proportionality for a given statewide vote share $s$ (usually $\half$).
+In practice, the partisan symmetry framework is applied via the partisan bias metric $\beta(s)$, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020], i.e., the number of seats that would need to be reallocated to reach partisan symmetry for a given statewide vote share $s$ (usually $\half$).
 
-More recently, particularly in the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
+More recently, and particularly in the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
 The *mean-median* difference looks to relationship between the vote share of the median seat in a delegation and the mean vote share of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
@@
 [^5]: This has been called by some the "Utah paradox" [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
+    Even when interpreted carefully, however, partisan bias requires significant extrapolation from observed election outcomes in lopsided states---how realistic is it to evaluate the counterfactual where Democrats run even with Republicans statewide in Utah?
 
@@
 ```
-## Racial Gerrymandering
+## Racial Gerrymandering {#subsec:racial}
 
@@
 New Jersey, with an average Democratic two-party vote share of 56.6% across our two baseline elections (2016 presidential and 2018 senatorial), will demonstrate that this is not a problem unique to lopsided states.
-Additionally, New Jersey has 12 congressional districts, which is enough to produce some variability in seats outcomes across counterfactual plans, but not so many as to render trivial the challenges of discreteness.
 
@@
 
-Differential partisan harm identifies each gerrymander as such, with a large negative value for the Democratic gerrymander and a large positive value for the Republican gerrymander.
+We first examine the values of each metric on the two example plans alone.
+Differential partisan harm correctly identifies each gerrymander as such, with a large negative value for the Democratic gerrymander and a large positive value for the Republican gerrymander.
 These values are relatively directly interpretable; for example, the differential harm score of $-0.221$ means that an average Democratic voter is 22.1% less likely to be misrepresented than their average Republican counter-partisan.
@@
 
-Considering the distribution of these statistics across the distribution of sampled plans doesn't help much, either.
+Next, we consider the distribution of these statistics across the sampled plans, but this doesn't help much, either.
 The expected number of Democratic seats naturally identifies the gerrymanders (two-sided $p=0.0002$) and the direction of bias, and as expected from Proposition \ref{prop:fairness} this measure is highly correlated with differential partisan harm.
@@
 
-We next turn to litigation in Alabama, @2022merrill, where civil rights groups sued Alabama arguing that a second Black opportunity-to-elect district should have been drawn.
-For litigation under Section 2 of the Voting Rights Act of 1965 (VRA), plaintiffs submit one or more remedial maps in order to meet the preconditions set under @1986gingles.
-Remedial maps demonstrate that a new plan could be drawn to comply with the VRA.
-We use four remedial plans submitted by plaintiff's expert, Moon Duchin, which each create a moderately compact, second Black opportunity district while complying with other state redistricting criteria.
-We opt to use remedial plans, as they offer a relevant alternative set of plans ($\tilde{Q}$) to which courts already look when establishing liability.
+We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights have argued that the existing Black majority district packs Black voters, and a second Black opportunity-to-elect district should be drawn.
 
-Figure \ref{fig:al-maps} displays the results of this process.
-Figures \ref{fig:al-maps}(a) and (b) display the partisanship and Black voting age percentages of Alabama.
-Unsurprisingly, Black population and Democratic vote share are highly correlated.
-The remedial plans create a Black opportunity district by connecting Mobile to Montgomery, via the Black Belt region.
-Figure \ref{fig:al-maps}(c) highlights the total harm from the enacted Alabama plan against the remedial plans.
-As we would expect, the harm largely follows the Black belt, where the Black opportunity district was drawn in each of the remedial plans.
+As discussed in Section \ref{subsec:racial}, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect plan could be drawn while meeting all other redistricting requirements.
+In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure \ref{fig:al-maps}(e).
+These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure \ref{fig:al-maps}(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.[^8]
 
-```{r al-maps, fig.width=6.5}
+[^8]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
+
+This analysis is slightly more complicated than in New Jersey, since in order to understand the racial distribution of harm we need to estimate voting preferences by race.
+To do so, we use an ecological inference model which jointly estimates turnout and vote choice by race and precinct.
+The model and its outputs are detailed in Appendix \ref{app:alei}.
+Unsurprisingly, voting is highly racially polarized outside of the cities of Birmingham and Montgomery.
+
+With these estimates in hand, we can calculate the average number of voters harmed in each precinct and racial group.
+Figure \ref{fig:al-maps}(c) shows the geographic distribution of harm.
+Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures \ref{fig:al-maps}(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
+It is precisely these areas that are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.
+
+```{r al-maps, fig.width=6.75}
 #| fig.cap: |
-#|      Baseline vote patterns (a),
+#|      Top row: baseline vote patterns (a),
 #|      fraction of the voting-age population which is Black (b),
-#|      and the expected fraction of voters in each precinct which are harmed by the
-#|      2020 enacted congressional districts versus the set of four remedial plans (c).
+#|      the expected fraction of voters in each precinct which are harmed by 
+#|      the 2020 enacted congressional districts versus the set of four remedial 
+#|      plans (c).
 #|      2020 enacted district boundaries are overlaid
-#|      in black on the left map and in white on the right maps.
+#|      in black on the leftmost map and in white on the other maps.
+#|      Bottom row: the challenged plan and a submitted remedial plan 
+#|      (d) and (e) colored by each district's Black share of the voting-age
+#|      population, and the expected number of harmed voters in each 
+#|      race/partisan group, relative to the total number of such voters (f).
 include_graphics("figures/al_maps.pdf")
@@
 
-First, this gives us that the average total harmed is 162,390 people.
-This is equivalent to a total harm of 8.9%.
-Geographically, we can see where voters are harmed.
-For example, in the north of Alabama, near Huntsville, despite a larger number of Democratic voters in a Republican district, none of them are harmed here.
-That is, they are also in Republican districts in the remedy plans.
-Along the entirety of the Black Belt, there is substantial harm, most of which comes from Black Democrats.
-There is also a pocket of harmed voters in Tuscaloosa, above the left shoulder of district 7.
+This intuition is confirmed by a closer analysis, summarized in Figure \ref{fig:al-maps}(f).
+Across the state, the average harm is 0.075, which translates to around 151,000 voters.
+Of these harmed voters, around 128,000 are Black, while 23,000 are White.
+Similarly, 140,000 are Democratic, while only 11,000 are Republican.
+Of course, the total number of each of these groups of voters is very different.
+Normalizing by group sizes, the expected harm for a Black Democrat is 0.23, for a Black Republican is 0.026, for a White Democrat is 0.084, and for a White Republican is 0.0082.
+Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have a differential harm of 0.149 versus White Democrats and a differential harm of 0.207 versus Black Republicans.
 
-We can then break down the harm into components by party.
-We find that an average of 154,368 Democrats are harmed by the enacted plan.
-Yet, on average, only 8,022 Republicans are harmed by the enacted plan.
-After normalizing by group sizes, this gives a substantial differential harm of 0.209.
-Thus, compared to the remedy plans, the enacted plan is much more likely to harm Democrats than Republicans.
-
-Taking this further, we can estimate the harm by race as well.
-To do so, we fit an ecological inference model to estimate party-by-race counts for each precinct.
-Using the estimates, we can see that the harm to Democrats is carried largely by Black Democrats over non-Black Democrats.
-Specifically, 125,635 Black Democrats are harmed, compared to 28,476 white Democrats and 257 other race Democrats.
-In comparison, only 147 Black Republicans, 7,867 white Republicans, and 7 (8?) other race Republicans are harmed on average.
-Thus, once normalized by group size, the differential racial harm is 0.237, indicating that Black voters are harmed at a substantially higher rate than white voters.
+These calculations are made by aggregating harm within racial groups at the precinct level, and so they automatically take into account individual voter preferences.
+Unlike in a traditional EI/*Gingles* analysis, there is no need to identify a particular party as the statewide "choice" of the minority racial group, nor is there any cause for concern because of lower racial voting polarization in urban areas.
+The harm measure treats voters of every race and location equally, recording just their inability to elect their preferred candidate.
+Only once we have aggregated these individual harms does it become apparent that the challenged plan concentrates the harm among certain groups of voters.
 
@@
 Nine of the councillors are elected from contiguous city council districts, and the remaining four councillors are elected at-large, with all councillors serving two-year terms.
-Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^8]
+Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^9]
 This section will study the extent to which the mixed at-large and district council system harms voters in comparison to a fully district-based system (as was the case before 1909 and from 1924-1951), either overall or differentially by racial group.
@@
 
-[^8]: City of Boston charter.
+[^9]: City of Boston charter.
     Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
@@
 
-To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^9] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
+To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^10] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
 Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail and contains additional diagnostic plots.
 
-[^9]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+[^10]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
 
@@
 
-In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^10]
+In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^11]
 Throughout this appendix, let $j$ index the set of electoral precincts or other units.
@@
 
-[^10]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+[^11]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
     Population information comes from Census blocks, which are practically never split.
@@
 
-The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^11]
+The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^12]
 This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
@@
 
-[^11]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
+[^12]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
 
@@
 \gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^12]
+$$ which is the inverse CDF of a Laplace distribution with rate 2.[^13]
 Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
@@
 
-[^12]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
+[^13]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
 
@@
 
+# Ecological Inference Model for Alabama {#app:alei}
+
 # Latent Factor Ecological Inference Model for Boston Primary Elections {#app:lfei}

### 2022-05-09 R/00_setup_utils.R

@@ -78,6 +78,12 @@ plot_cds = function(map, pl, county, abbr, qty="ndv") {
         rvote = map$nrv
         qty = expr(dem)
         scale = scale_fill_party_c(limits=c(0.3, 0.7))
+    } else if (qty == "bvap") {
+        dvote = map$vap_black
+        rvote = map$vap - map$vap_black
+        qty = expr(dem)
+        scale = scale_fill_wa_b("sea", name="BVAP", labels=percent,
+                                limits=c(0.1, 0.6), breaks=c(0.2, 0.3, 0.4, 0.5))
     } else {
         qty = expr(.plan)
         scale = scale_fill_manual(values=PAL, guide="none")
@@ -97,6 +103,7 @@ plot_cds = function(map, pl, county, abbr, qty="ndv") {
         st_as_sf() %>%
         group_by(.distr) %>%
         summarize(.plan = .plan[1],
+                  dem = 1 / (1 + sum(rvote) / sum(dvote)),
                   dem = 1 / (1 + sum(rvote) / sum(dvote)),
                   is_coverage=TRUE) %>%
         ggplot(aes(fill={{ qty }})) +
@@ -121,6 +128,7 @@ expl_vars = function(pl, labels, refs=character(0), rasterize=TRUE,...) {
         y <- h$counts; y <- y/max(y)
         rect(breaks[-nB], 0, breaks[-1], y, family="Times",
              col="#888888", border="#88888800", lwd=0)
+        abline(v=0, lty="dashed")
         for (i in seq_len(n_ref)) {
             abline(v=x[i], col=refs[i], lwd=3.0)
         }

### 2022-05-09 R/01_metrics.R

@@ -71,7 +71,7 @@ calc_plans_stats = function(plans, map, dem, gop, ker=k_t(), max_harm=200L) {
                comp = distr_compactness(map),
                dem = group_frac(map, dvote, dvote+rvote),
                decl = part_decl(., map, dvote, rvote),
-               egap = part_egap(., map, rvote, dvote))
+               egap = part_egap(., map, dvote, rvote))
 
     ker_t = k_t()
     pl_sum = plans %>%

### 2022-05-09 R/06_alabama.R

@@ -7,8 +7,7 @@ if (!file.exists(path <- here("data/AL_cd_final_vtd_20.rds"))) {
     al_map = read_rds(path)
 }
 
-
-d_votes = d %>%
+d_votes = al_map %>%
     as_tibble() %>%
     mutate(vap_other = vap - vap_white - vap_black) %>%
     select(GEOID20, pre_16_rep_tru:uss_20_dem_jon) %>%
@@ -20,7 +19,7 @@ d_votes = d %>%
            elec = fct_inorder(elec))
 tbls = make_tables_alabama(d_votes, al_map)
 
-fit = fit_ei(tbls, algorithm="vb", init=0, eta=0.1, adapt_engaged=F, tol_rel_obj=0.002)
+fit = fit_ei(tbls, algorithm="vb", init=0, eta=0.1, adapt_engaged=F, tol_rel_obj=0.0002)
 
 #  EI outputs--------
 cat("Turnout correlation matrix:\n")
@@ -34,6 +33,41 @@ p2 = plot(al_map, median(fit$turnout)[, "black"]) + scale_fill_wa_c(name="Turnou
 p3 = plot(al_map, median(fit$support)[, "white"]) + scale_fill_party_c() + labs(title="White Dem. support")
 p4 = plot(al_map, median(fit$support)[, "black"]) + scale_fill_party_c() + labs(title="Black Dem. support")
 p1 + p2 + p3 + p4 + plot_layout(nrow=1, guides="collect") & theme_repr_map() & theme(legend.position="bottom")
+
+
+# appendix output
+d_est = bind_cols(
+    rename_with(as_tibble(median(fit$turnout)), ~ str_c("turn_", .)),
+    rename_with(as_tibble(median(fit$support)), ~ str_c("supp_", .)),
+    as_tibble(tbls$race)
+) %>%
+    mutate(vap = tbls$vap)
+
+
+p1 = ggplot(d_est, aes(turn_white, turn_black, size=vap, color=black)) +
+    geom_point() +
+    scale_x_continuous("White turnout", trans="logit", labels=percent,
+                       breaks=c(0.1, 0.2, 0.4, 0.6, 0.8, 0.9)) +
+    scale_y_continuous("Black turnout", trans="logit", labels=percent,
+                       breaks=c(0.1, 0.2, 0.4, 0.6, 0.8, 0.9)) +
+    scale_color_wa_c("sea", name="BVAP", labels=percent, limits=c(0, 1)) +
[…]

### 2022-05-09 paper/paper.Rmd
AL appendix

@@
 
+To estimate voter preferences by race and precinct in Alabama, we use 2020 U.S.
+Census demographic data and election returns from seven recent statewide elections: the 2016 races for president and U.S.
+senator, the 2018 races for governor, attorney general, and secretary of state, and the 2020 races for president and U.S.
+senator.
+Our ecological inference model used to produce the estimates is broadly similar to that of @king1997solution.
+However, it jointly estimates both turnout and Democratic support across all seven elections, rather than estimating each quantity separately for each election.
+
+Adapting the notation from the previous appendix, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $\votes_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jv}$ denote the Democratic vote share, and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Other}\}$.
+Our main parameters of interest are $\turn_{jr}$ and $\supp_{jr}$, the baseline turnout and Democratic support rate by race in the precinct.
+
+```{r alabama-ei, fig.width=6.5}
+#| fig.cap: |
+#|      EI model estimates of precinct-level turnout by race (a) 
+#|      and Democratic support by race (b).
+#|      Points are sized in proportion to their voting-age population, 
+#|      and colored according to their Black share of the voting-age population.
+#|      All axes are transformed to a logit scale.
+include_graphics("figures/al_ei.pdf")
+```
+
+We model the observed vote counts as
+\begin{align*}
+    \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
+        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
+    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
+        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
+    s_{jv}\votes_{jv} \mid \supp_{jr}, \turn_{jr}, \beta^{(s)}_v
+        &\sim \Binlog\qty(\votes_{jv}, \logit\qty(
+        \supp_j^\top \frac{\turn_j\circ\vb p_j}{\turn_j^\top\vb p_j}) + \beta^{(s)}_v) \\
+    \logit(\supp_j) \mid \mu^{(s)}, \Sigma^{(s)}
+        &\sim \Norm(\mu^{(s)}, \Sigma^{(t)}),
+\end{align*}
+where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability paramter specified on the logit scale, and $\circ$ indicates the elementwise (Hadamard) product.
+We give $\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior,
+and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior, and $\mu^{(s)}$ an i.i.d. $\Norm(0, 1.0^2)$ prior.
+We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$; the former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).
+Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.
+
+The model was fitted with a mean-field variational approximation for computational efficiency.
+Posterior medians of $\turn_{jr}$ and $\supp_{jr}$ are plotted in Figure \ref{fig:alabama-ei}.
+These estimates broadly track with the findings of the voting rights expert retained in *Merrill*.
+
 # Latent Factor Ecological Inference Model for Boston Primary Elections {#app:lfei}
@@
 #| fig.cap: |
-#|      Model predictions of general election vote shares and turnout, compared to actual general election results.
+#|      Model predictions of general election vote shares and turnout, 
+#|      compared to actual general election results.
 #|      Red reference line indicates perfect agreement.
-#|      Actual and estimated results have Spearman correlations of 0.87 (vote share) and 0.84 (turnout rate).
+#|      Actual and estimated results have Spearman correlations of 
+#|      0.87 (vote share) and 0.84 (turnout rate).
 include_graphics("figures/boston_gen_pred.pdf")

### 2022-05-10 paper/paper.Rmd
push
dorchester orientation

@@
 These calculations are made by aggregating harm within racial groups at the precinct level, and so they automatically take into account individual voter preferences.
-Unlike in a traditional EI/*Gingles* analysis, there is no need to identify a particular party as the statewide "choice" of the minority racial group, nor is there any cause for concern because of lower racial voting polarization in urban areas.
+Unlike in a traditional ecological/*Gingles* analysis, there is no need to identify a particular party as the statewide "choice" of the minority racial group, nor is there any cause for concern because of lower racial voting polarization in urban areas.
 The harm measure treats voters of every race and location equally, recording just their inability to elect their preferred candidate.
@@
 #| fig.cap: |
-#|      Racial composition of the city of Boston (left),
+#|      Racial composition of the city of Boston (left)
 #|      and 2010 city coucil districts (right).
@@
 #| fig.cap: |
-#|      EI model estimates of precinct-level turnout by race (a) 
+#|      Model estimates of precinct-level turnout by race (a) 
 #|      and Democratic support by race (b).
@@
 
-We model the observed vote counts as
-\begin{align*}
+We model the observed vote counts as \begin{align*}
     \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
@@
         &\sim \Norm(\mu^{(s)}, \Sigma^{(t)}),
-\end{align*}
-where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability paramter specified on the logit scale, and $\circ$ indicates the elementwise (Hadamard) product.
-We give $\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior,
-and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior, and $\mu^{(s)}$ an i.i.d. $\Norm(0, 1.0^2)$ prior.
+\end{align*} where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability paramter specified on the logit scale, and $\circ$ indicates the elementwise (Hadamard) product.
+We give $\beta^{(t)}$ an i.i.d.
+$\Norm(0, 0.6^2)$ prior,and $\beta^{(s)}$ an i.i.d.
+$\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d.
+$\Norm(0, 1.5^2)$ prior, and $\mu^{(s)}$ an i.i.d.
+$\Norm(0, 1.0^2)$ prior.
 We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$; the former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).

### 2022-05-10 R/07_boston.R

@@ -17,9 +17,9 @@ d_votes =  d %>%
     mutate(precinct = fct_inorder(precinct),
            elec = fct_inorder(elec),
            candidate = fct_inorder(candidate))
-tbls = make_tables_boston(d_votes, cand_thresh=0.02)
+tbls = make_tables_boston(d_votes, cand_thresh=0.01)
 
-if (!file.exists(fit_path <- here("data-raw/boston_fit_2.rds"))) {
+if (!file.exists(fit_path <- here("data-raw/boston_fit.rds"))) {
     id = list(
         loc = list(sanders=c(-1, 0), biden=c(1, 0), janey=c(-0.25, -1)),
         scale = list(sanders=c(0.1, 0.25), biden=c(1, 0.5), janey=c(0.5, 2)),
@@ -75,15 +75,31 @@ p1 = ggplot(map, aes(fill=maj_grp, alpha=maj_share)) +
           legend.justification=c(1, 0),
           legend.direction="horizontal")
 
-set.seed(123)
-p2 = ggplot(map, aes(fill=sample(1:9)[ccd_2010])) +
-    geom_sf(size=0, color=NA, alpha=0.7) +
+d_distr = st_drop_geometry(map) %>%
+    as_tibble() %>%
+    group_by(ccd_2010) %>%
+    summarize(across(starts_with("vap"), sum)) %>%
+    mutate(vap_other = vap - vap_white - vap_black - vap_hisp) %>%
+    select(ccd_2010, vap, vap_white:vap_hisp, vap_other) %>%
+    mutate(across(vap_white:vap_other, ~ . / vap)) %>%
+    rowwise() %>%
+    transmute(ccd_2010=ccd_2010,
+              maj_grp = str_sub(colnames(.)[3:7][
+                  which.max(c_across(vap_white:vap_other))], 5),
+              maj_share = max(c_across(vap_white:vap_other)))
+
+p2 = map %>%
+    select(-maj_grp, -maj_share) %>%
+    left_join(d_distr, by="ccd_2010") %>%
+ggplot(aes(fill=maj_grp, alpha=maj_share)) +
+    geom_sf(size=0, color=NA) +
     geom_sf(data=geom_dist, inherit.aes=F, fill=NA, size=0.5, color="black") +
[…]

### 2022-05-11 paper/paper.Rmd
finish drafting boston main text

@@
 This analysis is slightly more complicated than in New Jersey, since in order to understand the racial distribution of harm we need to estimate voting preferences by race.
-To do so, we use an ecological inference model which jointly estimates turnout and vote choice by race and precinct.
-The model and its outputs are detailed in Appendix \ref{app:alei}.
+As is custom in VRA analyses, we use an ecological inference model.
+The model, detailed in Appendix \ref{app:alei}, jointly estimates turnout and vote choice by race and precinct across seven statewide elections.
 Unsurprisingly, voting is highly racially polarized outside of the cities of Birmingham and Montgomery.
@@
 Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^9]
-This section will study the extent to which the mixed at-large and district council system harms voters in comparison to a fully district-based system (as was the case before 1909 and from 1924-1951), either overall or differentially by racial group.
+This section will study the extent to which the current district boundaries differentially harm voters of different racial groups, in comparison to a completely race-neutral districting process.
 Figure \ref{fig:boston-map} shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.
@@
 
-To apply our proposed framework to study the council system and and its district boundaries, we need to be able to predict individual voting preferences under counterfactual districting plans and systems.
-This is more difficult than in our study of New Jersey, for two reasons.
+To apply our proposed framework to study the council district boundaries, we need to be able to predict individual voting preferences under counterfactual districting plans.
+This is more difficult than in both New Jersey and Alabama, for two reasons.
 First, the city council is officially nonpartisan and city politics are dominated by the Democratic Party, so voting preferences are more complex than a simple partisan preference.
 To estimate election outcomes, we will need some way of mapping observed support for particular candidates in one election onto counterfactual support for a hypothetical candidate in another election.
-Second, we aim to examine the racial impact of the district system, and thus need to estimate these voting preferences by race.
+Second, as in Alabama, we aim to examine the racial impact of the district system, and thus need to estimate these voting preferences by race.
 Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone tell us how different racial groups voted.
@@
 Yet we believe that the model is a useful simplification of reality that captures the major electoral dynamics that are relevant to our question of interest.
-For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates that advanced to the general election, we were able to predict the precinct-level general election turnout and vote shares with considerable accuracy (see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
+For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates that advanced to the general election, we were able to predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
 We stress that the latent factor ecological inference model is not necessary for this kind of analysis; local survey data that collected ideological preferences and racial information would also suffice.
@@
 ## Results
+
+As Figure \ref{fig:boston-factors} shows, primary voting preferences in Boston are polarized by race, especially for the second latent dimension.
+But there is significant variation within racial groups as well, such as the split between more conservative White voters in South Boston and more liberal White voters in Jamaica Plain (the first latent dimension appears to capture a progressive-moderate divide).
+It is certainly possible, then that the particular way districts are drawn will have an effect on the ability of different racial groups to elect their candidate of choice.
+
+We consider two different scenarios: one in which all of the council races are contested along the first latent dimension, and one in which the races are contested along the second.
+Specifically, we draw voter choices from the posterior predictive distribution of our model, setting the candidates' loadings to --1 and 1 on the chosen dimension and 0 on the other dimension.
+We draw the overall level of support from a $\Beta(15, 15)$ distribution, which means that we will study election scenarios where the overall vote share for each bloc of candidates ranges from roughly 30 to 70 percent.
+The posterior predictive draws yield counts of voters for each candidate in each precinct by race, from which we can calculate winners at the district level, under any configuration of districts.
+
+We sample 2,000 redistricting plans across two independent runs of the SMC algorithm of @mccartan2020, setting constraints to favor compact plans that respect neighborhood boundaries.
+Since we do not provide any racial information to the algorithm, this set of plans forms a good approximation of a race-neutral districting process.
+
+For the first scenario, where council races are contested along the ideological axis, we find that the enacted plan is quite fair, differentially harming White voters at a rate of only 0.0079 compared to the race-neutral baseline.
+For the second scenario, where the contests are oriented along the racial dimension, we estimate that the enacted plan differentially harms White voters at a rate of 0.041, compared to the race-neutral baseline.
+
+We can also flip the setup and compute harm by averaging over the simulated plans and using the enacted plan as a counterfactual.
+This will produce a different answer, since harm is not symmetric between the study plan and the counterfactual.
+Along the first dimension, a race-neutral baseline would differentially harm Black voters at a rate of 0.031 compared to the enacted plan, while along the second dimension, the differential harm would be 0.018.
+
+Taken together, these findings suggest that despite the dominance of the Democratic party in city politics, and the shared preferences between certain groups of Black and White voters, race-aware districting still serves an important role in empowering Black voters across the city to elect their chosen candidates.
+Further study could focus on the racial impact of the four at-large members in comparison to a system with 13 districts and no at-large members.
 

### 2022-05-16 paper/paper.Rmd
boston appendix

@@
 
+To estimate voter preferences by race and precinct in Boston, we use 2020 U.S.
+Census demographic data and election returns from the March 2020 presidential primary and the September 2021 mayoral preliminary elections.
+Our ecological inference model used to produce the estimates expands on the model in Appendix \ref{app:alei}, jointly estimating both turnout and vote choice in the two elections.
+However, it estimates voter preferences in a two-dimensional latent space and also estimates candidates' positions in this space.
+
+Continuing the notation from above, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $\votes_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jk}$ denote the vote share for candidate $k$ (who implicitly belongs to a particular election $v_k$), and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Hispanic},\text{Other}\}$ ("Other" consisting primarily of Asian voters in Boston).
+Our main parameters of interest are $\turn_{jr}$, the baseline turnout by race and precinct, $\pref_{jqr}$, the preferences in latent dimension $q$ by race and precinct, and $L_{kq}$, the loading of each candidate on latent dimension $q$.
+Here $q\in\{1,2\}$, since preliminary analyses found no improved fit when expanding to three dimensions.
+
 ```{r boston-gen-pred, fig.width=6.5}
@@
 #|      Red reference line indicates perfect agreement.
-#|      Actual and estimated results have Spearman correlations of 
-#|      0.87 (vote share) and 0.84 (turnout rate).
 include_graphics("figures/boston_gen_pred.pdf")
 ```
+
+We model the observed vote counts as \begin{align*}
+    \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
+        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
+    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
+        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
+    \logit(s_{jk}) \mid p_{jk}, \eps &\sim \Norm(p_{jk}, \frac{\eps}{1+\votes_{jv_k}}) \\ 
+    p_{jk} &= \beta^{(s)}_k + \alpha_k
+        \pref_{j}^\top(\turn_j\circ \vb p_j)\mathrm{diag}(\vb*\gamma)\vb L_k  \\
+    \pref_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
+\end{align*} Note that $\pref_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $\pref_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
+The matrix $\mathrm{diag}(\vb*\gamma)$ captures the relative scale or importance of each dimension to the overall outcomes.
+
+Because of the latent factor model, we do not require that vote shares by race and precinct, when tallied by race, necessary equal the observed vote share in the precinct.
+As a result we choose to model the vote shares directly, which is more computationally efficient.
+Since these vote shares come from Binomial observations, they have variance inversely proportional to the total number of votes in the precinct (the additional 1 is to avoid divide-by-zero errors).
+The multiplicative factor $\eps$ captures the additional variance that isn't accounted for by the latent factors.
+The $\vb*\alpha$ parameters are "fudge factors" that adjust the scale of the deviations from each candidate's average support.
+
+We give $\beta^{(t)}$ an i.i.d. $\Norm(0, 0.5^2)$ prior,and $\beta^{(s)}$ an i.i.d. $\Norm(0, 4.0^2)$ prior, and $\logit(\mu^{(t)})$ an i.i.d. $\Beta(10, 30)$ prior.
+We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$; the former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
+A larger LKJ shape parameter places more weight on correlation matrices closer to diagonal, i.e., it regularizes away from high correlations.
+The scale parameters $\vb*\gamma$ have i.i.d. $\Gam(1.5, 1.5/4.0)$ priors, the fudge factors $\vb*\alpha$ have i.i.d. $\Gam(5.0, 5.0)$ priors, and $\eps$ has an $\Expo(1/4.0)$ prior.
+Overall, these priors weakly identify the location and scale of the parameters.
+Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.
+
+Like all probabilistic factor models, ours is unidentifiable due to the rotation, location, and scale invariance of the factors in the likelihood.
+To identify the location and scale, we place $\Norm(0,1)$ priors on most components of $\vb L$.
+To identify the rotation, we place more informative priors on the loadings of certain candidates.
+This was accomplished through somewhat of an interactive process, where a preliminary principal components decomposition of the precinct vote shares helped suggest the latent factors that might be found by the model.
+We chose to make align the first dimension with the Sanders-Biden axis, and chose to make the second dimension roughly orthogonal along the Janey-Buttigieg axis.
+This translated to a prior of $\Norm((-1, 0), (0.1^2, 0.25^2))$ for Sanders,
+$\Norm((1, 0), (1.0^2, 0.5^2))$ for Biden, and $\Norm((-0.25, -1), (0.5^2, 2^2))$ for Janey.
+We additionally set the prior correlation between Sanders' and Biden's loadings to be $(-0.5, 0.8)$.
+
+The estimated loading matrix $\vb L$ is visualized on the left of Figure \ref{fig:boston-factors}, and $\pref$ is visualized for Black and White voters on the right.
+
+To test the suitability of the model for representing the outcomes of hypothetical future elections, we use the model fitted to the primary elections to predict the outcome of the mayoral general election in November 2021.
+Michelle Wu and Annissa Essaibi-George advanced to the general from the preliminary election.
+We fix the model's loading matrix $\vb L$ to the the posterior mean of their estimated loadings.
+We draw $s_{jk}$ from a $\Beta(20\bar s, 20(1-\bar s))$ distribution, where $\bar s$ is each candidate's share of the two-candidate vote from the preliminary election (59.7% Wu -- 40.3 Essaibi George).
+
+The model's predictions (posterior medians) for the turnout and vote share in each precinct are shown in Figure \ref{fig:boston-gen-pred}.
+Considering the differences between a primary and general election, and the additional campaign time elapsed between the elections, the predictions are remarkably accurate.
+The predicted turnout has a Spearman correlation of 0.84 with the actual turnout, and the predicted and actual vote shares have a correlation 0.87.
+The turnout and support levels are also in relatively close agreement.

### 2022-05-16 paper/paper.Rmd
AL priors

@@
 This intuition is confirmed by a closer analysis, summarized in Figure \ref{fig:al-maps}(f).
-Across the state, the average harm is 0.075, which translates to around 151,000 voters.
-Of these harmed voters, around 128,000 are Black, while 23,000 are White.
-Similarly, 140,000 are Democratic, while only 11,000 are Republican.
+Across the state, the average harm is 0.082, which translates to around 161,000 voters.
+Of these harmed voters, around 136,000 are Black, while 25,000 are White.
+Similarly, 151,000 are Democratic, while only 10,000 are Republican.
 Of course, the total number of each of these groups of voters is very different.
-Normalizing by group sizes, the expected harm for a Black Democrat is 0.23, for a Black Republican is 0.026, for a White Democrat is 0.084, and for a White Republican is 0.0082.
-Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have a differential harm of 0.149 versus White Democrats and a differential harm of 0.207 versus Black Republicans.
+Normalizing by group sizes, the expected harm for a Black Democrat is 0.25, for a Black Republican is 0.015, for a White Democrat is 0.091, and for a White Republican is 0.0073.
+Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have a differential harm of 0.158 versus White Democrats and a differential harm of 0.235 versus Black Republicans.
 
@@
 We fix the model's loading matrix $\vb L$ to the the posterior mean of their estimated loadings.
-We draw $s_{jk}$ from a $\Beta(20\bar s, 20(1-\bar s))$ distribution, where $\bar s$ is each candidate's share of the two-candidate vote from the preliminary election (59.7% Wu -- 40.3 Essaibi George).
+We draw $s_{jk}$ from a $\Beta(20\bar s, 20(1-\bar s))$ distribution, where $\bar s$ is each candidate's share of the two-candidate vote from the preliminary election (59.7% Wu -- 40.3% Essaibi George).
 

### 2022-05-16 R/05_fit_ei.R

@@ -2,17 +2,25 @@ library(cmdstanr)
 library(posterior)
 library(tidybayes)
 
-fit_ei = function(tbls, recompile=FALSE, algorithm="vb",
+fit_ei = function(tbls, recompile=FALSE, algorithm="vb", prior=NULL,
                   chains=4, warmup=1000, iter=500, adapt_delta=0.8, init=0, refresh=100, ...) {
+    n_race =  ncol(tbls$race)
+    if (is.null(prior)) {
+        prior = list(loc = rep(0, n_race), scale = rep(1, n_race))
+    }
+
     stan_d = list(
         N = nrow(tbls$votes),
         L = ncol(tbls$votes),
-        R = ncol(tbls$race),
+        R = n_race,
 
         vap = tbls$vap,
         votes = tbls$votes,
         dem_votes = tbls$dem_votes,
-        vap_race = tbls$race
+        vap_race = tbls$race,
+
+        prior_supp_loc=prior$loc,
+        prior_supp_scale=prior$scale
     )
 
     path_model = here("R/ei.stan")
@@ -32,6 +40,8 @@ fit_ei = function(tbls, recompile=FALSE, algorithm="vb",
                         init=init, refresh=refresh, adapt_delta=adapt_delta,...)
     } else if (algorithm == "vb") {
         fit = sm$variational(data=stan_d, algorithm="meanfield", init=init, ...)
+    } else if (algorithm == "optim") {
+        fit = sm$optimize(data=stan_d, init=init, ...)
     } else {
         stop("Algorithm should be `hmc` or `vb`.")
     }

### 2022-05-17 paper/paper.Rmd
corrections

@@
 
-We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights have argued that the existing Black majority district packs Black voters, and a second Black opportunity-to-elect district should be drawn.
+We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.
 
@@
 \Pr(\win_{q(I)}(V, q)=\dem\mid I\in j)
-= \Pr(S_{q^{-1}(j),V} > \half)
+= \Pr\qty(S_{q^{-1}(j),V} > \half)
 = \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}),
 $$ where $\Phi$ is the standard Normal CDF.
-To estimate this probability, we simply replace $\Phi$ with the CDF of the $t_{22}$ distribution ($F_{t_22}$), and $\sigma^2_\mathrm{hist}$ with $\hat\sigma^2_\mathrm{hist}$.
+To estimate this probability, we simply replace $\Phi$ with the CDF of the $t_{22}$ distribution ($F_{t_{22}}$), and $\sigma^2_\mathrm{hist}$ with $\hat\sigma^2_\mathrm{hist}$.
 Extending one step further, for two plans $q$ and $\tilde q$, we have
@@
 \Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j)
-= \qty(\Phi\qty(\frac{\gamma(\bar S_{\tilde q^{-1}(j)})}{\sigma^2_\mathrm{hist}})
-- \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
+= \max\qty{\Phi\qty(\frac{\gamma(\bar S_{\tilde q^{-1}(j)})}{\sigma^2_\mathrm{hist}})
+- \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}), 0}.
 \end{equation}
 ```
-where $(\cdot)_+\dfeq \max\{0, \cdot\}$.
 If we condition on $V(I)=\dem$, then \eqref{eq:appharm2} is exactly the expected harm to voter $I$ under plans $q$ and $\tilde q$.
-The calculation is analogous, up to sign changes, for the expected harm conditioning on $V(I)=\rep$.
+The calculation is the same, up to sign changes, for the expected harm conditioning on $V(I)=\rep$.
 This completes the first step: averaging harm within precinct and groups and across elections.
@@
     \dfeq \frac{1}{N_\text{sim}} \sum_{k=1}^{N_\text{sim}}
-    \qty(\Phi\qty(\frac{\gamma(\bar S_{\tilde q_k^{-1}(j)})}{\sigma^2_\mathrm{hist}})
-        - \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}))_+,
+    \max\qty{F_{t_{22}}\qty(\frac{\gamma(\bar S_{\tilde q_k^{-1}(j)})}{\sigma^2_\mathrm{hist}})
+        - F_{t_{22}}\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}), 0},
 $$ where $\hat h_{j,\dem}(q)$ is our estimate of the average harm to a Democratic voter in precinct $j$, averaging across $V$ and $\tilde Q$.
@@
 
-We model the observed vote counts as \begin{align*}
+We model the observed turnout $\votes_{jv}$ and vote counts $s_{jk}\votes_{jv}$  as \begin{align*}
     \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
@@
 \end{align*} where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability paramter specified on the logit scale, and $\circ$ indicates the elementwise (Hadamard) product.
-We give $\beta^{(t)}$ an i.i.d.
-$\Norm(0, 0.6^2)$ prior,and $\beta^{(s)}$ an i.i.d.
-$\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d.
-$\Norm(0, 1.5^2)$ prior, and $\mu^{(s)}$ an i.i.d.
-$\Norm(0, 1.0^2)$ prior.
+We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior. 
 We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$; the former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).
 Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.
+
+The mean support $\mu^{(s)}$ receives an informative prior based on the racially polarized voting estimates of @kuriwaki2021geography, with $\mu^{(s)}\sim\Norm(\logit((0.18, 0.90, 0.35)), (0.04^2, 0.04^2, 0.2^2))$.
+These estimates were derived through multilevel regression and poststratification from survey data, calibrated to actual election results.
 
@@
 
-We model the observed vote counts as \begin{align*}
+We model the observed turnout $\votes_{jv}$ and vote shares $s_{jk}$ as \begin{align*}
     \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
@@
         &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
-    \logit(s_{jk}) \mid p_{jk}, \eps &\sim \Norm(p_{jk}, \frac{\eps}{1+\votes_{jv_k}}) \\ 
-    p_{jk} &= \beta^{(s)}_k + \alpha_k
-        \pref_{j}^\top(\turn_j\circ \vb p_j)\mathrm{diag}(\vb*\gamma)\vb L_k  \\
+    \logit(s_{jk}) \mid \pi_{jk}, \eps &\sim \Norm(\pi_{jk}, \frac{\eps}{1+\votes_{jv_k}}) \\ 
+    \pi_{jk} &= \beta^{(s)}_k + \alpha_k
+        \pref_{j}^\top(\turn_j\circ \vb p_j)\mathrm{diag}(\vb*\rho)\vb L_k  \\
     \pref_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
 \end{align*} Note that $\pref_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $\pref_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
-The matrix $\mathrm{diag}(\vb*\gamma)$ captures the relative scale or importance of each dimension to the overall outcomes.
+The matrix $\mathrm{diag}(\vb*\rho)$ captures the relative scale or importance of each dimension to the overall outcomes.
 
@@
 
-We give $\beta^{(t)}$ an i.i.d. $\Norm(0, 0.5^2)$ prior,and $\beta^{(s)}$ an i.i.d. $\Norm(0, 4.0^2)$ prior, and $\logit(\mu^{(t)})$ an i.i.d. $\Beta(10, 30)$ prior.
+We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.5^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 4.0^2)$ prior, and $\logit(\mu^{(t)})$ an i.i.d. $\Beta(10, 30)$ prior.
 We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$; the former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
 A larger LKJ shape parameter places more weight on correlation matrices closer to diagonal, i.e., it regularizes away from high correlations.
-The scale parameters $\vb*\gamma$ have i.i.d. $\Gam(1.5, 1.5/4.0)$ priors, the fudge factors $\vb*\alpha$ have i.i.d. $\Gam(5.0, 5.0)$ priors, and $\eps$ has an $\Expo(1/4.0)$ prior.
+The scale parameters $\vb*\rho$ have i.i.d. $\Gam(1.5, 1.5/4.0)$ priors, the fudge factors $\vb*\alpha$ have i.i.d. $\Gam(5.0, 5.0)$ priors, and $\eps$ has an $\Expo(1/4.0)$ prior.
 Overall, these priors weakly identify the location and scale of the parameters.

### 2022-05-17 R/06_alabama.R

@@ -139,7 +139,7 @@ m_grps = with(al_map, matrix(c(
     `rownames<-`(c("White", "Black", "Other")) %>%
     `colnames<-`(c("Dem.", "Rep."))
 print(round(m_harm, 1))
-print(round(m_harm/m_grps, 3))
+print(round(m_harm/m_grps, 4))
 }
 
 # plots out

### 2022-05-23 paper/paper.Rmd
rough draft intro

@@
   We propose a unified framework for evaluating the impact of a districting plan on voters overall and in different groups.
-Considering the harm done to individual voters by their inability to elect their chosen candiate, and taking into account the space of counterfactual possibilities, we derive a set of measures that flexibly adapts to local geographic constraints, different types of elections, and different groups of study.
+Considering the harm done to individual voters by their inability to elect their chosen candiate, and taking into account the range of counterfactual possibilities, we derive a set of measures that flexibly adapts to local geographic constraints, different types of elections, and different groups of study.
 We provide recommendations for estimating these measures from historical election returns and discuss connections to commonly used redistricting metrics.
@@
 
--   why counterfactuals?
+For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
+Partisan gerrymandering, whether in the form of rotten boroughs or the salamander-shaped districts of Elbridge Gerry [cite], is perhaps the most consistently visible manifestation of this tendency.
+The academic study of partisan gerrymandering has largely centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@katz2020, need more].
+In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [cite].
+The Voting Rights Act of 1965 (VRA) provided a framework for legal challenges to racial gerrymanders that has also shaped the academic literature, which is characterized by a focus on identifying the voting preferences of racial groups and observing the extent to which these are realized or frustrated through district boundaries and other features of the electoral system [cites].
+At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the electoral system may be more subtle or may favor a narrower or less well-defined group [cite].
+Local redistricting has been less extensively studied, in part for this reason and in part because of data limitations [but see \_\_\_\_].
 
-    -   many factors influence the performance and qualities of an electoral system
+No matter the source or context, all of these cases involve the concentration or dilution of voting power to favor or disfavor a particular group.
+This paper proposes a new framework of *individual and differential harm* for the unified study of all forms of gerrymandering by way of measuring this concentration and dilution.
 
-    -   understanding how different components of the system contribute requires reaching in and manipulating those components
+The harm framework combines two key ideas.
+First, we operate at the level of the individual voter, and examine whether each voter is represented by the candidate of their choice.
+Doing so both unifies our approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public [cites on why districts are good].
 
-    -   just like causal inference, you are inherently interested in the counterfactual where some part of the electoral system is different
+Second, the proposed framework moves beyond a single electoral system under study to also consider *counterfactual* electoral systems.
+Many factors influence the performance and qualities of an electoral system.
+Without a well-defined comparison system, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain.
+As we demonstrate, modern redistricting simulation tools can provide a realistic set of counterfactuals that make these kinds of comparisons possible.
 
--   why representational harm?
+The combination of individual voter choices and counterfactual electoral systems defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different electoral system.
+And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
 
-    -   lit cites on why districts are good
-
-    -   district-based electoral system working well if people are represented by the candidate of their choice
-
-    -   this motivates *representational harm*
-
-    -   we argue that harm is actually a great unifying framework for studying district-based electoral systems
-
--   what makes a good redistricting plan?
-
--   on the partisan side, lots of attention on the fairness to parties
-
-    -   partisan symmetry [@katz2020]
-
-    -   seats-votes curve
-
--   on the racial side, ... ?
-    (VRA focus? looking at race of candidates vs choice? what? I don't know)
-
--   courts, on the other hand, have focused on individual standing [@1964reynolds; @1962baker]
-
--   **we argue that individual standing/harm is an excellent lens to unify the study and measurement of districting plans**
-
--   core idea: counterfactuals + voter choice.
-    Is a voter more likely to be represented by the candidate of their choice under a different districting plan?
-    If so, they've been harmed.
-
-    -   modern simulation methods allow us to produce realistic counterfactuals
-
-    -   by analyzing how harms fall differentially along party, race, ideological, etc. lines, we can connect individual harm to more global measures based on parties, etc.
-
-<!--# Yet, this suffers from a core inflexibility: it does not allow for variation in districting.
-Districting is the fundamental question of district-based democracies.
-For whom can you vote?
-Who do you vote with?
-Our proposed framework rectifies this inflexibility, by shifting to a focus on individual representation.
-The natural measurement approach to this directly incorporates alternatives under the logic of a counterfactual.
--->
+This framework has important similarities with the common-law approach to voting rights litigation.
+To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker].
+This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidates [case cite].
 
@@
 
-Section \ref{sec:motivation} motivates our framework for measuring harm and illustrates the key concepts in the context of partisan gerrymandering for three simple examples.
+In Section \ref{sec:motivation}, we motivate our framework for measuring harm and illustrate the key concepts with three simple examples.
 Section \ref{sec:indiv-harm} formalizes the intuition of Section \ref{sec:motivation} and defines a general measure of individual and differential harm.
@@
 Which is correct?
-Democrats win 8.47 seats on average across the simulations, which is more than the 6.79 seats that would be expected under a completely proportional outcome.
+It depends on what one means by "natural bias," though we submit that differential harm is one good way to measure it.
+Democrats do win 8.47 seats on average across the simulations, which is more than the 6.79 seats that would be expected under a completely proportional outcome.
 
@@
 To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^10] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
-Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail and contains additional diagnostic plots.
+Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail.
 
@@
 We chose to make align the first dimension with the Sanders-Biden axis, and chose to make the second dimension roughly orthogonal along the Janey-Buttigieg axis.
-This translated to a prior of $\Norm((-1, 0), (0.1^2, 0.25^2))$ for Sanders,
-$\Norm((1, 0), (1.0^2, 0.5^2))$ for Biden, and $\Norm((-0.25, -1), (0.5^2, 2^2))$ for Janey.
+This translated to a prior of $\Norm((-1, 0), (0.1^2, 0.25^2))$ for Sanders, $\Norm((1, 0), (1.0^2, 0.5^2))$ for Biden, and $\Norm((-0.25, -1), (0.5^2, 2^2))$ for Janey.
 We additionally set the prior correlation between Sanders' and Biden's loadings to be $(-0.5, 0.8)$.

### 2022-05-23 paper/paper.Rmd
small revisions to whole paper

@@
 For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
-Partisan gerrymandering, whether in the form of rotten boroughs or the salamander-shaped districts of Elbridge Gerry [cite], is perhaps the most consistently visible manifestation of this tendency.
+Partisan gerrymandering, whether in the form of rotten boroughs or Elbridge Gerry's salamander-shaped district [cite], is perhaps the most consistently visible manifestation of this tendency.
 The academic study of partisan gerrymandering has largely centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@katz2020, need more].
@@
 
-The harm framework combines two key ideas.
+The harm framework combines two key ideas, which are motivated at greater length in Section \ref{sec:motivation}.
 First, we operate at the level of the individual voter, and examine whether each voter is represented by the candidate of their choice.
@@
 And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
+Section \ref{sec:motivation} illustrates this with three simple examples, and the general definitions are formalized in Section \ref{sec:indiv-harm}.
+In Section \ref{sec:estimating}, we discuss how to estimate these measures from observed election results.
 
@@
 To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker].
-This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidates [case cite].
+This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidates [@2004vieth; @2018league; @2019lewis].
+A more detailed discussion of the connections between the proposed framework existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.
 
-## Outline of the Paper
-
-In Section \ref{sec:motivation}, we motivate our framework for measuring harm and illustrate the key concepts with three simple examples.
-Section \ref{sec:indiv-harm} formalizes the intuition of Section \ref{sec:motivation} and defines a general measure of individual and differential harm.
-Section \ref{sec:estimating} discusses estimating these measures from observed election results, and provides what we believe is a sensible and efficient way to calculate the measures in the special case of studying partisan gerrymandering.
-Section \ref{sec:connections} connects our proposed measure to existing practices for analyzing racial and partisan gerrymandering.
-Sections \ref{sec:nj}, \ref{sec:alabama}, and \ref{sec:boston} apply the proposed measures to partisan gerrymandering of New Jersey congressional districts, voting rights litigation for Alabama congressional districts, and racial impacts of district boundaries on Boston City Council primary elections.
+Harm and differential harm are more than a useful conceptual unification---they are practically useful in a wide variety of redistricting analyses, including partisan gerrymandering of New Jersey congressional districts (Section \ref{sec:nj}), voting rights litigation for Alabama congressional districts (Section \ref{sec:alabama}), and racial impacts of district boundaries on Boston City Council primary elections (Section \ref{sec:boston}).
 Section \ref{sec:discussion} concludes.
@@
 
-To illustrate the concepts and measures we introduce, we will make use of three examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichusetts.
+To motivate the concepts of individual and differential harm, we will make use of three examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichusetts.
 Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
@@
 Schematics for these examples are shown in Figure \ref{fig:mini-schematic}(a)--(c).
+
+## Minichusetts: counterfactuals matter
+
+Every precinct in Minichusetts has a Democratic majority.
+As a result, it is impossible to draw a districting plan built from precincts that creates even a single Republican district.
+This mirrors the situation in the actual state of Massachusetts, where a high level of partisan integration and a solid statewide Democratic majority leads to Republican inability to win any congressional seats under recent historical voting patterns [@duch:18].
 
@@
 ```
-
-## Minichusetts: counterfactuals matter
-
-Since every precinct in Minichusetts has a Democratic majority, it is impossible to draw a districting plan built from precincts that creates even a single Republican district.
-This mirrors the situation in the actual state of Massachusetts, where a high level of partisan integration and a solid statewide Democratic majority leads to Republican inability to win any congressional seats under recent historical voting patterns [@duch:18].
 
@@
 Consider the two plans in Figure \ref{fig:harm-mininois}(a)-(b).
-A Democratic gerrymander (a) effectuates Democratic control by splitting its city into with the blades of a pinwheel, reaching out into the more Republican and less population-dense remainder of the state.
-We see that the Democratic gerrymander would elect zero Republicans and three Democrats, whereas the proportional map (b) would elect two Democrats and one Republican.
+A Democratic gerrymander (a) effectuates Democratic control by splitting its city, pinwheel-like, into three districts reaching out into the more Republican and sparsely populated remainder of the state.
+This gerrymander would elect zero Republicans and three Democrats, whereas the proportional map (b) would elect two Democrats and one Republican.
 
@@
 Who does this decision harm, if anyone?
-We offer that a voter is harmed if they are not represented by their chosen party, but would have been so counterfactually.
+Precisely those voters who are not represented by their chosen party, but would have been counterfactually.
 Using plan (b) as a counterfactual, we can see that rural Republicans in the eastern half of Mininois are harmed by plan (a), since under plan (b) they would be represented by a Republican.
-Republicans in the city and in the southwest are unharmed, as are all Democratic voters,, since they are represented by the same party under both plans.
-Figure \ref{fig:harm-mininois}, Panel C illustrates this by plotting the number of voters harmed in each precinct.
+Republicans in the city and in the southwest are unharmed, as are all Democratic voters, since they are represented by the same party under both plans.
+Figure \ref{fig:harm-mininois}, Panel C plots the number of voters harmed in each precinct.
 
@@
 We could quantify the harm a Democratic gerrymander imposed on voters in Mininois by comparing election outcomes to those under a counterfactual proportional plan.
-But our results said nothing about the direction of the harm.
+But the results said nothing about the direction of the harm.
 While it is apparent that the Democratic gerrymander indeed benefits Democrats, tallying up the total harm done to all voters is not enough to prove this.
@@
 As in Mininois, we can tally the total number of voters harmed in each precinct ((d) and (f)) in comparison to a proportional counterfactual plan (b).
-As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.
+In both cases, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.
 
@@
 
-# Individual Representational Harm {#sec:indiv-harm}
+# Individual and Differential Harm {#sec:indiv-harm}
 
 This section will formalize and generalize the concepts in Section \ref{sec:motivation}, beginning with some notation.
-Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of legislative districts, indexed by $\Distrs$.
+Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of districts, indexed by $\Distrs$.
 
@@
 The key intuition from Section \ref{sec:motivation} was that to be harmed by a decision a voter must have a better outcome in the counterfactual world where another decision was made.
-To operationalize this, we need to introduce a counterfactual districting plan $\tilde q$.
+To operationalize this, we need to have a counterfactual districting plan $\tilde q$.
 Then we may define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as $$
@@
         \ind\{v(i) = \win_{\tilde q(i)}(v, \tilde q)\},
-$$that is, a voter is harmed only if they are not represented by the candidate of their choice and they would have been represented by their chosen candidate under the counterfactual plan.
+$$that is, a voter is harmed only if they are not represented by the candidate of their choice under $q$ and they would have been represented by their chosen candidate under the counterfactual plan $\tilde q$.
 
@@
 Ballot-level ranked-choice vote data, or candidate ideology scores, could be used to estimate such a cost function.
-Here, to avoid making further assumptions, we limit ourselves to the indicator-based cost function.
-Since many American elections are between just two candidates, this limitation should not prove too restrictive.
+Here, for simplicity and to avoid making further assumptions, we limit ourselves to the indicator-based cost function.
 Section \ref{sec:boston} further discusses the implications of this particular specification in the context of primary elections.
@@
 
-As defined above, $h_i(v,q,\tilde q)$ is defined for a single election and single counterfactual districting plan $\tilde q$, but in general we will be interested in outcomes over multiple elections, and possibly a much larger set of counterfactual plans.
-Ideally we would like to be able to average individual harm over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans, or even over a set of individuals with some shared characteristic.
+As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to a single counterfactual districting plan $\tilde q$.
+In general we will be interested in outcomes over multiple elections, and possibly a much larger set of counterfactual plans.
+Ideally, we would like to be able to average individual harm over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans, or even over a set of individuals with some shared characteristic.
 
@@
 The randomness in districting plans specifically may be conceptualized as the constraints and preferences of the plan-drawing process (i.e., plans which better respect the preferences are given a higher probability).
-As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V$ and $\tilde Q$ are mutually independent, though this is not required for our theoretical analysis.
+As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V$ and $\tilde Q$ are independent, though this is not required for our theoretical analysis.
 Finally, we will also replace $i$ with an individual $I$ taken uniformly at random from $\Voters$, with $I\indep V,\tilde Q$.
@@
 
-Once we have defined a set of groups, we say that a plan $q$ is *fair* in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.
+Once we have defined a set of groups, we say that a plan $q$ is *fair* for groups $\Groups$ in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.
 
@@
 
-The measures $H(q)$ and $\Delta H(q)$ defined above require specification of a probability distribution on elections and districting plans.
+Harm $H(q)$ and differential harm $\Delta H(q)$ as defined above require specification of a probability distribution on elections and districting plans.
 Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
@@
 In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined; there is certainly an element of subjectivity in sampling a set of districting plans.
-Nonetheless, sampling methods are widely used in practice.
+Nonetheless, sampling methods are widely used in practice [see e.g., @2019rucho; @2018league; @2022merrill; @becker2021computational; @chen:rodd:13].
 
@@
 Election results present a different set of modeling challenges, given the large number of factors that influence the outcome, and the general unpredictability of future events.
-As far as estimating measures like $H(q)$ and $\Delta H(q)$, we would ideally use a relatively simple model that captures the primary drivers of election outcomes.
-In the interest of fairness, the model should not make any biasing assumptions about the future (e.g., that demographic shifts will favor one party, or that past trends in election returns will continue in the same direction).
-
-We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting.
-Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
+We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
 Multiparty or primary elections pose additional modeling challenges.
@@
 
-Motivated primarily by computational convenience, we use a *proportionate uniform shift* model, which focuses on swings in the national political environment as the main determinant of election outcomes.
+Motivated primarily by computational convenience and simplicity, we use a *proportionate uniform shift* model, which focuses on swings in the national political environment as the main determinant of election outcomes.
 Election outcomes are modeled as a random deviation from a baseline outcome, and these deviations are assumed to consist of a constant fraction of voters of one party in each precinct switching to the other.
@@
 Partisan symmetry implies that the seats, $s$ won by a party with vote share $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
-In practice, the partisan symmetry framework is applied via the partisan bias metric $\beta(s)$, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020], i.e., the number of seats that would need to be reallocated to reach partisan symmetry for a given statewide vote share $s$ (usually $\half$).
+In practice, the partisan symmetry framework is applied via the partisan bias metric $\beta(v)$, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020], i.e., the number of seats that would need to be reallocated to reach partisan symmetry for a given statewide vote share $v$ (usually $\half$).
 
@@
 
-But as those authors recognize, in evaluating a districting plan, it is critical to consider "if there is a fairer alternative." The inability of partisan bias and other existing measures to account for the space of counterfactual districting plans means that none of them may be interpreted in a vacuum.
+But as those authors recognize, in evaluating a districting plan, it is critical to consider "if there is a fairer alternative." The inability of partisan bias and other existing measures to account for counterfactual districting plans means that none of them may be interpreted in a vacuum.
 No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the district-based electoral system.
@@
 
-These problems have led @deford2020symmetry, among others, to recommend evaluating the presence of a gerrymander simply by comparing the expected number of seats won under the plan versus the simulations We agree with this recommendation, as it turns out that in the partisan case, differential harm is in fact closely related to the overall number of seats won by each party.
-Proposition \ref{prop:fairness} records this formally below.
-However, unlike a simple measure of seats, differential harm is shifted so that a value of zero corresponds to fairness, and rescaled to be directly interpretable as a fraction of voters harmed.
+These problems have led @deford2020symmetry, among others, to recommend evaluating the presence of a gerrymander simply by comparing the expected number of seats won under the plan versus the simulations We agree with this recommendation, as it turns out that in the partisan case, differential harm is in practice tightly related to the overall number of seats won by each party, as Section \ref{sec:nj} and Appendix \ref{app:corr} demonstrate.
+However, unlike a direct measure of seats, differential harm is shifted so that a value of zero corresponds to fairness, and rescaled to be directly interpretable as a fraction of voters harmed.
 
-```{=tex}
-\begin{proprep} \label{prop:fairness}
-Suppose $\Groups=\{\dem,\rep\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
-\begin{align*}
-    \Delta_{\rep-\dem} H(q) = \E[\seats(\dem, V, \tilde Q)] - \E[\seats(\dem, V, q)].
-\end{align*}
-\textbf{CM note: define $\seats$ or replace with sum.}
-\end{proprep}
-\begin{proof}
-\end{proof}
-```
 ## Racial Gerrymandering {#subsec:racial}
@@
 
-In both these cases, the proposed measures will quantify the extent to which a challenged districting plan differentially harms minority voters, taking into account their geographic and political cohesion, and the individual preferences of every voter.
+In both these cases, the proposed harm measures will quantify the extent to which a challenged districting plan differentially harms minority voters, taking into account their geographic and political cohesion, and the individual preferences of every voter.
 Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.
@@
 #|      and one randomly sampled plan (d).
-#|      Census-designated places are shaded in gray and county lines in white.
+#|      County lines are overlaid in white.
 include_graphics("figures/nj_maps.pdf")
@@
 
-[^7]: As in section \ref{sec:connections}, this is $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
+[^7]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
 
@@
 The efficiency gap and declination measures also correctly identify the gerrymanders, though they are much more discrete, and declination gives a much larger score in absolute value to the Democratic plan than to the Republican.
-In contrast, the partisan bias and mean-median measures incorrectly classify the Republican gerrymander as a fair plan and a Democratic-leaning outlier.
-These measures are weakly correlated with the expected number of Democratic seats, and show substantial variation even for fixed values of total seats won at baseline.
+In contrast, the partisan bias and mean-median measures incorrectly classify the Republican gerrymander as a fair plan and a Democratic-leaning outlier, respectively.
 
 Next, we consider the distribution of these statistics across the sampled plans, but this doesn't help much, either.
-The expected number of Democratic seats naturally identifies the gerrymanders (two-sided $p=0.0002$) and the direction of bias, and as expected from Proposition \ref{prop:fairness} this measure is highly correlated with differential partisan harm.
+The expected number of Democratic seats naturally identifies the gerrymanders (two-sided $p=0.0002$) and the direction of bias, and as discussed above this measure is extremely correlated with differential partisan harm.
 The gerrymanders are also an outlier under the efficiency gap and declination measures also (both $p=0.0002$).
 While the gerrymanders are both an outlier for the mean-median score, only the Democratic plan is an outlier according to the partisan bias measure ($p=0.0022$, versus $p=0.099$ for the Republican plan).
+These measures are weakly correlated with the expected number of Democratic seats, and show substantial variation even for fixed values of total seats won at baseline.
 
@@
 
+New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
+Appendix \ref{app:corr} calculates these correlations across a number of states.
+The most consistent correlation is that between our estimate of differential partisan harm and the expected number of seats won by each party, which is always above \_\_ in the states we study.
+
 # Application to Voting Rights Litigation in Alabama {#sec:alabama}
 
-We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.
-
-As discussed in Section \ref{subsec:racial}, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect plan could be drawn while meeting all other redistricting requirements.
-In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure \ref{fig:al-maps}(e).
-These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure \ref{fig:al-maps}(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.[^8]
+We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.[^8]
 
 [^8]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
+    One of the authors (Kenny) provided research support to an expert witness for the plaintiffs.
+
+As discussed in Section \ref{subsec:racial}, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
+In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure \ref{fig:al-maps}(e).
+These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure \ref{fig:al-maps}(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
 
@@
 The model, detailed in Appendix \ref{app:alei}, jointly estimates turnout and vote choice by race and precinct across seven statewide elections.
-Unsurprisingly, voting is highly racially polarized outside of the cities of Birmingham and Montgomery.
+Voting is highly polarized by race outside of the cities of Birmingham and Montgomery.
 
-With these estimates in hand, we can calculate the average number of voters harmed in each precinct and racial group.
-Figure \ref{fig:al-maps}(c) shows the geographic distribution of harm.
+With these estimates in hand, we can estimate the average number of voters harmed in each precinct and racial group.
+Figure \ref{fig:al-maps}(c) shows the estimated geographic distribution of harm.
 Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures \ref{fig:al-maps}(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
-It is precisely these areas that are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.
+These areas are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.
 
@@
 This intuition is confirmed by a closer analysis, summarized in Figure \ref{fig:al-maps}(f).
-Across the state, the average harm is 0.082, which translates to around 161,000 voters.
+Across the state, the estimated average harm is 0.082, which translates to around 161,000 voters.
 Of these harmed voters, around 136,000 are Black, while 25,000 are White.
@@
 First, the city council is officially nonpartisan and city politics are dominated by the Democratic Party, so voting preferences are more complex than a simple partisan preference.
-To estimate election outcomes, we will need some way of mapping observed support for particular candidates in one election onto counterfactual support for a hypothetical candidate in another election.
-Second, as in Alabama, we aim to examine the racial impact of the district system, and thus need to estimate these voting preferences by race.
-Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone tell us how different racial groups voted.
+To estimate election outcomes, we require some way of mapping observed support for particular candidates in one election onto counterfactual support for a hypothetical candidate in another election.
+Second, as in Alabama, we aim to examine the racial impact of the district system, and thus must estimate these voting preferences by race.
+Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.
 
@@
 We consider two different scenarios: one in which all of the council races are contested along the first latent dimension, and one in which the races are contested along the second.
-Specifically, we draw voter choices from the posterior predictive distribution of our model, setting the candidates' loadings to --1 and 1 on the chosen dimension and 0 on the other dimension.
+Specifically, we draw voter choices between two candidates from the posterior predictive distribution of our model, setting the candidates' loadings to --1 and 1 on the chosen dimension and 0 on the other dimension.
 We draw the overall level of support from a $\Beta(15, 15)$ distribution, which means that we will study election scenarios where the overall vote share for each bloc of candidates ranges from roughly 30 to 70 percent.
 The posterior predictive draws yield counts of voters for each candidate in each precinct by race, from which we can calculate winners at the district level, under any configuration of districts.
+
+We limit our hypothetical elections to two candidates, to simplify the challenges of measuring harm with many candidates, over which voters may have different ranked preferences.
+However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
+One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption that all candidates the voter did not vote for are exchangeable.
 
@@
 
-For the first scenario, where council races are contested along the ideological axis, we find that the enacted plan is quite fair, differentially harming White voters at a rate of only 0.0079 compared to the race-neutral baseline.
+For the first scenario, where council races are contested along the ideological axis, we estimate that the enacted plan is quite fair, differentially harming White voters at a rate of only 0.0079 compared to the race-neutral baseline.
 For the second scenario, where the contests are oriented along the racial dimension, we estimate that the enacted plan differentially harms White voters at a rate of 0.041, compared to the race-neutral baseline.
@@
 This will produce a different answer, since harm is not symmetric between the study plan and the counterfactual.
-Along the first dimension, a race-neutral baseline would differentially harm Black voters at a rate of 0.031 compared to the enacted plan, while along the second dimension, the differential harm would be 0.018.
+Along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 0.031 compared to the enacted plan, while along the second dimension, the differential harm would be 0.018.
 
@@
 
-For the menagerie, we proffer a new beast.
-*can we make the whole discussion a zoo metaphor* This paper is an attempt to bridge the quantitative gap between votes and representation.
+Discussion outline suggestion:
+
+-   Recap goal and advantages.
+    Unified framework for studying redistricting.
+    Considers counterfactuals.
+    Centers vote choice, and separates vote choice from group membership.
+
+-   Recap strengths.
+    New normative standard.
+    Separation of target and estimation.
+    Flexible framework.
+    Application to local elections and places without strong parties.
+
+-   Discuss limitations.
+
+    -   Definition of counterfactual subjective & has a big impact.
+        Some kind of bounds analysis where we look at best/worst case counterfactuals?
+
+    -   Limited for multi-candidate settings where voters have ordinal preferences
+
+    -   Election models limited
+
+    -   Simplyfying estimating assumptions like uniform swing
+
+    -   Computational complexity, especially versus status quo measures which are cheap to compute.
+
+-   Final ¶: This stuff is complex.
+    Best we can do is to have a modular & flexible framework where people can tinker with assumptions and decompose how various pieces contribute to observed patterns and impacts.
+    Researchers can and should push our framework further to apply it better to multi-candidate elections, more realistic models, and beyond district-based electoral systems.
+
+This paper is an attempt to bridge the quantitative gap between votes and representation.
 We break from existing redistricting literature and focus on the relationship between votes and representation by copartisans, rather than the relationship between votes and seats.
@@
 
+# Correlation Between Partisan Gerrymandering Metrics Across Multiple States {#app:corr}
+
+The table below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section \ref{sec:nj} for \_\_ states which cover a wide range of partisan leans and number of seats.
+The simulations come from [alarm cite].
+
 # Ecological Inference Model for Alabama {#app:alei}
@@
 
+Like all ecological inference models, this one makes several assumptions, verifiable and unverifiable.
+We assume no spatial correlation in residuals, when some is almost certainly present.
+We place distributional assumptions on how voter preferences by races vary across precincts (importantly, that these preferences are centered around a common location).
+And we of course impose a latent two-dimensional factor structure, which must always be a major simplification of reality.
+
 To test the suitability of the model for representing the outcomes of hypothetical future elections, we use the model fitted to the primary elections to predict the outcome of the mayoral general election in November 2021.

### 2022-05-24 paper/paper.Rmd
work on citations

@@
 For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
-Partisan gerrymandering, whether in the form of rotten boroughs or Elbridge Gerry's salamander-shaped district [cite], is perhaps the most consistently visible manifestation of this tendency.
-The academic study of partisan gerrymandering has largely centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@katz2020, need more].
-In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [cite].
-The Voting Rights Act of 1965 (VRA) provided a framework for legal challenges to racial gerrymanders that has also shaped the academic literature, which is characterized by a focus on identifying the voting preferences of racial groups and observing the extent to which these are realized or frustrated through district boundaries and other features of the electoral system [cites].
-At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the electoral system may be more subtle or may favor a narrower or less well-defined group [cite].
-Local redistricting has been less extensively studied, in part for this reason and in part because of data limitations [but see \_\_\_\_].
+Partisan gerrymandering, whether in the form of rotten boroughs or Elbridge Gerry's salamander-shaped district [@griffith1907rise, pp.
+72--73], is perhaps the most consistently visible manifestation of this tendency.
+The academic study of partisan gerrymandering has largely centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
+Other types of gerrymandering exist as well.
+In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [@gilliland1969racial].
+The Voting Rights Act of 1965 (VRA) provided a framework for legal challenges to racial gerrymanders that has also shaped the academic literature, which is characterized by a focus on identifying the voting preferences of racial groups and observing the extent to which these are realized or frustrated through district boundaries and other features of the electoral system [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
+At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the electoral system may be more subtle or may favor a narrower or less well-defined group [@cain2002mapmaking; @halberstam2012process].
+Local redistricting has been less extensively studied, in part for this reason and in part because of data limitations [but see @siegel2013educational; @vargas2021racial].
 
@@
 To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker].
-This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidates [@2004vieth; @2018league; @2019lewis].
+This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
 A more detailed discussion of the connections between the proposed framework existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.
@@
 [^8]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
-    One of the authors (Kenny) provided research support to an expert witness for the plaintiffs.
 
@@
 The table below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section \ref{sec:nj} for \_\_ states which cover a wide range of partisan leans and number of seats.
-The simulations come from [alarm cite].
+The simulations are provided by the 50-State Simulation Project [@mccartan2022fiftystates].
 

### 2022-05-25 paper/paper.Rmd
correlation appendix

@@
 First, we operate at the level of the individual voter, and examine whether each voter is represented by the candidate of their choice.
-Doing so both unifies our approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public [cites on why districts are good].
+Doing so both unifies our approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [@brunell2010; other cites on why districts are good.
+Federalist No.
+52?].
 
@@
 New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
-Appendix \ref{app:corr} calculates these correlations across a number of states.
-The most consistent correlation is that between our estimate of differential partisan harm and the expected number of seats won by each party, which is always above \_\_ in the states we study.
+Appendix \ref{app:corr} calculates these correlations across eight states.
+The most consistent correlation is that between our estimate of differential partisan harm and the expected number of seats won by each party, which is usually around 0.98, and always above 0.66 in the eight states we study.
 
@@
 \renewcommand\thefigure{\thesection.\arabic{figure}}
+\renewcommand\thetable{\thesection.\arabic{table}}
 \setcounter{figure}{0}
+\setcounter{table}{0}
 ```
@@
 
-The table below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section \ref{sec:nj} for \_\_ states which cover a wide range of partisan leans and number of seats.
+Table \ref{tab:corr} below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section \ref{sec:nj} for eight states which cover a wide range of partisan leans and number of seats:
+
+- Utah, with 4 districts and a Democratic vote share baseline of 33%
+- Mississippi, with 4 districts and a Democratic vote share baseline of 43%
+- Iowa, with 4 districts and a Democratic vote share baseline of 46%
+- North Carolina, with 14 districts and a Democratic vote share baseline of 50%
+- Wisconsin, with 8 districts and a Democratic vote share baseline of 51%
+- New Jersey, with 12 districts and a Democratic vote share baseline of 57%
+- Massachusetts, with 9 districts and a Democratic vote share baseline of 61%
+- California, with 52 districts and a Democratic vote share baseline of 65%
+
+The correlations are calculated across 5,000 simulated plans for each state.
 The simulations are provided by the 50-State Simulation Project [@mccartan2022fiftystates].
+In particular, the New Jersey simulations are not the same as those used in Section \ref{sec:nj}.
+
+\begin{table}[h!]
+\centering
+\input{figures/corr_table.tex}
+\vspace{0.1in}
+\caption{Correlations between five gerrymandering metrics and partisan differential harm, for simulated districting plans from eight states.
+Declination is NA for Massachusetts since there are no Republican-leaning districts at baseline under any simulated plan.}
+\label{tab:corr}
+\end{table}
 
@@
 
-```{r boston-gen-pred, fig.width=6.5}
-#| fig.cap: |
-#|      Model predictions of general election vote shares and turnout, 
-#|      compared to actual general election results.
-#|      Red reference line indicates perfect agreement.
-include_graphics("figures/boston_gen_pred.pdf")
-```
-
 We model the observed turnout $\votes_{jv}$ and vote shares $s_{jk}$ as \begin{align*}
@@
 
+```{r boston-gen-pred, fig.width=6.5}
+#| fig.cap: |
+#|      Model predictions of general election vote shares and turnout, 
+#|      compared to actual general election results.
+#|      Red reference line indicates perfect agreement.
+include_graphics("figures/boston_gen_pred.pdf")
+```
+
 Like all ecological inference models, this one makes several assumptions, verifiable and unverifiable.

### 2022-05-25 R/08_correlation.R

@@ -1,0 +1,74 @@
+if (!requireNamespace("alarmdata", quietly=TRUE)) {
+    remotes::install_github("alarm-redist/alarmdata")
+}
+library(alarmdata)
+library(kableExtra)
+
+states = c("IA", "NC", "WI", "MA", "UT", "NJ", "MS", "CA")
+
+ker_t = k_t()
+corr_state = function(abbr) {
+    st_map = alarm_50state_map(abbr)
+    dvote = st_map$ndv
+    rvote = st_map$nrv
+    statewide = sum(dvote) / sum(dvote + rvote)
+    nd = attr(st_map, "ndists")
+
+    st_plans = alarm_50state_plans(abbr, stats=TRUE)
+    attr(st_plans, "ndists") = nd
+    cat("Downloaded", abbr, "\n")
+
+    idx_2 = sample(seq_len(ncol(as.matrix(st_plans)))[-1], 200)
+    hh = harm(st_plans, ndshare, st_map$ndv, st_map$nrv, idx_2=idx_2, ker=ker_t)
+
+    pl_corr = st_plans %>%
+        subset_sampled() %>%
+        mutate(decl = part_decl(., st_map, dvote, rvote)) %>%
+        group_by(draw) %>%
+        summarize(e_dem = sum(ker_t(ndshare)),
+                  pbias = pbias[1],
+                  egap = egap[1],
+                  decl = decl[1],
+                  mean_med = median(ndshare) - mean(ndshare)) %>%
+        mutate(h_dem = hh[1, -1],
+               h_rep = hh[2, -1],
+               dh = h_dem - h_rep,
+               h = hh[3, -1]) %>%
+        as_tibble() %>%
+        select(dh, e_dem, egap, pbias, mean_med, decl) %>%
+        as.matrix() %>%
+        cor(use="pairwise.complete.obs")
+
+    tibble(state=censable::match_name(abbr),
+           statewide=statewide,
+           ndists=nd,
+           stat=c("Exp. Dem. seats", "Efficiency gap", "Partisan bias",
+                  "Mean-median", "Declination"),
+           corr_dh=pl_corr[-1, "dh"])
+
+}
+
+res = map_dfr(states, corr_state)
+
+# make table
+tbl_body = res %>%
+    # mutate(state = str_glue("{state} ({round(100*statewide)}% D, {ndists} districts)")) %>%
+    arrange(statewide) %>%
+    select(-statewide, -ndists) %>%
[…]

### 2022-05-25 R/08_sf.R

@@ -1,94 +1,0 @@
-if (!exists("fit_lfei")) source(here("R/05_fit_ei.R"))
-
-# Load and fit -------------
-
-d = read_rds(here("data/sf_elec.rds"))
-map = redist_map(d, existing_plan=sup_2020, pop_tol=0.05, adj=d$adj)
-
-d_votes =  d %>%
-    as_tibble() %>%
-    select(precinct, atty_loftus:p21_no) %>%
-    pivot_longer(atty_loftus:p21_no, names_to="cand", values_to="votes") %>%
-    separate(cand, c("elec", "candidate"), sep="_", extra="merge") %>%
-    mutate(precinct = fct_inorder(precinct),
-           elec = fct_inorder(elec),
-           candidate = if_else(elec %in% c("p19", "p21"),
-                               str_c(elec, "_", candidate), candidate),
-           candidate = fct_inorder(candidate))
-tbls = make_tables_sf(d_votes, cand_thresh=0.02)
-
-if (!file.exists(fit_path <- here("data-raw/sf_fit.rds"))) {
-    id = list(
-        loc = list(p21_yes=c(-1, 0), p21_no=c(1, 0), biden=c(1, 0.5), buttigieg=c(0, 1)),
-        scale = list(p21_yes=c(0.5, 0.25), p21_no=c(0.5, 0.5), biden=c(1, 1), buttigieg=c(1, 0.5))
-    )
-    # fit = fit_lfei(tbls, n_comp=2L, id=id, algorithm="vb", eta=0.5, adapt_engaged=F, tol_rel_obj=0.005)
-
-    fit = fit_lfei(tbls, n_comp=2L, id=id, algorithm="hmc", step_size=0.1,
-                   inv_metric=read_rds(here("data-raw/sf_metric.rds")),
-                   warmup=500, iter=1000, thin=2)
-    save_fit(fit, fit_path)
-} else {
-    fit = read_rds(fit_path)
-}
-
-
-
-# VALIDATION ---------------
-
-library(bayesplot)
-rhats = unlist(lapply(fit, posterior::rhat))
-qplot(rhats, bins=50)
-mcmc_intervals(as_draws_matrix(fit$support))
-mcmc_intervals(as_draws_matrix(fit$alpha))
-mcmc_intervals(as_draws_matrix(fit$loading[, 1]))
[…]

### 2022-05-25 paper/paper.Rmd
edits minus discusison

@@
 This framework has important similarities with the common-law approach to voting rights litigation.
-To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker].
+To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98].
 This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
@@
 
-Measuring the harm of the Democratic gerrymander required comparing it to a fair counterfactual plan.
+Measuring the harm of the Democratic gerrymander required comparing it to a *fair* counterfactual plan. *ck: either delete or italicize fair?*
 The choice of counterfactual might seem hopelessly subjective.
@@
 
-This section will formalize and generalize the concepts in Section \ref{sec:motivation}, beginning with some notation.
+We formalize and generalize the concepts in Section \ref{sec:motivation}, beginning with some notation.
 Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of districts, indexed by $\Distrs$.
@@
 
-**CHRIS: review please.**
-
-The study of racial gerrymandering is considerably more complicated than that of partisan gerrymandering, due to the difficulty of understanding racial voting preferences when individual votes are secret, and as such no simple formulas exist to measure the racial impacts of a districting plan.
+The study of racial gerrymandering and vote dilution is considerably more complicated than that of partisan gerrymandering, due to the difficulty of understanding racial voting preferences when individual votes are secret.
+As such, no simple formulas exist to measure the precise racial impacts of a districting plan.
 Indeed, much of the academic literature has focused on the statistical challenge of estimating vote preferences and turnout by race [@king1997solution; @greiner2006ecological; @imai2016improving].
 
-The modern analysis of racial gerrymandering is inseparable from the Voting Rights Act of 1965 (VRA).
-The VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
+The modern analysis of racial gerrymandering and vote diluation are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th Amendment.
+Section 2 of the VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
 Under the current framework established by @1986gingles, to do so, plaintiffs must prove that (a) the minority group is large and geographically compact enough to form a majority in some hypothetical district, (b) the minority group is "politically cohesive," and (c) the majority group can and does vote to deny the minority group its choice of candidate [@hood2018legal] (the so-called *Gingles* criteria).
@@
 Democrats do win 8.47 seats on average across the simulations, which is more than the 6.79 seats that would be expected under a completely proportional outcome.
+The key benefit here is that differential harm does not push for localized gerrymanders to overcome the geographic imbalance in partisanship.
 

### 2022-05-25 paper/paper.Rmd
draft discussion

@@
 
-Measuring the harm of the Democratic gerrymander required comparing it to a *fair* counterfactual plan. *ck: either delete or italicize fair?*
+Measuring the harm of the Democratic gerrymander required comparing it to a neutral counterfactual plan.
 The choice of counterfactual might seem hopelessly subjective.
@@
 
+We offer a new framework which unifies varied strains of thoughts from common law standing to the logic of counterfactuals for causal inference.
+By using counterfactual redistricting plans, we reset the focus on voters who could be better represented.
+This imbues the long-standing understandings of standing into the quantity of interest.
+Once measured, this appreciably increases the accuracy in identifying partisan gerrymanders---both in sign and in existence.
+It goes beyond partisanship, too, allowing us to identify differential harms in other groups, such as by race.
+Yet, by centering the framework on voters and vote choice, we can calculate the harm to individuals beyond their group memberships.
+For any estimated relationship between the voters and their vote choice, such as their within-party ideology, we can estimate the harm to each individual.
+
+This flexibility stems from the development of a novel standard of fairness in redistricting.
+Moving away from partisan symmetry, we look to the rate at which voters are deprived of their ability to elect their chosen candidate.
+We offer measurement strategies for this quantity, building largely out of advances in simulation methodologies.
+Taken together, this gives one unified and feasible framework for studying redistricting.
+
+Yet, this is surely not the end of the menagerie.
+Flexibility comes at a cost, the largest being the specification of a counterfactual.
+While a representative sample of legally feasible plans may be a reasonable starting point, full inclusion of discretionary choices can be subjective.
+When subjective choices are not represented consistently, the analysis may differ.
+Further, modelling future elections is simply hard.
+While incorporating model free estimates based on past results is common, there is no assurance in future validity.
+When we do write a model, we often rely on very powerful simplifying assumptions, such as uniform swing models.
+Computationally, bringing all of these parts together remains difficult, as you need the ability to compute everything for each counterfactual plan.
+Each of these limitations present opportunities for future study.
+
+Redistricting is difficult to study.
+At its core, it is a high dimensional problem that must balance normative, legal, and administrative concerns to allow citizens to exercise their rights.
+We don't believe that this framework fixes the dimensionality problem.
+But, we do offer a modular and flexible framework to allow researchers to study more dimensions together.
+This framework allows people to tinker with their basic assumptions and decompose how various pieces contribute to observed patterns and impacts.
+Researchers can and should push our framework further to apply it better to multi-candidate elections, more realistic models, and beyond district-based electoral systems.
+
 Discussion outline suggestion:
@@
     -   Limited for multi-candidate settings where voters have ordinal preferences
+    Even given a model for two candidates, extending this approach to more complex elections, say multi-candidate election with ordinal preferences, is not yet done.
 

### 2022-05-25 paper/paper.Rmd
footnote pedantry
good to go

--- 2022-05-25 paper/paper.Rmd
@@
       Research Workshop and Imai Research Group  for helpful comments and discussion.
+      Replication code and data are available at <https://github.com/CoryMcCartan/representation>.
     department: Department of Statistics
@@
   We propose a unified framework for evaluating the impact of a districting plan on voters overall and in different groups.
-Considering the harm done to individual voters by their inability to elect their chosen candiate, and taking into account the range of counterfactual possibilities, we derive a set of measures that flexibly adapts to local geographic constraints, different types of elections, and different groups of study.
+Considering the harm done to individual voters by their inability to elect their chosen candidate, and taking into account the range of counterfactual possibilities, we derive a set of measures that flexibly adapts to local geography and constraints, different types of elections, and different groups of study.
 We provide recommendations for estimating these measures from historical election returns and discuss connections to commonly used redistricting metrics.
@@
 First, we operate at the level of the individual voter, and examine whether each voter is represented by the candidate of their choice.
-Doing so both unifies our approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [@brunell2010; other cites on why districts are good.
-Federalist No.
-52?].
+Doing so both unifies our approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [see also @brunell2010].
 
@@
 This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
-A more detailed discussion of the connections between the proposed framework existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.
+A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.
 
@@
 
-Panel (e) then splits the voters by party, and computes, for each party, the share of its voters which harmed.
+Panel (e) then splits the voters by party, and computes, for each party, the share of its voters which are harmed.
 The Democratic gerrymander (a) harms Republicans at a much higher rate than Democrats, and the opposite can be said of the Republican gerrymander (c).
@@
 
-## Harm
+## Harm {#subsec:harm}
 
@@
 A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between candidates he or she did not vote for.
-For example, a voter may consider candidates A and B nearly exchangeable, but may strongly dislike candidate C.
+For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
 If the voter ultimately voted for candidate A, this harm measure makes no distinction between candidate B or candidate C winning the voter's district.
@@
 To the extent that voters reward or punish an incumbent party for the districting plan they adopt, or that an adopted districting plan affects which candidates run in a district, this assumption is violated.
-But as our suggested model for election outcomes $V$ will be quite coarse, not accounting for candidate factors or policy preferences, the marginal effect of this independence assumptions on our conclusions will be minimal.
+But as our suggested model for election outcomes $V$ will be quite coarse, not accounting for candidate factors or policy preferences, the marginal effect of this independence assumption on our conclusions will be minimal.
 
@@
 The canonical framework for considering partisan fairness is partisan symmetry [@king1987democratic].
-Partisan symmetry implies that the seats, $s$ won by a party with vote share $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
+Partisan symmetry implies that the seats $s$ won by a party with vote share $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
 In practice, the partisan symmetry framework is applied via the partisan bias metric $\beta(v)$, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020], i.e., the number of seats that would need to be reallocated to reach partisan symmetry for a given statewide vote share $v$ (usually $\half$).
@@
 More recently, and particularly in the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
-The *mean-median* difference looks to relationship between the vote share of the median seat in a delegation and the mean vote share of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
+The *mean-median* difference looks at the relationship between the vote share of the median seat in a delegation and the mean vote share of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
 The *declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
@@
 
-The modern analysis of racial gerrymandering and vote diluation are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th Amendment.
+The modern analysis of racial gerrymandering and vote dilution are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
 Section 2 of the VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
@@
 #|      Racial composition of the city of Boston (left)
-#|      and 2010 city coucil districts (right).
+#|      and 2010 city council districts (right).
 #|      Neighborhood boundaries are overlaid in black on the
@@
 The city of Boston is governed by a mayor and a nonpartisan 13-member city council.
-Nine of the councillors are elected from contiguous city council districts, and the remaining four councillors are elected at-large, with all councillors serving two-year terms.
+Nine of the councilors are elected from contiguous city council districts, and the remaining four councilors are elected at-large, with all councilors serving two-year terms.
 Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^9]
@@
 But there is significant variation within racial groups as well, such as the split between more conservative White voters in South Boston and more liberal White voters in Jamaica Plain (the first latent dimension appears to capture a progressive-moderate divide).
-It is certainly possible, then that the particular way districts are drawn will have an effect on the ability of different racial groups to elect their candidate of choice.
+It is certainly possible, then, that the particular way districts are drawn will have an effect on the ability of different racial groups to elect their candidate of choice.
 
@@
 However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
-One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption that all candidates the voter did not vote for are exchangeable.
+One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption that all candidates the voter did not vote for are interchangeable.
 
@@
 
-Discussion outline suggestion:
+This paper offers a new framework for studying redistricting that brings together a focus on individual voter preferences and the causal-inference logic of counterfactuals.
+The combination of these ideas imbues the analysis of districting plans with the historical understandings of legal standing and harm.
+This focus naturally leads to a new normative standard of plan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters.
 
--   Recap goal and advantages.
-    Unified framework for studying redistricting.
-    Considers counterfactuals.
-    Centers vote choice, and separates vote choice from group membership.
+Compared to existing alternatives, differential harm more accurately identifies partisan gerrymanders---both in sign and in existence.
+The exact same approach applies beyond partisanship and allows for the calculation of differential harms by race and other protected groups.
+By centering the framework on individual voters' choices, we sidestep the need to blindly associate voters' preferences with their group memberships.
+Analysts can model vote choice separately, estimate the harm to each individual, and then aggregate by any group of interest.
+As Section \ref{sec:nj}--\ref{sec:boston} showed, this opens up space for more nuanced analyses that can extend to the local level as well.
 
--   Recap strengths.
-    New normative standard.
-    Separation of target and estimation.
-    Flexible framework.
-    Application to local elections and places without strong parties.
+The proposed framework is far from the end of the road.
+Its flexibility comes at a cost---most notably, the need to specify a set of counterfactual districting plans, the choice of which has a significant impact on conclusions.
+While modern simulation methods are a valuable starting point, these methods involve analyst discretion and interpretation of often-murky legal standards.
+When subjective choices are not represented consistently, different analyses may arrive at different conclusions about which voters are harmed and at what rate.
 
--   Discuss limitations.
+Modeling future elections poses another set of challenges.
+Simple models are easy to build from commonly-available data and are computationally tractable, but are gross simplifications of reality.
+More realistic models require more analyst thought and can add significant burden to the calculation of our harm measures.
+And as discussed in Section \ref{subsec:harm}, multi-candidate elections where voters have strong ordinal preferences provide further challenges that the proposed framework does not fully address.
+Each of these limitations present opportunities for future study.
 
-    -   Definition of counterfactual subjective & has a big impact.
-        Some kind of bounds analysis where we look at best/worst case counterfactuals?
-
-    -   Limited for multi-candidate settings where voters have ordinal preferences
-
-    -   Election models limited
-
-    -   Simplyfying estimating assumptions like uniform swing
-
-    -   Computational complexity, especially versus status quo measures which are cheap to compute.
-
--   Final ¶: This stuff is complex.
-    Best we can do is to have a modular & flexible framework where people can tinker with assumptions and decompose how various pieces contribute to observed patterns and impacts.
-    Researchers can and should push our framework further to apply it better to multi-candidate elections, more realistic models, and beyond district-based electoral systems.
-
-This paper is an attempt to bridge the quantitative gap between votes and representation.
-We break from existing redistricting literature and focus on the relationship between votes and representation by copartisans, rather than the relationship between votes and seats.
-This focus naturally leads to a new normative standard of partisan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters of each party.
-This may occur regardless of whether the districting plan achieves partisan symmetry.
-
-This approach offers new purchase into the foundational questions of how to quantify the harms and benefits that accrue to voters in district-based democracies, where the rules of defining districts are fixed, but the choice of how to how to draw the districts remains a policy question.
-While the overall framework is flexible, we offer a selection of simple yet interesting measures for improving the individual partisan implications of a districting plan.
-Further work is needed to characterize how existing redistricting requirements and practices, such as the number or size of districts, or encouragement of competitive districts, interact with these measures.
+At its core, redistricting is a multifaceted and high-dimensional problem that must balance political, legal, and administrative constraints to allow citizens to exercise their rights.
+A districting plan can never be boiled down to a single number.
+But we hope our approach provides a modular and flexible set of measures that can help decompose how various components of the electoral system combine to produce observed patterns and election outcomes.
+Researchers can and should tinker with our assumptions and push our framework further to better apply it to multi-candidate elections, more realistic election models, and beyond district-based electoral systems.
 
@@
 [^11]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
-    Population information comes from Census blocks, which are practically never split.
+    Population information comes from Census blocks, which, when populated, are practically never split.
 
@@
 
+```{=tex}
 \begin{table}[h!]
@@
 \end{table}
-
+```
 # Ecological Inference Model for Alabama {#app:alei}
@@
 
-```{r alabama-ei, fig.width=6.5}
-#| fig.cap: |
-#|      Model estimates of precinct-level turnout by race (a) 
-#|      and Democratic support by race (b).
-#|      Points are sized in proportion to their voting-age population, 
-#|      and colored according to their Black share of the voting-age population.
-#|      All axes are transformed to a logit scale.
-include_graphics("figures/al_ei.pdf")
-```
-
 We model the observed turnout $\votes_{jv}$ and vote counts $s_{jk}\votes_{jv}$ as \begin{align*}
@@
         &\sim \Norm(\mu^{(s)}, \Sigma^{(t)}),
-\end{align*} where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability paramter specified on the logit scale, and $\circ$ indicates the elementwise (Hadamard) product.
+\end{align*} where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale, and $\circ$ indicates the elementwise (Hadamard) product.
 We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior.
@@
 Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.
+
+```{r alabama-ei, fig.width=6.5}
+#| fig.cap: |
+#|      Model estimates of precinct-level turnout by race (a) 
+#|      and Democratic support by race (b).
+#|      Points are sized in proportion to their voting-age population, 
+#|      and colored according to their Black share of the voting-age population.
+#|      All axes are transformed to a logit scale.
+include_graphics("figures/al_ei.pdf")
+```
 
@@
 
-Because of the latent factor model, we do not require that vote shares by race and precinct, when tallied by race, necessary equal the observed vote share in the precinct.
+Because of the latent factor model, we do not require that vote shares by race and precinct, when tallied by race, necessarily equal the observed vote share in the precinct.
 As a result we choose to model the vote shares directly, which is more computationally efficient.
@@
 Michelle Wu and Annissa Essaibi-George advanced to the general from the preliminary election.
-We fix the model's loading matrix $\vb L$ to the the posterior mean of their estimated loadings.
+We fix the model's loading matrix $\vb L$ to the posterior mean of their estimated loadings.
 We draw $s_{jk}$ from a $\Beta(20\bar s, 20(1-\bar s))$ distribution, where $\bar s$ is each candidate's share of the two-candidate vote from the preliminary election (59.7% Wu -- 40.3% Essaibi George).

### 2022-05-25 paper/paper.Rmd
edit

@@
 New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
-Appendix \ref{app:corr} calculates these correlations across eight states.
-The most consistent correlation is that between our estimate of differential partisan harm and the expected number of seats won by each party, which is usually around 0.98, and always above 0.66 in the eight states we study.
+Appendix \ref{app:corr} shows these correlations across simulated plans in eight states.
+The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which is usually around 0.98, and always above 0.66 in the eight states we study.
 

### 2022-06-17 paper/paper.Rmd
repo rename
Update paper.Rmd
abstract + PO notation

@@
       Research Workshop and Imai Research Group  for helpful comments and discussion.
-      Replication code and data are available at <https://github.com/CoryMcCartan/representation>.
+      Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
     department: Department of Statistics
@@
 abstract: |
-  We propose a unified framework for evaluating the impact of a districting plan on voters overall and in different groups.
-Considering the harm done to individual voters by their inability to elect their chosen candidate, and taking into account the range of counterfactual possibilities, we derive a set of measures that flexibly adapts to local geography and constraints, different types of elections, and different groups of study.
-We provide recommendations for estimating these measures from historical election returns and discuss connections to commonly used redistricting metrics.
-Finally, we apply our framework to study partisan gerrymandering in New Jersey, voting rights litigation in Alabama, and the racial impacts of city council district boundaries in Boston.
+  Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
+But these measures cannot be easily adapted to other groups, including those defined by race, class, or geography.
+Nor are they applicable to single- or no-party contexts such as local redistricting.
+To overcome these limitations, we propose a unified framework of harm for evaluating the impacts of a districting plan on individual voters and the groups to which they belong.
+We consider a voter harmed if their chosen candidate is not elected under the current plan, but would be under a different plan.
+Harm improves on existing measures by both focusing on the choices of individual voters and directly incorporating counterfactual plans.
+We discuss strategies for estimating harm, and demonstrate the utility of our framework through analyses of partisan gerrymandering in New Jersey, voting rights litigation in Alabama, and racial dynamics of Boston City Council elections.
 keywords:
@@
 For an election $v$, we can tally the votes for each candidate; the candidate with the most votes wins.
-We denote the winner in district $d$ of districting plan $q$ as $\win_d(v,q)$.
+Using potential outcomes notation, we denote the winner of voter $i$'s district under districting plan $q$ and election $v$ as $\win_i(v,q)$.
 
@@
 Then we may define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as $$
-h_i(v,q,\tilde q) = \ind\{v(i) \neq \win_{q(i)}(v, q)\}
-        \ind\{v(i) = \win_{\tilde q(i)}(v, \tilde q)\},
+h_i(v,q,\tilde q) = \ind\{v(i) \neq \win_i(v, q)\}
+        \ind\{v(i) = \win_i(v, \tilde q)\},
 $$that is, a voter is harmed only if they are not represented by the candidate of their choice under $q$ and they would have been represented by their chosen candidate under the counterfactual plan $\tilde q$.
@@
 Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in election $v$ becomes$$
-\E[H(q)\mid V=v] = \Pr[v(I)\neq\win_{q(I)}(v, q)
-    \text{ and }v(I)=\win_{\tilde Q(I)}(v, \tilde Q)],
+\E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q)
+    \text{ and }v(I)=\win_I(v, \tilde Q)],
 $$that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
@@
 For a random voter $I$ in precinct $j$, we have for every plan $q$ that the probability of a Democrat winning their district is $$
-\Pr(\win_{q(I)}(V, q)=\dem\mid I\in j)
+\Pr(\win_I(V, q)=\dem\mid I\in j)
 = \Pr\qty(S_{q^{-1}(j),V} > \half)
@@
 \begin{equation}\label{eq:appharm1}
-\Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j)
+\Pr(\win_I(V, q)=\rep\text{ and }\win_I(V, \tilde q)=\dem\mid I\in j)
 = \Pr\qty(S_{q^{-1}(j),V} < \half, S_{\tilde q^{-1}(j),V} > \half).
@@
 \begin{equation}\label{eq:appharm2}
-\Pr(\win_{q(I)}(V, q)=\rep\text{ and }\win_{\tilde q(I)}(V, \tilde q)=\dem\mid I\in j)
+\Pr(\win_I(V, q)=\rep\text{ and }\win_I(V, \tilde q)=\dem\mid I\in j)
 = \max\qty{\Phi\qty(\frac{\gamma(\bar S_{\tilde q^{-1}(j)})}{\sigma^2_\mathrm{hist}})

### 2022-06-17 explore_elec.R

@@ -1,0 +1,50 @@
+library(alarmdata)
+library(lme4)
+
+d_house = read_csv(here("data-raw/1976-2020-house.csv"), show_col_types=FALSE) |>
+    group_by(year, state, district) |>
+    filter(sum(str_starts(party, "DEMOCRAT")) == 1,
+           sum(str_starts(party, "REPUBLICAN")) == 1) |>
+    ungroup() |>
+    filter(str_starts(party, "(DEMOCRAT|REPUBLICAN)")) |>
+    transmute(year = year,
+              state = state_po,
+              district = str_c(state, "-", district, "-", floor(year/10)*10),
+              party = str_to_lower(str_sub(party, 1, 3)),
+              votes = candidatevotes) |>
+    pivot_wider(names_from=party, values_from=votes) |>
+    mutate(dshare = dem / (dem + rep))
+
+m_distr = lmer(dshare ~ (1 | district) + (1 | year), data=d_house)
+
+ranef_distr = ranef(m_distr)$district |>
+    rownames_to_column("dist") |>
+    as_tibble() |>
+    rename(eff_dist = `(Intercept)`)
+ranef_year = ranef(m_distr)$year |>
+    rownames_to_column("year") |>
+    as_tibble() |>
+    rename(eff_year = `(Intercept)`)
+
+
+nj = alarm_50state_map("NJ")
+
+d = as_tibble(nj) |>
+    transmute(prec = GEOID,
+              dist = cd_2010,
+              turn_16 = pre_16_dem_cli + pre_16_rep_tru,
+              turn_18 = uss_18_dem_men + uss_18_rep_hug,
+              dshare_16 = pre_16_dem_cli / turn_16,
+              dshare_18 = uss_18_dem_men / turn_18) |>
+    pivot_longer(-prec:-dist, names_to=c("var", "year"), names_sep="_") |>
+    pivot_wider(names_from=var) |>
+    drop_na() |>
+    mutate(dvote = as.integer(round(dshare * turn)),
+           year = str_c("20", year),
+           dist = str_c("NJ-", str_pad(2, 3, pad="0"), "-2010")) |>
+    left_join(ranef_distr, by="dist") |>
+    left_join(ranef_year, by="year") |>
+    mutate(dshare_adj = dshare - eff_dist - eff_year)
+
+
+m_prec = lmer(dshare ~ (1 | prec) + (1|dist:year), data=d)

### 2022-06-17 paper/paper.Rmd
acknowledgements

@@
       To whom correspondence should be addressed.
-      The authors thank participants of the Harvard American Politics
-      Research Workshop and Imai Research Group  for helpful comments and discussion.
+      The authors thank Kosuke Imai, Gary King, Shiro Kuriwaki, 
+      Tyler Simko, Nicholas Stephanopolous, and 
+      participants of the Harvard American Politics Research Workshop 
+      and Imai Research Group for helpful comments and discussion.
       Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.

### 2022-06-17 paper/paper.Rmd
Update paper.Rmd

--- 2022-06-17 paper/paper.Rmd
@@
       To whom correspondence should be addressed.
-      The authors thank participants of the Harvard American Politics
-      Research Workshop and Imai Research Group  for helpful comments and discussion.
+      The authors thank Kosuke Imai, Gary King, Shiro Kuriwaki, 
+      Tyler Simko, Nick Stephanopolous, and 
+      participants of the Harvard American Politics Research Workshop 
+      and Imai Research Group for helpful comments and discussion.
       Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.

### 2022-07-01 paper/paper.Rmd
new elec model; revise NJ

@@
 bibliography: references.bib
-biblio-style: chicago
+biblio-style: apsr
 output:
@@
 To illustrate the connections in the preceding section and to demonstrate the value of the proposed measures in a real-world setting, we provide a brief analysis of congressional districting in the state of New Jersey.
-As discussed above, many gerrymandering measures function well in highly competitive states, but may produce more counterintuitive results in other states.
+As discussed above, many gerrymandering measures often function well in highly competitive states, but may produce more counterintuitive results in other states.
 New Jersey, with an average Democratic two-party vote share of 56.6% across our two baseline elections (2016 presidential and 2018 senatorial), will demonstrate that this is not a problem unique to lopsided states.
@@
 For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^7] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
-The distribution of these measures, along with their pairwise relationships, are shown in Figure \ref{fig:nj-pairs}.
+The distribution of these measures is shown in Figure \ref{fig:nj-meas}.
 
@@
 
-```{r nj-pairs, fig.width=7, fig.height=7}
+```{r nj-meas, fig.width=7}
 #| fig.cap: |
-#|      Scatterplots for all pairs of measures.
-#|      The Democratic and Republican gerrymanders are shown
-#|      with the blue and red lines and crosses.
+#|      The distribution of the existing and proposed measures across a sample of plans,
+#|      with the values for two gerrymanders overlaid as vertical lines.
 #|      The dashed line indicates a value of zero.
-#|      Below the diagonal, numbers indicate the absolute value of the correlation between variables.
-include_graphics("figures/nj_pairs.pdf")
+include_graphics("figures/nj_meas.pdf")
 ```
@@
 We first examine the values of each metric on the two example plans alone.
-Differential partisan harm correctly identifies each gerrymander as such, with a large negative value for the Democratic gerrymander and a large positive value for the Republican gerrymander.
-These values are relatively directly interpretable; for example, the differential harm score of $-0.221$ means that an average Democratic voter is 22.1% less likely to be misrepresented than their average Republican counter-partisan.
-The efficiency gap and declination measures also correctly identify the gerrymanders, though they are much more discrete, and declination gives a much larger score in absolute value to the Democratic plan than to the Republican.
-In contrast, the partisan bias and mean-median measures incorrectly classify the Republican gerrymander as a fair plan and a Democratic-leaning outlier, respectively.
+The estimate of differential partisan harm correctly identifies the partisan skew of each gerrymander, with a large negative value for the Democratic gerrymander and a small positive value for the Republican gerrymander.
+These values are relatively directly interpretable; for example, the differential harm score of $-0.109$ means that an average Democratic voter is 10.9% less likely to be misrepresented than their average Republican counter-partisan.
+The efficiency gap and declination measures also correctly identify the gerrymanders, though they are much more discrete, and the efficiency gap gives roughly equal scores in absolute value to the Democratic and Republican plans, in comparison to differential harm and declination, which identify the Democratic gerrymander as more extreme.
+In contrast, the usual partisan bias estimator classifies the Republican gerrymander as favoring Democrats and vice versa.
+While the mean-median measure identifies the Democratic gerrymander, it classifies the Republican gerrymander as a an even more extreme Democratic-leaning outlier.
 
-Next, we consider the distribution of these statistics across the sampled plans, but this doesn't help much, either.
-The expected number of Democratic seats naturally identifies the gerrymanders (two-sided $p=0.0002$) and the direction of bias, and as discussed above this measure is extremely correlated with differential partisan harm.
-The gerrymanders are also an outlier under the efficiency gap and declination measures also (both $p=0.0002$).
-While the gerrymanders are both an outlier for the mean-median score, only the Democratic plan is an outlier according to the partisan bias measure ($p=0.0022$, versus $p=0.099$ for the Republican plan).
-These measures are weakly correlated with the expected number of Democratic seats, and show substantial variation even for fixed values of total seats won at baseline.
+Considering the distribution of these statistics across the sampled plans doesn't help much, either.
+The expected number of Democratic seats naturally identifies the gerrymanders (two-sided $p=0.0002$) and the direction of bias.
+The gerrymanders are also an outlier under the efficiency gap and declination measures also (both $p=0.0002$), and in the correct direction.
+While the Republican gerrymander is an outlier for the partisan bias and mean-median score ($p=0.0224$ and $p=0.0004$), the Democratic gerrymander is only an outlier for the mean-median score($p=0.0004$ compared to $p=0.349$ for partisan bias).
+All of the plans identified as outliers under these two metrics are identified in the wrong direction.
 
-There is also disagreement between the measures about the natural bias of districting plans in New Jersey: the median sampled plan has a partisan bias of $0.047$, indicating a natural Republican bias, but the median mean-median score is just $0.0069$, and the median differential partisan harm is $-0.13$, indicating a natural Democratic bias.
+There is also disagreement between the measures about the natural bias of districting plans in New Jersey: the median sampled plan has a partisan bias of $0.020$, indicating a natural Republican bias, but the median mean-median score is just $0.0040$, and the median differential partisan harm is $-0.032$, indicating a natural Democratic bias.
 Which is correct?
 It depends on what one means by "natural bias," though we submit that differential harm is one good way to measure it.
-Democrats do win 8.47 seats on average across the simulations, which is more than the 6.79 seats that would be expected under a completely proportional outcome.
-The key benefit here is that differential harm does not push for localized gerrymanders to overcome the geographic imbalance in partisanship.
+Democrats do win 8.4 seats on average across the simulations, which is more than the 6.8 seats that would be expected under a completely proportional outcome.
+It is important to note that a value of zero estimated differential harm does not necessarily correspond to proportionality; the simulated plan with an estimated differential harm closest to zero yieled 8.1 expected Democratic seats.
 
-Both gerrymanders excessively harm the average voter, with the Democratic plan doing so at a moderately higher rate ($p=0.0002$ for the Democratic plan, $p=0.0004$ for the Republican plan).
+Both gerrymanders excessively harm the average voter, with the Democratic plan doing so at a moderately higher rate ($p=0.0004$ for the Democratic plan, $p=0.0008$ for the Republican plan).
 Notably, the plans which minimize total harm are also those which have near-zero differential partisan harm.
@@
 
+Figure \ref{fig:nj-pairs} in the appendix expands on Figure \ref{fig:nj-meas} to also show the correlations between these various measures.
+Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
+Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats. 
 New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
 Appendix \ref{app:corr} shows these correlations across simulated plans in eight states.
-The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which is usually around 0.98, and always above 0.66 in the eight states we study.
+The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which is usually around 0.82, and always above 0.66 in the eight states we study.
 
@@
 
+Figure \ref{fig:nj-pairs} expands on Figure \ref{fig:nj-meas} in the main text with pairwise scatterplots and correlation summaries.
+
+```{r nj-pairs, fig.width=7, fig.height=7}
+#| fig.cap: |
+#|      Scatterplots for all pairs of measures.
+#|      The Democratic and Republican gerrymanders are shown
+#|      with the blue and red lines and crosses.
+#|      The dashed line indicates a value of zero.
+#|      Below the diagonal, numbers indicate the absolute value of the correlation between variables.
+include_graphics("figures/nj_pairs.pdf")
+```
+
 Table \ref{tab:corr} below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section \ref{sec:nj} for eight states which cover a wide range of partisan leans and number of seats:

### 2022-07-01 R/00_setup_utils.R

@@ -19,14 +19,15 @@ NAMES_RACE = c("White", "Black", "Hispanic", "other")
 GOP_DEM = c("#A0442C", "#B25D4C", "#C27568", "#D18E84", "#DFA8A0",
             "#EBC2BC",  "#F6DCD9", "#F9F9F9", "#DAE2F4", "#BDCCEA",
             "#9FB6DE", "#82A0D2", "#638BC6", "#3D77BB", "#0063B1")
+GOP = "#CA685C"
+DEM = "#166AB0"
 scale_fill_party_c = function(name="Democratic share", midpoint=0.5, limits=0:1,
                               labels=percent, oob=squish, ...) {
     scale_fill_gradient2(name=name, ..., low = GOP_DEM[1], high = GOP_DEM[15],
                          midpoint=midpoint, limits=limits, labels=labels, oob=oob)
 }
 scale_color_party_d = function(...) {
-    scale_color_manual(..., values=c(GOP_DEM[2], GOP_DEM[14]),
-                       labels=c("Rep.", "Dem."))
+    scale_color_manual(..., values=c(GOP, DEM), labels=c("Rep.", "Dem."))
 }
 
 scale_fill_ideo = \(...) scale_fill_wa_c(palette="stuart", ...)
@@ -40,6 +41,7 @@ pos_part = function(x) {
     x[x < 0] = 0
     x
 }
+
 # historical congressional national shifts
 d_hist = local({
     dem_hist = c(0.575, 0.559, 0.537, 0.505, 0.552, 0.521, 0.543, 0.533, 0.521, 0.501, 0.447, 0.482, 0.473, 0.471, 0.452, 0.468, 0.523, 0.532, 0.449, 0.488, 0.455, 0.48, 0.534)
@@ -48,8 +50,8 @@ d_hist = local({
     list(sd=sd(dem_tpp), df=length(dem_hist)-1)
 })
 sd_app = d_hist$sd * sd(rt(1e5, df=d_hist$df))
-k_t = function(sd=d_hist$sd) {
-    function(x) pt(qlapl(x)/sd, df=d_hist$df)
+k_t = function(sd=d_hist$sd, df=d_hist$df) {
+    function(x) pt(qlapl(x)/sd, df=df)
 }
 e_approx = function(a, b) {
     y = sd_app * (dnorm(a/sd_app) - dnorm(b/sd_app)) / (pnorm(b/sd_app) - pnorm(a/sd_app))
[…]

### 2022-07-01 R/01_metrics.R

@@ -13,25 +13,32 @@ district_group = function(plans, group) {
     mat_by_prec(as.matrix(plans), m_grp)
 }
 
-#' Voter harm for local representation
+#' Partisan harm
 #'
-#' @param map a `redist_map` object.
-#' @param group_pop column of `map` containing the group population of each precinct.
-#' @param distr_grp a matrix from `district_group()`
-#' @param idx_1,idx_2 indices for each of the comparison groups
-#' @param kernel a vectorized function mapping [0, 1] to [0, 1]
+#' @param plans a `redist_plans` object.
+#' @param dem_share column of `map` containing the Democratic share of each precinct.
+#' @param dvote the Democratic vote in each precinct
+#' @param rvote the Republican vote in each precinct
+#' @param elec_model_spec A list with elements `year` and `resid` containing the
+#'   standard deviation for year effects (constant across districts) and
+#'   residual effects (vary by district)
+#' @param idx_1 indices of the district to evaluate the harm form
+#' @param idx_2 indices of the counterfactual districts
 #'
 #' @returns a numeric matrix matching `idx_1`: first row is harm to in-group,
-#'   second row is out-group.
-harm = function(plans, group_share, dvote, rvote,
-                idx_1=seq_len(ncol(as.matrix(plans))),
-                idx_2=seq_len(ncol(as.matrix(plans))),
-                ker=k_step) {
-    x = matrix(
-        ker(eval_tidy(enquo(group_share), plans)),
-        nrow=attr(plans, "ndists")
-    )
-    harm_helper(as.matrix(plans), x, dvote, rvote, idx_1, idx_2)
+#'   second row is out-group, third row is total harm.
+partisan_harm = function(plans, dem_share, dvote, rvote,
+                         elec_model_spec=list(distr=0.1586, year=0.02659, resid=0.079858),
+                         idx_1=seq_len(ncol(as.matrix(plans))),
+                         idx_2=seq_len(ncol(as.matrix(plans)))) {
+    ndists = attr(plans, "ndists")
+    m_dem = matrix(eval_tidy(enquo(dem_share), plans), nrow=ndists)
+    shift_elec = rnorm(length(idx_2), sd=elec_model_spec$year)
[…]

### 2022-07-01 R/03_sim_nj.R

@@ -1,18 +1,22 @@
 nj = read_rds(here("data/NJ_cd_final_vtd_20.rds")) %>%
     redist_map(pop_tol=0.01, ndists=12, adj=.$adj)
 
-N_sim = 2500
+N_sim = 10e3
+n_runs = 4
 if (!file.exists(sim_path <- here("data/nj_sims.rds"))) {
-    plans = redist_smc(nj, N_sim, counties=county, pop_temper=0.005,
-                       runs=4, verbose=TRUE)
+    plans = redist_smc(nj, N_sim/n_runs, counties=county, pop_temper=0.005,
+                       runs=n_runs, verbose=TRUE)
 
     # gerrymanders
     set.seed(5118)
-    opt_dem = redist_shortburst(nj, scorer_group_pct(nj, ndv, ndv+nrv, 11), max_bursts=800)
+    opt_dem = redist_shortburst(nj, scorer_group_pct(nj, ndv, ndv+nrv, 11),
+                                max_bursts=1000, burst_size=15, return_all=FALSE)
     set.seed(5118)
-    opt_rep = redist_shortburst(nj, scorer_group_pct(nj, nrv, ndv+nrv, 6), max_bursts=800)
+    opt_rep = redist_shortburst(nj, scorer_group_pct(nj, nrv, ndv+nrv, 6),
+                                max_bursts=1000, burst_size=15, return_all=FALSE)
 
     plans = plans %>%
+        subset_sampled() %>%
         add_reference(last_plan(opt_dem), "dem_gerry") %>%
         add_reference(last_plan(opt_rep), "rep_gerry")
     write_rds(plans, sim_path, compress="xz")
@@ -23,19 +27,19 @@ if (!file.exists(sim_path <- here("data/nj_sims.rds"))) {
 statewide = with(nj, sum(ndv)/sum(ndv+nrv))
 prop_seats = round(attr(nj, "ndists") * statewide)
 
-pl = calc_plans_stats(plans, nj, ndv, nrv)
+pl = calc_plans_stats(plans, nj, ndv, nrv, elec_model_spec)
 pl$plan$chain = by_plan(plans$chain, ndists=12)
 
 # get 100% accurate harm for 2 gerrymanders
-hh = harm(pl$distr, dem, nj$ndv, nj$nrv, idx_1=1:2, idx_2=(2+1:N_sim), ker=k_t())
[…]

### 2022-07-01 R/08_correlation.R

@@ -6,7 +6,7 @@ library(kableExtra)
 
 states = c("IA", "NC", "WI", "MA", "UT", "NJ", "MS", "CA")
 
-ker_t = k_t()
+ker_t = k_t(sd=with(elec_model_spec, sqrt(year^2 + resid^2)))
 corr_state = function(abbr) {
     st_map = alarm_50state_map(abbr)
     dvote = st_map$ndv
@@ -18,8 +18,8 @@ corr_state = function(abbr) {
     attr(st_plans, "ndists") = nd
     cat("Downloaded", abbr, "\n")
 
-    idx_2 = sample(seq_len(ncol(as.matrix(st_plans)))[-1], 200)
-    hh = harm(st_plans, ndshare, st_map$ndv, st_map$nrv, idx_2=idx_2, ker=ker_t)
+    idx_2 = sample(seq_len(ncol(as.matrix(st_plans)))[-1], 250)
+    hh = partisan_harm(st_plans, ndshare, st_map$ndv, st_map$nrv, elec_model_spec, idx_2=idx_2)
 
     pl_corr = st_plans %>%
         subset_sampled() %>%

### 2022-07-01 explore_elec.R

@@ -1,50 +1,0 @@
-library(alarmdata)
-library(lme4)
-
-d_house = read_csv(here("data-raw/1976-2020-house.csv"), show_col_types=FALSE) |>
-    group_by(year, state, district) |>
-    filter(sum(str_starts(party, "DEMOCRAT")) == 1,
-           sum(str_starts(party, "REPUBLICAN")) == 1) |>
-    ungroup() |>
-    filter(str_starts(party, "(DEMOCRAT|REPUBLICAN)")) |>
-    transmute(year = year,
-              state = state_po,
-              district = str_c(state, "-", district, "-", floor(year/10)*10),
-              party = str_to_lower(str_sub(party, 1, 3)),
-              votes = candidatevotes) |>
-    pivot_wider(names_from=party, values_from=votes) |>
-    mutate(dshare = dem / (dem + rep))
-
-m_distr = lmer(dshare ~ (1 | district) + (1 | year), data=d_house)
-
-ranef_distr = ranef(m_distr)$district |>
-    rownames_to_column("dist") |>
-    as_tibble() |>
-    rename(eff_dist = `(Intercept)`)
-ranef_year = ranef(m_distr)$year |>
-    rownames_to_column("year") |>
-    as_tibble() |>
-    rename(eff_year = `(Intercept)`)
-
-
-nj = alarm_50state_map("NJ")
-
-d = as_tibble(nj) |>
-    transmute(prec = GEOID,
-              dist = cd_2010,
-              turn_16 = pre_16_dem_cli + pre_16_rep_tru,
-              turn_18 = uss_18_dem_men + uss_18_rep_hug,
-              dshare_16 = pre_16_dem_cli / turn_16,
-              dshare_18 = uss_18_dem_men / turn_18) |>
-    pivot_longer(-prec:-dist, names_to=c("var", "year"), names_sep="_") |>
-    pivot_wider(names_from=var) |>
-    drop_na() |>
-    mutate(dvote = as.integer(round(dshare * turn)),
-           year = str_c("20", year),
-           dist = str_c("NJ-", str_pad(2, 3, pad="0"), "-2010")) |>
-    left_join(ranef_distr, by="dist") |>
-    left_join(ranef_year, by="year") |>
-    mutate(dshare_adj = dshare - eff_dist - eff_year)
-
-
-m_prec = lmer(dshare ~ (1 | prec) + (1|dist:year), data=d)

### 2022-07-01 paper/paper.Rmd
stashed changes

@@
       Tyler Simko, Nick Stephanopolous, and 
-      participants of the Harvard American Politics Research Workshop 
+      participants of the Fall 2021 Harvard American Politics Research Workshop 
       and Imai Research Group for helpful comments and discussion.
@@
 
-## Racial Gerrymandering {#subsec:racial}
+## Vote Dilution {#subsec:racial}
 

### 2022-07-01 paper/paper.Rmd
elec model writeup

--- 2022-07-01 paper/paper.Rmd
@@
 
+```{=html}
+<!--
 While not strictly necessary, modeling of $V$ and $\tilde Q$ independently will simplify the situation considerably.
@@
 But as our suggested model for election outcomes $V$ will be quite coarse, not accounting for candidate factors or policy preferences, the marginal effect of this independence assumption on our conclusions will be minimal.
-
+-->
+```
 ## Modeling Districting Plans
@@
 
-Motivated primarily by computational convenience and simplicity, we use a *proportionate uniform shift* model, which focuses on swings in the national political environment as the main determinant of election outcomes.
-Election outcomes are modeled as a random deviation from a baseline outcome, and these deviations are assumed to consist of a constant fraction of voters of one party in each precinct switching to the other.
-The baseline itself can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
-The proportionate uniform shift model is nearly identical to the uniform partisan swing model of @butler1951swing for most precincts and districts, but is more computationally convenient and avoids issues of vote shares exceeding 100% or falling below 0%.
-Despite their simplicity, uniform-swing--type models are highly accurate for out-of-sample prediction [@katz2020], and are intuitive and easy to implement.
-More details about our proportionate uniform shift model and how it may be used to efficiently estimate the harm measures may be found in Appendix \ref{app:formulas}.
+Motivated primarily by computational convenience and simplicity, we adopt the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as $$
+    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
+$$ where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
+The baseline $\bar{s}_d$ can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
+Since $\bar{s}_d$ is calculated for the particular configuration of district $d$, this model takes into account the specific districting plan under consideration.
+We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
+Appendix \ref{app:elecmodel} contains more details about how we estimate and apply this model.
 
 More complicated election outcome models are certainly possible and straightforward to use with the measures proposed here.
-For example, one could model district deviations as a systematic national shift plus a local idiosyncratic effect, as is done in the Stochastic Uniform Partisan Swing model of @gelman1994unified.
-For ease and efficiency of computation, we stick to the simple proportionate uniform shift model here.
+For example, one could operate on the logit scale, or add terms to model incumbency effects.
 
@@
 With a model for election outcomes and districting plans in hand, calculation of $H(q)$ and $\Delta H(q)$ is straightforward.
-For each sample $\tilde q$ of a counterfactual districting plan, we use the election model to come up with a probability that voters in each precinct are represented by the candidate of their choice under $\tilde q$ but not under $q$.
-When the groups used in $\Delta H(q)$ vary within precincts, we must repeat these computations for each group.
-Then by summing across precincts and/or groups and dividing by the appropriate totals, we arrive at values for $H(q)$ and $\Delta H(q)$.
-Appendix \ref{app:formulas} contains more details, including the specific formulas used to compute $H(q)$ and $\Delta H(q)$ for two-party elections from a set of past election returns and a sample of districting plans from a redistricting sampling algorithm.
+For each sample $\tilde q$ of a counterfactual districting plan, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
+We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^3]
+We tally the total harm done to Democratic and Republican voters across the map and divide by the appropriate totals to get a single draw of $H(q)\mid I\in\dem$ and $H(q)\mid I\in\rep$.
+When other groups (such as racial groups) are used in $\Delta H(q)$, we must repeat these computations for each group.
+Finally, we average across all of the samples, and add or subtract across groups, to produce estimates of $E[H(q)]$ and $\Delta_{\dem-\rep} H(q)$.
+
+[^3]: For this, we use the baseline count of Democratic or Republican votes.
+    This is a slight simplification, since because the vote share changes across random draws, the number of Democratic and Republican voters does, too.
+    To deal with this issue completely necessitates modeling elections at the precinct level, which requires much more data and computational expenditure.
 
@@
 The *declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
-Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015], which measures the difference in the number of the votes wasted by each party.[^3]
+Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015], which measures the difference in the number of the votes wasted by each party.[^4]
 
-[^3]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.
+[^4]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.
 
@@
 As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
-This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^4]
+This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^5]
 One advantage of the simpler metrics such as the efficiency gap is that they may quickly calculated using district vote shares alone.
 
-[^4]: For partisan bias, both researchers and practitioners have largely relied on simple uniform-shift models similar to the one we use in this work.
+[^5]: For partisan bias, both researchers and practitioners have largely relied on simple uniform-shift models similar to the one we use in this work.
 
@@
 It is this confounding which explains many of the concerns that have been raised over the existing measures.
-The standard calculation of partisan bias, $\beta(\half)$, can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.[^5]
+The standard calculation of partisan bias, $\beta(\half)$, can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.[^6]
 And none of the measures take into account any aspect of political geography, and therefore suffer in states like Massachusetts, where, for example, the efficiency gap takes a value generally indicating an extreme Democratic gerrymander.
 
-[^5]: This has been called by some the "Utah paradox" [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
+[^6]: This has been called by some the "Utah paradox" [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
     Even when interpreted carefully, however, partisan bias requires significant extrapolation from observed election outcomes in lopsided states---how realistic is it to evaluate the counterfactual where Democrats run even with Republicans statewide in Utah?
@@
 Despite these differences, there are also key empirical similarities in many cases.
-The efficiency gap and differential partisan harm are often highly correlated (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
+The efficiency gap and differential partisan harm are often highly correlated (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^7]
 Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with differential partisan harm as well.
@@
 
-[^6]: The differences stem from the efficiency gap considering a vote to be wasted if it is in excess of the 50% threshold needed to win, while the harm framework does not penalize such votes, since every such voter is happily represented by their candidate of choice.
+[^7]: The differences stem from the efficiency gap considering a vote to be wasted if it is in excess of the 50% threshold needed to win, while the harm framework does not penalize such votes, since every such voter is happily represented by their candidate of choice.
     This definition of wasted votes is also at the root of other criticisms of the efficiency gap, which note its tendency to penalize proportionality and reward 3-to-1 districts [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
@@
 One such plan is shown in Figure \ref{fig:nj-maps}(d).
-For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^7] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
+For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^8] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
 The distribution of these measures is shown in Figure \ref{fig:nj-meas}.
 
-[^7]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
+[^8]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
 
@@
 
-We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.[^8]
+We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.[^9]
 
-[^8]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
+[^9]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
 
@@
 Nine of the councilors are elected from contiguous city council districts, and the remaining four councilors are elected at-large, with all councilors serving two-year terms.
-Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^9]
+Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^10]
 This section will study the extent to which the current district boundaries differentially harm voters of different racial groups, in comparison to a completely race-neutral districting process.
@@
 
-[^9]: City of Boston charter.
+[^10]: City of Boston charter.
     Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
@@
 
-To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^10] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
+To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^11] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
 Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail.
 
-[^10]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+[^11]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
 
@@
 ```
-# Computing Harm Measures from Precinct-level Election Data {#app:formulas}
+# Two-party Election Model {#app:elecmodel}
 
-In practice, districting plans are built from discrete geographic units (often precincts), and election returns are reported at this level as well.[^11]
-Throughout this appendix, let $j$ index the set of electoral precincts or other units.
-For concreteness here and without loss of generality we will use the Democratic party as the reference party, letting $s_{jv}$ denote the Democratic two-party vote share in precinct $j$ in election $v$.
-Write-in or minor third-party candidates can be safely ignored, since the overall winner will be precisely the two-party vote winner.
+The election model used in estimating harm is $$
+    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
+    \qquad \delta_v\sim\Norm(0, \sigma^2_\delta),
+    \quad \varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon).
+$$ As noted in the main text, $\bar{s}_d$ is computed by averaging historical election returns.
+This is done at the precinct level, since this is the most granular level that districting plans are usually built from and election returns are reported at.[^12]
 
-[^11]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+[^12]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
     Population information comes from Census blocks, which, when populated, are practically never split.
 
-## Proportionate Uniform Shift Details
+To estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$, we use historical U.S.
+congressional election returns from 1976-2020, provided by the MIT Election Data Science Lab (<https://doi.org/10.7910/DVN/IG0UN2>).
+We discard elections which were uncontested by either party, or which had more than one Democratic or Republican candidate, leaving 5,931 total elections.
+The data do not have complete coverage, and some states are missing or represented very sparsely.
+However, there are 38 states with at least 20 elections.
 
-The proportionate uniform shift model assumes that from one election to another, a constant fraction of *voters of one party* in each precinct switch to the other party, and that per-precinct turnout is constant across elections.[^12]
-This has several advantages over the uniform swing model, where a constant fraction of *all voters* shifts parties.
-First, the proportionate shift prevents hypothetical party vote shares from exceeding 1 or falling below 0 in any precinct.
-Second, as it has the effect of scaling vote shares across precincts by a constant multiple, it preserves the relative balance of a party's voters across precincts.
-For $\Delta_{\dem-\rep}H(q)$, which considers voters in each party separately, this helpfully simplifies computations.
+For each election, we compute the two-party Democratic vote share.
+We fit a linear mixed-effects model to these vote shares, with an overall intercept fixed effect, and random effects for districts and election years.
+To account for the fact that district boundaries change every decade, the district identifiers are built as a concatenation of the state, district number, and redistricting.
+For example, the Washington's 7th congressional district receives one random effect from 2002-2010 and a different random effect from 2012-2020, since new district boundaries were enacted in the 2010-2012 period.
+We do not account for any mid-decade redistricting which may occur as a result of litigation.
 
-[^12]: Everything carries through if one allows for shifts in the overall turnout level across elections, as long as (a) the relative turnout across precincts is fixed, and (b) turnout shifts are independent of shifts in party preferences.
-
-The proportionate shift may be mathematically expressed as linear shift on a transformed scale, not unlike a generalized linear model.
-Specifically, we define $\gamma:[0,1]\to\R$, $$
-\gamma(x) = -\thalf\sgn(x-\thalf)\log(1-2\qty|x-\thalf|),
-$$ which is the inverse CDF of a Laplace distribution with rate 2.[^13]
-Then the Democratic vote share in precinct $P_j$ and election $v$, notated here as $s_{jv}$, is modeled as $$
-\gamma(s_{jv}) = \gamma(\overline s_j) + \eps_v, \qquad \eps_v \iid \Norm(0, \sigma^2_{\mathrm{hist}}),
-$$ where $\overline{s}_j$ is the party baseline in the precinct, and $\sigma^2_\mathrm{hist}$ is estimated from the historical variability in nationwide popular vote shares.
-For congressional elections in the United States from 1974--2018, we have $\widehat\sigma^2_\mathrm{hist}=0.03723$, and so draw predictive deviations from $\mathrm{t}_{22}(0, \widehat\sigma^2_\mathrm{hist})$.
-
-[^13]: A rate of 2 is chosen so that the slope of the link function at $\half$ is exactly 1, meaning that transformed vote shares will be on approximately the same scale as the original shares.
-
-It will simplify matters following to have notation for the overall Democratic vote share in a district.
-For a precinct $j$, let $q^{-1}(j)$ denote the district it belongs to.
-Then let $$
-    S_{q^{-1}(j), v}=\frac{\sum_{l\in q^{-1}(j)} s_{lv}\avgvote_l}{\sum_{l\in q^{-1}(j)} \avgvote_l}
-$$ denote the vote share in the district, where $\avgvote_l$ counts the total number of votes in precinct $l$ at baseline.
-The baseline vote share in the district is then $\bar S_{q^{-1}(j)}$.
-
-## Estimating $H(q)$ and $\Delta H_{\dem-\rep}(q)$
-
-We will build up estimates of the overall $H(q)$ and $\Delta H_{\dem-\rep}(q)$ from precinct-by-group harm estimates.
-Since we assumed that elections are independent of districting plans, we can first average across elections and then across counterfactual plans.
-For a random voter $I$ in precinct $j$, we have for every plan $q$ that the probability of a Democrat winning their district is $$
-\Pr(\win_I(V, q)=\dem\mid I\in j)
-= \Pr\qty(S_{q^{-1}(j),V} > \half)
-= \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}),
-$$ where $\Phi$ is the standard Normal CDF.
-To estimate this probability, we simply replace $\Phi$ with the CDF of the $t_{22}$ distribution ($F_{t_{22}}$), and $\sigma^2_\mathrm{hist}$ with $\hat\sigma^2_\mathrm{hist}$.
-Extending one step further, for two plans $q$ and $\tilde q$, we have
-
-```{=tex}
-\begin{equation}\label{eq:appharm1}
-\Pr(\win_I(V, q)=\rep\text{ and }\win_I(V, \tilde q)=\dem\mid I\in j)
-= \Pr\qty(S_{q^{-1}(j),V} < \half, S_{\tilde q^{-1}(j),V} > \half).
-\end{equation}
-```
-Because of the uniform shift, for any election $v$, $S_{q^{-1}(j),v} - S_{\tilde q^{-1}(j),v}$ is constant.
-Therefore if $S_{q^{-1}(j),v}>S_{\tilde q^{-1}(j),v}$, this probability is zero (if a Democrat wins under $\tilde q$ they will definitely win under $q$, which has a higher Democratic vote share).
-But if $S_{q^{-1}(j),v}<S_{\tilde q^{-1}(j),v}$, we can rewrite \eqref{eq:appharm1} as the probability that $\eps_{V}$ falls in a particular range.
-Specifically, we have in general that
-
-```{=tex}
-\begin{equation}\label{eq:appharm2}
-\Pr(\win_I(V, q)=\rep\text{ and }\win_I(V, \tilde q)=\dem\mid I\in j)
-= \max\qty{\Phi\qty(\frac{\gamma(\bar S_{\tilde q^{-1}(j)})}{\sigma^2_\mathrm{hist}})
-- \Phi\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}), 0}.
-\end{equation}
-```
-If we condition on $V(I)=\dem$, then \eqref{eq:appharm2} is exactly the expected harm to voter $I$ under plans $q$ and $\tilde q$.
-The calculation is the same, up to sign changes, for the expected harm conditioning on $V(I)=\rep$.
-This completes the first step: averaging harm within precinct and groups and across elections.
-
-We can then estimate the average across counterfactual plans $\tilde Q$ with the usual Monte Carlo estimator: $$
-    \hat h_{j,\dem}(q)
-    \dfeq \frac{1}{N_\text{sim}} \sum_{k=1}^{N_\text{sim}}
-    \max\qty{F_{t_{22}}\qty(\frac{\gamma(\bar S_{\tilde q_k^{-1}(j)})}{\sigma^2_\mathrm{hist}})
-        - F_{t_{22}}\qty(\frac{\gamma(\bar S_{q^{-1}(j)})}{\sigma^2_\mathrm{hist}}), 0},
-$$ where $\hat h_{j,\dem}(q)$ is our estimate of the average harm to a Democratic voter in precinct $j$, averaging across $V$ and $\tilde Q$.
-The corresponding $\hat h_{j,\rep}(q)$ is computed similarly.
-
-Our final estimates $\hat H(q)$ and $\Delta_{\dem-\rep}\hat H(q)$ may then be straightforwardly calculated as weighted averages:
-
-```{=tex}
-\begin{align*}
-\hat H(q) &\dfeq
-\frac{\sum_{j} (\bar s_j\hat h_{j,\dem}(q) +
-(1-\bar s_j)\hat h_{j,\rep}(q))\avgvote_j}{\sum_{j} \avgvote_j} \qand \\
-\Delta_{\dem-\rep}\hat H(q) &\dfeq
-\frac{\sum_{j} (\hat h_{j,\dem}(q) - \hat h_{j,\rep}(q))\avgvote_j}{\sum_{j} \avgvote_j}.
-\end{align*}
-```
-Here, we rely on the fact that even across election shifts, the relative balance of each party's voters across precincts is preserved.
+The estimated standard deviations of the random effects are $\hat{\sigma}_\delta=0.0266$ for the election-year shocks and $\hat{\sigma}_\varepsilon=0.0653$ for the election-district-specific effects.
+We substitute these point estimates directly into our election model, which fails to account for uncertainty in the estimates.
 
@@
 
-Adapting the notation from the previous appendix, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $\votes_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jv}$ denote the Democratic vote share, and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Other}\}$.
+For precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $\votes_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jv}$ denote the Democratic vote share, and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Other}\}$.
 Our main parameters of interest are $\turn_{jr}$ and $\supp_{jr}$, the baseline turnout and Democratic support rate by race in the precinct.

### 2022-07-01 paper/paper.Rmd
re-run AL

@@
 
-This analysis is slightly more complicated than in New Jersey, since in order to understand the racial distribution of harm we need to estimate voting preferences by race.
-As is custom in VRA analyses, we use an ecological inference model.
-The model, detailed in Appendix \ref{app:alei}, jointly estimates turnout and vote choice by race and precinct across seven statewide elections.
-Voting is highly polarized by race outside of the cities of Birmingham and Montgomery.
-
-With these estimates in hand, we can estimate the average number of voters harmed in each precinct and racial group.
-Figure \ref{fig:al-maps}(c) shows the estimated geographic distribution of harm.
-Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures \ref{fig:al-maps}(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
-These areas are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.
-
 ```{r al-maps, fig.width=6.75}
@@
 
+This analysis is slightly more complicated than in New Jersey, since in order to understand the racial distribution of harm we need to estimate voting preferences by race.
+As is custom in VRA analyses, we use an ecological inference model.
+The model, detailed in Appendix \ref{app:alei}, jointly estimates turnout and vote choice by race and precinct across seven statewide elections.
+Voting is highly polarized by race outside of the cities of Birmingham and Montgomery.
+
+With these estimates in hand, we can estimate the average number of voters harmed in each precinct and racial group.
+Figure \ref{fig:al-maps}(c) shows the estimated geographic distribution of harm.
+Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures \ref{fig:al-maps}(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
+These areas are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.
+
 This intuition is confirmed by a closer analysis, summarized in Figure \ref{fig:al-maps}(f).
-Across the state, the estimated average harm is 0.082, which translates to around 161,000 voters.
-Of these harmed voters, around 136,000 are Black, while 25,000 are White.
-Similarly, 151,000 are Democratic, while only 10,000 are Republican.
+Across the state, the estimated average harm is 0.079, which translates to around 159,000 voters.
+Of these harmed voters, around 123,000 are Black, while 36,000 are White.
+Similarly, 137,000 are Democratic, while only 22,000 are Republican.
 Of course, the total number of each of these groups of voters is very different.
-Normalizing by group sizes, the expected harm for a Black Democrat is 0.25, for a Black Republican is 0.015, for a White Democrat is 0.091, and for a White Republican is 0.0073.
-Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have a differential harm of 0.158 versus White Democrats and a differential harm of 0.235 versus Black Republicans.
+Normalizing by group sizes, the expected harm for a Black Democrat is 0.22, for a Black Republican is 0.037, for a White Democrat is 0.083, and for a White Republican is 0.017.
+Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have a differential harm of 0.138 versus White Democrats and a differential harm of 0.183 versus Black Republicans.
 
@@
 
-The mean support $\mu^{(s)}$ receives an informative prior based on the racially polarized voting estimates of @kuriwaki2021geography, with $\mu^{(s)}\sim\Norm(\logit((0.18, 0.90, 0.35)), (0.04^2, 0.04^2, 0.2^2))$.
+The mean support $\mu^{(s)}$ receives an informative prior based on the racially polarized voting estimates of @kuriwaki2021geography, with $\mu^{(s)}\sim\Norm(\logit((0.18, 0.90, 0.35)), (0.03^2, 0.03^2, 0.2^2))$.
 These estimates were derived through multilevel regression and poststratification from survey data, calibrated to actual election results.

### 2022-07-01 R/06_alabama.R

@@ -20,7 +20,7 @@ d_votes = al_map %>%
 tbls = make_tables_alabama(d_votes, al_map)
 
 prior = list(loc=qlogis(c(0.18, 0.90, 0.35)), # Kuriwaki et al. RPV estimates
-             scale=c(0.04, 0.04, 0.2))
+             scale=c(0.03, 0.03, 0.2))
 fit = fit_ei(tbls, algorithm="vb", prior=prior,
              init=0, eta=0.15, adapt_engaged=F, tol_rel_obj=0.0002)
 
@@ -88,44 +88,32 @@ al_map = select(al_map, GEOID20:adj) %>%
 
 # Analysis --------
 
+N_sim = 2000
+
 m_plans = as_tibble(al_map) %>%
     select(starts_with("cd_pet")) %>%
     as.matrix() %>%
     `colnames<-`(NULL)
+idx = rep(1:4, each=N_sim/4)
+m_plans = m_plans[, idx]
 
+ker_t = k_t(sd=with(elec_model_spec, sqrt(year^2 + resid^2)))
 plans = redist_plans(m_plans, al_map, algorithm="none", wgt=NULL) %>%
     add_reference(al_map$cd_2020) %>%
     mutate(total_vap = tally_var(al_map, vap),
            vap_white = tally_var(al_map, vap_white),
            vap_black = tally_var(al_map, vap_black),
            dem = group_frac(al_map, ndv, ndv + nrv),
-           pr_dem = k_t()(dem))
+           pr_dem = ker_t(dem))
 
-m_pr = district_group(plans, pr_dem)
-h_prec_dem = rowMeans(pos_part(m_pr[, -1] - m_pr[, 1]))
-h_prec_rep = rowMeans(pos_part(-m_pr[, -1] + m_pr[, 1]))
+hh_white = partisan_harm(plans, dem, al_map$ndv_white, al_map$nrv_white,
+                         elec_model_spec, idx_1=1, idx_2=1+seq_len(N_sim))
+hh_black = partisan_harm(plans, dem, al_map$ndv_black, al_map$nrv_black,
+                         elec_model_spec, idx_1=1, idx_2=1+seq_len(N_sim))
+hh_other = partisan_harm(plans, dem, al_map$ndv_other, al_map$nrv_other,
+                         elec_model_spec, idx_1=1, idx_2=1+seq_len(N_sim))
 
[…]

### 2022-07-02 paper/paper.Rmd
update fig minis

@@
 
-# Motivation {#sec:motivation}
+# Minivation {#sec:motivation}
 
-To motivate the concepts of individual and differential harm, we will make use of three examples, which are stylized representations of U.S. states: Mininois, Minissouri, and Minichusetts.
-Mininois has one large Democratic city which can electorally dominate the rest of the state; Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority; and Minichusetts has a uniform 3--2 Democratic split in every precinct.
-Each example state has the same population and should be divided into 3 districts with population between 54 and 66.
+To introduce and motivate the concepts of individual and differential harm, we will make use of Minissouri, a stylized representations of Missouri.
+Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority.
+Minissouri has a population of 180 and should be divided into 3 districts with population between 54 and 66.
 Voters' preferences are assumed to be stable across elections, and turnout is 100%.
-Schematics for these examples are shown in Figure \ref{fig:mini-schematic}(a)--(c).
+A schematic for this example is shown in Figure \ref{fig:mini-schematic}.
 

### 2022-07-02 R/02_run_toy.R

@@ -1,1 +1,191 @@
-lapply(sort(Sys.glob(here::here("R/toy/*.R"))), source)
+# helpers ----
+
+## make grid ----
+bb <- st_sfc(st_polygon(list(rbind(c(0,0), c(1,0), c(1,1), c(0,0)))))
+grid <- st_as_sf(st_make_grid(bb, n = 6, crs = 3857)) %>%
+    rename(geometry=x)
+
+
+minissouri <- grid %>%
+    mutate(row = rep(6:1, each=6),
+           col = rep(1:6, 6),
+           city1 = row <= 2 & col >= 5,
+           city2 = row >= 5 & col <= 2,
+           pop = 4,
+           pop = if_else(city1, 10, pop),
+           pop = if_else(city2, 7, pop),
+           dem = 1,
+           dem = if_else(city1, 8, dem),
+           dem = if_else(city2, 5, dem),
+           rep = 3,
+           rep = if_else(city1, 2, rep),
+           rep = if_else(city2, 2, rep),
+           circle = round((col + row + city1) / 15 * pop)) %>%
+    relocate(geometry, .after=everything()) %>%
+    redist_map(ndists=3, pop_tol=0.1)
+
+
+## people for plotting -------
+
+make_people = function(pop, dem, circle, geometry, row, col, ...) {
+    if (!file.exists(path <- str_glue("data-raw/csq{pop}.txt"))) {
+        url = str_glue("http://hydra.nat.uni-magdeburg.de/packing/csq/txt/csq{pop}.txt")
+        download.file(url, path)
+    }
+    offset = as.numeric(st_centroid(geometry))
+    dem_vec = c(rep(TRUE, dem), rep(FALSE, pop-dem))
+    circ_vec = c(rep(TRUE, circle), rep(FALSE, pop-circle))
+    idx = sample(pop)
+    read_table(path, col_names=F, col_types=cols(.default="d")) %>%
+        select(-X1) %>%
+        as.matrix() %>%
+        apply(1, function(x) st_point(x/6.28 + offset), simplify=FALSE) %>%
+        st_sfc(crs=3857) %>%
+        st_as_sf() %>%
+        rename(geometry=x) %>%
+        mutate(row = row, col = col,
+               dem = dem_vec[idx],
+               circle = circ_vec[idx])
+}
+
+minissouri_people = minissouri %>%
[…]

### 2022-07-02 R/toy/00_toy_setup.R

@@ -1,91 +1,0 @@
-# make grid ----
-bb <- st_sfc(st_polygon(list(rbind(c(0,0), c(1,0), c(1,1), c(0,0)))))
-grid <- st_as_sf(st_make_grid(bb, n = 6, crs = 3857)) %>%
-    rename(geometry=x)
-
-
-minissouri <- grid %>%
-    mutate(row = rep(6:1, each=6),
-           col = rep(1:6, 6),
-           city1 = row <= 2 & col >= 5,
-           city2 = row >= 5 & col <= 2,
-           pop = 4,
-           pop = if_else(city1, 10, pop),
-           pop = if_else(city2, 7, pop),
-           dem = 1,
-           dem = if_else(city1, 8, dem),
-           dem = if_else(city2, 5, dem),
-           rep = 3,
-           rep = if_else(city1, 2, rep),
-           rep = if_else(city2, 2, rep),
-           circle = round((col + row + city1) / 15 * pop)) %>%
-    relocate(geometry, .after=everything()) %>%
-    redist_map(ndists=3, pop_tol=0.1)
-
-
-# people for plotting -------
-
-make_people = function(pop, dem, circle, geometry, row, col, ...) {
-    if (!file.exists(path <- str_glue("data-raw/csq{pop}.txt"))) {
-        url = str_glue("http://hydra.nat.uni-magdeburg.de/packing/csq/txt/csq{pop}.txt")
-        download.file(url, path)
-    }
-    offset = as.numeric(st_centroid(geometry))
-    dem_vec = c(rep(TRUE, dem), rep(FALSE, pop-dem))
-    circ_vec = c(rep(TRUE, circle), rep(FALSE, pop-circle))
-    idx = sample(pop)
-    read_table(path, col_names=F, col_types=cols(.default="d")) %>%
-        select(-X1) %>%
-        as.matrix() %>%
-        apply(1, function(x) st_point(x/6.28 + offset), simplify=FALSE) %>%
-        st_sfc(crs=3857) %>%
-        st_as_sf() %>%
-        rename(geometry=x) %>%
-        mutate(row = row, col = col,
-               dem = dem_vec[idx],
-               circle = circ_vec[idx])
-}
-
-minissouri_people = minissouri %>%
-    pmap_dfr(make_people)
-
-# functions and helpers -------
-plot_state = function(map, people, plan=NULL, qty=NULL, ppl_size=1.75, ...) {
-    if (is.null(people$harm)) people$harm = FALSE
[…]

### 2022-07-02 R/toy/01_harm_mininois.R

@@ -1,54 +1,0 @@
-# Mininois
-pl_fair = c(2, 2, 3, 3, 3, 3,
-            2, 2, 3, 3, 3, 3,
-            2, 2, 3, 3, 3, 3,
-            2, 2, 2, 3, 3, 3,
-            2, 1, 1, 3, 3, 3,
-            1, 1, 1, 3, 3, 3)
-pl_gerry = c(2, 2, 2, 3, 3, 1,
-             2, 2, 2, 3, 3, 1,
-             2, 2, 2, 3, 3, 1,
-             2, 2, 3, 3, 3, 1,
-             2, 3, 3, 3, 1, 1,
-             1, 1, 1, 1, 1, 1)
-
-dem_fair = with(mininois, tapply(dem, pl_fair, sum)/tapply(pop, pl_fair, sum))[pl_fair]
-dem_gerry = with(mininois, tapply(dem, pl_gerry, sum)/tapply(pop, pl_gerry, sum))[pl_gerry]
-
-coords_fair = list(x=c(1/4, 1/6, 2/3),
-                   y=c(1.04, -0.04, -0.04))
-coords_gerry = list(x=c(1/2, 2/3, 1/4),
-                    y=c(1.04, -0.04, -0.04))
-
-suppressMessages({
-# Dem gerry plot
-p1 = plot_state(mininois, mininois_people,
-           pl_gerry, dem_gerry) +
-    labs(title="(a) Democratic gerrymander")
-for (i in 1:3) {
-    p1 = add_lab(p1, fmt_lean(dem_gerry[as.character(i)]),
-                coords_gerry$x[i], coords_gerry$y[i])
-}
-p1 = p1 + guides(fill="none")
-
-# Fair plan plot
-p2 = plot_state(mininois, mininois_people,
-                pl_fair, dem_fair) +
-    labs(title="(b) Proportional plan")
-for (i in 1:3) {
-    p2 = add_lab(p2, fmt_lean(dem_fair[as.character(i)]),
-                 coords_fair$x[i], coords_fair$y[i])
-}
-
-# Harm plot
-p3 = plot_state(mininois, mininois_people,
-           pl_gerry, mininois$rep*round(1 - dem_fair)) +
-    scale_fill_gradient("Total\nharm", low="#e5e6e5", high="#666666") +
-    labs(title="(c) Voter harm, (a) vs. (b)") +
-    theme(legend.key.width=unit(0.2, "cm"))
[…]

### 2022-07-04 paper/harm.Rmd
big revisions

--- 2022-07-02 paper/paper.Rmd
@@
 
-# Minivation {#sec:motivation}
+# Motivation {#sec:motivation}
 
@@
 Voters' preferences are assumed to be stable across elections, and turnout is 100%.
-A schematic for this example is shown in Figure \ref{fig:mini-schematic}.
+A schematic for this example is shown in Figure \ref{fig:minissouri}.
 
@@
 This mirrors the situation in the actual state of Massachusetts, where a high level of partisan integration and a solid statewide Democratic majority leads to Republican inability to win any congressional seats under recent historical voting patterns [@duch:18].
-
-```{r mini-schematic}
-#| fig.cap: |
-#|      Schematic maps of the example states,
-#|      showing the partisan composition of each of the 36 precincts.
-#|      The total voting population for each example is 180.
-include_graphics("figures/minis_schematic.pdf")
-```
 
@@
 
-Consider the two plans in Figure \ref{fig:harm-mininois}(a)-(b).
+Consider the two plans in Figure XX(a)-(b).
 A Democratic gerrymander (a) effectuates Democratic control by splitting its city, pinwheel-like, into three districts reaching out into the more Republican and sparsely populated remainder of the state.
 This gerrymander would elect zero Republicans and three Democrats, whereas the proportional map (b) would elect two Democrats and one Republican.
-
-```{r harm-mininois}
-#| fig.cap: |
-#|      Two potential districting plans for Mininois:
-#|      a Democratic gerrymander (a) and a close-to-proportional plan (b).
-#|      Panel (c) shows the total harm to voters in each precinct done by (a), using (b) as a counterfactual.
-include_graphics("figures/harm_mininois.pdf")
-```
 
@@
 Republicans in the city and in the southwest are unharmed, as are all Democratic voters, since they are represented by the same party under both plans.
-Figure \ref{fig:harm-mininois}, Panel C plots the number of voters harmed in each precinct.
+Figure XX, Panel C plots the number of voters harmed in each precinct.
 
@@
 If a plan disproportionately harms voters of one group, then we might conclude that that group was the target of a gerrymander.
-Figure \ref{fig:diffharm-minissouri} illustrates this idea for two hypothetical gerrymanders ((a) and (c)) of Minissouri.
+Figure \ref{fig:minissouri} illustrates this idea for two hypothetical gerrymanders ((a) and (c)) of Minissouri.
 As in Mininois, we can tally the total number of voters harmed in each precinct ((d) and (f)) in comparison to a proportional counterfactual plan (b).
@@
 
-```{r diffharm-minissouri}
+```{r minissouri, fig.width=6.75}
 #| fig.cap: |
-#|      Three potential districting plans for Minissouri:
-#|      a Democratic gerrymander (a),
-#|      a close-to-proportional plan (b), and a Republican gerrymander (c).
-#|      Panels (d) and (f) show the total harm to voters in each precinct done
-#|      by their corresponding gerrymander, using (b) as a counterfactual.
-#|      Panel (e) summarizes this harm by party and across precincts, showing
-#|      the average harm done to voters of each party by each gerrymander.
-include_graphics("figures/diffharm_minissouri.pdf")
+#|      Two potential districting plans for Minissouri,
+#|      a Democratic gerrymander (a), and a close-to-proportional plan (b).
+#|      Panel (c) shows which voters are harmed by the Democratic gerrymander,
+#|      using the proportional plan (b) as a counterfactual.
+#|      Panel (d) summarizes this across precincts, showing
+#|      the average harm done to voters of each partisan and racial group.
+include_graphics("figures/minissouri.pdf")
 ```
@@
 The differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander is $0.450$, where positive values indicate Republican bias.
-In other words, under the Democratic gerrymander, the average Republican voter in Mininois is 41.5% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan in Figure \ref{fig:diffharm-minissouri}(b).
+In other words, under the Democratic gerrymander, the average Republican voter in Mininois is 41.5% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan in Figure \ref{fig:minissouri}(b).
 
@@
 To illustrate the connections in the preceding section and to demonstrate the value of the proposed measures in a real-world setting, we provide a brief analysis of congressional districting in the state of New Jersey.
-As discussed above, many gerrymandering measures often function well in highly competitive states, but may produce more counterintuitive results in other states.
-New Jersey, with an average Democratic two-party vote share of 56.6% across our two baseline elections (2016 presidential and 2018 senatorial), will demonstrate that this is not a problem unique to lopsided states.
+Since 1995, New Jersey congressional redistricting has been handled by a 13-member commisison, with 6 political appointees from each party and one independent chair with *de facto* tiebreaking authority.
+For the 2020 redistricting cycle, each party submitted a proposed plan to the chair, John Wallace Jr.
+Wallace ultimately selected the Democratic plan for enactment, "simply because in the last redistricting, the map was drawn by Republicans."[^8] Here, in contrast, we evaluate the two plans on the basis of existing gerrymandering metrics as well as through the proposed harm framework.
 
-```{r nj-maps, fig.width=6.5}
+[^8]: See Matt Friedman, "Democrats prevail in New Jersey redistricting with map that could sacrifice Malinowski," *Politico,* December 22, 2021.
+    <https://www.politico.com/news/2021/12/22/new-jersey-redistricting-map-malinowski-525983> (accessed July 3, 2022).
+
+```{r nj-maps, fig.width=6.25}
 #| fig.cap: |
@@
 
-Figure \ref{fig:nj-maps} shows the baseline Democratic vote share by precinct, along with two hypothetical gerrymanders of the state, which were generated using the short-burst optimization approach of @cannon2020.
-The Democratic gerrymander has 11 Democratic seats and 1 Republican seat at baseline, while the Republican gerrymander has 6 Democratic and 6 Republican seats.
+Figure \ref{fig:nj-maps} shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^9]
+Using an average of the 2016 presidential and 2018 senatorial elections as a baseline, the Democratic proposal yields 9 Democratic seats and 3 Republican seats, while the Republican proposal has 7 Democratic and 5 Republican seats.
 
-We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, setting constraints to favor compact plans with district boundaries that tend to follow county lines.
+[^9]: The enacted plan is sourced from [All About Redistricting](https://redistricting.lls.edu/state/new-jersey/?cycle=2020&level=Congress&startdate=2021-12-22), and the GOP plan was provided by Harrison Neely, executive director for the GOP delegation to the New Jersey Congressional Redistricting Committe, following an Open Public Records Act request.
+
+We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, as implemented in the R package `redist` [@redist], setting constraints to favor compact plans with district boundaries that tend to follow county lines.
 One such plan is shown in Figure \ref{fig:nj-maps}(d).
-For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^8] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
+For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^10] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
 The distribution of these measures is shown in Figure \ref{fig:nj-meas}.
 
-[^8]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
+[^10]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
+
+We first examine the values of each metric on the two proposals alone.
+The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-0.036$ for the Democratic proposal and $0.020$ for the Republican proposal.
+These values are relatively directly interpretable; for example, the differential harm score of $-0.036$ means that an average Democratic voter is 3.6% less likely to be misrepresented than their average Republican counter-partisan.
+The efficiency gap, declination, and partisan bias measures agree with the differential harm estimates in this regard, though the former two are much more discrete.
+The efficiency gap and partisan bias estimate give roughly equal scores in absolute value to the Democratic and Republican plans, in comparison to differential harm and declination, which identify the Democratic proposal as slightly more extreme.
+While the mean-median measure identifies the Republican proposal as favoring Republicans, it also scores the Democratic proposal as favoring the Republicans, albeit by less.
 
@@
 #| fig.cap: |
-#|      The distribution of the existing and proposed measures across a sample of plans,
+#|      The distribution of the existing and proposed measures across 
+#|      a sample of 10,000 plans,
 #|      with the values for two gerrymanders overlaid as vertical lines.
@@
 
-We first examine the values of each metric on the two example plans alone.
-The estimate of differential partisan harm correctly identifies the partisan skew of each gerrymander, with a large negative value for the Democratic gerrymander and a small positive value for the Republican gerrymander.
-These values are relatively directly interpretable; for example, the differential harm score of $-0.109$ means that an average Democratic voter is 10.9% less likely to be misrepresented than their average Republican counter-partisan.
-The efficiency gap and declination measures also correctly identify the gerrymanders, though they are much more discrete, and the efficiency gap gives roughly equal scores in absolute value to the Democratic and Republican plans, in comparison to differential harm and declination, which identify the Democratic gerrymander as more extreme.
-In contrast, the usual partisan bias estimator classifies the Republican gerrymander as favoring Democrats and vice versa.
-While the mean-median measure identifies the Democratic gerrymander, it classifies the Republican gerrymander as a an even more extreme Democratic-leaning outlier.
+These values can be better contextualized in comparison the distribution across the sampled plans.
+The Democratic proposal is expected to produce an average of 8.55 Democratic seats under our election model, more than 79% of all sample plans.
+The Republican proposal is an outlier, yielding fewer seats on average than every single sampled plan.
+This pattern is repeated for all of the four common gerrymandering measures, where the Republican plan is a clear outlier (all $p=0.0002$, except $p=0.002$ for partisan bias) while the Democratic plan is not ( $p=0.34$ for the efficiency gap, $0.50$ for declination, $0.88$ for partisan bias, and $0.17$ for the mean-median score).
+However, both the partisan bias estimate and the mean-median score again misidentify the direction of the bias of the Republican plan, classifying it as an extreme Democratic outlier.
 
-Considering the distribution of these statistics across the sampled plans doesn't help much, either.
-The expected number of Democratic seats naturally identifies the gerrymanders (two-sided $p=0.0002$) and the direction of bias.
-The gerrymanders are also an outlier under the efficiency gap and declination measures also (both $p=0.0002$), and in the correct direction.
-While the Republican gerrymander is an outlier for the partisan bias and mean-median score ($p=0.0224$ and $p=0.0004$), the Democratic gerrymander is only an outlier for the mean-median score($p=0.0004$ compared to $p=0.349$ for partisan bias).
-All of the plans identified as outliers under these two metrics are identified in the wrong direction.
+The estimate of differential harm already uses the simulation output, and so comparing the proposals to the sampling distribution is not strictly required.
+However, it is notable that while the Democratic proposal has a more extreme estimate of differential partisan harm (1.6 percentage points higher), its value of $-0.037$ is not an outlier with respect to the distribution of differential harm for a randomly sampled plan ($p=0.43$, using the same set of samples as a counterfactual), while the Republican plan is ($p=0.0002$).
+In other words, a *typical* nonpartisan districting plan is not necessarily a *fair* one, even in comparison to other similarly generated districting plans.
+The sampled plan which minimizes differential harm yields 8.18 expected Democratic seats, compared to a sample average of 8.41 seats, and the 6.8 seats that would be required for an outcome completely proportional to the parties' statewide vote share.
 
-There is also disagreement between the measures about the natural bias of districting plans in New Jersey: the median sampled plan has a partisan bias of $0.020$, indicating a natural Republican bias, but the median mean-median score is just $0.0040$, and the median differential partisan harm is $-0.032$, indicating a natural Democratic bias.
-Which is correct?
-It depends on what one means by "natural bias," though we submit that differential harm is one good way to measure it.
-Democrats do win 8.4 seats on average across the simulations, which is more than the 6.8 seats that would be expected under a completely proportional outcome.
-It is important to note that a value of zero estimated differential harm does not necessarily correspond to proportionality; the simulated plan with an estimated differential harm closest to zero yieled 8.1 expected Democratic seats.
-
-Both gerrymanders excessively harm the average voter, with the Democratic plan doing so at a moderately higher rate ($p=0.0004$ for the Democratic plan, $p=0.0008$ for the Republican plan).
+Both proposals harm the average voter at a higher rate than most of the sampled plans, with the Republican plan again more extreme in this regard ($p=0.0008$ versus $p=0.061$ for the Democratic plan).
 Notably, the plans which minimize total harm are also those which have near-zero differential partisan harm.
 Thus, at least in New Jersey, overall harm reduction is consistent with a fair map on a partisan basis.
-Minimizing bias with other measures, however, could lead one astray: plans which score an efficiency gap of zero can have a large average harm, which may not be distributed fairly between parties.
-Plans with zero partisan bias or mean-median score actually have higher-than-average harm and tend to differentially harm Republicans.
+Minimizing bias with other measures, however, could lead one astray: plans which score an efficiency gap of zero can have average harm as high as 0.172 and differential harm as extreme as $-0.061$, nearly twice the value of the Democratic proposal.
+Plans with zero partisan bias or mean-median score likewise have higher-than-average harm and tend to differentially harm Republicans.
 
@@
 Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats.
+Thus there exist many plans which give Democrats a higher-than-average share of the seats, but which have partisan bias and mean-median score favoring Republicans.
 New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
@@
 
-We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.[^9]
+We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.[^11]
 
-[^9]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
+[^11]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
 
@@
 Nine of the councilors are elected from contiguous city council districts, and the remaining four councilors are elected at-large, with all councilors serving two-year terms.
-Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^10]
+Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^12]
 This section will study the extent to which the current district boundaries differentially harm voters of different racial groups, in comparison to a completely race-neutral districting process.
@@
 
-[^10]: City of Boston charter.
+[^12]: City of Boston charter.
     Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
@@
 
-To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^11] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
+To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^13] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
 Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail.
 
-[^11]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+[^13]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
 
@@
 
-We sample 2,000 redistricting plans across two independent runs of the SMC algorithm of @mccartan2020, setting constraints to favor compact plans that respect neighborhood boundaries.
+We sample 2,000 districting plans across two independent runs of the SMC algorithm of @mccartan2020, setting constraints to favor compact plans that respect neighborhood boundaries.
 Since we do not provide any racial information to the algorithm, this set of plans forms a good approximation of a race-neutral districting process.
@@
 $$ As noted in the main text, $\bar{s}_d$ is computed by averaging historical election returns.
-This is done at the precinct level, since this is the most granular level that districting plans are usually built from and election returns are reported at.[^12]
+This is done at the precinct level, since this is the most granular level that districting plans are usually built from and election returns are reported at.[^14]
 
-[^12]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+[^14]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
     Population information comes from Census blocks, which, when populated, are practically never split.

### 2022-07-04 R/03_sim_nj.R

@@ -1,40 +1,19 @@
 nj = read_rds(here("data/NJ_cd_final_vtd_20.rds")) %>%
     redist_map(pop_tol=0.01, ndists=12, adj=.$adj)
 
-if (!'dem_comm' %in% names(nj)) {
-    nj_enacted <- read_sf('https://redistricting.lls.edu/wp-content/uploads/nj_2020_congress_2021-12-22_2031-06-30.json')
-    nj_rep_comm <- read_sf('data-raw/NJ CD 2021 GOP submission/NJ CD 2021 GOP V5_Hospital_shoreline.shp')
-    nj$dem_comm <- as.integer(nj_enacted$District[geomander::geo_match(from = nj, to = nj_enacted, method = 'area')])
-    nj$rep_comm <- as.integer(nj_rep_comm$DISTRICT[geomander::geo_match(from = nj, to = nj_rep_comm, method = 'area')])
-}
-
 N_sim = 10e3
 n_runs = 4
 if (!file.exists(sim_path <- here("data/nj_sims.rds"))) {
     plans = redist_smc(nj, N_sim/n_runs, counties=county, pop_temper=0.005,
-                       runs=n_runs, verbose=TRUE)
+                       runs=n_runs, verbose=TRUE) %>%
+        add_reference(nj$dem_comm) %>%
+        add_reference(nj$rep_comm)
 
-    # gerrymanders
-    set.seed(5118)
-    opt_dem = redist_shortburst(nj, scorer_group_pct(nj, ndv, ndv+nrv, 11),
-                                max_bursts=1000, burst_size=15, return_all=FALSE)
-    set.seed(5118)
-    opt_rep = redist_shortburst(nj, scorer_group_pct(nj, nrv, ndv+nrv, 6),
-                                max_bursts=1000, burst_size=15, return_all=FALSE)
-
-    plans = plans %>%
-        subset_sampled() %>%
-        add_reference(last_plan(opt_dem), "dem_gerry") %>%
-        add_reference(last_plan(opt_rep), "rep_gerry")
     write_rds(plans, sim_path, compress="xz")
 } else {
     plans = read_rds(sim_path)
 }
 
-plans <- plans %>%
-    subset_sampled() %>%
-    add_reference(ref_plan = nj$dem_comm, name = 'dem_comm') %>%
[…]

### 2022-07-04 src/harm.cpp

@@ -71,3 +71,48 @@ arma::mat harm_helper(const arma::imat pl, const arma::mat vs,
 
     return out;
 }
+
+
+// [[Rcpp::export(rng = false)]]
+arma::mat prec_harm_helper(const arma::imat pl, const arma::mat vs,
+                      const arma::vec wt_a, const arma::vec wt_b,
+                      const arma::vec shift_elec,
+                      const arma::mat shift_distr_1,
+                      const arma::mat shift_distr_2,
+                      const arma::ivec idx_1, const arma::ivec idx_2) {
+    int n1 = idx_1.n_elem;
+    int n2 = idx_2.n_elem;
+    int prec = pl.n_rows;
+
+    arma::mat out(prec, n1);
+
+    RcppThread::ProgressBar bar(n1, 1);
+    RcppThread::parallelFor(0, n1, [&] (int i) {
+        int i1 = idx_1(i) - 1;
+        for (int k = 0; k < prec; k++) {
+            int pl_ki = pl(k, i1) - 1;
+            double vs_base_ik = vs(pl_ki, i1);
+            double wt_a_k = wt_a(k) / n2;
+            double wt_b_k = wt_b(k) / n2;
+            out(k, i) = 0;
+            for (int j = 0; j < n2; j++) {
+                int j2 = idx_2(j) - 1;
+                int pl_kj = pl(k, j2) - 1;
+                double vs_ik = vs_base_ik + shift_elec(j) + shift_distr_1(pl_ki, j);
+                double vs_jk = vs(pl_kj, j2) + shift_elec(j) + shift_distr_2(pl_kj, j);
+                if (vs_jk > 0.5) {
+                    if (vs_ik <= 0.5) { // A win under counterfactual but not original
+                        out(k, i) += wt_a_k;
+                    }
+                } else if (vs_ik > 0.5) { // B win under counterfactual but not original
+                    out(k, i) += wt_b_k;
+                }
+            }
+        }
+
+        RcppThread::checkUserInterrupt();
+        bar++;
+    });
+
+    return out;
+}

### 2022-07-04 paper/harm.Rmd
update correlation table

@@
 Appendix \ref{app:corr} shows these correlations across simulated plans in eight states.
-The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which is usually around 0.82, and always above 0.66 in the eight states we study.
+The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
 

### 2022-07-15 paper/harm.Rmd
revise most of gary+kosuke

@@
 $$that is, a voter is harmed only if they are not represented by the candidate of their choice under $q$ and they would have been represented by their chosen candidate under the counterfactual plan $\tilde q$.
+Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$.
+Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion; in general, $q$ will harm some individuals versus $\tilde q$, and $\tilde q$ will harm a distinct set of individuals versus $q$.
+A "symmetrized" version of $h_i(v,q,\tilde q)$ would have to combine both of these groups individuals: those harmed as well as those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
+In practice, voters, courts, or politicians may care more about preventing harm than maximizing gains.
 
-This definition of harm is limited in two key regards.
+Our definition of harm is limited in two key regards.
 First, it considers only individuals who vote; if a person abstains from voting then their calculated harm is zero.
@@
 In general we will be interested in outcomes over multiple elections, and possibly a much larger set of counterfactual plans.
-Ideally, we would like to be able to average individual harm over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans, or even over a set of individuals with some shared characteristic.
+Ideally, we would like to be able to average individual harm over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans which are consistent with districting criteria, or even over a set of individuals with some shared characteristic.
 
-To allow for this, we replace the fixed election and counterfactual districting plans with random variables $V\in\Elecs$ and $\tilde Q\in\Plans$.
-The randomness here is primarily a conceptual device to simplify the notation, but one can also think of it as reflecting uncertainty or variation in voter preferences, the political environment, candidate quality, campaign events, and other factors.
-The randomness in districting plans specifically may be conceptualized as the constraints and preferences of the plan-drawing process (i.e., plans which better respect the preferences are given a higher probability).
-As we will discuss in Section \ref{sec:estimating}, generally we will assume that $V$ and $\tilde Q$ are independent, though this is not required for our theoretical analysis.
-Finally, we will also replace $i$ with an individual $I$ taken uniformly at random from $\Voters$, with $I\indep V,\tilde Q$.
+To allow for this, we replace the fixed election $v$, counterfactual districting plans $\tilde q$, and individual voter $i$ with random variables $V\in\Elecs$, $\tilde Q\in\Plans$, and $I\in\Voters$.
+The random variable $V$ can be conceptualized as ranging over possible future elections---one realization might correspond to a heavily Democratic-favoring midterm year, while another might involve incumbents underperforming particularly badly.
+The random variable $\tilde Q$ captures the range of potential counterfactual districting plans.
+Rarely would an analyst have a single counterfactual plan in mind; different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
+Finally, $I$ is defined to be a voter taken uniformly at random from $\Voters$, with $I\indep V,\tilde Q$.
 
-Then we may define the harm for a (randomly selected) individual under plan $q$ as $H(q) \dfeq h_I(V, q, \tilde Q)$.
-Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in election $v$ becomes$$
+The randomness in $V$ and $\tilde Q$ allows us to compare the enacted districting plan to many plausible counterfactual alternatives, and across a range of electoral environments.
+At times we may wish to study on a particular election or a particular counterfactual plan; treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
+A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section \ref{sec:estimating}.
+
+With these definitions, the harm for a (randomly selected) individual under plan $q$ can be written as $$
+H(q) \dfeq h_I(V, q, \tilde Q).
+$$$H(q)$ is itself a random variable, representing the harm to *some* individual in *some* election compared to *some* counterfactual plan.
+The random variable $I$ here serves as a kind of "veil of ignorance" [@rawls2004theory]; a districting plan $q$ which tends to produce a low $H(q)$ would be expected to be preferred over a plan which tended to produce a higher $H(q)$.
+
+$H(q)$ actually contains all the relevant information about the harm created by a plan $q$.
+We can "zoom in" or summarize various aspects of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
+Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in a particular election $v$ becomes$$
 \E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q)
@@
 Once we have defined a set of groups, we say that a plan $q$ is *fair* for groups $\Groups$ in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.
+Fairness is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
+For algorithmic fairness, the implicit counterfactual is a perfect predictor---compared to this perfect predictor, the actual decision process harms individuals whenever it makes a false positive or false negative error.
+Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.
 
@@
 
-While applicable to the study of redistricting beyond racial and partisan gerrymandering, the proposed harm measures provide an improvement over existing analytic approaches in both of these important contexts.
+A primary advantage of the proposed harm measures is that they are defined at the individual level, allowing for the study of redistricting far beyond racial and partisan gerrymandering.
+However, the measures provide an improvement over existing analytic approaches in both of these important contexts.
 
-## Partisan Gerrymandering
-
-The canonical framework for considering partisan fairness is partisan symmetry [@king1987democratic].
-Partisan symmetry implies that the seats $s$ won by a party with vote share $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
-In practice, the partisan symmetry framework is applied via the partisan bias metric $\beta(v)$, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020], i.e., the number of seats that would need to be reallocated to reach partisan symmetry for a given statewide vote share $v$ (usually $\half$).
-
-More recently, and particularly in the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
-The *mean-median* difference looks at the relationship between the vote share of the median seat in a delegation and the mean vote share of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
-The *declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
-Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015], which measures the difference in the number of the votes wasted by each party.[^4]
-
-[^4]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.
-
-The latter four metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in a larger theoretical framework or normative standard (in particular, they do not measure partisan symmetry [@katz2020]).
-As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
-This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^5]
-One advantage of the simpler metrics such as the efficiency gap is that they may quickly calculated using district vote shares alone.
-
-[^5]: For partisan bias, both researchers and practitioners have largely relied on simple uniform-shift models similar to the one we use in this work.
-
-All of these existing measures are constructed so that a value of zero indicates no bias, and a value of zero is at least implicitly held up as a normative standard for districting plans.
-This poses a challenge for the measures which do not estimate a theoretical quantity, because it is unclear what the implicit standard is.
-For partisan symmetry this is more explicit; @katz2020 write that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric, if there is a fairer alternative, would not normally be regarded as a legitimate outcome."
-
-But as those authors recognize, in evaluating a districting plan, it is critical to consider "if there is a fairer alternative." The inability of partisan bias and other existing measures to account for counterfactual districting plans means that none of them may be interpreted in a vacuum.
-No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the district-based electoral system.
-
-It is this confounding which explains many of the concerns that have been raised over the existing measures.
-The standard calculation of partisan bias, $\beta(\half)$, can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.[^6]
-And none of the measures take into account any aspect of political geography, and therefore suffer in states like Massachusetts, where, for example, the efficiency gap takes a value generally indicating an extreme Democratic gerrymander.
-
-[^6]: This has been called by some the "Utah paradox" [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
-    Even when interpreted carefully, however, partisan bias requires significant extrapolation from observed election outcomes in lopsided states---how realistic is it to evaluate the counterfactual where Democrats run even with Republicans statewide in Utah?
-
-Despite these differences, there are also key empirical similarities in many cases.
-The efficiency gap and differential partisan harm are often highly correlated (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^7]
-Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with differential partisan harm as well.
-Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section \ref{sec:nj}.
-
-[^7]: The differences stem from the efficiency gap considering a vote to be wasted if it is in excess of the 50% threshold needed to win, while the harm framework does not penalize such votes, since every such voter is happily represented by their candidate of choice.
-    This definition of wasted votes is also at the root of other criticisms of the efficiency gap, which note its tendency to penalize proportionality and reward 3-to-1 districts [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
-
-Finally, it is worth comparing the proposed measures to the increasingly common application redistricting simulation techniques to cases of partisan gerrymandering.
-Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics.
-While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
-In the worst cases, this can lead to directionally wrong conclusions, as we will see in Section \ref{sec:nj}.
-
-These problems have led @deford2020symmetry, among others, to recommend evaluating the presence of a gerrymander simply by comparing the expected number of seats won under the plan versus the simulations We agree with this recommendation, as it turns out that in the partisan case, differential harm is in practice tightly related to the overall number of seats won by each party, as Section \ref{sec:nj} and Appendix \ref{app:corr} demonstrate.
-However, unlike a direct measure of seats, differential harm is shifted so that a value of zero corresponds to fairness, and rescaled to be directly interpretable as a fraction of voters harmed.
-
-## Vote Dilution {#subsec:racial}
+## Racial Vote Dilution {#subsec:racial}
 
@@
 
+## Partisan Gerrymandering
+
+The canonical framework for considering partisan fairness is partisan symmetry [@king1987democratic].
+Partisan symmetry implies that the seats $s$ won by a party with vote share $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
+In practice, the partisan symmetry framework is applied via the partisan bias metric $\beta(v)$, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020], i.e., the number of seats that would need to be reallocated to reach partisan symmetry for a given statewide vote share $v$ (usually $\half$).
+
+More recently, and particularly in the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
+The *mean-median* difference looks at the relationship between the vote share of the median seat in a delegation and the mean vote share of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
+The *declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
+Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015], which measures the difference in the number of the votes wasted by each party.[^4]
+
+[^4]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.
+
+The latter four metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in a larger theoretical framework (in particular, they do not measure partisan symmetry [@katz2020]).
+As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
+This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^5]
+This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
+
+[^5]: For partisan bias, both researchers and practitioners have largely relied on simple uniform-shift models similar to the one we use in this work.
+
+Since none of the existing measures are centered on individual voters, they offer far less granular insight than the proposed harm framework, even in the case of studying partisan gerrymandering.
+For example, differential harm can be estimated at the precinct level and then plotted on a map to show not just *who* is harmed by a gerrymander, but *where* they reside.
+
+All of these existing measures are constructed so that a value of zero indicates no bias, and a value of zero is at least implicitly held up as a normative standard for districting plans.
+In doing so, these measures allude to *implicit* counterfactuals.
+For example, with partisan bias, the implicit electoral counterfactual is an election in which the parties' vote shares have been reversed, and the implicit counterfactual districting plan is one which assigns equal vote shares to the parties under both the real-world and counterfactual electoral scenarios.
+Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partissan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome."
+Similar implicit counterfactuals can be thought up for the other measures.
+
+These implicit counterfactuals differ from the explicit counterfactuals we adopt in this work in that there is no guarantee they are possible. 
+@katz2020 and others recognize that a nonzero value in a measure like partisan bias can only be considered problematic "if there is a fairer alternative."
+The inability of partisan bias and other existing measures to explicitly account for realistic counterfactual districting plans means that none of them may be interpreted in a vacuum.
+No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the district-based electoral system.
+
+It is this confounding which explains many of the concerns that have been raised over the existing measures.
+The standard calculation of partisan bias, $\beta(\half)$, can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.[^6]
+And none of the measures take into account any aspect of political geography, and therefore suffer in states like Massachusetts, where, for example, the efficiency gap takes a value generally indicating an extreme Democratic gerrymander.
+
+[^6]: This has been called by some the "Utah paradox" [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
+    Even when interpreted carefully, however, partisan bias requires significant extrapolation from observed election outcomes in lopsided states---how realistic is it to evaluate the counterfactual where Democrats run even with Republicans statewide in Utah?
+
+Despite these differences, there are also key empirical similarities in many cases.
+The efficiency gap and is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^7]
+Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
+Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section \ref{sec:nj}.
+
+[^7]: The differences stem from the efficiency gap considering a vote to be wasted if it is in excess of the 50% threshold needed to win, while the harm framework does not penalize such votes, since every such voter is happily represented by their candidate of choice.
+    This definition of wasted votes is also at the root of other criticisms of the efficiency gap, which note its tendency to penalize proportionality and reward 3-to-1 districts [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
+
+Finally, it is worth comparing the proposed measures to the increasingly common application redistricting simulation techniques to cases of partisan gerrymandering.
+Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics.
+While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
+In the worst cases, this can lead to directionally wrong conclusions, as we will see in Section \ref{sec:nj}.
+
 # Application to Partisan Gerrymandering in New Jersey {#sec:nj}
@@
 To illustrate the connections in the preceding section and to demonstrate the value of the proposed measures in a real-world setting, we provide a brief analysis of congressional districting in the state of New Jersey.
-Since 1995, New Jersey congressional redistricting has been handled by a 13-member commisison, with 6 political appointees from each party and one independent chair with *de facto* tiebreaking authority.
+Since 1995, New Jersey congressional redistricting has been handled by a 13-member commission, with 6 political appointees from each party and one independent chair with *de facto* tiebreaking authority.
 For the 2020 redistricting cycle, each party submitted a proposed plan to the chair, John Wallace Jr.
@@
 One such plan is shown in Figure \ref{fig:nj-maps}(d).
-For each of these sampled plans and the two gerrymanders, we calculate the average harm and differential partisan harm versus the simulated baseline,[^10] as well as the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
+For each of these sampled plans and the two gerrymanders, we estimate the average harm and differential partisan harm versus the simulated baseline,[^10] and calculate the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
 The distribution of these measures is shown in Figure \ref{fig:nj-meas}.
@@
 We first examine the values of each metric on the two proposals alone.
-The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-0.036$ for the Democratic proposal and $0.020$ for the Republican proposal.
-These values are relatively directly interpretable; for example, the differential harm score of $-0.036$ means that an average Democratic voter is 3.6% less likely to be misrepresented than their average Republican counter-partisan.
+The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-3.6%$ for the Democratic proposal and $2.0%$ for the Republican proposal.
+These values are relatively directly interpretable; for example, the estimated differential harm of $-3.6%$ means that an average Democratic voter is 3.6 percentage points less likely to be misrepresented than their average Republican counter-partisan.
 The efficiency gap, declination, and partisan bias measures agree with the differential harm estimates in this regard, though the former two are much more discrete.
@@
 The Republican proposal is an outlier, yielding fewer seats on average than every single sampled plan.
-This pattern is repeated for all of the four common gerrymandering measures, where the Republican plan is a clear outlier (all $p=0.0002$, except $p=0.002$ for partisan bias) while the Democratic plan is not ( $p=0.34$ for the efficiency gap, $0.50$ for declination, $0.88$ for partisan bias, and $0.17$ for the mean-median score).
+This pattern is repeated for all of the four common gerrymandering measures, where the Republican plan is a clear outlier (all $p=0.0002$, except $p=0.002$ for partisan bias) while the Democratic plan is not ($p=0.34$ for the efficiency gap, $0.50$ for declination, $0.88$ for partisan bias, and $0.17$ for the mean-median score).
 However, both the partisan bias estimate and the mean-median score again misidentify the direction of the bias of the Republican plan, classifying it as an extreme Democratic outlier.
@@
 In other words, a *typical* nonpartisan districting plan is not necessarily a *fair* one, even in comparison to other similarly generated districting plans.
-The sampled plan which minimizes differential harm yields 8.18 expected Democratic seats, compared to a sample average of 8.41 seats, and the 6.8 seats that would be required for an outcome completely proportional to the parties' statewide vote share.
+The sampled plan which minimizes estimated differential harm yields 8.18 expected Democratic seats, compared to a sample average of 8.41 seats, and the 6.8 seats that would be required for an outcome completely proportional to the parties' statewide vote share.
 
-Both proposals harm the average voter at a higher rate than most of the sampled plans, with the Republican plan again more extreme in this regard ($p=0.0008$ versus $p=0.061$ for the Democratic plan).
-Notably, the plans which minimize total harm are also those which have near-zero differential partisan harm.
+Both proposals are estimated to harm the average voter at a higher rate than most of the sampled plans, with the Republican plan again more extreme in this regard ($p=0.0008$ versus $p=0.061$ for the Democratic plan).
+Notably, the plans which minimize estimated total harm are also those which have near-zero estimated differential partisan harm.
 Thus, at least in New Jersey, overall harm reduction is consistent with a fair map on a partisan basis.
-Minimizing bias with other measures, however, could lead one astray: plans which score an efficiency gap of zero can have average harm as high as 0.172 and differential harm as extreme as $-0.061$, nearly twice the value of the Democratic proposal.
-Plans with zero partisan bias or mean-median score likewise have higher-than-average harm and tend to differentially harm Republicans.
+Minimizing bias with other measures, however, could lead one astray: plans which score an efficiency gap of zero can have estimated average harm as high as 0.172 and estimated differential harm as extreme as $-0.061$, nearly twice the value of the Democratic proposal.
+Plans with zero partisan bias or mean-median score likewise are estmated to have higher-than-average harm and to tend to differentially harm Republicans.
 
@@
 This intuition is confirmed by a closer analysis, summarized in Figure \ref{fig:al-maps}(f).
-Across the state, the estimated average harm is 0.079, which translates to around 159,000 voters.
+Across the state, the estimated average harm is 7.9%, which translates to around 159,000 voters.
 Of these harmed voters, around 123,000 are Black, while 36,000 are White.
@@
 Of course, the total number of each of these groups of voters is very different.
-Normalizing by group sizes, the expected harm for a Black Democrat is 0.22, for a Black Republican is 0.037, for a White Democrat is 0.083, and for a White Republican is 0.017.
-Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have a differential harm of 0.138 versus White Democrats and a differential harm of 0.183 versus Black Republicans.
+Normalizing by group sizes, the expected harm for a Black Democrat is 22%, for a Black Republican is 3.7%, for a White Democrat is 8.3%, and for a White Republican is 1.7%.
+Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have an estimated differential harm of 13.8% versus White Democrats and of 18.3% versus Black Republicans.
 
 These calculations are made by aggregating harm within racial groups at the precinct level, and so they automatically take into account individual voter preferences.
-Unlike in a traditional ecological/*Gingles* analysis, there is no need to identify a particular party as the statewide "choice" of the minority racial group, nor is there any cause for concern because of lower racial voting polarization in urban areas.
+Unlike in a traditional ecological/*Gingles* analysis, there is no need to identify a particular party as the uniform statewide "choice" of the minority racial group (though vote choice by race and geography must still be estimated), nor is there any cause for concern because of lower racial voting polarization in urban areas.
 The harm measure treats voters of every race and location equally, recording just their inability to elect their preferred candidate.
@@
 
-For the first scenario, where council races are contested along the ideological axis, we estimate that the enacted plan is quite fair, differentially harming White voters at a rate of only 0.0079 compared to the race-neutral baseline.
-For the second scenario, where the contests are oriented along the racial dimension, we estimate that the enacted plan differentially harms White voters at a rate of 0.041, compared to the race-neutral baseline.
+For the first scenario, where council races are contested along the ideological axis, we estimate that the enacted plan is quite fair, differentially harming White voters at a rate of only 0.79% compared to the race-neutral baseline.
+For the second scenario, where the contests are oriented along the racial dimension, we estimate that the enacted plan differentially harms White voters at a rate of 4.1%, compared to the race-neutral baseline.
 
@@
 This will produce a different answer, since harm is not symmetric between the study plan and the counterfactual.
-Along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 0.031 compared to the enacted plan, while along the second dimension, the differential harm would be 0.018.
+Comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline; the reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
+For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan
+Along the second dimension, the differential harm to Black voters would be 1.8%.
 
-Taken together, these findings suggest that despite the dominance of the Democratic party in city politics, and the shared preferences between certain groups of Black and White voters, race-aware districting still serves an important role in empowering Black voters across the city to elect their chosen candidates.
+Taken together, these findings suggest that despite the dominance of the Democratic party in city politics, and the shared preferences between certain groups of Black and White voters, race-aware districting still serves an important role in empowering Black voters across the city to elect their chosen candidates, and that a switch to race-neutral boundaries would have a material impact on Black voting power.
 Further study could focus on the racial impact of the four at-large members in comparison to a system with 13 districts and no at-large members.
@@
 The estimated standard deviations of the random effects are $\hat{\sigma}_\delta=0.0266$ for the election-year shocks and $\hat{\sigma}_\varepsilon=0.0653$ for the election-district-specific effects.
-We substitute these point estimates directly into our election model, which fails to account for uncertainty in the estimates.
+We substitute these point estimates directly into our election model.
+While this has the advantage of simplicity, it does not account for uncertainty in the estimates.
 

### 2022-08-01 paper/harm.Rmd
minivation

@@
 
-# Motivation {#sec:motivation}
+# Minivation {#sec:motivation}
 
@@
 A schematic for this example is shown in Figure \ref{fig:minissouri}.
-
-## Minichusetts: counterfactuals matter
-
-Every precinct in Minichusetts has a Democratic majority.
-As a result, it is impossible to draw a districting plan built from precincts that creates even a single Republican district.
-This mirrors the situation in the actual state of Massachusetts, where a high level of partisan integration and a solid statewide Democratic majority leads to Republican inability to win any congressional seats under recent historical voting patterns [@duch:18].
-
-This Democratic dominance poses a problem for existing measures of partisan gerrymandering.
-One of the most common measures, the efficiency gap of @stephanopoulos2015, takes a value of $-0.311$ or $-0.317$ for Minichusetts, depending on the configuration of the districts.
-This would make any Minichusetts districting plan an extreme Democratic gerrymander in comparison to the historical range of efficiency gaps observed in states nationwide.[^1]
-
-[^1]: See, e.g., the historical distribution calculated by the Campaign Legal Center's PlanScore tool: <https://planscore.campaignlegal.org/metrics/efficiencygap/>.
-
-But it is impossible to gerrymander if every manipulation of district boundaries produces no change in partisan outcomes.
-What the efficiency gap and other gerrymandering measures lack is an awareness of the space of possible districting plans, plans that *counterfactually* might have been selected.
-Once a set of districting rules has been agreed on (in Minichusetts, that districts must be contiguous and composed of whole precincts), what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
-There can be no gerrymander, and no harm to individual voters, if no improvement in partisan fairness is possible.
-
-## Mininois: individual harm
-
-To explore how to properly integrate counterfactual districting plans into the analysis of a potential gerrymander, we turn to the more interesting state of Mininois, where, in a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions [@chen:rodd:13; @enos:brown:21].
-
-Consider the two plans in Figure XX(a)-(b).
-A Democratic gerrymander (a) effectuates Democratic control by splitting its city, pinwheel-like, into three districts reaching out into the more Republican and sparsely populated remainder of the state.
-This gerrymander would elect zero Republicans and three Democrats, whereas the proportional map (b) would elect two Democrats and one Republican.
-
-Suppose now that plan (a) is enacted into law.
-Who does this decision harm, if anyone?
-Precisely those voters who are not represented by their chosen party, but would have been counterfactually.
-Using plan (b) as a counterfactual, we can see that rural Republicans in the eastern half of Mininois are harmed by plan (a), since under plan (b) they would be represented by a Republican.
-Republicans in the city and in the southwest are unharmed, as are all Democratic voters, since they are represented by the same party under both plans.
-Figure XX, Panel C plots the number of voters harmed in each precinct.
-
-Measuring the harm of the Democratic gerrymander required comparing it to a neutral counterfactual plan.
-The choice of counterfactual might seem hopelessly subjective.
-We discuss in Section \ref{sec:estimating} principled ways to define counterfactuals, but for now we hope the value in measuring harm through counterfactuals is clear even if the operationalization is not.
-
-## Minissouri: differential harm
-
-We could quantify the harm a Democratic gerrymander imposed on voters in Mininois by comparing election outcomes to those under a counterfactual proportional plan.
-But the results said nothing about the direction of the harm.
-While it is apparent that the Democratic gerrymander indeed benefits Democrats, tallying up the total harm done to all voters is not enough to prove this.
-
-However, we can partition voters into groups, and compare how the harm done by a plan falls across those groups.
-If a plan disproportionately harms voters of one group, then we might conclude that that group was the target of a gerrymander.
-Figure \ref{fig:minissouri} illustrates this idea for two hypothetical gerrymanders ((a) and (c)) of Minissouri.
-As in Mininois, we can tally the total number of voters harmed in each precinct ((d) and (f)) in comparison to a proportional counterfactual plan (b).
-In both cases, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.
+Each person is a square or circle, identifying some group membership or characteristic.
+Each person is then shaded red or blue to correspond to Republicans and Democrats, respectively.
 
@@
 
-Panel (e) then splits the voters by party, and computes, for each party, the share of its voters which are harmed.
-The Democratic gerrymander (a) harms Republicans at a much higher rate than Democrats, and the opposite can be said of the Republican gerrymander (c).
+Now, consider the two plans in Figure \ref{fig:minissouri}(a)-(b).
+Like in much of the US, in Minissouri Democrats are sorted asymmetrically into the cities while Republicans dominate the rural regions [@chen:rodd:13; @enos:brown:21].
+A Democratic gerrymander (a) effectuates Democratic control by splitting its cities, connecting them through thin regions around the boundary of the state.
+This gerrymander would elect one Republican and two Democrats, whereas the proportional map (b) would elect two Republicans and one Democrat.
+
+If we take the Democratic gerrymander as the reference plan and the proportional plan as the counterfactual, Figure \ref{fig:minissouri}(c) highlights who is harmed.
+Most of the highlighted people are the Republicans who were placed in a Democratic district in the gerrymander, but are now in a Republican district in the proportional plan.
+Note that the Republicans in the north-east city are not harmed here.
+In both plans, those Republicans are in a Democratic district in both the reference and the counterfactual.
+There are two blue people who are harmed, as they were placed into a Republican district in the gerrymander, but now move to a Democratic district.
+We can say then that 44 Republicans and 2 Democrats are individually harmed.
+
+There's a second dimension here to individual harm.
+Among the people, some are circles and some are squares.
+Squares are highly correlated with the Republican choice as circles are with the Democratic choice.
+Based on their vote choice, as before, we can identify who is harmed.
+Now, there are circles harmed from both parties.
+2 blue circles, 11 red circles, 33 red squares, and no blue squares are individually harmed.
+
+We can use these groups, to compare how the harm done by a plan falls across those groups.
+If a plan disproportionately harms voters of one group, then we might conclude that that group was the target of a gerrymander.
+
+The Democratic gerrymander (a) harms Republicans at a much higher rate than Democrats.
 We call the gap between these two rates the *differential harm,* and propose it as a general way to account for the impact of a districting plan on particular groups.
+If there is no gap, then we have *fairness*.
 
-The differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander is $0.450$, where positive values indicate Republican bias.
-In other words, under the Democratic gerrymander, the average Republican voter in Mininois is 41.5% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan in Figure \ref{fig:minissouri}(b).
+The partisan differential harm of the Democratic gerrymander of Minissouri is $-0.415$, where positive values indicate Republican bias.
+In other words, under the Democratic gerrymander, the average Republican voter in Mininois is 41.5% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan.
+The shape-based differential harm of the gerrymander is $-0.2$, where positive values indicate pro-square bias.
+On account of being a square and not a circle, a square is 20% less likely to be represented by their candidate of choice.
 

### 2022-08-01 R/02_run_toy.R

@@ -215,3 +215,8 @@ p1 + p2 + p3 + p4 + plot_layout(nrow=1, widths=c(0.26, 0.26, 0.26, 0.22)) &
     guides(fill="none") & theme(plot.margin=margin(0, 0, 0, 0))
 
 ggsave(here("paper/figures/minissouri.pdf"), width=7.25, height=2.25)
+
+p1 + p2 + p3 + p4 + plot_layout(nrow=2) &
+    guides(fill="none") & theme(plot.margin=margin(0, 0, 0, 0))
+ggsave(here("paper/figures/minissouri_2x2.pdf"), width=8, height=8)
+

### 2022-08-01 paper/figures/corr_table.tex
missing corr table

\begin{tabular}[t]{l>{}r>{}r>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{5}{c}{Correlation with differential harm and:} \\
\cmidrule(l{3pt}r{3pt}){2-6}
State & Exp. Dem. seats & Efficiency gap & Partisan bias & Mean-median & Declination\\
\midrule
Utah & \cellcolor[HTML]{8A671F}{-0.983} & \cellcolor[HTML]{75689E}{0.957} & \cellcolor[HTML]{927421}{-0.893} & \cellcolor[HTML]{8E6C20}{-0.945} & \cellcolor[HTML]{947722}{-0.876}\\
Mississippi & \cellcolor[HTML]{9D8627}{-0.772} & \cellcolor[HTML]{BFD9D5}{0.207} & \cellcolor[HTML]{E8EEE3}{0.031} & \cellcolor[HTML]{E7E5C0}{-0.096} & \cellcolor[HTML]{A89735}{-0.656}\\
Iowa & \cellcolor[HTML]{AD9F3D}{-0.597} & \cellcolor[HTML]{A0B5C2}{0.449} & \cellcolor[HTML]{DED99D}{-0.185} & \cellcolor[HTML]{C4DCD7}{0.184} & \cellcolor[HTML]{A9C2C9}{0.365}\\
North Carolina & \cellcolor[HTML]{8E6C20}{-0.943} & \cellcolor[HTML]{8A88AE}{0.740} & \cellcolor[HTML]{9AAABC}{0.520} & \cellcolor[HTML]{8C8DB0}{0.704} & \cellcolor[HTML]{9BAABC}{0.519}\\
Wisconsin & \cellcolor[HTML]{998024}{-0.816} & \cellcolor[HTML]{929BB4}{0.621} & \cellcolor[HTML]{C2DBD7}{0.191} & \cellcolor[HTML]{9096B2}{0.652} & \cellcolor[HTML]{A8C1C8}{0.374}\\
New Jersey & \cellcolor[HTML]{9B8425}{-0.789} & \cellcolor[HTML]{9FB4C1}{0.459} & \cellcolor[HTML]{DCE8DF}{0.082} & \cellcolor[HTML]{EEF0E0}{-0.014} & \cellcolor[HTML]{A7C0C7}{0.381}\\
Massachusetts & \cellcolor[HTML]{987E23}{-0.826} & \cellcolor[HTML]{E9E9CA}{-0.069} & \cellcolor[HTML]{E1DDA7}{-0.158} & \cellcolor[HTML]{CADFD9}{0.159} & \cellcolor[HTML]{808080}{NA}\\
California & \cellcolor[HTML]{AEA03E}{-0.590} & \cellcolor[HTML]{939BB4}{0.616} & \cellcolor[HTML]{DCD798}{-0.199} & \cellcolor[HTML]{D1C97B}{-0.302} & \cellcolor[HTML]{AAC3C9}{0.359}\\
\bottomrule
\end{tabular}

### 2022-08-01 paper/harm.Rmd
line break abstract for any future editing
visual fmt

@@
   Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
-But these measures cannot be easily adapted to other groups, including those defined by race, class, or geography.
+  But these measures cannot be easily adapted to other groups, 
+  including those defined by race, class, or geography.
   Nor are they applicable to single- or no-party contexts such as local redistricting.
-To overcome these limitations, we propose a unified framework of harm for evaluating the impacts of a districting plan on individual voters and the groups to which they belong.
-We consider a voter harmed if their chosen candidate is not elected under the current plan, but would be under a different plan.
-Harm improves on existing measures by both focusing on the choices of individual voters and directly incorporating counterfactual plans.
-We discuss strategies for estimating harm, and demonstrate the utility of our framework through analyses of partisan gerrymandering in New Jersey, voting rights litigation in Alabama, and racial dynamics of Boston City Council elections.
+  To overcome these limitations, we propose a unified framework of harm for evaluating 
+  the impacts of a districting plan on individual voters and the groups to which they belong. 
+  We consider a voter harmed if their chosen candidate is not elected under the current plan, 
+  but would be under a different plan.
+  Harm improves on existing measures by both focusing on the choices of individual voters 
+  and directly incorporating counterfactual plans.
+  We discuss strategies for estimating harm, and demonstrate the utility of our framework 
+  through analyses of partisan gerrymandering in New Jersey, voting rights 
+  litigation in Alabama, and racial dynamics of Boston City Council elections.
 keywords:
@@
 To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
-In general we will not require the group identities to be fixed with respect to elections.[^2]
+In general we will not require the group identities to be fixed with respect to elections.[^1]
 This is because some group identities, like partisan preference, may themselves change election-to-election.
 
-[^2]: Formally, we have random indicator variables $\ind\{i\in G_k\}$ for all $i$ and $k$, subject to the constraint that $\sum_k \ind\{i\in G_k\}=1$ for all $i$.
+[^1]: Formally, we have random indicator variables $\ind\{i\in G_k\}$ for all $i$ and $k$, subject to the constraint that $\sum_k \ind\{i\in G_k\}=1$ for all $i$.
     These indicator variables may be correlated with $V$ and/or $\tilde Q$, in general.
@@
 For each sample $\tilde q$ of a counterfactual districting plan, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
-We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^3]
+We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^2]
 We tally the total harm done to Democratic and Republican voters across the map and divide by the appropriate totals to get a single draw of $H(q)\mid I\in\dem$ and $H(q)\mid I\in\rep$.
@@
 
-[^3]: For this, we use the baseline count of Democratic or Republican votes.
+[^2]: For this, we use the baseline count of Democratic or Republican votes.
     This is a slight simplification, since because the vote share changes across random draws, the number of Democratic and Republican voters does, too.
@@
 The *declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
-Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015], which measures the difference in the number of the votes wasted by each party.[^4]
+Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015], which measures the difference in the number of the votes wasted by each party.[^3]
 
-[^4]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.
+[^3]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.
 
@@
 As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
-This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^5]
+This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^4]
 This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
 
-[^5]: For partisan bias, both researchers and practitioners have largely relied on simple uniform-shift models similar to the one we use in this work.
+[^4]: For partisan bias, both researchers and practitioners have largely relied on simple uniform-shift models similar to the one we use in this work.
 
@@
 For example, with partisan bias, the implicit electoral counterfactual is an election in which the parties' vote shares have been reversed, and the implicit counterfactual districting plan is one which assigns equal vote shares to the parties under both the real-world and counterfactual electoral scenarios.
-Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partissan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome."
-Similar implicit counterfactuals can be thought up for the other measures.
+Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partissan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome." Similar implicit counterfactuals can be thought up for the other measures.
 
 These implicit counterfactuals differ from the explicit counterfactuals we adopt in this work in that there is no guarantee they are possible.
-@katz2020 and others recognize that a nonzero value in a measure like partisan bias can only be considered problematic "if there is a fairer alternative."
-The inability of partisan bias and other existing measures to explicitly account for realistic counterfactual districting plans means that none of them may be interpreted in a vacuum.
+@katz2020 and others recognize that a nonzero value in a measure like partisan bias can only be considered problematic "if there is a fairer alternative." The inability of partisan bias and other existing measures to explicitly account for realistic counterfactual districting plans means that none of them may be interpreted in a vacuum.
 No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the district-based electoral system.
@@
 It is this confounding which explains many of the concerns that have been raised over the existing measures.
-The standard calculation of partisan bias, $\beta(\half)$, can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.[^6]
+The standard calculation of partisan bias, $\beta(\half)$, can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.[^5]
 And none of the measures take into account any aspect of political geography, and therefore suffer in states like Massachusetts, where, for example, the efficiency gap takes a value generally indicating an extreme Democratic gerrymander.
 
-[^6]: This has been called by some the "Utah paradox" [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
+[^5]: This has been called by some the "Utah paradox" [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
     Even when interpreted carefully, however, partisan bias requires significant extrapolation from observed election outcomes in lopsided states---how realistic is it to evaluate the counterfactual where Democrats run even with Republicans statewide in Utah?
@@
 Despite these differences, there are also key empirical similarities in many cases.
-The efficiency gap and is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^7]
+The efficiency gap and is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
 Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
@@
 
-[^7]: The differences stem from the efficiency gap considering a vote to be wasted if it is in excess of the 50% threshold needed to win, while the harm framework does not penalize such votes, since every such voter is happily represented by their candidate of choice.
+[^6]: The differences stem from the efficiency gap considering a vote to be wasted if it is in excess of the 50% threshold needed to win, while the harm framework does not penalize such votes, since every such voter is happily represented by their candidate of choice.
     This definition of wasted votes is also at the root of other criticisms of the efficiency gap, which note its tendency to penalize proportionality and reward 3-to-1 districts [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
@@
 For the 2020 redistricting cycle, each party submitted a proposed plan to the chair, John Wallace Jr.
-Wallace ultimately selected the Democratic plan for enactment, "simply because in the last redistricting, the map was drawn by Republicans."[^8] Here, in contrast, we evaluate the two plans on the basis of existing gerrymandering metrics as well as through the proposed harm framework.
+Wallace ultimately selected the Democratic plan for enactment, "simply because in the last redistricting, the map was drawn by Republicans."[^7] Here, in contrast, we evaluate the two plans on the basis of existing gerrymandering metrics as well as through the proposed harm framework.
 
-[^8]: See Matt Friedman, "Democrats prevail in New Jersey redistricting with map that could sacrifice Malinowski," *Politico,* December 22, 2021.
+[^7]: See Matt Friedman, "Democrats prevail in New Jersey redistricting with map that could sacrifice Malinowski," *Politico,* December 22, 2021.
     <https://www.politico.com/news/2021/12/22/new-jersey-redistricting-map-malinowski-525983> (accessed July 3, 2022).
@@
 
-Figure \ref{fig:nj-maps} shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^9]
+Figure \ref{fig:nj-maps} shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^8]
 Using an average of the 2016 presidential and 2018 senatorial elections as a baseline, the Democratic proposal yields 9 Democratic seats and 3 Republican seats, while the Republican proposal has 7 Democratic and 5 Republican seats.
 
-[^9]: The enacted plan is sourced from [All About Redistricting](https://redistricting.lls.edu/state/new-jersey/?cycle=2020&level=Congress&startdate=2021-12-22), and the GOP plan was provided by Harrison Neely, executive director for the GOP delegation to the New Jersey Congressional Redistricting Committe, following an Open Public Records Act request.
+[^8]: The enacted plan is sourced from [All About Redistricting](https://redistricting.lls.edu/state/new-jersey/?cycle=2020&level=Congress&startdate=2021-12-22), and the GOP plan was provided by Harrison Neely, executive director for the GOP delegation to the New Jersey Congressional Redistricting Committe, following an Open Public Records Act request.
 
@@
 One such plan is shown in Figure \ref{fig:nj-maps}(d).
-For each of these sampled plans and the two gerrymanders, we estimate the average harm and differential partisan harm versus the simulated baseline,[^10] and calculate the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
+For each of these sampled plans and the two gerrymanders, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
 The distribution of these measures is shown in Figure \ref{fig:nj-meas}.
 
-[^10]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
+[^9]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
 
@@
 
-We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.[^11]
+We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.[^10]
 
-[^11]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
+[^10]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
 
@@
 Nine of the councilors are elected from contiguous city council districts, and the remaining four councilors are elected at-large, with all councilors serving two-year terms.
-Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^12]
+Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^11]
 This section will study the extent to which the current district boundaries differentially harm voters of different racial groups, in comparison to a completely race-neutral districting process.
@@
 
-[^12]: City of Boston charter.
+[^11]: City of Boston charter.
     Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
@@
 
-To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^13] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
+To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^12] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
 Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail.
 
-[^13]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+[^12]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
 
@@
 Comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline; the reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
-For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan
-Along the second dimension, the differential harm to Black voters would be 1.8%.
+For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan Along the second dimension, the differential harm to Black voters would be 1.8%.
 
@@
 $$ As noted in the main text, $\bar{s}_d$ is computed by averaging historical election returns.
-This is done at the precinct level, since this is the most granular level that districting plans are usually built from and election returns are reported at.[^14]
+This is done at the precinct level, since this is the most granular level that districting plans are usually built from and election returns are reported at.[^13]
 
-[^14]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+[^13]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
     Population information comes from Census blocks, which, when populated, are practically never split.

### 2022-08-01 paper/harm.Rmd
motivation edits

@@
 
-# Minivation {#sec:motivation}
+# Motivation {#sec:motivation}
 
-To introduce and motivate the concepts of individual and differential harm, we will make use of Minissouri, a stylized representations of Missouri.
-Minissouri has two medium-sized cities which are not large enough to deliver a statewide Democratic majority.
-Minissouri has a population of 180 and should be divided into 3 districts with population between 54 and 66.
+To introduce and motivate the concepts of individual and differential harm, we will make use of Minissouri, a stylized representation of Missouri.
+The state's total population is 180, and it should be divided into three districts with population between 54 and 66.
+Minissouri has two medium-sized cities and a substantial rural population which an overall partisan breakdown of 55.6% R -- 44.4% D.
 Voters' preferences are assumed to be stable across elections, and turnout is 100%.
 A schematic for this example is shown in Figure \ref{fig:minissouri}.
-Each person is a square or circle, identifying some group membership or characteristic.
-Each person is then shaded red or blue to correspond to Republicans and Democrats, respectively.
+In addition to partisan affiliation indicated by red or blue coloring, each voter also belongs to a racial group indicated by squares and circles.
 
@@
 
-Now, consider the two plans in Figure \ref{fig:minissouri}(a)-(b).
-Like in much of the US, in Minissouri Democrats are sorted asymmetrically into the cities while Republicans dominate the rural regions [@chen:rodd:13; @enos:brown:21].
-A Democratic gerrymander (a) effectuates Democratic control by splitting its cities, connecting them through thin regions around the boundary of the state.
-This gerrymander would elect one Republican and two Democrats, whereas the proportional map (b) would elect two Republicans and one Democrat.
+Now suppose the redistricting plan in Figure \ref{fig:minissorui}(a) has been enacted.
+Despite the statewide popular vote favoring Republicans, under this plan Democrats would win two of the three seats.
+In contrast, a more proportional plan \ref{fig:minissouri}(b) would elect two Republicans and one Democrat.
 
-If we take the Democratic gerrymander as the reference plan and the proportional plan as the counterfactual, Figure \ref{fig:minissouri}(c) highlights who is harmed.
-Most of the highlighted people are the Republicans who were placed in a Democratic district in the gerrymander, but are now in a Republican district in the proportional plan.
-Note that the Republicans in the north-east city are not harmed here.
-In both plans, those Republicans are in a Democratic district in both the reference and the counterfactual.
-There are two blue people who are harmed, as they were placed into a Republican district in the gerrymander, but now move to a Democratic district.
-We can say then that 44 Republicans and 2 Democrats are individually harmed.
+Our intuition tells us that plan (a) unfairly packs Republicans into a rural district, depriving them of one seat.
+To turn this intuition into something more solid, we can consider which voters are harmed by the enactment of plan (a) over the alternative (b).
+These voters are highlighted in Figure \ref{fig:minissouri}(c).
+For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
+In contrast, Republican voters in the northeast city are not harmed, since they reside in a Democratic district in both the enacted and counterfactual plan.
 
-There's a second dimension here to individual harm.
-Among the people, some are circles and some are squares.
-Squares are highly correlated with the Republican choice as circles are with the Democratic choice.
-Based on their vote choice, as before, we can identify who is harmed.
-Now, there are circles harmed from both parties.
-2 blue circles, 11 red circles, 33 red squares, and no blue squares are individually harmed.
+After identifying which voters are harmed by plan (a) (with plan (b) as a counterfactual), it is natural to examine the overall composition of these harmed voters.
+Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
+These represent 44% of all Republicans and 2.5% of all Democrats.
+We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
+Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan.
 
-We can use these groups, to compare how the harm done by a plan falls across those groups.
-If a plan disproportionately harms voters of one group, then we might conclude that that group was the target of a gerrymander.
-
-The Democratic gerrymander (a) harms Republicans at a much higher rate than Democrats.
-We call the gap between these two rates the *differential harm,* and propose it as a general way to account for the impact of a districting plan on particular groups.
-If there is no gap, then we have *fairness*.
-
-The partisan differential harm of the Democratic gerrymander of Minissouri is $-0.415$, where positive values indicate Republican bias.
-In other words, under the Democratic gerrymander, the average Republican voter in Mininois is 41.5% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan.
-The shape-based differential harm of the gerrymander is $-0.2$, where positive values indicate pro-square bias.
-On account of being a square and not a circle, a square is 20% less likely to be represented by their candidate of choice.
+We need not limit our consideration to partisanship.
+We can also tally harm by race, which in Minissouri is correlated with both geography and partisanship, finding that 13 circle voters and 33 square voters are harmed.
+This corresponds to a differential racial harm of $-20\%$, where positive values indicate pro-square bias.
+Square voters are 20 percentage points more likely to be harmed than circle voters.
+While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or merely aimed to maximize Democratic representation, it does quantify the impact of plan (a) on these voters versus the alternative plan (b).
 
@@
 We first examine the values of each metric on the two proposals alone.
-The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-3.6%$ for the Democratic proposal and $2.0%$ for the Republican proposal.
-These values are relatively directly interpretable; for example, the estimated differential harm of $-3.6%$ means that an average Democratic voter is 3.6 percentage points less likely to be misrepresented than their average Republican counter-partisan.
+The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-3.6\%$ for the Democratic proposal and $2.0\%$ for the Republican proposal.
+These values are relatively directly interpretable; for example, the estimated differential harm of $-3.6\%$ means that an average Democratic voter is 3.6 percentage points less likely to be misrepresented than their average Republican counter-partisan.
 The efficiency gap, declination, and partisan bias measures agree with the differential harm estimates in this regard, though the former two are much more discrete.
@@
 The estimate of differential harm already uses the simulation output, and so comparing the proposals to the sampling distribution is not strictly required.
-However, it is notable that while the Democratic proposal has a more extreme estimate of differential partisan harm (1.6 percentage points higher), its value of $-0.037$ is not an outlier with respect to the distribution of differential harm for a randomly sampled plan ($p=0.43$, using the same set of samples as a counterfactual), while the Republican plan is ($p=0.0002$).
+However, it is notable that while the Democratic proposal has a more extreme estimate of differential partisan harm (1.6 percentage points higher), its value of $-3.7\%$ is not an outlier with respect to the distribution of differential harm for a randomly sampled plan ($p=0.43$, using the same set of samples as a counterfactual), while the Republican plan is ($p=0.0002$).
 In other words, a *typical* nonpartisan districting plan is not necessarily a *fair* one, even in comparison to other similarly generated districting plans.
@@
 Thus, at least in New Jersey, overall harm reduction is consistent with a fair map on a partisan basis.
-Minimizing bias with other measures, however, could lead one astray: plans which score an efficiency gap of zero can have estimated average harm as high as 0.172 and estimated differential harm as extreme as $-0.061$, nearly twice the value of the Democratic proposal.
+Minimizing bias with other measures, however, could lead one astray: plans which score an efficiency gap of zero can have estimated average harm as high as 17.2% and estimated differential harm as extreme as $-6.1\%$, nearly twice the value of the Democratic proposal.
 Plans with zero partisan bias or mean-median score likewise are estmated to have higher-than-average harm and to tend to differentially harm Republicans.

### 2022-08-01 R/00_setup_utils.R

@@ -232,5 +232,4 @@ pval = function(x, n_ref=redist:::get_n_ref(plans)) {
 Rcpp::sourceCpp("src/harm.cpp", cacheDir="src")
 
 if (!exists("harm")) source(here("R/01_metrics.R"))
-# if (!exists("mininois")) source(here("R/toy/00_toy_setup.R"))
 source(here("R/04_appl_data.R"))

### 2022-08-18 paper/harm.Rmd
explain EI models

@@
 
-Now suppose the redistricting plan in Figure \ref{fig:minissorui}(a) has been enacted.
+Now suppose the redistricting plan in Figure \ref{fig:minissouri}(a) has been enacted.
 Despite the statewide popular vote favoring Republicans, under this plan Democrats would win two of the three seats.
@@
 
-The latter four metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in a larger theoretical framework (in particular, they do not measure partisan symmetry [@katz2020]).
+The latter four metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in the larger theoretical framework of partisan symmetry [@katz2020].
 As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
@@
 For example, with partisan bias, the implicit electoral counterfactual is an election in which the parties' vote shares have been reversed, and the implicit counterfactual districting plan is one which assigns equal vote shares to the parties under both the real-world and counterfactual electoral scenarios.
-Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partissan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome." Similar implicit counterfactuals can be thought up for the other measures.
+Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partisan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome." Similar implicit counterfactuals can be thought up for the other measures.
 
@@
 We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.[^10]
+Measuring differential harm by both race and party will allow us to cleanly separate the racial and partisan effects of the Alabama districting plan, something existing approaches are unable to do.
 
@@
 As is custom in VRA analyses, we use an ecological inference model.
-The model, detailed in Appendix \ref{app:alei}, jointly estimates turnout and vote choice by race and precinct across seven statewide elections.
-Voting is highly polarized by race outside of the cities of Birmingham and Montgomery.
+The model detailed in Appendix \ref{app:alei}, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
+For example, we model precinct $j$'s observed turnout $n_{jv}$ in election $v$ and baseline turnout rates $\turn_{jr}$ by race $r$ as \begin{align*}
+    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
+        &\sim \Binlog\qty(
+        \overbrace{\vap_j,}^{\substack{\text{voting-age}\\\text{population}}}
+        \hspace{0.5em}
+        \logit(\turn_j^\top
+        \overbrace{\vb p_j}^{\mathclap{\substack{\text{precinct racial}\\\text{proportions}}}}) + 
+        \underbrace{\beta^{(t)}_v}_{\mathllap{\text{election-specific turnout shift}}}) \\
+    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
+        &\sim \mathcal{N}\underbrace{\qty(\mu^{(t)}, \Sigma^{(t)})}_{\mathrlap{
+        \text{where to cluster the precinct turnout rates}}},
+\end{align*} where $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale.
+We model Democratic support rates $\supp_{jr}$ by precinct and race similarly.
 
-With these estimates in hand, we can estimate the average number of voters harmed in each precinct and racial group.
+The model estimates, as expected, reveal that voting is highly polarized by race outside of the cities of Birmingham and Montgomery.
+With these estimates in hand, we can calculate the average number of voters harmed in each precinct and racial group.
 Figure \ref{fig:al-maps}(c) shows the estimated geographic distribution of harm.
@@
 To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^12] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
-Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in detail.
+The turnout component of the model is identical to that used for Alabama.
+The core of the voter choice model is the following specification for the $\pi_{jk}$, the preference of precinct $j$ for candidate $k$ on the logit scale:[^13] \begin{align*}
+    \pi_{jk} &= \underbrace{\beta^{(s)}_k}_{\substack{\text{citywide support}\\\text{for candidate}}} +
+    \pref_{j}^\top(\overbrace{P_j}^{\mathclap{\substack{\text{diagonal matrix of}\\\text{precinct racial proportions}}}}
+    \underbrace{\turn_j}_{\mathclap{\substack{\text{turnout rate parameter}\\\text{from turnout model}}}}
+    )^\top\vb L_k.
+\end{align*} Here, $\pref_j$ is a matrix of preferences by race and latent dimension within precinct $j$, so $\pref_{j}^\top(P_j\turn_j)^\top$ is a vector with elements containing the preferences of the precinct's voters in each latent dimension, after marginalizing out race.
+The matrix $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $\pref_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
+Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
 
 [^12]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+
+[^13]: For simplicity of presentation, the parameterization shown here has been altered slightly from that detailed in the appendix, though the ultimate model can be made identical after an adjustment to the prior on $\vb L_k$.
 
@@
 However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
-One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption that all candidates the voter did not vote for are interchangeable.
+One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption in the definition of harm that all candidates the voter did not vote for are interchangeable.
 
@@
 $$ As noted in the main text, $\bar{s}_d$ is computed by averaging historical election returns.
-This is done at the precinct level, since this is the most granular level that districting plans are usually built from and election returns are reported at.[^13]
+This is done at the precinct level, since this is the most granular level that districting plans are usually built from and election returns are reported at.[^14]
 
-[^13]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+[^14]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
     Population information comes from Census blocks, which, when populated, are practically never split.
@@
 
-For precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $\votes_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jv}$ denote the Democratic vote share, and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Other}\}$.
+For precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $n_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jv}$ denote the Democratic vote share, and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Other}\}$.
 Our main parameters of interest are $\turn_{jr}$ and $\supp_{jr}$, the baseline turnout and Democratic support rate by race in the precinct.
 
-We model the observed turnout $\votes_{jv}$ and vote counts $s_{jk}\votes_{jv}$ as \begin{align*}
-    \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
+We model the observed turnout $n_{jv}$ and vote counts $s_{jk}n_{jv}$ as \begin{align*}
+    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
         &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
@@
         &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
-    s_{jv}\votes_{jv} \mid \supp_{jr}, \turn_{jr}, \beta^{(s)}_v
-        &\sim \Binlog\qty(\votes_{jv}, \logit\qty(
+    s_{jv}n_{jv} \mid \supp_{jr}, \turn_{jr}, \beta^{(s)}_v
+        &\sim \Binlog\qty(n_{jv}, \logit\qty(
         \supp_j^\top \frac{\turn_j\circ\vb p_j}{\turn_j^\top\vb p_j}) + \beta^{(s)}_v) \\
@@
 
-Continuing the notation from above, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $\votes_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jk}$ denote the vote share for candidate $k$ (who implicitly belongs to a particular election $v_k$), and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Hispanic},\text{Other}\}$ ("Other" consisting primarily of Asian voters in Boston).
+Continuing the notation from above, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $n_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jk}$ denote the vote share for candidate $k$ (who implicitly belongs to a particular election $v_k$), and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Hispanic},\text{Other}\}$ ("Other" consisting primarily of Asian voters in Boston).
 Our main parameters of interest are $\turn_{jr}$, the baseline turnout by race and precinct, $\pref_{jqr}$, the preferences in latent dimension $q$ by race and precinct, and $L_{kq}$, the loading of each candidate on latent dimension $q$.
@@
 
-We model the observed turnout $\votes_{jv}$ and vote shares $s_{jk}$ as \begin{align*}
-    \votes_{jv} \mid \turn_{jr}, \beta^{(t)}_v
+We model the observed turnout $n_{jv}$ identically as in Alabama, above: \begin{align*}
+    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
         &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
     \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
-        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
-    \logit(s_{jk}) \mid \pi_{jk}, \eps &\sim \Norm(\pi_{jk}, \frac{\eps}{1+\votes_{jv_k}}) \\ 
+        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}).
+\end{align*}
+
+Because of the latent factor model, we do not require that vote shares by race and precinct, when tallied by race, necessarily equal the observed vote share in the precinct.
+As a result we choose to model the vote shares directly, rather than the vote counts, which is more computationally efficient.
+Because each precinct has dozens or hundreds of voters, a Normal approximation to the Binomial likelihood is appopriate, and so we model the vote shares $s_{jk}$ as \begin{align*}
+    \logit(s_{jk}) \mid \pi_{jk}, \eps &\sim \Norm(\pi_{jk}, \frac{\eps}{1+n_{jv_k}}),
+\end{align*} where $\pi_{jk}$ is the preference of precinct $j$ for candidate $k$ on the logit scale, as detailed below.
+The Normal approximation variance is inversely proportional to the total number of votes in the precinct (the additional 1 is to avoid divide-by-zero errors).
+The multiplicative factor $\eps$ captures the additional variance that isn't accounted for by the latent factors.
+
+The logit candidate preference itself is modeled as \begin{align*}
     \pi_{jk} &= \beta^{(s)}_k + \alpha_k
-        \pref_{j}^\top(\turn_j\circ \vb p_j)\mathrm{diag}(\vb*\rho)\vb L_k  \\
+        \pref_{j}^\top(\turn_j\circ \vb p_j)^\top\mathrm{diag}(\vb*\rho)\vb L_k  \\
     \pref_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
@@
 The matrix $\mathrm{diag}(\vb*\rho)$ captures the relative scale or importance of each dimension to the overall outcomes.
-
-Because of the latent factor model, we do not require that vote shares by race and precinct, when tallied by race, necessarily equal the observed vote share in the precinct.
-As a result we choose to model the vote shares directly, which is more computationally efficient.
-Since these vote shares come from Binomial observations, they have variance inversely proportional to the total number of votes in the precinct (the additional 1 is to avoid divide-by-zero errors).
-The multiplicative factor $\eps$ captures the additional variance that isn't accounted for by the latent factors.
 The $\vb*\alpha$ parameters are "fudge factors" that adjust the scale of the deviations from each candidate's average support.

### 2022-08-18 paper/harm.Rmd
add some interp of model output

@@
 The *mean-median* difference looks at the relationship between the vote share of the median seat in a delegation and the mean vote share of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
-The *declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
+*Declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
 Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015], which measures the difference in the number of the votes wasted by each party.[^3]
@@
 
-The latter four metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in the larger theoretical framework of partisan symmetry [@katz2020].
+The latter three metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in the larger theoretical framework of partisan symmetry [@katz2020].
 As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
@@
 This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
+Plugging in past election results without using an election model places assumes too much about the stability of electoral dynamics and consequently can significantly understate the uncertainty in how a particular districting plan will perform into the future.
 
@@
 While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
-In the worst cases, this can lead to directionally wrong conclusions, as we will see in Section \ref{sec:nj}.
+In the worst cases, this can lead to directionally wrong conclusions, as we will see in the next section.
 
@@
 
-The proposed harm measures are particularly easy to calculate and interpret in a partisan context, since the preferences of each kind of partisan are directly observable as partisan vote shares.
+The proposed harm measures are particularly easy to calculate and interpret in two-party elections, since the preferences of each kind of partisan are directly observable as partisan vote shares.
 However, the same framework can be fruitfully applied to more complex settings, providing insights that go beyond those available through traditional methods.
@@
 The turnout component of the model is identical to that used for Alabama.
-The core of the voter choice model is the following specification for the $\pi_{jk}$, the preference of precinct $j$ for candidate $k$ on the logit scale:[^13] \begin{align*}
+The core of the voter choice model is the following specification for $\pi_{jk}$, the preference of precinct $j$ for candidate $k$ on the logit scale:[^13] \begin{align*}
     \pi_{jk} &= \underbrace{\beta^{(s)}_k}_{\substack{\text{citywide support}\\\text{for candidate}}} +
@@
 \end{align*} Here, $\pref_j$ is a matrix of preferences by race and latent dimension within precinct $j$, so $\pref_{j}^\top(P_j\turn_j)^\top$ is a vector with elements containing the preferences of the precinct's voters in each latent dimension, after marginalizing out race.
-The matrix $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $\pref_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
+The vector $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $\pref_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
 Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
+As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
+Biden and Wu opponent Anissa Essaibi-George on the right.
+The second dimension appears primarily to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city score opposite values on the second dimension.
 
@@
 Its flexibility comes at a cost---most notably, the need to specify a set of counterfactual districting plans, the choice of which has a significant impact on conclusions.
-While modern simulation methods are a valuable starting point, these methods involve analyst discretion and interpretation of often-murky legal standards.
+Fundamentally, harm and differential harm are unable to measure plans against an absolute benchmark, the way that other measures such as partisan symmetry are able to do.
+And while modern simulation methods are a valuable starting point to defining counterfactuals, these methods involve analyst discretion and interpretation of often-murky legal standards.
 When subjective choices are not represented consistently, different analyses may arrive at different conclusions about which voters are harmed and at what rate.
@@
 ```{=tex}
-\begin{table}[h!]
+\begin{table}[ht!]
 \centering

### 2022-08-18 paper/harm.Rmd
further tweaks

@@
     \pi_{jk} &= \underbrace{\beta^{(s)}_k}_{\substack{\text{citywide support}\\\text{for candidate}}} +
-    \pref_{j}^\top(\overbrace{P_j}^{\mathclap{\substack{\text{diagonal matrix of}\\\text{precinct racial proportions}}}}
+    F_{j}^\top(\overbrace{P_j}^{\mathclap{\substack{\text{diagonal matrix of}\\\text{precinct racial proportions}}}}
     \underbrace{\turn_j}_{\mathclap{\substack{\text{turnout rate parameter}\\\text{from turnout model}}}}
     )^\top\vb L_k.
-\end{align*} Here, $\pref_j$ is a matrix of preferences by race and latent dimension within precinct $j$, so $\pref_{j}^\top(P_j\turn_j)^\top$ is a vector with elements containing the preferences of the precinct's voters in each latent dimension, after marginalizing out race.
-The vector $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $\pref_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
+\end{align*} Here, $F_j$ is a matrix of preferences by race and latent dimension within precinct $j$, so $F_{j}^\top(P_j\turn_j)^\top$ is a vector with elements containing the preferences of the precinct's voters in each latent dimension, after marginalizing out race.
+The vector $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $F_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
 Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
@@
 Biden and Wu opponent Anissa Essaibi-George on the right.
-The second dimension appears primarily to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city score opposite values on the second dimension.
+The second dimension appears primarily to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.
 
@@
 Continuing the notation from above, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $n_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jk}$ denote the vote share for candidate $k$ (who implicitly belongs to a particular election $v_k$), and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Hispanic},\text{Other}\}$ ("Other" consisting primarily of Asian voters in Boston).
-Our main parameters of interest are $\turn_{jr}$, the baseline turnout by race and precinct, $\pref_{jqr}$, the preferences in latent dimension $q$ by race and precinct, and $L_{kq}$, the loading of each candidate on latent dimension $q$.
+Our main parameters of interest are $\turn_{jr}$, the baseline turnout by race and precinct, $F_{jqr}$, the preferences in latent dimension $q$ by race and precinct, and $L_{kq}$, the loading of each candidate on latent dimension $q$.
 Here $q\in\{1,2\}$, since preliminary analyses found no improved fit when expanding to three dimensions.
@@
     \pi_{jk} &= \beta^{(s)}_k + \alpha_k
-        \pref_{j}^\top(\turn_j\circ \vb p_j)^\top\mathrm{diag}(\vb*\rho)\vb L_k  \\
-    \pref_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
-\end{align*} Note that $\pref_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $\pref_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
+        F_{j}^\top(\turn_j\circ \vb p_j)^\top\mathrm{diag}(\vb*\rho)\vb L_k  \\
+    F_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
+\end{align*} Note that $F_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $F_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
 The matrix $\mathrm{diag}(\vb*\rho)$ captures the relative scale or importance of each dimension to the overall outcomes.
@@
 
-The estimated loading matrix $\vb L$ is visualized on the left of Figure \ref{fig:boston-factors}, and $\pref$ is visualized for Black and White voters on the right.
+The estimated loading matrix $\vb L$ is visualized on the left of Figure \ref{fig:boston-factors}, and $F$ is visualized for Black and White voters on the right.
 

### 2022-08-27 paper/harm.Rmd
section 2 tenses

@@
   including those defined by race, class, or geography.
-  Nor are they applicable to single- or no-party contexts such as local redistricting.
+  Nor are they applicable to single- or no-party contexts, such as local redistricting.
   To overcome these limitations, we propose a unified framework of harm for evaluating 
@@
 
-To introduce and motivate the concepts of individual and differential harm, we will make use of Minissouri, a stylized representation of Missouri.
+To motivate individual and differential harm, we use *Minissouri*, a stylized representation of Missouri.
 The state's total population is 180, and it should be divided into three districts with population between 54 and 66.
@@
 
-Now suppose the redistricting plan in Figure \ref{fig:minissouri}(a) has been enacted.
-Despite the statewide popular vote favoring Republicans, under this plan Democrats would win two of the three seats.
-In contrast, a more proportional plan \ref{fig:minissouri}(b) would elect two Republicans and one Democrat.
+Now suppose the redistricting plan in Figure \ref{fig:minissouri}(a) is the enacted plan.
+Despite the statewide popular vote favoring Republicans, under this plan Democrats win two of the three seats.
+In contrast, a more proportional plan \ref{fig:minissouri}(b) elect two Republicans and one Democrat.
 
 Our intuition tells us that plan (a) unfairly packs Republicans into a rural district, depriving them of one seat.
-To turn this intuition into something more solid, we can consider which voters are harmed by the enactment of plan (a) over the alternative (b).
+To turn this intuition into something more concrete, we can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
 These voters are highlighted in Figure \ref{fig:minissouri}(c).
@@
 
-After identifying which voters are harmed by plan (a) (with plan (b) as a counterfactual), it is natural to examine the overall composition of these harmed voters.
+After identifying which voters are harmed by plan (a) compared to counterfactual plan (b), it is natural to examine the overall composition of these harmed voters.
 Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
@@
 We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
-Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan.
+Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points more likely to be represented by a Democrat than the average Democrat is to be represented by a Republican, compared to plan (b).
 
 We need not limit our consideration to partisanship.
-We can also tally harm by race, which in Minissouri is correlated with both geography and partisanship, finding that 13 circle voters and 33 square voters are harmed.
+We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship, finding that 13 circle voters and 33 square voters are harmed.
 This corresponds to a differential racial harm of $-20\%$, where positive values indicate pro-square bias.

### 2022-08-27 paper/harm.Rmd
Update harm.Rmd

@@
 We formalize and generalize the concepts in Section \ref{sec:motivation}, beginning with some notation.
-Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of districts, indexed by $\Distrs$.
+Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration.
+Individuals vote to select a candidate to represent each of a set of districts, indexed by $\Distrs$.
 
@@
 
-Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
+Every voter resides in some legislative district.,
+This assignment defines a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
 The set of all possible districting plans is $\Plans$.
 
-For an election $v$, we can tally the votes for each candidate; the candidate with the most votes wins.
+For an election $v$, we can tally the votes for each candidate and the candidate with the most votes wins.
 Using potential outcomes notation, we denote the winner of voter $i$'s district under districting plan $q$ and election $v$ as $\win_i(v,q)$.
@@
 
-The key intuition from Section \ref{sec:motivation} was that to be harmed by a decision a voter must have a better outcome in the counterfactual world where another decision was made.
+The key intuition from Section \ref{sec:motivation} is that *to be harmed*, a voter must have a better outcome in the counterfactual world where a different decision was made.
 To operationalize this, we need to have a counterfactual districting plan $\tilde q$.
@@
         \ind\{v(i) = \win_i(v, \tilde q)\},
-$$that is, a voter is harmed only if they are not represented by the candidate of their choice under $q$ and they would have been represented by their chosen candidate under the counterfactual plan $\tilde q$.
+$$that is, a voter is harmed only if they are not represented by the candidate of their choice under $q$ but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$.
 Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$.
-Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion; in general, $q$ will harm some individuals versus $\tilde q$, and $\tilde q$ will harm a distinct set of individuals versus $q$.
-A "symmetrized" version of $h_i(v,q,\tilde q)$ would have to combine both of these groups individuals: those harmed as well as those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
-In practice, voters, courts, or politicians may care more about preventing harm than maximizing gains.
+Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion.
+In general, $q$ will harm some individuals while $\tilde q$, and $\tilde q$ will harm a different set of individuals.
+A symmetric version of $h_i(v,q,\tilde q)$ would need to combine both of these groups: those harmed and those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
+In practice, voters, courts, or politicians may find different balances in preventing harm and maximizing gains.
 
 Our definition of harm is limited in two key regards.
-First, it considers only individuals who vote; if a person abstains from voting then their calculated harm is zero.
-A natural extension of this definition would be to separate an individual's candidate preference from their vote choice; the latter would be used in calculating election winners, but the former could provide a more broadly-based harm measure.
-Naturally, this approach requires auxiliary data to implement (such as a large-scale survey), since candidate preferences for nonvoters cannot be directly inferred from election returns.
+First, it considers only individuals who vote: people who abstain from voting have a harm of zero.
+A natural extension of this definition is to separate individual candidate preference from vote choice. 
+The latter defines election winners, but the former provides a more-broad harm measure.
+Naturally, this approach requires auxiliary data to implement, such as a large-scale survey, as candidate preferences for non-voters are not directly inferred from election returns.
 
-A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between candidates he or she did not vote for.
+A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between the candidates that he or she did not vote for.
 For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
 If the voter ultimately voted for candidate A, this harm measure makes no distinction between candidate B or candidate C winning the voter's district.
-This reflects the simplistic nature of the cost function that is implicit in the harm measure.
-The problem could be remedied by replacing the indicator functions with other cost functions that consider ordinal preferences.
+This reflects the simplistic nature of the cost function assumed in single-choice elections.
+The problem could be remedied by replacing the indicator functions in the harm measure with other cost functions that consider ordinal preferences.
 Ballot-level ranked-choice vote data, or candidate ideology scores, could be used to estimate such a cost function.
-Here, for simplicity and to avoid making further assumptions, we limit ourselves to the indicator-based cost function.
-Section \ref{sec:boston} further discusses the implications of this particular specification in the context of primary elections.
+Here, for simplicity and to avoid making further assumptions, we focus on the indicator-based cost function.
+Section \ref{sec:boston} further discusses the implications of this particular specification, in the context of primary elections.
 
@@
 
-As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to a single counterfactual districting plan $\tilde q$.
-In general we will be interested in outcomes over multiple elections, and possibly a much larger set of counterfactual plans.
-Ideally, we would like to be able to average individual harm over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans which are consistent with districting criteria, or even over a set of individuals with some shared characteristic.
+As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
+In general, we are interested in outcomes over multiple elections and a much larger set of counterfactual plans.
+Ideally, we would average individual harm (1) over a set of future elections under the enacted districting plan and (2) over a set of hypothetical districting plans which are consistent with districting criteria.
+We could even consider a set of individuals with some shared characteristic(s).
 
 To allow for this, we replace the fixed election $v$, counterfactual districting plans $\tilde q$, and individual voter $i$ with random variables $V\in\Elecs$, $\tilde Q\in\Plans$, and $I\in\Voters$.
-The random variable $V$ can be conceptualized as ranging over possible future elections---one realization might correspond to a heavily Democratic-favoring midterm year, while another might involve incumbents underperforming particularly badly.
+The random variable $V$ can be conceptualized as ranging over possible future elections---one realization could correspond to a heavily Democratic-favoring midterm year, while another could involve incumbents underperforming particularly badly.
 The random variable $\tilde Q$ captures the range of potential counterfactual districting plans.
-Rarely would an analyst have a single counterfactual plan in mind; different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
+Rarely would an analyst have a single counterfactual plan in mind.
+Different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
 Finally, $I$ is defined to be a voter taken uniformly at random from $\Voters$, with $I\indep V,\tilde Q$.
@@
 The randomness in $V$ and $\tilde Q$ allows us to compare the enacted districting plan to many plausible counterfactual alternatives, and across a range of electoral environments.
-At times we may wish to study on a particular election or a particular counterfactual plan; treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
+At times we may wish to study on a particular election or a particular counterfactual plan.
+If we treat $V$ and $\tilde Q$ as random variables, we can then condition on particular values of each.
 A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section \ref{sec:estimating}.
 
-With these definitions, the harm for a (randomly selected) individual under plan $q$ can be written as $$
-H(q) \dfeq h_I(V, q, \tilde Q).
-$$$H(q)$ is itself a random variable, representing the harm to *some* individual in *some* election compared to *some* counterfactual plan.
-The random variable $I$ here serves as a kind of "veil of ignorance" [@rawls2004theory]; a districting plan $q$ which tends to produce a low $H(q)$ would be expected to be preferred over a plan which tended to produce a higher $H(q)$.
+With these definitions, the harm for a (randomly selected) individual under plan $q$ can be written as 
+$$H(q) \dfeq h_I(V, q, \tilde Q).$$
 
-$H(q)$ actually contains all the relevant information about the harm created by a plan $q$.
-We can "zoom in" or summarize various aspects of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
-Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in a particular election $v$ becomes$$
-\E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q)
-    \text{ and }v(I)=\win_I(v, \tilde Q)],
-$$that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
+$H(q)$ is itself a random variable, representing the harm to *some* individual in *some* election compared to *some* counterfactual plan.
+The random variable $I$ here serves as a kind of "veil of ignorance" [@rawls2004theory]: a districting plan $q$ which tends to produce a low $H(q)$ would be expected to be preferred over a plan which tended to produce a higher $H(q)$.
+
+$H(q)$ describes the harm created by a plan $q$.
+We can zoom in to specific dimensions of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
+Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in a particular election $v$ becomes
+
+$$\E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q) \text{ and }v(I)=\win_I(v, \tilde Q)],$$
+    that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
 
@@
 
-Single-member districts are often motivated by the value in representation to specific communities with shared interests; indeed, preserving "communities of interest" is one of the most common statutory and constitutional requirements imposed by states on the plan-drawing process [@ncslcriteria].
+Single-member districts are often motivated by the value in representation to specific communities with shared interests.
+Indeed, preserving "communities of interest" is one of the most common statutory and constitutional requirements imposed by states on the plan-drawing process [@ncslcriteria].
 It is therefore natural to consider not just the average harm to voters done by a plan, but also how this harm is distributed across certain groups of voters, defined by race, party, or other attributes of interest.
@@
 
-Fairness is a strict condition; any real-life plan will never be exactly fair.
+Fairness is a strict condition: real-life plans are incredibly unlikely to be exactly fair.
 To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
@@
 If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
-In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined; there is certainly an element of subjectivity in sampling a set of districting plans.
+In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined.
+There is certainly an element of subjectivity in sampling a set of districting plans.
 Nonetheless, sampling methods are widely used in practice [see e.g., @2019rucho; @2018league; @2022merrill; @becker2021computational; @chen:rodd:13].
@@
 [^2]: For this, we use the baseline count of Democratic or Republican votes.
-    This is a slight simplification, since because the vote share changes across random draws, the number of Democratic and Republican voters does, too.
-    To deal with this issue completely necessitates modeling elections at the precinct level, which requires much more data and computational expenditure.
+    This is a slight simplification, since the vote share changes across random draws, the number of Democratic and Republican voters changes too.
+    To deal with this issue completely necessitates modeling elections at the precinct level, which requires more data and computational costs.
 
@@
 
-A primary advantage of the proposed harm measures is that they are defined at the individual level, allowing for the study of redistricting far beyond racial and partisan gerrymandering.
-However, the measures provide an improvement over existing analytic approaches in both of these important contexts.
+A primary advantage of our proposed harm measures is that they our measures are defined at the individual level.
+This allows for the study of redistricting far beyond racial and partisan gerrymandering.
+The measures provide an improvement over existing analytical approaches in both of these important contexts.
 
@@
 While the specific legal requirements for a vote dilution claim will obviously remain front-and-center for court challenges of racial gerrymanders, it is notable that the proposed harm measures unify all three *Gingles* criteria.
-Criteria (b) and (c) captures the misrepresentation of minority voters by a candidate they do not support, while (a) requires the existence of a counterfactual where this would not be the case; together, (a) and (c) establish that minority voters are harmed.
+Criteria (b) and (c) capture the misrepresentation of minority voters by a candidate they do not support, while (a) requires the existence of a counterfactual where this would not be the case and (a) and (c) together establish that minority voters are harmed.
 But the proposed measures can also tackle trickier situations where the *Gingles* criteria may not be clearly met.
@@
 The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-3.6\%$ for the Democratic proposal and $2.0\%$ for the Republican proposal.
-These values are relatively directly interpretable; for example, the estimated differential harm of $-3.6\%$ means that an average Democratic voter is 3.6 percentage points less likely to be misrepresented than their average Republican counter-partisan.
+These values are relatively directly interpretable.
+For example, the estimated differential harm of $-3.6\%$ means that an average Democratic voter is 3.6 percentage points less likely to be misrepresented than their average Republican counter-partisan.
 The efficiency gap, declination, and partisan bias measures agree with the differential harm estimates in this regard, though the former two are much more discrete.
@@
 This will produce a different answer, since harm is not symmetric between the study plan and the counterfactual.
-Comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline; the reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
+Comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline.
+The reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
 For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan Along the second dimension, the differential harm to Black voters would be 1.8%.
@@
 We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior.
-We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$; the former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).
-Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.
+We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$. 
+The former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).
+Several reparametrizations are made for computational efficiency.
+The full Stan model is included in the replication code.
 
@@
 We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.5^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 4.0^2)$ prior, and $\logit(\mu^{(t)})$ an i.i.d. $\Beta(10, 30)$ prior.
-We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$; the former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
+We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$.
+The former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
 A larger LKJ shape parameter places more weight on correlation matrices closer to diagonal, i.e., it regularizes away from high correlations.
@@
 Overall, these priors weakly identify the location and scale of the parameters.
-Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.
+Several reparametrizations were made for computational efficiency.
+The full Stan code for the model is available as part of the replication code.
 

### 2022-08-27 paper/harm.Rmd
editing

@@
   Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
-  But these measures cannot be easily adapted to other groups, 
-  including those defined by race, class, or geography.
+  But these measures are not easily adapted to other groups, 
+  including groups defined by race, class, or geography.
   Nor are they applicable to single- or no-party contexts, such as local redistricting.
@@
 
-We formalize and generalize the concepts in Section \ref{sec:motivation}, beginning with some notation.
-Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration.
-Individuals vote to select a candidate to represent each of a set of districts, indexed by $\Distrs$.
+We formalize and generalize the concepts in Section \ref{sec:motivation}.
+Let $\Voters$ be the set of individuals, or *voters*, in the region and time period.
+Individuals vote for a candidate to represent districts.
+Districts are indexed by $\Distrs$.
 
@@
 
-Every voter resides in some legislative district.,
-This assignment defines a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
+Every voter resides in a legislative district.,
+Each assignment defines a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
 The set of all possible districting plans is $\Plans$.
 
-For an election $v$, we can tally the votes for each candidate and the candidate with the most votes wins.
+For election $v$, we tally the votes for each candidate.
+The candidate with the most votes wins.
 Using potential outcomes notation, we denote the winner of voter $i$'s district under districting plan $q$ and election $v$ as $\win_i(v,q)$.
@@
 The key intuition from Section \ref{sec:motivation} is that *to be harmed*, a voter must have a better outcome in the counterfactual world where a different decision was made.
-To operationalize this, we need to have a counterfactual districting plan $\tilde q$.
-Then we may define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as $$
-h_i(v,q,\tilde q) = \ind\{v(i) \neq \win_i(v, q)\}
-        \ind\{v(i) = \win_i(v, \tilde q)\},
-$$that is, a voter is harmed only if they are not represented by the candidate of their choice under $q$ but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$.
+To operationalize this, we rely on a counterfactual districting plan $\tilde q$.
+Then we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as
+
+$$h_i(v,q,\tilde q) = \ind\{v(i) \neq \win_i(v, q)\} \ind\{v(i) = \win_i(v, \tilde q)\},$$
+
+that is, a voter is harmed only if they are not represented by the candidate of their choice under $q$ but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$.
 Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$.

### 2022-08-27 paper/harm.Rmd
Update harm.Rmd

@@
 
-$$h_i(v,q,\tilde q) = \ind\{v(i) \neq \win_i(v, q)\} \ind\{v(i) = \win_i(v, \tilde q)\},$$
+$$h_i(v,q,\tilde q) = \ind\{v(i) \neq \win_i(v, q)\} \ind\{v(i) = \win_i(v, \tilde q)\}.$$
 
-that is, a voter is harmed only if they are not represented by the candidate of their choice under $q$ but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$.
+A voter is harmed only if they are not represented by the candidate of their choice under $q$ but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$.
 Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$.

### 2022-08-27 paper/harm.Rmd
add clarity to the limitations ~ some are cute but useless

@@
 Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion.
-In general, $q$ will harm some individuals while $\tilde q$, and $\tilde q$ will harm a different set of individuals.
+In general, $q$ harms some individuals while $\tilde q$, and $\tilde q$ harms a distinct, though not necessarily disjoint, set of individuals.
 A symmetric version of $h_i(v,q,\tilde q)$ would need to combine both of these groups: those harmed and those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
@@
 
-Our definition of harm is limited in two key regards.
+Our definition of harm is limited in two notable regards.
 First, it considers only individuals who vote: people who abstain from voting have a harm of zero.
-A natural extension of this definition is to separate individual candidate preference from vote choice. 
-The latter defines election winners, but the former provides a more-broad harm measure.
-Naturally, this approach requires auxiliary data to implement, such as a large-scale survey, as candidate preferences for non-voters are not directly inferred from election returns.
+Empirically, this changes nothing, but limits the measure in a theoretical sense.
+People who do not voter express no formal preference, even if they have informal preferences.
+The natural theoretical extension of this definition is to then separate individual preferences from vote choice. 
+Naturally, extending this approach requires auxiliary data to implement, such as an absurdly large-scale survey, as candidate preferences for non-voters are not directly inferred from election returns.
+As surveys are typically benchmarked against electoral outcomes, we do not endorse this type of extension.
 

### 2022-08-27 paper/harm.Rmd
editing outside intro

@@
 First, it considers only individuals who vote: people who abstain from voting have a harm of zero.
-Empirically, this changes nothing, but limits the measure in a theoretical sense.
-People who do not voter express no formal preference, even if they have informal preferences.
+People who do not voter express no formal preference in electoral outcomes, even if they have informal preferences.
 The natural theoretical extension of this definition is to then separate individual preferences from vote choice. 
+The latter defines election winners, but the former provides a more-broad harm measure.
 Naturally, extending this approach requires auxiliary data to implement, such as an absurdly large-scale survey, as candidate preferences for non-voters are not directly inferred from election returns.
@@
 Ideally, we would average individual harm (1) over a set of future elections under the enacted districting plan and (2) over a set of hypothetical districting plans which are consistent with districting criteria.
-We could even consider a set of individuals with some shared characteristic(s).
+We could even consider a set of individuals with some shared characteristics.
 
@@
 Election results present a different set of modeling challenges, given the large number of factors that influence the outcome, and the general unpredictability of future events.
-We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
+We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting. 
+Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
 Multiparty or primary elections pose additional modeling challenges.
@@
 
-The modern analysis of racial gerrymandering and vote dilution are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
+The modern analysis of vote dilution and racial gerrymandering are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
 Section 2 of the VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
@@
 For example, with partisan bias, the implicit electoral counterfactual is an election in which the parties' vote shares have been reversed, and the implicit counterfactual districting plan is one which assigns equal vote shares to the parties under both the real-world and counterfactual electoral scenarios.
-Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partisan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome." Similar implicit counterfactuals can be thought up for the other measures.
+Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partisan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome."
+Similar implicit counterfactuals can be thought up for the other measures.
 
@@
 
-Taken together, these findings suggest that despite the dominance of the Democratic party in city politics, and the shared preferences between certain groups of Black and White voters, race-aware districting still serves an important role in empowering Black voters across the city to elect their chosen candidates, and that a switch to race-neutral boundaries would have a material impact on Black voting power.
+Taken together, these findings suggest that despite the dominance of the Democratic party in city politics, and the shared preferences between certain groups of Black and White voters, race-aware districting still serves an important role in empowering Black voters across the city to elect their chosen candidates.
+A switch to race-neutral boundaries would have a material impact on Black voting power.
 Further study could focus on the racial impact of the four at-large members in comparison to a system with 13 districts and no at-large members.

### 2022-08-28 paper/harm.Rmd
premerge

--- 2022-08-27 paper/harm.Rmd
@@
   Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
-  But these measures are not easily adapted to other groups, 
-  including groups defined by race, class, or geography.
-  Nor are they applicable to single- or no-party contexts, such as local redistricting.
+  But these measures cannot be easily adapted to other groups, 
+  including those defined by race, class, or geography.
+  Nor are they applicable to single- or no-party contexts such as local redistricting.
   To overcome these limitations, we propose a unified framework of harm for evaluating 
@@
 
-To motivate individual and differential harm, we use *Minissouri*, a stylized representation of Missouri.
+To introduce and motivate the concepts of individual and differential harm, we will make use of Minissouri, a stylized representation of Missouri.
 The state's total population is 180, and it should be divided into three districts with population between 54 and 66.
@@
 
-Now suppose the redistricting plan in Figure \ref{fig:minissouri}(a) is the enacted plan.
-Despite the statewide popular vote favoring Republicans, under this plan Democrats win two of the three seats.
-In contrast, a more proportional plan \ref{fig:minissouri}(b) elect two Republicans and one Democrat.
+Now suppose the redistricting plan in Figure \ref{fig:minissouri}(a) has been enacted.
+Despite the statewide popular vote favoring Republicans, under this plan Democrats would win two of the three seats.
+In contrast, a more proportional plan \ref{fig:minissouri}(b) would elect two Republicans and one Democrat.
 
 Our intuition tells us that plan (a) unfairly packs Republicans into a rural district, depriving them of one seat.
-To turn this intuition into something more concrete, we can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
+To turn this intuition into something more solid, we can consider which voters are harmed by the enactment of plan (a) over the alternative (b).
 These voters are highlighted in Figure \ref{fig:minissouri}(c).
@@
 
-After identifying which voters are harmed by plan (a) compared to counterfactual plan (b), it is natural to examine the overall composition of these harmed voters.
+After identifying which voters are harmed by plan (a) (with plan (b) as a counterfactual), it is natural to examine the overall composition of these harmed voters.
 Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
@@
 We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
-Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points more likely to be represented by a Democrat than the average Democrat is to be represented by a Republican, compared to plan (b).
+Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the proportional plan.
 
 We need not limit our consideration to partisanship.
-We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship, finding that 13 circle voters and 33 square voters are harmed.
+We can also tally harm by race, which in Minissouri is correlated with both geography and partisanship, finding that 13 circle voters and 33 square voters are harmed.
 This corresponds to a differential racial harm of $-20\%$, where positive values indicate pro-square bias.
@@
 
-We formalize and generalize the concepts in Section \ref{sec:motivation}.
-Let $\Voters$ be the set of individuals, or *voters*, in the region and time period.
-Individuals vote for a candidate to represent districts.
-Districts are indexed by $\Distrs$.
+We formalize and generalize the concepts in Section \ref{sec:motivation}, beginning with some notation.
+Let $\Voters$ be the set of individuals, or voters, in the region and time period under consideration, who vote to select a candidate to represent each of a set of districts, indexed by $\Distrs$.
 
@@
 
-Every voter resides in a legislative district.,
-Each assignment defines a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
+Every voter resides in some legislative district; this assignment is referred to as a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
 The set of all possible districting plans is $\Plans$.
 
-For election $v$, we tally the votes for each candidate.
-The candidate with the most votes wins.
+For an election $v$, we can tally the votes for each candidate; the candidate with the most votes wins.
 Using potential outcomes notation, we denote the winner of voter $i$'s district under districting plan $q$ and election $v$ as $\win_i(v,q)$.
@@
 
-The key intuition from Section \ref{sec:motivation} is that *to be harmed*, a voter must have a better outcome in the counterfactual world where a different decision was made.
-To operationalize this, we rely on a counterfactual districting plan $\tilde q$.
-Then we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as
-
-$$h_i(v,q,\tilde q) = \ind\{v(i) \neq \win_i(v, q)\} \ind\{v(i) = \win_i(v, \tilde q)\}.$$
-
-A voter is harmed only if they are not represented by the candidate of their choice under $q$ but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$.
+The key intuition from Section \ref{sec:motivation} was that to be harmed by a decision a voter must have a better outcome in the counterfactual world where another decision was made.
+To operationalize this, we need to have a counterfactual districting plan $\tilde q$.
+Then we may define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as $$
+h_i(v,q,\tilde q) = \ind\{v(i) \neq \win_i(v, q)\}
+        \ind\{v(i) = \win_i(v, \tilde q)\},
+$$that is, a voter is harmed only if they are not represented by the candidate of their choice under $q$ and they would have been represented by their chosen candidate under the counterfactual plan $\tilde q$.
 Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$.
-Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion.
-In general, $q$ will harm some individuals while $\tilde q$, and $\tilde q$ will harm a different set of individuals.
-A symmetric version of $h_i(v,q,\tilde q)$ would need to combine both of these groups: those harmed and those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
-In practice, voters, courts, or politicians may find different balances in preventing harm and maximizing gains.
+Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion; in general, $q$ will harm some individuals versus $\tilde q$, and $\tilde q$ will harm a distinct set of individuals versus $q$.
+A "symmetrized" version of $h_i(v,q,\tilde q)$ would have to combine both of these groups individuals: those harmed as well as those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
+In practice, voters, courts, or politicians may care more about preventing harm than maximizing gains.
 
 Our definition of harm is limited in two key regards.
-First, it considers only individuals who vote: people who abstain from voting have a harm of zero.
-A natural extension of this definition is to separate individual candidate preference from vote choice. 
-The latter defines election winners, but the former provides a more-broad harm measure.
-Naturally, this approach requires auxiliary data to implement, such as a large-scale survey, as candidate preferences for non-voters are not directly inferred from election returns.
+First, it considers only individuals who vote; if a person abstains from voting then their calculated harm is zero.
+A natural extension of this definition would be to separate an individual's candidate preference from their vote choice; the latter would be used in calculating election winners, but the former could provide a more broadly-based harm measure.
+Naturally, this approach requires auxiliary data to implement (such as a large-scale survey), since candidate preferences for nonvoters cannot be directly inferred from election returns.
 
-A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between the candidates that he or she did not vote for.
+A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between candidates he or she did not vote for.
 For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
 If the voter ultimately voted for candidate A, this harm measure makes no distinction between candidate B or candidate C winning the voter's district.
-This reflects the simplistic nature of the cost function assumed in single-choice elections.
-The problem could be remedied by replacing the indicator functions in the harm measure with other cost functions that consider ordinal preferences.
+This reflects the simplistic nature of the cost function that is implicit in the harm measure.
+The problem could be remedied by replacing the indicator functions with other cost functions that consider ordinal preferences.
 Ballot-level ranked-choice vote data, or candidate ideology scores, could be used to estimate such a cost function.
-Here, for simplicity and to avoid making further assumptions, we focus on the indicator-based cost function.
-Section \ref{sec:boston} further discusses the implications of this particular specification, in the context of primary elections.
+Here, for simplicity and to avoid making further assumptions, we limit ourselves to the indicator-based cost function.
+Section \ref{sec:boston} further discusses the implications of this particular specification in the context of primary elections.
 
@@
 
-As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
-In general, we are interested in outcomes over multiple elections and a much larger set of counterfactual plans.
-Ideally, we would average individual harm (1) over a set of future elections under the enacted districting plan and (2) over a set of hypothetical districting plans which are consistent with districting criteria.
-We could even consider a set of individuals with some shared characteristic(s).
+As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to a single counterfactual districting plan $\tilde q$.
+In general we will be interested in outcomes over multiple elections, and possibly a much larger set of counterfactual plans.
+Ideally, we would like to be able to average individual harm over a set of future elections (under the same districting plan), or over a set of hypothetical districting plans which are consistent with districting criteria, or even over a set of individuals with some shared characteristic.
 
 To allow for this, we replace the fixed election $v$, counterfactual districting plans $\tilde q$, and individual voter $i$ with random variables $V\in\Elecs$, $\tilde Q\in\Plans$, and $I\in\Voters$.
-The random variable $V$ can be conceptualized as ranging over possible future elections---one realization could correspond to a heavily Democratic-favoring midterm year, while another could involve incumbents underperforming particularly badly.
+The random variable $V$ can be conceptualized as ranging over possible future elections---one realization might correspond to a heavily Democratic-favoring midterm year, while another might involve incumbents underperforming particularly badly.
 The random variable $\tilde Q$ captures the range of potential counterfactual districting plans.
-Rarely would an analyst have a single counterfactual plan in mind.
-Different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
+Rarely would an analyst have a single counterfactual plan in mind; different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
 Finally, $I$ is defined to be a voter taken uniformly at random from $\Voters$, with $I\indep V,\tilde Q$.
@@
 The randomness in $V$ and $\tilde Q$ allows us to compare the enacted districting plan to many plausible counterfactual alternatives, and across a range of electoral environments.
-At times we may wish to study on a particular election or a particular counterfactual plan.
-If we treat $V$ and $\tilde Q$ as random variables, we can then condition on particular values of each.
+At times we may wish to study on a particular election or a particular counterfactual plan; treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
 A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section \ref{sec:estimating}.
 
-With these definitions, the harm for a (randomly selected) individual under plan $q$ can be written as 
-$$H(q) \dfeq h_I(V, q, \tilde Q).$$
+With these definitions, the harm for a (randomly selected) individual under plan $q$ can be written as $$
+H(q) \dfeq h_I(V, q, \tilde Q).
+$$$H(q)$ is itself a random variable, representing the harm to *some* individual in *some* election compared to *some* counterfactual plan.
+The random variable $I$ here serves as a kind of "veil of ignorance" [@rawls2004theory]; a districting plan $q$ which tends to produce a low $H(q)$ would be expected to be preferred over a plan which tended to produce a higher $H(q)$.
 
-$H(q)$ is itself a random variable, representing the harm to *some* individual in *some* election compared to *some* counterfactual plan.
-The random variable $I$ here serves as a kind of "veil of ignorance" [@rawls2004theory]: a districting plan $q$ which tends to produce a low $H(q)$ would be expected to be preferred over a plan which tended to produce a higher $H(q)$.
-
-$H(q)$ describes the harm created by a plan $q$.
-We can zoom in to specific dimensions of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
-Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in a particular election $v$ becomes
-
-$$\E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q) \text{ and }v(I)=\win_I(v, \tilde Q)],$$
-    that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
+$H(q)$ actually contains all the relevant information about the harm created by a plan $q$.
+We can "zoom in" or summarize various aspects of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
+Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in a particular election $v$ becomes$$
+\E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q)
+    \text{ and }v(I)=\win_I(v, \tilde Q)],
+$$that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
 
@@
 
-Single-member districts are often motivated by the value in representation to specific communities with shared interests.
-Indeed, preserving "communities of interest" is one of the most common statutory and constitutional requirements imposed by states on the plan-drawing process [@ncslcriteria].
+Single-member districts are often motivated by the value in representation to specific communities with shared interests; indeed, preserving "communities of interest" is one of the most common statutory and constitutional requirements imposed by states on the plan-drawing process [@ncslcriteria].
 It is therefore natural to consider not just the average harm to voters done by a plan, but also how this harm is distributed across certain groups of voters, defined by race, party, or other attributes of interest.
@@
 
-Fairness is a strict condition: real-life plans are incredibly unlikely to be exactly fair.
+Fairness is a strict condition; any real-life plan will never be exactly fair.
 To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
@@
 If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
-In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined.
-There is certainly an element of subjectivity in sampling a set of districting plans.
+In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined; there is certainly an element of subjectivity in sampling a set of districting plans.
 Nonetheless, sampling methods are widely used in practice [see e.g., @2019rucho; @2018league; @2022merrill; @becker2021computational; @chen:rodd:13].
@@
 [^2]: For this, we use the baseline count of Democratic or Republican votes.
-    This is a slight simplification, since the vote share changes across random draws, the number of Democratic and Republican voters changes too.
-    To deal with this issue completely necessitates modeling elections at the precinct level, which requires more data and computational costs.
+    This is a slight simplification, since because the vote share changes across random draws, the number of Democratic and Republican voters does, too.
+    To deal with this issue completely necessitates modeling elections at the precinct level, which requires much more data and computational expenditure.
 
@@
 
-A primary advantage of our proposed harm measures is that they our measures are defined at the individual level.
-This allows for the study of redistricting far beyond racial and partisan gerrymandering.
-The measures provide an improvement over existing analytical approaches in both of these important contexts.
+A primary advantage of the proposed harm measures is that they are defined at the individual level, allowing for the study of redistricting far beyond racial and partisan gerrymandering.
+However, the measures provide an improvement over existing analytic approaches in both of these important contexts.
 
@@
 While the specific legal requirements for a vote dilution claim will obviously remain front-and-center for court challenges of racial gerrymanders, it is notable that the proposed harm measures unify all three *Gingles* criteria.
-Criteria (b) and (c) capture the misrepresentation of minority voters by a candidate they do not support, while (a) requires the existence of a counterfactual where this would not be the case and (a) and (c) together establish that minority voters are harmed.
+Criteria (b) and (c) captures the misrepresentation of minority voters by a candidate they do not support, while (a) requires the existence of a counterfactual where this would not be the case; together, (a) and (c) establish that minority voters are harmed.
 But the proposed measures can also tackle trickier situations where the *Gingles* criteria may not be clearly met.
@@
 The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-3.6\%$ for the Democratic proposal and $2.0\%$ for the Republican proposal.
-These values are relatively directly interpretable.
-For example, the estimated differential harm of $-3.6\%$ means that an average Democratic voter is 3.6 percentage points less likely to be misrepresented than their average Republican counter-partisan.
+These values are relatively directly interpretable; for example, the estimated differential harm of $-3.6\%$ means that an average Democratic voter is 3.6 percentage points less likely to be misrepresented than their average Republican counter-partisan.
 The efficiency gap, declination, and partisan bias measures agree with the differential harm estimates in this regard, though the former two are much more discrete.
@@
 For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates that advanced to the general election, we were able to predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
+The accuracy of these validation predictions tracks with previous work that also used factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
 We stress that the latent factor ecological inference model is not necessary for this kind of analysis; local survey data that collected ideological preferences and racial information would also suffice.
@@
 This will produce a different answer, since harm is not symmetric between the study plan and the counterfactual.
-Comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline.
-The reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
+Comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline; the reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
 For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan Along the second dimension, the differential harm to Black voters would be 1.8%.
@@
 We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior.
-We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$. 
-The former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).
-Several reparametrizations are made for computational efficiency.
-The full Stan model is included in the replication code.
+We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$; the former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).
+Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.
 
@@
 We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.5^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 4.0^2)$ prior, and $\logit(\mu^{(t)})$ an i.i.d. $\Beta(10, 30)$ prior.
-We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$.
-The former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
+We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$; the former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
 A larger LKJ shape parameter places more weight on correlation matrices closer to diagonal, i.e., it regularizes away from high correlations.
@@
 Overall, these priors weakly identify the location and scale of the parameters.
-Several reparametrizations were made for computational efficiency.
-The full Stan code for the model is available as part of the replication code.
+Several reparametrizations were made for computational efficiency; the full Stan code for the model is available as part of the replication code.
 

### 2022-08-28 paper/harm.Rmd
redraft intro + other edits

--- 2022-08-27 paper/harm.Rmd
@@
 72--73], is perhaps the most consistently visible manifestation of this tendency.
-The academic study of partisan gerrymandering has largely centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
-Other types of gerrymandering exist as well.
+But gerrymandering takes other forms as well.
 In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [@gilliland1969racial].
-The Voting Rights Act of 1965 (VRA) provided a framework for legal challenges to racial gerrymanders that has also shaped the academic literature, which is characterized by a focus on identifying the voting preferences of racial groups and observing the extent to which these are realized or frustrated through district boundaries and other features of the electoral system [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
 At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the electoral system may be more subtle or may favor a narrower or less well-defined group [@cain2002mapmaking; @halberstam2012process].
-Local redistricting has been less extensively studied, in part for this reason and in part because of data limitations [but see @siegel2013educational; @vargas2021racial].
+The diversity of types of gerrymandering is further apparent from a survey of major litigation over the past decades, which includes allegations of gerrymandering involving party [@2018league; @2019rucho], race [@1993shaw; @2022merrill], religion [@1977ujo; @2020ercsd], incarceration status [@2015calvin; @2016davidson], and residency or citizenship [@1966burns; @2016evenwel].
 
-No matter the source or context, all of these cases involve the concentration or dilution of voting power to favor or disfavor a particular group.
-This paper proposes a new framework of *individual and differential harm* for the unified study of all forms of gerrymandering by way of measuring this concentration and dilution.
+The academic study of gerrymandering has unfortunately not reflected this diversity.
+The vast majority of research attention has been on partisan gerrymandering, and has particularly centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
+A large number of formulas for quantifying or identifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18].
+The study of racial gerrymandering has been largely shaped by the U.S.
+legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA).
+This literature is characterized by a focus on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
+Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @vargas2021racial].
+
+The dominance of the partisan framework for analyzing redistricting poses a problem even in cases of pure partisan gerrymandering.
+Existing tools, by focusing only on the effect of district lines on parties, are unable to examine the effect of manipulated lines on other groups.
+A redistricting plan that diminishes Democratic voting power will naturally have an effect on the relative voting power of White and Black voters, poor and wealthy voters, young and old voters, and urban or rural voters.
+These effects on individuals will be missed by any framework that only studies parties.
+
+This paper addresses all of these limitations with a new framework of individual and differential harm for the unified study of all forms of gerrymandering.
+Fundamentally, gerrymandering involves the concentration or dilution of voting power to favor or disfavor a particular group.
+Our proposed framework focuses directly on measuring this concentration and dilution, and in doing so avoids the pitfalls of party-centric approaches.
 
@@
 
-Second, the proposed framework moves beyond a single electoral system under study to also consider *counterfactual* electoral systems.
+Second, the proposed framework moves beyond a single districting plan under study to also consider *counterfactual* districting plans.
 Many factors influence the performance and qualities of an electoral system.
-Without a well-defined comparison system, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain.
-As we demonstrate, modern redistricting simulation tools can provide a realistic set of counterfactuals that make these kinds of comparisons possible.
+Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
+Many existing partisan gerrymandering metrics make use of an implicit counterfactual, which may not be of interest or even feasible.
+By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.
 
@@
 And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
-Section \ref{sec:motivation} illustrates this with three simple examples, and the general definitions are formalized in Section \ref{sec:indiv-harm}.
-In Section \ref{sec:estimating}, we discuss how to estimate these measures from observed election results.
+Section \ref{sec:motivation} illustrates this with a simple example, and the general definitions are formalized in Section \ref{sec:indiv-harm}.
 
-This framework has important similarities with the common-law approach to voting rights litigation.
+A key feature of the proposed framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022].
+The challenge of estimating harm and differential harm from observed data is discussed in Section \ref{sec:estimating}.
+Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
+These algorithmically-generated sets of districting plans can provide a realistic set of counterfactuals for our measures.
+
+The harm framework has important similarities with the common-law approach to voting rights litigation.
 To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98].
@@
 
-Harm and differential harm are more than a useful conceptual unification---they are practically useful in a wide variety of redistricting analyses, including partisan gerrymandering of New Jersey congressional districts (Section \ref{sec:nj}), voting rights litigation for Alabama congressional districts (Section \ref{sec:alabama}), and racial impacts of district boundaries on Boston City Council primary elections (Section \ref{sec:boston}).
+Harm and differential harm are more than a useful conceptual unification---they are practically useful in a wide variety of redistricting analyses.
+A brief study of partisan gerrymandering of New Jersey congressional districts in Section \ref{sec:nj} shows how existing gerrymandering metrics can make incorrect conclusions about the magnitude and even the direction of partisan gerrymanders.
+In Section \ref{sec:alabama}, we study voting rights litigation for Alabama congressional districts, and use differential harm to separate the racial and partisan effects of the enacted districting plan.
+Section \ref{sec:boston} demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
 Section \ref{sec:discussion} concludes.
@@
 
-To motivate individual and differential harm, we use *Minissouri*, a stylized representation of Missouri.
+The core idea motivating individual and differential harm is a simple one.
+Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure \ref{fig:minissouri}.
 The state's total population is 180, and it should be divided into three districts with population between 54 and 66.
@@
 Voters' preferences are assumed to be stable across elections, and turnout is 100%.
-A schematic for this example is shown in Figure \ref{fig:minissouri}.
 In addition to partisan affiliation indicated by red or blue coloring, each voter also belongs to a racial group indicated by squares and circles.
@@
 
-Now suppose the redistricting plan in Figure \ref{fig:minissouri}(a) is the enacted plan.
+Now suppose the redistricting plan in Figure \ref{fig:minissouri}(a) has been enacted.
 Despite the statewide popular vote favoring Republicans, under this plan Democrats win two of the three seats.
@@
 
-Every voter resides in a legislative district.,
-Each assignment defines a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
+Every voter resides in a legislative district., Each assignment defines a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
 The set of all possible districting plans is $\Plans$.
 
-For election $v$, we tally the votes for each candidate.
-The candidate with the most votes wins.
+For an election $v$, we tally the votes for each candidate; the candidate with the most votes wins.
 Using potential outcomes notation, we denote the winner of voter $i$'s district under districting plan $q$ and election $v$ as $\win_i(v,q)$.
@@
 
-The key intuition from Section \ref{sec:motivation} is that *to be harmed*, a voter must have a better outcome in the counterfactual world where a different decision was made.
+The key intuition from Section \ref{sec:motivation} is that to be harmed, a voter must have a better outcome in the counterfactual world where a different decision was made.
 To operationalize this, we rely on a counterfactual districting plan $\tilde q$.
@@
 A symmetric version of $h_i(v,q,\tilde q)$ would need to combine both of these groups: those harmed and those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
-In practice, voters, courts, or politicians may find different balances in preventing harm and maximizing gains.
+In practice, voters, courts, or politicians may differ in their views on how to balance preventing harm and maximizing gains.
 
 Our definition of harm is limited in two notable regards.
-First, it considers only individuals who vote: people who abstain from voting have a harm of zero.
-People who do not voter express no formal preference in electoral outcomes, even if they have informal preferences.
-The natural theoretical extension of this definition is to then separate individual preferences from vote choice. 
-The latter defines election winners, but the former provides a more-broad harm measure.
-Naturally, extending this approach requires auxiliary data to implement, such as an absurdly large-scale survey, as candidate preferences for non-voters are not directly inferred from election returns.
-As surveys are typically benchmarked against electoral outcomes, we do not endorse this type of extension.
+First, it considers only individuals who vote: people who abstain from voting have experience zero harm.
+People who do not vote express no formal preference in electoral outcomes, even if they have informal preferences.
+The natural theoretical extension of this definition is to then separate individuals' candidate preferences from their vote choice; the latter defines election winners, but the former provides a more-broad harm measure.
+Naturally, extending this approach requires auxiliary data to implement, such as a large-scale survey, as candidate preferences for non-voters cannot be directly inferred from election returns.
 
-A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between the candidates that he or she did not vote for.
+A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between the candidates that they did not vote for.
 For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
 If the voter ultimately voted for candidate A, this harm measure makes no distinction between candidate B or candidate C winning the voter's district.
-This reflects the simplistic nature of the cost function assumed in single-choice elections.
-The problem could be remedied by replacing the indicator functions in the harm measure with other cost functions that consider ordinal preferences.
+This reflects the simplistic nature of the cost function implicitly assumed in our proposed measure.
+The problem could be remedied by replacing the indicator functions in the measure with other cost functions that consider ordinal preferences.
 Ballot-level ranked-choice vote data, or candidate ideology scores, could be used to estimate such a cost function.
@@
 To allow for this, we replace the fixed election $v$, counterfactual districting plans $\tilde q$, and individual voter $i$ with random variables $V\in\Elecs$, $\tilde Q\in\Plans$, and $I\in\Voters$.
-The random variable $V$ can be conceptualized as ranging over possible future elections---one realization could correspond to a heavily Democratic-favoring midterm year, while another could involve incumbents underperforming particularly badly.
+The random variable $V$ can be conceptualized as ranging over possible future elections---one realization could correspond to a heavily Democratic-favoring midterm year, while another could involve incumbents under-performing particularly badly.
 The random variable $\tilde Q$ captures the range of potential counterfactual districting plans.
-Rarely would an analyst have a single counterfactual plan in mind.
-Different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
+Rarely would an analyst have a single counterfactual plan in mind: different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
 Finally, $I$ is defined to be a voter taken uniformly at random from $\Voters$, with $I\indep V,\tilde Q$.
@@
 At times we may wish to study on a particular election or a particular counterfactual plan.
-If we treat $V$ and $\tilde Q$ as random variables, we can then condition on particular values of each.
+Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
 A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section \ref{sec:estimating}.
 
-With these definitions, the harm for a (randomly selected) individual under plan $q$ can be written as 
-$$H(q) \dfeq h_I(V, q, \tilde Q).$$
-
-$H(q)$ is itself a random variable, representing the harm to *some* individual in *some* election compared to *some* counterfactual plan.
+With these definitions, the harm for a (randomly selected) individual under plan $q$ can be written as $$
+H(q) \dfeq h_I(V, q, \tilde Q).
+$$ $H(q)$ is itself a random variable, representing the harm to *some* individual in *some* election compared to *some* counterfactual plan.
 The random variable $I$ here serves as a kind of "veil of ignorance" [@rawls2004theory]: a districting plan $q$ which tends to produce a low $H(q)$ would be expected to be preferred over a plan which tended to produce a higher $H(q)$.
 
-$H(q)$ describes the harm created by a plan $q$.
+Despite being only a scalar quantity, $H(q)$ actually fully captures the harm created by a plan $q$.
 We can zoom in to specific dimensions of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
-Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in a particular election $v$ becomes
-
-$$\E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q) \text{ and }v(I)=\win_I(v, \tilde Q)],$$
-    that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
+Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in a particular election $v$ becomes $$
+    \E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q) \text{ and }v(I)=\win_I(v, \tilde Q)],
+$$ that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
 
@@
 
-```{=html}
-<!--
-While not strictly necessary, modeling of $V$ and $\tilde Q$ independently will simplify the situation considerably.
-Practically, this translates to the assumption that individual voters would make the same candidate choice no matter the districting plan.
-To the extent that voters reward or punish an incumbent party for the districting plan they adopt, or that an adopted districting plan affects which candidates run in a district, this assumption is violated.
-But as our suggested model for election outcomes $V$ will be quite coarse, not accounting for candidate factors or policy preferences, the marginal effect of this independence assumption on our conclusions will be minimal.
--->
-```
 ## Modeling Districting Plans
@@
 [^2]: For this, we use the baseline count of Democratic or Republican votes.
-    This is a slight simplification, since the vote share changes across random draws, the number of Democratic and Republican voters changes too.
-    To deal with this issue completely necessitates modeling elections at the precinct level, which requires more data and computational costs.
+    This is a slight simplification, since the vote share changes across random draws, and so the number of Democratic and Republican voters changes too.
+    To deal with this issue completely, one must model elections at the precinct level, which requires more data and computational costs.
 
@@
 For example, with partisan bias, the implicit electoral counterfactual is an election in which the parties' vote shares have been reversed, and the implicit counterfactual districting plan is one which assigns equal vote shares to the parties under both the real-world and counterfactual electoral scenarios.
-Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partisan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome."
-Similar implicit counterfactuals can be thought up for the other measures.
+Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partisan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome." Similar implicit counterfactuals can be thought up for the other measures.
 
@@
 
-[^8]: The enacted plan is sourced from [All About Redistricting](https://redistricting.lls.edu/state/new-jersey/?cycle=2020&level=Congress&startdate=2021-12-22), and the GOP plan was provided by Harrison Neely, executive director for the GOP delegation to the New Jersey Congressional Redistricting Committe, following an Open Public Records Act request.
+[^8]: The enacted plan is sourced from [All About Redistricting](https://redistricting.lls.edu/state/new-jersey/?cycle=2020&level=Congress&startdate=2021-12-22), and the GOP plan was provided by Harrison Neely, executive director for the GOP delegation to the New Jersey Congressional Redistricting Committee, following an Open Public Records Act request.
 
@@
 Minimizing bias with other measures, however, could lead one astray: plans which score an efficiency gap of zero can have estimated average harm as high as 17.2% and estimated differential harm as extreme as $-6.1\%$, nearly twice the value of the Democratic proposal.
-Plans with zero partisan bias or mean-median score likewise are estmated to have higher-than-average harm and to tend to differentially harm Republicans.
+Plans with zero partisan bias or mean-median score likewise are estimated to have higher-than-average harm and to tend to differentially harm Republicans.
 
@@
     )^\top\vb L_k.
-\end{align*} Here, $F_j$ is a matrix of preferences by race and latent dimension within precinct $j$, so $F_{j}^\top(P_j\turn_j)^\top$ is a vector with elements containing the preferences of the precinct's voters in each latent dimension, after marginalizing out race.
-The vector $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $F_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
-Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
-As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
-Biden and Wu opponent Anissa Essaibi-George on the right.
-The second dimension appears primarily to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.
+\end{align*}
 
@@
 [^13]: For simplicity of presentation, the parameterization shown here has been altered slightly from that detailed in the appendix, though the ultimate model can be made identical after an adjustment to the prior on $\vb L_k$.
+
+Here, $F_j$ is a matrix of preferences by race and latent dimension within precinct $j$, so $F_{j}^\top(P_j\turn_j)^\top$ is a vector with elements containing the preferences of the precinct's voters in each latent dimension, after marginalizing out race.
+The vector $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $F_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
+Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
+As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
+Biden and Wu opponent Annissa Essaibi-George on the right.
+The second dimension appears primarily to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.
 
@@
 
-Figure \ref{fig:nj-pairs} expands on Figure \ref{fig:nj-meas} in the main text with pairwise scatterplots and correlation summaries.
+Figure \ref{fig:nj-pairs} expands on Figure \ref{fig:nj-meas} in the main text with pairwise scatter plots and correlation summaries.
 
@@
         &\sim \Norm(\mu^{(s)}, \Sigma^{(t)}),
-\end{align*} where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale, and $\circ$ indicates the elementwise (Hadamard) product.
+\end{align*} where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale, and $\circ$ indicates the element-wise (Hadamard) product.
 We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior.
@@
 The mean support $\mu^{(s)}$ receives an informative prior based on the racially polarized voting estimates of @kuriwaki2021geography, with $\mu^{(s)}\sim\Norm(\logit((0.18, 0.90, 0.35)), (0.03^2, 0.03^2, 0.2^2))$.
-These estimates were derived through multilevel regression and poststratification from survey data, calibrated to actual election results.
+These estimates were derived through multilevel regression and post-stratification from survey data, calibrated to actual election results.
 
@@
 As a result we choose to model the vote shares directly, rather than the vote counts, which is more computationally efficient.
-Because each precinct has dozens or hundreds of voters, a Normal approximation to the Binomial likelihood is appopriate, and so we model the vote shares $s_{jk}$ as \begin{align*}
+Because each precinct has dozens or hundreds of voters, a Normal approximation to the Binomial likelihood is appropriate, and so we model the vote shares $s_{jk}$ as \begin{align*}
     \logit(s_{jk}) \mid \pi_{jk}, \eps &\sim \Norm(\pi_{jk}, \frac{\eps}{1+n_{jv_k}}),

### 2022-08-28 paper/harm.Rmd
add MA

@@
 Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
-Many existing partisan gerrymandering metrics make use of an implicit counterfactual, which may not be of interest or even feasible.
+For example, in Massachusetts, it is currently impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
+The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
+While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, it may not be of substantive interest or even an actually feasible plan at all.
 By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

### 2022-08-28 paper/harm.Rmd
grammar

@@
 The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
-While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, it may not be of substantive interest or even an actually feasible plan at all.
+While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
 By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.

### 2022-08-30 paper/harm.Rmd
add abbott
edits

@@
 This literature is characterized by a focus on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
-Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @vargas2021racial].
+Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @abott2020large; @vargas2021racial].
 
@@
 
-This paper addresses all of these limitations with a new framework of individual and differential harm for the unified study of all forms of gerrymandering.
+This paper addresses these limitations with a new framework of *individual and differential harm* for the unified study of all forms of gerrymandering.
 Fundamentally, gerrymandering involves the concentration or dilution of voting power to favor or disfavor a particular group.
@@
 The harm framework combines two key ideas, which are motivated at greater length in Section \ref{sec:motivation}.
-First, we operate at the level of the individual voter, and examine whether each voter is represented by the candidate of their choice.
-Doing so both unifies our approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [see also @brunell2010].
+First, it operates at the level of the individual voter, and examine whether each voter is represented by the candidate of their choice.
+Doing so both unifies the approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [see also @brunell2010].
 
@@
 The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
+Once a set
+of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
 While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
@@
 
-The combination of individual voter choices and counterfactual electoral systems defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different electoral system.
+The combination of individual voter choices and counterfactual electoral systems defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different districting plan.
 And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
@@
 A brief study of partisan gerrymandering of New Jersey congressional districts in Section \ref{sec:nj} shows how existing gerrymandering metrics can make incorrect conclusions about the magnitude and even the direction of partisan gerrymanders.
-In Section \ref{sec:alabama}, we study voting rights litigation for Alabama congressional districts, and use differential harm to separate the racial and partisan effects of the enacted districting plan.
+Section \ref{sec:alabama}, studies voting rights litigation for Alabama congressional districts, and uses differential harm to separate the racial and partisan effects of the enacted districting plan.
 Section \ref{sec:boston} demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
@@
 Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure \ref{fig:minissouri}.
-The state's total population is 180, and it should be divided into three districts with population between 54 and 66.
-Minissouri has two medium-sized cities and a substantial rural population which an overall partisan breakdown of 55.6% R -- 44.4% D.
+The state's total population is 180, and it is divided into three districts with population between 54 and 66.
+Minissouri has two medium-sized cities and a substantial rural population which an overall partisan breakdown of 55.6% R to 44.4% D.
 Voters' preferences are assumed to be stable across elections, and turnout is 100%.
@@
 Now suppose the redistricting plan in Figure \ref{fig:minissouri}(a) has been enacted.
-Despite the statewide popular vote favoring Republicans, under this plan Democrats win two of the three seats.
-In contrast, a more proportional plan \ref{fig:minissouri}(b) elect two Republicans and one Democrat.
+Despite the statewide popular vote favoring Republicans, under this plan Democrats will always win two of the three seats.
+In contrast, a more proportional plan \ref{fig:minissouri}(b) elects two Republicans and one Democrat.
 
-Our intuition tells us that plan (a) unfairly packs Republicans into a rural district, depriving them of one seat.
+Intuitively, plan (a) unfairly packs Republicans into a rural district, depriving them of one seat.
 To turn this intuition into something more concrete, we can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
@@
 We need not limit our consideration to partisanship.
-We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship, finding that 13 circle voters and 33 square voters are harmed.
-This corresponds to a differential racial harm of $-20\%$, where positive values indicate pro-square bias.
-Square voters are 20 percentage points more likely to be harmed than circle voters.
-While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or merely aimed to maximize Democratic representation, it does quantify the impact of plan (a) on these voters versus the alternative plan (b).
+We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship.
+Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of \$-20\\%\$, where positive values indicate pro-square bias.
+In other words, square voters are 20 percentage points more likely to be harmed than circle voters.
+While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or merely aimed to maximize Democratic representation, it does quantify the impact of plan (a) on these voters versus the alternative plan (b), and it does so on the same scale as differential partisan harm.
 
@@
 Let $\Voters$ be the set of individuals, or *voters*, in the region and time period.
-Individuals vote for a candidate to represent districts.
-Districts are indexed by $\Distrs$.
+Individuals vote for a candidate to represent districts, which are indexed by $\Distrs$.
 
-Voters vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
+Individuals vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
 In an election, voters either vote for a candidate running in their district, or abstain from voting.
 We write $v(i)$ to mean the candidate that individual $i$ votes for in election $v$.
+The set of all individual vote choices (which may include the choice to abstain from voting) fully determines the specific election outcome.
 
-Every voter resides in a legislative district., Each assignment defines a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
+Every voter resides in some legislative district.
+This assignment defines a *districting plan* and is represented here by a function $q(i)$ indicating the district that individual $i$ lives in.
 The set of all possible districting plans is $\Plans$.
 
-For an election $v$, we tally the votes for each candidate; the candidate with the most votes wins.
+In an election $v$, the votes are tallied and the candidate with the most votes wins.
 Using potential outcomes notation, we denote the winner of voter $i$'s district under districting plan $q$ and election $v$ as $\win_i(v,q)$.
-
 This notation obscures the complex relationships and feedback effects between individuals, elections, and districting plans.
@@
 
-The key intuition from Section \ref{sec:motivation} is that to be harmed, a voter must have a better outcome in the counterfactual world where a different decision was made.
-To operationalize this, we rely on a counterfactual districting plan $\tilde q$.
-Then we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as
+The key intuition from Section \ref{sec:motivation} is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
+Letting $\tilde q$ denote a counterfactual districting plan, we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as
 
@@
 A voter is harmed only if they are not represented by the candidate of their choice under $q$ but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$.
+
 Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$.
@@
 
-Our definition of harm is limited in two notable regards.
+Our definition of harm is limited in two regards.
 First, it considers only individuals who vote: people who abstain from voting have experience zero harm.
@@
 The natural theoretical extension of this definition is to then separate individuals' candidate preferences from their vote choice; the latter defines election winners, but the former provides a more-broad harm measure.
-Naturally, extending this approach requires auxiliary data to implement, such as a large-scale survey, as candidate preferences for non-voters cannot be directly inferred from election returns.
+Naturally, implementing this extension in practice would require auxiliary data, such as a large-scale survey, since candidate preferences for non-voters cannot be directly inferred from election returns.
 
@@
 Here, for simplicity and to avoid making further assumptions, we focus on the indicator-based cost function.
-Section \ref{sec:boston} further discusses the implications of this particular specification, in the context of primary elections.
+Section \ref{sec:boston} further discusses the implications of this particular specification in the context of primary elections.
 
@@
 In general, we are interested in outcomes over multiple elections and a much larger set of counterfactual plans.
-Ideally, we would average individual harm (1) over a set of future elections under the enacted districting plan and (2) over a set of hypothetical districting plans which are consistent with districting criteria.
-We could even consider a set of individuals with some shared characteristics.
+Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans which are consistent with districting criteria.
+We could also consider averaging over a set of individuals with some shared characteristics.
 
-To allow for this, we replace the fixed election $v$, counterfactual districting plans $\tilde q$, and individual voter $i$ with random variables $V\in\Elecs$, $\tilde Q\in\Plans$, and $I\in\Voters$.
+To allow for such averaging, we replace the fixed election $v$, counterfactual districting plan $\tilde q$, and individual voter $i$ with random variables $V\in\Elecs$, $\tilde Q\in\Plans$, and $I\in\Voters$.
 The random variable $V$ can be conceptualized as ranging over possible future elections---one realization could correspond to a heavily Democratic-favoring midterm year, while another could involve incumbents under-performing particularly badly.
@@
 $$ $H(q)$ is itself a random variable, representing the harm to *some* individual in *some* election compared to *some* counterfactual plan.
-The random variable $I$ here serves as a kind of "veil of ignorance" [@rawls2004theory]: a districting plan $q$ which tends to produce a low $H(q)$ would be expected to be preferred over a plan which tended to produce a higher $H(q)$.
+The random variable $I$ here serves as a kind of "veil of ignorance" [@rawls2004theory]: without picking a specific individual, a districting plan $q$ which tends to produce a low $H(q)$ would be expected to be preferred over a plan which tended to produce a higher $H(q)$.
 
-Despite being only a scalar quantity, $H(q)$ actually fully captures the harm created by a plan $q$.
+Despite being only a scalar quantity, $H(q)$ actually fully captures the harm created by the plan $q$.
 We can zoom in to specific dimensions of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
-Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$; and the electorate's average harm in a particular election $v$ becomes $$
+Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$, and the electorate's average harm in a particular election $v$ becomes $$
     \E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q) \text{ and }v(I)=\win_I(v, \tilde Q)],
@@
 Single-member districts are often motivated by the value in representation to specific communities with shared interests.
-Indeed, preserving "communities of interest" is one of the most common statutory and constitutional requirements imposed by states on the plan-drawing process [@ncslcriteria].
-It is therefore natural to consider not just the average harm to voters done by a plan, but also how this harm is distributed across certain groups of voters, defined by race, party, or other attributes of interest.
+Preserving "communities of interest" is one of the most common statutory and constitutional requirements imposed by states on the plan-drawing process [@ncslcriteria].
+It is therefore natural to consider not just the average harm to voters done by a plan, but also how this harm is distributed across certain groups of voters, defined by race, party, or other attributes.
 
@@
 
-Fairness is a strict condition: real-life plans are incredibly unlikely to be exactly fair.
+Fairness is a strict condition: real-life plans are highly unlikely to be exactly fair.
 To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
@@
 
-It is this measure of differential harm that we propose as a unified framework for evaluating a districting plan across a variety of dimensions.
+Differential harm provides a unified framework for evaluating a districting plan across a variety of dimensions.
 It is derived from an individual-level measure of representation and it accounts for the space of counterfactual possibilities.
-Importantly, it separates *vote choice* from *group membership*, allowing one to discuss how partisan election outcomes under a specific districting plan create harm for specific groups or their interactions, simply by conditioning $H(q)$ on a set of individuals.
+Importantly, it separates *vote choice* from *group membership*, allowing one to discuss how partisan election outcomes under a specific districting plan create harm for specific groups or their interactions, simply by conditioning $H(q)$ on different sets of voters.
 
@@
 Harm $H(q)$ and differential harm $\Delta H(q)$ as defined above require specification of a probability distribution on elections and districting plans.
-Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must aim to estimate.
+Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must estimate.
 Either way, past election data and an enacted districting plan alone are not enough to calculate these measures.
@@
 
-For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans which are geographically contiguous, have district populations within a deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
+For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans which are geographically contiguous, have district populations within a specified deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
 These maximum-entropy distributions put as few conditions as possible on the districting sampling process, as long as a set of constraints imposed by the researcher are met.
-If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
+If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan or other considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
 In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined.
@@
 
-Election results present a different set of modeling challenges, given the large number of factors that influence the outcome, and the general unpredictability of future events.
+Election results present a different set of modeling challenges, given the large number of factors that influence election outcomes and the general unpredictability of future events.
 We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting.
@@
 
-A primary advantage of our proposed harm measures is that they our measures are defined at the individual level.
-This allows for the study of redistricting far beyond racial and partisan gerrymandering.
-The measures provide an improvement over existing analytical approaches in both of these important contexts.
+A primary advantage of our proposed harm measures is that they our measures are defined at the individual level, which allows for the study of redistricting far beyond racial and partisan gerrymandering.
+However, even for these important contexts, the measures provide an improvement over existing analytical approaches.
 
@@
 
-The canonical framework for considering partisan fairness is partisan symmetry [@king1987democratic].
+The canonical framework for considering partisan fairness in redistricting is partisan symmetry [@king1987democratic].
 Partisan symmetry implies that the seats $s$ won by a party with vote share $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
@@
 Since none of the existing measures are centered on individual voters, they offer far less granular insight than the proposed harm framework, even in the case of studying partisan gerrymandering.
-For example, differential harm can be estimated at the precinct level and then plotted on a map to show not just *who* is harmed by a gerrymander, but *where* they reside.
+For example, differential harm can be estimated at the precinct level and then plotted on a map to show not just *which party* is harmed by a gerrymander, but *where* this harm occurs.
 
@@
 
-Despite these differences, there are also key empirical similarities in many cases.
+Despite these differences, there are also empirical similarities in many cases.
 The efficiency gap and is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
@@
 Since 1995, New Jersey congressional redistricting has been handled by a 13-member commission, with 6 political appointees from each party and one independent chair with *de facto* tiebreaking authority.
-For the 2020 redistricting cycle, each party submitted a proposed plan to the chair, John Wallace Jr.
+For the 2020 redistricting cycle, each party submitted a proposed plan to John Wallace Jr, the chair.
 Wallace ultimately selected the Democratic plan for enactment, "simply because in the last redistricting, the map was drawn by Republicans."[^7] Here, in contrast, we evaluate the two plans on the basis of existing gerrymandering metrics as well as through the proposed harm framework.
@@
 Both proposals are estimated to harm the average voter at a higher rate than most of the sampled plans, with the Republican plan again more extreme in this regard ($p=0.0008$ versus $p=0.061$ for the Democratic plan).
-Notably, the plans which minimize estimated total harm are also those which have near-zero estimated differential partisan harm.
+The plans which minimize estimated total harm are also those which have near-zero estimated differential partisan harm.
 Thus, at least in New Jersey, overall harm reduction is consistent with a fair map on a partisan basis.
@@
 New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
-Appendix \ref{app:corr} shows these correlations across simulated plans in eight states.
+Appendix \ref{app:corr} records these correlations across simulated plans in eight states.
 The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
@@
 As is custom in VRA analyses, we use an ecological inference model.
-The model detailed in Appendix \ref{app:alei}, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
+The model, detailed in Appendix \ref{app:alei}, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
 For example, we model precinct $j$'s observed turnout $n_{jv}$ in election $v$ and baseline turnout rates $\turn_{jr}$ by race $r$ as \begin{align*}
@@
 However, the same framework can be fruitfully applied to more complex settings, providing insights that go beyond those available through traditional methods.
+
+A major question in voting rights law and research is the extent to which voting must be racially polarized before race should enter into the redistricting equation and statutory protections like the VRA should kick in [@elmendorf2016racially].
+This question is particularly acute in cities, where general elections are characterized by Democratic dominance and low levels of racial polarized voting.
+When the locus of political conflict shifts to primary elections, racial patterns are harder to observe and generalize.
+Thus there remains uncertainty about the extent that race plays a factor in primary elections, and consequently the role it should play in designing urban legislative maps.
 
@@
 The vector $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $F_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
+
 Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
@@
 This paper offers a new framework for studying redistricting that brings together a focus on individual voter preferences and the causal-inference logic of counterfactuals.
-The combination of these ideas imbues the analysis of districting plans with the historical understandings of legal standing and harm.
+The combination of these ideas imbues the analysis of districting plans with historical understandings of legal standing and harm.
 This focus naturally leads to a new normative standard of plan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters.

### 2022-08-30 paper/harm.Rmd
word

@@
 When the locus of political conflict shifts to primary elections, racial patterns are harder to observe and generalize.
-Thus there remains uncertainty about the extent that race plays a factor in primary elections, and consequently the role it should play in designing urban legislative maps.
+Thus there remains uncertainty about the extent that race is a factor in primary elections, and consequently the role it should play in designing urban legislative maps.
 

### 2022-08-30 paper/harm.Rmd
draft edits

@@
       To whom correspondence should be addressed.
-      The authors thank Kosuke Imai, Gary King, Shiro Kuriwaki, 
+      The authors thank Steve Ansolabehere, Kosuke Imai, Gary King, Shiro Kuriwaki, 
       Tyler Simko, Nick Stephanopolous, and 
@@
 The harm framework combines two key ideas, which are motivated at greater length in Section \ref{sec:motivation}.
-First, it operates at the level of the individual voter, and examine whether each voter is represented by the candidate of their choice.
+First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
 Doing so both unifies the approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [see also @brunell2010].
@@
 Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
-For example, in Massachusetts, it is currently impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
+For example, in Massachusetts, it was impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
 The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
@@
 A key feature of the proposed framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022].
+This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry].
 The challenge of estimating harm and differential harm from observed data is discussed in Section \ref{sec:estimating}.
@@
 
-```{r minissouri, fig.width=6.75}
+```{r minissouri, fig.width=6.75, fig.pos='b'}
 #| fig.cap: |
@@
 Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points more likely to be represented by a Democrat than the average Democrat is to be represented by a Republican, compared to plan (b).
+In direct terms, this means, on account of their partisan affiliation, Democrats would be 41.5% less likely to be represented by a Democrat than Republicans are to be represented by a Republican.
 
@@
 We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship.
-Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of \$-20\\%\$, where positive values indicate pro-square bias.
+Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of $-20\%$, where positive values indicate pro-square bias.
 In other words, square voters are 20 percentage points more likely to be harmed than circle voters.
@@
 
-A voter is harmed only if they are not represented by the candidate of their choice under $q$ but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$.
+A voter is harmed only if they are not represented by the candidate of their choice under $q$ (i.e.
+$\ind\{v(i) \neq \win_i(v, q)\}$) but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$ (i.e.
+$\ind\{v(i) = \win_i(v, \tilde q)\}$).
 
-Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$.
+Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$,
+as $\ind\{v(i) \neq \win_i(v, q)\} \neq \ind\{v(i) \neq \win_i(v, \tilde q)\}$.
 Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion.
+
 In general, $q$ harms some individuals while $\tilde q$, and $\tilde q$ harms a distinct, though not necessarily disjoint, set of individuals.
@@
 Fairness is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
-For algorithmic fairness, the implicit counterfactual is a perfect predictor---compared to this perfect predictor, the actual decision process harms individuals whenever it makes a false positive or false negative error.
+In the algorithmic fairness literature, the implicit counterfactual is a perfect predictor and the actual decision process harms individuals whenever it makes a false positive or false negative error.
 Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.
@@
 Fairness is a strict condition: real-life plans are highly unlikely to be exactly fair.
-To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
+To quantify deviations from *fairness*, we introduce the concept of *differential harm*, defined as $$
   \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
@@
 We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
-Appendix \ref{app:elecmodel} contains more details about how we estimate and apply this model.
+Appendix \ref{app:elecmodel} contains more details on how we estimate and apply this model.
 
-More complicated election outcome models are certainly possible and straightforward to use with the measures proposed here.
-For example, one could operate on the logit scale, or add terms to model incumbency effects.
+More complicated election outcome models are certainly possible and are straightforward to use with the measures proposed here.
+For example, one could operate on the logit scale, model incumbency effects, or add terms to account for local factors.
 
@@
 With a model for election outcomes and districting plans in hand, calculation of $H(q)$ and $\Delta H(q)$ is straightforward.
-For each sample $\tilde q$ of a counterfactual districting plan, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
+For each sampled counterfactual districting plan, $\tilde q$, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
 We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^2]
@@
 
-A primary advantage of our proposed harm measures is that they our measures are defined at the individual level, which allows for the study of redistricting far beyond racial and partisan gerrymandering.
-However, even for these important contexts, the measures provide an improvement over existing analytical approaches.
+A primary advantage of our proposed harm measures is that they our measures are defined at the individual level. 
+As such, we can study redistricting for any defined group memberships.
+For social science, perhaps the two most important group memberships are that of race and party.
+For these important contexts, we begin by showing that our measures provide an improvement over existing analytical approaches.
 
@@
 The study of racial gerrymandering and vote dilution is considerably more complicated than that of partisan gerrymandering, due to the difficulty of understanding racial voting preferences when individual votes are secret.
-As such, no simple formulas exist to measure the precise racial impacts of a districting plan.
-Indeed, much of the academic literature has focused on the statistical challenge of estimating vote preferences and turnout by race [@king1997solution; @greiner2006ecological; @imai2016improving].
+No simple formulas exist to measure the precise racial impacts of a districting plan.
+Indeed, much of the academic literature has focused on the statistical challenges of estimating vote preferences and turnout by race [@king1997solution; @greiner2006ecological; @imai2016improving].
 
-The modern analysis of vote dilution and racial gerrymandering are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
+Modern analyses of vote dilution and racial gerrymandering are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
 Section 2 of the VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
-Under the current framework established by @1986gingles, to do so, plaintiffs must prove that (a) the minority group is large and geographically compact enough to form a majority in some hypothetical district, (b) the minority group is "politically cohesive," and (c) the majority group can and does vote to deny the minority group its choice of candidate [@hood2018legal] (the so-called *Gingles* criteria).
+Under the current framework for vote dilution established by @1986gingles, plaintiffs must prove that (a) the minority group is large and geographically compact enough to form a majority in some hypothetical district, (b) the minority group is "politically cohesive," and (c) the majority group can and does vote to deny the minority group its choice of candidate [@hood2018legal] (the so-called *Gingles* criteria).
 Criteria (a) is generally proven through a single example districting plan, while criteria (b) and (c) are shown using the statistical estimates of vote preferences and turnout by race.
 
-While the specific legal requirements for a vote dilution claim will obviously remain front-and-center for court challenges of racial gerrymanders, it is notable that the proposed harm measures unify all three *Gingles* criteria.
+While the specific legal requirements for a vote dilution claim will obviously remain front-and-center for court challenges of racial gerrymanders, our proposed harm measures unifies all three *Gingles* criteria.
 Criteria (b) and (c) capture the misrepresentation of minority voters by a candidate they do not support, while (a) requires the existence of a counterfactual where this would not be the case and (a) and (c) together establish that minority voters are harmed.
-But the proposed measures can also tackle trickier situations where the *Gingles* criteria may not be clearly met.
+But our proposed measures can also tackle trickier situations where the *Gingles* criteria may not be clearly met.
 
@@
 
-In both these cases, the proposed harm measures will quantify the extent to which a challenged districting plan differentially harms minority voters, taking into account their geographic and political cohesion, and the individual preferences of every voter.
+In both these cases, the proposed harm measures quantify the extent to which a challenged districting plan differentially harms minority voters.
+We can account for geographic sorting, political cohesion, and the individual preferences of every voter.
 Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.
@@
 [^3]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.
+They are exactly the votes which don't contribute to winning a seat.
 
@@
 
-Since none of the existing measures are centered on individual voters, they offer far less granular insight than the proposed harm framework, even in the case of studying partisan gerrymandering.
+Since none of the existing measures are centered on individual voters, they offer far less granular insights than the proposed harm framework, even in the case of studying partisan gerrymandering.
 For example, differential harm can be estimated at the precinct level and then plotted on a map to show not just *which party* is harmed by a gerrymander, but *where* this harm occurs.
@@
 Despite these differences, there are also empirical similarities in many cases.
-The efficiency gap and is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
+The efficiency gap is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
 Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
@@
 While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
-In the worst cases, this can lead to directionally wrong conclusions, as we will see in the next section.
+In the worst cases, this can lead to directionally wrong conclusions, as we see in the next section.
 
@@
 The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
+Seats are, of course, more discrete and cannot directly identify the local areas where gains and losses are derived.
 
@@
 
-To apply our proposed framework to study the council district boundaries, we need to be able to predict individual voting preferences under counterfactual districting plans.
+To apply our proposed framework to study the council district boundaries, we predict individual voting preferences under counterfactual districting plans.
 This is more difficult than in both New Jersey and Alabama, for two reasons.
-First, the city council is officially nonpartisan and city politics are dominated by the Democratic Party, so voting preferences are more complex than a simple partisan preference.
-To estimate election outcomes, we require some way of mapping observed support for particular candidates in one election onto counterfactual support for a hypothetical candidate in another election.
-Second, as in Alabama, we aim to examine the racial impact of the district system, and thus must estimate these voting preferences by race.
+First, the city council is officially nonpartisan, but city politics are dominated by the Democratic Party, so voting preferences are more complex than a simple partisan preference.
+To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election.
+Second, as in Alabama, we aim to examine the racial impact of the district system, and so must estimate these voting preferences by race.
 Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.
@@
 As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
-Biden and Wu opponent Annissa Essaibi-George on the right.
-The second dimension appears primarily to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.
+Biden and Wu's opponent, Annissa Essaibi-George, on the right.
+The second dimension appears to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.
 
-Like all ecological inference models, ours rests on several unverifiable assumptions.
+Like all ecological inference models, ours relies on several unverifiable assumptions.
 The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
-As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not strictly met.
-Yet we believe that the model is a useful simplification of reality that captures the major electoral dynamics that are relevant to our question of interest.
-For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates that advanced to the general election, we were able to predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
-The accuracy of these validation predictions tracks with previous work that also used factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
-We stress that the latent factor ecological inference model is not necessary for this kind of analysis; local survey data that collected ideological preferences and racial information would also suffice.
+As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not *strictly* met.
+Yet, the model is a useful simplification of reality which captures the major electoral dynamics relevant to our question of interest.
+For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
+The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
+The latent factor ecological inference model is not necessary for this kind of analysis; local survey data with ideological preferences and racial information would also suffice.
 
@@
 
-We limit our hypothetical elections to two candidates, to simplify the challenges of measuring harm with many candidates, over which voters may have different ranked preferences.
+Our hypothetical elections are focused on cases with two candidates, to simplify the challenges of measuring harm with multiple candidates, over which voters may have different ranked preferences.
 However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
@@
 
-We can also flip the setup and compute harm by averaging over the simulated plans and using the enacted plan as a counterfactual.
-This will produce a different answer, since harm is not symmetric between the study plan and the counterfactual.
-Comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline.
+Recall that harm is asymmetric.
+While that may be an undesirable property in some areas, here we can leverage that fact.
+We flip the setup and compute harm by averaging over the simulated plans with the enacted plan as the counterfactual.
+Here, comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline.
 The reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
@@
 This focus naturally leads to a new normative standard of plan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters.
+In other words, your personal characteristics should not impact your ability to be represented.
 
@@
 Fundamentally, harm and differential harm are unable to measure plans against an absolute benchmark, the way that other measures such as partisan symmetry are able to do.
-And while modern simulation methods are a valuable starting point to defining counterfactuals, these methods involve analyst discretion and interpretation of often-murky legal standards.
+And while modern simulation methods are a valuable method for defining counterfactuals, these approaches involve analyst discretion and interpretation of often-murky legal standards.
 When subjective choices are not represented consistently, different analyses may arrive at different conclusions about which voters are harmed and at what rate.

### 2022-08-30 paper/harm.Rmd
add gill cite
edit

@@
 Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
-For example, in Massachusetts, it was impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
+For example, in Massachusetts, it is currently impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
 The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
-Once a set
-of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
+Once a set of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
 While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
@@
 We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
-Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points more likely to be represented by a Democrat than the average Democrat is to be represented by a Republican, compared to plan (b).
-In direct terms, this means, on account of their partisan affiliation, Democrats would be 41.5% less likely to be represented by a Democrat than Republicans are to be represented by a Republican.
+Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points less likely to be represented by a Republican than the average Democrat is to be represented by a Democrat, *compared to plan (b)*.
 
@@
 
-A voter is harmed only if they are not represented by the candidate of their choice under $q$ (i.e.
-$\ind\{v(i) \neq \win_i(v, q)\}$) but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$ (i.e.
-$\ind\{v(i) = \win_i(v, \tilde q)\}$).
+A voter is harmed only if they are not represented by the candidate of their choice under $q$ ($\ind\{v(i) \neq \win_i(v, q)\}$) but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$ ($\ind\{v(i) = \win_i(v, \tilde q)\}$).
 
-Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$,
-as $\ind\{v(i) \neq \win_i(v, q)\} \neq \ind\{v(i) \neq \win_i(v, \tilde q)\}$.
+Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$, as $\ind\{v(i) \neq \win_i(v, q)\} \neq \ind\{v(i) \neq \win_i(v, \tilde q)\}$.
 Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion.
@@
 In general, we are interested in outcomes over multiple elections and a much larger set of counterfactual plans.
-Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans which are consistent with districting criteria.
+Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans that are consistent with districting criteria.
 We could also consider averaging over a set of individuals with some shared characteristics.
@@
 Fairness is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
-In the algorithmic fairness literature, the implicit counterfactual is a perfect predictor and the actual decision process harms individuals whenever it makes a false positive or false negative error.
+There, the implicit counterfactual is a perfect predictor.
+Compared to the perfect predictor, the actual decision process harms individuals whenever it makes a false positive or false negative error.
 Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.
@@
 Fairness is a strict condition: real-life plans are highly unlikely to be exactly fair.
-To quantify deviations from *fairness*, we introduce the concept of *differential harm*, defined as $$
+To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
   \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
@@
 
-For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans which are geographically contiguous, have district populations within a specified deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
+For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans that are geographically contiguous, have district populations within a specified deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
 These maximum-entropy distributions put as few conditions as possible on the districting sampling process, as long as a set of constraints imposed by the researcher are met.
@@
 More complicated election outcome models are certainly possible and are straightforward to use with the measures proposed here.
-For example, one could operate on the logit scale, model incumbency effects, or add terms to account for local factors.
+For example, one could operate on the logit scale, model incumbency effects, or include additional data from local elections.
 
@@
 With a model for election outcomes and districting plans in hand, calculation of $H(q)$ and $\Delta H(q)$ is straightforward.
-For each sampled counterfactual districting plan, $\tilde q$, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
+For each sampled counterfactual districting plan $\tilde q$, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
 We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^2]
@@
 
-A primary advantage of our proposed harm measures is that they our measures are defined at the individual level. 
-As such, we can study redistricting for any defined group memberships.
-For social science, perhaps the two most important group memberships are that of race and party.
-For these important contexts, we begin by showing that our measures provide an improvement over existing analytical approaches.
+A primary advantage of our proposed harm measures is that they our measures are defined at the individual level,
+allowing for the study of redistricting for any defined group memberships.
+For political science, two of the most important group memberships are perhaps race and party.
+Even in these well-studied contexts, our measures provide an improvement over existing analytical approaches.
 
@@
 In both these cases, the proposed harm measures quantify the extent to which a challenged districting plan differentially harms minority voters.
-We can account for geographic sorting, political cohesion, and the individual preferences of every voter.
+They account for geographic sorting, political cohesion, and the individual preferences of every voter.
 Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.
@@
 
-To apply our proposed framework to study the council district boundaries, we predict individual voting preferences under counterfactual districting plans.
+To apply our proposed framework to study the council district boundaries, must be able to we predict individual voting preferences under counterfactual districting plans.
 This is more difficult than in both New Jersey and Alabama, for two reasons.
-First, the city council is officially nonpartisan, but city politics are dominated by the Democratic Party, so voting preferences are more complex than a simple partisan preference.
+First, the city council is officially nonpartisan, and city politics are dominated by the Democratic Party, so voting choices are more complex than a simple partisan preference.
 To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election.
@@
 As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not *strictly* met.
-Yet, the model is a useful simplification of reality which captures the major electoral dynamics relevant to our question of interest.
+However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
 For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
 The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
-The latent factor ecological inference model is not necessary for this kind of analysis; local survey data with ideological preferences and racial information would also suffice.
+We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one choice out of many model which could provide estimated vote choices for individuals.
+Local survey data with ideological preferences and racial information would be a sufficient alternative.
 
@@
 
-Our hypothetical elections are focused on cases with two candidates, to simplify the challenges of measuring harm with multiple candidates, over which voters may have different ranked preferences.
+Our hypothetical elections involve only two candidates, to simplify the challenges of measuring harm with multiple candidates, over which voters may have different ranked preferences.
 However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
@@
 
-Recall that harm is asymmetric.
-While that may be an undesirable property in some areas, here we can leverage that fact.
-We flip the setup and compute harm by averaging over the simulated plans with the enacted plan as the counterfactual.
-Here, comparing the enacted plan to the the simulations shows who is harmed by the current plan versus a race-neutral baseline.
-The reverse comparison shows who would be harmed by a race-neutral baseline in comparison to the current plan.
+These two harm estimates reflect the harm of the current boundaries compared to a race-neutral baseline.
+The more policy-relevant direction might be in evaluating a change from the status quo.
+What is the harm of switching to race-neutral districting from the current race-aware boundaries?
+We can flip the above setup and compute harm by averaging over the simulated plans with the enacted plan as the counterfactual.
 For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan Along the second dimension, the differential harm to Black voters would be 1.8%.
@@
 Fundamentally, harm and differential harm are unable to measure plans against an absolute benchmark, the way that other measures such as partisan symmetry are able to do.
-And while modern simulation methods are a valuable method for defining counterfactuals, these approaches involve analyst discretion and interpretation of often-murky legal standards.
+And while modern simulation methods are a valuable starting point for defining counterfactuals, these approaches involve analyst discretion and interpretation of often-murky legal standards.
 When subjective choices are not represented consistently, different analyses may arrive at different conclusions about which voters are harmed and at what rate.

### 2022-08-30 paper/harm.Rmd
typos

@@
   and directly incorporating counterfactual plans.
-  We discuss strategies for estimating harm, and demonstrate the utility of our framework 
+  We discuss strategies for estimating harm, and demonstrate the utility of our framework, 
   through analyses of partisan gerrymandering in New Jersey, voting rights 
@@
 This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
+Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
 A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.
@@
 The state's total population is 180, and it is divided into three districts with population between 54 and 66.
-Minissouri has two medium-sized cities and a substantial rural population which an overall partisan breakdown of 55.6% R to 44.4% D.
+Minissouri has two medium-sized cities and a substantial rural population with an overall partisan breakdown of 55.6% R to 44.4% D.
 Voters' preferences are assumed to be stable across elections, and turnout is 100%.
@@
 Our definition of harm is limited in two regards.
-First, it considers only individuals who vote: people who abstain from voting have experience zero harm.
+First, it considers only individuals who vote: people who abstain from voting have experienced zero harm.
 People who do not vote express no formal preference in electoral outcomes, even if they have informal preferences.
@@
 
-A primary advantage of our proposed harm measures is that they our measures are defined at the individual level,
-allowing for the study of redistricting for any defined group memberships.
+A primary advantage of our proposed harm measures is that they are defined at the individual level, allowing for the study of redistricting for any defined group memberships.
 For political science, two of the most important group memberships are perhaps race and party.
@@
 This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
-Plugging in past election results without using an election model places assumes too much about the stability of electoral dynamics and consequently can significantly understate the uncertainty in how a particular districting plan will perform into the future.
+Plugging in past election results without using an election model assumes too much about the stability of electoral dynamics and consequently can significantly understate the uncertainty in how a particular districting plan will perform into the future.
 
@@
 
-These values can be better contextualized in comparison the distribution across the sampled plans.
+These values can be better contextualized in comparison to the distribution across the sampled plans.
 The Democratic proposal is expected to produce an average of 8.55 Democratic seats under our election model, more than 79% of all sample plans.
@@
 
-To apply our proposed framework to study the council district boundaries, must be able to we predict individual voting preferences under counterfactual districting plans.
+To apply our proposed framework to study the council district boundaries, we must be able to predict individual voting preferences under counterfactual districting plans.
 This is more difficult than in both New Jersey and Alabama, for two reasons.
@@
 The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
-We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one choice out of many model which could provide estimated vote choices for individuals.
+We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one out of several approaches which could provide estimated vote choices for individuals.
 Local survey data with ideological preferences and racial information would be a sufficient alternative.
@@
 Simple models are easy to build from commonly-available data and are computationally tractable, but are gross simplifications of reality.
-More realistic models require more analyst thought and can add significant burden to the calculation of our harm measures.
+More realistic models require more analyst thought and can add a significant burden to the calculation of our harm measures.
 And as discussed in Section \ref{subsec:harm}, multi-candidate elections where voters have strong ordinal preferences provide further challenges that the proposed framework does not fully address.
@@
 To account for the fact that district boundaries change every decade, the district identifiers are built as a concatenation of the state, district number, and redistricting.
-For example, the Washington's 7th congressional district receives one random effect from 2002-2010 and a different random effect from 2012-2020, since new district boundaries were enacted in the 2010-2012 period.
+For example, Washington's 7th congressional district receives one random effect from 2002-2010 and a different random effect from 2012-2020, since new district boundaries were enacted in the 2010-2012 period.
 We do not account for any mid-decade redistricting which may occur as a result of litigation.
@@
 The mean support $\mu^{(s)}$ receives an informative prior based on the racially polarized voting estimates of @kuriwaki2021geography, with $\mu^{(s)}\sim\Norm(\logit((0.18, 0.90, 0.35)), (0.03^2, 0.03^2, 0.2^2))$.
-These estimates were derived through multilevel regression and post-stratification from survey data, calibrated to actual election results.
+These estimates were derived through multilevel regression and poststratification from survey data, calibrated to actual election results.
 
@@
 This was accomplished through somewhat of an interactive process, where a preliminary principal components decomposition of the precinct vote shares helped suggest the latent factors that might be found by the model.
-We chose to make align the first dimension with the Sanders-Biden axis, and chose to make the second dimension roughly orthogonal along the Janey-Buttigieg axis.
+We chose to align the first dimension with the Sanders-Biden axis, and chose to make the second dimension roughly orthogonal along the Janey-Buttigieg axis.
 This translated to a prior of $\Norm((-1, 0), (0.1^2, 0.25^2))$ for Sanders, $\Norm((1, 0), (1.0^2, 0.5^2))$ for Biden, and $\Norm((-0.25, -1), (0.5^2, 2^2))$ for Janey.

### 2022-08-30 paper/harm.Rmd
Update harm.Rmd

@@
   and directly incorporating counterfactual plans.
-  We discuss strategies for estimating harm, and demonstrate the utility of our framework, 
+  We discuss strategies for estimating harm, and demonstrate the utility of our framework 
   through analyses of partisan gerrymandering in New Jersey, voting rights 

### 2022-12-06 paper/harm.Rmd
str_replace electoral system

@@
 In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [@gilliland1969racial].
-At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the electoral system may be more subtle or may favor a narrower or less well-defined group [@cain2002mapmaking; @halberstam2012process].
+At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the districting plan may be more subtle or may favor a narrower or less well-defined group [@cain2002mapmaking; @halberstam2012process].
 The diversity of types of gerrymandering is further apparent from a survey of major litigation over the past decades, which includes allegations of gerrymandering involving party [@2018league; @2019rucho], race [@1993shaw; @2022merrill], religion [@1977ujo; @2020ercsd], incarceration status [@2015calvin; @2016davidson], and residency or citizenship [@1966burns; @2016evenwel].
@@
 The academic study of gerrymandering has unfortunately not reflected this diversity.
-The vast majority of research attention has been on partisan gerrymandering, and has particularly centered on how electoral systems treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
+The vast majority of research attention has been on partisan gerrymandering, and has particularly centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
 A large number of formulas for quantifying or identifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18].
@@
 Second, the proposed framework moves beyond a single districting plan under study to also consider *counterfactual* districting plans.
-Many factors influence the performance and qualities of an electoral system.
+Many factors influence the performance and qualities of a districting plan.
 Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
@@
 
-The combination of individual voter choices and counterfactual electoral systems defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different districting plan.
+The combination of individual voter choices and counterfactual districting plans defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different districting plan.
 And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
@@
 @katz2020 and others recognize that a nonzero value in a measure like partisan bias can only be considered problematic "if there is a fairer alternative." The inability of partisan bias and other existing measures to explicitly account for realistic counterfactual districting plans means that none of them may be interpreted in a vacuum.
-No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the district-based electoral system.
+No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the electoral system's geography.
 
@@
 A districting plan can never be boiled down to a single number.
-But we hope our approach provides a modular and flexible set of measures that can help decompose how various components of the electoral system combine to produce observed patterns and election outcomes.
+But we hope our approach provides a modular and flexible set of measures that can help decompose how various components of the districting plan combine to produce observed patterns and election outcomes.
 Researchers can and should tinker with our assumptions and push our framework further to better apply it to multi-candidate elections, more realistic election models, and beyond district-based electoral systems.

### 2022-12-06 paper/harm.Rmd
update thanks

@@
       To whom correspondence should be addressed.
-      The authors thank Steve Ansolabehere, Kosuke Imai, Gary King, Shiro Kuriwaki, 
-      Tyler Simko, Nick Stephanopolous, and 
-      participants of the Fall 2021 Harvard American Politics Research Workshop 
+      The authors thank Steve Ansolabehere, Matt Blackwell, Kosuke Imai, Gary King, 
+      Shiro Kuriwaki, Tyler Simko, Nick Stephanopolous, and 
+      participants of the Fall 2021 Harvard American Politics Research Workshop,
+      Fall 2022 Harvard Applied Statistics Workshop, 
       and Imai Research Group for helpful comments and discussion.

### 2022-12-06 paper/harm.Rmd
AL appendx + pres figures

--- 2022-08-30 paper/harm.Rmd
@@
 These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure \ref{fig:al-maps}(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
+In Appendix \ref{app:alsampling}, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.
 
@@
 
+# Additional Sampling Analysis for Alabama {#app:alsampling}
+
+Here, we perform the same analysis as in Section \ref{sec:alabama}, but using a randomly sampled set of plans as a counterfactual, rather than the four plans used in litigation.
+Specifically, we use 2,000 plans provided by the 50-State Simulation Project [@mccartan2022fiftystates].
+These plans are designed to discourage packing of Black voters and maintain at least the same number of majority-opportunity districts as the enacted 2020 plan.
+
+```{r alabama-harm-sample, fig.width=6.5}
+#| fig.cap: |
+#|      On thel eft, the expected fraction of voters in each precinct which are
+#|      harmed by the 2020 enacted congressional districts versus the sampled 
+#|      counterfactual set.
+#|      On the right, the expected number of harmed voters in each race/partisan 
+#|      group, relative to the total number of such voters.
+include_graphics("figures/al_harm_sample.pdf")
+```
+
+Figure \@ref(fig:alabama-harm-sample) summarizes the results, which are mostly similar to the results in in Section \ref{sec:alabama}.
+Across the state, the estimated average harm is 6.9%, which translates to around 144,000 voters.
+Of these harmed voters, around 56,000 are Black, while 88,000 are White.
+Similarly, 73,000 are Democratic, while only 71,000 are Republican.
+Normalizing by group sizes, the expected harm for a Black Democrat is 9.6%, for a Black Republican is 16.8%, for a White Democrat is 9.7%, and for a White Republican is 5.2%.
+So the challenged plan differentially harms Democratic voters, but does not differentially harm Black Democrats versus White Democrats.
+However, it it especially harms Black Republicans, who have an estimated differential harm of 7.2% versus Black Democrats and of 11.6% versus White Republicans.
+
 # Latent Factor Ecological Inference Model for Boston Primary Elections {#app:lfei}

### 2022-12-06 pres/toy_pres.R

@@ -1,0 +1,206 @@
+# helpers ----
+
+## make grid ----
+bb <- st_sfc(st_polygon(list(rbind(c(0,0), c(1,0), c(1,1), c(0,0)))))
+grid <- st_as_sf(st_make_grid(bb, n = 6, crs = 3857)) %>%
+    rename(geometry=x)
+
+
+minissouri <- grid %>%
+    mutate(row = rep(6:1, each=6),
+           col = rep(1:6, 6),
+           city1 = row <= 2 & col >= 5,
+           city2 = row >= 5 & col <= 2,
+           pop = 4,
+           pop = if_else(city1, 10, pop),
+           pop = if_else(city2, 7, pop),
+           dem = 1,
+           dem = if_else(city1, 8, dem),
+           dem = if_else(city2, 5, dem),
+           rep = 3,
+           rep = if_else(city1, 2, rep),
+           rep = if_else(city2, 2, rep),
+           circle = round((col + row + city1) / 15 * pop)) %>%
+    relocate(geometry, .after=everything()) %>%
+    redist_map(ndists=3, pop_tol=0.1)
+
+
+## people for plotting -------
+
+make_people = function(pop, dem, circle, geometry, row, col, ...) {
+    if (!file.exists(path <- str_glue("data-raw/csq{pop}.txt"))) {
+        url = str_glue("http://hydra.nat.uni-magdeburg.de/packing/csq/txt/csq{pop}.txt")
+        download.file(url, path)
+    }
+    offset = as.numeric(st_centroid(geometry))
+    dem_vec = c(rep(TRUE, dem), rep(FALSE, pop-dem))
+    circ_vec = c(rep(TRUE, circle), rep(FALSE, pop-circle))
+    idx = sample(pop)
+    read_table(path, col_names=F, col_types=cols(.default="d")) %>%
+        select(-X1) %>%
+        as.matrix() %>%
+        apply(1, function(x) st_point(x/6.28 + offset), simplify=FALSE) %>%
+        st_sfc(crs=3857) %>%
+        st_as_sf() %>%
+        rename(geometry=x) %>%
+        mutate(row = row, col = col,
+               dem = dem_vec[idx],
+               circle = circ_vec[idx])
+}
+
+minissouri_people = minissouri %>%
+    pmap_dfr(make_people)
+
[…]

### 2022-12-06 paper/harm.Rmd
readability
typo

@@
 
-In general, $q$ harms some individuals while $\tilde q$, and $\tilde q$ harms a distinct, though not necessarily disjoint, set of individuals.
+In general, $q$ harms some individuals versus $\tilde q$, and $\tilde q$ harms a distinct, though not necessarily disjoint, set of individuals, versus $q$.
 A symmetric version of $h_i(v,q,\tilde q)$ would need to combine both of these groups: those harmed and those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
@@
 #| fig.cap: |
-#|      On thel eft, the expected fraction of voters in each precinct which are
+#|      On the left, the expected fraction of voters in each precinct which are
 #|      harmed by the 2020 enacted congressional districts versus the sampled 

### 2022-12-06 paper/harm.Rmd
template

@@
 output:
-  rticles::arxiv_article:
+  bookdown::pdf_document2:
+    template: "apsr.tex"
     keep_tex: true

### 2022-12-06 paper/harm.Rmd
add rank and addresses

--- 2022-12-06 paper/harm.Rmd
@@
       Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
-    department: Department of Statistics
+    department: Ph.D. Candidate, Department of Statistics
     affiliation: Harvard University
-    location: Cambridge, MA
+    location: 1 Oxford St, Cambridge, MA
     email: cmccartan@fas.harvard.edu
   - name: Christopher T. Kenny
-    department: Department of Government
+    department: Ph.D. Candidate, Department of Government
     affiliation: Harvard University
-    location: Cambridge, MA
+    location: 1737 Cambridge St., Cambridge, MA
     email: christopherkenny@fas.harvard.edu

### 2022-12-06 paper/harm.Rmd
change font size and double space

@@
     latex_engine: pdflatex
+fontsize: 12pt
 editor_options:

### 2022-12-06 paper/harm.Rmd
fix fig ref
anon

@@
 authors:
-  - name: Cory McCartan
-    thanks: |
-      To whom correspondence should be addressed.
-      The authors thank Steve Ansolabehere, Matt Blackwell, Kosuke Imai, Gary King, 
-      Shiro Kuriwaki, Tyler Simko, Nick Stephanopolous, and 
-      participants of the Fall 2021 Harvard American Politics Research Workshop,
-      Fall 2022 Harvard Applied Statistics Workshop, 
-      and Imai Research Group for helpful comments and discussion.
-      Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
-    department: Ph.D. Candidate, Department of Statistics
-    affiliation: Harvard University
-    location: 1 Oxford St, Cambridge, MA
-    email: cmccartan@fas.harvard.edu
-  - name: Christopher T. Kenny
-    department: Ph.D. Candidate, Department of Government
-    affiliation: Harvard University
-    location: 1737 Cambridge St., Cambridge, MA
-    email: christopherkenny@fas.harvard.edu
+  - name: Anonymized Author(s)
+    affiliation: Anonymized Institution
 abstract: |
@@
 
-Figure \@ref(fig:alabama-harm-sample) summarizes the results, which are mostly similar to the results in in Section \ref{sec:alabama}.
+Figure \ref{fig:alabama-harm-sample} summarizes the results, which are mostly similar to the results in in Section \ref{sec:alabama}.
 Across the state, the estimated average harm is 6.9%, which translates to around 144,000 voters.

### 2023-06-02 paper/harm-short.Rmd
version without boston

--- 2022-12-06 paper/harm.Rmd
@@
   We discuss strategies for estimating harm, and demonstrate the utility of our framework 
-  through analyses of partisan gerrymandering in New Jersey, voting rights 
-  litigation in Alabama, and racial dynamics of Boston City Council elections.
+  through analyses of partisan gerrymandering in New Jersey and voting rights 
+  litigation in Alabama.
 keywords:
@@
 Section \ref{sec:alabama}, studies voting rights litigation for Alabama congressional districts, and uses differential harm to separate the racial and partisan effects of the enacted districting plan.
-Section \ref{sec:boston} demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
 Section \ref{sec:discussion} concludes.
@@
 Here, for simplicity and to avoid making further assumptions, we focus on the indicator-based cost function.
-Section \ref{sec:boston} further discusses the implications of this particular specification in the context of primary elections.
 
@@
 Multiparty or primary elections pose additional modeling challenges.
-In Section \ref{sec:boston} we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.
 
@@
 They account for geographic sorting, political cohesion, and the individual preferences of every voter.
-Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.
+Sections \ref{sec:alabama} demonstrates this approach in practice.
 
@@
 
-# Application to Primary Elections in Boston {#sec:boston}
-
-The proposed harm measures are particularly easy to calculate and interpret in two-party elections, since the preferences of each kind of partisan are directly observable as partisan vote shares.
-However, the same framework can be fruitfully applied to more complex settings, providing insights that go beyond those available through traditional methods.
-
-A major question in voting rights law and research is the extent to which voting must be racially polarized before race should enter into the redistricting equation and statutory protections like the VRA should kick in [@elmendorf2016racially].
-This question is particularly acute in cities, where general elections are characterized by Democratic dominance and low levels of racial polarized voting.
-When the locus of political conflict shifts to primary elections, racial patterns are harder to observe and generalize.
-Thus there remains uncertainty about the extent that race is a factor in primary elections, and consequently the role it should play in designing urban legislative maps.
-
-```{r boston-map, fig.width=6.75, fig.height=3.375}
-#| fig.cap: |
-#|      Racial composition of the city of Boston (left)
-#|      and 2010 city council districts (right).
-#|      Neighborhood boundaries are overlaid in black on the
-#|      left map and in white on the right map.
-include_graphics("figures/boston_map.pdf")
-```
-
-The city of Boston is governed by a mayor and a nonpartisan 13-member city council.
-Nine of the councilors are elected from contiguous city council districts, and the remaining four councilors are elected at-large, with all councilors serving two-year terms.
-Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^11]
-This section will study the extent to which the current district boundaries differentially harm voters of different racial groups, in comparison to a completely race-neutral districting process.
-Figure \ref{fig:boston-map} shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.
-
-[^11]: City of Boston charter.
-    Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
-
-## Approach
-
-To apply our proposed framework to study the council district boundaries, we must be able to predict individual voting preferences under counterfactual districting plans.
-This is more difficult than in both New Jersey and Alabama, for two reasons.
-First, the city council is officially nonpartisan, and city politics are dominated by the Democratic Party, so voting choices are more complex than a simple partisan preference.
-To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election.
-Second, as in Alabama, we aim to examine the racial impact of the district system, and so must estimate these voting preferences by race.
-Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.
-
-```{r boston-factors, fig.cap="", fig.width=7, fig.height=3.5}
-#| fig.cap: |
-#|      Candidate factor loadings, with approximate 80\% credible intervals shown as
-#|      ellipses (left), and geographic variation in factor values by race (right).
-#|      Neighborhood boundaries overlaid in black.
-#|      Factor values in the Black second-dimension map are truncated so that variation in the other maps is visible.
-include_graphics("figures/boston_factors.pdf")
-```
-
-To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^12] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
-The turnout component of the model is identical to that used for Alabama.
-The core of the voter choice model is the following specification for $\pi_{jk}$, the preference of precinct $j$ for candidate $k$ on the logit scale:[^13] \begin{align*}
-    \pi_{jk} &= \underbrace{\beta^{(s)}_k}_{\substack{\text{citywide support}\\\text{for candidate}}} +
-    F_{j}^\top(\overbrace{P_j}^{\mathclap{\substack{\text{diagonal matrix of}\\\text{precinct racial proportions}}}}
-    \underbrace{\turn_j}_{\mathclap{\substack{\text{turnout rate parameter}\\\text{from turnout model}}}}
-    )^\top\vb L_k.
-\end{align*}
-
-[^12]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
-
-[^13]: For simplicity of presentation, the parameterization shown here has been altered slightly from that detailed in the appendix, though the ultimate model can be made identical after an adjustment to the prior on $\vb L_k$.
-
-Here, $F_j$ is a matrix of preferences by race and latent dimension within precinct $j$, so $F_{j}^\top(P_j\turn_j)^\top$ is a vector with elements containing the preferences of the precinct's voters in each latent dimension, after marginalizing out race.
-The vector $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $F_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
-
-Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
-As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
-Biden and Wu's opponent, Annissa Essaibi-George, on the right.
-The second dimension appears to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.
-
-Like all ecological inference models, ours relies on several unverifiable assumptions.
-The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
-As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not *strictly* met.
-However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
-For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
-The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
-We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one out of several approaches which could provide estimated vote choices for individuals.
-Local survey data with ideological preferences and racial information would be a sufficient alternative.
-
-## Results
-
-As Figure \ref{fig:boston-factors} shows, primary voting preferences in Boston are polarized by race, especially for the second latent dimension.
-But there is significant variation within racial groups as well, such as the split between more conservative White voters in South Boston and more liberal White voters in Jamaica Plain (the first latent dimension appears to capture a progressive-moderate divide).
-It is certainly possible, then, that the particular way districts are drawn will have an effect on the ability of different racial groups to elect their candidate of choice.
-
-We consider two different scenarios: one in which all of the council races are contested along the first latent dimension, and one in which the races are contested along the second.
-Specifically, we draw voter choices between two candidates from the posterior predictive distribution of our model, setting the candidates' loadings to --1 and 1 on the chosen dimension and 0 on the other dimension.
-We draw the overall level of support from a $\Beta(15, 15)$ distribution, which means that we will study election scenarios where the overall vote share for each bloc of candidates ranges from roughly 30 to 70 percent.
-The posterior predictive draws yield counts of voters for each candidate in each precinct by race, from which we can calculate winners at the district level, under any configuration of districts.
-
-Our hypothetical elections involve only two candidates, to simplify the challenges of measuring harm with multiple candidates, over which voters may have different ranked preferences.
-However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
-One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption in the definition of harm that all candidates the voter did not vote for are interchangeable.
-
-We sample 2,000 districting plans across two independent runs of the SMC algorithm of @mccartan2020, setting constraints to favor compact plans that respect neighborhood boundaries.
-Since we do not provide any racial information to the algorithm, this set of plans forms a good approximation of a race-neutral districting process.
-
-For the first scenario, where council races are contested along the ideological axis, we estimate that the enacted plan is quite fair, differentially harming White voters at a rate of only 0.79% compared to the race-neutral baseline.
-For the second scenario, where the contests are oriented along the racial dimension, we estimate that the enacted plan differentially harms White voters at a rate of 4.1%, compared to the race-neutral baseline.
-
-These two harm estimates reflect the harm of the current boundaries compared to a race-neutral baseline.
-The more policy-relevant direction might be in evaluating a change from the status quo.
-What is the harm of switching to race-neutral districting from the current race-aware boundaries?
-We can flip the above setup and compute harm by averaging over the simulated plans with the enacted plan as the counterfactual.
-For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan Along the second dimension, the differential harm to Black voters would be 1.8%.
-
-Taken together, these findings suggest that despite the dominance of the Democratic party in city politics, and the shared preferences between certain groups of Black and White voters, race-aware districting still serves an important role in empowering Black voters across the city to elect their chosen candidates.
-A switch to race-neutral boundaries would have a material impact on Black voting power.
-Further study could focus on the racial impact of the four at-large members in comparison to a system with 13 districts and no at-large members.
-
 # Discussion {#sec:discussion}
@@
 Analysts can model vote choice separately, estimate the harm to each individual, and then aggregate by any group of interest.
-As Section \ref{sec:nj}--\ref{sec:boston} showed, this opens up space for more nuanced analyses that can extend to the local level as well.
+As Section \ref{sec:nj} and \ref{sec:alabama} showed, this opens up space for more nuanced analyses that can extend to the local level as well.
 
@@
 However, it it especially harms Black Republicans, who have an estimated differential harm of 7.2% versus Black Democrats and of 11.6% versus White Republicans.
-
-# Latent Factor Ecological Inference Model for Boston Primary Elections {#app:lfei}
-
-To estimate voter preferences by race and precinct in Boston, we use 2020 U.S.
-Census demographic data and election returns from the March 2020 presidential primary and the September 2021 mayoral preliminary elections.
-Our ecological inference model used to produce the estimates expands on the model in Appendix \ref{app:alei}, jointly estimating both turnout and vote choice in the two elections.
-However, it estimates voter preferences in a two-dimensional latent space and also estimates candidates' positions in this space.
-
-Continuing the notation from above, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $n_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jk}$ denote the vote share for candidate $k$ (who implicitly belongs to a particular election $v_k$), and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Hispanic},\text{Other}\}$ ("Other" consisting primarily of Asian voters in Boston).
-Our main parameters of interest are $\turn_{jr}$, the baseline turnout by race and precinct, $F_{jqr}$, the preferences in latent dimension $q$ by race and precinct, and $L_{kq}$, the loading of each candidate on latent dimension $q$.
-Here $q\in\{1,2\}$, since preliminary analyses found no improved fit when expanding to three dimensions.
-
-We model the observed turnout $n_{jv}$ identically as in Alabama, above: \begin{align*}
-    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
-        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
-    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
-        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}).
-\end{align*}
-
-Because of the latent factor model, we do not require that vote shares by race and precinct, when tallied by race, necessarily equal the observed vote share in the precinct.
-As a result we choose to model the vote shares directly, rather than the vote counts, which is more computationally efficient.
-Because each precinct has dozens or hundreds of voters, a Normal approximation to the Binomial likelihood is appropriate, and so we model the vote shares $s_{jk}$ as \begin{align*}
-    \logit(s_{jk}) \mid \pi_{jk}, \eps &\sim \Norm(\pi_{jk}, \frac{\eps}{1+n_{jv_k}}),
-\end{align*} where $\pi_{jk}$ is the preference of precinct $j$ for candidate $k$ on the logit scale, as detailed below.
-The Normal approximation variance is inversely proportional to the total number of votes in the precinct (the additional 1 is to avoid divide-by-zero errors).
-The multiplicative factor $\eps$ captures the additional variance that isn't accounted for by the latent factors.
-
-The logit candidate preference itself is modeled as \begin{align*}
-    \pi_{jk} &= \beta^{(s)}_k + \alpha_k
-        F_{j}^\top(\turn_j\circ \vb p_j)^\top\mathrm{diag}(\vb*\rho)\vb L_k  \\
-    F_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
-\end{align*} Note that $F_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $F_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
-The matrix $\mathrm{diag}(\vb*\rho)$ captures the relative scale or importance of each dimension to the overall outcomes.
-The $\vb*\alpha$ parameters are "fudge factors" that adjust the scale of the deviations from each candidate's average support.
-
-We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.5^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 4.0^2)$ prior, and $\logit(\mu^{(t)})$ an i.i.d. $\Beta(10, 30)$ prior.
-We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$.
-The former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
-A larger LKJ shape parameter places more weight on correlation matrices closer to diagonal, i.e., it regularizes away from high correlations.
-The scale parameters $\vb*\rho$ have i.i.d. $\Gam(1.5, 1.5/4.0)$ priors, the fudge factors $\vb*\alpha$ have i.i.d. $\Gam(5.0, 5.0)$ priors, and $\eps$ has an $\Expo(1/4.0)$ prior.
-Overall, these priors weakly identify the location and scale of the parameters.
-Several reparametrizations were made for computational efficiency.
-The full Stan code for the model is available as part of the replication code.
-
-Like all probabilistic factor models, ours is unidentifiable due to the rotation, location, and scale invariance of the factors in the likelihood.
-To identify the location and scale, we place $\Norm(0,1)$ priors on most components of $\vb L$.
-To identify the rotation, we place more informative priors on the loadings of certain candidates.
-This was accomplished through somewhat of an interactive process, where a preliminary principal components decomposition of the precinct vote shares helped suggest the latent factors that might be found by the model.
-We chose to align the first dimension with the Sanders-Biden axis, and chose to make the second dimension roughly orthogonal along the Janey-Buttigieg axis.
-This translated to a prior of $\Norm((-1, 0), (0.1^2, 0.25^2))$ for Sanders, $\Norm((1, 0), (1.0^2, 0.5^2))$ for Biden, and $\Norm((-0.25, -1), (0.5^2, 2^2))$ for Janey.
-We additionally set the prior correlation between Sanders' and Biden's loadings to be $(-0.5, 0.8)$.
-
-The estimated loading matrix $\vb L$ is visualized on the left of Figure \ref{fig:boston-factors}, and $F$ is visualized for Black and White voters on the right.
-
-```{r boston-gen-pred, fig.width=6.5}
-#| fig.cap: |
-#|      Model predictions of general election vote shares and turnout, 
-#|      compared to actual general election results.
-#|      Red reference line indicates perfect agreement.
-include_graphics("figures/boston_gen_pred.pdf")
-```
-
-Like all ecological inference models, this one makes several assumptions, verifiable and unverifiable.
-We assume no spatial correlation in residuals, when some is almost certainly present.
-We place distributional assumptions on how voter preferences by races vary across precincts (importantly, that these preferences are centered around a common location).
-And we of course impose a latent two-dimensional factor structure, which must always be a major simplification of reality.
-
-To test the suitability of the model for representing the outcomes of hypothetical future elections, we use the model fitted to the primary elections to predict the outcome of the mayoral general election in November 2021.
-Michelle Wu and Annissa Essaibi-George advanced to the general from the preliminary election.
-We fix the model's loading matrix $\vb L$ to the posterior mean of their estimated loadings.
-We draw $s_{jk}$ from a $\Beta(20\bar s, 20(1-\bar s))$ distribution, where $\bar s$ is each candidate's share of the two-candidate vote from the preliminary election (59.7% Wu -- 40.3% Essaibi George).
-
-The model's predictions (posterior medians) for the turnout and vote share in each precinct are shown in Figure \ref{fig:boston-gen-pred}.
-Considering the differences between a primary and general election, and the additional campaign time elapsed between the elections, the predictions are remarkably accurate.
-The predicted turnout has a Spearman correlation of 0.84 with the actual turnout, and the predicted and actual vote shares have a correlation 0.87.
-The turnout and support levels are also in relatively close agreement.

### 2023-07-13 paper/harm-short.Rmd
call merrill allen

@@
 As discussed in Section \ref{subsec:racial}, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
-In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure \ref{fig:al-maps}(e).
+In *Allen,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure \ref{fig:al-maps}(e).
 These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure \ref{fig:al-maps}(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
@@
 Posterior medians of $\turn_{jr}$ and $\supp_{jr}$ are plotted in Figure \ref{fig:alabama-ei}.
-These estimates broadly track with the findings of the voting rights expert retained in *Merrill*.
+These estimates broadly track with the findings of the voting rights expert retained in *Allen*.
 

### 2023-07-14 paper/harm-short.Rmd
edits and formatting

@@
       Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
-    department: Ph.D. Candidate, Department of Statistics
-    affiliation: Harvard University
-    location: 1 Oxford St, Cambridge, MA
-    email: cmccartan@fas.harvard.edu
+    department: Faculty Fellow, Center for Data Science
+    affiliation: New York University
+    location: 60 5th Ave, New York, NY
+    email: corymccartan@nyu.edu
   - name: Christopher T. Kenny
@@
 biblio-style: apsr
+fontsize: 12pt
 output:
@@
 ---
+
+\doublespacing
+
+\begin{center}\bfseries 9,970 words\end{center}
 
@@
 Analysts can model vote choice separately, estimate the harm to each individual, and then aggregate by any group of interest.
-As Section \ref{sec:nj} and \ref{sec:alabama} showed, this opens up space for more nuanced analyses that can extend to the local level as well.
 

### 2023-07-14 paper/harm-short.Rmd
title page

@@
 title: Individual and Differential Harm in Redistricting
-authors:
-  - name: Cory McCartan
-    thanks: |
-      To whom correspondence should be addressed.
-      The authors thank Steve Ansolabehere, Matt Blackwell, Kosuke Imai, Gary King, 
-      Shiro Kuriwaki, Tyler Simko, Nick Stephanopolous, and 
-      participants of the Fall 2021 Harvard American Politics Research Workshop,
-      Fall 2022 Harvard Applied Statistics Workshop, 
-      and Imai Research Group for helpful comments and discussion.
-      Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
-    department: Faculty Fellow, Center for Data Science
-    affiliation: New York University
-    location: 60 5th Ave, New York, NY
-    email: corymccartan@nyu.edu
-  - name: Christopher T. Kenny
-    department: Ph.D. Candidate, Department of Government
-    affiliation: Harvard University
-    location: 1737 Cambridge St., Cambridge, MA
-    email: christopherkenny@fas.harvard.edu
+# authors:
+#   - name: Cory McCartan
+#     thanks: |
+#       To whom correspondence should be addressed.
+#       The authors thank Steve Ansolabehere, Matt Blackwell, Kosuke Imai, Gary King, 
+#       Shiro Kuriwaki, Tyler Simko, Nick Stephanopolous, and 
+#       participants of the Fall 2021 Harvard American Politics Research Workshop,
+#       Fall 2022 Harvard Applied Statistics Workshop, 
+#       and Imai Research Group for helpful comments and discussion.
+#       Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
+#     department: Faculty Fellow, Center for Data Science
+#     affiliation: New York University
+#     location: 60 5th Ave, New York, NY
+#     email: corymccartan@nyu.edu
+#   - name: Christopher T. Kenny
+#     department: Ph.D. Candidate, Department of Government
+#     affiliation: Harvard University
+#     location: 1737 Cambridge St., Cambridge, MA
+#     email: christopherkenny@fas.harvard.edu
 abstract: |
@@
 \begin{center}\bfseries 9,970 words\end{center}
+\clearpage
 

### 2023-07-14 paper/harm-short.Rmd
anonymize

--- 2023-07-14 paper/harm-short.Rmd
@@
 authors:
-  - name: Cory McCartan
-    thanks: |
-      To whom correspondence should be addressed.
-      The authors thank Steve Ansolabehere, Matt Blackwell, Kosuke Imai, Gary King, 
-      Shiro Kuriwaki, Tyler Simko, Nick Stephanopolous, and 
-      participants of the Fall 2021 Harvard American Politics Research Workshop,
-      Fall 2022 Harvard Applied Statistics Workshop, 
-      and Imai Research Group for helpful comments and discussion.
-      Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
-    department: Faculty Fellow, Center for Data Science
-    affiliation: New York University
-    location: 60 5th Ave, New York, NY
-    email: corymccartan@nyu.edu
-  - name: Christopher T. Kenny
-    department: Ph.D. Candidate, Department of Government
-    affiliation: Harvard University
-    location: 1737 Cambridge St., Cambridge, MA
-    email: christopherkenny@fas.harvard.edu
+  - name: Anonymous Authors
 abstract: |
@@
 
+\vspace{1in}
 \begin{center}\bfseries 9,970 words\end{center}
+\clearpage
 

### 2024-12-04 paper/harm.qmd
rerender

--- 2022-12-06 paper/harm.Rmd
@@
 title: Individual and Differential Harm in Redistricting
+date: last-modified
 authors:
-  - name: Anonymized Author(s)
-    affiliation: Anonymized Institution
+  - name: Cory McCartan
+    affiliations:
+      - name: Pennsylvania State University
+        department: Department of Statistics
+        address: 461 Pollock Road
+        city: University Park
+        region: PA
+        postal-code: 16801
+    email: mccartan@psu.edu
+    url: https://corymccartan.com/
+    corresponding: true
+  - name: Christopher T. Kenny
+    affiliations:
+    - name: Harvard University
+      department: Department of Government
+      address: 1737 Cambridge Street
+      city: Cambridge
+      region: MA
+      postal-code: "02138 "
+    email: christopherkenny@fas.harvard.edu
+    url: https://christophertkenny.com
 abstract: |
@@
   litigation in Alabama, and racial dynamics of Boston City Council elections.
+thanks: |
+  The authors thank Steve Ansolabehere, Matt Blackwell, Kosuke Imai, Gary King, 
+  Shiro Kuriwaki, Tyler Simko, Nick Stephanopolous, and 
+  participants of the Fall 2021 Harvard American Politics Research Workshop,
+  Fall 2022 Harvard Applied Statistics Workshop, 
+  and Imai Research Group for helpful comments and discussion.
+  Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
 keywords:
@@
 bibliography: references.bib
+format: 
+    cmc-article-pdf:
+        fontsize: 11pt
         biblio-style: apsr
-output:
-  rticles::arxiv_article:
-    keep_tex: true
-    includes:
-      in_header: "header.tex"
-    latex_engine: pdflatex
-fontsize: 12pt
+        include-in-header: "header.tex"
 editor_options:
@@
 
-The harm framework combines two key ideas, which are motivated at greater length in Section \ref{sec:motivation}.
+The harm framework combines two key ideas, which are motivated at greater length in Section \ref{sec-motivation}.
 First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
@@
 And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
-Section \ref{sec:motivation} illustrates this with a simple example, and the general definitions are formalized in Section \ref{sec:indiv-harm}.
+Section \ref{sec-motivation} illustrates this with a simple example, and the general definitions are formalized in Section \ref{sec-indiv-harm}.
 
@@
 This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry].
-The challenge of estimating harm and differential harm from observed data is discussed in Section \ref{sec:estimating}.
+The challenge of estimating harm and differential harm from observed data is discussed in Section \ref{sec-estimating}.
 Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
@@
 Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
-A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.
+A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec-connections}.
 
 Harm and differential harm are more than a useful conceptual unification---they are practically useful in a wide variety of redistricting analyses.
-A brief study of partisan gerrymandering of New Jersey congressional districts in Section \ref{sec:nj} shows how existing gerrymandering metrics can make incorrect conclusions about the magnitude and even the direction of partisan gerrymanders.
-Section \ref{sec:alabama}, studies voting rights litigation for Alabama congressional districts, and uses differential harm to separate the racial and partisan effects of the enacted districting plan.
-Section \ref{sec:boston} demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
-Section \ref{sec:discussion} concludes.
+A brief study of partisan gerrymandering of New Jersey congressional districts in Section \ref{sec-nj} shows how existing gerrymandering metrics can make incorrect conclusions about the magnitude and even the direction of partisan gerrymanders.
+Section \ref{sec-alabama}, studies voting rights litigation for Alabama congressional districts, and uses differential harm to separate the racial and partisan effects of the enacted districting plan.
+Section \ref{sec-boston} demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
+Section \ref{sec-discussion} concludes.
 
-# Motivation {#sec:motivation}
+# Motivation {#sec-motivation}
 
 The core idea motivating individual and differential harm is a simple one.
-Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure \ref{fig:minissouri}.
+Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure \ref{fig-minissouri}.
 The state's total population is 180, and it is divided into three districts with population between 54 and 66.
@@
 
-```{r minissouri, fig.width=6.75, fig.pos='b'}
+```{r fig-minissouri, fig.width=6.75}
 #| fig.cap: |
@@
 
-Now suppose the redistricting plan in Figure \ref{fig:minissouri}(a) has been enacted.
+Now suppose the redistricting plan in Figure \ref{fig-minissouri}(a) has been enacted.
 Despite the statewide popular vote favoring Republicans, under this plan Democrats will always win two of the three seats.
-In contrast, a more proportional plan \ref{fig:minissouri}(b) elects two Republicans and one Democrat.
+In contrast, a more proportional plan \ref{fig-minissouri}(b) elects two Republicans and one Democrat.
 
@@
 To turn this intuition into something more concrete, we can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
-These voters are highlighted in Figure \ref{fig:minissouri}(c).
+These voters are highlighted in Figure \ref{fig-minissouri}(c).
 For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
@@
 
-# Individual and Differential Harm {#sec:indiv-harm}
+# Individual and Differential Harm {#sec-indiv-harm}
 
-We formalize and generalize the concepts in Section \ref{sec:motivation}.
+We formalize and generalize the concepts in Section \ref{sec-motivation}.
 Let $\Voters$ be the set of individuals, or *voters*, in the region and time period.
@@
 For example, different districting plans may lead to different candidates running, depending on district partisanship and incumbent residency.
-The assumptions we make on these relationships will be made more explicit in Section \ref{sec:estimating}, where we discuss how to estimate our measures in practice.
+The assumptions we make on these relationships will be made more explicit in Section \ref{sec-estimating}, where we discuss how to estimate our measures in practice.
 
-## Harm {#subsec:harm}
+## Harm {#subsec-harm}
 
-The key intuition from Section \ref{sec:motivation} is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
+The key intuition from Section \ref{sec-motivation} is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
 Letting $\tilde q$ denote a counterfactual districting plan, we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as
@@
 Here, for simplicity and to avoid making further assumptions, we focus on the indicator-based cost function.
-Section \ref{sec:boston} further discusses the implications of this particular specification in the context of primary elections.
+Section \ref{sec-boston} further discusses the implications of this particular specification in the context of primary elections.
 
@@
 Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
-A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section \ref{sec:estimating}.
+A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section \ref{sec-estimating}.
 
@@
 
-## Differential Harm and Fairness {#subsec:diff-harm}
+## Differential Harm and Fairness {#subsec-diff-harm}
 
@@
 
-# Estimating the Harm Measures {#sec:estimating}
+# Estimating the Harm Measures {#sec-estimating}
 
@@
 Multiparty or primary elections pose additional modeling challenges.
-In Section \ref{sec:boston} we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.
+In Section \ref{sec-boston} we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.
 
@@
 
-# Connections to Current Practice for Analyzing Partisan and Racial Gerrymandering {#sec:connections}
+# Connections to Current Practice for Analyzing Partisan and Racial Gerrymandering {#sec-connections}
 
@@
 
-## Racial Vote Dilution {#subsec:racial}
+## Racial Vote Dilution {#subsec-racial}
 
@@
 They account for geographic sorting, political cohesion, and the individual preferences of every voter.
-Sections \ref{sec:alabama} and \ref{sec:boston} demonstrate this approach in practice.
+Sections \ref{sec-alabama} and \ref{sec-boston} demonstrate this approach in practice.
 
@@
 As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
-This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^4]
+This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec-estimating}.[^4]
 This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
@@
 Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
-Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section \ref{sec:nj}.
+Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section \ref{sec-nj}.
 
@@
 
-# Application to Partisan Gerrymandering in New Jersey {#sec:nj}
+# Application to Partisan Gerrymandering in New Jersey {#sec-nj}
 
@@
 
-```{r nj-maps, fig.width=6.25}
+```{r fig-nj-maps, fig.width=6.25}
 #| fig.cap: |
@@
 
-Figure \ref{fig:nj-maps} shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^8]
+Figure \ref{fig-nj-maps} shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^8]
 Using an average of the 2016 presidential and 2018 senatorial elections as a baseline, the Democratic proposal yields 9 Democratic seats and 3 Republican seats, while the Republican proposal has 7 Democratic and 5 Republican seats.
@@
 We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, as implemented in the R package `redist` [@redist], setting constraints to favor compact plans with district boundaries that tend to follow county lines.
-One such plan is shown in Figure \ref{fig:nj-maps}(d).
+One such plan is shown in Figure \ref{fig-nj-maps}(d).
 For each of these sampled plans and the two gerrymanders, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
-The distribution of these measures is shown in Figure \ref{fig:nj-meas}.
+The distribution of these measures is shown in Figure \ref{fig-nj-meas}.
 
@@
 
-```{r nj-meas, fig.width=7}
+```{r fig-nj-meas, fig.width=7}
 #| fig.cap: |
@@
 
-Figure \ref{fig:nj-pairs} in the appendix expands on Figure \ref{fig:nj-meas} to also show the correlations between these various measures.
+Figure \ref{fig-nj-pairs} in the appendix expands on Figure \ref{fig-nj-meas} to also show the correlations between these various measures.
 Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
@@
 
-# Application to Voting Rights Litigation in Alabama {#sec:alabama}
+# Application to Voting Rights Litigation in Alabama {#sec-alabama}
 
@@
 
-As discussed in Section \ref{subsec:racial}, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
-In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure \ref{fig:al-maps}(e).
-These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure \ref{fig:al-maps}(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
+As discussed in Section \ref{subsec-racial}, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
+In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure \ref{fig-al-maps}(e).
+These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure \ref{fig-al-maps}(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
 In Appendix \ref{app:alsampling}, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.
 
-```{r al-maps, fig.width=6.75}
+```{r fig-al-maps, fig.width=6.75}
 #| fig.cap: |
@@
 With these estimates in hand, we can calculate the average number of voters harmed in each precinct and racial group.
-Figure \ref{fig:al-maps}(c) shows the estimated geographic distribution of harm.
-Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures \ref{fig:al-maps}(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
+Figure \ref{fig-al-maps}(c) shows the estimated geographic distribution of harm.
+Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures \ref{fig-al-maps}(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
 These areas are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.
 
-This intuition is confirmed by a closer analysis, summarized in Figure \ref{fig:al-maps}(f).
+This intuition is confirmed by a closer analysis, summarized in Figure \ref{fig-al-maps}(f).
 Across the state, the estimated average harm is 7.9%, which translates to around 159,000 voters.
@@
 
-# Application to Primary Elections in Boston {#sec:boston}
+# Application to Primary Elections in Boston {#sec-boston}
 
@@
 
-```{r boston-map, fig.width=6.75, fig.height=3.375}
+```{r fig-boston-map, fig.width=6.75, fig.height=3.375}
 #| fig.cap: |
@@
 This section will study the extent to which the current district boundaries differentially harm voters of different racial groups, in comparison to a completely race-neutral districting process.
-Figure \ref{fig:boston-map} shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.
+Figure \ref{fig-boston-map} shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.
 
@@
 
-```{r boston-factors, fig.cap="", fig.width=7, fig.height=3.5}
+```{r fig-boston-factors, fig.cap="", fig.width=7, fig.height=3.5}
 #| fig.cap: |
@@
 
-Figure \ref{fig:boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
+Figure \ref{fig-boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
 As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
 Biden and Wu's opponent, Annissa Essaibi-George, on the right.
-The second dimension appears to capture racial differences in support---as the factor value maps in Figure \ref{fig:boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.
+The second dimension appears to capture racial differences in support---as the factor value maps in Figure \ref{fig-boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.
 
@@
 However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
-For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
+For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig-boston-gen-pred} in Appendix \ref{app:lfei}).
 The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
@@
 
-As Figure \ref{fig:boston-factors} shows, primary voting preferences in Boston are polarized by race, especially for the second latent dimension.
+As Figure \ref{fig-boston-factors} shows, primary voting preferences in Boston are polarized by race, especially for the second latent dimension.
 But there is significant variation within racial groups as well, such as the split between more conservative White voters in South Boston and more liberal White voters in Jamaica Plain (the first latent dimension appears to capture a progressive-moderate divide).
@@
 
-# Discussion {#sec:discussion}
+# Discussion {#sec-discussion}
 
@@
 Analysts can model vote choice separately, estimate the harm to each individual, and then aggregate by any group of interest.
-As Section \ref{sec:nj}--\ref{sec:boston} showed, this opens up space for more nuanced analyses that can extend to the local level as well.
+As Section \ref{sec-nj}--\ref{sec-boston} showed, this opens up space for more nuanced analyses that can extend to the local level as well.
 
@@
 More realistic models require more analyst thought and can add a significant burden to the calculation of our harm measures.
-And as discussed in Section \ref{subsec:harm}, multi-candidate elections where voters have strong ordinal preferences provide further challenges that the proposed framework does not fully address.
+And as discussed in Section \ref{subsec-harm}, multi-candidate elections where voters have strong ordinal preferences provide further challenges that the proposed framework does not fully address.
 Each of these limitations present opportunities for future study.
@@
 
-Figure \ref{fig:nj-pairs} expands on Figure \ref{fig:nj-meas} in the main text with pairwise scatter plots and correlation summaries.
+Figure \ref{fig-nj-pairs} expands on Figure \ref{fig-nj-meas} in the main text with pairwise scatter plots and correlation summaries.
 
-```{r nj-pairs, fig.width=7, fig.height=7}
+```{r fig-nj-pairs, fig.width=7, fig.height=7}
 #| fig.cap: |
@@
 
-Table \ref{tab:corr} below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section \ref{sec:nj} for eight states which cover a wide range of partisan leans and number of seats:
+Table \ref{tab:corr} below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section \ref{sec-nj} for eight states which cover a wide range of partisan leans and number of seats:
 
@@
 The simulations are provided by the 50-State Simulation Project [@mccartan2022fiftystates].
-In particular, the New Jersey simulations are not the same as those used in Section \ref{sec:nj}.
+In particular, the New Jersey simulations are not the same as those used in Section \ref{sec-nj}.
 
@@
 
-```{r alabama-ei, fig.width=6.5}
+```{r fig-alabama-ei, fig.width=6.5}
 #| fig.cap: |
@@
 The model was fitted with a mean-field variational approximation for computational efficiency.
-Posterior medians of $\turn_{jr}$ and $\supp_{jr}$ are plotted in Figure \ref{fig:alabama-ei}.
+Posterior medians of $\turn_{jr}$ and $\supp_{jr}$ are plotted in Figure \ref{fig-alabama-ei}.
 These estimates broadly track with the findings of the voting rights expert retained in *Merrill*.
@@
 
-Here, we perform the same analysis as in Section \ref{sec:alabama}, but using a randomly sampled set of plans as a counterfactual, rather than the four plans used in litigation.
+Here, we perform the same analysis as in Section \ref{sec-alabama}, but using a randomly sampled set of plans as a counterfactual, rather than the four plans used in litigation.
 Specifically, we use 2,000 plans provided by the 50-State Simulation Project [@mccartan2022fiftystates].
@@
 
-```{r alabama-harm-sample, fig.width=6.5}
+```{r fig-alabama-harm-sample, fig.width=6.5}
 #| fig.cap: |
@@
 
-Figure \ref{fig:alabama-harm-sample} summarizes the results, which are mostly similar to the results in in Section \ref{sec:alabama}.
+Figure \ref{fig-alabama-harm-sample} summarizes the results, which are mostly similar to the results in in Section \ref{sec-alabama}.
 Across the state, the estimated average harm is 6.9%, which translates to around 144,000 voters.
@@
 
-The estimated loading matrix $\vb L$ is visualized on the left of Figure \ref{fig:boston-factors}, and $F$ is visualized for Black and White voters on the right.
+The estimated loading matrix $\vb L$ is visualized on the left of Figure \ref{fig-boston-factors}, and $F$ is visualized for Black and White voters on the right.
 
-```{r boston-gen-pred, fig.width=6.5}
+```{r fig-boston-gen-pred, fig.width=6.5}
 #| fig.cap: |
@@
 
-The model's predictions (posterior medians) for the turnout and vote share in each precinct are shown in Figure \ref{fig:boston-gen-pred}.
+The model's predictions (posterior medians) for the turnout and vote share in each precinct are shown in Figure \ref{fig-boston-gen-pred}.
 Considering the differences between a primary and general election, and the additional campaign time elapsed between the elections, the predictions are remarkably accurate.

### 2024-12-17 paper/harm.qmd
submission

@@
     cmc-article-pdf:
+        journal:
+            blinded: true
         fontsize: 11pt
@@
 
+# Data availability {.unnumbered}
+All data are publicly available and are included in the replication package for the manuscript.
+
+# Code availability {.unnumbered}
+All code is included in the replication package for the manuscript.
+
 # References {.unnumbered}

### 2025-04-28 paper/harm-short.Rmd

@@
 biblio-style: apsr
+citeproc: false
 fontsize: 12pt

### 2025-04-28 paper/harm.qmd
use current crossref syntax

@@
 
-The harm framework combines two key ideas, which are motivated at greater length in Section \ref{sec-motivation}.
+The harm framework combines two key ideas, which are motivated at greater length in Section @sec-motivation.
 First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
@@
 And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
-Section \ref{sec-motivation} illustrates this with a simple example, and the general definitions are formalized in Section \ref{sec-indiv-harm}.
+Section @sec-motivation illustrates this with a simple example, and the general definitions are formalized in Section @sec-indiv-harm.
 
@@
 This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry].
-The challenge of estimating harm and differential harm from observed data is discussed in Section \ref{sec-estimating}.
+The challenge of estimating harm and differential harm from observed data is discussed in Section @sec-estimating.
 Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
@@
 Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
-A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec-connections}.
+A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section @sec-connections.
 
 Harm and differential harm are more than a useful conceptual unification---they are practically useful in a wide variety of redistricting analyses.
-A brief study of partisan gerrymandering of New Jersey congressional districts in Section \ref{sec-nj} shows how existing gerrymandering metrics can make incorrect conclusions about the magnitude and even the direction of partisan gerrymanders.
-Section \ref{sec-alabama}, studies voting rights litigation for Alabama congressional districts, and uses differential harm to separate the racial and partisan effects of the enacted districting plan.
-Section \ref{sec-boston} demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
-Section \ref{sec-discussion} concludes.
+A brief study of partisan gerrymandering of New Jersey congressional districts in Section @sec-nj shows how existing gerrymandering metrics can make incorrect conclusions about the magnitude and even the direction of partisan gerrymanders.
+Section @sec-alabama, studies voting rights litigation for Alabama congressional districts, and uses differential harm to separate the racial and partisan effects of the enacted districting plan.
+Section @sec-boston demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
+Section @sec-discussion concludes.
 
@@
 The core idea motivating individual and differential harm is a simple one.
-Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure \ref{fig-minissouri}.
+Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure @fig-minissouri.
 The state's total population is 180, and it is divided into three districts with population between 54 and 66.
@@
 
-Now suppose the redistricting plan in Figure \ref{fig-minissouri}(a) has been enacted.
+Now suppose the redistricting plan in Figure @fig-minissouri(a) has been enacted.
 Despite the statewide popular vote favoring Republicans, under this plan Democrats will always win two of the three seats.
-In contrast, a more proportional plan \ref{fig-minissouri}(b) elects two Republicans and one Democrat.
+In contrast, a more proportional plan @fig-minissouri(b) elects two Republicans and one Democrat.
 
@@
 To turn this intuition into something more concrete, we can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
-These voters are highlighted in Figure \ref{fig-minissouri}(c).
+These voters are highlighted in Figure @fig-minissouri(c).
 For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
@@
 
-We formalize and generalize the concepts in Section \ref{sec-motivation}.
+We formalize and generalize the concepts in Section @sec-motivation.
 Let $\Voters$ be the set of individuals, or *voters*, in the region and time period.
@@
 For example, different districting plans may lead to different candidates running, depending on district partisanship and incumbent residency.
-The assumptions we make on these relationships will be made more explicit in Section \ref{sec-estimating}, where we discuss how to estimate our measures in practice.
+The assumptions we make on these relationships will be made more explicit in Section @sec-estimating, where we discuss how to estimate our measures in practice.
 
-## Harm {#subsec-harm}
+## Harm {#sec-harm}
 
-The key intuition from Section \ref{sec-motivation} is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
+The key intuition from Section @sec-motivation is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
 Letting $\tilde q$ denote a counterfactual districting plan, we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as
@@
 Here, for simplicity and to avoid making further assumptions, we focus on the indicator-based cost function.
-Section \ref{sec-boston} further discusses the implications of this particular specification in the context of primary elections.
+Section @sec-boston further discusses the implications of this particular specification in the context of primary elections.
 
@@
 Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
-A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section \ref{sec-estimating}.
+A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section @sec-estimating.
 
@@
 
-## Differential Harm and Fairness {#subsec-diff-harm}
+## Differential Harm and Fairness {#sec-diff-harm}
 
@@
 Multiparty or primary elections pose additional modeling challenges.
-In Section \ref{sec-boston} we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.
+In Section @sec-boston we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.
 
@@
 We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
-Appendix \ref{app:elecmodel} contains more details on how we estimate and apply this model.
+Appendix @app:elecmodel contains more details on how we estimate and apply this model.
 
@@
 
-## Racial Vote Dilution {#subsec-racial}
+## Racial Vote Dilution {#sec-racial}
 
@@
 They account for geographic sorting, political cohesion, and the individual preferences of every voter.
-Sections \ref{sec-alabama} and \ref{sec-boston} demonstrate this approach in practice.
+Sections @sec-alabama and @sec-boston demonstrate this approach in practice.
 
@@
 As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
-This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec-estimating}.[^4]
+This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section @sec-estimating.[^4]
 This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
@@
 Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
-Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section \ref{sec-nj}.
+Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section @sec-nj.
 
@@
 
-Figure \ref{fig-nj-maps} shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^8]
+Figure @fig-nj-maps shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^8]
 Using an average of the 2016 presidential and 2018 senatorial elections as a baseline, the Democratic proposal yields 9 Democratic seats and 3 Republican seats, while the Republican proposal has 7 Democratic and 5 Republican seats.
@@
 We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, as implemented in the R package `redist` [@redist], setting constraints to favor compact plans with district boundaries that tend to follow county lines.
-One such plan is shown in Figure \ref{fig-nj-maps}(d).
+One such plan is shown in Figure @fig-nj-maps(d).
 For each of these sampled plans and the two gerrymanders, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
-The distribution of these measures is shown in Figure \ref{fig-nj-meas}.
+The distribution of these measures is shown in Figure @fig-nj-meas.
 
@@
 
-Figure \ref{fig-nj-pairs} in the appendix expands on Figure \ref{fig-nj-meas} to also show the correlations between these various measures.
+Figure @fig-nj-pairs in the appendix expands on Figure @fig-nj-meas to also show the correlations between these various measures.
 Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
@@
 New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
-Appendix \ref{app:corr} records these correlations across simulated plans in eight states.
+Appendix @app:corr records these correlations across simulated plans in eight states.
 The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
@@
 
-As discussed in Section \ref{subsec-racial}, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
-In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure \ref{fig-al-maps}(e).
-These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure \ref{fig-al-maps}(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
-In Appendix \ref{app:alsampling}, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.
+As discussed in Section @sec-racial, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
+In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure @fig-al-maps(e).
+These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure @fig-al-maps(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
+In Appendix @app:alsampling, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.
 
@@
 As is custom in VRA analyses, we use an ecological inference model.
-The model, detailed in Appendix \ref{app:alei}, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
+The model, detailed in Appendix @app:alei, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
 For example, we model precinct $j$'s observed turnout $n_{jv}$ in election $v$ and baseline turnout rates $\turn_{jr}$ by race $r$ as \begin{align*}
@@
 With these estimates in hand, we can calculate the average number of voters harmed in each precinct and racial group.
-Figure \ref{fig-al-maps}(c) shows the estimated geographic distribution of harm.
-Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures \ref{fig-al-maps}(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
+Figure @fig-al-maps(c) shows the estimated geographic distribution of harm.
+Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures @fig-al-maps(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
 These areas are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.
 
-This intuition is confirmed by a closer analysis, summarized in Figure \ref{fig-al-maps}(f).
+This intuition is confirmed by a closer analysis, summarized in Figure @fig-al-maps(f).
 Across the state, the estimated average harm is 7.9%, which translates to around 159,000 voters.
@@
 This section will study the extent to which the current district boundaries differentially harm voters of different racial groups, in comparison to a completely race-neutral districting process.
-Figure \ref{fig-boston-map} shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.
+Figure @fig-boston-map shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.
 
@@
 
-Figure \ref{fig-boston-factors} shows the key outputs from this model, and Appendix \ref{app:lfei} explains the model in more detail.
+Figure @fig-boston-factors shows the key outputs from this model, and Appendix @app:lfei explains the model in more detail.
 As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
 Biden and Wu's opponent, Annissa Essaibi-George, on the right.
-The second dimension appears to capture racial differences in support---as the factor value maps in Figure \ref{fig-boston-factors} show, minority-majority and White areas of the city take opposite values on the second dimension.
+The second dimension appears to capture racial differences in support---as the factor value maps in Figure @fig-boston-factors show, minority-majority and White areas of the city take opposite values on the second dimension.
 
@@
 The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
-As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not *strictly* met.
+As Appendix @app:lfei discusses, these assumptions are almost certainly not *strictly* met.
 However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
-For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig-boston-gen-pred} in Appendix \ref{app:lfei}).
+For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure @fig-boston-gen-pred in Appendix @app:lfei).
 The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
@@
 
-As Figure \ref{fig-boston-factors} shows, primary voting preferences in Boston are polarized by race, especially for the second latent dimension.
+As Figure @fig-boston-factors shows, primary voting preferences in Boston are polarized by race, especially for the second latent dimension.
 But there is significant variation within racial groups as well, such as the split between more conservative White voters in South Boston and more liberal White voters in Jamaica Plain (the first latent dimension appears to capture a progressive-moderate divide).
@@
 Analysts can model vote choice separately, estimate the harm to each individual, and then aggregate by any group of interest.
-As Section \ref{sec-nj}--\ref{sec-boston} showed, this opens up space for more nuanced analyses that can extend to the local level as well.
+As Section @sec-nj--@sec-boston showed, this opens up space for more nuanced analyses that can extend to the local level as well.
 
@@
 More realistic models require more analyst thought and can add a significant burden to the calculation of our harm measures.
-And as discussed in Section \ref{subsec-harm}, multi-candidate elections where voters have strong ordinal preferences provide further challenges that the proposed framework does not fully address.
+And as discussed in Section @sec-harm, multi-candidate elections where voters have strong ordinal preferences provide further challenges that the proposed framework does not fully address.
 Each of these limitations present opportunities for future study.
@@
 All code is included in the replication package for the manuscript.
-
-# References {.unnumbered}
 
@@
 
-Figure \ref{fig-nj-pairs} expands on Figure \ref{fig-nj-meas} in the main text with pairwise scatter plots and correlation summaries.
+Figure @fig-nj-pairs expands on Figure @fig-nj-meas in the main text with pairwise scatter plots and correlation summaries.
 
@@
 
-Table \ref{tab:corr} below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section \ref{sec-nj} for eight states which cover a wide range of partisan leans and number of seats:
+Table @tab:corr below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section @sec-nj for eight states which cover a wide range of partisan leans and number of seats:
 
@@
 The simulations are provided by the 50-State Simulation Project [@mccartan2022fiftystates].
-In particular, the New Jersey simulations are not the same as those used in Section \ref{sec-nj}.
+In particular, the New Jersey simulations are not the same as those used in Section @sec-nj.
 
@@
 The model was fitted with a mean-field variational approximation for computational efficiency.
-Posterior medians of $\turn_{jr}$ and $\supp_{jr}$ are plotted in Figure \ref{fig-alabama-ei}.
+Posterior medians of $\turn_{jr}$ and $\supp_{jr}$ are plotted in Figure @fig-alabama-ei.
 These estimates broadly track with the findings of the voting rights expert retained in *Merrill*.
@@
 
-Here, we perform the same analysis as in Section \ref{sec-alabama}, but using a randomly sampled set of plans as a counterfactual, rather than the four plans used in litigation.
+Here, we perform the same analysis as in Section @sec-alabama, but using a randomly sampled set of plans as a counterfactual, rather than the four plans used in litigation.
 Specifically, we use 2,000 plans provided by the 50-State Simulation Project [@mccartan2022fiftystates].
@@
 
-Figure \ref{fig-alabama-harm-sample} summarizes the results, which are mostly similar to the results in in Section \ref{sec-alabama}.
+Figure @fig-alabama-harm-sample summarizes the results, which are mostly similar to the results in in Section @sec-alabama.
 Across the state, the estimated average harm is 6.9%, which translates to around 144,000 voters.
@@
 Census demographic data and election returns from the March 2020 presidential primary and the September 2021 mayoral preliminary elections.
-Our ecological inference model used to produce the estimates expands on the model in Appendix \ref{app:alei}, jointly estimating both turnout and vote choice in the two elections.
+Our ecological inference model used to produce the estimates expands on the model in Appendix @app:alei, jointly estimating both turnout and vote choice in the two elections.
 However, it estimates voter preferences in a two-dimensional latent space and also estimates candidates' positions in this space.
@@
 
-The estimated loading matrix $\vb L$ is visualized on the left of Figure \ref{fig-boston-factors}, and $F$ is visualized for Black and White voters on the right.
+The estimated loading matrix $\vb L$ is visualized on the left of Figure @fig-boston-factors, and $F$ is visualized for Black and White voters on the right.
 
@@
 
-The model's predictions (posterior medians) for the turnout and vote share in each precinct are shown in Figure \ref{fig-boston-gen-pred}.
+The model's predictions (posterior medians) for the turnout and vote share in each precinct are shown in Figure @fig-boston-gen-pred.
 Considering the differences between a primary and general election, and the additional campaign time elapsed between the elections, the predictions are remarkably accurate.

### 2025-04-28 paper/harm.qmd
fix app refs too

@@
 We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
-Appendix @app:elecmodel contains more details on how we estimate and apply this model.
+Appendix @sec-app-elecmodel contains more details on how we estimate and apply this model.
 
@@
 New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
-Appendix @app:corr records these correlations across simulated plans in eight states.
+Appendix @sec-app-corr records these correlations across simulated plans in eight states.
 The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
@@
 These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure @fig-al-maps(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
-In Appendix @app:alsampling, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.
+In Appendix @sec-app-alsampling, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.
 
@@
 As is custom in VRA analyses, we use an ecological inference model.
-The model, detailed in Appendix @app:alei, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
+The model, detailed in Appendix @sec-app-alei, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
 For example, we model precinct $j$'s observed turnout $n_{jv}$ in election $v$ and baseline turnout rates $\turn_{jr}$ by race $r$ as \begin{align*}
@@
 
-Figure @fig-boston-factors shows the key outputs from this model, and Appendix @app:lfei explains the model in more detail.
+Figure @fig-boston-factors shows the key outputs from this model, and Appendix @sec-app-lfei explains the model in more detail.
 As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
@@
 The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
-As Appendix @app:lfei discusses, these assumptions are almost certainly not *strictly* met.
+As Appendix @sec-app-lfei discusses, these assumptions are almost certainly not *strictly* met.
 However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
-For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure @fig-boston-gen-pred in Appendix @app:lfei).
+For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure @fig-boston-gen-pred in Appendix @sec-app-lfei).
 The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
@@
 ```
-# Two-party Election Model {#app:elecmodel}
+# Two-party Election Model {#sec-app-elecmodel}
 
@@
 
-# Correlation Between Partisan Gerrymandering Metrics Across Multiple States {#app:corr}
+# Correlation Between Partisan Gerrymandering Metrics Across Multiple States {#sec-app-corr}
 
@@
 ```
-# Ecological Inference Model for Alabama {#app:alei}
+# Ecological Inference Model for Alabama {#sec-app-alei}
 
@@
 
-# Additional Sampling Analysis for Alabama {#app:alsampling}
+# Additional Sampling Analysis for Alabama {#sec-app-alsampling}
 
@@
 
-# Latent Factor Ecological Inference Model for Boston Primary Elections {#app:lfei}
+# Latent Factor Ecological Inference Model for Boston Primary Elections {#sec-app-lfei}
 
@@
 Census demographic data and election returns from the March 2020 presidential primary and the September 2021 mayoral preliminary elections.
-Our ecological inference model used to produce the estimates expands on the model in Appendix @app:alei, jointly estimating both turnout and vote choice in the two elections.
+Our ecological inference model used to produce the estimates expands on the model in Appendix @sec-app-alei, jointly estimating both turnout and vote choice in the two elections.
 However, it estimates voter preferences in a two-dimensional latent space and also estimates candidates' positions in this space.

### 2025-07-25 paper/harm.qmd
move §4 to app

@@
 
-With these definitions, the harm for a (randomly selected) individual under plan $q$ can be written as $$
+With these definitions, the harm for a (randomly selected) individual under plan $q$ can be written as 
+$$
 H(q) \dfeq h_I(V, q, \tilde Q).
-$$ $H(q)$ is itself a random variable, representing the harm to *some* individual in *some* election compared to *some* counterfactual plan.
+$$ 
+$H(q)$ is itself a random variable, representing the harm to *some* individual in *some* election compared to *some* counterfactual plan.
 The random variable $I$ here serves as a kind of "veil of ignorance" [@rawls2004theory]: without picking a specific individual, a districting plan $q$ which tends to produce a low $H(q)$ would be expected to be preferred over a plan which tended to produce a higher $H(q)$.
@@
 We can zoom in to specific dimensions of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
-Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$, and the electorate's average harm in a particular election $v$ becomes $$
+Thus, for example, the harm for a single individual $i$ becomes $H(q)\mid I=i$, and the electorate's average harm in a particular election $v$ becomes 
+$$
     \E[H(q)\mid V=v] = \Pr[v(I)\neq\win_I(v, q) \text{ and }v(I)=\win_I(v, \tilde Q)],
-$$ that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
+$$ 
+that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
 
@@
 Fairness is a strict condition: real-life plans are highly unlikely to be exactly fair.
-To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
+To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as 
+$$
   \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
-$$ for two groups $G_k$ and $G_l$.
+$$ 
+for two groups $G_k$ and $G_l$.
 So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
@@
 
-Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe next.
-
-# Estimating the Harm Measures {#sec-estimating}
-
-Harm $H(q)$ and differential harm $\Delta H(q)$ as defined above require specification of a probability distribution on elections and districting plans.
-Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must estimate.
-Either way, past election data and an enacted districting plan alone are not enough to calculate these measures.
-
-## Modeling Districting Plans
-
-It is difficult to construct a tractable probability distribution for districting plans, since the space of plans is discrete and lacks a natural metric structure.
-Thankfully, in recent years a number of algorithms have been developed which sample districting plans from specific distributions [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
-These distributions can generally be characterized as maximum-entropy distributions on the space of all valid districting plans, in combination with several moment constraints.
-
-For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans that are geographically contiguous, have district populations within a specified deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
-These maximum-entropy distributions put as few conditions as possible on the districting sampling process, as long as a set of constraints imposed by the researcher are met.
-If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan or other considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
-In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined.
-There is certainly an element of subjectivity in sampling a set of districting plans.
-Nonetheless, sampling methods are widely used in practice [see e.g., @2019rucho; @2018league; @2022merrill; @becker2021computational; @chen:rodd:13].
-
-## Modeling Election Outcomes
-
-Election results present a different set of modeling challenges, given the large number of factors that influence election outcomes and the general unpredictability of future events.
-We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting.
-Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
-Multiparty or primary elections pose additional modeling challenges.
-In Section @sec-boston we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.
-
-Motivated primarily by computational convenience and simplicity, we adopt the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as $$
-    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
-$$ where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
-The baseline $\bar{s}_d$ can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
-Since $\bar{s}_d$ is calculated for the particular configuration of district $d$, this model takes into account the specific districting plan under consideration.
-We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
-Appendix @sec-app-elecmodel contains more details on how we estimate and apply this model.
-
-More complicated election outcome models are certainly possible and are straightforward to use with the measures proposed here.
-For example, one could operate on the logit scale, model incumbency effects, or include additional data from local elections.
-
-## Calculating Harm and Differential Harm
-
-With a model for election outcomes and districting plans in hand, calculation of $H(q)$ and $\Delta H(q)$ is straightforward.
-For each sampled counterfactual districting plan $\tilde q$, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
-We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^2]
-We tally the total harm done to Democratic and Republican voters across the map and divide by the appropriate totals to get a single draw of $H(q)\mid I\in\dem$ and $H(q)\mid I\in\rep$.
-When other groups (such as racial groups) are used in $\Delta H(q)$, we must repeat these computations for each group.
-Finally, we average across all of the samples, and add or subtract across groups, to produce estimates of $E[H(q)]$ and $\Delta_{\dem-\rep} H(q)$.
-
-[^2]: For this, we use the baseline count of Democratic or Republican votes.
-    This is a slight simplification, since the vote share changes across random draws, and so the number of Democratic and Republican voters changes too.
-    To deal with this issue completely, one must model elections at the precinct level, which requires more data and computational costs.
+Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe in @sec-methods.
 
@@
 
+# Methods {#sec-methods}
+
+This section discusses estimating the proposed harm measures from real data, and provides details on the analyses conducted for the three applications in the main text.
+
+## Estimating the Harm Measures {#sec-estimating}
+
+Harm $H(q)$ and differential harm $\Delta H(q)$ as defined above require specification of a probability distribution on elections and districting plans.
+Depending on one's viewpoint, this is either a theoretical fiction which must be specified by a model, or an actual data-generating process we must estimate.
+Either way, past election data and an enacted districting plan alone are not enough to calculate these measures.
+
+### Modeling Districting Plans
+
+It is difficult to construct a tractable probability distribution for districting plans, since the space of plans is discrete and lacks a natural metric structure.
+Thankfully, in recent years a number of algorithms have been developed which sample districting plans from specific distributions [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
+These distributions can generally be characterized as maximum-entropy distributions on the space of all valid districting plans, in combination with several moment constraints.
+
+For example, the algorithms of @carter2019 and @mccartan2020 sample from the maximum-entropy distribution of districting plans that are geographically contiguous, have district populations within a specified deviation from equality, and which have a certain average compactness value (according to a particular graph-theoretic compactness measure).
+These maximum-entropy distributions put as few conditions as possible on the districting sampling process, as long as a set of constraints imposed by the researcher are met.
+If these constraints are precisely those which are required by law or traditional redistricting criteria, then samples from this distribution provide a natural set of counterfactual districting plans, uninfluenced by superfluous partisan or other considerations, to use in estimating $H(q)$ and $\Delta H(q)$.
+In practice, the translation from tradition and law into mathematical formulations of constraints is not always clear or even well-defined.
+There is certainly an element of subjectivity in sampling a set of districting plans.
+Nonetheless, sampling methods are widely used in practice [see e.g., @2019rucho; @2018league; @2022merrill; @becker2021computational; @chen:rodd:13].
+
+### Modeling Election Outcomes
+
+Election results present a different set of modeling challenges, given the large number of factors that influence election outcomes and the general unpredictability of future events.
+We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting.
+Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
+Multiparty or primary elections pose additional modeling challenges.
+In Section @sec-boston we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.
+
+Motivated primarily by computational convenience and simplicity, we adopt the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as
+$$
+    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
+$$ 
+where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
+The baseline $\bar{s}_d$ can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
+Since $\bar{s}_d$ is calculated for the particular configuration of district $d$, this model takes into account the specific districting plan under consideration.
+We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
+Appendix @sec-app-elecmodel contains more details on how we estimate and apply this model.
+
+More complicated election outcome models are certainly possible and are straightforward to use with the measures proposed here.
+For example, one could operate on the logit scale, model incumbency effects, or include additional data from local elections.
+
+### Calculating Harm and Differential Harm
+
+With a model for election outcomes and districting plans in hand, calculation of $H(q)$ and $\Delta H(q)$ is straightforward.
+For each sampled counterfactual districting plan $\tilde q$, we use the election model to simulate vote shares in each district of $q$ and $\tilde q$.
+We can then identify, for each precinct, how many Democratic or Republican voters were harmed.[^2]
+We tally the total harm done to Democratic and Republican voters across the map and divide by the appropriate totals to get a single draw of $H(q)\mid I\in\dem$ and $H(q)\mid I\in\rep$.
+When other groups (such as racial groups) are used in $\Delta H(q)$, we must repeat these computations for each group.
+Finally, we average across all of the samples, and add or subtract across groups, to produce estimates of $E[H(q)]$ and $\Delta_{\dem-\rep} H(q)$.
+
+[^2]: For this, we use the baseline count of Democratic or Republican votes.
+    This is a slight simplification, since the vote share changes across random draws, and so the number of Democratic and Republican voters changes too.
+    To deal with this issue completely, one must model elections at the precinct level, which requires more data and computational costs.
+
 # Data availability {.unnumbered}
@@
 
-The election model used in estimating harm is $$
+The election model used in estimating harm is 
+$$
     s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
@@
     \quad \varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon).
-$$ As noted in the main text, $\bar{s}_d$ is computed by averaging historical election returns.
+$$ 
+As noted in the main text, $\bar{s}_d$ is computed by averaging historical election returns.
 This is done at the precinct level, since this is the most granular level that districting plans are usually built from and election returns are reported at.[^14]
@@
 
-We model the observed turnout $n_{jv}$ and vote counts $s_{jk}n_{jv}$ as \begin{align*}
+We model the observed turnout $n_{jv}$ and vote counts $s_{jk}n_{jv}$ as 
+\begin{align*}
     n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
@@
         &\sim \Norm(\mu^{(s)}, \Sigma^{(t)}),
-\end{align*} where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale, and $\circ$ indicates the element-wise (Hadamard) product.
+\end{align*} 
+where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale, and $\circ$ indicates the element-wise (Hadamard) product.
 We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior.
@@
 
-We model the observed turnout $n_{jv}$ identically as in Alabama, above: \begin{align*}
+We model the observed turnout $n_{jv}$ identically as in Alabama, above: 
+\begin{align*}
     n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
@@
 As a result we choose to model the vote shares directly, rather than the vote counts, which is more computationally efficient.
-Because each precinct has dozens or hundreds of voters, a Normal approximation to the Binomial likelihood is appropriate, and so we model the vote shares $s_{jk}$ as \begin{align*}
+Because each precinct has dozens or hundreds of voters, a Normal approximation to the Binomial likelihood is appropriate, and so we model the vote shares $s_{jk}$ as 
+\begin{align*}
     \logit(s_{jk}) \mid \pi_{jk}, \eps &\sim \Norm(\pi_{jk}, \frac{\eps}{1+n_{jv_k}}),
-\end{align*} where $\pi_{jk}$ is the preference of precinct $j$ for candidate $k$ on the logit scale, as detailed below.
+\end{align*} 
+where $\pi_{jk}$ is the preference of precinct $j$ for candidate $k$ on the logit scale, as detailed below.
 The Normal approximation variance is inversely proportional to the total number of votes in the precinct (the additional 1 is to avoid divide-by-zero errors).
@@
 
-The logit candidate preference itself is modeled as \begin{align*}
+The logit candidate preference itself is modeled as 
+\begin{align*}
     \pi_{jk} &= \beta^{(s)}_k + \alpha_k
@@
     F_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
-\end{align*} Note that $F_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $F_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
+\end{align*} 
+Note that $F_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $F_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
 The matrix $\mathrm{diag}(\vb*\rho)$ captures the relative scale or importance of each dimension to the overall outcomes.

### 2025-07-29 paper/harm.qmd
trimming

@@
 For as long as there have been district-based electoral systems, groups, individuals, and parties have manipulated them for political gain.
-Partisan gerrymandering, whether in the form of rotten boroughs or Elbridge Gerry's salamander-shaped district [@griffith1907rise, pp.
-72--73], is perhaps the most consistently visible manifestation of this tendency.
+Partisan gerrymandering, going back to Elbridge Gerry's salamander-shaped district [@griffith1907rise, pp. 72--73], is perhaps the most consistently visible manifestation of this tendency.
 But gerrymandering takes other forms as well.
-In the United States, many states have historically used racial gerrymandering to maintain White political supremacy [@gilliland1969racial].
-At the local level, where most redistricting occurs, partisan dynamics are not always as salient, and the manipulation of the districting plan may be more subtle or may favor a narrower or less well-defined group [@cain2002mapmaking; @halberstam2012process].
+In the United States, states historically used racial gerrymandering to maintain White political supremacy [@gilliland1969racial].
+At the local level, where most redistricting occurs, partisanship is less salient, and manipulation may be more subtle or may favor a narrower group [@cain2002mapmaking; @halberstam2012process].
 The diversity of types of gerrymandering is further apparent from a survey of major litigation over the past decades, which includes allegations of gerrymandering involving party [@2018league; @2019rucho], race [@1993shaw; @2022merrill], religion [@1977ujo; @2020ercsd], incarceration status [@2015calvin; @2016davidson], and residency or citizenship [@1966burns; @2016evenwel].
 
-The academic study of gerrymandering has unfortunately not reflected this diversity.
-The vast majority of research attention has been on partisan gerrymandering, and has particularly centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
-A large number of formulas for quantifying or identifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18].
-The study of racial gerrymandering has been largely shaped by the U.S.
-legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA).
+The academic study of gerrymandering has not reflected this diversity.
+Nearly all research has been on partisan gerrymandering, and has centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
+A large number of formulas for quantifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18].
+The study of racial gerrymandering has been largely shaped by the legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA).
 This literature is characterized by a focus on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
@@
 
-The dominance of the partisan framework for analyzing redistricting poses a problem even in cases of pure partisan gerrymandering.
-Existing tools, by focusing only on the effect of district lines on parties, are unable to examine the effect of manipulated lines on other groups.
-A redistricting plan that diminishes Democratic voting power will naturally have an effect on the relative voting power of White and Black voters, poor and wealthy voters, young and old voters, and urban or rural voters.
-These effects on individuals will be missed by any framework that only studies parties.
+The dominance of partisan frameworks limits redistricting analysis, even in purely partisan cases.
+By focusing only on party effects, existing tools overlook how partisan lines also have an effect of relative voting power by race, wealth, age, and urbanicity.
 
-This paper addresses these limitations with a new framework of *individual and differential harm* for the unified study of all forms of gerrymandering.
-Fundamentally, gerrymandering involves the concentration or dilution of voting power to favor or disfavor a particular group.
-Our proposed framework focuses directly on measuring this concentration and dilution, and in doing so avoids the pitfalls of party-centric approaches.
+This paper addresses these limitations with a new framework of *individual and differential harm*.
+Our proposed framework focuses on measuring the concentration or dilution of voting power to favor or disfavor a particular group.
+This avoids the pitfalls of party-centric approaches.
 
-The harm framework combines two key ideas, which are motivated at greater length in Section @sec-motivation.
+The harm framework combines two key ideas, further motivated in Section @sec-motivation.
 First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
-Doing so both unifies the approach to gerrymandering across contexts and also reflects a core goal of representative democracy: that elected officials represent the preferences of the voting public to the greatest extent possible [see also @brunell2010].
+This unifies gerrymandering analyses across contexts and reflects the goal that elected officials represent the preferences of the voting public [see also @brunell2010].
 
-Second, the proposed framework moves beyond a single districting plan under study to also consider *counterfactual* districting plans.
+Second, the proposed framework considers *counterfactual* districting plans, not just a single enacted plan.
 Many factors influence the performance and qualities of a districting plan.
-Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are intrinsic consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
-For example, in Massachusetts, it is currently impossible to draw a set of congressional districts that produces even one reliable Republican seat, despite Republican voters making up roughly 32% of the state's voters [@duch:18].
-The 9-0 Democratic composition of the Massachusetts congressional delegation therefore reflects the homogeneity of the state's political geography and not any partisan gerrymandering.
+Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
+For example, in Massachusetts, it is currently impossible to draw even one reliable Republican seat, despite 32% of voters being Republican [@duch:18].
+The 9-0 Democratic composition reflects the homogeneity of the state's political geography, not partisan gerrymandering.
 Once a set of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
-While many existing partisan gerrymandering metrics make use of an *implicit* counterfactual plan, such a plan may not be of substantive interest or even feasible at all.
+Many metrics rely on *implicit* counterfactuals, which may be irrelevant or infeasible.
 By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.
 
-The combination of individual voter choices and counterfactual districting plans defines our notion of *individual harm*: informally, a voter is harmed if they are more likely to be represented by the candidate of their choice under a different districting plan.
-And by analyzing how harms fall differentially along partisan, racial, ideological, and other lines, we can connect individual harm to traditional notions of gerrymandering.
-Section @sec-motivation illustrates this with a simple example, and the general definitions are formalized in Section @sec-indiv-harm.
+The combination of individual voter choices and counterfactual districting plans defines *individual harm*: informally, a voter is harmed if they would be represented by their candidate of choice under a counterfactual plan.
+By examining how harm varies across partisan, racial, or ideological lines, we link it back to traditional gerrymandering.
+Section @sec-motivation gives a simple example Section @sec-indiv-harm formalizes it.
 
-A key feature of the proposed framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022].
+A key feature of our framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022].
 This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry].
@@
 Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
+
+Our proposed harm measures improve on exististing racial and partisan gerrymandering analytical approaches.
+The harm approach corresponds well to measuring the Gingles criteria, can directly use data from ecological inference, and handles complex cases like multi-minority coalitions [@stephanopoulos2014south]
+Further, we explicitly model the counterfactuals and their feasibility, allowing us to improve on existing issues with other measures such as partisan bias [@king1987democratic], efficiency gap [@stephanopoulos2015], and declination [@warrington18].
 A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section @sec-connections.
 
-Harm and differential harm are more than a useful conceptual unification---they are practically useful in a wide variety of redistricting analyses.
-A brief study of partisan gerrymandering of New Jersey congressional districts in Section @sec-nj shows how existing gerrymandering metrics can make incorrect conclusions about the magnitude and even the direction of partisan gerrymanders.
-Section @sec-alabama, studies voting rights litigation for Alabama congressional districts, and uses differential harm to separate the racial and partisan effects of the enacted districting plan.
-Section @sec-boston demonstrates the application of our framework to local redistricting, and quantifies the benefits of race-aware city council district boundaries to Black voters in the city of Boston.
+Harm and differential harm offer practical value across diverse contexts.
+Section @sec-nj analyzes New Jersey congressional districts, showing how traditional metrics can misidentify both the magnitude and direction of partisan gerrymanders.
+Section @sec-alabama uses differential harm to disentangle racial and partisan effects in Alabama's districting.
+Section @sec-boston applies the framework to local redistricting, quantifying the benefits of race-aware boundaries for Black voters in Boston.
 Section @sec-discussion concludes.
@@
 
-The core idea motivating individual and differential harm is a simple one.
 Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure @fig-minissouri.
-The state's total population is 180, and it is divided into three districts with population between 54 and 66.
-Minissouri has two medium-sized cities and a substantial rural population with an overall partisan breakdown of 55.6% R to 44.4% D.
+Its total population is 180, and it is divided into three districts with population between 54 and 66.
+Minissouri has two cities and a substantial rural population with a partisan breakdown of 55.6% R to 44.4% D.
 Voters' preferences are assumed to be stable across elections, and turnout is 100%.
-In addition to partisan affiliation indicated by red or blue coloring, each voter also belongs to a racial group indicated by squares and circles.
+In addition to partisan affiliation (red or blue), each voter also belongs to a racial group (squares or circles).
 
@@
 
-Now suppose the redistricting plan in Figure @fig-minissouri(a) has been enacted.
-Despite the statewide popular vote favoring Republicans, under this plan Democrats will always win two of the three seats.
-In contrast, a more proportional plan @fig-minissouri(b) elects two Republicans and one Democrat.
+Suppose the redistricting plan in Figure @fig-minissouri(a) has been enacted.
+Though Republicans win the popular vote, Democrats win two of three seats.
+In contrast, @fig-minissouri(b) is more proportional, electing two Republicans and one Democrat.
 
-Intuitively, plan (a) unfairly packs Republicans into a rural district, depriving them of one seat.
-To turn this intuition into something more concrete, we can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
+Intuitively, plan (a) unfairly packs Republicans into a rural district.
+We can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
 These voters are highlighted in Figure @fig-minissouri(c).
 For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
-In contrast, Republican voters in the northeast city are not harmed, since they reside in a Democratic district in both the enacted and counterfactual plan.
+In contrast, Republican voters in the northeast city are not harmed, since they reside in a Democratic district under both plans.
 
@@
 We can also tally harm by *race*, which in Minissouri is correlated with both geography and partisanship.
-Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of $-20\%$, where positive values indicate pro-square bias.
-In other words, square voters are 20 percentage points more likely to be harmed than circle voters.
-While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or merely aimed to maximize Democratic representation, it does quantify the impact of plan (a) on these voters versus the alternative plan (b), and it does so on the same scale as differential partisan harm.
+Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of $-20\%$.
+This means square voters are 20 percentage points more likely to be harmed than circle voters.
+While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or aimed to maximize Democratic representation, it quantifies the impact of plan (a) on these voters versus the alternative plan (b), on the same scale as differential partisan harm.
 
@@
 Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe in @sec-methods.
-
-# Connections to Current Practice for Analyzing Partisan and Racial Gerrymandering {#sec-connections}
-
-A primary advantage of our proposed harm measures is that they are defined at the individual level, allowing for the study of redistricting for any defined group memberships.
-For political science, two of the most important group memberships are perhaps race and party.
-Even in these well-studied contexts, our measures provide an improvement over existing analytical approaches.
-
-## Racial Vote Dilution {#sec-racial}
-
-The study of racial gerrymandering and vote dilution is considerably more complicated than that of partisan gerrymandering, due to the difficulty of understanding racial voting preferences when individual votes are secret.
-No simple formulas exist to measure the precise racial impacts of a districting plan.
-Indeed, much of the academic literature has focused on the statistical challenges of estimating vote preferences and turnout by race [@king1997solution; @greiner2006ecological; @imai2016improving].
-
-Modern analyses of vote dilution and racial gerrymandering are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
-Section 2 of the VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
-Under the current framework for vote dilution established by @1986gingles, plaintiffs must prove that (a) the minority group is large and geographically compact enough to form a majority in some hypothetical district, (b) the minority group is "politically cohesive," and (c) the majority group can and does vote to deny the minority group its choice of candidate [@hood2018legal] (the so-called *Gingles* criteria).
-Criteria (a) is generally proven through a single example districting plan, while criteria (b) and (c) are shown using the statistical estimates of vote preferences and turnout by race.
-
-While the specific legal requirements for a vote dilution claim will obviously remain front-and-center for court challenges of racial gerrymanders, our proposed harm measures unifies all three *Gingles* criteria.
-Criteria (b) and (c) capture the misrepresentation of minority voters by a candidate they do not support, while (a) requires the existence of a counterfactual where this would not be the case and (a) and (c) together establish that minority voters are harmed.
-But our proposed measures can also tackle trickier situations where the *Gingles* criteria may not be clearly met.
-
-One such case arises when there are several distinct minority racial groups with similar preferences.
-In these cases, some courts have found that (a) may be weakened to consider "coalition" districts that still allow minority voters overall to elect candidates of their choice, but there is no clear rule [@stephanopoulos2014south].
-Or when urban White voters have similar partisan preferences to minority voters, (b) and (c) may not be strictly satisfied, especially for two-party general elections.
-Some argue, however, that courts should still rely on the VRA in these cases to protect minority voters, especially in light of differing racial preferences in primary elections [@becker2021computational].
-
-In both these cases, the proposed harm measures quantify the extent to which a challenged districting plan differentially harms minority voters.
-They account for geographic sorting, political cohesion, and the individual preferences of every voter.
-Sections @sec-alabama and @sec-boston demonstrate this approach in practice.
-
-## Partisan Gerrymandering
-
-The canonical framework for considering partisan fairness in redistricting is partisan symmetry [@king1987democratic].
-Partisan symmetry implies that the seats $s$ won by a party with vote share $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
-In practice, the partisan symmetry framework is applied via the partisan bias metric $\beta(v)$, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020], i.e., the number of seats that would need to be reallocated to reach partisan symmetry for a given statewide vote share $v$ (usually $\half$).
-
-More recently, and particularly in the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
-The *mean-median* difference looks at the relationship between the vote share of the median seat in a delegation and the mean vote share of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
-*Declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
-Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015], which measures the difference in the number of the votes wasted by each party.[^3]
-
-[^3]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.
-    They are exactly the votes which don't contribute to winning a seat.
-
-The latter three metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in the larger theoretical framework of partisan symmetry [@katz2020].
-As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
-This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section @sec-estimating.[^4]
-This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
-Plugging in past election results without using an election model assumes too much about the stability of electoral dynamics and consequently can significantly understate the uncertainty in how a particular districting plan will perform into the future.
-
-[^4]: For partisan bias, both researchers and practitioners have largely relied on simple uniform-shift models similar to the one we use in this work.
-
-Since none of the existing measures are centered on individual voters, they offer far less granular insights than the proposed harm framework, even in the case of studying partisan gerrymandering.
-For example, differential harm can be estimated at the precinct level and then plotted on a map to show not just *which party* is harmed by a gerrymander, but *where* this harm occurs.
-
-All of these existing measures are constructed so that a value of zero indicates no bias, and a value of zero is at least implicitly held up as a normative standard for districting plans.
-In doing so, these measures allude to *implicit* counterfactuals.
-For example, with partisan bias, the implicit electoral counterfactual is an election in which the parties' vote shares have been reversed, and the implicit counterfactual districting plan is one which assigns equal vote shares to the parties under both the real-world and counterfactual electoral scenarios.
-Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partisan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome." Similar implicit counterfactuals can be thought up for the other measures.
-
-These implicit counterfactuals differ from the explicit counterfactuals we adopt in this work in that there is no guarantee they are possible.
-@katz2020 and others recognize that a nonzero value in a measure like partisan bias can only be considered problematic "if there is a fairer alternative." The inability of partisan bias and other existing measures to explicitly account for realistic counterfactual districting plans means that none of them may be interpreted in a vacuum.
-No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the electoral system's geography.
-
-It is this confounding which explains many of the concerns that have been raised over the existing measures.
-The standard calculation of partisan bias, $\beta(\half)$, can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.[^5]
-And none of the measures take into account any aspect of political geography, and therefore suffer in states like Massachusetts, where, for example, the efficiency gap takes a value generally indicating an extreme Democratic gerrymander.
-
-[^5]: This has been called by some the "Utah paradox" [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
-    Even when interpreted carefully, however, partisan bias requires significant extrapolation from observed election outcomes in lopsided states---how realistic is it to evaluate the counterfactual where Democrats run even with Republicans statewide in Utah?
-
-Despite these differences, there are also empirical similarities in many cases.
-The efficiency gap is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
-Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
-Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section @sec-nj.
-
-[^6]: The differences stem from the efficiency gap considering a vote to be wasted if it is in excess of the 50% threshold needed to win, while the harm framework does not penalize such votes, since every such voter is happily represented by their candidate of choice.
-    This definition of wasted votes is also at the root of other criticisms of the efficiency gap, which note its tendency to penalize proportionality and reward 3-to-1 districts [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
-
-Finally, it is worth comparing the proposed measures to the increasingly common application redistricting simulation techniques to cases of partisan gerrymandering.
-Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics.
-While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
-In the worst cases, this can lead to directionally wrong conclusions, as we see in the next section.
 
@@
 
-```{r fig-boston-factors, fig.cap="", fig.width=7, fig.height=3.5}
+```{r fig-boston-factors, fig.width=7, fig.height=3.5}
 #| fig.cap: |
@@
     To deal with this issue completely, one must model elections at the precinct level, which requires more data and computational costs.
+
+# Connections to Current Practice for Analyzing Partisan and Racial Gerrymandering {#sec-connections}
+
+A primary advantage of our proposed harm measures is that they are defined at the individual level, allowing for the study of redistricting for any defined group memberships.
+For political science, two of the most important group memberships are perhaps race and party.
+Even in these well-studied contexts, our measures provide an improvement over existing analytical approaches.
+
+## Racial Vote Dilution {#sec-racial}
+
+The study of racial gerrymandering and vote dilution is considerably more complicated than that of partisan gerrymandering, due to the difficulty of understanding racial voting preferences when individual votes are secret.
+No simple formulas exist to measure the precise racial impacts of a districting plan.
+Indeed, much of the academic literature has focused on the statistical challenges of estimating vote preferences and turnout by race [@king1997solution; @greiner2006ecological; @imai2016improving].
+
+Modern analyses of vote dilution and racial gerrymandering are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
+Section 2 of the VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
+Under the current framework for vote dilution established by @1986gingles, plaintiffs must prove that (a) the minority group is large and geographically compact enough to form a majority in some hypothetical district, (b) the minority group is "politically cohesive," and (c) the majority group can and does vote to deny the minority group its choice of candidate [@hood2018legal] (the so-called *Gingles* criteria).
+Criteria (a) is generally proven through a single example districting plan, while criteria (b) and (c) are shown using the statistical estimates of vote preferences and turnout by race.
+
+While the specific legal requirements for a vote dilution claim will obviously remain front-and-center for court challenges of racial gerrymanders, our proposed harm measures unifies all three *Gingles* criteria.
+Criteria (b) and (c) capture the misrepresentation of minority voters by a candidate they do not support, while (a) requires the existence of a counterfactual where this would not be the case and (a) and (c) together establish that minority voters are harmed.
+But our proposed measures can also tackle trickier situations where the *Gingles* criteria may not be clearly met.
+
+One such case arises when there are several distinct minority racial groups with similar preferences.
+In these cases, some courts have found that (a) may be weakened to consider "coalition" districts that still allow minority voters overall to elect candidates of their choice, but there is no clear rule [@stephanopoulos2014south].
+Or when urban White voters have similar partisan preferences to minority voters, (b) and (c) may not be strictly satisfied, especially for two-party general elections.
+Some argue, however, that courts should still rely on the VRA in these cases to protect minority voters, especially in light of differing racial preferences in primary elections [@becker2021computational].
+
+In both these cases, the proposed harm measures quantify the extent to which a challenged districting plan differentially harms minority voters.
+They account for geographic sorting, political cohesion, and the individual preferences of every voter.
+Sections @sec-alabama and @sec-boston demonstrate this approach in practice.
+
+## Partisan Gerrymandering
+
+The canonical framework for considering partisan fairness in redistricting is partisan symmetry [@king1987democratic].
+Partisan symmetry implies that the seats $s$ won by a party with vote share $v$ would be satisfied if $s(v) = 1 - s(1 - v)$.
+In practice, the partisan symmetry framework is applied via the partisan bias metric $\beta(v)$, which measures the seat deviation from symmetry [@tufte1973pb; @katz2020], i.e., the number of seats that would need to be reallocated to reach partisan symmetry for a given statewide vote share $v$ (usually $\half$).
+
+More recently, and particularly in the years since @2004vieth, researchers have developed a veritable zoo of measures to identify partisan gerrymanders.
+The *mean-median* difference looks at the relationship between the vote share of the median seat in a delegation and the mean vote share of the delegation to make claims about how a map is skewed [@mcdonald2015unfair].
+*Declination* describes the relative rate at which districts become less competitive for each party [@warrington18].
+Perhaps the new measure which has reached the most acclaim is the *efficiency gap* [@stephanopoulos2015], which measures the difference in the number of the votes wasted by each party.[^3]
+
+[^3]: Wasted votes are those cast for a losing candidate, as well as those cast for a winning candidate beyond the required 50% to win the election.
+    They are exactly the votes which don't contribute to winning a seat.
+
+The latter three metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in the larger theoretical framework of partisan symmetry [@katz2020].
+As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
+This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section @sec-estimating.[^4]
+This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
+Plugging in past election results without using an election model assumes too much about the stability of electoral dynamics and consequently can significantly understate the uncertainty in how a particular districting plan will perform into the future.
+
+[^4]: For partisan bias, both researchers and practitioners have largely relied on simple uniform-shift models similar to the one we use in this work.
+
+Since none of the existing measures are centered on individual voters, they offer far less granular insights than the proposed harm framework, even in the case of studying partisan gerrymandering.
+For example, differential harm can be estimated at the precinct level and then plotted on a map to show not just *which party* is harmed by a gerrymander, but *where* this harm occurs.
+
+All of these existing measures are constructed so that a value of zero indicates no bias, and a value of zero is at least implicitly held up as a normative standard for districting plans.
+In doing so, these measures allude to *implicit* counterfactuals.
+For example, with partisan bias, the implicit electoral counterfactual is an election in which the parties' vote shares have been reversed, and the implicit counterfactual districting plan is one which assigns equal vote shares to the parties under both the real-world and counterfactual electoral scenarios.
+Indeed, @katz2020 reference an implicit counterfactual in discussing the normative implications of partisan symmetry, writing that "[o]bviously, a process designed to be 'fair' that turns out to be substantially asymmetric [i.e., have nonzero partisan bias], if there is a fairer alternative, would not normally be regarded as a legitimate outcome." Similar implicit counterfactuals can be thought up for the other measures.
+
+These implicit counterfactuals differ from the explicit counterfactuals we adopt in this work in that there is no guarantee they are possible.
+@katz2020 and others recognize that a nonzero value in a measure like partisan bias can only be considered problematic "if there is a fairer alternative." The inability of partisan bias and other existing measures to explicitly account for realistic counterfactual districting plans means that none of them may be interpreted in a vacuum.
+No particular value can be used as a threshold past which a plan becomes a gerrymander, because the measures confound the map-drawer's bias with the bias inherent to the electoral system's geography.
+
+It is this confounding which explains many of the concerns that have been raised over the existing measures.
+The standard calculation of partisan bias, $\beta(\half)$, can give unintuitive results in more lopsided states, which, if not interpreted carefully, can lead practitioners to make incorrect conclusions about the presence and direction of a partisan gerrymander.[^5]
+And none of the measures take into account any aspect of political geography, and therefore suffer in states like Massachusetts, where, for example, the efficiency gap takes a value generally indicating an extreme Democratic gerrymander.
+
+[^5]: This has been called by some the "Utah paradox" [@deford2020symmetry], though the paradoxical nature is contested as a mistake of using it for description, rather than inference [@katz2022].
+    Even when interpreted carefully, however, partisan bias requires significant extrapolation from observed election outcomes in lopsided states---how realistic is it to evaluate the counterfactual where Democrats run even with Republicans statewide in Utah?
+
+Despite these differences, there are also empirical similarities in many cases.
+The efficiency gap is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
+Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
+Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section @sec-nj.
+
+[^6]: The differences stem from the efficiency gap considering a vote to be wasted if it is in excess of the 50% threshold needed to win, while the harm framework does not penalize such votes, since every such voter is happily represented by their candidate of choice.
+    This definition of wasted votes is also at the root of other criticisms of the efficiency gap, which note its tendency to penalize proportionality and reward 3-to-1 districts [@bernstein2017egap; @tamcho2017egap; @deford2020symmetry].
+
+Finally, it is worth comparing the proposed measures to the increasingly common application redistricting simulation techniques to cases of partisan gerrymandering.
+Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics.
+While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.
+In the worst cases, this can lead to directionally wrong conclusions.
 

### 2025-07-30 paper/figures/corr_table.tex
cut lots

@@
 \midrule
-Utah & \cellcolor[HTML]{8A671F}{-0.983} & \cellcolor[HTML]{75689E}{0.957} & \cellcolor[HTML]{927421}{-0.893} & \cellcolor[HTML]{8E6C20}{-0.945} & \cellcolor[HTML]{947722}{-0.876}\\
-Mississippi & \cellcolor[HTML]{9D8627}{-0.772} & \cellcolor[HTML]{BFD9D5}{0.207} & \cellcolor[HTML]{E8EEE3}{0.031} & \cellcolor[HTML]{E7E5C0}{-0.096} & \cellcolor[HTML]{A89735}{-0.656}\\
-Iowa & \cellcolor[HTML]{AD9F3D}{-0.597} & \cellcolor[HTML]{A0B5C2}{0.449} & \cellcolor[HTML]{DED99D}{-0.185} & \cellcolor[HTML]{C4DCD7}{0.184} & \cellcolor[HTML]{A9C2C9}{0.365}\\
-North Carolina & \cellcolor[HTML]{8E6C20}{-0.943} & \cellcolor[HTML]{8A88AE}{0.740} & \cellcolor[HTML]{9AAABC}{0.520} & \cellcolor[HTML]{8C8DB0}{0.704} & \cellcolor[HTML]{9BAABC}{0.519}\\
-Wisconsin & \cellcolor[HTML]{998024}{-0.816} & \cellcolor[HTML]{929BB4}{0.621} & \cellcolor[HTML]{C2DBD7}{0.191} & \cellcolor[HTML]{9096B2}{0.652} & \cellcolor[HTML]{A8C1C8}{0.374}\\
-New Jersey & \cellcolor[HTML]{9B8425}{-0.789} & \cellcolor[HTML]{9FB4C1}{0.459} & \cellcolor[HTML]{DCE8DF}{0.082} & \cellcolor[HTML]{EEF0E0}{-0.014} & \cellcolor[HTML]{A7C0C7}{0.381}\\
-Massachusetts & \cellcolor[HTML]{987E23}{-0.826} & \cellcolor[HTML]{E9E9CA}{-0.069} & \cellcolor[HTML]{E1DDA7}{-0.158} & \cellcolor[HTML]{CADFD9}{0.159} & \cellcolor[HTML]{808080}{NA}\\
-California & \cellcolor[HTML]{AEA03E}{-0.590} & \cellcolor[HTML]{939BB4}{0.616} & \cellcolor[HTML]{DCD798}{-0.199} & \cellcolor[HTML]{D1C97B}{-0.302} & \cellcolor[HTML]{AAC3C9}{0.359}\\
+Utah & \cellcolor[HTML]{8A671F}{$-0.983$} & \cellcolor[HTML]{75689E}{$0.957$} & \cellcolor[HTML]{927421}{$-0.893$} & \cellcolor[HTML]{8E6C20}{$-0.945$} & \cellcolor[HTML]{947722}{$-0.876$}\\
+Mississippi & \cellcolor[HTML]{9D8627}{$-0.772$} & \cellcolor[HTML]{BFD9D5}{$0.207$} & \cellcolor[HTML]{E8EEE3}{$0.031$} & \cellcolor[HTML]{E7E5C0}{$-0.096$} & \cellcolor[HTML]{A89735}{$-0.656$}\\
+Iowa & \cellcolor[HTML]{AD9F3D}{$-0.597$} & \cellcolor[HTML]{A0B5C2}{$0.449$} & \cellcolor[HTML]{DED99D}{$-0.185$} & \cellcolor[HTML]{C4DCD7}{$0.184$} & \cellcolor[HTML]{A9C2C9}{$0.365$}\\
+North Carolina & \cellcolor[HTML]{8E6C20}{$-0.943$} & \cellcolor[HTML]{8A88AE}{$0.740$} & \cellcolor[HTML]{9AAABC}{$0.520$} & \cellcolor[HTML]{8C8DB0}{$0.704$} & \cellcolor[HTML]{9BAABC}{$0.519$}\\
+Wisconsin & \cellcolor[HTML]{998024}{$-0.816$} & \cellcolor[HTML]{929BB4}{$0.621$} & \cellcolor[HTML]{C2DBD7}{$0.191$} & \cellcolor[HTML]{9096B2}{$0.652$} & \cellcolor[HTML]{A8C1C8}{$0.374$}\\
+New Jersey & \cellcolor[HTML]{9B8425}{$-0.789$} & \cellcolor[HTML]{9FB4C1}{$0.459$} & \cellcolor[HTML]{DCE8DF}{$0.082$} & \cellcolor[HTML]{EEF0E0}{$-0.014$} & \cellcolor[HTML]{A7C0C7}{$0.381$}\\
+Massachusetts & \cellcolor[HTML]{987E23}{$-0.826$} & \cellcolor[HTML]{E9E9CA}{$-0.069$} & \cellcolor[HTML]{E1DDA7}{$-0.158$} & \cellcolor[HTML]{CADFD9}{$0.159$} & \cellcolor[HTML]{808080}{NA}\\
+California & \cellcolor[HTML]{AEA03E}{$-0.590$} & \cellcolor[HTML]{939BB4}{$0.616$} & \cellcolor[HTML]{DCD798}{$-0.199$} & \cellcolor[HTML]{D1C97B}{$-0.302$} & \cellcolor[HTML]{AAC3C9}{$0.359$}\\
 \bottomrule

### 2025-07-30 paper/harm.qmd
cut lots

@@
 The study of racial gerrymandering has been largely shaped by the legal framework for challenging such gerrymanders: the Voting Rights Act of 1965 (VRA).
-This literature is characterized by a focus on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
+This literature focuses on identifying the voting preferences of racial groups through techniques such as ecological inference [@oloughlin1982id; @grofman1991identifying; @king1997solution; @greiner2006ecological].
 Work on local redistricting is even more sparse, in part because of data limitations [but see @siegel2013educational; @abott2020large; @vargas2021racial].
@@
 The dominance of partisan frameworks limits redistricting analysis, even in purely partisan cases.
-By focusing only on party effects, existing tools overlook how partisan lines also have an effect of relative voting power by race, wealth, age, and urbanicity.
-
+By focusing only on party effects, existing tools overlook how partisan lines also have an effect of relative voting power on axes such as race, wealth, and urbanicity.
 This paper addresses these limitations with a new framework of *individual and differential harm*.
@@
 
-The harm framework combines two key ideas, further motivated in Section @sec-motivation.
-First, it operates at the level of the individual voter and examines whether each voter is represented by the candidate of their choice.
+The harm framework combines two key ideas, further motivated in @sec-motivation.
+First, it operates at the level of the individual voter and measures whether each voter is represented by the candidate of their choice.
 This unifies gerrymandering analyses across contexts and reflects the goal that elected officials represent the preferences of the voting public [see also @brunell2010].
@@
 Second, the proposed framework considers *counterfactual* districting plans, not just a single enacted plan.
-Many factors influence the performance and qualities of a districting plan.
-Without a well-defined comparison, it is impossible to differentiate whether certain aspects of the system, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
-For example, in Massachusetts, it is currently impossible to draw even one reliable Republican seat, despite 32% of voters being Republican [@duch:18].
-The 9-0 Democratic composition reflects the homogeneity of the state's political geography, not partisan gerrymandering.
-Once a set of districting rules has been agreed on, what matters is not how a plan measures up to a fixed ideal, but how it performs in comparison to counterfactual plans.
-Many metrics rely on *implicit* counterfactuals, which may be irrelevant or infeasible.
+Many factors influence the qualities of a districting plan, and without a well-defined comparison, it is impossible to differentiate which qualities, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain @chen:rodd:13.
+For example, in Massachusetts in 2010, it was impossible to draw even one majority-Republican seat, despite 32% of voters preferring Republicans @duch:18.
+Thus the 9-0 Democratic Congressional delegation reflects the homogeneity of the state's political geography, not partisan gerrymandering.
+Many existing districting metrics rely on *implicit* counterfactuals, which may be irrelevant or infeasible, as we discuss in [Appendix @sec-connections].
 By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.
@@
 The combination of individual voter choices and counterfactual districting plans defines *individual harm*: informally, a voter is harmed if they would be represented by their candidate of choice under a counterfactual plan.
-By examining how harm varies across partisan, racial, or ideological lines, we link it back to traditional gerrymandering.
-Section @sec-motivation gives a simple example Section @sec-indiv-harm formalizes it.
+Once measured for each voter, researchers can examine the distribution of harm by geography, partisanship, race, or other factors.
+@sec-motivation gives a simple example of this idea; @sec-indiv-harm formalizes it.
 
 A key feature of our framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022].
-This distinction is also missed by many critics of the efficiency gap [@veomett2018efficiency; @deford2020symmetry].
-The challenge of estimating harm and differential harm from observed data is discussed in Section @sec-estimating.
+The methods section discusses the challenge of estimating harm and differential harm from observed data.
 Recent advances in redistricting simulation, which can generate a large number of random districting plans that respect relevant legal constraints, are critical to this effort [@fifield2020mcmc; @deford2019; @carter2019; @mccartan2020].
@@
 Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
+The harm measure also captures or generalizes the three *Gingles* preconditions for vote dilution litigation under the Voting Rights Act of 1965, as discussed further in [Appendix @sec-connections].
 
-Our proposed harm measures improve on exististing racial and partisan gerrymandering analytical approaches.
-The harm approach corresponds well to measuring the Gingles criteria, can directly use data from ecological inference, and handles complex cases like multi-minority coalitions [@stephanopoulos2014south]
-Further, we explicitly model the counterfactuals and their feasibility, allowing us to improve on existing issues with other measures such as partisan bias [@king1987democratic], efficiency gap [@stephanopoulos2015], and declination [@warrington18].
-A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section @sec-connections.
-
-Harm and differential harm offer practical value across diverse contexts.
-Section @sec-nj analyzes New Jersey congressional districts, showing how traditional metrics can misidentify both the magnitude and direction of partisan gerrymanders.
-Section @sec-alabama uses differential harm to disentangle racial and partisan effects in Alabama's districting.
-Section @sec-boston applies the framework to local redistricting, quantifying the benefits of race-aware boundaries for Black voters in Boston.
-Section @sec-discussion concludes.
+We demonstrate the practical value of the proposed framework in three diverse contexts.
+@sec-nj analyzes New Jersey congressional districts, showing how traditional metrics can misidentify both the magnitude and direction of partisan gerrymanders.
+@sec-alabama uses differential harm to disentangle racial and partisan effects in Alabama's districting.
+@sec-boston applies the framework to local redistricting, quantifying the benefits of race-aware boundaries for Black voters in Boston.
+@sec-discussion concludes.
 
@@
 
-Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure @fig-minissouri.
+Consider *Minissouri*, a stylized representation of Missouri which is shown in @fig-minissouri.
 Its total population is 180, and it is divided into three districts with population between 54 and 66.
@@
 
-Suppose the redistricting plan in Figure @fig-minissouri(a) has been enacted.
+Suppose the redistricting plan in @fig-minissouri(a) has been enacted.
 Though Republicans win the popular vote, Democrats win two of three seats.
@@
 
-Intuitively, plan (a) unfairly packs Republicans into a rural district.
-We can identify which voters are harmed by the enactment of plan (a) over the alternative (b).
-These voters are highlighted in Figure @fig-minissouri(c).
+Compared to plan (b), plan (a) packs Republicans into a rural district.
+@fig-minissouri highlights the voters who are harmed by the enactment of plan (a) over the alternative (b).
 For example, in the upper leftmost precinct, there are three Republican square voters who are placed into a Democratic district in plan (a) but would have been placed into a Republican district in plan (b), and thus are harmed.
@@
 Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
-These represent 44% of all Republicans and 2.5% of all Democrats.
+These represent $44\%$ of all Republicans and $2.5\%$ of all Democrats.
 We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
-Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is 41.5 percentage points less likely to be represented by a Republican than the average Democrat is to be represented by a Democrat, *compared to plan (b)*.
+Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is $41.5$pp less likely to be represented by a Republican than the average Democrat is to be represented by a Democrat, *compared to plan (b)*.
 
@@
 Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of $-20\%$.
-This means square voters are 20 percentage points more likely to be harmed than circle voters.
-While this number cannot establish whether the drawers of plan (a) intended to harm square voters, or aimed to maximize Democratic representation, it quantifies the impact of plan (a) on these voters versus the alternative plan (b), on the same scale as differential partisan harm.
+This means square voters are $20$pp more likely to be harmed than circle voters.
+While this comparison alone cannot establish whether the drawers of plan (a) intended to harm square voters versus maximize Democratic representation, it quantifies the impact of plan (a) on these voters versus the alternative plan (b), on the same scale as differential partisan harm.
 
@@
 
-We formalize and generalize the concepts in Section @sec-motivation.
+We formalize and generalize the concepts in @sec-motivation.
 Let $\Voters$ be the set of individuals, or *voters*, in the region and time period.
@@
 Individuals vote in elections, each of which produces a specific outcome $v$ from the set $\Elecs$ of all possible election outcomes.
-In an election, voters either vote for a candidate running in their district, or abstain from voting.
+<!-- In an election, voters either vote for a candidate running in their district, or abstain from voting. -->
 We write $v(i)$ to mean the candidate that individual $i$ votes for in election $v$.
@@
 For example, different districting plans may lead to different candidates running, depending on district partisanship and incumbent residency.
-The assumptions we make on these relationships will be made more explicit in Section @sec-estimating, where we discuss how to estimate our measures in practice.
+The assumptions we make on these relationships are made explicit by the statistical models used to estimate the harm measure; these models may be as simple or complex as the researcher chooses, as we discuss in the methods section.
 
@@
 
-The key intuition from Section @sec-motivation is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
+The key intuition from @sec-motivation is that to be harmed, a voter must have a better outcome in the counterfactual world where a different districting plan was in place.
 Letting $\tilde q$ denote a counterfactual districting plan, we define the *harm* to an individual $i$ in election $v$ held under districting plan $q$ as
@@
 
-A voter is harmed only if they are not represented by the candidate of their choice under $q$ ($\ind\{v(i) \neq \win_i(v, q)\}$) but they would have been represented by their candidate of choice under the counterfactual plan $\tilde q$ ($\ind\{v(i) = \win_i(v, \tilde q)\}$).
+Harm requires both terms to be true: the first measuring if a voter is not represented by their chosen candidate under $q$, and the second term measuring if they would have been represented by their chosen candidate under the counterfactual plan $\tilde q$.
 
@@
 In practice, voters, courts, or politicians may differ in their views on how to balance preventing harm and maximizing gains.
-
-Our definition of harm is limited in two regards.
-First, it considers only individuals who vote: people who abstain from voting have experienced zero harm.
-People who do not vote express no formal preference in electoral outcomes, even if they have informal preferences.
-The natural theoretical extension of this definition is to then separate individuals' candidate preferences from their vote choice; the latter defines election winners, but the former provides a more-broad harm measure.
-Naturally, implementing this extension in practice would require auxiliary data, such as a large-scale survey, since candidate preferences for non-voters cannot be directly inferred from election returns.
-
-A second limitation is that $h_i(v,q,\tilde q)$ ignores any preferences a voter may have between the candidates that they did not vote for.
-For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
-If the voter ultimately voted for candidate A, this harm measure makes no distinction between candidate B or candidate C winning the voter's district.
-This reflects the simplistic nature of the cost function implicitly assumed in our proposed measure.
-The problem could be remedied by replacing the indicator functions in the measure with other cost functions that consider ordinal preferences.
-Ballot-level ranked-choice vote data, or candidate ideology scores, could be used to estimate such a cost function.
-Here, for simplicity and to avoid making further assumptions, we focus on the indicator-based cost function.
-Section @sec-boston further discusses the implications of this particular specification in the context of primary elections.
 
@@
 Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
-A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in Section @sec-estimating.
+A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in the methods section.
 
@@
 
-Despite being only a scalar quantity, $H(q)$ actually fully captures the harm created by the plan $q$.
+<!-- Despite being only a scalar quantity, $H(q)$ actually fully captures the harm created by the plan $q$.
 We can zoom in to specific dimensions of the plan by conditioning on values of $V$, $\tilde Q$, and $I$, or by taking expectations.
@@
 $$
-that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual.
+that is, the probability that a randomly selected voter in this election is misrepresented under the adopted plan but would not have been under a randomly selected counterfactual. -->
 
-## Differential Harm and Fairness {#sec-diff-harm}
+## Differential Harm {#sec-diff-harm}
 
@@
 
-Once we have defined a set of groups, we say that a plan $q$ is *fair* for groups $\Groups$ in comparison to a counterfactual $\tilde Q$ if $\E[H(q)\mid I\in G_k]$ is constant across groups $G_k\in\Groups$.
-Fairness is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
+Once we have defined a set of groups, we define the *differential harm* for two groups $G_k$ and $G_l$ as
+$$
+  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l].
+$$
+So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
+
+Zero differential harm between groups is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
 There, the implicit counterfactual is a perfect predictor.
@@
 Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.
-
-Fairness is a strict condition: real-life plans are highly unlikely to be exactly fair.
-To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as
-$$
-  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
-$$
-for two groups $G_k$ and $G_l$.
-So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
-Using our definition of fairness, $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.
 
@@
 
-Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe in @sec-methods.
+Estimating these quantities from available data is another matter, of course, and requires additional assumptions, as we describe in the methods section.
 
@@
 
-To illustrate the connections in the preceding section and to demonstrate the value of the proposed measures in a real-world setting, we provide a brief analysis of congressional districting in the state of New Jersey.
+To demonstrate the proposed framework in a classic partisan gerrymandering context, we provide a brief analysis of congressional districting in the state of New Jersey.
+We show that differential harm correctly identifies possible gerrymandering, in contrast to some common redistricting metrics.
+Appendices [-@sec-app-nj] and [-@sec-app-corr] contain further comparisons between gerrymandering metrics.
+
 Since 1995, New Jersey congressional redistricting has been handled by a 13-member commission, with 6 political appointees from each party and one independent chair with *de facto* tiebreaking authority.
-For the 2020 redistricting cycle, each party submitted a proposed plan to John Wallace Jr, the chair.
-Wallace ultimately selected the Democratic plan for enactment, "simply because in the last redistricting, the map was drawn by Republicans."[^7] Here, in contrast, we evaluate the two plans on the basis of existing gerrymandering metrics as well as through the proposed harm framework.
+@fig-nj-maps shows the baseline Democratic vote share by precinct, along with the Democratic and Republican commission proposals.[^8]
+In a typical election, the Democratic proposal yields 9 Democratic and 3 Republican seats, while the Republican proposal gives a 7--5 split.
 
-[^7]: See Matt Friedman, "Democrats prevail in New Jersey redistricting with map that could sacrifice Malinowski," *Politico,* December 22, 2021.
-    <https://www.politico.com/news/2021/12/22/new-jersey-redistricting-map-malinowski-525983> (accessed July 3, 2022).
+[^8]: The Democratic plan was enacted.
+The GOP plan was provided by Harrison Neely, executive director for the GOP delegation to the New Jersey Congressional Redistricting Committee, following an Open Public Records Act request.
 
@@
 #|      Baseline vote patterns (a), two hypothetical gerrymanders (b) and (c),
-#|      and one randomly sampled plan (d).
+#|      and one randomly sampled counterfactual plan (d).
 #|      County lines are overlaid in white.
@@
 
-Figure @fig-nj-maps shows the baseline Democratic vote share by precinct, along with the Democratic and Republican proposals.[^8]
-Using an average of the 2016 presidential and 2018 senatorial elections as a baseline, the Democratic proposal yields 9 Democratic seats and 3 Republican seats, while the Republican proposal has 7 Democratic and 5 Republican seats.
+To provide a neutral comparison baseline, we generate 10,000 districting plans across using the SMC algorithm of [@mccartan2020; implemented in @redist] setting constraints to favor compact plans that respect county lines.
+@fig-nj-maps(d) shows one such plan.
+For each of these sampled plans and the two proposals, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats and several common partisan metrics.
+These measures, and their distribution in the basline, are shown in @fig-nj-meas.
 
-[^8]: The enacted plan is sourced from [All About Redistricting](https://redistricting.lls.edu/state/new-jersey/?cycle=2020&level=Congress&startdate=2021-12-22), and the GOP plan was provided by Harrison Neely, executive director for the GOP delegation to the New Jersey Congressional Redistricting Committee, following an Open Public Records Act request.
+[^9]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so positive values correspond to a Republican-favoring plan.
 
-We sample 10,000 districting plans across four independent runs of the SMC algorithm of @mccartan2020, as implemented in the R package `redist` [@redist], setting constraints to favor compact plans with district boundaries that tend to follow county lines.
-One such plan is shown in Figure @fig-nj-maps(d).
-For each of these sampled plans and the two gerrymanders, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats, the efficiency gap, partisan bias, the mean-median score, and partisan declination.
-The distribution of these measures is shown in Figure @fig-nj-meas.
-
-[^9]: We choose to calculate $\Delta_{\dem-\rep}H(q)$, so that positive values correspond to a plan that favors Republicans.
-
-We first examine the values of each metric on the two proposals alone.
 The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-3.6\%$ for the Democratic proposal and $2.0\%$ for the Republican proposal.
-These values are relatively directly interpretable.
-For example, the estimated differential harm of $-3.6\%$ means that an average Democratic voter is 3.6 percentage points less likely to be misrepresented than their average Republican counter-partisan.
-The efficiency gap, declination, and partisan bias measures agree with the differential harm estimates in this regard, though the former two are much more discrete.
-The efficiency gap and partisan bias estimate give roughly equal scores in absolute value to the Democratic and Republican plans, in comparison to differential harm and declination, which identify the Democratic proposal as slightly more extreme.
-While the mean-median measure identifies the Republican proposal as favoring Republicans, it also scores the Democratic proposal as favoring the Republicans, albeit by less.
+This means that, for example, an average Democratic voter is $3.6$pp less likely to be misrepresented than their average Republican counter-partisan.
+Overall, the Democratic plan has lower average harm.
 
@@
 
-These values can be better contextualized in comparison to the distribution across the sampled plans.
-The Democratic proposal is expected to produce an average of 8.55 Democratic seats under our election model, more than 79% of all sample plans.
-The Republican proposal is an outlier, yielding fewer seats on average than every single sampled plan.
-This pattern is repeated for all of the four common gerrymandering measures, where the Republican plan is a clear outlier (all $p=0.0002$, except $p=0.002$ for partisan bias) while the Democratic plan is not ($p=0.34$ for the efficiency gap, $0.50$ for declination, $0.88$ for partisan bias, and $0.17$ for the mean-median score).
-However, both the partisan bias estimate and the mean-median score again misidentify the direction of the bias of the Republican plan, classifying it as an extreme Democratic outlier.
+Most of the other metrics, with the notable exception of the mean-median measure, agree with differential harm that each party's proposal favors itself.
+The efficiency gap and partisan bias estimate give roughly equal scores in absolute value to the Democratic and Republican plans, in comparison to differential harm and declination, which identify the Democratic proposal as slightly more extreme.
 
-The estimate of differential harm already uses the simulation output, and so comparing the proposals to the sampling distribution is not strictly required.
-However, it is notable that while the Democratic proposal has a more extreme estimate of differential partisan harm (1.6 percentage points higher), its value of $-3.7\%$ is not an outlier with respect to the distribution of differential harm for a randomly sampled plan ($p=0.43$, using the same set of samples as a counterfactual), while the Republican plan is ($p=0.0002$).
-In other words, a *typical* nonpartisan districting plan is not necessarily a *fair* one, even in comparison to other similarly generated districting plans.
-The sampled plan which minimizes estimated differential harm yields 8.18 expected Democratic seats, compared to a sample average of 8.41 seats, and the 6.8 seats that would be required for an outcome completely proportional to the parties' statewide vote share.
-
-Both proposals are estimated to harm the average voter at a higher rate than most of the sampled plans, with the Republican plan again more extreme in this regard ($p=0.0008$ versus $p=0.061$ for the Democratic plan).
-The plans which minimize estimated total harm are also those which have near-zero estimated differential partisan harm.
-Thus, at least in New Jersey, overall harm reduction is consistent with a fair map on a partisan basis.
-Minimizing bias with other measures, however, could lead one astray: plans which score an efficiency gap of zero can have estimated average harm as high as 17.2% and estimated differential harm as extreme as $-6.1\%$, nearly twice the value of the Democratic proposal.
-Plans with zero partisan bias or mean-median score likewise are estimated to have higher-than-average harm and to tend to differentially harm Republicans.
-
-Figure @fig-nj-pairs in the appendix expands on Figure @fig-nj-meas to also show the correlations between these various measures.
-Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
-Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats.
-Thus there exist many plans which give Democrats a higher-than-average share of the seats, but which have partisan bias and mean-median score favoring Republicans.
-New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
-Appendix @sec-app-corr records these correlations across simulated plans in eight states.
-The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
-Seats are, of course, more discrete and cannot directly identify the local areas where gains and losses are derived.
+The raw metrics can be better contextualized in comparison to the distribution across the sampled plans, something we discuss in [Appendix @sec-app-nj].
+That appendix also compares the correlations between the various metrics in the simulated sample.
+We find strong correlations of differential harm with the number of seats won by each party, while several common metrics like the mean-median gap and partisan bias are very weakly correlated with seats won.
 
@@
 
-We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.[^10]
+We next apply our proposed framework to @2022merrill, ongoing VRA litigation in which civil rights groups have argued that the existing Black majority district packs Black voters, and so a second Black opportunity-to-elect district should be drawn.
 Measuring differential harm by both race and party will allow us to cleanly separate the racial and partisan effects of the Alabama districting plan, something existing approaches are unable to do.
 
-[^10]: Relevant court filings are available at <https://redistricting.lls.edu/case/milligan-v-merrill/>.
-
-As discussed in Section @sec-racial, plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
-In *Merrill,* the plaintiff's expert, Moon Duchin, submitted four such remedial plans, one of which is shown in Figure @fig-al-maps(e).
-These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (Figure @fig-al-maps(d)), especially since the federal district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
-In Appendix @sec-app-alsampling, we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.
+Plaintiffs had to satisfy the first *Gingles* criterion by submitting one or more remedial maps demonstrating that a new Black opportunity-to-elect district could be drawn while meeting all other redistricting requirements.
+In *Allen*, the plaintiff's expert, Moon Duchin, submitted four such plans, one of which is shown in @fig-al-maps(e).
+These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (@fig-al-maps(d)), especially since the district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
+In [Appendix @sec-app-alsampling], we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.
 
@@
 
-This analysis is slightly more complicated than in New Jersey, since in order to understand the racial distribution of harm we need to estimate voting preferences by race.
-As is custom in VRA analyses, we use an ecological inference model.
-The model, detailed in Appendix @sec-app-alei, jointly estimates two turnout and vote choice by race and precinct across seven statewide elections using a setup following @king1997solution.
-For example, we model precinct $j$'s observed turnout $n_{jv}$ in election $v$ and baseline turnout rates $\turn_{jr}$ by race $r$ as \begin{align*}
-    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
-        &\sim \Binlog\qty(
-        \overbrace{\vap_j,}^{\substack{\text{voting-age}\\\text{population}}}
-        \hspace{0.5em}
-        \logit(\turn_j^\top
-        \overbrace{\vb p_j}^{\mathclap{\substack{\text{precinct racial}\\\text{proportions}}}}) +
-        \underbrace{\beta^{(t)}_v}_{\mathllap{\text{election-specific turnout shift}}}) \\
-    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
-        &\sim \mathcal{N}\underbrace{\qty(\mu^{(t)}, \Sigma^{(t)})}_{\mathrlap{
-        \text{where to cluster the precinct turnout rates}}},
-\end{align*} where $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale.
-We model Democratic support rates $\supp_{jr}$ by precinct and race similarly.
+This analysis is slightly more complicated than in New Jersey, since in order to understand the racial distribution of harm we need to estimate turnout and voting preferences by race.
+As we describe in the methods section, this is accomplished with an ecological inference model.
+The estimates confirm that voting is highly polarized by race outside of the cities of Birmingham and Montgomery.
 
-The model estimates, as expected, reveal that voting is highly polarized by race outside of the cities of Birmingham and Montgomery.
-With these estimates in hand, we can calculate the average number of voters harmed in each precinct and racial group.
-Figure @fig-al-maps(c) shows the estimated geographic distribution of harm.
-Even without the ecological estimates, it is immediately clear, with reference to the partisan and racial patterns shown in Figures @fig-al-maps(a) and (b), that harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
+With these estimates in hand, @fig-al-maps(c) shows the average number of voters harmed in each precinct, and @fig-al-maps(f) summarizes harm by party and race.
+Harm is concentrated among Black and Democratic voters in Birmingham, Montgomery, and the southern end of the Black Belt region.
 These areas are excluded from the Black-majority district under the challenged plan, but are included in a Black-majority district in the average remedial plan.
 
-This intuition is confirmed by a closer analysis, summarized in Figure @fig-al-maps(f).
-Across the state, the estimated average harm is 7.9%, which translates to around 159,000 voters.
-Of these harmed voters, around 123,000 are Black, while 36,000 are White.
-Similarly, 137,000 are Democratic, while only 22,000 are Republican.
-Of course, the total number of each of these groups of voters is very different.
-Normalizing by group sizes, the expected harm for a Black Democrat is 22%, for a Black Republican is 3.7%, for a White Democrat is 8.3%, and for a White Republican is 1.7%.
-Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have an estimated differential harm of 13.8% versus White Democrats and of 18.3% versus Black Republicans.
+Across the state, the estimated average harm is $7.9\%$.
+The expected harm for a Black Democrat is $22\%$, for a Black Republican is $3.7\%$, for a White Democrat is $8.3\%$, and for a White Republican is $1.7\%$.
+Thus while the challenged plan differentially harms both Black voters and Democratic voters, it *especially* harms Black Democrats, who have an estimated differential harm of $13.8\%$ versus White Democrats and of $18.3\%$ versus Black Republicans.
 
 These calculations are made by aggregating harm within racial groups at the precinct level, and so they automatically take into account individual voter preferences.
-Unlike in a traditional ecological/*Gingles* analysis, there is no need to identify a particular party as the uniform statewide "choice" of the minority racial group (though vote choice by race and geography must still be estimated), nor is there any cause for concern because of lower racial voting polarization in urban areas.
-The harm measure treats voters of every race and location equally, recording just their inability to elect their preferred candidate.
-Only once we have aggregated these individual harms does it become apparent that the challenged plan concentrates the harm among certain groups of voters.
+Unlike in a traditional *Gingles* analysis, there is no need to identify a particular party as the uniform statewide "choice" of the minority racial group, nor is there any cause for concern because of lower racial voting polarization in urban areas.
+The harm measure also treats voters of every race and location equally, recording just their inability to elect their preferred candidate.
+<!-- Only once we have aggregated these individual harms does it become apparent that the challenged plan concentrates the harm among certain groups of voters. -->
 
@@
 
-The proposed harm measures are particularly easy to calculate and interpret in two-party elections, since the preferences of each kind of partisan are directly observable as partisan vote shares.
-However, the same framework can be fruitfully applied to more complex settings, providing insights that go beyond those available through traditional methods.
+The proposed harm measures are particularly easy to calculate in two-party elections, since the preferences of each kind of partisan are directly observable as partisan vote shares.
+However, the same framework can be fruitfully applied to more complex settings, providing insights unavailable with traditional methods.
 
-A major question in voting rights law and research is the extent to which voting must be racially polarized before race should enter into the redistricting equation and statutory protections like the VRA should kick in [@elmendorf2016racially].
-This question is particularly acute in cities, where general elections are characterized by Democratic dominance and low levels of racial polarized voting.
-When the locus of political conflict shifts to primary elections, racial patterns are harder to observe and generalize.
-Thus there remains uncertainty about the extent that race is a factor in primary elections, and consequently the role it should play in designing urban legislative maps.
+A challenge in voting rights law and research are settings where racially polarized voting occurs in primary elections or nonpartisan races, where it is difficult to observe due to the lack of party labels.
+This is the case in many cities, where the Democratic primary often is the *de facto* general election.
+The resulting uncertainty leads to arguments about the proper role of race in drawing urban legislative districts.
+
+This section studies how the current Boston city council district boundaries differentially harm Black voters, versus a race-neutral districting process.
+In Boston, nine councilors are elected from contiguous districts, shown in @fig-boston-map alongside a demographic map, and the remaining four are elected at-large.
 
@@
 
-The city of Boston is governed by a mayor and a nonpartisan 13-member city council.
-Nine of the councilors are elected from contiguous city council districts, and the remaining four councilors are elected at-large, with all councilors serving two-year terms.
-Council district boundaries are redrawn every ten years following the decennial census, and must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^11]
-This section will study the extent to which the current district boundaries differentially harm voters of different racial groups, in comparison to a completely race-neutral districting process.
-Figure @fig-boston-map shows the racial composition of the city as well as the boundaries of the 2010 city council districts and the city's official neighborhoods.
-
-[^11]: City of Boston charter.
-    Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
-
-## Approach
-
-To apply our proposed framework to study the council district boundaries, we must be able to predict individual voting preferences under counterfactual districting plans.
-This is more difficult than in both New Jersey and Alabama, for two reasons.
+To conduct our analysis, we must be able to predict individual preferences under counterfactual districting plans.
+This is more difficult than in both New Jersey and Alabama.
 First, the city council is officially nonpartisan, and city politics are dominated by the Democratic Party, so voting choices are more complex than a simple partisan preference.
-To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election.
+<!-- To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election. -->
 Second, as in Alabama, we aim to examine the racial impact of the district system, and so must estimate these voting preferences by race.
-Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.
+<!-- Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.
+-->
 
@@
 #| fig.cap: |
-#|      Candidate factor loadings, with approximate 80\% credible intervals shown as
-#|      ellipses (left), and geographic variation in factor values by race (right).
+#|      Candidate factor loadings from the latent EI model,
+#|      with approximate 80\% credible intervals shown as ellipses (left),
+#|      and geographic variation in factor values by race (right).
 #|      Neighborhood boundaries overlaid in black.
@@
 
-To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space,[^12] and (c) voter preferences in this space by precinct and race, using precinct-level demographic data and vote totals from the March 2020 presidential primary and September 2021 mayoral preliminary elections.
-The turnout component of the model is identical to that used for Alabama.
-The core of the voter choice model is the following specification for $\pi_{jk}$, the preference of precinct $j$ for candidate $k$ on the logit scale:[^13] \begin{align*}
-    \pi_{jk} &= \underbrace{\beta^{(s)}_k}_{\substack{\text{citywide support}\\\text{for candidate}}} +
-    F_{j}^\top(\overbrace{P_j}^{\mathclap{\substack{\text{diagonal matrix of}\\\text{precinct racial proportions}}}}
-    \underbrace{\turn_j}_{\mathclap{\substack{\text{turnout rate parameter}\\\text{from turnout model}}}}
-    )^\top\vb L_k.
-\end{align*}
+To solve both of these problems, we develop a *latent factor ecological inference model*, which simultaneously estimates (a) turnout by precinct and race, (b) candidate positions in a latent two-dimensional political space, and (c) voter preferences in this space by precinct and race.
+@fig-boston-factors shows the key outputs from this model, and the methods section explains the model in more detail.
 
-[^12]: Exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+The first dimension primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
+Biden and Wu's opponent, Annissa Essaibi-George, on the right.
+The second dimension appears to capture racial differences in support---as the factor value maps in @fig-boston-factors show, minority-majority and White areas of the city take opposite values on the second dimension, demonstrating racial polarization.
 
-[^13]: For simplicity of presentation, the parameterization shown here has been altered slightly from that detailed in the appendix, though the ultimate model can be made identical after an adjustment to the prior on $\vb L_k$.
+To evaluate harm in a typical council election, we consider two scenarios, one in which all of the council races are contested along the ideological (first) dimension, and one in which the races are contested along the racial (second) dimension.
+We use a simulated set of council maps as our race-neutral counterfactual.
 
-Here, $F_j$ is a matrix of preferences by race and latent dimension within precinct $j$, so $F_{j}^\top(P_j\turn_j)^\top$ is a vector with elements containing the preferences of the precinct's voters in each latent dimension, after marginalizing out race.
-The vector $\vb L_k$ contains the loading of candidate $k$ onto the latent space: When $\vb L_k$ and $F_{j}$ are nearby in latent space, then their product will be large, and the precinct's support for the candidate, $\pi_{jk}$, will be large.
+For the first scenario, we estimate that the enacted plan differentially harms White voters at a rate of only $0.79\%$ compared to the race-neutral baseline.
+For the second scenario, we estimate that the enacted plan differentially harms White voters at a rate of $4.1\%$.
 
-Figure @fig-boston-factors shows the key outputs from this model, and Appendix @sec-app-lfei explains the model in more detail.
-As expected one dimension (the first) primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders, Elizabeth Warren, and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
-Biden and Wu's opponent, Annissa Essaibi-George, on the right.
-The second dimension appears to capture racial differences in support---as the factor value maps in Figure @fig-boston-factors show, minority-majority and White areas of the city take opposite values on the second dimension.
+These findings suggest that despite the dominance of the Democratic party in city politics, and the shared preferences between certain groups of Black and White voters, race-aware districting still serves an important role in empowering Black voters across the city to elect their chosen candidates.
+<!-- Further study could focus on the racial impact of the four at-large members in comparison to a system with 13 districts and no at-large members. -->
 
-Like all ecological inference models, ours relies on several unverifiable assumptions.
-The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
-As Appendix @sec-app-lfei discusses, these assumptions are almost certainly not *strictly* met.
-However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
-For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure @fig-boston-gen-pred in Appendix @sec-app-lfei).
-The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
-We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one out of several approaches which could provide estimated vote choices for individuals.
-Local survey data with ideological preferences and racial information would be a sufficient alternative.
-
-## Results
-
-As Figure @fig-boston-factors shows, primary voting preferences in Boston are polarized by race, especially for the second latent dimension.
-But there is significant variation within racial groups as well, such as the split between more conservative White voters in South Boston and more liberal White voters in Jamaica Plain (the first latent dimension appears to capture a progressive-moderate divide).
-It is certainly possible, then, that the particular way districts are drawn will have an effect on the ability of different racial groups to elect their candidate of choice.
-
-We consider two different scenarios: one in which all of the council races are contested along the first latent dimension, and one in which the races are contested along the second.
-Specifically, we draw voter choices between two candidates from the posterior predictive distribution of our model, setting the candidates' loadings to --1 and 1 on the chosen dimension and 0 on the other dimension.
-We draw the overall level of support from a $\Beta(15, 15)$ distribution, which means that we will study election scenarios where the overall vote share for each bloc of candidates ranges from roughly 30 to 70 percent.
-The posterior predictive draws yield counts of voters for each candidate in each precinct by race, from which we can calculate winners at the district level, under any configuration of districts.
-
-Our hypothetical elections involve only two candidates, to simplify the challenges of measuring harm with multiple candidates, over which voters may have different ranked preferences.
-However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
-One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption in the definition of harm that all candidates the voter did not vote for are interchangeable.
-
-We sample 2,000 districting plans across two independent runs of the SMC algorithm of @mccartan2020, setting constraints to favor compact plans that respect neighborhood boundaries.
-Since we do not provide any racial information to the algorithm, this set of plans forms a good approximation of a race-neutral districting process.
-
-For the first scenario, where council races are contested along the ideological axis, we estimate that the enacted plan is quite fair, differentially harming White voters at a rate of only 0.79% compared to the race-neutral baseline.
-For the second scenario, where the contests are oriented along the racial dimension, we estimate that the enacted plan differentially harms White voters at a rate of 4.1%, compared to the race-neutral baseline.
-
-These two harm estimates reflect the harm of the current boundaries compared to a race-neutral baseline.
+<!-- These two harm estimates reflect the harm of the current boundaries compared to a race-neutral baseline.
 The more policy-relevant direction might be in evaluating a change from the status quo.
@@
 For this reverse comparison, along the first dimension, we estimate that a race-neutral baseline would differentially harm Black voters at a rate of 3.1% compared to the enacted plan Along the second dimension, the differential harm to Black voters would be 1.8%.
-
-Taken together, these findings suggest that despite the dominance of the Democratic party in city politics, and the shared preferences between certain groups of Black and White voters, race-aware districting still serves an important role in empowering Black voters across the city to elect their chosen candidates.
-A switch to race-neutral boundaries would have a material impact on Black voting power.
-Further study could focus on the racial impact of the four at-large members in comparison to a system with 13 districts and no at-large members.
+-->
 
@@
 
-This paper offers a new framework for studying redistricting that brings together a focus on individual voter preferences and the causal-inference logic of counterfactuals.
+We offer a new framework for studying redistricting that brings together a focus on individual voter preferences and the causal-inference logic of counterfactuals.
 The combination of these ideas imbues the analysis of districting plans with historical understandings of legal standing and harm.
-This focus naturally leads to a new normative standard of plan fairness, which is achieved when the necessary harms of a districting plan are distributed evenly across voters.
-In other words, your personal characteristics should not impact your ability to be represented.
 
-Compared to existing alternatives, differential harm more accurately identifies partisan gerrymanders---both in sign and in existence.
-The exact same approach applies beyond partisanship and allows for the calculation of differential harms by race and other protected groups.
+A key advantage of the harm framework is that it applies without modification beyond partisanship, so that harms can be measured by race or other protected groups.
 By centering the framework on individual voters' choices, we sidestep the need to blindly associate voters' preferences with their group memberships.
-Analysts can model vote choice separately, estimate the harm to each individual, and then aggregate by any group of interest.
-As Section @sec-nj--@sec-boston showed, this opens up space for more nuanced analyses that can extend to the local level as well.
+As @sec-boston showed, this opens up space for more nuanced analyses that can extend to the local level as well.
 
-The proposed framework is far from the end of the road.
-Its flexibility comes at a cost---most notably, the need to specify a set of counterfactual districting plans, the choice of which has a significant impact on conclusions.
-Fundamentally, harm and differential harm are unable to measure plans against an absolute benchmark, the way that other measures such as partisan symmetry are able to do.
-And while modern simulation methods are a valuable starting point for defining counterfactuals, these approaches involve analyst discretion and interpretation of often-murky legal standards.
-When subjective choices are not represented consistently, different analyses may arrive at different conclusions about which voters are harmed and at what rate.
+The proposed framework is not without important limitations, however, which present opportunities for future study.
+The flexibility of choosing a model for counterfactual plans and hypothetical election outcomes also creates possibly large sensitivity to those choices.
+<!-- Fundamentally, harm and differential harm are unable to measure plans against an absolute benchmark, the way that other measures such as partisan symmetry are able to do. -->
+Simple election models are computationally tractable and require minimal data, but grossly simplify reality.
+More realistic models can add a significant burden to the calculation of our harm measures.
+And while modern simulation methods are a valuable starting point for defining counterfactual plans, these approaches involve analyst discretion and interpretation of often-murky legal standards.
 
-Modeling future elections poses another set of challenges.
-Simple models are easy to build from commonly-available data and are computationally tractable, but are gross simplifications of reality.
-More realistic models require more analyst thought and can add a significant burden to the calculation of our harm measures.
-And as discussed in Section @sec-harm, multi-candidate elections where voters have strong ordinal preferences provide further challenges that the proposed framework does not fully address.
-Each of these limitations present opportunities for future study.
+One limitation of the harm definition itself is that considers only individuals who vote; abstainers experience zero harm.
+A richer framework and the use of auxiliary survey data might allow consideration of the informal preferences of the whole population, even when those preferences are not expressed at the ballot box.
+
+A second limitation is that the harm definition ignores any preferences a voter may have between the candidates that they did not vote for.
+For example, a voter may consider candidates A and B nearly interchangeable, but may strongly dislike candidate C.
+The simple cost function implicitly assumed in our proposed measure does not differentiate between candidates B or C winning the district.
+The problem could be remedied by replacing the indicator functions in the measure with other cost functions that consider ordinal preferences.
 
@@
 
-## Estimating the Harm Measures {#sec-estimating}
+## Estimating the harm measures {#sec-estimating}
 
@@
 Multiparty or primary elections pose additional modeling challenges.
-In Section @sec-boston we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.
+In @sec-boston we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.
 
@@
 where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
+
 The baseline $\bar{s}_d$ can be established by averaging the results of all available statewide elections over the past decade or other suitable period.
+This is done at the precinct level, since this is the most granular level that districting plans are usually built from and election returns are reported at.[^14]
 Since $\bar{s}_d$ is calculated for the particular configuration of district $d$, this model takes into account the specific districting plan under consideration.
-We estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$ by fitting a linear mixed-effects model to 5,931 contested U.S. congressional elections from 1976 to 2020.
-Appendix @sec-app-elecmodel contains more details on how we estimate and apply this model.
+
+[^14]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
+    Population information comes from Census blocks, which, when populated, are practically never split.
+
+To estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$, we use historical U.S.
+congressional election returns from 1976-2020, provided by the MIT Election Data Science Lab (<https://doi.org/10.7910/DVN/IG0UN2>).
+We discard elections which were uncontested by either party, or which had more than one Democratic or Republican candidate, leaving 5,931 total elections.
+The data do not have complete coverage, and some states are missing or represented very sparsely.
+However, there are 38 states with at least 20 elections.
+
+For each election, we compute the two-party Democratic vote share.
+We fit a linear mixed-effects model to these vote shares, with an overall intercept fixed effect, and random effects for districts and election years.
+To account for the fact that district boundaries change every decade, the district identifiers are built as a concatenation of the state, district number, and redistricting.
+For example, Washington's 7th congressional district receives one random effect from 2002-2010 and a different random effect from 2012-2020, since new district boundaries were enacted in the 2010-2012 period.
+We do not account for any mid-decade redistricting which may occur as a result of litigation.
+
+The estimated standard deviations of the random effects are $\hat{\sigma}_\delta=0.0266$ for the election-year shocks and $\hat{\sigma}_\varepsilon=0.0653$ for the election-district-specific effects.
+We substitute these point estimates directly into our election model.
+While this has the advantage of simplicity, it does not account for uncertainty in the estimates.
 
@@
 
-# Connections to Current Practice for Analyzing Partisan and Racial Gerrymandering {#sec-connections}
+## Ecological inference model for Alabama {#sec-alei}
+
+To estimate voter preferences by race and precinct in Alabama, we use 2020 U.S.
+Census demographic data and election returns from seven recent statewide elections: the 2016 races for president and U.S.
+senator, the 2018 races for governor, attorney general, and secretary of state, and the 2020 races for president and U.S.
+senator.
+Our ecological inference model used to produce the estimates is broadly similar to that of @king1997solution.
+However, it jointly estimates both turnout and Democratic support across all seven elections, rather than estimating each quantity separately for each election.
+
+For precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $n_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jv}$ denote the Democratic vote share, and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Other}\}$.
+Our main parameters of interest are $\turn_{jr}$ and $\supp_{jr}$, the baseline turnout and Democratic support rate by race in the precinct.
+
+We model the observed turnout $n_{jv}$ and vote counts $s_{jk}n_{jv}$ as
+\begin{align*}
+    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
+        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
+    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
+        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
+    s_{jv}n_{jv} \mid \supp_{jr}, \turn_{jr}, \beta^{(s)}_v
+        &\sim \Binlog\qty(n_{jv}, \logit\qty(
+        \supp_j^\top \frac{\turn_j\circ\vb p_j}{\turn_j^\top\vb p_j}) + \beta^{(s)}_v) \\
+    \logit(\supp_j) \mid \mu^{(s)}, \Sigma^{(s)}
+        &\sim \Norm(\mu^{(s)}, \Sigma^{(t)}),
+\end{align*}
+where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale, and $\circ$ indicates the element-wise (Hadamard) product.
+We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior.
+We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$.
+The former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).
+Several reparametrizations are made for computational efficiency.
+The full Stan model is included in the replication code.
+
+```{r fig-alabama-ei, fig.width=6.5}
+#| fig.cap: |
+#|      Model estimates of precinct-level turnout by race (a)
+#|      and Democratic support by race (b).
+#|      Points are sized in proportion to their voting-age population,
+#|      and colored according to their Black share of the voting-age population.
+#|      All axes are transformed to a logit scale.
+include_graphics("figures/al_ei.pdf")
+```
+
+The mean support $\mu^{(s)}$ receives an informative prior based on the racially polarized voting estimates of @kuriwaki2021geography, with $\mu^{(s)}\sim\Norm(\logit((0.18, 0.90, 0.35)), (0.03^2, 0.03^2, 0.2^2))$.
+These estimates were derived through multilevel regression and poststratification from survey data, calibrated to actual election results.
+
+The model was fitted with a mean-field variational approximation for computational efficiency.
+Posterior medians of $\turn_{jr}$ and $\supp_{jr}$ are plotted in @fig-alabama-ei.
+These estimates broadly track with the findings of the voting rights expert retained in *Allen*.
+
+## Boston application methods {#sec-bos-methods}
+
+### Latent factor ecological inference model {#sec-lfei}
+
+To estimate voter preferences by race and precinct in Boston, we use 2020 U.S.
+Census demographic data and election returns from the March 2020 presidential primary and the September 2021 mayoral preliminary elections.
+Our ecological inference model used to produce the estimates expands on the model in @sec-alei, jointly estimating both turnout and vote choice in the two elections.
+However, it estimates voter preferences in a two-dimensional latent space and also estimates candidates' positions in this space.
+
+Continuing the notation from above, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $n_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jk}$ denote the vote share for candidate $k$ (who implicitly belongs to a particular election $v_k$), and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Hispanic},\text{Other}\}$ ("Other" consisting primarily of Asian voters in Boston).
+Our main parameters of interest are $\turn_{jr}$, the baseline turnout by race and precinct, $F_{jqr}$, the preferences in latent dimension $q$ by race and precinct, and $L_{kq}$, the loading of each candidate on latent dimension $q$.
+Here $q\in\{1,2\}$, since preliminary analyses found no improved fit when expanding to three dimensions.
+
+We model the observed turnout $n_{jv}$ identically as in Alabama, above:
+\begin{align*}
+    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
+        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
+    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
+        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}).
+\end{align*}
+
+Because of the latent factor model, we do not require that vote shares by race and precinct, when tallied by race, necessarily equal the observed vote share in the precinct.
+As a result we choose to model the vote shares directly, rather than the vote counts, which is more computationally efficient.
+Because each precinct has dozens or hundreds of voters, a Normal approximation to the Binomial likelihood is appropriate, and so we model the vote shares $s_{jk}$ as
+\begin{align*}
+    \logit(s_{jk}) \mid \pi_{jk}, \eps &\sim \Norm(\pi_{jk}, \frac{\eps}{1+n_{jv_k}}),
+\end{align*}
+where $\pi_{jk}$ is the preference of precinct $j$ for candidate $k$ on the logit scale, as detailed below.
+The Normal approximation variance is inversely proportional to the total number of votes in the precinct (the additional 1 is to avoid divide-by-zero errors).
+The multiplicative factor $\eps$ captures the additional variance that isn't accounted for by the latent factors.
+
+The logit candidate preference itself is modeled as
+\begin{align*}
+    \pi_{jk} &= \beta^{(s)}_k + \alpha_k
+        F_{j}^\top(\turn_j\circ \vb p_j)^\top\mathrm{diag}(\vb*\rho)\vb L_k  \\
+    F_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
+\end{align*}
+We use a two-dimensional factor model, since exploratory analyses found that three latent dimensions did not provide a substantially better fit to the data.
+Note that $F_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $F_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
+The matrix $\mathrm{diag}(\vb*\rho)$ captures the relative scale or importance of each dimension to the overall outcomes.
+The $\vb*\alpha$ parameters are "fudge factors" that adjust the scale of the deviations from each candidate's average support.
+
+We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.5^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 4.0^2)$ prior, and $\logit(\mu^{(t)})$ an i.i.d. $\Beta(10, 30)$ prior.
+We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$.
+The former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
+A larger LKJ shape parameter places more weight on correlation matrices closer to diagonal, i.e., it regularizes away from high correlations.
+The scale parameters $\vb*\rho$ have i.i.d. $\Gam(1.5, 1.5/4.0)$ priors, the fudge factors $\vb*\alpha$ have i.i.d. $\Gam(5.0, 5.0)$ priors, and $\eps$ has an $\Expo(1/4.0)$ prior.
+Overall, these priors weakly identify the location and scale of the parameters.
+Several reparametrizations were made for computational efficiency.
+The full Stan code for the model is available as part of the replication code.
+
+Like all probabilistic factor models, ours is unidentifiable due to the rotation, location, and scale invariance of the factors in the likelihood.
+To identify the location and scale, we place $\Norm(0,1)$ priors on most components of $\vb L$.
+To identify the rotation, we place more informative priors on the loadings of certain candidates.
+This was accomplished through somewhat of an interactive process, where a preliminary principal components decomposition of the precinct vote shares helped suggest the latent factors that might be found by the model.
+We chose to align the first dimension with the Sanders-Biden axis, and chose to make the second dimension roughly orthogonal along the Janey-Buttigieg axis.
+This translated to a prior of $\Norm((-1, 0), (0.1^2, 0.25^2))$ for Sanders, $\Norm((1, 0), (1.0^2, 0.5^2))$ for Biden, and $\Norm((-0.25, -1), (0.5^2, 2^2))$ for Janey.
+We additionally set the prior correlation between Sanders' and Biden's loadings to be $(-0.5, 0.8)$.
+
+The estimated loading matrix $\vb L$ is visualized on the left of @fig-boston-factors, and $F$ is visualized for Black and White voters on the right.
+
+```{r fig-boston-gen-pred, fig.width=6.5}
+#| fig.cap: |
+#|      Model predictions of general election vote shares and turnout,
+#|      compared to actual general election results.
+#|      Red reference line indicates perfect agreement.
+include_graphics("figures/boston_gen_pred.pdf")
+```
+
+Like all ecological inference models, this one makes several assumptions, verifiable and unverifiable.
+We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one out of several approaches which could provide estimated vote choices for individuals.
+Local survey data with ideological preferences and racial information would be a sufficient alternative.
+We assume no spatial correlation in residuals, when some is almost certainly present.
+We place distributional assumptions on how voter preferences by races vary across precincts (importantly, that these preferences are centered around a common location).
+And we of course impose a latent two-dimensional factor structure, which must always be a major simplification of reality.
+
+To test the suitability of the model for representing the outcomes of hypothetical future elections, we use the model fitted to the primary elections to predict the outcome of the mayoral general election in November 2021.
+Michelle Wu and Annissa Essaibi-George advanced to the general from the preliminary election.
+We fix the model's loading matrix $\vb L$ to the posterior mean of their estimated loadings.
+We draw $s_{jk}$ from a $\Beta(20\bar s, 20(1-\bar s))$ distribution, where $\bar s$ is each candidate's share of the two-candidate vote from the preliminary election (59.7% Wu -- 40.3% Essaibi George).
+
+The model's predictions (posterior medians) for the turnout and vote share in each precinct are shown in @fig-boston-gen-pred.
+Considering the differences between a primary and general election, and the additional campaign time elapsed between the elections, the predictions are remarkably accurate.
+The predicted turnout has a Spearman correlation of 0.84 with the actual turnout, and the predicted and actual vote shares have a correlation 0.87.
+The turnout and support levels are also in relatively close agreement.
+The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
+
+We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one out of several approaches which could provide estimated vote choices for individuals.
+Local survey data with ideological preferences and racial information would be a sufficient alternative.
+
+### Election scenarios
+
+We analyze two different scenarios: one in which all of the council races are contested along the first latent dimension, and one in which the races are contested along the second.
+Specifically, we draw voter choices between two candidates from the posterior predictive distribution of our model, setting the candidates' loadings to --1 and 1 on the chosen dimension and 0 on the other dimension.
+We draw the overall level of support from a $\Beta(15, 15)$ distribution, which means that we will study election scenarios where the overall vote share for each bloc of candidates ranges from roughly 30 to 70 percent.
+The posterior predictive draws yield counts of voters for each candidate in each precinct by race, from which we can calculate winners at the district level, under any configuration of districts.
+
+Our hypothetical elections involve only two candidates, to simplify the challenges of measuring harm with multiple candidates, over which voters may have different ranked preferences.
+However, our use of latent factors provides one way to extend our harm framework to the multiple-candidate setting.
+One could weight voter harm by the distance in latent space between each voter's preferred candidate and the elected candidate, thus dropping the implicit assumption in the definition of harm that all candidates the voter did not vote for are interchangeable.
+
+We sample 2,000 districting plans across two independent runs of the SMC algorithm of @mccartan2020
+In Boston, council districts must be equipopulous, be compact, and preserve neighborhood and community boundaries.[^11]
+In sampling, we therefore set constraints to favor compact plans that respect neighborhood boundaries.
+Since we do not provide any racial information to the algorithm, this set of plans forms a good approximation of a race-neutral districting process.
+
+[^11]: City of Boston charter.
+    Available at <https://www.boston.gov/sites/default/files/embed/b/boston-city-charter-2007.pdf>.
+
+# Data availability {.unnumbered}
+All data are publicly available and are included in the replication package for the manuscript.
+
+# Code availability {.unnumbered}
+All code is included in the replication package for the manuscript.
+
+::: {#refs}
+:::
+
+```{=tex}
+\vspace{1in}
+\appendix
+\renewcommand\thefigure{\thesection.\arabic{figure}}
+\renewcommand\thetable{\thesection.\arabic{table}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+```
+
+# Connections to Current Practice for Analyzing Partisan and Racial Gerrymandering {#sec-connections .appendix}
 
@@
 As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
-This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section @sec-estimating.[^4]
+This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in @sec-estimating.[^4]
 This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
@@
 Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
-Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in Section @sec-nj.
+Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in @sec-nj.
 
@@
 
-# Data availability {.unnumbered}
-All data are publicly available and are included in the replication package for the manuscript.
+# Additional New Jersey discussion {#sec-app-nj}
 
-# Code availability {.unnumbered}
-All code is included in the replication package for the manuscript.
-
-::: {#refs}
-:::
-
-```{=tex}
-\vspace{1in}
-\appendix
-\renewcommand\thefigure{\thesection.\arabic{figure}}
-\renewcommand\thetable{\thesection.\arabic{table}}
-\setcounter{figure}{0}
-\setcounter{table}{0}
-```
-# Two-party Election Model {#sec-app-elecmodel}
-
-The election model used in estimating harm is
-$$
-    s_{dv} = \bar{s}_d + \delta_v + \varepsilon_{dv},
-    \qquad \delta_v\sim\Norm(0, \sigma^2_\delta),
-    \quad \varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon).
-$$
-As noted in the main text, $\bar{s}_d$ is computed by averaging historical election returns.
-This is done at the precinct level, since this is the most granular level that districting plans are usually built from and election returns are reported at.[^14]
-
-[^14]: When district boundaries split precincts, practitioners generally impute the returns in each component of the split precinct by population-weighted interpolation.
-    Population information comes from Census blocks, which, when populated, are practically never split.
-
-To estimate $\sigma^2_\delta$ and $\sigma^2_\varepsilon$, we use historical U.S.
-congressional election returns from 1976-2020, provided by the MIT Election Data Science Lab (<https://doi.org/10.7910/DVN/IG0UN2>).
-We discard elections which were uncontested by either party, or which had more than one Democratic or Republican candidate, leaving 5,931 total elections.
-The data do not have complete coverage, and some states are missing or represented very sparsely.
-However, there are 38 states with at least 20 elections.
-
-For each election, we compute the two-party Democratic vote share.
-We fit a linear mixed-effects model to these vote shares, with an overall intercept fixed effect, and random effects for districts and election years.
-To account for the fact that district boundaries change every decade, the district identifiers are built as a concatenation of the state, district number, and redistricting.
-For example, Washington's 7th congressional district receives one random effect from 2002-2010 and a different random effect from 2012-2020, since new district boundaries were enacted in the 2010-2012 period.
-We do not account for any mid-decade redistricting which may occur as a result of litigation.
-
-The estimated standard deviations of the random effects are $\hat{\sigma}_\delta=0.0266$ for the election-year shocks and $\hat{\sigma}_\varepsilon=0.0653$ for the election-district-specific effects.
-We substitute these point estimates directly into our election model.
-While this has the advantage of simplicity, it does not account for uncertainty in the estimates.
-
-# Correlation Between Partisan Gerrymandering Metrics Across Multiple States {#sec-app-corr}
-
-Figure @fig-nj-pairs expands on Figure @fig-nj-meas in the main text with pairwise scatter plots and correlation summaries.
+This section compares the partisan metrics from @sec-nj across the simulated baseline.
+@fig-nj-pairs expands on @fig-nj-meas in the main text with pairwise scatter plots and correlation summaries.
 
@@
 
-Table @tab:corr below reports the correlation between differential harm and the partisan gerrymandering metrics used in Section @sec-nj for eight states which cover a wide range of partisan leans and number of seats:
+The Democratic proposal is expected to produce an average of 8.55 Democratic seats under our election model, more than 79% of all sample plans.
+The Republican proposal is an outlier, yielding fewer seats on average than every single sampled plan.
+This pattern is repeated for all of the four common gerrymandering measures, where the Republican plan is a clear outlier (all $p=0.0002$, except $p=0.002$ for partisan bias) while the Democratic plan is not ($p=0.34$ for the efficiency gap, $0.50$ for declination, $0.88$ for partisan bias, and $0.17$ for the mean-median score).
 
--   Utah, with 4 districts and a Democratic vote share baseline of 33%
--   Mississippi, with 4 districts and a Democratic vote share baseline of 43%
--   Iowa, with 4 districts and a Democratic vote share baseline of 46%
--   North Carolina, with 14 districts and a Democratic vote share baseline of 50%
--   Wisconsin, with 8 districts and a Democratic vote share baseline of 51%
--   New Jersey, with 12 districts and a Democratic vote share baseline of 57%
--   Massachusetts, with 9 districts and a Democratic vote share baseline of 61%
--   California, with 52 districts and a Democratic vote share baseline of 65%
+The estimate of differential harm already uses the simulation output, and so comparing the proposals to the sampling distribution is not strictly required.
+However, it is notable that while the Democratic proposal has a more extreme estimate of differential partisan harm (1.6 percentage points higher), its value of $-3.7\%$ is not an outlier with respect to the distribution of differential harm for a randomly sampled plan ($p=0.43$, using the same set of samples as a counterfactual), while the Republican plan is ($p=0.0002$).
+In other words, a *typical* nonpartisan districting plan is not necessarily a *fair* one, even in comparison to other similarly generated districting plans.
+The sampled plan which minimizes estimated differential harm yields 8.18 expected Democratic seats, compared to a sample average of 8.41 seats, and the 6.8 seats that would be required for an outcome completely proportional to the parties' statewide vote share.
+
+Both proposals are estimated to harm the average voter at a higher rate than most of the sampled plans, with the Republican plan again more extreme in this regard ($p=0.0008$ versus $p=0.061$ for the Democratic plan).
+The plans which minimize estimated total harm are also those which have near-zero estimated differential partisan harm.
+Thus, at least in New Jersey, overall harm reduction is consistent with a fair map on a partisan basis.
+Minimizing bias with other measures, however, could lead one astray: plans which score an efficiency gap of zero can have estimated average harm as high as 17.2% and estimated differential harm as extreme as $-6.1\%$, nearly twice the value of the Democratic proposal.
+Plans with zero partisan bias or mean-median score likewise are estimated to have higher-than-average harm and to tend to differentially harm Republicans.
+
+@fig-nj-pairs appendix expands on @fig-nj-meas to also show the correlations between these various measures.
+Expected Democratic seats, the differential harm estimates, the efficiency gap, and declination are all moderately correlated with each other.
+Similarly, partisan bias and the mean-median score are moderately correlated, but very weakly correlated with the expected number of Democratic seats.
+Thus there exist many plans which give Democrats a higher-than-average share of the seats, but which have partisan bias and mean-median score favoring Republicans.
+New Jersey is just one state, and while some of the correlations between measures here are found in other states, some are not.
+[Appendix @sec-app-corr] records these correlations across simulated plans in eight states.
+The most consistent correlation is between estimated differential partisan harm and the expected number of seats won by each party, which averages 0.79 in the eight states we study.
+Seats are, of course, more discrete and cannot directly identify the local areas where gains and losses are derived.
+
+# Correlation Between Partisan Gerrymandering Metrics Across Multiple States {#sec-app-corr}
+
+@tbl-corr reports the correlation between differential harm and the partisan gerrymandering metrics used in @sec-nj for eight states which cover a wide range of partisan leans and number of seats:
+
+-   Utah, with 4 districts and a Democratic vote share baseline of $33$%
+-   Mississippi, with 4 districts and a Democratic vote share baseline of $43$%
+-   Iowa, with 4 districts and a Democratic vote share baseline of $46$%
+-   North Carolina, with 14 districts and a Democratic vote share baseline of $50$%
+-   Wisconsin, with 8 districts and a Democratic vote share baseline of $51$%
+-   New Jersey, with 12 districts and a Democratic vote share baseline of $57$%
+-   Massachusetts, with 9 districts and a Democratic vote share baseline of $61$%
+-   California, with 52 districts and a Democratic vote share baseline of $65$%
 
@@
 The simulations are provided by the 50-State Simulation Project [@mccartan2022fiftystates].
-In particular, the New Jersey simulations are not the same as those used in Section @sec-nj.
+In particular, the New Jersey simulations are not the same as those used in @sec-nj.
+
+::: {#tbl-corr}
 
 ```{=tex}
-\begin{table}[ht!]
-\centering
+\begin{center}
 \input{figures/corr_table.tex}
-\vspace{0.1in}
-\caption{Correlations between five gerrymandering metrics and partisan differential harm, for simulated districting plans from eight states.
-Declination is NA for Massachusetts since there are no Republican-leaning districts at baseline under any simulated plan.}
-\label{tab:corr}
-\end{table}
-```
-# Ecological Inference Model for Alabama {#sec-app-alei}
-
-To estimate voter preferences by race and precinct in Alabama, we use 2020 U.S.
-Census demographic data and election returns from seven recent statewide elections: the 2016 races for president and U.S.
-senator, the 2018 races for governor, attorney general, and secretary of state, and the 2020 races for president and U.S.
-senator.
-Our ecological inference model used to produce the estimates is broadly similar to that of @king1997solution.
-However, it jointly estimates both turnout and Democratic support across all seven elections, rather than estimating each quantity separately for each election.
-
-For precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $n_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jv}$ denote the Democratic vote share, and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Other}\}$.
-Our main parameters of interest are $\turn_{jr}$ and $\supp_{jr}$, the baseline turnout and Democratic support rate by race in the precinct.
-
-We model the observed turnout $n_{jv}$ and vote counts $s_{jk}n_{jv}$ as
-\begin{align*}
-    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
-        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
-    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
-        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}) \\
-    s_{jv}n_{jv} \mid \supp_{jr}, \turn_{jr}, \beta^{(s)}_v
-        &\sim \Binlog\qty(n_{jv}, \logit\qty(
-        \supp_j^\top \frac{\turn_j\circ\vb p_j}{\turn_j^\top\vb p_j}) + \beta^{(s)}_v) \\
-    \logit(\supp_j) \mid \mu^{(s)}, \Sigma^{(s)}
-        &\sim \Norm(\mu^{(s)}, \Sigma^{(t)}),
-\end{align*}
-where the $\vb*\beta^{(t)}$ and $\vb*\beta^{(s)}$ model election-specific logit shifts, $\Binlog$ indicates the Binomial distribution with the probability parameter specified on the logit scale, and $\circ$ indicates the element-wise (Hadamard) product.
-We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.6^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 0.4^2)$ prior, and $\mu^{(t)}$ an i.i.d. $\Norm(0, 1.5^2)$ prior.
-We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(s)}$.
-The former received $\Gam(1.5, 1.5/0.5)$ and $\Gam(1.5, 1.5/0.25)$ priors for turnout and support, while the latter received uniform priors (i.e., LKJ priors with shape parameter 1).
-Several reparametrizations are made for computational efficiency.
-The full Stan model is included in the replication code.
-
-```{r fig-alabama-ei, fig.width=6.5}
-#| fig.cap: |
-#|      Model estimates of precinct-level turnout by race (a)
-#|      and Democratic support by race (b).
-#|      Points are sized in proportion to their voting-age population,
-#|      and colored according to their Black share of the voting-age population.
-#|      All axes are transformed to a logit scale.
-include_graphics("figures/al_ei.pdf")
+\end{center}
 ```
 
-The mean support $\mu^{(s)}$ receives an informative prior based on the racially polarized voting estimates of @kuriwaki2021geography, with $\mu^{(s)}\sim\Norm(\logit((0.18, 0.90, 0.35)), (0.03^2, 0.03^2, 0.2^2))$.
-These estimates were derived through multilevel regression and poststratification from survey data, calibrated to actual election results.
-
-The model was fitted with a mean-field variational approximation for computational efficiency.
-Posterior medians of $\turn_{jr}$ and $\supp_{jr}$ are plotted in Figure @fig-alabama-ei.
-These estimates broadly track with the findings of the voting rights expert retained in *Merrill*.
+Correlations between five gerrymandering metrics and partisan differential harm, for simulated districting plans from eight states.
+Declination is NA for Massachusetts since there are no Republican-leaning districts at baseline under any simulated plan.
+:::
 
@@
 
-Here, we perform the same analysis as in Section @sec-alabama, but using a randomly sampled set of plans as a counterfactual, rather than the four plans used in litigation.
+Here, we perform the same analysis as in @sec-alabama, but using a randomly sampled set of plans as a counterfactual, rather than the four plans used in litigation.
 Specifically, we use 2,000 plans provided by the 50-State Simulation Project [@mccartan2022fiftystates].
@@
 ```
-
-Figure @fig-alabama-harm-sample summarizes the results, which are mostly similar to the results in in Section @sec-alabama.
+@fig-alabama-harm-sample summarizes the results, which are mostly similar to the results in in @sec-alabama.
 Across the state, the estimated average harm is 6.9%, which translates to around 144,000 voters.
@@
 However, it it especially harms Black Republicans, who have an estimated differential harm of 7.2% versus Black Democrats and of 11.6% versus White Republicans.
-
-# Latent Factor Ecological Inference Model for Boston Primary Elections {#sec-app-lfei}
-
-To estimate voter preferences by race and precinct in Boston, we use 2020 U.S.
-Census demographic data and election returns from the March 2020 presidential primary and the September 2021 mayoral preliminary elections.
-Our ecological inference model used to produce the estimates expands on the model in Appendix @sec-app-alei, jointly estimating both turnout and vote choice in the two elections.
-However, it estimates voter preferences in a two-dimensional latent space and also estimates candidates' positions in this space.
-
-Continuing the notation from above, for precinct $j$ in election $v$, let $\vap_j$ denote the voting-age population, $n_{jv}$ denote the total number of votes cast for major-party candidates, $s_{jk}$ denote the vote share for candidate $k$ (who implicitly belongs to a particular election $v_k$), and $p_{jr}$ denote the fraction of the voting-age population belonging to racial group $r$, where here $r\in\{\text{White},\text{Black},\text{Hispanic},\text{Other}\}$ ("Other" consisting primarily of Asian voters in Boston).
-Our main parameters of interest are $\turn_{jr}$, the baseline turnout by race and precinct, $F_{jqr}$, the preferences in latent dimension $q$ by race and precinct, and $L_{kq}$, the loading of each candidate on latent dimension $q$.
-Here $q\in\{1,2\}$, since preliminary analyses found no improved fit when expanding to three dimensions.
-
-We model the observed turnout $n_{jv}$ identically as in Alabama, above:
-\begin{align*}
-    n_{jv} \mid \turn_{jr}, \beta^{(t)}_v
-        &\sim \Binlog\qty(\vap_j, \logit(\turn_j^\top\vb p_j) + \beta^{(t)}_v) \\
-    \logit(\turn_j) \mid \mu^{(t)}, \Sigma^{(t)}
-        &\sim \Norm(\mu^{(t)}, \Sigma^{(t)}).
-\end{align*}
-
-Because of the latent factor model, we do not require that vote shares by race and precinct, when tallied by race, necessarily equal the observed vote share in the precinct.
-As a result we choose to model the vote shares directly, rather than the vote counts, which is more computationally efficient.
-Because each precinct has dozens or hundreds of voters, a Normal approximation to the Binomial likelihood is appropriate, and so we model the vote shares $s_{jk}$ as
-\begin{align*}
-    \logit(s_{jk}) \mid \pi_{jk}, \eps &\sim \Norm(\pi_{jk}, \frac{\eps}{1+n_{jv_k}}),
-\end{align*}
-where $\pi_{jk}$ is the preference of precinct $j$ for candidate $k$ on the logit scale, as detailed below.
-The Normal approximation variance is inversely proportional to the total number of votes in the precinct (the additional 1 is to avoid divide-by-zero errors).
-The multiplicative factor $\eps$ captures the additional variance that isn't accounted for by the latent factors.
-
-The logit candidate preference itself is modeled as
-\begin{align*}
-    \pi_{jk} &= \beta^{(s)}_k + \alpha_k
-        F_{j}^\top(\turn_j\circ \vb p_j)^\top\mathrm{diag}(\vb*\rho)\vb L_k  \\
-    F_{jq} \mid \Sigma^{(p)} &\sim \Norm(0, \Sigma^{(p)}).
-\end{align*}
-Note that $F_j$ is a $q\times r$ matrix of preferences by race and dimension within precinct $j$, and $\turn_j\circ \vb p_j$ is a vector of the percentages of the total VAP in the precinct that belongs to voters of each race, so $F_{j}^\top(\turn_j\circ \vb p_j)$ is a vector with $q$ elements containing the preferences by dimension of the precinct's voters, after marginalizing out race.
-The matrix $\mathrm{diag}(\vb*\rho)$ captures the relative scale or importance of each dimension to the overall outcomes.
-The $\vb*\alpha$ parameters are "fudge factors" that adjust the scale of the deviations from each candidate's average support.
-
-We give $\vb*\beta^{(t)}$ an i.i.d. $\Norm(0, 0.5^2)$ prior,and $\vb*\beta^{(s)}$ an i.i.d. $\Norm(0, 4.0^2)$ prior, and $\logit(\mu^{(t)})$ an i.i.d. $\Beta(10, 30)$ prior.
-We set priors independently on the variance and correlation components of $\Sigma^{(t)}$ and $\Sigma^{(p)}$.
-The former has $\Gam(1.5, 1.5/0.25)$ and $\Gam(1.5, 1.5/1.0)$ priors for turnout and support, while the latter have LKJ priors with shape parameters 4.0 and 20.0.
-A larger LKJ shape parameter places more weight on correlation matrices closer to diagonal, i.e., it regularizes away from high correlations.
-The scale parameters $\vb*\rho$ have i.i.d. $\Gam(1.5, 1.5/4.0)$ priors, the fudge factors $\vb*\alpha$ have i.i.d. $\Gam(5.0, 5.0)$ priors, and $\eps$ has an $\Expo(1/4.0)$ prior.
-Overall, these priors weakly identify the location and scale of the parameters.
-Several reparametrizations were made for computational efficiency.
-The full Stan code for the model is available as part of the replication code.
-
-Like all probabilistic factor models, ours is unidentifiable due to the rotation, location, and scale invariance of the factors in the likelihood.
-To identify the location and scale, we place $\Norm(0,1)$ priors on most components of $\vb L$.
-To identify the rotation, we place more informative priors on the loadings of certain candidates.
-This was accomplished through somewhat of an interactive process, where a preliminary principal components decomposition of the precinct vote shares helped suggest the latent factors that might be found by the model.
-We chose to align the first dimension with the Sanders-Biden axis, and chose to make the second dimension roughly orthogonal along the Janey-Buttigieg axis.
-This translated to a prior of $\Norm((-1, 0), (0.1^2, 0.25^2))$ for Sanders, $\Norm((1, 0), (1.0^2, 0.5^2))$ for Biden, and $\Norm((-0.25, -1), (0.5^2, 2^2))$ for Janey.
-We additionally set the prior correlation between Sanders' and Biden's loadings to be $(-0.5, 0.8)$.
-
-The estimated loading matrix $\vb L$ is visualized on the left of Figure @fig-boston-factors, and $F$ is visualized for Black and White voters on the right.
-
-```{r fig-boston-gen-pred, fig.width=6.5}
-#| fig.cap: |
-#|      Model predictions of general election vote shares and turnout,
-#|      compared to actual general election results.
-#|      Red reference line indicates perfect agreement.
-include_graphics("figures/boston_gen_pred.pdf")
-```
-
-Like all ecological inference models, this one makes several assumptions, verifiable and unverifiable.
-We assume no spatial correlation in residuals, when some is almost certainly present.
-We place distributional assumptions on how voter preferences by races vary across precincts (importantly, that these preferences are centered around a common location).
-And we of course impose a latent two-dimensional factor structure, which must always be a major simplification of reality.
-
-To test the suitability of the model for representing the outcomes of hypothetical future elections, we use the model fitted to the primary elections to predict the outcome of the mayoral general election in November 2021.
-Michelle Wu and Annissa Essaibi-George advanced to the general from the preliminary election.
-We fix the model's loading matrix $\vb L$ to the posterior mean of their estimated loadings.
-We draw $s_{jk}$ from a $\Beta(20\bar s, 20(1-\bar s))$ distribution, where $\bar s$ is each candidate's share of the two-candidate vote from the preliminary election (59.7% Wu -- 40.3% Essaibi George).
-
-The model's predictions (posterior medians) for the turnout and vote share in each precinct are shown in Figure @fig-boston-gen-pred.
-Considering the differences between a primary and general election, and the additional campaign time elapsed between the elections, the predictions are remarkably accurate.
-The predicted turnout has a Spearman correlation of 0.84 with the actual turnout, and the predicted and actual vote shares have a correlation 0.87.
-The turnout and support levels are also in relatively close agreement.

### 2025-08-25 paper/harm.qmd
update intro

@@
 Second, the proposed framework considers *counterfactual* districting plans, not just a single enacted plan.
-Many factors influence the qualities of a districting plan, and without a well-defined comparison, it is impossible to differentiate which qualities, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain @chen:rodd:13.
-For example, in Massachusetts in 2010, it was impossible to draw even one majority-Republican seat, despite 32% of voters preferring Republicans @duch:18.
+While all measures can be used with counterfactuals, our approach uses them within the definition of the measure itself.
+Many factors influence the qualities of a districting plan, and without a well-defined comparison, it is impossible to differentiate which qualities, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
+For example, in Massachusetts in 2010, it was impossible to draw even one majority-Republican seat, despite 32% of voters preferring Republicans [@duch:18].
 Thus the 9-0 Democratic Congressional delegation reflects the homogeneity of the state's political geography, not partisan gerrymandering.
 Many existing districting metrics rely on *implicit* counterfactuals, which may be irrelevant or infeasible, as we discuss in [Appendix @sec-connections].
-By making the comparison set explicit, we can better reflect legal and geographic realities, while making the interpretation of our measures clearer.
+By making the comparison set explicit, our approach improves on existing work by better reflecting legal and geographic realities, while making the interpretation of our measures clearer.
 
@@
 Once measured for each voter, researchers can examine the distribution of harm by geography, partisanship, race, or other factors.
+This combination of individual-level measurement and counterfactual comparison is the key innovation of our framework.
 @sec-motivation gives a simple example of this idea; @sec-indiv-harm formalizes it.

### 2025-08-25 paper/harm.qmd
revisions

--- 2025-07-30 paper/harm.qmd
@@
 In *Allen*, the plaintiff's expert, Moon Duchin, submitted four such plans, one of which is shown in @fig-al-maps(e).
-These plans form a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (@fig-al-maps(d)), especially since the district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
+A uniform distribution over these plans forms a natural counterfactual $\tilde Q$ to use in analyzing the challenged districting plan (@fig-al-maps(d)), especially since the district court ruled in favor of the plaintiffs and found that their remedial plans satisfied the first *Gingles* criterion.
 In [Appendix @sec-app-alsampling], we repeat the analysis performed here on a different counterfactual consisting of plans sampled with a redistricting simulation algorithm; the results are broadly similar.
@@
 This is more difficult than in both New Jersey and Alabama.
-First, the city council is officially nonpartisan, and city politics are dominated by the Democratic Party, so voting choices are more complex than a simple partisan preference.
+First, the lack of party labels in city council races means that voting choices are more complex than a simple partisan preference.
 <!-- To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election. -->
@@
 @fig-boston-factors shows the key outputs from this model, and the methods section explains the model in more detail.
+This is a different modeling approach than the stochastic uniform swing model used in the previous two applications, but the harm framework is designed to accommodate such modeling flexibility.
 
-The first dimension primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
+The first dimension estimated by the model primarily aligns with a traditional liberal-conservative spectrum, with progressives like Bernie Sanders and current Boston mayor Michelle Wu on the left, and centrists like Joseph R.
 Biden and Wu's opponent, Annissa Essaibi-George, on the right.
@@
 
-Motivated primarily by computational convenience and simplicity, we adopt the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as
+Motivated primarily by computational convenience and simplicity, we recommend for most cases the stochastic uniform partisan swing model of @gelman1994unified, modeling the Democratic two-party vote share in district $d$ and election $v$ as
 $$
@@
 where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
+This can be thought of as a hiererchical extension of the classic uniform swing model.
 
@@
 For example, one could operate on the logit scale, model incumbency effects, or include additional data from local elections.
+
+However, the simple model presented here does a reasonable job of capturing the distribution in future elections.
+@kenny2023widespread (Appendix D) validated the model in a similar redistricting context and found it to perform well.
 
@@
 Finally, it is worth comparing the proposed measures to the increasingly common application redistricting simulation techniques to cases of partisan gerrymandering.
-Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics.
+Standard practice in such litigation, for example, is to calculate a battery of gerrymandering metrics on the simulated plans, and show that the challenged plan is an outlier on all or most of the metrics: see, e.g., @2019rucho (report of Dr.
+Jonathan Mattingly) and @2018league (report of Dr.
+Wesley Pegden).
 While this can help contextualize the challenged plan, it is not foolproof---if the underlying metric is misleading, it will still be misleading when applied to thousands of simulated plans.

### 2025-08-25 paper/harm.qmd
intro

@@
 This unifies gerrymandering analyses across contexts and reflects the goal that elected officials represent the preferences of the voting public [see also @brunell2010].
+Few existing measures, with the exception of partisan disclocation [@defoeubarodd21], are focused on the invididual voter this way.
 
 Second, the proposed framework considers *counterfactual* districting plans, not just a single enacted plan.
-While all measures can be used with counterfactuals, our approach uses them within the definition of the measure itself.
 Many factors influence the qualities of a districting plan, and without a well-defined comparison, it is impossible to differentiate which qualities, like its partisan performance, are consequences of laws and political geography or result from improper manipulation of the system for political gain [@chen:rodd:13].
-For example, in Massachusetts in 2010, it was impossible to draw even one majority-Republican seat, despite 32% of voters preferring Republicans [@duch:18].
-Thus the 9-0 Democratic Congressional delegation reflects the homogeneity of the state's political geography, not partisan gerrymandering.
-Many existing districting metrics rely on *implicit* counterfactuals, which may be irrelevant or infeasible, as we discuss in [Appendix @sec-connections].
+For example, Massachusetts' 2010 districts had an efficiency gap that would normally qualify as a Democratic gerrymander, but was in fact due to the homogeneity of the state's political geography [@duch:18].
+While consideration of counterfactual plans is now commonplace with redistricting simulation algorithms [e.g., @deford2019], our approach is different in that it incorporates them within the definition of the measure itself.
+As far as districting metrics themselves, many rely on *implicit* counterfactuals, which may be irrelevant or infeasible, as we discuss in [Appendix @sec-connections].
 By making the comparison set explicit, our approach improves on existing work by better reflecting legal and geographic realities, while making the interpretation of our measures clearer.
 
-The combination of individual voter choices and counterfactual districting plans defines *individual harm*: informally, a voter is harmed if they would be represented by their candidate of choice under a counterfactual plan.
+The *combination* of individual voter choices and counterfactual districting plans is the key innovation of our framework of *individual harm*: informally, a voter is harmed if they would be represented by their candidate of choice under a counterfactual plan.
 Once measured for each voter, researchers can examine the distribution of harm by geography, partisanship, race, or other factors.
-This combination of individual-level measurement and counterfactual comparison is the key innovation of our framework.
 @sec-motivation gives a simple example of this idea; @sec-indiv-harm formalizes it.

### 2025-08-25 paper/harm.qmd
edits

@@
 Consider *Minissouri*, a stylized representation of Missouri which is shown in @fig-minissouri.
-Its total population is 180, and it is divided into three districts with population between 54 and 66.
-Minissouri has two cities and a substantial rural population with a partisan breakdown of 55.6% R to 44.4% D.
-Voters' preferences are assumed to be stable across elections, and turnout is 100%.
+Minissouri has two cities and a substantial rural population with a partisan breakdown of 55.6% R to 44.4% D, and is divided into three districts.
+<!-- Voters' preferences are assumed to be stable across elections, and turnout is 100%. -->
 In addition to partisan affiliation (red or blue), each voter also belongs to a racial group (squares or circles).
@@
 
-After identifying which voters are harmed by plan (a) compared to counterfactual plan (b), it is natural to examine the overall composition of these harmed voters.
+After identifying which voters are harmed by plan (a) compared to counterfactual (b), it is natural to examine the overall composition of these harmed voters.
 Of the 46 harmed voters, 44 are Republicans and only 2 are Democrats.
@@
 We call the gap between these two rates the *differential harm*, and propose it as a general way to account for the impact of a districting plan on particular groups.
-Here, then the partisan differential harm of plan (a) is $-41.5\%$, meaning that the average Republican voter in Mininois is $41.5$pp less likely to be represented by a Republican than the average Democrat is to be represented by a Democrat, *compared to plan (b)*.
+Here, then the partisan differential harm is $-41.5\%$, meaning that the average Republican voter is $41.5$pp less likely to be represented by a Republican than the average Democrat is to be represented by a Democrat, *compared to plan (b)*.
 
@@
 Plan (a) harms 13 circle voters and 33 square voters---a differential racial harm of $-20\%$.
-This means square voters are $20$pp more likely to be harmed than circle voters.
+<!-- This means square voters are $20$pp more likely to be harmed than circle voters. -->
 While this comparison alone cannot establish whether the drawers of plan (a) intended to harm square voters versus maximize Democratic representation, it quantifies the impact of plan (a) on these voters versus the alternative plan (b), on the same scale as differential partisan harm.
+
+We stress that the results here are tied the specific counterfactual plan (b).
+Indeed, this is a feature of the harm framework: different research questions may require different counterfactuals.
+We illustrate this aspect of the framework throughout the applications below.
 
@@
 
-To provide a neutral comparison baseline, we generate 10,000 districting plans across using the SMC algorithm of [@mccartan2020; implemented in @redist] setting constraints to favor compact plans that respect county lines.
+For differential harm to measure gerrymandering, a counterfactual $\tilde Q$ drawn without partisan intent is needed.
+To do so, we generate 10,000 districting plans without any partisan data, using the SMC algorithm of [@mccartan2020; implemented in @redist] setting constraints to favor compact plans that respect county lines.
 @fig-nj-maps(d) shows one such plan.
@@
 To evaluate harm in a typical council election, we consider two scenarios, one in which all of the council races are contested along the ideological (first) dimension, and one in which the races are contested along the racial (second) dimension.
-We use a simulated set of council maps as our race-neutral counterfactual.
+Matching the research question, our counterfactual plans are drawn from a race-blind districting simulation.
 

### 2025-08-25 paper/harm.qmd
more dislocation discussion

--- 2025-08-25 paper/harm.qmd
@@
 
-Most of the other metrics, with the notable exception of the mean-median measure, agree with differential harm that each party's proposal favors itself.
+Most of the other metrics, with the notable exceptions of the mean-median and partisan dislocation measure, agree with differential harm that each party's proposal favors itself.
 The efficiency gap and partisan bias estimate give roughly equal scores in absolute value to the Democratic and Republican plans, in comparison to differential harm and declination, which identify the Democratic proposal as slightly more extreme.
@@
 
-Since none of the existing measures are centered on individual voters, they offer far less granular insights than the proposed harm framework, even in the case of studying partisan gerrymandering.
+A quite different approach to measuring partisan gerrymandering is the partisan dislocation [@defoeubarodd21].
+This measure uses a $k$-nearest-neighbor approach, where $k$ is the size of a district, to identify when a voter's district is politically different from their assigned district.
+This can be repeated for  voter in a state or region, demonstrating at a local level when people been placed in a district which is atypical for who their nearest neighbors are.
+When aggregated to the precinct-level, this can produce local measures of partisan gerrymandering, similar to our proposed differential partisan harm measure.
+
+Since none of the existing measures, other than partisan dislocation, are centered on individual voters, they offer far less granular insights than the proposed harm framework, even in the case of studying partisan gerrymandering.
 For example, differential harm can be estimated at the precinct level and then plotted on a map to show not just *which party* is harmed by a gerrymander, but *where* this harm occurs.
@@
 
-The Democratic proposal is expected to produce an average of 8.55 Democratic seats under our election model, more than 79% of all sample plans.
+The Democratic proposal is expected to produce an average of 3.45 Republican seats under our election model, more than 79% of all sample plans.
 The Republican proposal is an outlier, yielding fewer seats on average than every single sampled plan.
 This pattern is repeated for all of the four common gerrymandering measures, where the Republican plan is a clear outlier (all $p=0.0002$, except $p=0.002$ for partisan bias) while the Democratic plan is not ($p=0.34$ for the efficiency gap, $0.50$ for declination, $0.88$ for partisan bias, and $0.17$ for the mean-median score).
+
+The partisan dislocation measure [@defoeubarodd21] shows that the Democratic plan is typical ($p=0.24$) while the Republican plan is an outlier ($p=0.01$).
+We use the vote-weighted measure of the average dislocation to produce a statewide summary.
+We find that the sign is incorrect for the Democratic plan, which has a positive sign.
+As the sign is positive for the simulations as well, this difference may be related to the fact that this measure uses the $k$ nearest neighbors, which is not uniform over the voters in the state.
+Implicitly, voters in dense, urban areas will be counterd at a higher rate than those in rural areas.
+
+Note that the Ranked Marginal Deviation [@herschlag2020quantifying] is not a signed measure.
+All scores are at least zero, with low scores indicating that a plan is similar to the simulated plans.
+In this analysis, the Democratic plan sits within the range of the sampled plans, while the Republican plan is an outlier ($p=0.0002$).
 

### 2025-08-25 paper/harm.qmd
more appx

@@
 
+Even when the choice of constraints maps onto legal requirements, it is important to ensure that the counterfactual plans are similar to the enacted plan on discretionary criteria as well.
+For example, a mapdrawer may weakly respect something like compactness or respecting county boundaries without being required to do so.
+In this case, if these optional constraints are not matched in the counterfactual, it will be unclear whether any estimated harm is due to *actual* biases in a plan or the mismatch in criteria.
+As such, it is important to not just ensure that any counterfactual samples use methods with known sampling properties, but also that the sampled plans are similar to the enacted plan on these discretionary criteria.
+
 ### Modeling Election Outcomes

### 2025-08-25 paper/harm.qmd
discsussion sentence
edits

--- 2025-08-25 paper/harm.qmd
@@
 While consideration of counterfactual plans is now commonplace with redistricting simulation algorithms [e.g., @deford2019], our approach is different in that it incorporates them within the definition of the measure itself.
-As far as districting metrics themselves, many rely on *implicit* counterfactuals, which may be irrelevant or infeasible, as we discuss in [Appendix @sec-connections].
+Some existing districting rely on *implicit* counterfactuals, which may be irrelevant or infeasible, as we discuss in [Appendix @sec-connections].
 By making the comparison set explicit, our approach improves on existing work by better reflecting legal and geographic realities, while making the interpretation of our measures clearer.
@@
 The problem could be remedied by replacing the indicator functions in the measure with other cost functions that consider ordinal preferences.
+For example, in our Boston example, the distance in latent space between a voter's ideal point and the winning candidate's position could be used.
 
@@
 While this has the advantage of simplicity, it does not account for uncertainty in the estimates.
+
+One implication of the model presented here is that it rules out any mechanism whereby voters reward or punish the parties based on the plan enacted: $q$ itself does not show up in the model.
+Indirect effects, such as $q$ affecting the type of candidates who run, and therefore the vote share each party wins, are also ruled out, or at least on average: the residual $\varepsilon_{dv}$ captures candidate effects broadly, but it is assumed to be drawn independently of $q$.
+To the extent that a particular $q$ affects voter turnout or vote choice in these ways, the simple model presented here will not capture those effects.
 

### 2025-08-25 paper/harm.qmd

@@
 And while modern simulation methods are a valuable starting point for defining counterfactual plans, these approaches involve analyst discretion and interpretation of often-murky legal standards.
+Incorrect or incomplete interpretation of these standards can negatively influence the validity of the counterfactual plans.
 
@@
 Nonetheless, sampling methods are widely used in practice [see e.g., @2019rucho; @2018league; @2022merrill; @becker2021computational; @chen:rodd:13].
+
+Even when the choice of constraints maps onto legal requirements, it is important to ensure that the counterfactual plans are similar to the enacted plan on discretionary criteria as well.
+For example, a mapdrawer may weakly respect something like compactness or respecting county boundaries without being required to do so.
+In this case, if these optional constraints are not matched in the counterfactual, it will be unclear whether any estimated harm is due to *actual* biases in a plan or the mismatch in criteria.
+As such, it is important to not just ensure that any counterfactual samples use methods with known sampling properties, but also that the sampled plans are similar to the enacted plan on these discretionary criteria.
 

### 2025-08-25 paper/harm.qmd
remove use of \indep
cuts

@@
             blinded: true
+        keep-tex: true
         fontsize: 11pt
@@
 The academic study of gerrymandering has not reflected this diversity.
-Nearly all research has been on partisan gerrymandering, and has centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @gelman1994unified; @thomas2013estimating; @katz2020].
+Nearly all research has been on partisan gerrymandering, and has centered on how districting plans treat political parties, as captured in the *seats-votes curve* or notions of *partisan symmetry* [@tufte1973pb; @grofman1983measures; @king1987democratic; @king1989representation; @thomas2013estimating; @katz2020].
 A large number of formulas for quantifying partisan gerrymanders have also been developed [@stephanopoulos2015; @mcdonald2015unfair; @warrington18].
@@
 The harm framework has important similarities with the common-law approach to voting rights litigation.
-To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98].
-This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
+To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [e.g., @1962baker; @issakarl98].
+This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [e.g., @2004vieth].
 Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
@@
 
-Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$, as $\ind\{v(i) \neq \win_i(v, q)\} \neq \ind\{v(i) \neq \win_i(v, \tilde q)\}$.
-Exchanging the role of the enacted plan and the counterfactual will not simply flip the direction of our conclusion.
-
-In general, $q$ harms some individuals versus $\tilde q$, and $\tilde q$ harms a distinct, though not necessarily disjoint, set of individuals, versus $q$.
+Notice the fundamental asymmetry in this definition: $h_i(v,q,\tilde q)\neq h_i(v,\tilde q,q)$, since $\ind\{v(i) \neq \win_i(v, q)\} \neq \ind\{v(i) \neq \win_i(v, \tilde q)\}$.
+In general, $q$ harms some individuals versus $\tilde q$, and $\tilde q$ harms a distinct set of individuals versus $q$.
 A symmetric version of $h_i(v,q,\tilde q)$ would need to combine both of these groups: those harmed and those who gain by electing their chosen candidate under $q$ but not under the counterfactual $\tilde q$.
@@
 As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
-In general, we are interested in outcomes over multiple elections and a much larger set of counterfactual plans.
-Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans that are consistent with districting criteria.
-We could also consider averaging over a set of individuals with some shared characteristics.
+Researchers are generally interested in outcomes over multiple elections and a much larger set of counterfactual plans.
+<!-- Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans that are consistent with districting criteria.
+We could also consider averaging over a set of individuals with some shared characteristics. -->
 
-To allow for such averaging, we replace the fixed election $v$, counterfactual districting plan $\tilde q$, and individual voter $i$ with random variables $V\in\Elecs$, $\tilde Q\in\Plans$, and $I\in\Voters$.
-The random variable $V$ can be conceptualized as ranging over possible future elections---one realization could correspond to a heavily Democratic-favoring midterm year, while another could involve incumbents under-performing particularly badly.
-The random variable $\tilde Q$ captures the range of potential counterfactual districting plans.
+To allow for this broader scope, we replace the fixed election $v$, counterfactual districting plan $\tilde q$, and individual voter $i$ with random variables $V\in\Elecs$, $\tilde Q\in\Plans$, and $I\in\Voters$.
+The random variable $V$ represents a statistical model for possible future elections---one realization could correspond to a heavily Democratic-favoring midterm year, while another could involve incumbents under-performing particularly badly.
+The random variable $\tilde Q$ represents a model for potential counterfactual districting plans.
 Rarely would an analyst have a single counterfactual plan in mind: different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.

### 2025-08-25 paper/harm.qmd
edits from checking r1 comments

--- 2025-08-25 paper/harm.qmd
@@
 Rarely would an analyst have a single counterfactual plan in mind: different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
-Finally, $I$ is defined to be a voter taken uniformly at random from $\Voters$, with $I\indep V,\tilde Q$.
+Finally, $I$ is defined to be a voter taken uniformly at random from $\Voters$, with $I$ independent of $V,\tilde Q$.
 
@@
 
-Since none of the existing measures, other than partisan dislocation, are centered on individual voters, they offer far less granular insights than the proposed harm framework, even in the case of studying partisan gerrymandering.
+Since none of the existing measures, except partisan dislocation, are centered on individual voters, they offer far less granular insights than the proposed harm framework, even in the case of studying partisan gerrymandering.
 For example, differential harm can be estimated at the precinct level and then plotted on a map to show not just *which party* is harmed by a gerrymander, but *where* this harm occurs.
@@
 As the sign is positive for the simulations as well, this difference may be related to the fact that this measure uses the $k$ nearest neighbors, which is not uniform over the voters in the state.
-Implicitly, voters in dense, urban areas will be counterd at a higher rate than those in rural areas.
+Implicitly, voters in dense, urban areas will be counted at a higher rate than those in rural areas.
 

### 2025-08-25 paper/harm.qmd
typo
part discloc through sec 3

@@
 This unifies gerrymandering analyses across contexts and reflects the goal that elected officials represent the preferences of the voting public [see also @brunell2010].
-Few existing measures, with the exception of partisan disclocation [@defoeubarodd21], are focused on the invididual voter this way.
+Few existing measures are focused on individual voters this way.
+The exception is partisan dislocation [@defoeubarodd21], which compares the partisanship of a voter's district with the partisanship of their nearest neighbors.
 
@@
 
-Harm requires both terms to be true: the first measuring if a voter is not represented by their chosen candidate under $q$, and the second term measuring if they would have been represented by their chosen candidate under the counterfactual plan $\tilde q$.
+Harm involves two terms: the first measuring if a voter is not represented by their chosen candidate under $q$, and the second term measuring if they would have been represented by their chosen candidate under the counterfactual $\tilde q$.
+This comparison differs from the one that underlies partisan dislocation [@defoeubarodd21], where the comparison is against a group of neighboring voters rather than an alternative plan.
+We discuss this difference more in [Appendix @sec-connections], and compare the two measures empirically in @sec-nj.
 

### 2025-08-25 paper/harm.qmd
cuts

@@
 The random variable $\tilde Q$ represents a model for potential counterfactual districting plans.
-Rarely would an analyst have a single counterfactual plan in mind: different realizations of $\tilde Q$ correspond to different districting plans that are permissible under the relevant districting criteria.
-Finally, $I$ is defined to be a voter taken uniformly at random from $\Voters$, with $I$ independent of $V,\tilde Q$.
+Rarely would an analyst have a single counterfactual plan in mind: different realizations of $\tilde Q$ could correspond to different plans that respect the relevant districting criteria.
+Finally, $I$ is defined to be a voter taken uniformly at random from $\Voters$, independently of $V,\tilde Q$.
 
@@
 Single-member districts are often motivated by the value in representation to specific communities with shared interests.
-Preserving "communities of interest" is one of the most common statutory and constitutional requirements imposed by states on the plan-drawing process [@ncslcriteria].
+<!-- Preserving "communities of interest" is one of the most common statutory and constitutional requirements imposed by states on the plan-drawing process [@ncslcriteria]. -->
 It is therefore natural to consider not just the average harm to voters done by a plan, but also how this harm is distributed across certain groups of voters, defined by race, party, or other attributes.

### 2025-08-25 paper/harm.qmd
nj edits

@@
 
-To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a set of mutually exclusive groups of voters, so that for each voter $i$, we have $i\in G_k$ for some group $G_k$.
+To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a (random) partition of the set of voters.
 In general we will not require the group identities to be fixed with respect to elections.[^1]
@@
 
-Zero differential harm between groups is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and false negative rates across protected classes [@zafar2017fairness; @chouldechova2017fair].
+Zero differential harm between groups is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and negative rates across groups [@zafar2017fairness; @chouldechova2017fair].
 There, the implicit counterfactual is a perfect predictor.
-Compared to the perfect predictor, the actual decision process harms individuals whenever it makes a false positive or false negative error.
+Compared to the perfect predictor, the actual process harms individuals whenever it makes a false positive or negative error.
 Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.
@@
 @fig-nj-maps(d) shows one such plan.
-For each of these sampled plans and the two proposals, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats and several common partisan metrics.
+For each of these sampled plans and the two proposals, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats and several common partisan metrics, including partisan dislocation.
 These measures, and their distribution in the basline, are shown in @fig-nj-meas.
@@
 This means that, for example, an average Democratic voter is $3.6$pp less likely to be misrepresented than their average Republican counter-partisan.
-Overall, the Democratic plan has lower average harm.
+Overall, the Democratic plan has lower average harm, and a lower ranked marginal deviation [@herschlag2020quantifying].
 
@@
 
-Most of the other metrics, with the notable exceptions of the mean-median and partisan dislocation measure, agree with differential harm that each party's proposal favors itself.
-The efficiency gap and partisan bias estimate give roughly equal scores in absolute value to the Democratic and Republican plans, in comparison to differential harm and declination, which identify the Democratic proposal as slightly more extreme.
+Most of the other metrics agree with differential harm that each party's proposal favors itself.
+The efficiency gap and partisan bias score both plans almost equally in absolute value, in comparison to declination, which identifies the Republican proposal as slightly more extreme.
+However, the mean-median and partisan dislocation measures both classify the Democratic proposal as favoring Republicans, even though it produces more Democratic seats than the typical simulated plan.
 
-The raw metrics can be better contextualized in comparison to the distribution across the sampled plans, something we discuss in [Appendix @sec-app-nj].
+The raw metrics, and these two counterintuitive results, can be better contextualized in comparison to the distribution across the sampled plans, something we discuss in [Appendix @sec-app-nj].
 That appendix also compares the correlations between the various metrics in the simulated sample.

### 2025-08-25 paper/harm.qmd
push

@@
 More realistic models can add a significant burden to the calculation of our harm measures.
-And while modern simulation methods are a valuable starting point for defining counterfactual plans, these approaches involve analyst discretion and interpretation of often-murky legal standards.
+And while modern simulation methods are a valuable starting point for defining counterfactual plans, these approaches involve analyst discretion and interpretation of often-murky legal standards, and face their own computational challenges.
 Incorrect or incomplete interpretation of these standards can negatively influence the validity of the counterfactual plans.
@@
 Even when the choice of constraints maps onto legal requirements, it is important to ensure that the counterfactual plans are similar to the enacted plan on discretionary criteria as well.
-For example, a mapdrawer may weakly respect something like compactness or respecting county boundaries without being required to do so.
+For example, a mapdrawer may target a particular compactness metric, or draw districts that respect county boundaries, without being required to do so.
 In this case, if these optional constraints are not matched in the counterfactual, it will be unclear whether any estimated harm is due to *actual* biases in a plan or the mismatch in criteria.
+This includes the use of different compactness metrics in simulations than those used by the mapdrawer.
 As such, it is important to not just ensure that any counterfactual samples use methods with known sampling properties, but also that the sampled plans are similar to the enacted plan on these discretionary criteria.
@@
 
-A quite different approach to measuring partisan gerrymandering is the partisan dislocation [@defoeubarodd21].
-This measure uses a $k$-nearest-neighbor approach, where $k$ is the size of a district, to identify when a voter's district is politically different from their assigned district.
+A quite different approach to measuring partisan gerrymandering from symmetry-based methods is the partisan dislocation [@defoeubarodd21].
+This measure is based on a $k$-nearest-neighbor comparison, where $k$ is the size of a district, to identify when a voter's district is politically different from their assigned district.
 This can be repeated for  voter in a state or region, demonstrating at a local level when people been placed in a district which is atypical for who their nearest neighbors are.
-When aggregated to the precinct-level, this can produce local measures of partisan gerrymandering, similar to our proposed differential partisan harm measure.
+When aggregated to the precinct-level or beyond, this can produce local or statewide measures of partisan gerrymandering, similar to our proposed differential partisan harm measure.
+
+However, unlike the counterfactual _plan_ that is used in the harm framework, the comparison $k$-nearest-neighbors do not need to constitute a valid district.
+Moreover, harm considers the effect on election _outcomes_, while partisan dislocation measures changes in _vote share_.
+In our framework, if a Democrat wins in both the enacted and counterfactual plans, but by a smaller margin, there can be no harm.
+Partisan dislocation, in contrast, would measure the change in margin.
+This reflects the different quantities of interest: harm to individual voters, rather than changes in partisan configurations.
 
@@
 The partisan dislocation measure [@defoeubarodd21] shows that the Democratic plan is typical ($p=0.24$) while the Republican plan is an outlier ($p=0.01$).
-We use the vote-weighted measure of the average dislocation to produce a statewide summary.
-We find that the sign is incorrect for the Democratic plan, which has a positive sign.
-As the sign is positive for the simulations as well, this difference may be related to the fact that this measure uses the $k$ nearest neighbors, which is not uniform over the voters in the state.
-Implicitly, voters in dense, urban areas will be counted at a higher rate than those in rural areas.
+We average each voter's dislocation to produce the statewide summary value shown in the figure.
+We find that the sign is incorrect for the Democratic plan, which has positive dislocation (indicating an average shift towards Republicans).
+Notably, the sign is positive for the simulations as well, which suggests that such a statewide average of partisan dislocation may not reliably measure partisan bias.
+It is difficult to know exactly what causes this positive skew, but we suspect it is due to the interaction of the $k$-nearest neighbors comparison and the state's boundaries and varying population density.
 
@@
 In this analysis, the Democratic plan sits within the range of the sampled plans, while the Republican plan is an outlier ($p=0.0002$).
+This matches the distribution over Democratic seats, as well as average harm.
 

### 2025-08-25 paper/harm.qmd
add EI sentence, 5k words

@@
 As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
-Researchers are generally interested in outcomes over multiple elections and a much larger set of counterfactual plans.
+Researchers are generally interested in outcomes over multiple elections and a larger set of counterfactual plans.
 <!-- Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans that are consistent with districting criteria.
@@
 The randomness in $V$ and $\tilde Q$ allows us to compare the enacted districting plan to many plausible counterfactual alternatives, and across a range of electoral environments.
-At times we may wish to study on a particular election or a particular counterfactual plan.
+We may also wish to study a particular election or counterfactual plan.
 Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
@@
 To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a (random) partition of the set of voters.
-In general we will not require the group identities to be fixed with respect to elections.[^1]
+In general we do require the group identities to be fixed with respect to elections.[^1]
 This is because some group identities, like partisan preference, may themselves change election-to-election.
+Groups may need to be estimated with methods such as ecological inference [@king1997solution], as discussed in @sec-estimating.
 
@@
 Zero differential harm between groups is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and negative rates across groups [@zafar2017fairness; @chouldechova2017fair].
-There, the implicit counterfactual is a perfect predictor.
-Compared to the perfect predictor, the actual process harms individuals whenever it makes a false positive or negative error.
+There, the implicit counterfactual is a perfect predictor and the actual process harms individuals whenever it makes an incorrect prediction.
 Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.

### 2025-08-25 paper/harm.qmd
typos

@@
 
-We stress that the results here are tied the specific counterfactual plan (b).
+We stress that the results here are tied to the specific counterfactual plan (b).
 Indeed, this is a feature of the harm framework: different research questions may require different counterfactuals.
@@
 To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a (random) partition of the set of voters.
-In general we do require the group identities to be fixed with respect to elections.[^1]
-This is because some group identities, like partisan preference, may themselves change election-to-election.
-Groups may need to be estimated with methods such as ecological inference [@king1997solution], as discussed in @sec-estimating.
+In general we will not require the group identities to be fixed with respect to elections.[^1]
+This is because some identities, like partisan preference, may themselves change election-to-election.
 
@@
 So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
+The estimation of harm by group may require methods like ecological inference [@king1997solution], as discussed in @sec-estimating.
 
 Zero differential harm between groups is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and negative rates across groups [@zafar2017fairness; @chouldechova2017fair].
-There, the implicit counterfactual is a perfect predictor and the actual process harms individuals whenever it makes an incorrect prediction.
+There, the implicit counterfactual is a perfect predictor; the actual process harms individuals whenever it makes an incorrect prediction.
 Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.
@@
 For each of these sampled plans and the two proposals, we estimate the average harm and differential partisan harm versus the simulated baseline,[^9] and calculate the expected number of Democratic seats and several common partisan metrics, including partisan dislocation.
-These measures, and their distribution in the basline, are shown in @fig-nj-meas.
+These measures, and their distribution in the baseline, are shown in @fig-nj-meas.
 
@@
 
-One limitation of the harm definition itself is that considers only individuals who vote; abstainers experience zero harm.
+One limitation of the harm definition is that it considers only voters; abstainers experience zero harm.
 A richer framework and the use of auxiliary survey data might allow consideration of the informal preferences of the whole population, even when those preferences are not expressed at the ballot box.
@@
 where $\bar{s}_d$ is the baseline Democratic vote share in the districts, $\delta_v\sim\Norm(0, \sigma^2_\delta)$ is a election-specific uniform shift, and $\varepsilon_{dv}\sim\Norm(0, \sigma^2_\varepsilon)$ is a district-election residual error term.
-This can be thought of as a hiererchical extension of the classic uniform swing model.
+This can be thought of as a hierarchical extension of the classic uniform swing model.
 
@@
 This measure is based on a $k$-nearest-neighbor comparison, where $k$ is the size of a district, to identify when a voter's district is politically different from their assigned district.
-This can be repeated for  voter in a state or region, demonstrating at a local level when people been placed in a district which is atypical for who their nearest neighbors are.
+This can be repeated for each voter in a state or region, demonstrating at a local level when people have been placed in a district which is atypical for who their nearest neighbors are.
 When aggregated to the precinct-level or beyond, this can produce local or statewide measures of partisan gerrymandering, similar to our proposed differential partisan harm measure.
@@
 The efficiency gap is often highly correlated with estimates of differential partisan harm under a stochastic uniform partisan swing model (though the efficiency gap is much more discrete), as wasted votes and partisan harm are rooted in similar logic.[^6]
-Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates differential partisan harm as well.
+Similarly, in competitive states, where the baseline election has a statewide vote share near 50%, $\beta(\half)$ is highly correlated with the average number of seats won by each party and with estimates of differential partisan harm as well.
 Correlations, of course, are not everything: it is the universal scale and centering of our harm measure which makes it more useful in evaluating partisan gerrymanders, as we demonstrate in @sec-nj.
@@
 ```
-@fig-alabama-harm-sample summarizes the results, which are mostly similar to the results in in @sec-alabama.
+@fig-alabama-harm-sample summarizes the results, which are mostly similar to the results in @sec-alabama.
 Across the state, the estimated average harm is 6.9%, which translates to around 144,000 voters.

### 2025-08-25 paper/harm.qmd
edits

@@
 
-A key feature of our framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022].
+A core feature of our framework is its clear separation of theoretical quantities of interest and empirical strategies for estimating them [@katz2022].
 The methods section discusses the challenge of estimating harm and differential harm from observed data.
@@
 So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
-The estimation of harm by group may require methods like ecological inference [@king1997solution], as discussed in @sec-estimating.
+The estimation of harm by group may require methods like ecological inference, as discussed in the methods section.
 
@@
 
-The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-3.6\%$ for the Democratic proposal and $2.0\%$ for the Republican proposal.
+The estimate of differential partisan harm aligns with the expected partisan skew of each plan, with a value of $-3.6\%$ for the Democratic and $2.0\%$ for the Republican proposal.
 This means that, for example, an average Democratic voter is $3.6$pp less likely to be misrepresented than their average Republican counter-partisan.
